### 2020-07-01 paper.tex
template

\documentclass[twoside]{article}

\usepackage{aistats2020}
% If your paper is accepted, change the options for the package
% aistats2020 as follows:
%
% \usepackage[accepted]{aistats2020}
%
% This option will print headings for the title of your paper and
% headings for the authors names, plus a copyright note at the end of
% the first column of the first page.

% If you set papersize explicitly, activate the following three lines:
%\special{papersize = 8.5in, 11in}
%\setlength{\pdfpageheight}{11in}
%\setlength{\pdfpagewidth}{8.5in}

% If you use natbib package, activate the following three lines:
%\usepackage[round]{natbib}
%\renewcommand{\bibname}{References}
%\renewcommand{\bibsection}{\subsubsection*{\bibname}}

% If you use BibTeX in apalike style, activate the following line:
%\bibliographystyle{apalike}

\begin{document}

% If your paper is accepted and the title of your paper is very long,
% the style will print as headings an error message. Use the following
% command to supply a shorter title of your paper so that it can be
% used as headings.
%
%\runningtitle{I use this title instead because the last one was very long}

% If your paper is accepted and the number of authors is large, the
% style will print as headings an error message. Use the following
% command to supply a shorter version of the authors names so that
% they can be used as headings (for example, use only the surnames)
%
%\runningauthor{Surname 1, Surname 2, Surname 3, ...., Surname n}

\twocolumn[

\aistatstitle{Instructions for paper submissions to AISTATS 2020}

\aistatsauthor{ Author 1 \And Author 2 \And  Author 3 }

\aistatsaddress{ Institution 1 \And  Institution 2 \And Institution 3 } ]

\begin{abstract}
  The Abstract paragraph should be indented 0.25 inch (1.5 picas) on
  both left and right-hand margins. Use 10~point type, with a vertical
  spacing of 11~points. The \textbf{Abstract} heading must be centered,
  bold, and in point size 12. Two line spaces precede the
  Abstract. The Abstract must be limited to one paragraph.
\end{abstract}

\section{GENERAL FORMATTING INSTRUCTIONS}

The camera-ready versions of the accepted papers are 8 pages,
plus any additional pages needed for references.

Papers are in 2 columns with the overall line width of 6.75~inches (41~picas).
Each column is 3.25~inches wide (19.5~picas).  The space
between the columns is .25~inches wide (1.5~picas).  The left margin is 0.88~inches (5.28~picas).
Use 10~point type with a vertical spacing of
11~points. Please use US Letter size paper instead of A4.

Paper title is 16~point, caps/lc, bold, centered between 2~horizontal rules.
Top rule is 4~points thick and bottom rule is 1~point thick.
Allow 1/4~inch space above and below title to rules.

Author descriptions are center-justified, initial caps.  The lead
author is to be listed first (left-most), and the Co-authors are set
to follow.  If up to three authors, use a single row of author
descriptions, each one center-justified, and all set side by side;
with more authors or unusually long names or institutions, use more
rows.

Use one-half line space between paragraphs, with no indent.

\section{FIRST LEVEL HEADINGS}

First level headings are all caps, flush left, bold, and in point size
12. Use one line space before the first level heading and one-half line space
after the first level heading.

\subsection{Second Level Heading}

Second level headings are initial caps, flush left, bold, and in point
size 10. Use one line space before the second level heading and one-half line
space after the second level heading.

\subsubsection{Third Level Heading}

Third level headings are flush left, initial caps, bold, and in point
size 10. Use one line space before the third level heading and one-half line
space after the third level heading.

\paragraph{Fourth Level Heading}

Fourth level headings must be flush left, initial caps, bold, and
Roman type.  Use one line space before the fourth level heading, and
place the section text immediately after the heading with no line
break, but an 11 point horizontal space.

\subsection{CITATIONS, FIGURES, REFERENCES}

\subsubsection{Citations in Text}

Citations within the text should include the author's last name and
year, e.g., (Cheesman, 1985). Apart from including the author's last 
name and year, citations can follow any style, as long as the style 
is consistent throughout the paper.  Be sure that the sentence reads 
correctly if the citation is deleted: e.g., instead of ``As described 
by (Cheesman, 1985), we first frobulate the widgets,'' write ``As 
described by Cheesman (1985), we first frobulate the widgets.''  
%Be sure to avoid
%accidentally disclosing author identities through citations.

\subsubsection{Footnotes}

Indicate footnotes with a number\footnote{Sample of the first
  footnote.} in the text. Use 8 point type for footnotes. Place the
footnotes at the bottom of the column in which their markers appear,
continuing to the next column if required. Precede the footnote
section of a column with a 0.5 point horizontal rule 1~inch (6~picas)
long.\footnote{Sample of the second footnote.}

\subsubsection{Figures}

All artwork must be centered, neat, clean, and legible.  All lines
should be very dark for purposes of reproduction, and art work should
not be hand-drawn.  Figures may appear at the top of a column, at the
top of a page spanning multiple columns, inline within a column, or
with text wrapped around them, but the figure number and caption
always appear immediately below the figure.  Leave 2 line spaces
between the figure and the caption. The figure caption is initial caps
and each figure should be numbered consecutively.

Make sure that the figure caption does not get separated from the
figure. Leave extra white space at the bottom of the page rather than
splitting the figure and figure caption.
\begin{figure}[h]
\vspace{.3in}
\centerline{\fbox{This figure intentionally left non-blank}}
\vspace{.3in}
\caption{Sample Figure Caption}
\end{figure}

\subsubsection{Tables}

All tables must be centered, neat, clean, and legible. Do not use hand-drawn tables.
Table number and title always appear above the table.
See Table~\ref{sample-table}.

Use one line space before the table title, one line space after the table title,
and one line space after the table. The table title must be
initial caps and each table numbered consecutively.

\begin{table}[h]
\caption{Sample Table Title} \label{sample-table}
\begin{center}
\begin{tabular}{ll}
\textbf{PART}  &\textbf{DESCRIPTION} \\
\hline \\
Dendrite         &Input terminal \\
Axon             &Output terminal \\
Soma             &Cell body (contains cell nucleus) \\
\end{tabular}
\end{center}
\end{table}

\section{SUPPLEMENTARY MATERIAL}

If you need to include additional appendices during submission, you
can include them in the supplementary material file.

\section{INSTRUCTIONS FOR CAMERA-READY PAPERS}

For the camera-ready paper, if you are using \LaTeX, please make sure
that you follow these instructions.  (If you are not using \LaTeX,
please make sure to achieve the same effect using your chosen
typesetting package.)

\begin{enumerate}
    \item Download \texttt{fancyhdr.sty} -- the
    \texttt{aistats2020.sty} file will make use of it.
    \item Begin your document with
    \begin{flushleft}
    \texttt{\textbackslash documentclass[twoside]\{article\}}\\
    \texttt{\textbackslash usepackage[accepted]\{aistats2020\}}
    \end{flushleft}
    The \texttt{twoside} option for the class article allows the
    package \texttt{fancyhdr.sty} to include headings for even and odd
    numbered pages. The option \texttt{accepted} for the package
    \texttt{aistats2020.sty} will write a copyright notice at the end of
    the first column of the first page. This option will also print
    headings for the paper.  For the \emph{even} pages, the title of
    the paper will be used as heading and for \emph{odd} pages the
    author names will be used as heading.  If the title of the paper
    is too long or the number of authors is too large, the style will
    print a warning message as heading. If this happens additional
    commands can be used to place as headings shorter versions of the
    title and the author names. This is explained in the next point.
    \item  If you get warning messages as described above, then
    immediately after $\texttt{\textbackslash
    begin\{document\}}$, write
    \begin{flushleft}
    \texttt{\textbackslash runningtitle\{Provide here an alternative
    shorter version of the title of your paper\}}\\
    \texttt{\textbackslash runningauthor\{Provide here the surnames of
    the authors of your paper, all separated by commas\}}
    \end{flushleft}
    Note that the text that appears as argument in \texttt{\textbackslash
      runningtitle} will be printed as a heading in the \emph{even}
    pages. The text that appears as argument in \texttt{\textbackslash
      runningauthor} will be printed as a heading in the \emph{odd}
    pages.  If even the author surnames do not fit, it is acceptable
    to give a subset of author names followed by ``et al.''

    \item Use the file sample\_paper.tex as an example.

    \item The camera-ready versions of the accepted papers are 8
      pages, plus any additional pages needed for references.

    \item If you need to include additional appendices,
      you can include them in the supplementary
      material file.

    \item Please, don't change the layout given by the above
      instructions and by the style file.

\end{enumerate}

\subsubsection*{Acknowledgements}
All acknowledgments go at the end of the paper, including thanks to reviewers who gave useful comments, to colleagues who contributed to the ideas, and to funding agencies and corporate sponsors that provided financial support.

\subsubsection*{References}

References follow the acknowledgements.  Use an unnumbered third level
heading for the references section.  Please use the same font
size for references as for the body of the paper---remember that
references do not count against your page length total.

\begin{thebibliography}{}
\setlength{\itemindent}{-\leftmargin}
\makeatletter\renewcommand{\@biblabel}[1]{}\makeatother
\bibitem{} J.~Alspector, B.~Gupta, and R.~B.~Allen (1989).
    \newblock Performance of a stochastic learning microchip.
    \newblock In D. S. Touretzky (ed.),
    \textit{Advances in Neural Information Processing Systems 1}, 748--760.
    San Mateo, Calif.: Morgan Kaufmann.

\bibitem{} F.~Rosenblatt (1962).
    \newblock \textit{Principles of Neurodynamics.}
    \newblock Washington, D.C.: Spartan Books.

\bibitem{} G.~Tesauro (1989).
    \newblock Neurogammon wins computer Olympiad.
    \newblock \textit{Neural Computation} \textbf{1}(3):321--323.
\end{thebibliography}
\end{document}

### 2020-07-02 paper.tex
started formalism

@@
 
-\usepackage{aistats2020}
+%\usepackage{aistats2020}
 % If your paper is accepted, change the options for the package
@@
 %
-% \usepackage[accepted]{aistats2020}
+\usepackage[accepted]{aistats2020}
 %
@@
 % If you use natbib package, activate the following three lines:
-%\usepackage[round]{natbib}
-%\renewcommand{\bibname}{References}
-%\renewcommand{\bibsection}{\subsubsection*{\bibname}}
+\usepackage[round]{natbib}
+\renewcommand{\bibname}{References}
+\renewcommand{\bibsection}{\subsubsection*{\bibname}}
 
 % If you use BibTeX in apalike style, activate the following line:
-%\bibliographystyle{apalike}
+\bibliographystyle{apalike}
+
+% math and stuff
+\usepackage{amssymb}
+\usepackage{amsmath}
+\usepackage{amsthm}
+\usepackage{algorithmicx}
+\usepackage{algpseudocode}
+%\newcommand{\disc}{\mathrm{disc}}
+%\newcommand{\cont}{\mathrm{cont}}
+%\newcommand{\tr}{\mathtt{tr}}
+%\newcommand{\model}{\mathcal{P}}
+%\newcommand{\proposal}{\mathcal{Q}}
 
@@
 
-\aistatstitle{Instructions for paper submissions to AISTATS 2020}
+\aistatstitle{Automated Involutive MCMC}
 
-\aistatsauthor{ Author 1 \And Author 2 \And  Author 3 }
+\aistatsauthor{ Marco Cusumano-Towner \And Alexander K. Lew \And Vikash K. Mansinghka }
 
-\aistatsaddress{ Institution 1 \And  Institution 2 \And Institution 3 } ]
+\aistatsaddress{ Massachusetts Institute of Technology } ]
 
@@
 
-\section{GENERAL FORMATTING INSTRUCTIONS}
+\section{INTRODUCTION}
 
-The camera-ready versions of the accepted papers are 8 pages,
-plus any additional pages needed for references.
-
-Papers are in 2 columns with the overall line width of 6.75~inches (41~picas).
-Each column is 3.25~inches wide (19.5~picas).  The space
-between the columns is .25~inches wide (1.5~picas).  The left margin is 0.88~inches (5.28~picas).
-Use 10~point type with a vertical spacing of
-11~points. Please use US Letter size paper instead of A4.
-
-Paper title is 16~point, caps/lc, bold, centered between 2~horizontal rules.
-Top rule is 4~points thick and bottom rule is 1~point thick.
-Allow 1/4~inch space above and below title to rules.
-
-Author descriptions are center-justified, initial caps.  The lead
-author is to be listed first (left-most), and the Co-authors are set
-to follow.  If up to three authors, use a single row of author
-descriptions, each one center-justified, and all set side by side;
-with more authors or unusually long names or institutions, use more
-rows.
-
-Use one-half line space between paragraphs, with no indent.
-
-\section{FIRST LEVEL HEADINGS}
-
-First level headings are all caps, flush left, bold, and in point size
-12. Use one line space before the first level heading and one-half line space
-after the first level heading.
+This is the best paper~\citep{cusumano2019gen}.
 
@@
 
-Second level headings are initial caps, flush left, bold, and in point
-size 10. Use one line space before the second level heading and one-half line
-space after the second level heading.
-
 \subsubsection{Third Level Heading}
-
-Third level headings are flush left, initial caps, bold, and in point
-size 10. Use one line space before the third level heading and one-half line
-space after the third level heading.
 
@@
 
-Fourth level headings must be flush left, initial caps, bold, and
-Roman type.  Use one line space before the fourth level heading, and
-place the section text immediately after the heading with no line
-break, but an 11 point horizontal space.
+\section{BACKGROUND AND RELATED WORK}
 
-\subsection{CITATIONS, FIGURES, REFERENCES}
+\subsection{Traces}
 
-\subsubsection{Citations in Text}
+\section{INVOLUTIVE MCMC WITH TRACES}
 
-Citations within the text should include the author's last name and
-year, e.g., (Cheesman, 1985). Apart from including the author's last 
-name and year, citations can follow any style, as long as the style 
-is consistent throughout the paper.  Be sure that the sentence reads 
-correctly if the citation is deleted: e.g., instead of ``As described 
-by (Cheesman, 1985), we first frobulate the widgets,'' write ``As 
-described by Cheesman (1985), we first frobulate the widgets.''  
-%Be sure to avoid
-%accidentally disclosing author identities through citations.
+\begin{algorithmic}
+\Require Target distribution $p$, proposal distribution $q$, involution $h(u, t)$, previous state $x$.
+\Procedure{involutive-mcmc-move}{$p$, $q$, $f$, $x$}
+    \State $u \sim q(\cdot; x)$ \Comment{Sample auxiliary variables}
+    \State $(x', u') \gets f(x, u)$ \Comment{Apply involution}
+    \State $\alpha \gets
+        \frac{p(x') q(u'; x')}{p(x) q(u; x)} \left| \left[ \frac{\partial h(u, t)}{\partial (u, t)} \right] \right|$
+    \State $r \sim \mathrm{Uniform}(0, 1)$
+    \State \algorithmicif \, $r \le \alpha$ \algorithmicthen \, \Return $x'$ \algorithmicelse \, \Return $x$ 
+\EndProcedure
+\end{algorithmic}
 
-\subsubsection{Footnotes}
+% TODO PyTorch implementation??
 
-Indicate footnotes with a number\footnote{Sample of the first
-  footnote.} in the text. Use 8 point type for footnotes. Place the
-footnotes at the bottom of the column in which their markers appear,
-continuing to the next column if required. Precede the footnote
-section of a column with a 0.5 point horizontal rule 1~inch (6~picas)
-long.\footnote{Sample of the second footnote.}
+\section{AUTOMATIC ACCEPTANCE PROBABILITY CALCULATION}
 
-\subsubsection{Figures}
+\subsection{Exploiting Jacobian Sparsity}
 
-All artwork must be centered, neat, clean, and legible.  All lines
-should be very dark for purposes of reproduction, and art work should
-not be hand-drawn.  Figures may appear at the top of a column, at the
-top of a page spanning multiple columns, inline within a column, or
-with text wrapped around them, but the figure number and caption
-always appear immediately below the figure.  Leave 2 line spaces
-between the figure and the caption. The figure caption is initial caps
-and each figure should be numbered consecutively.
+\subsection{Exploiting Jacobian Sparsity}
 
-Make sure that the figure caption does not get separated from the
-figure. Leave extra white space at the bottom of the page rather than
-splitting the figure and figure caption.
-\begin{figure}[h]
-\vspace{.3in}
-\centerline{\fbox{This figure intentionally left non-blank}}
-\vspace{.3in}
-\caption{Sample Figure Caption}
-\end{figure}
+\subsection{Exploiting Jacobian Sparsity}
 
-\subsubsection{Tables}
+\section{A DYNAMIC CHECK FOR DETECTING BUGS}
 
-All tables must be centered, neat, clean, and legible. Do not use hand-drawn tables.
-Table number and title always appear above the table.
-See Table~\ref{sample-table}.
-
-Use one line space before the table title, one line space after the table title,
-and one line space after the table. The table title must be
-initial caps and each table numbered consecutively.
-
-\begin{table}[h]
-\caption{Sample Table Title} \label{sample-table}
-\begin{center}
-\begin{tabular}{ll}
-\textbf{PART}  &\textbf{DESCRIPTION} \\
-\hline \\
-Dendrite         &Input terminal \\
-Axon             &Output terminal \\
-Soma             &Cell body (contains cell nucleus) \\
-\end{tabular}
-\end{center}
-\end{table}
-
-\section{SUPPLEMENTARY MATERIAL}
-
-If you need to include additional appendices during submission, you
-can include them in the supplementary material file.
-
-\section{INSTRUCTIONS FOR CAMERA-READY PAPERS}
-
-For the camera-ready paper, if you are using \LaTeX, please make sure
-that you follow these instructions.  (If you are not using \LaTeX,
-please make sure to achieve the same effect using your chosen
-typesetting package.)
-
-\begin{enumerate}
-    \item Download \texttt{fancyhdr.sty} -- the
-    \texttt{aistats2020.sty} file will make use of it.
-    \item Begin your document with
-    \begin{flushleft}
-    \texttt{\textbackslash documentclass[twoside]\{article\}}\\
-    \texttt{\textbackslash usepackage[accepted]\{aistats2020\}}
-    \end{flushleft}
-    The \texttt{twoside} option for the class article allows the
-    package \texttt{fancyhdr.sty} to include headings for even and odd
-    numbered pages. The option \texttt{accepted} for the package
-    \texttt{aistats2020.sty} will write a copyright notice at the end of
-    the first column of the first page. This option will also print
-    headings for the paper.  For the \emph{even} pages, the title of
-    the paper will be used as heading and for \emph{odd} pages the
-    author names will be used as heading.  If the title of the paper
-    is too long or the number of authors is too large, the style will
-    print a warning message as heading. If this happens additional
-    commands can be used to place as headings shorter versions of the
-    title and the author names. This is explained in the next point.
-    \item  If you get warning messages as described above, then
-    immediately after $\texttt{\textbackslash
-    begin\{document\}}$, write
-    \begin{flushleft}
-    \texttt{\textbackslash runningtitle\{Provide here an alternative
-    shorter version of the title of your paper\}}\\
-    \texttt{\textbackslash runningauthor\{Provide here the surnames of
-    the authors of your paper, all separated by commas\}}
-    \end{flushleft}
-    Note that the text that appears as argument in \texttt{\textbackslash
-      runningtitle} will be printed as a heading in the \emph{even}
-    pages. The text that appears as argument in \texttt{\textbackslash
-      runningauthor} will be printed as a heading in the \emph{odd}
-    pages.  If even the author surnames do not fit, it is acceptable
-    to give a subset of author names followed by ``et al.''
-
-    \item Use the file sample\_paper.tex as an example.
-
-    \item The camera-ready versions of the accepted papers are 8
-      pages, plus any additional pages needed for references.
-
-    \item If you need to include additional appendices,
-      you can include them in the supplementary
-      material file.
-
-    \item Please, don't change the layout given by the above
-      instructions and by the style file.
-
-\end{enumerate}
+\section{EXAMPLES}
 
@@
 
-\subsubsection*{References}
+\bibliography{references}
 
-References follow the acknowledgements.  Use an unnumbered third level
-heading for the references section.  Please use the same font
-size for references as for the body of the paper---remember that
-references do not count against your page length total.
+\clearpage
+\onecolumn
+\section*{APPENDIX}
 
-\begin{thebibliography}{}
-\setlength{\itemindent}{-\leftmargin}
-\makeatletter\renewcommand{\@biblabel}[1]{}\makeatother
-\bibitem{} J.~Alspector, B.~Gupta, and R.~B.~Allen (1989).
-    \newblock Performance of a stochastic learning microchip.
-    \newblock In D. S. Touretzky (ed.),
-    \textit{Advances in Neural Information Processing Systems 1}, 748--760.
-    San Mateo, Calif.: Morgan Kaufmann.
+\subsection{Formalism}
+Let $(Z, \Sigma, \mu)$ be a measure space of pairs of traces of the model and proposal, where $\mu$ is $\sigma$-finite.
+Consider a $\mu$-measurable function $h : Z \to Z$ that is an involution ($h(h(z)) = z$) and let $\mu \circ h^{-1}$ denote the pushforward of $\mu$ under $h$, such that $\mu \circ h^{-1}$ is $\sigma$-finite and such that $\mu \circ h^{-1}$ is absolutely continuous with respect to $\mu$.
+% TODO can we simplify the conditions, using the fact that h is an involution?
+Then, the Radon-Nikodym derivative exists, and is denoted $(d (\mu \circ h^{-1}) / d \mu)(z)$.
 
-\bibitem{} F.~Rosenblatt (1962).
-    \newblock \textit{Principles of Neurodynamics.}
-    \newblock Washington, D.C.: Spartan Books.
+The goal is to prove stationarity with respect to the measure defined by density $p(z)$ with respect to $\mu$.
+The move starts with state $z \in Z$, applies the involution $h$, and then accepts (returns $h(z)$ with probability $\alpha(z)$) and otherwise returns $z$.
+Let $k_{z}$ denote the measure on new states induced by the involution $h$ and the previous state $z$:
+\begin{equation}
+k_{z}(A) = [h(z) \in A] \alpha(z) + [z \in A] (1 - \alpha(z))
+\end{equation}
+The stationarity condition is:
+\begin{equation}
+\int_Z p(z) k_{z}(A) \mu(dz) = p(A) = \int_Z p(z) [z \in A] \mu(dz) \mbox{ for all } A \in \Sigma
+\end{equation}
+Substituting in the definition of $k_z(A)$ and simplifying gives:
+\begin{equation} \label{eq:stationarity-requirement}
+\int_Z p(z) [h(z) \in A] \alpha(z) \mu(dz) = \int_Z p(z) [z \in A] \alpha(z) \mu(dz) \mbox{ for all } A \in \Sigma
+\end{equation}
+Note that for a $\mu$-measurable function $g$ such that $g \circ h$ is integrable with respect to $\mu$:
+\begin{equation} \label{eq:pushforward}
+\int_Z g(h(z)) \mu(dz) = \int_Z g(z) (\mu \circ h^{-1})(dz) = \int_Z g(z) \left( \frac{d (\mu \circ h^{-1})}{d \mu}(z) \right) \mu(dz)
+\end{equation}
+For $g(z) := p(h(z)) [z \in A] \alpha(h(z))$ we have $g(h(z)) = p(z) [h(z) \in A] \alpha(z)$, which is the integrand of the left-hand-side of Equation~(\ref{eq:stationarity-requirement}).
+Applying Equation~(\ref{eq:pushforward}) to the left-hand side of Equation~(\ref{eq:stationarity-requirement}) gives:
+\begin{equation}
+\int_Z p(z) [h(z) \in A] \alpha(z) \mu(dz) = \int_Z g(z) \left( \frac{d (\mu \circ h^{-1})}{d \mu}(z) \right) \mu(dz) = \int_Z p(h(z)) [z \in A] \alpha(h(z)) \left( \frac{d (\mu \circ h^{-1})}{d \mu}(z) \right) \mu(dz)
+\end{equation}
+Equating this with the right-hand side of Equation~(\ref{eq:stationarity-requirement}) gives the following sufficient condition for stationarity:
+\begin{equation}
+\int_Z p(h(z)) [z \in A] \alpha(h(z)) \left( \frac{d (\mu \circ h^{-1})}{d \mu}(z) \right) \mu(dz) = \int_Z p(z) [z \in A] \mu(dz) \mbox{ for all } A \in \Sigma
+\end{equation}
+It therefore suffices to find $\alpha$ such that:
+\begin{equation}
+p(h(z)) [z \in A] \alpha(h(z)) \left( \frac{d (\mu \circ h^{-1})}{d \mu}(z) \right) = p(z) [z \in A] \alpha(z) \mbox{ for all } z \in Z
+\end{equation}
+The following choice of $\alpha$ suffices:
+\begin{equation}
+\alpha(z) = \min\left\{ 1, \frac{p(h(z))}{p(z)} \left( \frac{d (\mu \circ h^{-1})}{d \mu}(z) \right) \right\}
+\end{equation}
 
-\bibitem{} G.~Tesauro (1989).
-    \newblock Neurogammon wins computer Olympiad.
-    \newblock \textit{Neural Computation} \textbf{1}(3):321--323.
-\end{thebibliography}
+All requirements:
+\begin{enumerate}
+\item $\mu$ is $\sigma$-finite
+\item $h$ is $\mu$-measurable
+\item $h$ is an involution
+\item the pushforward $\mu \circ h^{-1}$ is $\sigma$-finite
+\item the pushforward $\mu \circ h^{-1}$ is absolutely continuous with respect to $\mu$
+\item the function $g(z) := p(h(z)) [z \in A] \alpha(h(z))$ is $\mu$-measurable for all $A \in \Sigma$
+\item the function $(g \circ h)(z) = p(z) [h(z) \in A] \alpha(z)$ is integrable with respect to $\mu$ for all $A \in \Sigma$
+\end{enumerate}
+
 \end{document}

### 2020-07-02 build.sh

@@ -1,0 +1,9 @@
+#!/bin/bash
+set -u
+set -e
+set -x
+
+pdflatex paper
+bibtex paper
+pdflatex paper
+pdflatex paper

### 2020-07-02 clean.sh

@@ -1,0 +1,5 @@
+#!/bin/bash
+set -u
+set -e
+
+rm -i paper.bbl paper.blg paper.log paper.aux

### 2020-07-02 paper.tex
draft abstract (too long)

@@
 \bibliographystyle{apalike}
+
+% graphics
+\usepackage{float}
+\usepackage{graphicx}
 
@@
 \begin{abstract}
-  The Abstract paragraph should be indented 0.25 inch (1.5 picas) on
-  both left and right-hand margins. Use 10~point type, with a vertical
-  spacing of 11~points. The \textbf{Abstract} heading must be centered,
-  bold, and in point size 12. Two line spaces precede the
-  Abstract. The Abstract must be limited to one paragraph.
+Involutive MCMC is a unifying framework for MCMC that encompasses many standard
+and recent MCMC algorithms from the literature, from reversible jump MCMC to
+Hamiltonian Monte Carlo to kernels based on deep neural networks. The key idea
+in involutive MCMC is that a combination of (i) an auxiliary probability
+distribution and (ii) an involution on the extended state space that includes
+the model latent variables and the auxiliary variables results in MCMC kernels
+that satisfy detailed balance with respect to the target distribution. The
+involutive MCMC framework is appealing for its simplicity and generality, and
+promises to be a useful conceptual and mathematical tool for exploring the
+design space of MCMC kernels. However, like other Monte Carlo samplers,
+implementing involutive samplers is time consuming and error prone. This paper
+describes a technique for automatically generating the implementation of
+involutive MCMC samplers from two probabilistic programs that define the target
+distribution and the auxiliary probability distribution respectively, and a
+differentiable program that defines the involution. The technique also detects
+common conceptual and programming errors that arise when designing and
+specifying involutive MCMC algorithms. This paper describes the support for
+involutive MCMC in the Gen probabilistic system that uses this technique, and
+shows involutive samplers specified using Gen's high-level probabilistic and
+differentiable programming languages.
 \end{abstract}
@@
 
-This is the best paper~\citep{cusumano2019gen}.
+%This is the best paper~\citep{cusumano2019gen}.
 
@@
 \subsection{Traces}
+
+\input{headline-figure}
 
@@
 
-\subsection{Exploiting Jacobian Sparsity}
+% TODO can only show this for a special case -- discrete involution + continuous?, or some more general case?
 
-\subsection{Exploiting Jacobian Sparsity}
+% TODO show an performance table of moves per second with and without the sparsity reduction (?)
 
 \section{A DYNAMIC CHECK FOR DETECTING BUGS}
+
+% TODO show an example of the type of bug it can detect (an edge case?)
 
@@
 
-\subsection{Formalism}
+\subsection{Distributions on traces}
+Define target density $\pi$ (XXX below it is called $p$, fix this) on the joint space $X \times U$, where $X$ are traces of the model and $X$ are traces of the proposal, in terms of the density $p(x)$ and the family of densities $q_x(u)$ as $\pi(x, u) := p(x) q_u(x)$.
+
+% TODO rename the joint target density to \pi below
+We want to show stationarity with respect to $\pi$, using stationarity of the involution with respect to the joint density (below).
+\begin{equation}
+\int_X p(x) \ell_x(A) \mu(dx) = p(A) \mbox{ for all } A \in \Sigma
+\end{equation}
+where $\ell_x(A)$ is the probability of transitioning into set $A$ from $x$, and is given by:
+\begin{equation}
+\ell_x(A) := \int_U k_{(x, u)}(A \times U) \mu(du)
+\end{equation}
+where $k_{(x, u)}$ is the measure on $X \times U$ that is induced by applying the involution $h$ to the state $(x, u)$.
+% TODO checkme
+
+\subsection{Stationarity of the involution with respect to joint density}
 Let $(Z, \Sigma, \mu)$ be a measure space of pairs of traces of the model and proposal, where $\mu$ is $\sigma$-finite.
@@
 The goal is to prove stationarity with respect to the measure defined by density $p(z)$ with respect to $\mu$.
-The move starts with state $z \in Z$, applies the involution $h$, and then accepts (returns $h(z)$ with probability $\alpha(z)$) and otherwise returns $z$.
+The move starts with state $z \in Z$, applies the involution $h$, and then accepts (returns $h(z)$) with probability $\alpha(z)$ and otherwise returns $z$.
 Let $k_{z}$ denote the measure on new states induced by the involution $h$ and the previous state $z$:
@@
 
+\paragraph{Characterizing the pushforward $\mu \circ h^{-1}$ for involution $h$}
+Can we guarantee that it is $\sigma$-finite or absolutely continuous with respect to $\mu$?
+Are there simpler sufficient conditions we can give?
+$(\mu \circ h^{-1})$ is $\sigma$-finite if $Z$ is the countable union of $(mu \circ h^{-1})$-measurable sets with finite measure.
+
+$\sigma$-finite means
+
 \end{document}

### 2020-07-06 experiments/mixture.jl
split merge example checkpoint

@@ -1,0 +1,99 @@
+# univariate mixture model from richardson and green
+# https://people.maths.bris.ac.uk/~mapjg/papers/RichardsonGreenRSSB.pdf
+
+# modifications:
+# - no upper bound to k (kmax) -- why did they do this??
+# - only implement split/merge moves, not other moves..
+# - instead of restricting the means to be ordered, add explicit swap moves
+#   that permute the indices
+# - no hyperpriors
+
+# just test the split/merge move, from a reasonably chosen initial state, show
+# that it accepts and works
+
+using Gen
+using PyPlot
+import Random
+
+include("dirichlet.jl")
+
+@gen function model(n::Int)
+    k ~ poisson(2)
+    means = Vector{Float64}(undef, k)
+    vars = Vector{Float64}(undef, k)
+    for j in 1:k
+        means[j] = ({(:mu, j)} ~ normal(0, 10))
+        vars[j] = ({(:var, j)} ~ inv_gamma(1, 1))
+    end
+    probs ~ dirichlet(fill(2, k))
+    for i in 1:n
+        z = ({(:z, i)} ~ categorical(probs))
+        {(:x, i)} ~ normal(means[z], vars[z])
+    end
+end
+
+function get_n(trace)
+    return get_args(trace)[1]
+end
+
+function get_means(trace)
+    k = trace[:k]
+    return [trace[(:mu, i)] for i in 1:k]
+end
+
+function get_vars(trace)
+    k = trace[:k]
+    return [trace[(:var, i)] for i in 1:k]
+end
+
+function get_xs(trace)
+    n = get_n(trace)
+    return [trace[(:x, i)] for i in 1:n]
+end
+
+function marginal_density(k, probs, means, vars, x)
+    ls = zeros(k)
+    for j in 1:k
+        mu = means[j]
+        var = vars[j]
+        ls[j] = logpdf(normal, x, mu, var) + log(probs[j])
+    end
+    return exp(logsumexp(ls))
+end
+
+function get_densities_at(trace, xs)
+    k = trace[:k]
+    probs = trace[:probs]
+    means = get_means(trace)
+    vars = get_vars(trace)
+    return [marginal_density(k, probs, means, vars, x) for x in xs]
+end
+
+# simulate data and plot a histogram..
+function show_prior_data()
+    Random.seed!(3)
+
+    n = 1000
+    trace = simulate(model, (n,))
+    figure()
+    xmin = -30.0
[…]

### 2020-07-06 experiments/mixture.jl
richardson green split merge checkpoint

@@ -28,7 +28,7 @@ include("dirichlet.jl")
     probs ~ dirichlet(fill(2, k))
     for i in 1:n
         z = ({(:z, i)} ~ categorical(probs))
-        {(:x, i)} ~ normal(means[z], vars[z])
+        {(:x, i)} ~ normal(means[z], sqrt(vars[z]))
     end
 end
 
@@ -56,7 +56,7 @@ function marginal_density(k, probs, means, vars, x)
     for j in 1:k
         mu = means[j]
         var = vars[j]
-        ls[j] = logpdf(normal, x, mu, var) + log(probs[j])
+        ls[j] = logpdf(normal, x, mu, sqrt(var)) + log(probs[j])
     end
     return exp(logsumexp(ls))
 end
@@ -69,31 +69,123 @@ function get_densities_at(trace, xs)
     return [marginal_density(k, probs, means, vars, x) for x in xs]
 end
 
-# simulate data and plot a histogram..
-function show_prior_data()
-    Random.seed!(3)
-
-    n = 1000
-    trace = simulate(model, (n,))
-    figure()
-    xmin = -30.0
-    xmax = 30.0
-
+function render_trace(trace, xmin, xmax)
     # histogram
-    #subplot(1, 2, 1)
     xs = get_xs(trace)
     (hist_data, ) = hist(xs, bins=collect(range(xmin, stop=xmax, length=50)))
     (ymin, ymax) = gca().get_ylim()
 
     # density plot
-    #subplot(1, 2, 2)
     test_xs = collect(range(xmin, stop=xmax, length=1000))
     densities = get_densities_at(trace, test_xs)
     max_density = maximum(densities)
     scale = ymax / max_density
     plot(test_xs, densities * scale, color="orange")
+end
 
+# simulate data and plot a histogram..
+function show_prior_data()
+    Random.seed!(3)
+    n = 1000
+    trace = simulate(model, (n,))
+    figure()
+    xmin = -30.0
+    xmax = 30.0
+    render_trace(trace, xmin, xmax)
     savefig("prior_sample.png")
 end
 
 show_prior_data()
+
+function generate_synthetic_two_mixture_data()
+    Random.seed!(1)
+    n = 1000
+    constraints = choicemap()
+    constraints[:k] = 2
+    constraints[:probs] = [0.5, 0.5]
+    constraints[(:mu, 1)] = -10.0
[…]

### 2020-07-06 experiments/mixture.jl
collapsed mixture of normals

@@ -16,19 +16,15 @@ using PyPlot
 import Random
 
 include("dirichlet.jl")
+include("mixture_of_normals.jl")
 
 @gen function model(n::Int)
     k ~ poisson(2)
-    means = Vector{Float64}(undef, k)
-    vars = Vector{Float64}(undef, k)
-    for j in 1:k
-        means[j] = ({(:mu, j)} ~ normal(0, 10))
-        vars[j] = ({(:var, j)} ~ inv_gamma(1, 1))
-    end
-    probs ~ dirichlet(fill(2, k))
+    means = [({(:mu, j)} ~ normal(0, 10)) for j in 1:k]
+    vars = [({(:var, j)} ~ inv_gamma(1, 1)) for j in 1:k]
+    weights ~ dirichlet([2.0 for j in 1:k])
     for i in 1:n
-        z = ({(:z, i)} ~ categorical(probs))
-        {(:x, i)} ~ normal(means[z], sqrt(vars[z]))
+        {(:x, i)} ~ mixture_of_normals(weights, means, vars)
     end
 end
 
@@ -51,22 +47,22 @@ function get_xs(trace)
     return [trace[(:x, i)] for i in 1:n]
 end
 
-function marginal_density(k, probs, means, vars, x)
+function marginal_density(k, weights, means, vars, x)
     ls = zeros(k)
     for j in 1:k
         mu = means[j]
         var = vars[j]
-        ls[j] = logpdf(normal, x, mu, sqrt(var)) + log(probs[j])
+        ls[j] = logpdf(normal, x, mu, sqrt(var)) + log(weights[j])
     end
     return exp(logsumexp(ls))
 end
 
 function get_densities_at(trace, xs)
     k = trace[:k]
-    probs = trace[:probs]
+    weights = trace[:weights]
     means = get_means(trace)
     vars = get_vars(trace)
-    return [marginal_density(k, probs, means, vars, x) for x in xs]
+    return [marginal_density(k, weights, means, vars, x) for x in xs]
 end
 
 function render_trace(trace, xmin, xmax)
@@ -102,7 +98,7 @@ function generate_synthetic_two_mixture_data()
     n = 1000
     constraints = choicemap()
     constraints[:k] = 2
-    constraints[:probs] = [0.5, 0.5]
+    constraints[:weights] = [0.5, 0.5]
     constraints[(:mu, 1)] = -10.0
     constraints[(:mu, 2)] = 10.0
[…]

### 2020-07-06 experiments/mixture.jl
runs passes check

@@ -91,7 +91,7 @@ function show_prior_data()
     savefig("prior_sample.png")
 end
 
-show_prior_data()
+#show_prior_data()
 
 function generate_synthetic_two_mixture_data()
     Random.seed!(1)
@@ -109,9 +109,10 @@ function generate_synthetic_two_mixture_data()
     xmax = 30.0
     render_trace(trace, xmin, xmax)
     savefig("synthetic_data.png")
+    return trace
 end
 
-generate_synthetic_two_mixture_data()
+#generate_synthetic_two_mixture_data()
 
 function merge_weights(weights, j, k)
     w1 = weights[j]
@@ -132,6 +133,7 @@ function merge_mean_and_var(mu1, mu2, var1, var2, w1, w2, w)
 end
 
 function split_weights(weights, j, u1, k)
+    w = weights[j]
     w1 = w * u1
     w2 = w * (1 - u1)
     new_weights = [(i == j) ? w1 : (i == k + 1) ? w2 : weights[i] for i in 1:k+1]
@@ -154,31 +156,34 @@ end
 @gen function split_merge_proposal(trace)
     # decide whether to split or merge
     k = trace[:k]
-    if k > 0
+    if k > 1
         split ~ bernoulli(0.5)
     else
         split = true
     end
     if split
         # if split, pick random to split
-        j = uniform_discrete(1, k)
+        j ~ uniform_discrete(1, k)
         # then pick DoFs
         u1 ~ beta(2, 2)
         u2 ~ beta(2, 2)
         u3 ~ beta(1, 1)
     else
         # if merge, then pick two to merge
-        j1 = uniform_discrete(1, k-1)
+        @assert k > 1
+        j ~ uniform_discrete(1, k-1)
     end
     return split
 end
 
 @bijection function split_merge_inv(
         model_args, proposal_args, proposal_retval)
+    split = proposal_retval
     k = @read_discrete_from_model(:k)
     if split
 
         # split
+        println("splitting, starting from k=$k")
 
         j = @read_discrete_from_proposal(:j) # the cluster to split
         u1 = @read_continuous_from_proposal(:u1)
@@ -188,36 +193,44 @@ end
         mu = @read_continuous_from_model((:mu, j))
         var = @read_continuous_from_model((:var, j))
 
-        new_weights = split_weights(weights, j, u1)
+        new_weights = split_weights(weights, j, u1, k)
         (mu1, mu2) = split_means(mu, var, u2, new_weights[j], new_weights[k+1])
[…]

### 2020-07-06 experiments/mixture.jl
add example merge/split fig

@@ -18,10 +18,12 @@ import Random
 include("dirichlet.jl")
 include("mixture_of_normals.jl")
 
+@dist poisson_plus_one(rate) = poisson(rate) + 1
+
 @gen function model(n::Int)
-    k ~ poisson(2)
+    k ~ poisson_plus_one(1)
     means = [({(:mu, j)} ~ normal(0, 10)) for j in 1:k]
-    vars = [({(:var, j)} ~ inv_gamma(1, 1)) for j in 1:k]
+    vars = [({(:var, j)} ~ inv_gamma(1, 10)) for j in 1:k]
     weights ~ dirichlet([2.0 for j in 1:k])
     for i in 1:n
         {(:x, i)} ~ mixture_of_normals(weights, means, vars)
@@ -68,7 +70,7 @@ end
 function render_trace(trace, xmin, xmax)
     # histogram
     xs = get_xs(trace)
-    (hist_data, ) = hist(xs, bins=collect(range(xmin, stop=xmax, length=50)))
+    (hist_data, ) = hist(xs, bins=collect(range(xmin, stop=xmax, length=50)), color="gray")
     (ymin, ymax) = gca().get_ylim()
 
     # density plot
@@ -76,13 +78,23 @@ function render_trace(trace, xmin, xmax)
     densities = get_densities_at(trace, test_xs)
     max_density = maximum(densities)
     scale = ymax / max_density
-    plot(test_xs, densities * scale, color="orange")
+    plot(test_xs, densities * scale, color="orange", linewidth=4, zorder=1)
+    
+    # individual component density plot
+    #if trace[:k] > 1
+        for j in 1:trace[:k]
+            densities = [marginal_density(1, [1.0], [trace[(:mu, j)]], [trace[(:var, j)]], x) for x in test_xs]
+            max_density = maximum(densities)
+            scale = ymax / max_density
+            plot(test_xs, densities * scale, color="red", linewidth=2, zorder=2)
+        end
+    #end
 end
 
 # simulate data and plot a histogram..
 function show_prior_data()
     Random.seed!(3)
-    n = 1000
+    n = 100
     trace = simulate(model, (n,))
     figure()
     xmin = -30.0
@@ -95,20 +107,20 @@ end
 
 function generate_synthetic_two_mixture_data()
     Random.seed!(1)
-    n = 1000
+    n = 100
     constraints = choicemap()
     constraints[:k] = 2
     constraints[:weights] = [0.5, 0.5]
[…]

### 2020-07-06 experiments/mixture.jl
beautify code for fig

@@ -159,92 +159,82 @@ function split_means(mu, var, u2, w1, w2)
     return (mu1, mu2)
 end
 
-function split_vars(w, w1, w2, var, u2, u3)
-    var1 = u3 * (1 - u2^2) * var * w / w1
-    var2 = (1 - u3) * (1 - u2^2) * var * w / w2
+function split_vars(w1, w2, var, u2, u3)
+    var1 = u3 * (1 - u2^2) * var * (w1 + w2) / w1
+    var2 = (1 - u3) * (1 - u2^2) * var * (w1 + w2) / w2
     return (var1, var2)
 end
 
 @gen function split_merge_proposal(trace)
-    # decide whether to split or merge
-    k = trace[:k]
-    if k > 1
-        split ~ bernoulli(0.5)
-    else
-        split = true
-    end
-    if split
-        # if split, pick random to split
-        j ~ uniform_discrete(1, k)
-        # then pick DoFs
-        u1 ~ beta(2, 2)
-        u2 ~ beta(2, 2)
-        u3 ~ beta(1, 1)
-    else
-        # if merge, then pick two to merge
-        @assert k > 1
-        j ~ uniform_discrete(1, k-1)
-    end
-    return split
+   k = trace[:k]
+   split = (k == 1) ? true : ({:split} ~ bernoulli(0.5))
+   if split
+      # split; pick cluster to split and sample degrees of freedom
+      cluster_to_split ~ uniform_discrete(1, k)
+      u1 ~ beta(2, 2)
+      u2 ~ beta(2, 2)
+      u3 ~ beta(1, 1)
+   else
+      # merge; pick cluster to merge with last cluster
+      cluster_to_merge ~ uniform_discrete(1, k-1)
+   end
 end
 
-@bijection function split_merge_inv(
-        model_args, proposal_args, proposal_retval)
-    split = proposal_retval
+@bijection function split_merge_inv(_, _, _)
     k = @read_discrete_from_model(:k)
+    split = (k == 1) ? true : @read_discrete_from_proposal(:split)
     if split
 
-        # split
-        #println("splitting, starting from k=$k")
-
-        j = @read_discrete_from_proposal(:j) # the cluster to split
+        cluster_to_split = @read_discrete_from_proposal(:cluster_to_split)
         u1 = @read_continuous_from_proposal(:u1)
         u2 = @read_continuous_from_proposal(:u2)
         u3 = @read_continuous_from_proposal(:u3)
         weights = @read_continuous_from_model(:weights)
[…]

### 2020-07-06 experiments/mixture.jl
make it into a 3-clusdter example

@@ -70,7 +70,7 @@ end
 function render_trace(trace, xmin, xmax)
     # histogram
     xs = get_xs(trace)
-    (hist_data, ) = hist(xs, bins=collect(range(xmin, stop=xmax, length=50)), color="gray")
+    (hist_data, ) = hist(xs, bins=collect(range(xmin, stop=xmax, length=50)), color="LightGray")
     (ymin, ymax) = gca().get_ylim()
 
     # density plot
@@ -78,15 +78,16 @@ function render_trace(trace, xmin, xmax)
     densities = get_densities_at(trace, test_xs)
     max_density = maximum(densities)
     scale = ymax / max_density
-    plot(test_xs, densities * scale, color="orange", linewidth=4, zorder=1)
+    plot(test_xs, densities * scale, color="black", linewidth=4, zorder=1)
     
     # individual component density plot
+    colors = ["red", "orange", "blue"]
     #if trace[:k] > 1
         for j in 1:trace[:k]
-            densities = [marginal_density(1, [1.0], [trace[(:mu, j)]], [trace[(:var, j)]], x) for x in test_xs]
+            densities = [marginal_density(1, [1.0], [trace[(:mu, j)]], [trace[(:var, j)]], x) for x in test_xs] * trace[:weights][j]
             max_density = maximum(densities)
-            scale = ymax / max_density
-            plot(test_xs, densities * scale, color="red", linewidth=2, zorder=2)
+            #scale = ymax / max_density
+            plot(test_xs, densities * scale, color=colors[j], linewidth=2, zorder=2)
         end
     #end
 end
@@ -109,12 +110,14 @@ function generate_synthetic_two_mixture_data()
     Random.seed!(1)
     n = 100
     constraints = choicemap()
-    constraints[:k] = 2
-    constraints[:weights] = [0.5, 0.5]
+    constraints[:k] = 3
+    constraints[:weights] = [0.4, 0.4, 0.2]
     constraints[(:mu, 1)] = -10.0
     constraints[(:mu, 2)] = 10.0
+    constraints[(:mu, 3)] = 20.0
     constraints[(:var, 1)] = 50.0
     constraints[(:var, 2)] = 50.0
+    constraints[(:var, 3)] = 50.0
     trace, = generate(model, (n,), constraints)
     #figure()
     #xmin = -30.0
@@ -247,31 +250,49 @@ end
 
 function test_split_merge_move()
[…]

### 2020-07-08 paper.tex
add figs update code

@@
 
-\input{headline-figure}
+\begin{figure*}[ht]
+    \centering
+    \includegraphics[width=\textwidth]{figures/mixture.pdf}
+    \caption{
+Example of reversible jump MCMC~\citep{green1995reversible} implemented using involutive MCMC in Gen.
+The example implements a `split-merge move' in a infinite Gaussian mixture model~\citep{richardson1997bayesian} using three Gen programs:
+(1) a probabilistic program $\mathtt{p}$ encoding the generative model (shown in b),
+(2) a probabilistic program $\mathtt{q}$ encoding an auxiliary probability distribution (shown in c),
+and (3) a differentiable program $\mathtt{h}$ that encodes an involution on the space of pairs of traces of $\mathtt{p}$ and $\mathtt{q}$ (shown in d).
+Gen's involutive MCMC operator (shown in e) automatically computes the acceptance probability.
+}
+    \label{fig:mixture}
+\end{figure*}
+
+\begin{figure*}[ht]
+    \centering
+    \includegraphics[width=\textwidth]{figures/structure-learning.pdf}
+    \caption{
+A mixture kernel implemented using involutive MCMC in Gen, applied to infer the covariance function of a Gaussian process.
+The prior on covariance functions is based on a probabilistic context-free grammar.
+Each component kernel in the mixture replaces a subtree of the covariance function parse tree with a new subtree.
+The mixture kernel chooses a random subtree to replace via a random walk on the parse tree.
+The mixture kernel is composed from three Gen programs:
+(1) a probabilistic program $\mathtt{p}$ encoding the generative model (shown in a),
+(2) a probabilistic program $\mathtt{q}$ encoding an auxiliary probability distribution (shown in b), and
+(3) a differentiable program $\mathtt{h}$ that encodes an involution (shown in d).
+}
+    \label{fig:structure-learning}
+\end{figure*}
 

### 2020-07-08 experiments/gp_data.jl

@@ -1,0 +1,42 @@
+using Gen
+using PyPlot
+using LinearAlgebra: Symmetric, I
+
+function cov_fn(x, y)
+    return 0.452 + exp(-(x-y)^2/0.132) * (x - 0.085) * (y - 0.085)
+end
+
+function make_cov_matrix(xs)
+    n = length(xs)
+    mat = zeros(n, n)
+    for i in 1:n
+        for j in 1:n
+            mat[i, j] = cov_fn(xs[i], xs[j])
+        end
+        #mat[i, i] += 0.1
+    end
+    return mat
+end
+
+#xs = collect(range(-1, stop=1, length=30))
+xs = collect(range(-1, stop=1, length=100))
+println(xs)
+cov = make_cov_matrix(xs)
+println(cov)
+
+eye(n) = Matrix{Float64}(I, n, n)
+
+figure(figsize=(4, 2))
+for i in 1:10
+    ys = mvnormal(zeros(length(xs)), Symmetric(cov) .+ 0.001 * eye(length(xs)))
+    plot(ys, color="black", alpha=0.5)
+end
+yticks([-2, 0, 2])
+xlabel("x")
+ylabel("y")
+tight_layout()
+savefig("gp_samples.png", dpi=300)
+
+#####
+
+# just rescale the mixture weights in a 1D finite mixture model

### 2020-07-08 experiments/mixture.jl

@@ -67,7 +67,7 @@ function get_densities_at(trace, xs)
     return [marginal_density(k, weights, means, vars, x) for x in xs]
 end
 
-function render_trace(trace, xmin, xmax)
+function render_trace(trace, xmin, xmax, colors)
     # histogram
     xs = get_xs(trace)
     (hist_data, ) = hist(xs, bins=collect(range(xmin, stop=xmax, length=50)), color="LightGray")
@@ -81,7 +81,6 @@ function render_trace(trace, xmin, xmax)
     plot(test_xs, densities * scale, color="black", linewidth=4, zorder=1)
     
     # individual component density plot
-    colors = ["red", "orange", "blue"]
     #if trace[:k] > 1
         for j in 1:trace[:k]
             densities = [marginal_density(1, [1.0], [trace[(:mu, j)]], [trace[(:var, j)]], x) for x in test_xs] * trace[:weights][j]
@@ -100,7 +99,8 @@ function show_prior_data()
     figure()
     xmin = -30.0
     xmax = 30.0
-    render_trace(trace, xmin, xmax)
+    colors = ["red", "orange", "blue"]
+    render_trace(trace, xmin, xmax, colors)
     savefig("prior_sample.png")
 end
 
@@ -281,13 +281,13 @@ function test_split_merge_move()
     xmin = -30.0
     xmax = 30.0
     subplot(1, 3, 1)
-    render_trace(three_cluster_trace, xmin, xmax)
+    render_trace(three_cluster_trace, xmin, xmax, ["red", "purple", "blue"])
     gca().get_yaxis().set_visible(false)
     subplot(1, 3, 2)
-    render_trace(two_cluster_trace, xmin, xmax)
+    render_trace(two_cluster_trace, xmin, xmax, ["red", "orange", "blue"])
     gca().get_yaxis().set_visible(false)
     subplot(1, 3, 3)
-    render_trace(one_cluster_trace, xmin, xmax)
+    render_trace(one_cluster_trace, xmin, xmax, ["red", "orange", "blue"])
     gca().get_yaxis().set_visible(false)
     tight_layout()
     savefig("rjmcmc.png")

### 2020-07-08 paper.tex
progress on theory

@@
 \usepackage{amsthm}
+\usepackage{algorithm}
 \usepackage{algorithmicx}
@@
 %\newcommand{\proposal}{\mathcal{Q}}
+\newtheorem{theorem}{Theorem}[section]
+\newtheorem{corollary}{Corollary}[theorem]
+\newtheorem{lemma}[theorem]{Lemma}
 
@@
 
-\section{INVOLUTIVE MCMC WITH TRACES}
+\section{INVOLUTIVE MCMC}
+Let $p$ denote the target probability density with respect to some $\sigma$-finite measure $\mu_X$ on a set $X$.
+For each $x$ such that $p(x) > 0$, let $q_x$ denote a probability density with respect to some $\sigma$-finite measure $\mu_U$ on a set $U$.
+Let $\mu := \mu_X \times \mu_U$ denote the product measure on the space $X \times U$, and let $\pi(x, u) := p(x) q_x(u)$ and $Z := \{(x, u) : \pi(x, u) > 0\}$.
+Let $f : Z \to Z$ denote an involution ($f^{-1} = f$) such that the pushforward of $\mu$ under $f$, denoted $\mu \circ f^{-1}$, is absolutely continuous with respect to $\mu$, with Radon-Nikodym derivative $d (\mu \circ f^{-1}) / d\mu : Z \to [0, \infty)$.
 
+\begin{algorithm}[h]
 \begin{algorithmic}
-\Require Target distribution $p$, proposal distribution $q$, involution $h(u, t)$, previous state $x$.
 \Procedure{involutive-mcmc-move}{$p$, $q$, $f$, $x$}
@@
     \State $(x', u') \gets f(x, u)$ \Comment{Apply involution}
+    %\State $\alpha \gets
+        %\frac{p(x') q(u'; x')}{p(x) q(u; x)} \left| \left[ \frac{\partial h(u, t)}{\partial (u, t)} \right] \right|$
+    % TODO or should we show the Radon-Nikodym derivative?
     \State $\alpha \gets
-        \frac{p(x') q(u'; x')}{p(x) q(u; x)} \left| \left[ \frac{\partial h(u, t)}{\partial (u, t)} \right] \right|$
+        \displaystyle \frac{p(x') q_{x'}(u')}{p(x) q_{x}(u)} \left( \frac{d (\mu \circ f^{-1})}{d \mu} (x, u)\right)^{-1}$
+    % TODO check it
     \State $r \sim \mathrm{Uniform}(0, 1)$
@@
 \end{algorithmic}
+\caption{Involutive MCMC}
+\label{alg:involutive-mcmc}
+\end{algorithm}
+
+\begin{lemma}\label{lemma:involution-detailed-balance} % TODO cite
+For $p$, $q$, and $f$ satisfying the properties above, let $k'$ denote the following transition kernel on $Z$:
+\[
+k'_z(A) := [f(z) \in A] \alpha(z) + [z \in A](1 - \alpha(z))
+\]
+The transition kernel $k'$ on $Z$ defined by the involution $f$ satisfies detailed balance with respect to $\pi$:
+\[
+\int_B k'_z(A) \pi(dz) = \int_A k'_z(B) \pi(dz) \mbox{ for all } A, B \in \Sigma_Z
+\]
+And in particular it is stationary with respect to $\pi$:
+\[
+\int_Z k'_z(A) \pi(dz) = \pi(A) \mbox{ for all } A \in \Sigma_Z
+\]
+\end{lemma}
+\begin{proof}
+\end{proof}
+
+\begin{theorem}\label{theorem:stationarity}
+The transition kernel $k$ on $X$ defined by Algorithm~\ref{alg:involutive-mcmc} for valid $p$, $q$, and $f$:
+\[
+k_x(A) := \int_U k'_{x,u}(A \times U) q_x(u) \mu_U(du) \;\; \mbox{for all} \;\; A \in \Sigma_X, x \in X
+\]
+is stationary with respect to $p$.
+\end{theorem}
+\begin{proof}
+By Lemma~\citep{lemma:involution-detailed-balance}, the transition kernel $k'$ is stationary with respect to $\pi$:
+\[
+%\int_B k'_{z}(B) \pi(z) \mu(dz) = \pi(B) \mbox{ for all } z \in X \times U, B \in \Sigma_Z
+\]
+It suffices to show that
+\begin{align*}
+\int_X k_x(A) p(x) \mu_X(dx) &= p(A) \mbox{ for all } A \in \Sigma_X\\
+\int_X \left( \int_U k'_{x,u}(A \times U) q_x(u) \mu_U(du) \right) p(x) \mu_X(dx) &= p(A)\\
+\int_{X \times U} k'_{x,u}(A \times U) q_x(u) p(x) \mu(dz) &= p(A)
+\end{align*}
+By Lemma~\citep{lemma:involution-detailed-balance}, setting $B := X \times U$, the left-hand-side is:
+\[
+\pi(X \times U) = \int_X \int_U p(x) q_x(u) \mu(
+\]
+\end{proof}
+\paragraph{Detailed balance of the involution}
+%Let $q_
+
+\paragraph{Stationarity}
+
+\paragraph{Conditional distribution and observations fixed}
 
@@
 
+\section{Detailed balance}
+
+Can we employ Theorem 2 of Tierney 1998? How does this relate to how the algorithm is described.
+Tierney already employs it for the involution-based move for the joint space of model and proposal traces (see page 4, bullet point 2 'deterministic proposals').
+But employing it for the composite move that includes the proposal and involution is more complex.
+The elements are:
+\begin{enumerate}
+\item $x, y \in X$ and $(X, \Sigma_X, \mu_X)$, where $x$ are traces of the model, and $\pi(x)$ is the model (target) density.
+\item $u \in U$ and $(U, \Sigma_U, \mu_U)$, where $u$ are traces of the proposal and $q_{x}(u)$ is the proposal density.
+\item An involution $h : X \times U \to X \times U$ (properly restricted to the support of $\pi(x) q_{x}(u)$.
+\end{enumerate}
+We want to provide a detailed balance result on $(X, \Sigma_X, \mu_X)$, proving detailed balance with respect to $\pi(x)$.
+Tierney's proposal kernel is defined by a family of measures $Q_{x} : \Sigma_X \to [0, \infty)$.
+In our case, this is:
+\[
+Q_{x}(A) := \int_U [h(x, u)_1 \in A] q_{x}(du) = \int_U [h(x, u)_1 \in A] q_{x}(u) \mu_U(du)
+\]
+Tierney's transition kernel with acceptance probability $\alpha(x \to y)$ is defined by a family of measures $P_x : \Sigma_X \to [0, \infty)$ (Equation 1 of Tierney):
+\[
+P_x(A) = \int_A \alpha(x \to y) Q_x(dy) + [x \in A] \int_X (1 - \alpha(x \to y)) Q_x(dy) \mbox{ for } A \in \Sigma_X
+\]
+Detailed balance (Tierney equation 2) is expressed as an equality between two measures on $X \times X$, denoted $(\pi P)_{\to}$ and $(\pi P)_{\gets}$.
+\[
+(\pi P)_{\to}(A, B) = (\pi P)_{\gets}(A, B)
+\]
+where
+\[
+(\pi P)_{\to}(A, B) := \int_A \left( \int_B P_x(dy) \right) \pi(dx)
+\]
+and
+\[
+(\pi P)_{\gets}(A, B) := \int_B \left( \int_A P_x(dy) \right) \pi(dx)
+\]
+Tierney Proposition 1 relies on a $\sigma$-finite measure $\mu$ on the product space $(X \times X, \Sigma_X \times \Sigma_X$), which is constructed as
+the measure induced by first sampling $x$ according to $\pi$ and then sampling $y$ according to the proposal $Q_x$:
+\[
+\mu(A, B) := \int_A \left( \int_B Q_x(dy) \right) \pi(dx) \;\;\; \mu^T(A, B) := \mu(B, A)
+\]
+In our case, this is:
+\[
+\mu(A, B) := \int_A \left( \int_U [h(x, u)_1 \in B] q_{x}(du) \right) \pi(dx) = \int_A \left( \int_U [h(x, u) \in B \times U] q_{x}(du) \right) \pi(dx)
+\]
+\paragraph{Show that $\mu$ is $\sigma$-finite (a requirement for Tierney Proposition 1)}
+$X \times X$ needs to be the countable union of measurable sets with finite measure $\mu(X \times X)$.
+It should build on the fact that $\pi$ and $q_x$ are $\sigma$-finite.
+%Can we use the fact that the composition of two $\sigma$-finite kernels is itself $\sigma$-finite (where the first kernel is $\pi$ and has a trivial index space)?
+%Apparently 
+%If so, then it would suffice to show that $Q$ is a $\sigma$-finite kernel, which is true if all $Q_x$ are $\sigma$-finite measures.
+
+\paragraph{Characterize $R$, and $\mu_R$ is the restriction of $\mu$ to $R$, and $\mu_R^T$ is as above}
+
+\paragraph{Characterize the following Radon-Nikodym derivative of $\mu_R$ with respect to $\mu_R^T$}
+\[
+r(x, y) = \frac{d\mu_R}{d\mu_R^T}(x, y)
+\]
+We need to show that:
+\begin{enumerate}
+\item $\alpha$ is $\mu$-almost everywhere zero on $R^c$
+\item $\alpha(x, y) r(x, y) = \alpha(y, x)$
+\end{enumerate}
+Then, according to Tierney Proposition 1, there exists a (unique up to null-sets) symmetric set $R \in \Sigma_X \times \Sigma_X$ such that $\mu$ and $\mu^T$ are mutually absolutely continuous.
+%(TODO: constructing $R$ for our kernel is a key step).
+
+\subsection{Special case of involution on discrete + diffeomorphism}
+
+\section{Using Tierney's result for deterministic proposals}
+
+\paragraph{Step 1: determine the acceptance ratio based on Tierney's involution construction}
+(page 4 of Tierney 1998)
+$T$ is the involution and $E$ is the joint space $X \times U$.
+Define the measure $\pi'$ by $\pi'(A) = \pi(T^{-1}(A))$ for all $A \in \Sigma_X$
+and let $\nu(dx) = \pi(dx) + \pi'(dx)$.
+Let $h(x)$ be a density for $\pi(dx)$ with respect to $\nu(dx)$ (( TODO this is they key step )).
+
+\paragraph{Step 2: prove stationarity for the end-to-end kernel}
+Using stationarity of the involutive kernel, which derives from Step 1.
+
+\section{Proof of detailed balance for involution}
+Detailed balance is proven by applying a result from~\citet{tierney1998note} for general Metropolis-Hastings moves with deterministic proposals, and using the measure $h(z) := \pi(z) / (\pi(z) + \pi'(z))$ where $\pi'$
+% See top of page 4, bullet point 2, of Tierney 1998 note
+
+\section{Proof of stationarity for involution}
+Stationarity derives from detailed balance by setting $B := Z$ in the detailed balance equation, and using $k'_z(Z) = \alpha(z) + (1 - \alpha(z)) =1$, so the right-hand side simplifies to $\int_A \pi(dz) = \pi(A)$.
+
+\section{Proof of stationarity for end-to-end kernel}
+We are given that the involution is stationary with respect to (the measure induced by) $\pi(x, u) := p(x) q_x(u)$:
+\[
+\int_Z k'_z(A) \pi(z) \mu(dz) = \pi(A) \mbox{ for all } A \in \Sigma_Z
+\]
+It suffices to show that the end-to-end kernel $k$ is stationary with respect to (the measure induced by) $p(x)$:
+\[
+\int_X k_x(A) p(x) \mu_X(dx) = p(A) \mbox{ for all } A \in \Sigma_X
+\]
+The kernel $k$ on $X$ defined by Algorithm~\ref{alg:involutive-mcmc} for valid $p$, $q$, and $f$:
+\[
+k_x(A) := \int_U k'_{x,u}(A \times U) q_x(u) \mu_U(du) \;\; \mbox{for all} \;\; A \in \Sigma_X, x \in X
+\]
+is stationary with respect to $p$.
+\begin{align*}
+    \int_X k_x(A) p(x) \mu_X(dx) &= p(A) \mbox{ for all } A \in \Sigma_X\\
+    \int_X \left( \int_U k'_{x,u}(A \times U) q_x(u) \mu_U(du) \right) p(x) \mu_X(dx) &= p(A)\\
+    \int_{X \times U} k'_{x,u}(A \times U) q_x(u) p(x) \mu(dz) &= p(A)\\
+    \int_{X \times U} k'_{x,u}(A \times U) \pi(z) \mu(dz) &= p(A)\\
+    \pi(A \times U) &= p(A)\\
+    \int_{A \times U} q_x(u) p(x) \mu(dz) &= p(A)\\
+    \int_A \left( \int_U q_x(u) \mu_U(du) \right) p(x) \mu_X(dx) &= p(A)\\
+    \int_A p(x) \mu_X(dx) &= p(A)
+\end{align*}
+% TODO characterize the various conditions
+
 \end{document}

### 2020-07-08 paper.tex
progress on theory

@@
 
+\section{Derivation of pushforward Radon-Nikodym derivative for special case}
+TODO: Do some special case where it simplifies to a Jacobian determinant.
+Maybe the case where the involution is composed of an involution on discrete part and a diffeomorphism on the continuous part.
+
 \section{Proof of detailed balance for involution}
-Detailed balance is proven by applying a result from~\citet{tierney1998note} for general Metropolis-Hastings moves with deterministic proposals, and using the measure $h(z) := \pi(z) / (\pi(z) + \pi'(z))$ where $\pi'$
-% See top of page 4, bullet point 2, of Tierney 1998 note
+\citet{tierney1998note} gives a class of MCMC kernels based on involutions that satisfy detailed balance.
+Specifically, given an involution $T : Z \to Z$, they denote the pushforward of $\pi$ under $T$ by $\pi'$:
+$\pi'(A) := \pi(T^{-1}(A))$.
+They let $\nu(A) := \pi(A) + \pi'(A)$.
+They let $h(z)$ denote `a density for $\pi$ with respect to $\nu$'.
+Then, they set:
+\[
+\alpha(x \to T(x)) \frac{h(x)}{h(T(x))} = \alpha(T(x) \to x)
+\]
+To use this result, it suffices to show that $h$ simplifes to the following, when $\pi$ and $\pi'$ are $\sigma$-finite and have densities with respect to the same reference measure $\mu$:
+\[
+\frac{h(z)}{h(T(z))}= \frac{\pi(z)}{\pi(T(z))} \left( \frac{d (\mu \circ T^{-1})}{d \mu} \right)^{-1}
+\]
+(note that $T$ is our $f$).
+What is $h(z)$, the density of $\pi$ with respect to $\nu(A) := \pi(A) + \pi'(A)$?
+Well, if $\pi$ and $\pi'$ have densities with respect to $\mu$, then so does $\nu$, and it is $\nu(z) = \pi(z) + \pi'(z)$.
+Then, the density of $\pi$ with respect to $\nu$ is:
+\[
+h(z) = \frac{\pi(z)}{\nu(z)} = \frac{\pi(z)}{\pi(z) + \pi'(z)}
+\]
+It suffices to show that:
+\[
+\frac{h(z)}{h(T(z))} = \frac{\pi(z)}{\pi(T(z))} \frac{\pi(T(z)) + \pi'(T(z))}{\pi(z) + \pi'(z)}
+= \frac{\pi(z)}{\pi(T(z))} \left( \frac{d (\mu \circ T^{-1})}{d \mu} \right)^{-1}
+\]
+It suffices to show that:
+\[
+\frac{\pi(T(z)) + \pi'(T(z))}{\pi(z) + \pi'(z)} = \left( \frac{d (\mu \circ T^{-1})}{d \mu} \right)^{-1}
+\]
+Let $a(z) = \pi(z) + \pi'(z)$, and note that $a$ is a density with respect to $\mu$.
+Then, the ratio above is:
+\[
+\frac{\pi(T(z)) + \pi'(T(z))}{\pi(z) + \pi'(z)} = \frac{a(T(z))}{a(z)}
+\]
+Consider the measure $a$ on $Z$ and the pushforward measure $a \circ T^{-1}$ on $Z$.
+It suffices to show that
+\[ 
+1 = \frac{d (a \circ T^{-1})}{d a} = \frac{a(T(z))}{a(z)} \left( \frac{d (\mu \circ T^{-1})}{d \mu} \right)
+\]
+$\frac{d (a \circ T^{-1})}{d a} = 1$ follows from the fact that $a$ and $a \circ T^{-1}$ are actually the same measure:
+\begin{align}
+a(B) &= \pi(B) + \pi'(B) = \pi(B) + \pi(T^{-1}(B))\\
+(a \circ T^{-1})(B) &= \pi(T^{-1}(B)) + \pi(T{-1}(T^{-1}(B))) = \pi(T^{-1}(B)) + \pi(B)
+\end{align}
+Using the chain rule (TODO check this!):
+\[
+1 = \frac{d (a \circ T^{-1})}{d a}
+=   \frac{d (a \circ T^{-1})}{d (\mu \circ T^{-1})}
+    \frac{d (\mu \circ T^{-1})}{d \mu}
+    \frac{d \mu}{d a}
+= a(T(z))  \frac{d (\mu \circ T^{-1})}{d \mu} \frac{1}{a(z)}
+\]
+Note that $a$ is actually equivalent to Tierney's $\nu$.
+The non-obvious part is the first factor:
+\[
+\frac{d (a \circ T^{-1})}{d (\mu \circ T^{-1})}(z) = \frac{d a}{d \mu} (T(z))
+\]
+TODO: prove me!
 
 \section{Proof of stationarity for involution}
-Stationarity derives from detailed balance by setting $B := Z$ in the detailed balance equation, and using $k'_z(Z) = \alpha(z) + (1 - \alpha(z)) =1$, so the right-hand side simplifies to $\int_A \pi(dz) = \pi(A)$.
+Stationarity follows from detailed balance by setting $B := Z$ in the detailed balance equation, and using $k'_z(Z) = \alpha(z) + (1 - \alpha(z)) =1$, so the right-hand side simplifies to $\int_A \pi(dz) = \pi(A)$.
 
@@
 \end{align*}
-% TODO characterize the various conditions
+% TODO characterize the various assumptions that allow us to e.g. rearrange the order of integration, etc.
 

### 2020-07-09 paper.tex
Finish the proof sketch and fix inverse error

--- 2020-07-08 paper.tex
@@
     \State $\alpha \gets
-        \displaystyle \frac{p(x') q_{x'}(u')}{p(x) q_{x}(u)} \left( \frac{d (\mu \circ f^{-1})}{d \mu} (x, u)\right)^{-1}$
+        \displaystyle \frac{p(x') q_{x'}(u')}{p(x) q_{x}(u)} \left( \frac{d (\mu \circ f^{-1})}{d \mu} (x, u)\right)$
     % TODO check it
@@
 
+\section{Derivation of pushforward Radon-Nikodym derivative for special case}
+TODO: Do some special case where it simplifies to a Jacobian determinant.
+Maybe the case where the involution is composed of an involution on discrete part and a diffeomorphism on the continuous part.
+
 \section{Proof of detailed balance for involution}
-Detailed balance is proven by applying a result from~\citet{tierney1998note} for general Metropolis-Hastings moves with deterministic proposals, and using the measure $h(z) := \pi(z) / (\pi(z) + \pi'(z))$ where $\pi'$
-% See top of page 4, bullet point 2, of Tierney 1998 note
+\citet{tierney1998note} gives a class of MCMC kernels based on involutions that satisfy detailed balance.
+Specifically, given an involution $T : Z \to Z$, they denote the pushforward of $\pi$ under $T$ by $\pi'$:
+$\pi'(A) := \pi(T^{-1}(A))$.
+They let $\nu(A) := \pi(A) + \pi'(A)$.
+They let $h(z)$ denote `a density for $\pi$ with respect to $\nu$'.
+Then, they set:
+\[
+\alpha(x \to T(x)) \frac{h(x)}{h(T(x))} = \alpha(T(x) \to x)
+\]
+Our definition of $\alpha$ satisfies:
+\[
+\frac{\alpha(z \to T(z))}{\alpha(T(z) \to z)} = \frac{\pi(T(z))}{\pi(z)} \left( \frac{d (\mu \circ T^{-1})}{d \mu} (z) \right)
+\]
+Therefore, to use Tierney's result, it suffices to show that (when $\pi$ and $\pi'$ are $\sigma$-finite and have densities with respect to the same reference measure $\mu$):
+\[
+\frac{h(T(z))}{h(z)}= \frac{\pi(T(z))}{\pi(z)} \left( \frac{d (\mu \circ T^{-1})}{d \mu} (z) \right)
+\]
+(note that $T$ is our $f$).
+What is $h(z)$, the density of $\pi$ with respect to $\nu(A) := \pi(A) + \pi'(A)$?
+Well, if $\pi$ and $\pi'$ have densities with respect to $\mu$, then so does $\nu$, and it is $\nu(z) = \pi(z) + \pi'(z)$, so the density of $\pi$ with respect to $\nu$ is:
+\[
+h(z) = \frac{\pi(z)}{\nu(z)} %= \frac{\pi(z)}{\pi(z) + \pi'(z)}
+\]
+and therefore:
+\[
+\frac{h(T(z))}{h(z)} = \frac{\pi(T(z))}{\pi(z)} \cdot \frac{\nu(z)}{\nu(T(z))}
+\]
+It suffices to show that:
+\[
+\frac{\nu(T(z))}{\nu(z)} = \left( \frac{d (\mu \circ T^{-1})}{d \mu} \right)^{-1}
+\]
+First, note that $\nu$ and $\nu \circ T^{-1}$ are the same measure:
+\begin{align}
+\nu(A) &= \pi(A) + \pi'(A) = \pi(A) + \pi(T^{-1}(A))\\
+(\nu \circ T^{-1})(A) &= \pi(T^{-1}(A)) + \pi(T^{-1}(T^{-1}(A))) = \pi(T^{-1}(A)) + \pi(A)
+\end{align}
+Therefore,
+\[
+\frac{d (\nu \circ T^{-1})}{d \nu}(z) = 1 \mbox{ for all } z
+\]
+Expanding $d(\nu \circ T^{-1}) / d \nu$ using the chain rule:
+\begin{align}
+1 &= \frac{d (\nu \circ T^{-1})}{d \nu}(z)\\
+&=   \frac{d (\nu \circ T^{-1})}{d (\mu \circ T^{-1})}(z) \cdot
+    \frac{d (\mu \circ T^{-1})}{d \mu}(z) \cdot
+    \frac{d \mu}{d \nu}(z)\\
+&= \nu(T(z))  \frac{d (\mu \circ T^{-1})}{d \mu} \frac{1}{\nu(z)}
+\end{align}
+where in the second step we used that
+\[
+\frac{d (\nu \circ T^{-1})}{d (\mu \circ T^{-1})}(z) = \frac{d \nu}{d \mu}(T(z)) = \nu(T(z))
+\]
+To show this, use (i) $\nu \circ T^{-1} = \nu$, (ii) the definition of Radon-Nikodym derivative $d \nu / d \mu$ and (iii) and Theorem 3.6.1 in Bogachev with $g := d\nu / d\mu$:
+\begin{align}
+(\nu \circ T^{-1})(A) = \nu(A) &= \int_A \frac{d \nu}{d \mu}(z) \mu(dz) = \int_A \frac{d \nu}{d \mu} (T(z)) ( \mu \circ T^{-1} )(dz)
+\end{aling}
 
 \section{Proof of stationarity for involution}
-Stationarity derives from detailed balance by setting $B := Z$ in the detailed balance equation, and using $k'_z(Z) = \alpha(z) + (1 - \alpha(z)) =1$, so the right-hand side simplifies to $\int_A \pi(dz) = \pi(A)$.
+Stationarity follows from detailed balance by setting $B := Z$ in the detailed balance equation, and using $k'_z(Z) = \alpha(z) + (1 - \alpha(z)) =1$, so the right-hand side simplifies to $\int_A \pi(dz) = \pi(A)$.
 

### 2020-07-09 paper.tex
remove outdated

@@
 
-\section{Using Tierney's result for deterministic proposals}
-
-\paragraph{Step 1: determine the acceptance ratio based on Tierney's involution construction}
-(page 4 of Tierney 1998)
-$T$ is the involution and $E$ is the joint space $X \times U$.
-Define the measure $\pi'$ by $\pi'(A) = \pi(T^{-1}(A))$ for all $A \in \Sigma_X$
-and let $\nu(dx) = \pi(dx) + \pi'(dx)$.
-Let $h(x)$ be a density for $\pi(dx)$ with respect to $\nu(dx)$ (( TODO this is they key step )).
-
-\paragraph{Step 2: prove stationarity for the end-to-end kernel}
-Using stationarity of the involutive kernel, which derives from Step 1.
-
 \section{Derivation of pushforward Radon-Nikodym derivative for special case}

### 2020-07-09 paper.tex
finish proof for Jacobian

@@
 
-%\usepackage{aistats2020}
+\usepackage{aistats2020}
 % If your paper is accepted, change the options for the package
@@
 %
-\usepackage[accepted]{aistats2020}
+%\usepackage[accepted]{aistats2020}
 %
@@
 \subsection{Distributions on traces}
+% TODO rewrite this section..
 Define target density $\pi$ (XXX below it is called $p$, fix this) on the joint space $X \times U$, where $X$ are traces of the model and $X$ are traces of the proposal, in terms of the density $p(x)$ and the family of densities $q_x(u)$ as $\pi(x, u) := p(x) q_u(x)$.
@@
 
-\subsection{Stationarity of the involution with respect to joint density}
-Let $(Z, \Sigma, \mu)$ be a measure space of pairs of traces of the model and proposal, where $\mu$ is $\sigma$-finite.
-Consider a $\mu$-measurable function $h : Z \to Z$ that is an involution ($h(h(z)) = z$) and let $\mu \circ h^{-1}$ denote the pushforward of $\mu$ under $h$, such that $\mu \circ h^{-1}$ is $\sigma$-finite and such that $\mu \circ h^{-1}$ is absolutely continuous with respect to $\mu$.
-% TODO can we simplify the conditions, using the fact that h is an involution?
-Then, the Radon-Nikodym derivative exists, and is denoted $(d (\mu \circ h^{-1}) / d \mu)(z)$.
+\subsection{Derivation of the pushforward Radon-Nikodym derivative for a special case}
+We consider the special case in which the involution $f$ can be factored into an involution on a countable set $I$ and a family of bijections on $\mathbb{R}^{n_i}$ for some $n_i$ for each $i \in I$.
+Suppose $Z = \{(i, \mathbf{x}) : i \in I, \mathbf{x} \in \mathbb{R}^{n_i}\}$.
+Suppose $f_1$ is an involution on $I$ and $n_i = n_{f_1(i)}$ and $f_2$ is a family of continuously differentiable bijections indexed by $i \in I$, such that $f_{2,i} : \mathbb{R}^{n_i} \to \mathbb{R}^{n_i}$.
+Also suppose that $f_{2,i} = f_{2,f_1(i)}^{-1}$ for all $i \in I$. That is,
+\begin{equation}
+f_{2,i}(f_{2,f_1(i)}(\mathbf{x})) = \mathbf{x} \mbox{ for all } \mathbf{x} \in \mathbb{R}^{n_i} \mbox{ and all } i \in I
+\end{equation} 
+Then, $f : Z \to Z : (i, \mathbf{x}) \mapsto (f_1(i), f_{2,i}(\mathbf{x}))$ is an involution because:
+\begin{equation}
+f(f(i, \mathbf{x})) = f(f_1(i), f_{2,i}(\mathbf{x})) = (f_1(f_1(i)), f_{2,f_1(i)}(f_{2,i}(\mathbf{x}))) = (i, \mathbf{x})
+\end{equation}
+Let $\Sigma_n$ and $\mu_n$ denote the Lebesgue $\sigma$-algebra and Lebesgue measure on $\mathbb{R}^n$, respectively.
+Let $\Sigma \subset \mathcal{P}(Z)$ be the $\sigma$-algebra of sets of the form $\cup_{i \in I} \{(i, \mathbf{x}) : \mathbf{x} \in A_i\}$ for some $A_i \in \Sigma_{n_i}$ for each $i \in I$.
+%Let $\Sigma$ be the $\sigma$-algebra on $Z$ generated by the sets $G := \{ \{(i, \mathbf{x}) : \mathbf{x} \in A\} : i \in I, A \in \Sigma_{n_i} \}$.
+%Note that the sets $G$ form a $\pi$-system:
+%\begin{align}
+%\{(i, \mathbf{x}) : \mathbf{x} \in A\} \cap \{(j, \mathbf{x}) : \mathbf{x} \in B\} = \varnothing = \{(i, \mathbf{x}) : \mathbf{x} \in \varnothing \} \in G \mbox{ for } i \ne j\\
+%\{(i, \mathbf{x}) : \mathbf{x} \in A\} \cap \{(i, \mathbf{x}) : \mathbf{x} \in B\} = \{(i, \mathbf{x}) : \mathbf{x} \in A \cap B\} \in G \mbox{ because } A \cap B \in \Sigma_{n_i}
+%\end{align}
+Let $\mu$ denote the measure on measurable space $(Z, \Sigma)$ given by:%, uniquely defined by its value on the sets $G$:
+\begin{equation}
+\mu(\cup_{i \in I} \{(i, \mathbf{x}) : \mathbf{x} \in A_i\}) := \sum_{i \in I} \mu_{n_i}(A_i)
+\end{equation}
+We wish to show that the Radon-Nikodym derivative of the pushforward of $\mu$ by $f$ with respect to $\mu$, evaluated at $(i, \mathbf{x})$, is the absolute value of the Jacobian (determinant) of the function $f_{2,i}$ evaluated at $\mathbf{x}$, which is denoted $(J f_{2,i})(\mathbf{x})$:
+\begin{equation}
+\frac{d (\mu \circ f^{-1})}{d \mu}(i, \mathbf{x})
+= \left| (Jf_{2,i})(\mathbf{x})\right|
+\end{equation}
+Consider $(\mu \circ f^{-1})(A)$ for $A \in \Sigma$:
+\begin{align}
+(\mu \circ f^{-1})(\cup_{i \in I} \{ (i, \mathbf{x}) : \mathbf{x} \in A_i \})
+&= \mu(f^{-1}(\cup_{i \in I} \{ (i, \mathbf{x}) : \mathbf{x} \in A_i \}))\\
+&= \mu(\cup_{i \in I} f^{-1}(\{ (i, \mathbf{x}) : \mathbf{x} \in A_i \}))\\
+&= \mu(\cup_{i \in I} \{ (f_1(i), \mathbf{x}) : \mathbf{x} \in f_{2,i}(A_i) \})\\
+&= \sum_{i \in I} \mu_{n_i}(f_{2,i}(A_i))
+\end{align}
+It suffices to show that for all $A \in \Sigma$:
+\begin{equation}
+\int_A \left| (Jf_{2,i})(\mathbf{x})\right| \mu(dz) = \sum_{i \in I} \mu_{n_i}(f_{2,i}(A_i))
+\end{equation}
+Expanding the left-hand side:
+\begin{align}
+\int_A \left| (Jf_{2,i})(\mathbf{x})\right| \mu(dz)
+= \sum_{i \in I} \int_{A_i} \left| (Jf_{2,i})(\mathbf{x})\right| \mu_{n_i}(d \mathbf{x}) % TODO how to justify this? it seems obvious
+= \sum_{i \in I} \mu_{n_i}(f_{2,i}(A_i)) % Borgachev Theorem 4.7.1 
+\end{align}
+where the final step uses Borgachev Theorem 3.7.1 with $g := 1$, and $F := f_{2,i}$.
 
-The goal is to prove stationarity with respect to the measure defined by density $p(z)$ with respect to $\mu$.
-The move starts with state $z \in Z$, applies the involution $h$, and then accepts (returns $h(z)$) with probability $\alpha(z)$ and otherwise returns $z$.
-Let $k_{z}$ denote the measure on new states induced by the involution $h$ and the previous state $z$:
-\begin{equation}
-k_{z}(A) = [h(z) \in A] \alpha(z) + [z \in A] (1 - \alpha(z))
-\end{equation}
-The stationarity condition is:
-\begin{equation}
-\int_Z p(z) k_{z}(A) \mu(dz) = p(A) = \int_Z p(z) [z \in A] \mu(dz) \mbox{ for all } A \in \Sigma
-\end{equation}
-Substituting in the definition of $k_z(A)$ and simplifying gives:
-\begin{equation} \label{eq:stationarity-requirement}
-\int_Z p(z) [h(z) \in A] \alpha(z) \mu(dz) = \int_Z p(z) [z \in A] \alpha(z) \mu(dz) \mbox{ for all } A \in \Sigma
-\end{equation}
-Note that for a $\mu$-measurable function $g$ such that $g \circ h$ is integrable with respect to $\mu$:
-\begin{equation} \label{eq:pushforward}
-\int_Z g(h(z)) \mu(dz) = \int_Z g(z) (\mu \circ h^{-1})(dz) = \int_Z g(z) \left( \frac{d (\mu \circ h^{-1})}{d \mu}(z) \right) \mu(dz)
-\end{equation}
-For $g(z) := p(h(z)) [z \in A] \alpha(h(z))$ we have $g(h(z)) = p(z) [h(z) \in A] \alpha(z)$, which is the integrand of the left-hand-side of Equation~(\ref{eq:stationarity-requirement}).
-Applying Equation~(\ref{eq:pushforward}) to the left-hand side of Equation~(\ref{eq:stationarity-requirement}) gives:
-\begin{equation}
-\int_Z p(z) [h(z) \in A] \alpha(z) \mu(dz) = \int_Z g(z) \left( \frac{d (\mu \circ h^{-1})}{d \mu}(z) \right) \mu(dz) = \int_Z p(h(z)) [z \in A] \alpha(h(z)) \left( \frac{d (\mu \circ h^{-1})}{d \mu}(z) \right) \mu(dz)
-\end{equation}
-Equating this with the right-hand side of Equation~(\ref{eq:stationarity-requirement}) gives the following sufficient condition for stationarity:
-\begin{equation}
-\int_Z p(h(z)) [z \in A] \alpha(h(z)) \left( \frac{d (\mu \circ h^{-1})}{d \mu}(z) \right) \mu(dz) = \int_Z p(z) [z \in A] \mu(dz) \mbox{ for all } A \in \Sigma
-\end{equation}
-It therefore suffices to find $\alpha$ such that:
-\begin{equation}
-p(h(z)) [z \in A] \alpha(h(z)) \left( \frac{d (\mu \circ h^{-1})}{d \mu}(z) \right) = p(z) [z \in A] \alpha(z) \mbox{ for all } z \in Z
-\end{equation}
-The following choice of $\alpha$ suffices:
-\begin{equation}
-\alpha(z) = \min\left\{ 1, \frac{p(h(z))}{p(z)} \left( \frac{d (\mu \circ h^{-1})}{d \mu}(z) \right) \right\}
-\end{equation}
-
-All requirements:
-\begin{enumerate}
-\item $\mu$ is $\sigma$-finite
-\item $h$ is $\mu$-measurable
-\item $h$ is an involution
-\item the pushforward $\mu \circ h^{-1}$ is $\sigma$-finite
-\item the pushforward $\mu \circ h^{-1}$ is absolutely continuous with respect to $\mu$
-\item the function $g(z) := p(h(z)) [z \in A] \alpha(h(z))$ is $\mu$-measurable for all $A \in \Sigma$
-\item the function $(g \circ h)(z) = p(z) [h(z) \in A] \alpha(z)$ is integrable with respect to $\mu$ for all $A \in \Sigma$
-\end{enumerate}
-
-\paragraph{Characterizing the pushforward $\mu \circ h^{-1}$ for involution $h$}
-Can we guarantee that it is $\sigma$-finite or absolutely continuous with respect to $\mu$?
-Are there simpler sufficient conditions we can give?
-$(\mu \circ h^{-1})$ is $\sigma$-finite if $Z$ is the countable union of $(mu \circ h^{-1})$-measurable sets with finite measure.
-
-$\sigma$-finite means
-
-\section{Detailed balance}
-
-Can we employ Theorem 2 of Tierney 1998? How does this relate to how the algorithm is described.
-Tierney already employs it for the involution-based move for the joint space of model and proposal traces (see page 4, bullet point 2 'deterministic proposals').
-But employing it for the composite move that includes the proposal and involution is more complex.
-The elements are:
-\begin{enumerate}
-\item $x, y \in X$ and $(X, \Sigma_X, \mu_X)$, where $x$ are traces of the model, and $\pi(x)$ is the model (target) density.
-\item $u \in U$ and $(U, \Sigma_U, \mu_U)$, where $u$ are traces of the proposal and $q_{x}(u)$ is the proposal density.
-\item An involution $h : X \times U \to X \times U$ (properly restricted to the support of $\pi(x) q_{x}(u)$.
-\end{enumerate}
-We want to provide a detailed balance result on $(X, \Sigma_X, \mu_X)$, proving detailed balance with respect to $\pi(x)$.
-Tierney's proposal kernel is defined by a family of measures $Q_{x} : \Sigma_X \to [0, \infty)$.
-In our case, this is:
-\[
-Q_{x}(A) := \int_U [h(x, u)_1 \in A] q_{x}(du) = \int_U [h(x, u)_1 \in A] q_{x}(u) \mu_U(du)
-\]
-Tierney's transition kernel with acceptance probability $\alpha(x \to y)$ is defined by a family of measures $P_x : \Sigma_X \to [0, \infty)$ (Equation 1 of Tierney):
-\[
-P_x(A) = \int_A \alpha(x \to y) Q_x(dy) + [x \in A] \int_X (1 - \alpha(x \to y)) Q_x(dy) \mbox{ for } A \in \Sigma_X
-\]
-Detailed balance (Tierney equation 2) is expressed as an equality between two measures on $X \times X$, denoted $(\pi P)_{\to}$ and $(\pi P)_{\gets}$.
-\[
-(\pi P)_{\to}(A, B) = (\pi P)_{\gets}(A, B)
-\]
-where
-\[
-(\pi P)_{\to}(A, B) := \int_A \left( \int_B P_x(dy) \right) \pi(dx)
-\]
-and
-\[
-(\pi P)_{\gets}(A, B) := \int_B \left( \int_A P_x(dy) \right) \pi(dx)
-\]
-Tierney Proposition 1 relies on a $\sigma$-finite measure $\mu$ on the product space $(X \times X, \Sigma_X \times \Sigma_X$), which is constructed as
-the measure induced by first sampling $x$ according to $\pi$ and then sampling $y$ according to the proposal $Q_x$:
-\[
-\mu(A, B) := \int_A \left( \int_B Q_x(dy) \right) \pi(dx) \;\;\; \mu^T(A, B) := \mu(B, A)
-\]
-In our case, this is:
-\[
-\mu(A, B) := \int_A \left( \int_U [h(x, u)_1 \in B] q_{x}(du) \right) \pi(dx) = \int_A \left( \int_U [h(x, u) \in B \times U] q_{x}(du) \right) \pi(dx)
-\]
-\paragraph{Show that $\mu$ is $\sigma$-finite (a requirement for Tierney Proposition 1)}
-$X \times X$ needs to be the countable union of measurable sets with finite measure $\mu(X \times X)$.
-It should build on the fact that $\pi$ and $q_x$ are $\sigma$-finite.
-%Can we use the fact that the composition of two $\sigma$-finite kernels is itself $\sigma$-finite (where the first kernel is $\pi$ and has a trivial index space)?
-%Apparently 
-%If so, then it would suffice to show that $Q$ is a $\sigma$-finite kernel, which is true if all $Q_x$ are $\sigma$-finite measures.
-
-\paragraph{Characterize $R$, and $\mu_R$ is the restriction of $\mu$ to $R$, and $\mu_R^T$ is as above}
-
-\paragraph{Characterize the following Radon-Nikodym derivative of $\mu_R$ with respect to $\mu_R^T$}
-\[
-r(x, y) = \frac{d\mu_R}{d\mu_R^T}(x, y)
-\]
-We need to show that:
-\begin{enumerate}
-\item $\alpha$ is $\mu$-almost everywhere zero on $R^c$
-\item $\alpha(x, y) r(x, y) = \alpha(y, x)$
-\end{enumerate}
-Then, according to Tierney Proposition 1, there exists a (unique up to null-sets) symmetric set $R \in \Sigma_X \times \Sigma_X$ such that $\mu$ and $\mu^T$ are mutually absolutely continuous.
-%(TODO: constructing $R$ for our kernel is a key step).
-
-\subsection{Special case of involution on discrete + diffeomorphism}
-
-\section{Derivation of pushforward Radon-Nikodym derivative for special case}
-TODO: Do some special case where it simplifies to a Jacobian determinant.
-Maybe the case where the involution is composed of an involution on discrete part and a diffeomorphism on the continuous part.
-
-\section{Proof of detailed balance for involution}
+\subsection{Proof of detailed balance for involution}
 \citet{tierney1998note} gives a class of MCMC kernels based on involutions that satisfy detailed balance.
@@
 where in the second step we used that
-\[
+\begin{equation}
 \frac{d (\nu \circ T^{-1})}{d (\mu \circ T^{-1})}(z) = \frac{d \nu}{d \mu}(T(z)) = \nu(T(z))
-\]
-To show this, use (i) $\nu \circ T^{-1} = \nu$, (ii) the definition of Radon-Nikodym derivative $d \nu / d \mu$ and (iii) and Theorem 3.6.1 in Bogachev with $g := d\nu / d\mu$:
+\end{equation}
+To show this, we use the definition of the Radon-Nikodym derivative $d (\nu \circ T^{-1}) / d (\mu \circ T^{-1})$.
+To show that $z \mapsto (d \nu / d \mu)(T(z))$ is this Radon-Nikodym derivative, it suffices to show that for all measurable $A$:
+\begin{equation}
+\int_A \left( \frac{d \nu}{d \mu} (T(z)) \right) ( \mu \circ T^{-1} )(dz) = (\nu \circ T^{-1})(A) := \nu(T^{-1}(A))
+\end{equation}
+Applying Theorem 3.6.1 in Bogachev with $Y := A$, $y := z$, $x := z'$, $X := T^{-1}(A)$, $f := T$, and $g(y) := (d\nu / d\mu)(T(y))$:
 \begin{align}
-(\nu \circ T^{-1})(A) = \nu(A) &= \int_A \frac{d \nu}{d \mu}(z) \mu(dz) = \int_A \frac{d \nu}{d \mu} (T(z)) ( \mu \circ T^{-1} )(dz)
-\end{aling}
+\int_A \left( \frac{d \nu}{d \mu} (T(z)) \right) ( \mu \circ T^{-1} )(dz)
+&= \int_Y g(y) ( \mu \circ f^{-1} )(dy)\\
+&= \int_X g(f(x)) \mu(dx) \;\;\;\; [\mbox{Bogachev Theorem 3.6.1}]\\
+&= \int_{T^{-1}(A)} (d\nu / d\mu)(T(T(z'))) \mu(dz')\\
+&= \int_{T^{-1}(A)} (d\nu / d\mu)(z') \mu(dz')\\
+&= \int_{T^{-1}(A)} \nu(dz')
+= \nu(T^{-1}(A))
+\end{align}
+% NOTE Bogachev's theorem uses domains of integration which are the whole
+% sample spaces for the two sigma algebras.  I think that the theorem should
+% also apply to any measurable sets A and T(A) because one can construct a
+% sigma algebra and measure that are restricted to to any nonempty measurable
+% set.
 
-\section{Proof of stationarity for involution}
+\subsection{Proof of stationarity for involution}
 Stationarity follows from detailed balance by setting $B := Z$ in the detailed balance equation, and using $k'_z(Z) = \alpha(z) + (1 - \alpha(z)) =1$, so the right-hand side simplifies to $\int_A \pi(dz) = \pi(A)$.
 
-\section{Proof of stationarity for end-to-end kernel}
+\subsection{Proof of stationarity for end-to-end kernel}
 We are given that the involution is stationary with respect to (the measure induced by) $\pi(x, u) := p(x) q_x(u)$:

### 2020-07-10 paper.tex
Add some text to main paper

@@
 
-\subsection{Traces}
+%\subsection{Traces}
 
@@
 \end{algorithm}
+See Section~\ref{sec:radon-nikodym-special-case} for a derivation of the Radon-Nikodym derivative $d (\mu \circ f^{-1}) / d \mu$ for a special case, when the involution $f$ can be factored into (i) an involution on a countable partition of $Z$ and (ii) a family of continuously differentiable bijection on Euclidean spaces.
 
-\begin{lemma}\label{lemma:involution-detailed-balance} % TODO cite
-For $p$, $q$, and $f$ satisfying the properties above, let $k'$ denote the following transition kernel on $Z$:
-\[
-k'_z(A) := [f(z) \in A] \alpha(z) + [z \in A](1 - \alpha(z))
-\]
-The transition kernel $k'$ on $Z$ defined by the involution $f$ satisfies detailed balance with respect to $\pi$:
-\[
-\int_B k'_z(A) \pi(dz) = \int_A k'_z(B) \pi(dz) \mbox{ for all } A, B \in \Sigma_Z
-\]
-And in particular it is stationary with respect to $\pi$:
-\[
-\int_Z k'_z(A) \pi(dz) = \pi(A) \mbox{ for all } A \in \Sigma_Z
-\]
-\end{lemma}
+\begin{theorem}[Involutive MCMC is stationary]
+For $p$, $q$, and $f$ satisfying the requirements above, Algorithm~\ref{alg:involutive-mcmc} defines a probability kernel on $X$ that is stationary with respect to the distribution $p$.
+\end{theorem}
 \begin{proof}
+The proof is presented in stages in the appendix (see Section~\ref{sec:involution-detailed-balance}, Section~\ref{sec:involution-is-stationary}, and Section~\ref{sec:involutive-mcmc-is-stationary}).
 \end{proof}
 
-\begin{theorem}\label{theorem:stationarity}
-The transition kernel $k$ on $X$ defined by Algorithm~\ref{alg:involutive-mcmc} for valid $p$, $q$, and $f$:
-\[
-k_x(A) := \int_U k'_{x,u}(A \times U) q_x(u) \mu_U(du) \;\; \mbox{for all} \;\; A \in \Sigma_X, x \in X
-\]
-is stationary with respect to $p$.
-\end{theorem}
-\begin{proof}
-By Lemma~\citep{lemma:involution-detailed-balance}, the transition kernel $k'$ is stationary with respect to $\pi$:
-\[
-%\int_B k'_{z}(B) \pi(z) \mu(dz) = \pi(B) \mbox{ for all } z \in X \times U, B \in \Sigma_Z
-\]
-It suffices to show that
-\begin{align*}
-\int_X k_x(A) p(x) \mu_X(dx) &= p(A) \mbox{ for all } A \in \Sigma_X\\
-\int_X \left( \int_U k'_{x,u}(A \times U) q_x(u) \mu_U(du) \right) p(x) \mu_X(dx) &= p(A)\\
-\int_{X \times U} k'_{x,u}(A \times U) q_x(u) p(x) \mu(dz) &= p(A)
-\end{align*}
-By Lemma~\citep{lemma:involution-detailed-balance}, setting $B := X \times U$, the left-hand-side is:
-\[
-\pi(X \times U) = \int_X \int_U p(x) q_x(u) \mu(
-\]
-\end{proof}
-\paragraph{Detailed balance of the involution}
-%Let $q_
-
-\paragraph{Stationarity}
-
-\paragraph{Conditional distribution and observations fixed}
+%\begin{lemma}\label{lemma:involution-detailed-balance} % TODO cite
+%For $p$, $q$, and $f$ satisfying the properties above, let $k'$ denote the following transition kernel on $Z$:
+%\[
+%k'_z(A) := [f(z) \in A] \alpha(z) + [z \in A](1 - \alpha(z))
+%\]
+%The transition kernel $k'$ on $Z$ defined by the involution $f$ satisfies detailed balance with respect to $\pi$:
+%\[
+%\int_B k'_z(A) \pi(dz) = \int_A k'_z(B) \pi(dz) \mbox{ for all } A, B \in \Sigma_Z
+%\]
+%And in particular it is stationary with respect to $\pi$:
+%\[
+%\int_Z k'_z(A) \pi(dz) = \pi(A) \mbox{ for all } A \in \Sigma_Z
+%\]
+%\end{lemma}
+%\begin{proof}
+%\end{proof}
+%
+%\begin{theorem}\label{theorem:stationarity}
+%The transition kernel $k$ on $X$ defined by Algorithm~\ref{alg:involutive-mcmc} for valid $p$, $q$, and $f$:
+%\[
+%k_x(A) := \int_U k'_{x,u}(A \times U) q_x(u) \mu_U(du) \;\; \mbox{for all} \;\; A \in \Sigma_X, x \in X
+%\]
+%is stationary with respect to $p$.
+%\end{theorem}
+%\begin{proof}
+%By Lemma~\citep{lemma:involution-detailed-balance}, the transition kernel $k'$ is stationary with respect to $\pi$:
+%\[
+%%\int_B k'_{z}(B) \pi(z) \mu(dz) = \pi(B) \mbox{ for all } z \in X \times U, B \in \Sigma_Z
+%\]
+%It suffices to show that
+%\begin{align*}
+%\int_X k_x(A) p(x) \mu_X(dx) &= p(A) \mbox{ for all } A \in \Sigma_X\\
+%\int_X \left( \int_U k'_{x,u}(A \times U) q_x(u) \mu_U(du) \right) p(x) \mu_X(dx) &= p(A)\\
+%\int_{X \times U} k'_{x,u}(A \times U) q_x(u) p(x) \mu(dz) &= p(A)
+%\end{align*}
+%By Lemma~\citep{lemma:involution-detailed-balance}, setting $B := X \times U$, the left-hand-side is:
+%\[
+%\pi(X \times U) = \int_X \int_U p(x) q_x(u) \mu(
+%\]
+%\end{proof}
+%\paragraph{Detailed balance of the involution}
+%%Let $q_
+%
+%\paragraph{Stationarity}
+%
+%\paragraph{Conditional distribution and observations fixed}
 
@@
 
-\subsection{Distributions on traces}
-% TODO rewrite this section..
-Define target density $\pi$ (XXX below it is called $p$, fix this) on the joint space $X \times U$, where $X$ are traces of the model and $X$ are traces of the proposal, in terms of the density $p(x)$ and the family of densities $q_x(u)$ as $\pi(x, u) := p(x) q_u(x)$.
-
-% TODO rename the joint target density to \pi below
-We want to show stationarity with respect to $\pi$, using stationarity of the involution with respect to the joint density (below).
-\begin{equation}
-\int_X p(x) \ell_x(A) \mu(dx) = p(A) \mbox{ for all } A \in \Sigma
-\end{equation}
-where $\ell_x(A)$ is the probability of transitioning into set $A$ from $x$, and is given by:
-\begin{equation}
-\ell_x(A) := \int_U k_{(x, u)}(A \times U) \mu(du)
-\end{equation}
-where $k_{(x, u)}$ is the measure on $X \times U$ that is induced by applying the involution $h$ to the state $(x, u)$.
+%\subsection{Distributions on traces}
+%% TODO rewrite this section..
+%Define target density $\pi$ (XXX below it is called $p$, fix this) on the joint space $X \times U$, where $X$ are traces of the model and $X$ are traces of the proposal, in terms of the density $p(x)$ and the family of densities $q_x(u)$ as $\pi(x, u) := p(x) q_u(x)$.
+%
+%% TODO rename the joint target density to \pi below
+%We want to show stationarity with respect to $\pi$, using stationarity of the involution with respect to the joint density (below).
+%\begin{equation}
+%\int_X p(x) \ell_x(A) \mu(dx) = p(A) \mbox{ for all } A \in \Sigma
+%\end{equation}
+%where $\ell_x(A)$ is the probability of transitioning into set $A$ from $x$, and is given by:
+%\begin{equation}
+%\ell_x(A) := \int_U k_{(x, u)}(A \times U) \mu(du)
+%\end{equation}
+%where $k_{(x, u)}$ is the measure on $X \times U$ that is induced by applying the involution $h$ to the state $(x, u)$.
 % TODO checkme
 
-\subsection{Derivation of the pushforward Radon-Nikodym derivative for a special case}
-We consider the special case in which the involution $f$ can be factored into an involution on a countable set $I$ and a family of bijections on $\mathbb{R}^{n_i}$ for some $n_i$ for each $i \in I$.
+\subsection{Derivation of the pushforward Radon-Nikodym derivative for a special case} \label{sec:radon-nikodym-special-case}
+Implementing Algorithm~\ref{alg:involutive-mcmc} requires computing the Radon-Nikodym derivative $d (\mu \circ f^{-1}) / d \mu$.
+This section derives that function for the special case in which the involution $f$ can be factored into an involution on a countable set $I$ and a family of bijections on $\mathbb{R}^{n_i}$ for some $n_i$ for each $i \in I$.
 Suppose $Z = \{(i, \mathbf{x}) : i \in I, \mathbf{x} \in \mathbb{R}^{n_i}\}$.
@@
 
-\subsection{Proof of detailed balance for involution}
+\subsection{Proof of detailed balance for involution} \label{sec:involution-detailed-balance}
 \citet{tierney1998note} gives a class of MCMC kernels based on involutions that satisfy detailed balance.
-Specifically, given an involution $T : Z \to Z$, they denote the pushforward of $\pi$ under $T$ by $\pi'$:
-$\pi'(A) := \pi(T^{-1}(A))$.
-They let $\nu(A) := \pi(A) + \pi'(A)$.
-They let $h(z)$ denote `a density for $\pi$ with respect to $\nu$'.
-Then, they set:
-\[
-\alpha(x \to T(x)) \frac{h(x)}{h(T(x))} = \alpha(T(x) \to x)
-\]
-Our definition of $\alpha$ satisfies:
-\[
-\frac{\alpha(z \to T(z))}{\alpha(T(z) \to z)} = \frac{\pi(T(z))}{\pi(z)} \left( \frac{d (\mu \circ T^{-1})}{d \mu} (z) \right)
-\]
-Therefore, to use Tierney's result, it suffices to show that (when $\pi$ and $\pi'$ are $\sigma$-finite and have densities with respect to the same reference measure $\mu$):
-\[
-\frac{h(T(z))}{h(z)}= \frac{\pi(T(z))}{\pi(z)} \left( \frac{d (\mu \circ T^{-1})}{d \mu} (z) \right)
-\]
-(note that $T$ is our $f$).
-What is $h(z)$, the density of $\pi$ with respect to $\nu(A) := \pi(A) + \pi'(A)$?
-Well, if $\pi$ and $\pi'$ have densities with respect to $\mu$, then so does $\nu$, and it is $\nu(z) = \pi(z) + \pi'(z)$, so the density of $\pi$ with respect to $\nu$ is:
-\[
-h(z) = \frac{\pi(z)}{\nu(z)} %= \frac{\pi(z)}{\pi(z) + \pi'(z)}
-\]
-and therefore:
-\[
-\frac{h(T(z))}{h(z)} = \frac{\pi(T(z))}{\pi(z)} \cdot \frac{\nu(z)}{\nu(T(z))}
-\]
+We reproduce the result in our notation:
+\begin{lemma}[Detailed balance for involution move~\citep{tierney1998note}]
+Let $(Z, \Sigma, \pi)$ denote a measure space.
+Suppose $f$ is a one-to-one function from $Z$ onto $Z$ such that $f^{-1} = f$.
+Consider the probability kernel $k$ defined by $k_z(A) := [f(z) \in A] \alpha(z, f(z)) + [z \in A] (1 - \alpha(z, f(z)))$ (where $\alpha(z, f(z))$ gives the probability of accepting a proposed transition from $z$ to $f(z)$).
+Let $\nu(dz) := \pi(dz) + (\pi \circ f^{-1})(dz)$.
+Let $h(z)$ be a density for $\pi$ with respect to $\nu$.
+Let $A := \{z \in Z : h(z) > 0 \mbox{ and } h(f(z)) > 0\}$.
+$k$ satisfies detailed balance with respect to $\pi$ if and only if:
+\begin{enumerate}
+\item $\alpha(z, f(z)) = 0$ for $\pi$-almost all $z \not \in A$
+\item $\alpha(z, f(z)) \frac{h(z)}{h(f(z))} = \alpha(f(z), z)$
+\end{enumerate}
+\end{lemma}
+
+Now we apply Lemma~\ref{lemma:tierney-involution} to our setting where $\pi$ is $\sigma$-finite, there exists a $\sigma$-finite reference measure $\mu$ for measurable space $(Z, \Sigma)$ such that $\pi$ is absolutely continuous with respect to $\mu$, and where the pushforward of $\mu$ by $f$, denoted $\mu \circ f^{-1}$, is absolutely continuous with respect to $\mu$.
+(One step in the proof---Equation~\ref{eq:mu-is-absolutely-continuous-with-respect-to-nu}---will also require that $\mu$ is also absolutely continuous with respect to $\pi$, which is possible if we restrict $Z$ to the support of $\pi$, $\{z : (d \pi / d \mu)(z) > 0\}$, which must be in the $\sigma$-algebra $\Sigma$).
+
+In our setting, $\alpha$ is defined as:
+\begin{equation}
+\alpha(z, f(z)) := \mbox{min}\left\{1, \frac{\frac{d \pi}{d \mu}(f(z))}{\frac{d \pi}{d \mu}(z)} \cdot \frac{d (\mu \circ f^{-1})}{d \mu}(z)\right\}
+\end{equation}
+This definition of $\alpha$ satisfies:
+\begin{equation}
+\alpha(z, f(z)) \frac{\frac{d \pi}{d \mu}(z)}{\frac{d \pi}{d \mu}(f(z))} \cdot \left( \frac{d (\mu \circ f^{-1})}{d \mu} (z)\right)^{-1}= \alpha(f(z), z)
+%\frac{\alpha(z, f(z))}{\alpha(f(z), z)} = \frac{\pi(T(z))}{\pi(z)} \left( \frac{d (\mu \circ T^{-1})}{d \mu} (z) \right)
+\end{equation}
+Therefore, to apply Lemma~\ref{lemma:tierney-involution}, it suffices to show $\pi$ has density with respect to $\nu$ (denoted $h(z)$) such that:
+\begin{equation}
+\frac{h(z)}{h(f(z))} = \frac{\frac{d \pi}{d \mu}(z)}{\frac{d \pi}{d \mu}(f(z))} \left( \frac{d (\mu \circ f^{-1})}{d \mu} (z)\right)^{-1}
+\end{equation}
+(TODO: be careful with the zero cases and support.)\\ % $ TODO
+Since $\pi$ and $\pi \circ f^{-1}$ are both absolutely continuous with respect to $\mu$, $\nu$ is also absolutely continuous with respect to $\mu$, and has density:
+\begin{equation}
+\frac{d \nu}{d \mu}(z) = \frac{d \pi}{d \mu}(z) + \frac{d (\pi \circ f^{-1})}{d \mu}(z) 
+\end{equation}
+$\pi$ is absolutely continuous with respect to $\nu$, and therefore:
+\begin{equation}
+\frac{d \pi}{d \mu}(z) = \frac{d \pi}{d \nu}(z) \cdot \frac{d \nu}{d \mu}(z)
+\end{equation}
+Therefore,
+\begin{equation}
+h(z)
+:= \frac{d \pi}{d \nu}(z)
+= \frac{\frac{d \pi}{d \mu}(z)}{\frac{d \nu}{d \mu}(z)}
+\;\; \mbox{ for } z \mbox{ s.t. } (d \nu / d \mu)(z) > 0
+\end{equation}
+Therefore:%(TODO: be careful with zeros)
+\begin{equation}
+\frac{h(z)}{h(f(z))} = \frac{\frac{d \pi}{d \mu}(z)}{\frac{d \pi}{d \mu}(f(z))} \cdot \frac{\frac{d \nu}{d \mu}(f(z))}{\frac{d \nu}{d \mu}(z)}
+\end{equation}
 It suffices to show that:
 \[
-\frac{\nu(T(z))}{\nu(z)} = \left( \frac{d (\mu \circ T^{-1})}{d \mu} \right)^{-1}
+\frac{\frac{d \nu}{d \mu}(f(z))}{\frac{d \nu}{d \mu}(z)} = \left(\frac{d (\mu \circ f^{-1})}{d \mu}(z)\right)^{-1}
 \]
-First, note that $\nu$ and $\nu \circ T^{-1}$ are the same measure:
+
+First, we prove a Lemma:
+\begin{lemma} \label{lemma:reference-measure-pushforward}
+If $(Z, \Sigma)$ is a measurable space and $f : Z \to Z$ is a measurable function that is an involution, $\nu$ and $\mu$ are $\sigma$-finite measures such that $\nu$ is absolutely continuous with respect to $\mu$, and such that the pushforward measures $\nu \circ f^{-1}$ and $\mu \circ f^{-1}$ are both $\sigma$-finite, then
+$\nu \circ f^{-1}$ is absolutely continuous with respect to $\mu \circ f^{-1}$ and
+\begin{equation}
+\frac{d (\nu \circ f^{-1})}{d (\mu \circ f^{-1})}(z) = \frac{d \nu}{d \mu}(f(z))
+\end{equation}
+\end{lemma}
+\begin{proof}
+First, $\nu \circ f^{-1}$ is absolutely continuous with respect to $\mu \circ f^{-1}$ because
+$(\mu \circ f^{-1})(A) = 0$ implies $\mu(f^{-1}(A)) = 0$ implies $\nu(f^{-1}(A)) = 0$ implies $(\nu \circ f^{-1})(A) = 0$.
+To show that $z \mapsto (d \nu / d \mu)(f(z))$ is the Radon-Nikodym derivative $d (\nu \circ f^{-1}) / d (\mu \circ f^{-1})$, it suffices to show that for all $A \in \Sigma$:
+\begin{equation}
+\int_A \left( \frac{d \nu}{d \mu} (f(z)) \right) ( \mu \circ f^{-1} )(dz) = (\nu \circ f^{-1})(A) := \nu(f^{-1}(A))
+\end{equation}
+Applying Theorem 3.6.1 in Bogachev with $Y := A$, $y := z$, $x := z'$, $X := f^{-1}(A)$, and $g(y) := (d\nu / d\mu)(f(y))$:
 \begin{align}
-\nu(A) &= \pi(A) + \pi'(A) = \pi(A) + \pi(T^{-1}(A))\\
-(\nu \circ T^{-1})(A) &= \pi(T^{-1}(A)) + \pi(T^{-1}(T^{-1}(A))) = \pi(T^{-1}(A)) + \pi(A)
-\end{align}
-Therefore,
-\[
-\frac{d (\nu \circ T^{-1})}{d \nu}(z) = 1 \mbox{ for all } z
-\]
-Expanding $d(\nu \circ T^{-1}) / d \nu$ using the chain rule:
-\begin{align}
-1 &= \frac{d (\nu \circ T^{-1})}{d \nu}(z)\\
-&=   \frac{d (\nu \circ T^{-1})}{d (\mu \circ T^{-1})}(z) \cdot
-    \frac{d (\mu \circ T^{-1})}{d \mu}(z) \cdot
-    \frac{d \mu}{d \nu}(z)\\
-&= \nu(T(z))  \frac{d (\mu \circ T^{-1})}{d \mu} \frac{1}{\nu(z)}
-\end{align}
-where in the second step we used that
-\begin{equation}
-\frac{d (\nu \circ T^{-1})}{d (\mu \circ T^{-1})}(z) = \frac{d \nu}{d \mu}(T(z)) = \nu(T(z))
-\end{equation}
-To show this, we use the definition of the Radon-Nikodym derivative $d (\nu \circ T^{-1}) / d (\mu \circ T^{-1})$.
-To show that $z \mapsto (d \nu / d \mu)(T(z))$ is this Radon-Nikodym derivative, it suffices to show that for all measurable $A$:
-\begin{equation}
-\int_A \left( \frac{d \nu}{d \mu} (T(z)) \right) ( \mu \circ T^{-1} )(dz) = (\nu \circ T^{-1})(A) := \nu(T^{-1}(A))
-\end{equation}
-Applying Theorem 3.6.1 in Bogachev with $Y := A$, $y := z$, $x := z'$, $X := T^{-1}(A)$, $f := T$, and $g(y) := (d\nu / d\mu)(T(y))$:
-\begin{align}
-\int_A \left( \frac{d \nu}{d \mu} (T(z)) \right) ( \mu \circ T^{-1} )(dz)
+\int_A \left( \frac{d \nu}{d \mu} (f(z)) \right) ( \mu \circ f^{-1} )(dz)
 &= \int_Y g(y) ( \mu \circ f^{-1} )(dy)\\
 &= \int_X g(f(x)) \mu(dx) \;\;\;\; [\mbox{Bogachev Theorem 3.6.1}]\\
-&= \int_{T^{-1}(A)} (d\nu / d\mu)(T(T(z'))) \mu(dz')\\
-&= \int_{T^{-1}(A)} (d\nu / d\mu)(z') \mu(dz')\\
-&= \int_{T^{-1}(A)} \nu(dz')
-= \nu(T^{-1}(A))
+&= \int_{f^{-1}(A)} (d\nu / d\mu)(f(f(z'))) \mu(dz')\\
+&= \int_{f^{-1}(A)} (d\nu / d\mu)(z') \mu(dz')\\
+&= \int_{f^{-1}(A)} \nu(dz')
+= \nu(f^{-1}(A))
 \end{align}
@@
 
-\subsection{Proof of stationarity for involution}
+\end{proof}
+
+Now, note that $\nu$ and $\nu \circ f^{-1}$ are the same measure:
+\begin{align}
+\nu(A) &= \pi(A) + \pi'(A) = \pi(A) + \pi(f^{-1}(A))\\
+(\nu \circ f^{-1})(A) &= \pi(f^{-1}(A)) + \pi(f^{-1}(f^{-1}(A))) = \pi(f^{-1}(A)) + \pi(A)
+\end{align}
+Therefore,
+\[
+\frac{d (\nu \circ f^{-1})}{d \nu}(z) = 1 \mbox{ for all } z
+\]
+Expanding $d(\nu \circ f^{-1}) / d \nu$ using the chain rule:
+\begin{align}
+1 &= \frac{d (\nu \circ f^{-1})}{d \nu}(z)\\
+&=   \frac{d (\nu \circ f^{-1})}{d (\mu \circ f^{-1})}(z) \cdot
+    \frac{d (\mu \circ f^{-1})}{d \mu}(z) \cdot
+    \frac{d \mu}{d \nu}(z) \label{eq:mu-is-absolutely-continuous-with-respect-to-nu}\\
+&= \frac{d \nu}{d \mu}(f(z)) \cdot \frac{d (\mu \circ f^{-1})}{d \mu}(z) \cdot \frac{d \mu}{d \mu}(z) \;\;\;\; [\mbox{Lemma~\ref{lemma:reference-measure-pushforward}}]\\
+&= \frac{d \nu}{d \mu}(f(z)) \cdot \frac{d (\mu \circ f^{-1})}{d \mu}(z) \cdot \frac{1}{\frac{d \nu}{d \mu}(z)}\\
+\left(\frac{d (\mu \circ f^{-1})}{d \mu}(z)\right)^{-1} &= \frac{\frac{d \nu}{d \mu}(f(z))}{\frac{d \nu}{d \mu}(z)}
+\end{align}
+
+\subsection{Proof of stationarity for involution} \label{sec:involution-is-stationary}
 Stationarity follows from detailed balance by setting $B := Z$ in the detailed balance equation, and using $k'_z(Z) = \alpha(z) + (1 - \alpha(z)) =1$, so the right-hand side simplifies to $\int_A \pi(dz) = \pi(A)$.
 
-\subsection{Proof of stationarity for end-to-end kernel}
+\subsection{Proof of stationarity for end-to-end kernel} \label{sec:involutive-mcmc-is-stationary}
 We are given that the involution is stationary with respect to (the measure induced by) $\pi(x, u) := p(x) q_x(u)$:

### 2020-07-11 paper.tex
checkpoint on probabilistic programming with traces

@@
 \section{INVOLUTIVE MCMC}
+
+% problem: what if the return value of p determines the control flow of q_x?
+% then the reference measure on 
+%Let $(Z,
+
 Let $p$ denote the target probability density with respect to some $\sigma$-finite measure $\mu_X$ on a set $X$.
@@
 
-%\begin{lemma}\label{lemma:involution-detailed-balance} % TODO cite
-%For $p$, $q$, and $f$ satisfying the properties above, let $k'$ denote the following transition kernel on $Z$:
-%\[
-%k'_z(A) := [f(z) \in A] \alpha(z) + [z \in A](1 - \alpha(z))
-%\]
-%The transition kernel $k'$ on $Z$ defined by the involution $f$ satisfies detailed balance with respect to $\pi$:
-%\[
-%\int_B k'_z(A) \pi(dz) = \int_A k'_z(B) \pi(dz) \mbox{ for all } A, B \in \Sigma_Z
-%\]
-%And in particular it is stationary with respect to $\pi$:
-%\[
-%\int_Z k'_z(A) \pi(dz) = \pi(A) \mbox{ for all } A \in \Sigma_Z
-%\]
-%\end{lemma}
-%\begin{proof}
-%\end{proof}
-%
-%\begin{theorem}\label{theorem:stationarity}
-%The transition kernel $k$ on $X$ defined by Algorithm~\ref{alg:involutive-mcmc} for valid $p$, $q$, and $f$:
-%\[
-%k_x(A) := \int_U k'_{x,u}(A \times U) q_x(u) \mu_U(du) \;\; \mbox{for all} \;\; A \in \Sigma_X, x \in X
-%\]
-%is stationary with respect to $p$.
-%\end{theorem}
-%\begin{proof}
-%By Lemma~\citep{lemma:involution-detailed-balance}, the transition kernel $k'$ is stationary with respect to $\pi$:
-%\[
-%%\int_B k'_{z}(B) \pi(z) \mu(dz) = \pi(B) \mbox{ for all } z \in X \times U, B \in \Sigma_Z
-%\]
-%It suffices to show that
-%\begin{align*}
-%\int_X k_x(A) p(x) \mu_X(dx) &= p(A) \mbox{ for all } A \in \Sigma_X\\
-%\int_X \left( \int_U k'_{x,u}(A \times U) q_x(u) \mu_U(du) \right) p(x) \mu_X(dx) &= p(A)\\
-%\int_{X \times U} k'_{x,u}(A \times U) q_x(u) p(x) \mu(dz) &= p(A)
-%\end{align*}
-%By Lemma~\citep{lemma:involution-detailed-balance}, setting $B := X \times U$, the left-hand-side is:
-%\[
-%\pi(X \times U) = \int_X \int_U p(x) q_x(u) \mu(
-%\]
-%\end{proof}
-%\paragraph{Detailed balance of the involution}
-%%Let $q_
-%
-%\paragraph{Stationarity}
-%
-%\paragraph{Conditional distribution and observations fixed}
+\section{PROBABILISTIC PROGRAMMING WITH TRACES}
+
+Probabilistic programs are programs that define probability distributions.
+Probabilistic programs can be used to define standard joint probability distributions over vectors of random-variables, as well as more complex probability distributions with uncertainty over the structure in which the set of random variables itself random.
+Involutive MCMC can be applied for target probability distributions on non-standard reference measures, including models with structure uncertainty.
+For example, reversible jump MCMC, which is a special case of involutive MCMC, can be applied to switch between different model structures.
+This section describes a framework for probabilistic programming that is expressive enough to represent probability distributions with structure uncertainty.
+
+\paragraph{Traces}
+In our framework, a probabilistic program defines a probability distribution on finite dictionaries that map the name of a `random choice' to its value.
+Following standard terminology in probabilistic programming~\citep{?}, we call these dictionaries \emph{traces} because they can be sampled by running a probabilistic program, and `tracing' the execution the program by recording the value of eacn random choice encountered.
+In the probabilistic programming language used in this paper, the user assigns a unique name or \emph{address} to each random choice using a special syntax:
+
+TODO: include example of tilde syntax
+
+TODO: introduce the mathematical notation for a trace
+
+\paragraph{A probabilistic program encodes a probability distribution on traces}
+For probabilistic programs that only sample individual random choices from discrete probability distributions, the meaning of the program can be described by a probability mass function that assigns a probailtiy to every possible trace.
+
+TODO: Give a simple example of a probabilistic program with discrete random chocies and stochastic control flow, and the resulting probabiilty distribution.
+TODO: describe the two variants of tilde syntax
+
+Note that traces are always finite, but there can be no upper bound.
+For example, for this program, there is a countably infinite number of possible traces:
+
+TODO: show a discrete probabilistic prgram with the table of traces it could sample
+
+% measure version
+\paragraph{Handling continuous distributions}
+For programs that can sample from continuous probability distributions, the formalism needs to be extended.
+pecifically, we define the $\mathcal{S}$ to be the set of all possible \emph{trace structures} as a (possibly countably infinite) set of finite sets of addresses, denoted $\mathcal{S} \subseteq P(\mathcal{A})$, where $\mathcal{A}$ is the set of all addresses.
+For each address $a \in \mathcal{A}$ we assign a reference measure space $(X_a, \Sigma_a, \mu_a)$ where $\mu_a$ is $\sigma$-finite.
+The set of all traces of a probabilistic program is the set $\mathcal{T} := \cup_{S \in \mathcal{S}} \{(S, \mathbf{x}) : \mathbf{x} \in \times_{a \in S} X_a\}\}$.
+The \emph{reference measure space} for a probabilistic program is $(\mathcal{T}, \Sigma, \mu)$ where
+$\Sigma$ is the $\sigma$-algebra that contains sets $\cup_{S \in \mathcal{S}}\{ B_S \} \in \mathcal{P}(\mathcal{T})$ such that $B_S$ is an element of the product $\sigma$-algebra $\Sigma_S := \otimes_{a \in S} \Sigma_a$,
+and where $\mu(\cup_{S \in \mathcal{S}}\{ B_S \}) := \sum_{S \in \mathcal{S}} \mu_S(B_S)$ where $\mu_S$ is the product measure $\times_{a \in S} \mu_a$.
+
+We assume that a probabilistic program denotes a reference measure space $\mu$ and a density $p(t)$ on traces $t$ with respect to the reference measure, such that $p$ and $\mu$ are mutually absolutely continuous.
+We do not give the denotational semantics of a probabilistic programming language with this property, but instead refer to POPL paper for an example of a denotational semantics.
+% TODO what about requirement that the support of each random choice is constant?
+(this is implied by the definition of reference measure)
+
+TODO: Give an example of a probabilistic program in Gen's dynamic modeling language that includes continuous random choices, and the vbase measure and a density for it.
+
+% step 1: *assert* that the composition of p and then q defines a probabilistic program with ref meas mu and mut. abs. cont. pi(z) [ok] (( do this without ascribing separate semantics to p and q?? ))
+% step 2: show detailed balance on the joint space ... [ok]
+% step 3: show stationarity on the joint space [ok, easy from detailed balance]
+% step 4: show stationarity with respect to space of p.... ((requires ascribing semantics to p, but maybe not q??)
+% - this would be straightforward for a product space, but our space is more complicated?
+% - need to assert relationship between p and pi, ...
+
+\[
+p(t) \;\; \pi(z)
+\]
+
+\paragraph{Extending one probabilistic program with another}
+Given a
+
+% the only place where this matters is in the 'gibbs move' argument.. (8.4)
+
+We need $p$ to be a density with respect to some reference measure space $(T_P, \Sigma_{P}, \mu_P)$.
+We need $q_t$ to be a density with respect to some reference measure $\mu_{Q,t}$ for each $t \in T_P$.
+Such that the function $\pi(t, u) := p(x) q_t(u)$ is a density $\pi(x, u)$ with respect to some reference measure $\mu$, and $\pi$ is mutually absolutely continuous with $\mu$.
+(this can be constructed by using hierarchical addresses `model' and `proposal' for the random choics made in the two programs).
+
+\paragraph{Conditioning}
+
+\section{DIFFERENTIABLE PROGRAMMING WITH TRACES}
 
@@
 \section{AUTOMATIC ACCEPTANCE PROBABILITY CALCULATION}
-
-\subsection{Exploiting Jacobian Sparsity}
 
@@
 We reproduce the result in our notation:
-\begin{lemma}[Detailed balance for involution move~\citep{tierney1998note}]
+\begin{lemma}[Detailed balance for involution move~\citep{tierney1998note}] \label{lemma:tierney-involution}
 Let $(Z, \Sigma, \pi)$ denote a measure space.
@@
 
+\paragraph{Proof of stationarity}
+We need the composition of the two programs to define a valid probabilistic program with density $\pi$ that is absolutely continuous with respect to the refernce measure $\mu$.
+We also need there to be a particular relationship between $\pi$ and $p$.
+The involution part of the kernel is a kernel on the combined space $(Z, \Sigma_Z, \mu)$.
+\[
+k'_z(B_Z) = [f(z) \in B_Z] \alpha(z) + [z \in B_Z] (1 - \alpha(z))
+\]
+The kernel of Algorithm 1 is defined by:
+\[
+k_t(B_T) := \int_{U_t} q_{t}(u) k'_{(t, u)}(\{(t', u') \in Z : t' \in B_T\}) \mu_{Q,t}(du)
+\]
+We want to show it is stationary:
+\begin{align}
+\int_T k_t(B_T) p(t) \mu_P(dt) &= p(B_T) \mbox{ for all } B_T \in \Sigma_P\\
+\int_T \left( \int_{U_t} q_{t}(u) k'_{(t, u)}(\{(t', u') \in Z : t' \in B_T\}) \mu_{Q,t}(du) \right) p(t) \mu_P(dt) &= p(B_T)\\
+\sum_{i} \int_{T_C} \left( \int_{U_t} q_{t}(u) k'_{(t, u)}(\{(t', u') \in Z : t' \in B_T\}) \mu_{Q,t}(du) \right) p(i, t_c) \mu_{P_C,i}(dt_c) &= p(B_T)\\ % separate out discrete part of t (i) and continuous part of t (t_c)
+\sum_{i} \int_{T_C} \left( \int_{U_i} q_{t}(u) k'_{(t, u)}(\{(t', u') \in Z : t' \in B_T\}) \mu_{Q,i}(du) \right) p(i, t_c) \mu_{P_C,i}(dt_c) &= p(B_T)\\ % use that the reference measure mu_{Q,t} only depends on the discrete part
+..\\
+\int_{Z} \pi(t, u) k'_{t,u}(\{(t', u') \in Z : t' \in B_T\}) \mu(dz) &= p(B_T)\\
+\pi(\{(t', u') \in Z : t' \in B_T\}) &= p(B_T)\\
+\end{align}
+
+the set of addresses $\mathcal{A}$ is partitioned into two spaces $\mathcal{A}_1$ and $\mathcal{A}_2$\\
+
+each structure $S$ partitions into two structures $S_1$ and $S_2$
+
+then $Z := \cup_{S \in \mathcal{S}} \{(S, \mathbf{x}) : \mathbf{x} \in \times_{a \in S} X_a \}$\\
+
+Space for $Z$:
+\[
+\cup_{S \in \mathcal{S}} \{((\mathcal{A}_1 \cap S) \cup (\mathcal{A}_2 \cap S), \mathbf{x_1}, \mathbf{x_2}) : \mathbf{x_1} \in \times_{a \in S_1} X_a, \mathbf{x_2} \in \times_{a \in S_2} X_a\}
+\]
+
+Space for $T$:
+\[
+\cup_{S_1 \in \mathcal{S}} \{(S_1, \mathbf{x_1}) : \mathbf{x_1} \in \times_{a \in S_1}\}
+\]
+
+\begin{align}
+\pi(
+\cup_{S \in \mathcal{S} : A_1 \cap S = S_1} \{(S_1 \cup (\mathcal{A}_2 \cap S), \mathbf{x_1}, \mathbf{x_2}) : \mathbf{x_1} \in B_{S_1}, \mathbf{x_2} \in \times_{a \in S_2} X_a\}
+) &= p(B_T)\\
+\sum_{S \in \mathcal{S} : A_1 \cap S = S_1} \int_{B_{S_1} \times \otimes_{a \in S_2} X_a} \pi(S_1, S_2, \mathbf{x}_1, \mathbf{x}_2) \mu_{S_1 \cup S_2}(d \mathbf{x}_1 \mathbf{x}_2) &= p(B_T)\\
+\sum_{S \in \mathcal{S} : A_1 \cap S = S_1} \int_{B_{S_1} \times \otimes_{a \in S_2} X_a} p(S_1, \mathbf{x}_1) q_{(S_1,\mathbf{x}_1)}(S_2, \mathbf{x}_2) \mu_{S_1 \cup S_2}(d \mathbf{x}_1 \mathbf{x}_2) &= p(B_T)\\
+\sum_{S \in \mathcal{S} : A_1 \cap S = S_1} \int_{B_{S_1}} p(S_1, \mathbf{x}_1) \left( \int_{\otimes_{a \in S_2 X_a}} q_{(S_1,\mathbf{x}_1)}(S_2, \mathbf{x}_2) \mu_{S_2}(d \mathbf{x}_2) \right) \mu_{S_1}(d \mathbf{x}_1) &= p(B_T) [\mbox{Fubini / Tonnelli?}]\\
+\sum_{S_2 \subseteq A_2 : S_1 \cup S_2 \in \mathcal{S}} \int_{B_{S_1}} p(S_1, \mathbf{x}_1) \left( \int_{\otimes_{a \in S_2 X_a}} q_{(S_1,\mathbf{x}_1)}(S_2, \mathbf{x}_2) \mu_{S_2}(d \mathbf{x}_2) \right) \mu_{S_1}(d \mathbf{x}_1) &= p(B_T)\\
+\int_{B_{S_1}} p(S_1, \mathbf{x}_1) \left( \sum_{S_2 \subseteq A_2 : S_1 \cup S_2 \in \mathcal{S}} \int_{\otimes_{a \in S_2 X_a}} q_{(S_1,\mathbf{x}_1)}(S_2, \mathbf{x}_2) \mu_{S_2}(d \mathbf{x}_2) \right) \mu_{S_1}(d \mathbf{x}_1) &= p(B_T) [\mbox{Fubini / Tonnelli?}]\\
+\end{align}
+It suffices to show that
+\begin{align}
+\sum_{S_2 \subseteq A_2 : S_1 \cup S_2 \in \mathcal{S}} \int_{\otimes_{a \in S_2 X_a}} q_{(S_1,\mathbf{x}_1)}(S_2, \mathbf{x}_2) \mu_{S_2}(d \mathbf{x}_2) = 1 \mbox{ for all } (S_1, \mathbf{x}_1) \in T
+\end{align}
+It suffices to show that for each $(S_1, \mathbf{x}_1) \in T$, we have that $q_{(S_1, \mathbf{x}_1)}$ is the density of a (terminating with probability one) probabilistic program with structruers $\mathcal{S}_2$ such that
+\[
+\mathcal{S}_2 = \{S_2 \subseteq A_2 : S_1 \cup S_2 \in \mathcal{S}\}
+\]
+
+\[
+B_1 = \cup_{S_1 \in \mathcal{S}_1} \{(S_1, \mathbf{x}_1) : \mathbf{x}_1 \in B_{S_1}\}\;\;\; \mbox{where} \;\; B_{S_1} \in \otimes_{a \in S_1} \Sigma_{a} \;\;\; [\mbox{Measurable set } B_1]
+\]
+% TODO the proof above should have a sum over S_1
+
+% NOTE for each $t$ there is a different reference measure...
+% we need to restrict q so that for each discrete part of $t$, the reference measure is the same
+% (which implies that the same sets of random choices are made?)
+% that will allow us to
+
 \end{document}

### 2020-07-11 paper.tex
checkpt

@@
 
-% problem: what if the return value of p determines the control flow of q_x?
-% then the reference measure on 
-%Let $(Z,
+Let $(X, \Sigma_P, \mu_P)$ and $(Y, \Sigma_Q, \mu_Q)$ denote two measure spaces with $\sigma$-finite $\mu_P$ and $\mu_Q$.
+Let $p : X \to [0, \infty)$ be a probability density with respect to $\mu_P$ and for each $t \in X$ such that $p(t) > 0$ let $q_t : Y \to [0, \infty)$ be probability density with respect to $\mu_Q$.
+We call $p$ the \emph{target density} and we call $q_t$ \emph{auxiliary densities}.
+Let $\pi(t, u) := p(t) q_t(u)$ and $Z := \{(t, u) : \pi(t, u) > 0 \}$.
+Suppose $Z \in \Sigma_P \otimes \Sigma_Q$ (the product $\sigma$-algebra), so that $\Sigma := \{A \in \Sigma_P \otimes \Sigma_Q : A \subseteq Z\}$ is a $\sigma$-algebra on $Z$ and let $\mu : \Sigma \to [0, \infty)$ with $\mu(A) := (\mu_P \times \mu_Q)(A)$
+(note that $\mu$ is $\sigma$-finite since $\mu_P \times \mu_Q$ is $\sigma$-finite. % TODO: prove this 
+Suppose that $\pi$ is a probability density with respect to $\mu$. % TODO is this guaranteed?
 
-Let $p$ denote the target probability density with respect to some $\sigma$-finite measure $\mu_X$ on a set $X$.
-For each $x$ such that $p(x) > 0$, let $q_x$ denote a probability density with respect to some $\sigma$-finite measure $\mu_U$ on a set $U$.
-Let $\mu := \mu_X \times \mu_U$ denote the product measure on the space $X \times U$, and let $\pi(x, u) := p(x) q_x(u)$ and $Z := \{(x, u) : \pi(x, u) > 0\}$.
 Let $f : Z \to Z$ denote an involution ($f^{-1} = f$) such that the pushforward of $\mu$ under $f$, denoted $\mu \circ f^{-1}$, is absolutely continuous with respect to $\mu$, with Radon-Nikodym derivative $d (\mu \circ f^{-1}) / d\mu : Z \to [0, \infty)$.
+Let $u \sim q_t(\cdot)$ denote a random sample from the measure defined by $A \mapsto \int_A q_t(u) \mu_Q(du)$ for $A \in \Sigma_Q$.
 
@@
 \begin{algorithmic}
-\Procedure{involutive-mcmc-move}{$p$, $q$, $f$, $x$}
-    \State $u \sim q(\cdot; x)$ \Comment{Sample auxiliary variables}
+\Procedure{involutive-mcmc-move}{$p$, $q$, $f$, $t$}
+    \State $u \sim q_t(\cdot)$ \Comment{Sample auxiliary variables}
     \State $(x', u') \gets f(x, u)$ \Comment{Apply involution}
@@
 For programs that can sample from continuous probability distributions, the formalism needs to be extended.
-pecifically, we define the $\mathcal{S}$ to be the set of all possible \emph{trace structures} as a (possibly countably infinite) set of finite sets of addresses, denoted $\mathcal{S} \subseteq P(\mathcal{A})$, where $\mathcal{A}$ is the set of all addresses.
+pecifically, we define the $\mathcal{S}$ to be the set of all possible \emph{trace structures} as a (possibly countably infinite) set of finite sets of addresses, denoted $\mathcal{S} \subseteq \mathcal{P}(\mathcal{A})$, where $\mathcal{A}$ is the set of all addresses.
 For each address $a \in \mathcal{A}$ we assign a reference measure space $(X_a, \Sigma_a, \mu_a)$ where $\mu_a$ is $\sigma$-finite.
-The set of all traces of a probabilistic program is the set $\mathcal{T} := \cup_{S \in \mathcal{S}} \{(S, \mathbf{x}) : \mathbf{x} \in \times_{a \in S} X_a\}\}$.
-The \emph{reference measure space} for a probabilistic program is $(\mathcal{T}, \Sigma, \mu)$ where
-$\Sigma$ is the $\sigma$-algebra that contains sets $\cup_{S \in \mathcal{S}}\{ B_S \} \in \mathcal{P}(\mathcal{T})$ such that $B_S$ is an element of the product $\sigma$-algebra $\Sigma_S := \otimes_{a \in S} \Sigma_a$,
+The set of all traces of a probabilistic program is the set $T := \cup_{S \in \mathcal{S}} \{(S, \mathbf{x}) : \mathbf{x} \in \times_{a \in S} X_a\}\}$.
+The \emph{reference measure space} for a probabilistic program is $(T, \Sigma, \mu)$ where
+$\Sigma$ is the $\sigma$-algebra that contains sets $\cup_{S \in \mathcal{S}}\{ B_S \} \in \mathcal{P}(T)$ such that $B_S$ is an element of the product $\sigma$-algebra $\Sigma_S := \otimes_{a \in S} \Sigma_a$,
 and where $\mu(\cup_{S \in \mathcal{S}}\{ B_S \}) := \sum_{S \in \mathcal{S}} \mu_S(B_S)$ where $\mu_S$ is the product measure $\times_{a \in S} \mu_a$.
 
-We assume that a probabilistic program denotes a reference measure space $\mu$ and a density $p(t)$ on traces $t$ with respect to the reference measure, such that $p$ and $\mu$ are mutually absolutely continuous.
+We assume that a probabilistic program $P$ denotes a reference measure space $(T_P, \Sigma_P, \mu_P)$ and a measure $P(dt)$ that is mutually absolutely continuous with respect to $\mu_P$, with density
+$p : T \to [0, \infty)$ on traces $t \in T$.
 We do not give the denotational semantics of a probabilistic programming language with this property, but instead refer to POPL paper for an example of a denotational semantics.
@@
 \paragraph{Extending one probabilistic program with another}
+A second probabilistic program that is parametrized by (i.e. takes as input) traces of the first probabilistic program $P$.
+The reference measure space of the second probabilistic program for $t \in T_P$ is denoted $(T_{Q,t}, \Sigma_{Q,t}, \mu_{P,t})$.
+We define $Z := \{(t, u) : t \in T_P, u \in T_{Q,t}$ and we define $\pi(z) := p(t) q_t(u)$.
+We require that $(Z, \Sigma, \mu)$ is a probabilistic program reference measure and $\pi$ is a probability density with respect to $\mu$ that is mutually absolutely continuous.
+The combined program is $P \circ Q$.
+
 Given a
@@
 
-Now we apply Lemma~\ref{lemma:tierney-involution} to our setting where $\pi$ is $\sigma$-finite, there exists a $\sigma$-finite reference measure $\mu$ for measurable space $(Z, \Sigma)$ such that $\pi$ is absolutely continuous with respect to $\mu$, and where the pushforward of $\mu$ by $f$, denoted $\mu \circ f^{-1}$, is absolutely continuous with respect to $\mu$.
-(One step in the proof---Equation~\ref{eq:mu-is-absolutely-continuous-with-respect-to-nu}---will also require that $\mu$ is also absolutely continuous with respect to $\pi$, which is possible if we restrict $Z$ to the support of $\pi$, $\{z : (d \pi / d \mu)(z) > 0\}$, which must be in the $\sigma$-algebra $\Sigma$).
+Now we apply Lemma~\ref{lemma:tierney-involution} to our setting where $\pi$ is $\sigma$-finite, there exists a $\sigma$-finite reference measure $\mu$ for measurable space $(Z, \Sigma)$ such that $\pi$ is mutually absolutely continuous with respect to $\mu$, and where the pushforward of $\mu$ by $f$, denoted $\mu \circ f^{-1}$, is absolutely continuous with respect to $\mu$.
+%(One step in the proof---Equation~\ref{eq:mu-is-absolutely-continuous-with-respect-to-nu}---will also require that $\mu$ is also absolutely continuous with respect to $\pi$, which is possible if we restrict $Z$ to the support of $\pi$, $\{z : (d \pi / d \mu)(z) > 0\}$, which must be in the $\sigma$-algebra $\Sigma$).
 
@@
 \end{equation}
-(TODO: be careful with the zero cases and support.)\\ % $ TODO
 Since $\pi$ and $\pi \circ f^{-1}$ are both absolutely continuous with respect to $\mu$, $\nu$ is also absolutely continuous with respect to $\mu$, and has density:
@@
 \end{equation}
+Because $(d \pi / d \mu)(z) > 0$ for all $z \in Z$, $(d \nu / d \mu)(z) > 0$ for all $z \in Z$.
 Therefore,
@@
 = \frac{\frac{d \pi}{d \mu}(z)}{\frac{d \nu}{d \mu}(z)}
-\;\; \mbox{ for } z \mbox{ s.t. } (d \nu / d \mu)(z) > 0
 \end{equation}
-Therefore:%(TODO: be careful with zeros)
+Therefore:
 \begin{equation}
@@
 \subsection{Proof of stationarity for involution} \label{sec:involution-is-stationary}
-Stationarity follows from detailed balance by setting $B := Z$ in the detailed balance equation, and using $k'_z(Z) = \alpha(z) + (1 - \alpha(z)) =1$, so the right-hand side simplifies to $\int_A \pi(dz) = \pi(A)$.
+Detailed balance of the the involution kernel with respect to the measure induced by $\pi$ implies:
+\begin{equation}
+\int_B \pi(z) k'_z(A) \mu(dz) = \int_A \pi(z) k'_z(B) \mu(dz) \;\; \mbox{ for all } A, B \in \Sigma
+\end{equation}
+Stationarity with respect to $\pi$ follows by substituting $Z$ for $B$:
+\begin{align}
+\int_Z \pi(z) k'_z(A) \mu(dz) &= \int_A \pi(z) k'_z(Z) \mu(dz) = \int_A \pi(z) \mu(dz) \;\; \mbox{ for all } A \in \Sigma\\
+\end{align}
 
@@
 We are given that the involution is stationary with respect to (the measure induced by) $\pi(x, u) := p(x) q_x(u)$:
-\[
-\int_Z k'_z(A) \pi(z) \mu(dz) = \pi(A) \mbox{ for all } A \in \Sigma_Z
-\]
-It suffices to show that the end-to-end kernel $k$ is stationary with respect to (the measure induced by) $p(x)$:
-\[
-\int_X k_x(A) p(x) \mu_X(dx) = p(A) \mbox{ for all } A \in \Sigma_X
-\]
-The kernel $k$ on $X$ defined by Algorithm~\ref{alg:involutive-mcmc} for valid $p$, $q$, and $f$:
-\[
-k_x(A) := \int_U k'_{x,u}(A \times U) q_x(u) \mu_U(du) \;\; \mbox{for all} \;\; A \in \Sigma_X, x \in X
-\]
-is stationary with respect to $p$.
+\begin{equation}
+\int_Z k'_z(A) \pi(z) \mu(dz) = \int_A \pi(z) \mu(dz) \mbox{ for all } A \in \Sigma
+\end{equation}
+The end-to-end kernel is defined fir all $x \in X_P$ such that $p(x) > 0$ as:
+\begin{equation}
+k_x(A) := \int_U k'_{x,u}(A \times U) q_x(u) \mu_U(du) \;\; \mbox{for all} \;\; A \in \Sigma_P, x \in X
+\end{equation}
+Stationarity of the end-to-end kernel with respect to the measure induced by $p$ is:
+\begin{equation}
+\int_X k_x(A) p(x) \mu_P(dx) = \int_A p(x) \mu_P(dx) \mbox{ for all } A \in \Sigma_P
+\end{equation}
+Expanding:
 \begin{align*}
-    \int_X k_x(A) p(x) \mu_X(dx) &= p(A) \mbox{ for all } A \in \Sigma_X\\
-    \int_X \left( \int_U k'_{x,u}(A \times U) q_x(u) \mu_U(du) \right) p(x) \mu_X(dx) &= p(A)\\
-    \int_{X \times U} k'_{x,u}(A \times U) q_x(u) p(x) \mu(dz) &= p(A)\\
-    \int_{X \times U} k'_{x,u}(A \times U) \pi(z) \mu(dz) &= p(A)\\
-    \pi(A \times U) &= p(A)\\
-    \int_{A \times U} q_x(u) p(x) \mu(dz) &= p(A)\\
-    \int_A \left( \int_U q_x(u) \mu_U(du) \right) p(x) \mu_X(dx) &= p(A)\\
-    \int_A p(x) \mu_X(dx) &= p(A)
+    \int_X k_x(A) p(x) \mu_P(dx) &= \int_X \left( \int_U k'_{x,u}(A \times U) q_x(u) \mu_Q(du) \right) p(x) \mu_P(dx)\\
+    &= \int_{X \times U} k'_{x,u}(A \times U) q_x(u) p(x) (\mu_P \times \mu_Q)(dz) \;\; [\mbox{ Tonelli's theorem }]\\ % TODO check
+    &= \int_{Z} k'_{x,u}(\{(x', u') \in Z : x' \in A\}) \pi(z) \mu(dz)\\ % TODO justify
+    &= \int_{\{(x, u) \in Z : x \in A\}} \pi(z) \mu(dz)\;\;\; [\mbox{ Stationarity of $k'$ with respect to $\pi$}]\\
+    &= \int_{\{(x, u) \in Z : x \in A\}} q_x(u) p(x) \mu(dz)\\
+    &= \int_A \left( \int_U q_x(u) \mu_Q(du) \right) p(x) \mu_P(dx)\\
+    &= \int_A p(x) \mu_P(dx)
 \end{align*}
@@
 
-\paragraph{Proof of stationarity}
-We need the composition of the two programs to define a valid probabilistic program with density $\pi$ that is absolutely continuous with respect to the refernce measure $\mu$.
-We also need there to be a particular relationship between $\pi$ and $p$.
-The involution part of the kernel is a kernel on the combined space $(Z, \Sigma_Z, \mu)$.
-\[
-k'_z(B_Z) = [f(z) \in B_Z] \alpha(z) + [z \in B_Z] (1 - \alpha(z))
-\]
-The kernel of Algorithm 1 is defined by:
-\[
-k_t(B_T) := \int_{U_t} q_{t}(u) k'_{(t, u)}(\{(t', u') \in Z : t' \in B_T\}) \mu_{Q,t}(du)
-\]
-We want to show it is stationary:
-\begin{align}
-\int_T k_t(B_T) p(t) \mu_P(dt) &= p(B_T) \mbox{ for all } B_T \in \Sigma_P\\
-\int_T \left( \int_{U_t} q_{t}(u) k'_{(t, u)}(\{(t', u') \in Z : t' \in B_T\}) \mu_{Q,t}(du) \right) p(t) \mu_P(dt) &= p(B_T)\\
-\sum_{i} \int_{T_C} \left( \int_{U_t} q_{t}(u) k'_{(t, u)}(\{(t', u') \in Z : t' \in B_T\}) \mu_{Q,t}(du) \right) p(i, t_c) \mu_{P_C,i}(dt_c) &= p(B_T)\\ % separate out discrete part of t (i) and continuous part of t (t_c)
-\sum_{i} \int_{T_C} \left( \int_{U_i} q_{t}(u) k'_{(t, u)}(\{(t', u') \in Z : t' \in B_T\}) \mu_{Q,i}(du) \right) p(i, t_c) \mu_{P_C,i}(dt_c) &= p(B_T)\\ % use that the reference measure mu_{Q,t} only depends on the discrete part
-..\\
-\int_{Z} \pi(t, u) k'_{t,u}(\{(t', u') \in Z : t' \in B_T\}) \mu(dz) &= p(B_T)\\
-\pi(\{(t', u') \in Z : t' \in B_T\}) &= p(B_T)\\
-\end{align}
-
-the set of addresses $\mathcal{A}$ is partitioned into two spaces $\mathcal{A}_1$ and $\mathcal{A}_2$\\
-
-each structure $S$ partitions into two structures $S_1$ and $S_2$
-
-then $Z := \cup_{S \in \mathcal{S}} \{(S, \mathbf{x}) : \mathbf{x} \in \times_{a \in S} X_a \}$\\
-
-Space for $Z$:
-\[
-\cup_{S \in \mathcal{S}} \{((\mathcal{A}_1 \cap S) \cup (\mathcal{A}_2 \cap S), \mathbf{x_1}, \mathbf{x_2}) : \mathbf{x_1} \in \times_{a \in S_1} X_a, \mathbf{x_2} \in \times_{a \in S_2} X_a\}
-\]
-
-Space for $T$:
-\[
-\cup_{S_1 \in \mathcal{S}} \{(S_1, \mathbf{x_1}) : \mathbf{x_1} \in \times_{a \in S_1}\}
-\]
-
-\begin{align}
-\pi(
-\cup_{S \in \mathcal{S} : A_1 \cap S = S_1} \{(S_1 \cup (\mathcal{A}_2 \cap S), \mathbf{x_1}, \mathbf{x_2}) : \mathbf{x_1} \in B_{S_1}, \mathbf{x_2} \in \times_{a \in S_2} X_a\}
-) &= p(B_T)\\
-\sum_{S \in \mathcal{S} : A_1 \cap S = S_1} \int_{B_{S_1} \times \otimes_{a \in S_2} X_a} \pi(S_1, S_2, \mathbf{x}_1, \mathbf{x}_2) \mu_{S_1 \cup S_2}(d \mathbf{x}_1 \mathbf{x}_2) &= p(B_T)\\
-\sum_{S \in \mathcal{S} : A_1 \cap S = S_1} \int_{B_{S_1} \times \otimes_{a \in S_2} X_a} p(S_1, \mathbf{x}_1) q_{(S_1,\mathbf{x}_1)}(S_2, \mathbf{x}_2) \mu_{S_1 \cup S_2}(d \mathbf{x}_1 \mathbf{x}_2) &= p(B_T)\\
-\sum_{S \in \mathcal{S} : A_1 \cap S = S_1} \int_{B_{S_1}} p(S_1, \mathbf{x}_1) \left( \int_{\otimes_{a \in S_2 X_a}} q_{(S_1,\mathbf{x}_1)}(S_2, \mathbf{x}_2) \mu_{S_2}(d \mathbf{x}_2) \right) \mu_{S_1}(d \mathbf{x}_1) &= p(B_T) [\mbox{Fubini / Tonnelli?}]\\
-\sum_{S_2 \subseteq A_2 : S_1 \cup S_2 \in \mathcal{S}} \int_{B_{S_1}} p(S_1, \mathbf{x}_1) \left( \int_{\otimes_{a \in S_2 X_a}} q_{(S_1,\mathbf{x}_1)}(S_2, \mathbf{x}_2) \mu_{S_2}(d \mathbf{x}_2) \right) \mu_{S_1}(d \mathbf{x}_1) &= p(B_T)\\
-\int_{B_{S_1}} p(S_1, \mathbf{x}_1) \left( \sum_{S_2 \subseteq A_2 : S_1 \cup S_2 \in \mathcal{S}} \int_{\otimes_{a \in S_2 X_a}} q_{(S_1,\mathbf{x}_1)}(S_2, \mathbf{x}_2) \mu_{S_2}(d \mathbf{x}_2) \right) \mu_{S_1}(d \mathbf{x}_1) &= p(B_T) [\mbox{Fubini / Tonnelli?}]\\
-\end{align}
-It suffices to show that
-\begin{align}
-\sum_{S_2 \subseteq A_2 : S_1 \cup S_2 \in \mathcal{S}} \int_{\otimes_{a \in S_2 X_a}} q_{(S_1,\mathbf{x}_1)}(S_2, \mathbf{x}_2) \mu_{S_2}(d \mathbf{x}_2) = 1 \mbox{ for all } (S_1, \mathbf{x}_1) \in T
-\end{align}
-It suffices to show that for each $(S_1, \mathbf{x}_1) \in T$, we have that $q_{(S_1, \mathbf{x}_1)}$ is the density of a (terminating with probability one) probabilistic program with structruers $\mathcal{S}_2$ such that
-\[
-\mathcal{S}_2 = \{S_2 \subseteq A_2 : S_1 \cup S_2 \in \mathcal{S}\}
-\]
-
-\[
-B_1 = \cup_{S_1 \in \mathcal{S}_1} \{(S_1, \mathbf{x}_1) : \mathbf{x}_1 \in B_{S_1}\}\;\;\; \mbox{where} \;\; B_{S_1} \in \otimes_{a \in S_1} \Sigma_{a} \;\;\; [\mbox{Measurable set } B_1]
-\]
-% TODO the proof above should have a sum over S_1
+%\paragraph{Proof of stationarity}
+%We need the composition of the two programs to define a valid probabilistic program with density $\pi$ that is absolutely continuous with respect to the refernce measure $\mu$.
+%We also need there to be a particular relationship between $\pi$ and $p$.
+%The involution part of the kernel is a kernel on the combined space $(Z, \Sigma_Z, \mu)$.
+%\[
+%k'_z(B_Z) = [f(z) \in B_Z] \alpha(z) + [z \in B_Z] (1 - \alpha(z))
+%\]
+%The kernel of Algorithm 1 is defined by:
+%\[
+%k_t(B_T) := \int_{U_t} q_{t}(u) k'_{(t, u)}(\{(t', u') \in Z : t' \in B_T\}) \mu_{Q,t}(du)
+%\]
+%We want to show it is stationary:
+%\begin{align}
+%\int_T k_t(B_T) p(t) \mu_P(dt) &= p(B_T) \mbox{ for all } B_T \in \Sigma_P\\
+%\int_T \left( \int_{U_t} q_{t}(u) k'_{(t, u)}(\{(t', u') \in Z : t' \in B_T\}) \mu_{Q,t}(du) \right) p(t) \mu_P(dt) &= p(B_T)\\
+%\sum_{i} \int_{T_C} \left( \int_{U_t} q_{t}(u) k'_{(t, u)}(\{(t', u') \in Z : t' \in B_T\}) \mu_{Q,t}(du) \right) p(i, t_c) \mu_{P_C,i}(dt_c) &= p(B_T)\\ % separate out discrete part of t (i) and continuous part of t (t_c)
+%\sum_{i} \int_{T_C} \left( \int_{U_i} q_{t}(u) k'_{(t, u)}(\{(t', u') \in Z : t' \in B_T\}) \mu_{Q,i}(du) \right) p(i, t_c) \mu_{P_C,i}(dt_c) &= p(B_T)\\ % use that the reference measure mu_{Q,t} only depends on the discrete part
+%..\\
+%\int_{Z} \pi(t, u) k'_{t,u}(\{(t', u') \in Z : t' \in B_T\}) \mu(dz) &= p(B_T)\\
+%\pi(\{(t', u') \in Z : t' \in B_T\}) &= p(B_T)\\
+%\end{align}
+%
+%the set of addresses $\mathcal{A}$ is partitioned into two spaces $\mathcal{A}_1$ and $\mathcal{A}_2$\\
+%
+%each structure $S$ partitions into two structures $S_1$ and $S_2$
+%
+%then $Z := \cup_{S \in \mathcal{S}} \{(S, \mathbf{x}) : \mathbf{x} \in \times_{a \in S} X_a \}$\\
+%
+%Space for $Z$:
+%\[
+%\cup_{S \in \mathcal{S}} \{((\mathcal{A}_1 \cap S) \cup (\mathcal{A}_2 \cap S), \mathbf{x_1}, \mathbf{x_2}) : \mathbf{x_1} \in \times_{a \in S_1} X_a, \mathbf{x_2} \in \times_{a \in S_2} X_a\}
+%\]
+%
+%Space for $T$:
+%\[
+%\cup_{S_1 \in \mathcal{S}} \{(S_1, \mathbf{x_1}) : \mathbf{x_1} \in \times_{a \in S_1}\}
+%\]
+%
+%\begin{align}
+%\pi(
+%\cup_{S \in \mathcal{S} : A_1 \cap S = S_1} \{(S_1 \cup (\mathcal{A}_2 \cap S), \mathbf{x_1}, \mathbf{x_2}) : \mathbf{x_1} \in B_{S_1}, \mathbf{x_2} \in \times_{a \in S_2} X_a\}
+%) &= p(B_T)\\
+%\sum_{S \in \mathcal{S} : A_1 \cap S = S_1} \int_{B_{S_1} \times \otimes_{a \in S_2} X_a} \pi(S_1, S_2, \mathbf{x}_1, \mathbf{x}_2) \mu_{S_1 \cup S_2}(d \mathbf{x}_1 \mathbf{x}_2) &= p(B_T)\\
+%\sum_{S \in \mathcal{S} : A_1 \cap S = S_1} \int_{B_{S_1} \times \otimes_{a \in S_2} X_a} p(S_1, \mathbf{x}_1) q_{(S_1,\mathbf{x}_1)}(S_2, \mathbf{x}_2) \mu_{S_1 \cup S_2}(d \mathbf{x}_1 \mathbf{x}_2) &= p(B_T)\\
+%\sum_{S \in \mathcal{S} : A_1 \cap S = S_1} \int_{B_{S_1}} p(S_1, \mathbf{x}_1) \left( \int_{\otimes_{a \in S_2 X_a}} q_{(S_1,\mathbf{x}_1)}(S_2, \mathbf{x}_2) \mu_{S_2}(d \mathbf{x}_2) \right) \mu_{S_1}(d \mathbf{x}_1) &= p(B_T) [\mbox{Fubini / Tonnelli?}]\\
+%\sum_{S_2 \subseteq A_2 : S_1 \cup S_2 \in \mathcal{S}} \int_{B_{S_1}} p(S_1, \mathbf{x}_1) \left( \int_{\otimes_{a \in S_2 X_a}} q_{(S_1,\mathbf{x}_1)}(S_2, \mathbf{x}_2) \mu_{S_2}(d \mathbf{x}_2) \right) \mu_{S_1}(d \mathbf{x}_1) &= p(B_T)\\
+%\int_{B_{S_1}} p(S_1, \mathbf{x}_1) \left( \sum_{S_2 \subseteq A_2 : S_1 \cup S_2 \in \mathcal{S}} \int_{\otimes_{a \in S_2 X_a}} q_{(S_1,\mathbf{x}_1)}(S_2, \mathbf{x}_2) \mu_{S_2}(d \mathbf{x}_2) \right) \mu_{S_1}(d \mathbf{x}_1) &= p(B_T) [\mbox{Fubini / Tonnelli?}]\\
+%\end{align}
+%It suffices to show that
+%\begin{align}
+%\sum_{S_2 \subseteq A_2 : S_1 \cup S_2 \in \mathcal{S}} \int_{\otimes_{a \in S_2 X_a}} q_{(S_1,\mathbf{x}_1)}(S_2, \mathbf{x}_2) \mu_{S_2}(d \mathbf{x}_2) = 1 \mbox{ for all } (S_1, \mathbf{x}_1) \in T
+%\end{align}
+%It suffices to show that for each $(S_1, \mathbf{x}_1) \in T$, we have that $q_{(S_1, \mathbf{x}_1)}$ is the density of a (terminating with probability one) probabilistic program with structruers $\mathcal{S}_2$ such that
+%\[
+%\mathcal{S}_2 = \{S_2 \subseteq A_2 : S_1 \cup S_2 \in \mathcal{S}\}
+%\]
+%
+%\[
+%B_1 = \cup_{S_1 \in \mathcal{S}_1} \{(S_1, \mathbf{x}_1) : \mathbf{x}_1 \in B_{S_1}\}\;\;\; \mbox{where} \;\; B_{S_1} \in \otimes_{a \in S_1} \Sigma_{a} \;\;\; [\mbox{Measurable set } B_1]
+%\]
+%% TODO the proof above should have a sum over S_1
 

### 2020-07-11 paper.tex
checkpoint on probabilisic programming with traces

@@
 
-\section{INVOLUTIVE MCMC}
-
+\section{INVOLUTIVE MCMC} \label{sec:involutive-mcmc}
 Let $(X, \Sigma_P, \mu_P)$ and $(Y, \Sigma_Q, \mu_Q)$ denote two measure spaces with $\sigma$-finite $\mu_P$ and $\mu_Q$.
-Let $p : X \to [0, \infty)$ be a probability density with respect to $\mu_P$ and for each $t \in X$ such that $p(t) > 0$ let $q_t : Y \to [0, \infty)$ be probability density with respect to $\mu_Q$.
-We call $p$ the \emph{target density} and we call $q_t$ \emph{auxiliary densities}.
-Let $\pi(t, u) := p(t) q_t(u)$ and $Z := \{(t, u) : \pi(t, u) > 0 \}$.
-Suppose $Z \in \Sigma_P \otimes \Sigma_Q$ (the product $\sigma$-algebra), so that $\Sigma := \{A \in \Sigma_P \otimes \Sigma_Q : A \subseteq Z\}$ is a $\sigma$-algebra on $Z$ and let $\mu : \Sigma \to [0, \infty)$ with $\mu(A) := (\mu_P \times \mu_Q)(A)$
-(note that $\mu$ is $\sigma$-finite since $\mu_P \times \mu_Q$ is $\sigma$-finite. % TODO: prove this 
-Suppose that $\pi$ is a probability density with respect to $\mu$. % TODO is this guaranteed?
+Let $p : X \to [0, \infty)$ be a probability density with respect to $\mu_P$ and for each $x \in X$ such that $p(x) > 0$ let $q_x : Y \to [0, \infty)$ be probability density with respect to $\mu_Q$.
+We call $p$ the \emph{model density} and we call $q_x$ \emph{auxiliary densities}.
+Let $\pi(x, y) := p(x) q_x(y)$ and $Z := \{(x, y) \in X \times Y : \pi(x, y) > 0 \}$.
+Suppose that $Z \in \Sigma_P \otimes \Sigma_Q$ ($Z$ is a $\mu_P \times \mu_Q$-measurable set).
+Then $\Sigma := \{A \in \Sigma_P \otimes \Sigma_Q : A \subseteq Z\}$ is a $\sigma$-algebra on $Z$ and $\mu : \Sigma \to [0, \infty)$ with $\mu(A) := (\mu_P \times \mu_Q)(A)$ is a $\sigma$-finite measure on measurable space $(Z, \Sigma)$.
 
 Let $f : Z \to Z$ denote an involution ($f^{-1} = f$) such that the pushforward of $\mu$ under $f$, denoted $\mu \circ f^{-1}$, is absolutely continuous with respect to $\mu$, with Radon-Nikodym derivative $d (\mu \circ f^{-1}) / d\mu : Z \to [0, \infty)$.
-Let $u \sim q_t(\cdot)$ denote a random sample from the measure defined by $A \mapsto \int_A q_t(u) \mu_Q(du)$ for $A \in \Sigma_Q$.
+Let $y \sim q_x(\cdot)$ denote a random sample from the measure defined by $A \mapsto \int_A q_x(y) \mu_Q(dy)$ for $A \in \Sigma_Q$.
 
@@
 \begin{algorithmic}
-\Procedure{involutive-mcmc-move}{$p$, $q$, $f$, $t$}
-    \State $u \sim q_t(\cdot)$ \Comment{Sample auxiliary variables}
-    \State $(x', u') \gets f(x, u)$ \Comment{Apply involution}
-    %\State $\alpha \gets
-        %\frac{p(x') q(u'; x')}{p(x) q(u; x)} \left| \left[ \frac{\partial h(u, t)}{\partial (u, t)} \right] \right|$
-    % TODO or should we show the Radon-Nikodym derivative?
+\Procedure{involutive-mcmc-move}{$p$, $q$, $f$, $x$}
+    \State $y \sim q_x(\cdot)$ \Comment{Sample auxiliary variables}
+    \State $(x', y') \gets f(x, y)$ \Comment{Apply involution}
     \State $\alpha \gets
-        \displaystyle \frac{p(x') q_{x'}(u')}{p(x) q_{x}(u)} \left( \frac{d (\mu \circ f^{-1})}{d \mu} (x, u)\right)$
+        \displaystyle \frac{p(x') q_{x'}(y')}{p(x) q_{x}(y)} \left( \frac{d (\mu \circ f^{-1})}{d \mu} (x, y)\right)$
     % TODO check it
@@
 
-\paragraph{Traces}
+\subsection{Traces}
 In our framework, a probabilistic program defines a probability distribution on finite dictionaries that map the name of a `random choice' to its value.
@@
 
-TODO: introduce the mathematical notation for a trace
+TODO: introduce the mathematical notation for a concrete trace (i.e. $\{a \mapsto 1, b \mapsto 2\}$).
 
-\paragraph{A probabilistic program encodes a probability distribution on traces}
+\subsection{A probabilistic program encodes a probability distribution on traces}
 For probabilistic programs that only sample individual random choices from discrete probability distributions, the meaning of the program can be described by a probability mass function that assigns a probailtiy to every possible trace.
@@
 % measure version
-\paragraph{Handling continuous distributions}
+\subsection{Handling Continuous Distributions}
 For programs that can sample from continuous probability distributions, the formalism needs to be extended.
-pecifically, we define the $\mathcal{S}$ to be the set of all possible \emph{trace structures} as a (possibly countably infinite) set of finite sets of addresses, denoted $\mathcal{S} \subseteq \mathcal{P}(\mathcal{A})$, where $\mathcal{A}$ is the set of all addresses.
-For each address $a \in \mathcal{A}$ we assign a reference measure space $(X_a, \Sigma_a, \mu_a)$ where $\mu_a$ is $\sigma$-finite.
-The set of all traces of a probabilistic program is the set $T := \cup_{S \in \mathcal{S}} \{(S, \mathbf{x}) : \mathbf{x} \in \times_{a \in S} X_a\}\}$.
-The \emph{reference measure space} for a probabilistic program is $(T, \Sigma, \mu)$ where
-$\Sigma$ is the $\sigma$-algebra that contains sets $\cup_{S \in \mathcal{S}}\{ B_S \} \in \mathcal{P}(T)$ such that $B_S$ is an element of the product $\sigma$-algebra $\Sigma_S := \otimes_{a \in S} \Sigma_a$,
-and where $\mu(\cup_{S \in \mathcal{S}}\{ B_S \}) := \sum_{S \in \mathcal{S}} \mu_S(B_S)$ where $\mu_S$ is the product measure $\times_{a \in S} \mu_a$.
+Let $\mathcal{A}$ denote a set of addresses.
+For each address $a \in \mathcal{A}$ we assign a reference measure space $(V_a, \Sigma_a, \mu_a)$ where $\mu_a$ is $\sigma$-finite ($V$ stands for `values').
+Let $T := \cup_{A \subseteq \mathcal{A} : |A| < \infty} \{(A, \mathbf{x}) : \mathbf{x} \in \times_{a \in S} X_a\}\}$ denote the set of all finite traces (i.e.
+dictionaries) from addresses in $\mathcal{A}$ to their values.
+For each finite $A \subseteq \mathcal{A}$, let $\Sigma_A := \otimes_{a \in A} \Sigma_a$ (the product $\sigma$-algebra over $\Sigma_a$ for $a \in A$), and let $\mu_A := \times_{a \in A} \mu_a$ (the product measure).
+Let $\Sigma \subseteq \mathcal{P}(T)$ be the $\sigma$-algebra formed by sets of the form $B := \cup_{A \subseteq \mathcal{A}, |A| < \infty} \{(A, \mathbf{x}) : \mathbf{x} \in B_A\}$ for some $B_A \in \Sigma_A$ for each $A$.
+Let $\mu : \Sigma \to [0, \infty)$ be a measure on measurable space $(T, \Sigma)$ defined by $\mu(B) := \sum_{A \subseteq \mathcal{A}, |A| < \infty} \mu_A(B_A)$.
 
-We assume that a probabilistic program $P$ denotes a reference measure space $(T_P, \Sigma_P, \mu_P)$ and a measure $P(dt)$ that is mutually absolutely continuous with respect to $\mu_P$, with density
-$p : T \to [0, \infty)$ on traces $t \in T$.
-We do not give the denotational semantics of a probabilistic programming language with this property, but instead refer to POPL paper for an example of a denotational semantics.
-% TODO what about requirement that the support of each random choice is constant?
-(this is implied by the definition of reference measure)
+A probabilistic program $P$ denotes a probability density $p : T \to [0, \infty)$ with respect to a reference measure $\mu$ on traces.
+Describing the denotational semantics of a particular probabilistic programming language (the function mapping programs $P$ to density functions $p$) is out of scope for this paper.
+We refer to the reader to~\citet{popl} for an example of denotational semantics of a probabilistic programming language that uses explicit addresses for random-choices.
 
-TODO: Give an example of a probabilistic program in Gen's dynamic modeling language that includes continuous random choices, and the vbase measure and a density for it.
+This formalism is capable of expressing probabilistic programs that use continuous distributions by setting the reference measure space $(V_a, \Sigma_a, \mu_a)$ for some addresses $a$ to use the Lebesgue-measure.
+For example, consider the following probabilistic program:
+
+TODO: Give an example of a probabilistic program in Gen's dynamic modeling language that includes continuous random choices, and show the reference measure on traces and its density.
 
@@
 
-\[
-p(t) \;\; \pi(z)
-\]
+TODO: somewhere, talk about hierarchical addresses
+%\paragraph{Hierarchical addresses}
 
-\paragraph{Extending one probabilistic program with another}
-A second probabilistic program that is parametrized by (i.e. takes as input) traces of the first probabilistic program $P$.
-The reference measure space of the second probabilistic program for $t \in T_P$ is denoted $(T_{Q,t}, \Sigma_{Q,t}, \mu_{P,t})$.
-We define $Z := \{(t, u) : t \in T_P, u \in T_{Q,t}$ and we define $\pi(z) := p(t) q_t(u)$.
-We require that $(Z, \Sigma, \mu)$ is a probabilistic program reference measure and $\pi$ is a probability density with respect to $\mu$ that is mutually absolutely continuous.
-The combined program is $P \circ Q$.
+\subsection{Involutive MCMC with probabilistic programs}
 
-Given a
+To define model the auxiliary densitites used in involutive MCMC in Section~\ref{sec:involutive-mcmc} using probabilistic programs, we write one probabilistic program ($P$) to define the model's probability distribution, and another probabilistic program ($Q$) to define the set of auxiliary distributions.
+The model reference measure $(X, \Sigma_P, \mu_P)$ and density required in Section~\ref{sec:involutive-mcmc} are set to the probabilistic program's reference measure on traces as defined above and density $p$ (so that $X := T$).
+To define the auxiliary densities, we require a probabilistic program $Q$ that is parametrized by (i.e. takes as input) traces of the model program $P$.
+The set of possible inputs to $Q$ is $\{x \in T : p(x) > 0\}$, and for each such $x$ the probabilistic program $Q$ defines a density $q_x : T \to [0, \infty)$ with respect to the reference measure on traces.\footnote{
+Note that we the same measure space $(T, \Sigma, \mu)$ on traces for both the model and auxiliary probabilistic program.
+If both programs sample some random choice at the same address $a$ and require different measure spaces $(V_a, \Sigma_a, \mu_a)$ for the address, this can be handled formally by making the addresses sampled by the two programs disjoint by prepending either the prefix `p.' or `q.' to addresses.
+This is only required for book-keeping in the formalism, and these prefixes do not need to be actually added to the addresses used in the programs.}
 
-% the only place where this matters is in the 'gibbs move' argument.. (8.4)
-
-We need $p$ to be a density with respect to some reference measure space $(T_P, \Sigma_{P}, \mu_P)$.
-We need $q_t$ to be a density with respect to some reference measure $\mu_{Q,t}$ for each $t \in T_P$.
-Such that the function $\pi(t, u) := p(x) q_t(u)$ is a density $\pi(x, u)$ with respect to some reference measure $\mu$, and $\pi$ is mutually absolutely continuous with $\mu$.
-(this can be constructed by using hierarchical addresses `model' and `proposal' for the random choics made in the two programs).
+Full support requirement (that for all structures and assignments to discrete random variables within the support, the other addresses have complete support).
+Also for a pair of programs.
+Use it guarantee that $X := \{(t, u) : \pi(t, u) > 0\} \in \Sigma_P \otimes \Sigma_Q$?
 
@@
 \subsection{Proof of detailed balance for involution} \label{sec:involution-detailed-balance}
+
+\begin{lemma}
+Given $\sigma$-finite measures $(X_P, \Sigma_P, \mu_P)$ and $(X_Q, \Sigma_Q, \mu_Q)$,
+if $X \in \Sigma_P \otimes \Sigma_Q$ then $\Sigma = \{A \in \Sigma_P \otimes \Sigma_Q : A \subseteq X\}$ is a $\sigma$-algebra on $X$, and $\mu : \Sigma \to [0, \infty)$ given by $\mu(A) := (\mu_P \times \mu_Q)(A)$ is $\sigma$-finite measure on measurable space $(X, \Sigma)$.
+\end{lemma}
+\begin{proof}
+TODO: write out the detailed steps, proving each property of a $\sigma$-algebra%$ TODO
+\end{proof}
+
+\begin{lemma}
+$\pi : X \to [0, \infty)$ is a probability density with respect to $\mu$ and the measure induced by $\pi$ and the measure $\mu$ are mutually absolutely continuous 
+\end{lemma}
+\begin{proof}
+% TODO need to show that pi is measurable by the product measure?
+\begin{equation}
+\int_X \pi(x) \mu(dx) = \int_{X_P \times X_Q} \pi(x) (\mu_P \times \mu_Q)(dx) = \int_{X_P} p(t) \left( \int_{X_Q} q_t(u) \mu_Q(du) \right) \mu_P(dt) = 1
+\end{equation}
+
+%(note that $\mu$ is $\sigma$-finite since $\mu_P \times \mu_Q$ is $\sigma$-finite. % TODO: prove this 
+%Suppose that $\pi$ is a probability density with respect to $\mu$. % TODO is this guaranteed?
+\end{proof}
+
 \citet{tierney1998note} gives a class of MCMC kernels based on involutions that satisfy detailed balance.

### 2020-07-11 paper.tex
add lemma for support being measurable set

@@
 For each address $a \in \mathcal{A}$ we assign a reference measure space $(V_a, \Sigma_a, \mu_a)$ where $\mu_a$ is $\sigma$-finite ($V$ stands for `values').
-Let $T := \cup_{A \subseteq \mathcal{A} : |A| < \infty} \{(A, \mathbf{x}) : \mathbf{x} \in \times_{a \in S} X_a\}\}$ denote the set of all finite traces (i.e.
+Let $T := \cup_{A \subseteq \mathcal{A} : |A| < \infty} \{(A, \mathbf{x}) : \mathbf{x} \in \times_{a \in A} V_a\}\}$ denote the set of all finite traces (i.e.
 dictionaries) from addresses in $\mathcal{A}$ to their values.
@@
 To define model the auxiliary densitites used in involutive MCMC in Section~\ref{sec:involutive-mcmc} using probabilistic programs, we write one probabilistic program ($P$) to define the model's probability distribution, and another probabilistic program ($Q$) to define the set of auxiliary distributions.
-The model reference measure $(X, \Sigma_P, \mu_P)$ and density required in Section~\ref{sec:involutive-mcmc} are set to the probabilistic program's reference measure on traces as defined above and density $p$ (so that $X := T$).
-To define the auxiliary densities, we require a probabilistic program $Q$ that is parametrized by (i.e. takes as input) traces of the model program $P$.
+The model reference measure $(X, \Sigma_P, \mu_P)$ and density that were required in Section~\ref{sec:involutive-mcmc} are set to the probabilistic program's reference measure and density $p$ on traces as defined above (so that $X := T$).
+To define the auxiliary densities, we use a second probabilistic program $Q$ that is parametrized by (i.e. takes as input) traces of the model program $P$.
 The set of possible inputs to $Q$ is $\{x \in T : p(x) > 0\}$, and for each such $x$ the probabilistic program $Q$ defines a density $q_x : T \to [0, \infty)$ with respect to the reference measure on traces.\footnote{
@@
 
-Full support requirement (that for all structures and assignments to discrete random variables within the support, the other addresses have complete support).
-Also for a pair of programs.
-Use it guarantee that $X := \{(t, u) : \pi(t, u) > 0\} \in \Sigma_P \otimes \Sigma_Q$?
+The involutive MCMC framework of Section~\ref{sec:involutive-mcmc} requires that $Z := \{(x, y) \in X \times Y : \pi(x, y) > 0\}$ is $\mu_P \times \mu_Q$-measurable.
+We now give a sufficient condition for this to hold.
+Let $D \subseteq \mathcal{A}$ denote the subset of addresses that are discrete (i.e. where $V_a$ is a countable set and $\mu_a$ is the counting measure).
+For $\mathbf{x} \in \times_{a \in A} V_a$ let $\mathbf{x} = (\mathbf{d}, \mathbf{c})$ where $\mathbf{d} \in \times_{a \in A \cap D} V_a$ and $\mathbf{c} = \times_{a \in A \setminus D} V_a$, so that $\mathbf{d}$ is the discrete part of $\mathbf{x}$ and $\mathbf{c}$ is the non-discrete part.
+ %TODO make this a proposition
+\begin{lemma}
+Suppose that $p$ and $q$ are such that $p(A, (\mathbf{d}, \mathbf{c})) > 0$ implies
+$p(A, (\mathbf{d}, \mathbf{c}')) > 0$ for all $\mathbf{c}' \in \prod_{a \in A \setminus D} V_a$,
+and that $q_x(A, (\mathbf{d}, \mathbf{c})) > 0$ where $x = (\tilde{A}, \tilde{\mathbf{d}}, \tilde{\mathbf{c}})$ implies that
+$q_{x'}(A, (\mathbf{d}, \mathbf{c}')) > 0$ for all $\mathbf{c}' \in \prod_{a \in A \setminus D} V_a$ and all $x' = (\tilde{A}, \tilde{\mathbf{d}}, \tilde{\mathbf{c}}')$ where $\tilde{\mathbf{c}}' \in \prod_{a \in \tilde{A} \setminus D} V_a$.
+Then, $Z := \{(x, y) \in X \times Y : \pi(x, y) > 0\}$ is $\mu \times \mu$-measurable where $\mu$ is the reference measure on traces.
+\end{lemma}
+\begin{proof}
+For $p$ and $q$ satisfying these conditions,
+$Z = \cup_{(A_1,A_2,\mathbf{d}_1,\mathbf{d}_2) \in E} \{((A_1, (\mathbf{d}_1, \mathbf{c}_1)), (A_2, (\mathbf{d}_2, \mathbf{c}_2))) : \mathbf{c}_1 \in \times_{a \in A_1 \setminus D} V_a, \mathbf{c}_2 \in \times_{a \in A_2 \setminus D} V_a\}$ for some countable set $E \subseteq \{(A_1, A_2, \mathbf{d}_1, \mathbf{d}_2) : A_1, A_2 \subseteq \mathcal{A}, |A_1| < \infty, |A_2| < \infty, \mathbf{d}_1 \in \times_{a \in A_1 \cap D} V_a, \mathbf{d}_2 \in \times_{a \in A_2 \cap D} V_a\}$ of address sets and discrete choice values for both programs.
+The measure of $Z$ is
+$(\mu \times \mu)(Z) = \sum_{(A_1,A_2,\mathbf{d}_1,\mathbf{d}_2) \in E} \prod_{a \in (A_1 \setminus D) \cup (A_2 \setminus D)} \mu_a(V_a)$.
+% TODO double check me.,
+\end{proof}
 
-\paragraph{Conditioning}
+Intuitively, this requirement means that for both the model probabilistic program $P$ and the auxiliary probabilistic program $Q$, the support of a random choice that is not discrete cannot depend on the value of another non-discrete random choice.
+Additionally, the support of non-discrete random choices in $Q$ cannot depend on the value of non-discrete random choices in the input $x$, which is a trace of $P$.
+
+%\paragraph{Conditioning}
 

### 2020-07-12 paper.tex
add text on traces

@@
 \usepackage{graphicx}
+\usepackage[usenames,dvipsnames,svgnames]{xcolor}
 
@@
 \usepackage{amsthm}
+\usepackage{bm}
 \usepackage{algorithm}
@@
 \usepackage{algpseudocode}
+
+% use straight double quotes in math mode
+\DeclareMathSymbol{\mathdblquotechar}{\mathalpha}{letters}{`"}
+\newcommand{\mathdblquote}{\mathtt{\mathdblquotechar}}
+\begingroup\lccode`~=`"\lowercase{\endgroup
+  \let~\mathdblquote
+}
+\AtBeginDocument{\mathcode`"="8000 }
+
+% code listings
+\usepackage{listings}
+\lstset{
+    literate={~} {$\color{black} \bm{\sim}$}{1}
+}
+\usepackage[framemethod=tikz]{mdframed}
+\input{genlisting}
+%\definecolor{codegray}{gray}{0.96}Z
+\definecolor{addresscolor}{RGB}{22, 88, 12}
+\definecolor{distcolor}{RGB}{40, 12, 88}
+\definecolor{gencode}{RGB}{255,235,150}
+%^\definecolor{jlcode}{RGB}{255, 247, 247}
+\definecolor{jlcode}{RGB}{255,210,210}
+%\definecolor{dslcode}{RGB}{247, 255, 230}
+\definecolor{dslcode}{RGB}{220,242,255}
+\usepackage{colortbl}
+
 %\newcommand{\disc}{\mathrm{disc}}
@@
 \newtheorem{lemma}[theorem]{Lemma}
+\newcommand{\true}{{\mathtt{T}}}
+\newcommand{\false}{{\mathtt{F}}}
 
@@
 \end{algorithm}
+
 See Section~\ref{sec:radon-nikodym-special-case} for a derivation of the Radon-Nikodym derivative $d (\mu \circ f^{-1}) / d \mu$ for a special case, when the involution $f$ can be factored into (i) an involution on a countable partition of $Z$ and (ii) a family of continuously differentiable bijection on Euclidean spaces.
@@
 \begin{theorem}[Involutive MCMC is stationary]
-For $p$, $q$, and $f$ satisfying the requirements above, Algorithm~\ref{alg:involutive-mcmc} defines a probability kernel on $X$ that is stationary with respect to the distribution $p$.
+Algorithm~\ref{alg:involutive-mcmc} defines a probability kernel $k$ on $X$ that is stationary with respect to the model probability distribution.
+That is, $\int_X k_x(B) p(x) d\mu_X(dx) = \int_B p(x) d\mu_X(dx)$ for all $B \in \Sigma_X$.
 \end{theorem}
@@
 
-\subsection{Traces}
-In our framework, a probabilistic program defines a probability distribution on finite dictionaries that map the name of a `random choice' to its value.
-Following standard terminology in probabilistic programming~\citep{?}, we call these dictionaries \emph{traces} because they can be sampled by running a probabilistic program, and `tracing' the execution the program by recording the value of eacn random choice encountered.
-In the probabilistic programming language used in this paper, the user assigns a unique name or \emph{address} to each random choice using a special syntax:
+In our framework, a probabilistic program defines a probability distribution on finite dictionaries that map the identifer for a random choice (called its \emph{address}) to its value.
+Following standard terminology~\citep{?}, we call these dictionaries \emph{traces} because they are often sampled by running a probabilistic program, and `tracing' the execution the program by recording the value of eacn random choice encountered.
+The next section makes the concept of a trace more concrete using code in a probabilistic programming language.
 
-TODO: include example of tilde syntax
+\subsection{A Probabilistic Programming Language}
 
-TODO: introduce the mathematical notation for a concrete trace (i.e. $\{a \mapsto 1, b \mapsto 2\}$).
+The probabilistic programming language used in this paper is based on the dynamic modeling language in the Gen probabilistic programming system~\citep{gen-pldi}.
+In this language, user tags each random choice with an explicit address in the code, as part of a \emph{random-choice expression} of the following form:
+\noindent
+\begin{center}
+\begin{tabular}{c}
+{\begin{lstlisting}[basicstyle=\small\ttfamily]
+{<address>} ~ <distribution>
+\end{lstlisting}}
+\end{tabular}
+\end{center}
+The left-hand side of the expression (green) specifies the address of the choice, and the right-hand side specifies the probability distribution to be sampled from.
+For example, the following expression samples a non-negative integer from a geometric distribution and the value gets recorded in the trace at the address \texttt{"a"}:
+\noindent
+\begin{center}
+\begin{tabular}{c}
+{\begin{lstlisting}[basicstyle=\small\ttfamily]
+{"a"} ~ geometric(0.5)
+\end{lstlisting}}
+\end{tabular}
+\end{center}
+In addition to being recorded in the trace, the sampled value is returned to the running program.
+That is, the random choice expression evaluates to the sampled value.
+For example, the expression below evaluates to a an even non-negative integer:
+\noindent
+\begin{center}
+\begin{tabular}{c}
+{\begin{lstlisting}[basicstyle=\small\ttfamily]
+({"a"} ~ geometric(0.5)) * 2
+\end{lstlisting}}
+\end{tabular}
+\end{center}
+Probabilistic programs are typically constructed by combining multiple random choice expressions with regular Julia code (Gen is embedded in the Julia programming language~\citep{julia}).
+\noindent
+\begin{center}
+\begin{tabular}{c}
+{\begin{lstlisting}[basicstyle=\small\ttfamily]
+a = ({"a"} ~ geometric(0.5))
+{"b"} ~ bernoulli(a / (a + 1))
+\end{lstlisting}}
+\end{tabular}
+\end{center}
+This probabilistic program has an unbounded (countably infinite) number of possible traces it could produce.
+One possible trace samples the value 1 at $\texttt{"a"}$ and the Boolean value $\false$ at \texttt{"b"}, and we denote this trace by $\{\mathtt{"a"} \mapsto 1, \mathtt{"b"} \mapsto \false\}$.
+Note that the variable to which the first choice was assigned happens to be the same as the address.
+This is not required, and often the random choice has no variable name in the program, as in the second random choice with address \texttt{"b"}.
+But to reduce possibe unecessary typing, the following optional syntactic sugar will simultaneously assign the value to the program variable $\mathtt{a}$ and record it at trace address $\mathtt{a}$.\footnote{
+More precisely, the address is the Julia symbol (interned string) \texttt{a}, constructed with Julia syntax \texttt{:a}.}
+\noindent
+\begin{center}
+\begin{tabular}{c}
+{\begin{lstlisting}[basicstyle=\small\ttfamily]
+a ~ geometric(0.5)
+\end{lstlisting}}
+\end{tabular}
+\end{center}
+Probabilistic programs can make use of familiar control flow constructs, like loops and branching.
+Stochastic control flow is also possible:
+\noindent
+\begin{center}
+\begin{tabular}{c}
+{\begin{lstlisting}[basicstyle=\small\ttfamily]
+i = 1
+while ({i} ~ bernoulli(0.5))
+    i = i + 1
+end
+\end{lstlisting}}
+\end{tabular}
+\end{center}
+Note that for the probabilistic program to be valid, we require that every address is used at most once within every possible execution.
 
-\subsection{A probabilistic program encodes a probability distribution on traces}
-For probabilistic programs that only sample individual random choices from discrete probability distributions, the meaning of the program can be described by a probability mass function that assigns a probailtiy to every possible trace.
+For probabilistic programs that sample from discrete probability distributions like those used above, the meaning of the program is a probability mass function $p$ that assigns a probability to every possible trace.
+For example, for the program with the \texttt{while} loop above, the distribution is:
+\[
+\begin{array}{c|c}
+    x & p(x)\\
+    \hline
+    \{1 \mapsto \false\} & 0.5\\
+    \{1 \mapsto \true, 2 \mapsto \false\} & 0.5^2\\
+    \{1 \mapsto \true, 2 \mapsto \true, 3 \mapsto \false\} & 0.5^3\\
+    .. & ..
+\end{array}
+\]
+We assume that probabilistic programs halt with probability $1$, meaning that the sum of probabilities over all (finite) traces is one.
+Note that traces are always finite, but there may be an infinite number of possible traces, as in the previous example.
 
-TODO: Give a simple example of a probabilistic program with discrete random chocies and stochastic control flow, and the resulting probabiilty distribution.
-TODO: describe the two variants of tilde syntax
-
-Note that traces are always finite, but there can be no upper bound.
-For example, for this program, there is a countably infinite number of possible traces:
-
-TODO: show a discrete probabilistic prgram with the table of traces it could sample
+In order to construct more complex distributions, we wrap probabilistic programs in functions, called
+`generative functions' in Gen.
+These functions can then be invoked by other functions, including recursively, which allows programs encode probability distributions on more complex and combinatorial objects.
+When invoking such a function, we give an \emph{address namespace} under which every random choice made in that function should be recorded.
+This helps prevent addresses of choices defined in different functions from colliding.
+The syntax for calling a generative function and assigning the address namespace is the same as that of random choice expressions.
+For example:
+\begin{center}
+\begin{tabular}{c}
+\begin{lstlisting}[basicstyle=\small\ttfamily]
+@gen function g()
+  if ({"go"} ~ bernoulli(0.2))
+    n1 = ({"L"} ~ g())
+    n2 = ({"R"} ~ g())
+    return n1 + n2 + 1
+  else
+    return 1
+  end
+end
+\end{lstlisting}
+\end{tabular}
+\end{center}
+Because this is a recursive function, a the trace is populated with a hierarchy of addresses.
+For example, the first random choice in the first recursive call has address \texttt{"L"."go"}.
+When this function is invoked it encodes the following probability distribution on traces (quotation marks omitted from addresses to improve readability):
+\[
+\begin{array}{c|c}
+    x & p(x)\\
+    \hline
+    \left\{\mathtt{go} \mapsto \false\right\} & 0.8\\
+& \\
+    \left\{\begin{array}{c}
+        \mathtt{go} \mapsto \true, \mathtt{L}.\mathtt{go} \mapsto \false, \mathtt{R}.\mathtt{go} \mapsto \false
+    \end{array}\right\} & 0.2 \cdot 0.8^2\\
+& \\
+    \left\{\begin{array}{c}
+        \mathtt{go} \mapsto \true,
+        \mathtt{L}.\mathtt{go} \mapsto \true,
+        \mathtt{R}.\mathtt{go} \mapsto \false,\\
+        \mathtt{L}.\mathtt{L}.\mathtt{go} \mapsto \false,
+        \mathtt{L}.\mathtt{R}.\mathtt{go} \mapsto \false
+    \end{array}\right\}& 0.2^2 \cdot 0.8^3\\
+     .. & ..
+\end{array}
+\]
 
@@
 TODO: Give an example of a probabilistic program in Gen's dynamic modeling language that includes continuous random choices, and show the reference measure on traces and its density.
-
-% step 1: *assert* that the composition of p and then q defines a probabilistic program with ref meas mu and mut. abs. cont. pi(z) [ok] (( do this without ascribing separate semantics to p and q?? ))
-% step 2: show detailed balance on the joint space ... [ok]
-% step 3: show stationarity on the joint space [ok, easy from detailed balance]
-% step 4: show stationarity with respect to space of p.... ((requires ascribing semantics to p, but maybe not q??)
-% - this would be straightforward for a product space, but our space is more complicated?
-% - need to assert relationship between p and pi, ...
+\noindent
+\begin{center}
+\begin{tabular}{c}
+{\begin{lstlisting}[basicstyle=\small\ttfamily]
+@gen function foo(x1, x2, x3)
+  a = ({:a} ~ bernoulli(x1))
+  b = ({:b} ~ bernoulli(x2))
+  c = ({:c} ~ bernoulli(x3))
+  {:d} ~ bernoulli((a && b && c) ? 0.9 : 0.1)
+  return a && b && c && d
+end
+\end{lstlisting}}
+\end{tabular}
+\end{center}
 
@@
 
-To define model the auxiliary densitites used in involutive MCMC in Section~\ref{sec:involutive-mcmc} using probabilistic programs, we write one probabilistic program ($P$) to define the model's probability distribution, and another probabilistic program ($Q$) to define the set of auxiliary distributions.
+To define the model and auxiliary densitites used in involutive MCMC in Section~\ref{sec:involutive-mcmc} using probabilistic programs, we write one probabilistic program ($P$) to define the model's probability distribution, and another probabilistic program ($Q$) to define the set of auxiliary distributions.
 The model reference measure $(X, \Sigma_P, \mu_P)$ and density that were required in Section~\ref{sec:involutive-mcmc} are set to the probabilistic program's reference measure and density $p$ on traces as defined above (so that $X := T$).

### 2020-07-12 paper.tex
add example of a density

@@
 \newcommand{\false}{{\mathtt{F}}}
+\newcommand{\reals}{\mathbb{R}}
+\newcommand{\naturals}{\mathbb{N}}
+\newcommand{\integers}{\mathbb{Z}}
+\newcommand{\normal}{\mathcal{N}}
+\newcommand{\unif}{\mathcal{U}}
 
@@
 
-% measure version
 \subsection{Handling Continuous Distributions}
-For programs that can sample from continuous probability distributions, the formalism needs to be extended.
-Let $\mathcal{A}$ denote a set of addresses.
+For programs that can sample from continuous probability distributions, the formalism needs to be generalized.
+Instead of a probability mass function on traces, a probabilistic program defines a probability density function on traces with respect to a certain reference measure.
+
+Let $\mathcal{A}$ denote a countable set of all possible addresses.
 For each address $a \in \mathcal{A}$ we assign a reference measure space $(V_a, \Sigma_a, \mu_a)$ where $\mu_a$ is $\sigma$-finite ($V$ stands for `values').
@@
 A probabilistic program $P$ denotes a probability density $p : T \to [0, \infty)$ with respect to a reference measure $\mu$ on traces.
-Describing the denotational semantics of a particular probabilistic programming language (the function mapping programs $P$ to density functions $p$) is out of scope for this paper.
-We refer to the reader to~\citet{popl} for an example of denotational semantics of a probabilistic programming language that uses explicit addresses for random-choices.
+Describing the denotational semantics of a particular probabilistic programming language (the function mapping programs $P$ to density functions $p$) is out of scope for this paper, and we refer to the reader to~\citet{popl} for an example of denotational semantics of a probabilistic programming language that uses explicit addresses for random-choices.
 
-This formalism is capable of expressing probabilistic programs that use continuous distributions by setting the reference measure space $(V_a, \Sigma_a, \mu_a)$ for some addresses $a$ to use the Lebesgue-measure.
-For example, consider the following probabilistic program:
-
-TODO: Give an example of a probabilistic program in Gen's dynamic modeling language that includes continuous random choices, and show the reference measure on traces and its density.
+The measure-theoretic formalism above is capable of expressing probabilistic programs that use continuous distributions by setting the reference measure space $(V_a, \Sigma_a, \mu_a)$ for some addresses $a$ to use the Lebesgue-measure.
+For example, consider the following probabilistic program, which encodes a generative model for data generated from a polynomial with unknown degree and unknown coefficients:
 \noindent
@@
 {\begin{lstlisting}[basicstyle=\small\ttfamily]
-@gen function foo(x1, x2, x3)
-  a = ({:a} ~ bernoulli(x1))
-  b = ({:b} ~ bernoulli(x2))
-  c = ({:c} ~ bernoulli(x3))
-  {:d} ~ bernoulli((a && b && c) ? 0.9 : 0.1)
-  return a && b && c && d
+@gen function polynomials(x_coords)
+    degree ~ |geometric(0.5)|
+    noise ~ |inv_gamma(1, 1)|
+    coeffs = [ ({("c", i)} ~ |normal(0, 2)|)
+               for i in 0:degree ]
+    for i=1:length(x_coords)
+        x = x_coords[i]
+        mu = coeffs' * x.^(0:length(coeffs)-1)
+        {("y", i)} ~ |normal(mu, noise + 0.1)|
+    end
 end
@@
 \end{center}
-
-TODO: somewhere, talk about hierarchical addresses
-%\paragraph{Hierarchical addresses}
+The addresses used by this generative function are:
+\[
+\mathcal{A} := \begin{array}{c}
+\{\mathtt{:}\mathtt{noise}, \mathtt{:}\mathtt{degree}\} \bigcup
+\{(\mathtt{"c"}, i) : i \in \integers_{\ge 0}\} \bigcup\\
+\{(\mathtt{"y"}, i) : i \in \integers_{> 0}\}
+\end{array}
+\]
+(Note that $\mathcal{A}$ could contain additional addresses that the program never samples.)
+The random choice at address $\mathtt{:}\mathtt{degree}$ is discrete with $V_a = \integers_{\ge 0}$ and the other random choices are continuous.
+Address $\mathtt{:}\mathtt{:noise}$ has $V_a = \reals_{> 0}$ and $(\mathtt{"c"}, i)$ and $(\mathtt{"y"}, i)$ have $V_a = \reals$, and all of these use the Lebesgue measure. % TODO double-check this
+The density with respect to the reference measure $\mu$ is:
+\begin{align*}
+p(x) =
+\begin{array}{l}
+p_{\mathrm{geometric}(0.5)}(x[\mathtt{:}\mathtt{degree}]) \cdot\\
+p_{\mathrm{inversegamma}(1, 1)}(x[\mathtt{:}\mathtt{noise}]) \cdot\\
+\prod_{0=1}^{x[\mathtt{:}\mathtt{degree}]} p_{\mathrm{normal}(0,1)}(x[(\mathtt{"c"}, i)])\cdot\\
+\prod_{i=1}^n p_{\mathrm{normal}(m(t, i), 0.1 + x[\mathtt{noise}])}(x[(\mathtt{"y"}, i)])
+\end{array}
+\end{align*}
+where $m(t, i) := \sum_{j=0}^{x[\mathtt{:}\mathtt{degree}]} \mathtt{xcoords}[i]^j \cdot x[(\mathtt{"c"}, j)]$.
 

### 2020-07-12 paper.tex
add content for differentiable programming language

@@
 
-\section{PROBABILISTIC PROGRAMMING WITH TRACES}
+\section{PROBABILISTIC PROGRAMMING WITH TRACES} \label{sec:prob-prog-traces}
 
@@
 \subsection{Handling Continuous Distributions}
-For programs that can sample from continuous probability distributions, the formalism needs to be generalized.
-Instead of a probability mass function on traces, a probabilistic program defines a probability density function on traces with respect to a certain reference measure.
+To handle programs that sample continuous probability distributions the notion of a probability mass function on traces needs to be generalized to a probability density function on traces with respect to a certain reference measure.
+The probabilistic program still defines a probability distibution (i.e. measure) on traces, but viavia  a density instead of a probability mass function.
 
@@
 Describing the denotational semantics of a particular probabilistic programming language (the function mapping programs $P$ to density functions $p$) is out of scope for this paper, and we refer to the reader to~\citet{popl} for an example of denotational semantics of a probabilistic programming language that uses explicit addresses for random-choices.
+But, informally, the density of a trace in our formalism (and existing formalisms) is the product of densities of each individual random choice, as an example below illustrates.
 
@@
 
-\subsection{Involutive MCMC with probabilistic programs}
+\subsection{Specifying the Densities for Involutive MCMC with Probabilistic Programs}
 
@@
 Additionally, the support of non-discrete random choices in $Q$ cannot depend on the value of non-discrete random choices in the input $x$, which is a trace of $P$.
+This requirement defines a notion \emph{well-behavedness} for a probabilistic program ($P$) and an additional notion of well-behavedness for a pair of probabilistic programs that are sequenced one after the other ($P$ and $Q$).
+% TODO: prove that if Q satisfeies the criterion, then the composition P circ Q satisfies the single-program criterion..
 
-%\paragraph{Conditioning}
+\subsubsection{Conditional Distributions on Traces}
+The formalism in this section so far defines the model distribution (with respect to which stationarity of the kernel is required) as the distribution on traces of an (unconditioned) probabilistic program $P$ that encodes a generative model.
+However, typically in applications of MCMC the target distribution is obtained by conditioning on evidence.
+Our formalism does not distinguish between latent and observed random choices, and it is straightforward to show that the involution MCMC kernel is stationary with respect to a conditional distribution on the latent part of a partially observed trace as long as the involution does not modify the existence or values of observed random choices (when such conditional distribution is well-defined).
+Gen provides a mechanism for obtaining initial trace(s) with observed random choices fixed deterministically to the observed values.
+% TODO could add more content from my thesis here to make this more formal.
+% (define marginal density, then define conditional distribution)
+% could also do this in an updated version of the tech report..
 
-\section{DIFFERENTIABLE PROGRAMMING WITH TRACES}
+\section{DIFFERENTIABLE PROGRAMMING WITH TRACES} \label{sec:differentiable-programming-traces}
 
-% TODO PyTorch implementation??
+The previous section described how the probability densities ($p$ and $q$) required for involutive MCMC can be specified using probabilistic programs.
+This section introduces a differentiable programming language that can be used to specify the involution $f$ that is used in involutive MCMC.
+The interpreter for the language can compute the Radon-Nikodym derivative required for the acceptance probability using automatic differentiation, which will be discussed in Section~\ref{sec:automating}.
+The language has been implemented as part of the Gen system~\citep{Gen}.
+This section describes the language and informally introduces its semantics.
 
-\section{AUTOMATIC ACCEPTANCE PROBABILITY CALCULATION}
+In Algorithm~\ref{alg:involutive-mcmc} the involution $f : Z \to Z$ maps a pair $(x, y) \in Z$ to a new pair $(x', y') \in Z$.
+When the densities $p$ and $q$ are defined with probabilistic programs, each element of these pairs is a trace.
+Recall that a trace is mathematically, a finite dictionary that maps addresses to values.
+Therefore, the involution is a map from pairs of traces to pairs of traces.
+%To simplify the notation, we can represent a pair of traces $(x, y)$ as a single trace $z$, in which each address in $x$ and $x'$ is prefixed with some unique identifier (e.g.
+\texttt{model}) and each address in $y$ is prefixed with some unique identifier $y'$ (e.g.
+\texttt{auxiliary}).
+Then, $Z$ is a set of such trace pairs, and the involution $f$ is a function from $Z$ onto itself that is its own invverse.
+
+\subsection{A Differentiable Programming Language for Manipulating Traces}
+Like the probabilistic programming languages described above, our differentiable language for involutions on traces is embedded based on Julia.
+The differentiable language has specialized expressions for reading values from addresses in a trace, writing values to the addresses in a trace, and copying values from an address in one trace to an address in another trace.
+For example, to read a value from a trace, we use the \texttt{@read} keyword:
+\noindent
+\begin{center}
+\begin{tabular}{c}
+{\begin{lstlisting}[basicstyle=\small\ttfamily]
+    value = @read(<trace>, <address>, <type>)
+\end{lstlisting}}
+\end{tabular}
+\end{center}
+The first argument defines what trace we are reading from.
+The second argument is the address in that trace, which may be hierarchical.\footnote{
+Hierarchical addresses of the form $\mathtt{a.b.c}$ are constructed in the language using the syntax \texttt{a=>b=>c}.}
+The third argument is either $\mathtt{:}\mathtt{discrete}$ or $\mathtt{:}\mathtt{continuous}$, and informs the language runtime whether the random choice at that address is drawn from a discrete or continuous distribution (this information is used to support automatic differentiation, as discussed in Section~\ref{sec:automating}).
+Similarly, to write a value to a trace, we use a \texttt{@write} keyword:
+\noindent
+\begin{center}
+\begin{tabular}{c}
+{\begin{lstlisting}[basicstyle=\small\ttfamily]
+    @write(<trace>, <address>, <value>, <type>)
+\end{lstlisting}}
+\end{tabular}
+\end{center}
+The mathematical behavior of the function is defined simply by writing Julia code that implements it.
+For example, the following code reads the values of parameters in polar coordinates and writes out their values in cartesian coordinates:
+\noindent
+\begin{center}
+\begin{tabular}{c}
+{\begin{lstlisting}[basicstyle=\small\ttfamily]
+r = @read(trace1, :r, :continuous)
+theta = @read(trace1, :theta, :continuous)
+x = r * sin(theta)
+y = r * cos(theta)
+@write(trace2, :x, x, :continuous)
+@write(trace2, :y, y, :continuous)
+\end{lstlisting}}
+\end{tabular}
+\end{center}
+It is also possible to copy values from an address in one trace to an address in another trace:
+\noindent
+\begin{center}
+\begin{tabular}{c}
+{\begin{lstlisting}[basicstyle=\small\ttfamily]
+@copy(<from-trace>, <from-addr>,
+      <to-trace>, <to-addr>)
+\end{lstlisting}}
+\end{tabular}
+\end{center}
+As we will see in Section~\ref{sec:automating}, it is preferable to copy when possible instead of reading and then writing, as the additional information can improve the efficiency of automatic differentiation.
+
+The names of the input traces and output traces are defined as part of the function signature, which as the following form, where \texttt{h} is the name of the function being defined.
+\noindent
+\begin{center}
+\begin{tabular}{c}
+{\begin{lstlisting}[basicstyle=\small\ttfamily]
+@tracefn h from (tr1, tr2) to (tr3, tr4) begin
+    ..
+end
+\end{lstlisting}}
+\end{tabular}
+\end{center}
+ % TODO change it from trace_bijection to tracefn in the figures and implementation...
+Each input and output trace is given a name (e.g. \texttt{tr1}, \texttt{tr2}, \texttt{tr3}, and \texttt{tr4}) and these names must be unique.
+For involution MCMC, the function must have two input traces and two output traces.
+The input trace is the input model trace ($x$), the second input trace is the input auxiliary trace ($y$), the first output trace is the output model trace ($x'$) and the second output trace is the output auxiliary trace ($y'$).
+
+\subsection{Implicit Copies}
+
+If an address is present in the input model trace and is needed to form a complete output model trace but was not written to or copied to in by the function body, then it will be implicitly copied from the previous trace by the runtime implementation.
+This significantly reduces the amount of code that has to be written when only a small subset of the model trace is being modified during the application of the involution MCMC move (which is common in applications of reversible jump MCMC).
+As a simple example, if there are $N$ latent variables, and the move only updates 1 of them, then the involution needs to contain only $O(1)$ lines of code instead of $O(N)$ lines.
+
+\subsection{Reasoning About Involutions}
+Recall the ealier code snippet that maps from polar to cartesian coordinates.
+This code snippet defines a bijection from one coordinate system to another.
+However, it does not define an involution, because applying the function the first time produces a trace containing cartesian coordinates, and applying the function a second time to the resulting trace would fail, because the function reads in polar coordinates from its input trace.
+How can we make the function an involution?
+
+First, the function will need the ability to go from cartesian coordinates to polar coordinates, so the following code for the inverse of the forward transformation must appear in the function somewhere:
+\noindent
+\begin{center}
+\begin{tabular}{c}
+{\begin{lstlisting}[basicstyle=\small\ttfamily]
+r = (x^2 + y^2)
+theta = atan(y, x)
+\end{lstlisting}}
+\end{tabular}
+\end{center}
+To be an involution, the input and output traces must contain some information indicating whether they are using polar or cartesian coordinates, so the involution code knows which of the two transformations to read from, and which addresses to read from and write to.
+Below is a minimal example of the involution, that uses a Boolean address polar.
+Note that the involution negates the value of this address, to indicate that the coordinate system is switching with each application.
+\noindent
+\begin{center}
+\begin{tabular}{c}
+{\begin{lstlisting}[basicstyle=\small\ttfamily]
+@tracefn h from (tr1, tr2) to (tr3, tr4)
+   polar = @read(tr1, :polar, :discrete)
+   if polar
+      r = @read(tr1, :r, :continuous)
+      theta = @read(tr1, :theta, :continuous)
+      x = r * sin(theta)
+      y = r * cos(theta)
+      @write(tr3, :x, x, :continuous)
+      @write(tr3, :y, y, :continuous)
+   else
+      x = @read(tr1, :x, :continuous)
+      y = @read(tr1, :y, :continuous)
+      r = (x^2 + y^2)
+      theta = atan(y, x)
+      @write(tr3, :r, r, :continuous)
+      @write(tr3, :theta, theta, :continuous)
+   end
+   @write(tr3, :polar, !polar, :discrete)
+end
+\end{lstlisting}}
+\end{tabular}
+\end{center}
+In order to apply this function in the context of involution MCMC, we require probabilistic programs $P$ and $Q$ that define space of input and output traces for the involution.
+A minimal example does not require any auxiliary random choices, so $Q$ can be an empty probabilistic program.
+A minimal toy example of a probabilistic program $P$ that is compatible with the involution \texttt{h} defined above is:
+\noindent
+\begin{center}
+\begin{tabular}{c}
+{\begin{lstlisting}[basicstyle=\small\ttfamily]
+@gen function p()
+    polar ~ bernoulli(0.5)
+    if polar
+        r ~ gamma(1, 1)
+        theta ~ uniform(-pi, pi)
+    else
+        x ~ normal(0, 1)
+        y ~ normal(0, 1)
+    end
+end
+\end{lstlisting}}
+\end{tabular}
+\end{center}
+
+% TODO do a PyTorch implementation??
+
+\section{AUTOMATIC ACCEPTANCE PROBABILITY CALCULATION} \label{sec:automating}
 

### 2020-07-12 paper.tex
Add text on Radon-Nikodym to the differentiable language section

@@
 \section{INVOLUTIVE MCMC} \label{sec:involutive-mcmc}
+This section gives a measure-theoretic formulation of involutive MCMC on general state spaces.
+This generality is necessary for applications of involutive MCMC to models with structure uncertainty and for describing how involutive MCMC can be automated using probabilistic programming.
+
 Let $(X, \Sigma_P, \mu_P)$ and $(Y, \Sigma_Q, \mu_Q)$ denote two measure spaces with $\sigma$-finite $\mu_P$ and $\mu_Q$.
@@
 
+TODO: show an example
+
 \subsection{Reasoning About Involutions}
@@
 \end{center}
+
+\subsection{The Radon-Nikodym Derivative}
+
+The differentiable programming language introduced in this section assumes that the reference measure of each random choice address is either the counting measure (corresponding to the $\mathtt{:}\mathtt{discrete}$ tag in \texttt{@read} and \texttt{@write} operations) or a Lebesgue measure ($\mathtt{:}\mathtt{continuous}$ tag).
+Consider a bijection $f : Z \to Z$ (recall that every involution is a bijection).
+If the space $Z$ can be partitioned into a countable number of sets $Z_1, Z_2, \ldots$, such that the each element of the partition is in one-to-one correspondence with another element ($f_1 : \{Z_1, Z_2, \ldots\} \to \{Z_1, Z_2, \ldots\}$ for some bijection $f_1$), and such that $f_{2,i} : Z_i \to Z_{f_1(i)}$ is a continuously differentiable bijection, and such that $f(z) = f_{2,i(z)}(z)$, then a Radon-Nikodym can be constructed from the Jacobian of the function $f_{2,i(z)}$ where $i(z) = Z_j$ for $z \in Z_j$.
+See Section~\ref{sec:radon-nikodym-special-case} of the appendix for details.
+Applying this to the setting where $Z$ is a set of pairs of traces of probabilistic programs as formalized in the preivous section is straightforward when the partition of $Z$ is the countable disjoint union over all pairs $(A, \mathbf{d})$ where $A \subseteq \mathcal{A}$ are trace structures and $\mathbf{d}$ are assignments to discrete choices.
 

### 2020-07-12 paper.tex
add Jacobian sparsity section

@@
 
-\aistatstitle{Automated Involutive MCMC}
+\aistatstitle{Automating Involutive MCMC using Probabilistic and Differentiable Programming}
 
@@
 If the space $Z$ can be partitioned into a countable number of sets $Z_1, Z_2, \ldots$, such that the each element of the partition is in one-to-one correspondence with another element ($f_1 : \{Z_1, Z_2, \ldots\} \to \{Z_1, Z_2, \ldots\}$ for some bijection $f_1$), and such that $f_{2,i} : Z_i \to Z_{f_1(i)}$ is a continuously differentiable bijection, and such that $f(z) = f_{2,i(z)}(z)$, then a Radon-Nikodym can be constructed from the Jacobian of the function $f_{2,i(z)}$ where $i(z) = Z_j$ for $z \in Z_j$.
+We denote the Jacobian (determinant) of a function $g$ by $| J g |$.
 See Section~\ref{sec:radon-nikodym-special-case} of the appendix for details.
@@
 
-% TODO can only show this for a special case -- discrete involution + continuous?, or some more general case?
+The general form of the acceptance probability in Algorithm~\ref{alg:involutive-mcmc} is:
+\begin{equation}
+\min\left\{1, \frac{p(x') q_{x'}(y')}{p(x) q_{x}(y)} \left( \frac{d (\mu \circ f^{-1})}{d \mu} (x, y)\right) \right\}
+\end{equation}
+For the setting when $p$, $q$, and $f$ are defined using the probabilistic and differentiable programming languages of this section, the probability simplifies to:
+\begin{equation} \label{eq:accept-prob-jacobian}
+\min\left\{1, \frac{p(x') q_{x'}(y')}{p(x) q_{x}(y)} |J f_{2,i(x,y)} | \right\}
+\end{equation}
+Implementing the computation of this probability by hand is error-prone due to conceptual, arithmetic, and implementation errors; as well as time consuming and tedious.
+This section describes how the computation of this probabability is automated by Gen using probabilistic and differentiable programming techniques.
+
+\subsection{Automatically Computing Probability Density Ratios}
+
+When the model density function $p$ and the auxiliary density functions $q_x$ in Algorithm~\ref{alg:involutive-mcmc} are specified by writing probabilistic programs in the language of Section~\ref{sec:prob-prog-traces}, Gen automatically computes the following ratio of densities, given the previous traces $x$ and $y$ and the proposed traces $x'$ and $y'$:
+\begin{equation} \label{eq:density-ratio}
+\frac{p(x') q_{x'}(y')}{p(x) q_x(y)}
+\end{equation}
+The density $p$ of a probabilistic program with respect to the reference measure $\mu$, evaluated at a trace $x = (A, \mathbf{x})$, is the product of individual densities for each random choice (and similarly for the auxiliary densities and the auxiliary probabilistic program):
+\begin{equation} \label{eq:density-product}
+p((A, \mathbf{x})) = \prod_{a \in A} p_{\sigma_a}(\mathbf{x}[a])
+\end{equation}
+where $\mathbf{x}[a]$ denotes the value of address $a$ in the trace, $p_{\sigma_a}$ is the probability density function for distribution from which random choice $a$ is sampled, given that the state of the program when sampling is $\sigma_a$.
+(See~\citet{popl} for a formal proof for a somewhat more restricted probabilistic programming language.)
+Therefore, the probabilistic programming language can compute the density at a trace $x$ by executing the program, but at each random choice $a$, reading the value of the choice $\mathbf{x}[a]$ instead of sampling it, computing the density at $\mathbf{x}[a]$ and accumulating it into the product above.\footnote{Probabilistic programming implementations use sums of log-densities instead of products for numerical reasons.}
+
+To evaluate this density ratio, we utilize Gen's API for probabilistic programs and traces, which automatically computes the density of a trace of a probabilistic program, whether that trace was sampled from the probabilistic program (as in the case of computing $q_x(y)$ for $y \sim q_{x}(\cdot)$) or when that trace was produced externally from the probabilistic program (as in the case of computing $q_{x'}(y')$, where $y'$ was produced from the involution $f$).
+
+\subsubsection{Incremental Computation of Model Density Ratio}
+While it is possible to compute the ratio of densities in Equation~(\ref{eq:density-ratio}) by separately computing the four densities of the four traces, this scales linearly in the total number of random choices in the traces, because every random choice in each trace needs to be visited to compute the products in Equation~(\ref{eq:density-product}).
+Often, the number of random choices in the auxiliary traces $y$ and $y'$ are constant size $O(1)$ whereas the number of random choices in the model traces $x$ and $x'$ is $O(N)$ where $N$ is the number of data points, or latent components
+(see Figure~\ref{fig:mixture} for an example, where at most two clusters are affected by each application of the MCMC kernel).
+In these cases, the ratio of model trace densities $p(x') / p(x)$ dominates the asymptotic running time of computing the density ratio of Equation~\ref{eq:density-ratio}.
+However, due to conditional independencies in the model, the densities $p(x')$ and $p(x)$ often share common factors that cancel, which affords an opportunity for $O(1)$ computation of the ratio.
+Gen's trace API includes an \emph{incremental update} operation~\citep{gen-paper}, which takes a previous trace $x$, and a partial trace $u$ containing the values of random choices that should be changed (or added if there is stochastic structure in the model), and returns a new trace $x'$.
+This allows for $O(1)$ computation of $p(x') / p(x)$ when the modeling language runtime system is able to detect and exploit the cancellation.
+
+\subsection{Sparsity-Aware Automatic Jacobian Computation}
+
+It is possible to automatically compute the Jacobian determinant in Equation~(\ref{eq:accept-prob-jacobian}) using automatic differentiation of the involution $f$.
+Note that while the involution can manipulate both discrete and continuous random choices, the Jacobian is the matrix of partial derivatives associated with the continuous random choices only.
+The differentiable programming language implementation distinguishes between continuous and discrete choices with the help of the user-provided type tags.
+The sums of the number of continuous dimensions across all addresses in the input and the output traces respectively must match, because otherwise $f$ cannot factor into continuously differentiable bijections $f_{2,i}$, which is necessary for the involutive MCMC acceptance ratio to simplify to Equation~(\ref{eq:accept-prob-jacobian}) (see Section~\ref{sec:radon-nikodym-special-case}).
+
+The Jacobian is computed automatically by applying forward-mode automatic differentiation separately for each continuous random choice in the input traces, which scales as $O(N^2)$ where $N$ is the number of such choices.
+Common determinant implementations based on the LU-decomposition scale on the order $O(N^2)$ for an $N$-by-$N$ matrix.
+Therefore, the basic algorithm computes $|J f_{2,i(x,y)} |$ in $O(N^2)$ where $N$ is the number of continuous addresses in $x$ or $y$.
+
+To derive a more efficeint algorithm, we use the cofactor expansion of the determinant, and observe that any row or column in the Jacobian matrix that contains a single $1$ and all other ($N-1$) entries $0$ implies that the absolute value of the determinant is equivalent to the absolute value of the $N-1$-by-$N-1$ sub-matrix with the corresponding row and column omitted (even if that would remove other nonzero entries from the matrix).
+Applying this rule recursively allows a significant reduction in the size of the determinant computation when the Jacobian matrix possesses sparsity.
+Sparse Jacobians are common in applications of involutive MCMC, and reversible jump MCMC in particular.
+For example, consider the Jacobian for the split-merge move of~\citet{richardson1997bayesian}, implemented in Figure~\ref{fig:mixture}:
+\includegraphics[width=\linewidth]{figures/sparse-jacobian.pdf}
+In this case, the Jacobian determinant can be simplified to the determinant of a $6$-by-$6$ matrix that does not grow in size with the number of clusters and number of data points, because unaffected clusters and data points have their values \emph{copied} from the input model trace to the output model trace.
+While a reasonable hand-coded implementation of this algorithm would likely exploit this sparsity as well, it would require careful and error-prone reasoning and implementation, which our approach automates.
 

### 2020-07-13 paper.tex
Add Section 7, fix some typos

@@
 A probabilistic program $P$ denotes a probability density $p : T \to [0, \infty)$ with respect to a reference measure $\mu$ on traces.
-Describing the denotational semantics of a particular probabilistic programming language (the function mapping programs $P$ to density functions $p$) is out of scope for this paper, and we refer to the reader to~\citet{popl} for an example of denotational semantics of a probabilistic programming language that uses explicit addresses for random-choices.
+Describing the denotational semantics of a particular probabilistic programming language (the function mapping programs $P$ to density functions $p$) is out of scope for this paper, and we refer to the reader to~\citet{popltraces} for an example of denotational semantics of a probabilistic programming language that uses explicit addresses for random-choices.
 But, informally, the density of a trace in our formalism (and existing formalisms) is the product of densities of each individual random choice, as an example below illustrates.
@@
 \subsection{Reasoning About Involutions}
-Recall the ealier code snippet that maps from polar to cartesian coordinates.
+Recall the earlier code snippet that maps from polar to cartesian coordinates.
 This code snippet defines a bijection from one coordinate system to another.
@@
 Implementing the computation of this probability by hand is error-prone due to conceptual, arithmetic, and implementation errors; as well as time consuming and tedious.
-This section describes how the computation of this probabability is automated by Gen using probabilistic and differentiable programming techniques.
+This section describes how the computation of this probability is automated by Gen using probabilistic and differentiable programming techniques.
 
@@
 where $\mathbf{x}[a]$ denotes the value of address $a$ in the trace, $p_{\sigma_a}$ is the probability density function for distribution from which random choice $a$ is sampled, given that the state of the program when sampling is $\sigma_a$.
-(See~\citet{popl} for a formal proof for a somewhat more restricted probabilistic programming language.)
+(See~\citet{popltraces} for a formal proof for a somewhat more restricted probabilistic programming language.)
 Therefore, the probabilistic programming language can compute the density at a trace $x$ by executing the program, but at each random choice $a$, reading the value of the choice $\mathbf{x}[a]$ instead of sampling it, computing the density at $\mathbf{x}[a]$ and accumulating it into the product above.\footnote{Probabilistic programming implementations use sums of log-densities instead of products for numerical reasons.}
@@
 
-To derive a more efficeint algorithm, we use the cofactor expansion of the determinant, and observe that any row or column in the Jacobian matrix that contains a single $1$ and all other ($N-1$) entries $0$ implies that the absolute value of the determinant is equivalent to the absolute value of the $N-1$-by-$N-1$ sub-matrix with the corresponding row and column omitted (even if that would remove other nonzero entries from the matrix).
+To derive a more efficient algorithm, we use the cofactor expansion of the determinant, and observe that any row or column in the Jacobian matrix that contains a single $1$ and all other ($N-1$) entries $0$ implies that the absolute value of the determinant is equivalent to the absolute value of the $N-1$-by-$N-1$ sub-matrix with the corresponding row and column omitted (even if that would remove other nonzero entries from the matrix).
 Applying this rule recursively allows a significant reduction in the size of the determinant computation when the Jacobian matrix possesses sparsity.
@@
 For example, consider the Jacobian for the split-merge move of~\citet{richardson1997bayesian}, implemented in Figure~\ref{fig:mixture}:
-\includegraphics[width=\linewidth]{figures/sparse-jacobian.pdf}
+%\includegraphics[width=\linewidth]{figures/sparse-jacobian.pdf}
 In this case, the Jacobian determinant can be simplified to the determinant of a $6$-by-$6$ matrix that does not grow in size with the number of clusters and number of data points, because unaffected clusters and data points have their values \emph{copied} from the input model trace to the output model trace.
@@
 \section{A DYNAMIC CHECK FOR DETECTING BUGS}
+
+Due to the automation described in the previous section, users of Gen can implement involutive MCMC algorithms by writing probabilistic programs for $p$ and $q$ and a differentiable program for the involution $f$, avoiding the need to derive and implement the accept/reject formula by hand.
+This eliminates certain classes of errors that could otherwise be difficult to root out in hand-coded implementations of MCMC algorithms.
+However, it is still of course possible to implement $p$, $q$, or $f$ incorrectly, introducing bugs that may invalidate the correctness proofs for the involutive MCMC algorithm.
+
+Fortunately, the factorization of diverse algorithms into a common template, involving an explicitly represented model, auxiliary distribution, and involution, enables simple and automated debugging checks that can catch a wide variety of such errors dynamically.
+These checks are not complete, but anecdotally, we have found that they often alert users to subtle bugs in the design or implementation of an algorithm.
+The checks are summarized in Algorithm~\ref{alg:dynamic-check}.
+
+\begin{algorithm}[h]
+\begin{algorithmic}
+\Procedure{imcmc-dynamic-check}{$p$, $q$, $f$}
+    \State $x \sim p(\cdot)$ \Comment{Sample state from prior}
+    \State $y \sim q_x(\cdot)$ \Comment{Sample auxiliary variables}
+    \State $(x', y') \gets f(x, y)$ \Comment{Apply involution}
+    \State {\bf assert} $p(x')q_{x'}(y') > 0$ \Comment{Support check}
+    \State {\bf assert} $f(x', y') = (x, y)$ \Comment{Involution check}
+    \State {\bf assert} $|\text{Rd}^{\textit{cont}}_{f,x,y}| = |\text{Wr}^{\textit{cont}}_{f,x,y}|$ \Comment{Dimension check}
+\EndProcedure
+\end{algorithmic}
+\caption{Dynamic Check for Bugs}
+\label{alg:dynamic-check}
+\end{algorithm}
+
+To test the user's algorithm, Gen first samples a model trace $x$ from the model prior $p$, and auxiliary randomness $y$ from the kernel $q_x$. It then runs three tests:
+
+\begin{enumerate}
+\item \textbf{Support check.} The \textit{support check} runs the involution $f$ on $(x, y)$ and checks that the resulting pair of traces, $(x', y')$, are within the set $Z$ of positive-density elements for $\pi$.
+
+\item \textbf{Involution check.} The \textit{involution check} runs $f \circ f$ on $(x, y)$ and checks that the result is equal to $(x, y)$.
+If $f$ is an involution on $Z$, then this check will always succeed; if there is a set $\tilde{Z}$ of positive $\pi$-measure on which $f(f(z)) \neq z$ (i.e., $f$ is not an involution), then this check has positive probability of failing.
+
+\item \textbf{Dimension check.} The \textit{dimension check} runs the involution $f$ on $(x, y)$ and records the sets of addresses $\text{Rd}_{f,x,y}^{\textit{cont}}$ and $\text{Wr}_{f,x,y}^{\textit{cont}}$ at which $f$ performs continuous reads and writes, respectively.
+It then checks that the sizes of these sets match.
+(Addresses with an $n$-dimensional Lebesgue measure as their reference measure, corresponding to vector-valued random choices, are unpacked into $n$ separate addresses for the purposes of this check.)
+\end{enumerate}
+
+These checks each catch qualitatively different bugs in user programs. We now give several examples.
+
+\textit{Incorrectly transformed continuous variables.} Many bugs in the design or implementation of deterministic transformations of continuous variables are naturally detected by the involution check.
+Consider, for example, the Hamiltonian Monte Carlo algorithm~\citep{hmc}, which, as \citet{imcmc} observe, is an instance of the involutive MCMC framework.
+%$x$ is the model state, $y$ is the momentum, and the involution $f$ first runs a leapfrog integrator to obtain a new state $x'$ and final momentum $-y'$, then negates the final momentum and returns $(x', y')$.
+When applied to HMC, Gen's dynamic bug check involves:
+%Gen's dynamic checks will uncover mathematical errors in the implementation of the leapfrog integrator, and will also detect a failure to perform the final negation of the momentum.
+In particular, the dynamic check involves
+(a) sampling a model state $x$ from the prior; (b) sampling a momentum $y$; (c) running the leapfrog integrator forward to get a new state $x'$ and new momentum $-y'$, (d) running the leapfrog integrator backward from state $x'$ with momentum $y'$, and (e) checking that this results in the state-momentum pair $x, -y$.
+If the momentum is not properly negated, or if the leapfrog integrator is incorrectly implemented, this check can fail.
+
+As another, simpler example of this type of error, consider the \texttt{split\_params} and \texttt{merge\_params} functions invoked in Figure~\ref{fig:mixture}, as part of a reversible-jump MCMC kernel for learning a mixture of Gaussians with an unknown number of components.
+Considering only the $\mu$ parameter, suppose the \textit{split} move samples an auxiliary variable $u_1$ and computes $\mu_1 = \mu + u_1, \mu_2 = \mu - u_1$ as the means of the two new clusters.
+If the \textit{merge} move joins two clusters and assigns $\mu = \sqrt{\mu_1 \mu_2}$, there is a mismatch: a split move cannot be reversed by a corresponding merge, because in general, $\sqrt{(\mu+u_1)(\mu-u_1)} \neq \mu$.
+This error could be fixed either by changing the split to compute $\mu_1 = \mu u_1, \mu_2 = \frac{\mu}{u_1}$, or by changing the merge to compute $\mu = \frac{\mu_1 + \mu_2}{2}$.
+
+As a final, rather different example in this class, suppose approximate inverses $F_\theta, G_\theta : \mathbb{R}^d \rightarrow \mathbb{R}^d$ are learned neural networks trained via a Cycle GAN objective~\citep{cyclegan} to translate between the latent spaces of two models in a Bayesian model selection problem.
+A user might implement the following involution:
+
+\begin{center}
+\begin{tabular}{c}
+{\begin{lstlisting}[basicstyle=\small\ttfamily]
+@tracefn h from (tr1, tr2) to (tr3, tr4)
+   which_model = @read(tr1, :model, :discrete)
+   if which_model == 1
+      z = @read(tr1, :z, :continuous)
+      z_in_model_2 = F(z)
+      @write(tr3, :z, z_in_model_2, :continuous)
+      @write(tr3, :model, 2, :discrete)
+   else
+      z = @read(tr1, :z, :continuous)
+      z_in_model_1 = G(z)
+      @write(tr3, :z, z_in_model_1, :continuous)
+      @write(tr3, :model, 1, :discrete)
+   end
+end
+\end{lstlisting}}
+\end{tabular}
+\end{center}
+
+However, even though $F_\theta$ and $G_\theta$ are trained with
+an objective that encourages $F_\theta(G_\theta(z)) \approx z$, 
+this relationship is not exact equality, so the involutive MCMC
+framework cannot be directly applied, and the involution check
+will fail. This error could be fixed by sampling a proposed value
+\textit{near} $F_\theta(z)$ or $G_\theta(z)$, rather than directly applying $F_\theta$
+and $G_\theta$ to $z$.
+
+\textit{Discrete logic errors.} Another common class of errors is 
+for the discrete logic of an involution to be flawed. Consider the following incorrect 
+implementation of a \textit{birth-death} move for the mixture model in Figure~\ref{fig:mixture}, which either adds a new mixture component or selects an existing one
+at random to delete:
+
+\begin{center}
+\begin{tabular}{c}
+{\begin{lstlisting}[basicstyle=\small\ttfamily]
+@gen function q(trace)
+  current_k = trace[:k]
+  is_birth ~ bernoulli(current_k == 1 ? 1.0 : 0.5)
+  if is_birth
+    new_mu ~ normal(0, 10)
+    new_var ~ inv_gamma(1, 10)
+  else
+    deletion_idx ~ uniform_discrete(1, current_k)
+  end
+end
+
+@tracefn h from (tr1, tr2) to (tr3, tr4)
+   is_birth = @read(tr2, :is_birth, :discrete)
+   @write(tr4, :is_birth, !is_birth, :discrete)
+   k = @read(tr1, :k, :discrete)
+   weights = @read(tr1, :weights, :continuous)
+   if is_birth
+     @write(tr3, :k, k+1, :discrete)
+     new_mu = @read(tr2, :new_mu, :continuous)
+     new_var = @read(tr2, :new_var, :continuous)
+     @write(tr3, (:mu, k+1), new_mu, :continuous)
+     @write(tr3, (:var, k+1), new_var, :continuous)
+     new_weights = add_weight(weights)
+     @write(tr4, :deletion_idx, k+1, :discrete)
+   else
+     idx = @read(tr2, :deletion_idx, :discrete)
+     @copy(tr1, (:mu, idx), tr4, :new_mu)
+     @copy(tr1, (:var, idx), tr4, :new_var)
+     for i in (idx+1):k
+       @copy(tr1, (:mu, i), tr3, (:mu, i-1))
+       @copy(tr1, (:var, i), tr3, (:var, i-1))
+     end
+     @write(tr3, :k, k-1, :discrete)
+     new_weights = delete_weight(weights, i)
+   end
+   @write(tr3, :weights, new_weights, :continuous)
+end
+\end{lstlisting}}
+\end{tabular}
+\end{center}
+
+The flaw in this implementation is that although the death move can 
+delete any of the $k$ mixture components, the birth move can only add
+a new component to the \textit{end} (index $k+1$), so the move is not
+reversible. Gen's involution check will discover that the deletion of a
+component with index $i < k$ is not reversed by a corresponding birth move,
+and will thus raise an error.
+
+\textit{Other miscellaneous errors.} When implementing distributions as probabilistic programs, it is also possible for users to make more mundane errors, such as spelling the name of a random choice inconsistently, or characterizing random choices using the wrong \texttt{type} tags (\texttt{:continuous} and \texttt{:discrete}).
+Such errors can be difficult to detect statically, because the addresses at which a probabilistic program makes random choices, and the distributions of those choices, may change from sample to sample.
+(Previous work has explored static analyses based on types~\citep{popltraces} and abstract interpretation~\citep{verified_svi}, but these each work on limited subsets of the programs that Gen's full modeling language permits, and it is often precisely these more complex programs that require the flexibility of the involutive MCMC framework in the first place.) Gen's dynamic support and dimension checks can help to detect bugs like these.
+For example, if the involution $f$ writes to a misspelled address, the support check will determine that the resulting trace's density is 0.
 

### 2020-07-13 paper.tex
first pass on reversible jump example section

@@
 \section{EXAMPLES}
+
+\subsection{Reversible Jump MCMC}
+Reversible jump MCMC~\citep{green1995reversible,hastie2012model} is a special case of involutive MCMC, and 
+the implementation of reversible jump MCMC kernels can be automated using the probabilistic and differentiable programming languages presented in this paper.
+We now review reversible jump MCMC, then show how it can be automated using the techniques presented earlier, and give an example.
+
+The reversible jump MCMC framework involves a set of `models' $k \in \mathcal{K}$, and a prior distribution on models $p(k)$.
+For each model, there is a latent continuous parameter vector $\theta_k \in \mathbb{R}^{n(k)}$ where $n(k)$ is the dimension of model $k$, and a likelihood function $L_{D,k}(\theta_k)$ for each $k$ given data $D$.
+The latent state $x$ is a pair $(k, \theta_k)$ of model and continuous parameter.
+There is a set of \emph{move types} $\mathcal{M}$.
+Each move type $m \in \mathcal{M}$ is associated with an unordered pair of models $(k_1, k_2)$ and a dimensionality $d(m)$ such that $d(m) \ge n(k_1)$ and $d(m) \ge n(k_2)$ (zero, one, or more than one move types maybe assocaite with a given pair of models).
+For each latent state $x = (k, \theta_k)$, there is a probability distribution $q_{x}(m)$ on move types such that $q_{x}(m) > 0$ implies that $k$ is one of the models for move type $m$.
+For each move type $m \in \mathcal{K}$ between $k_1$ and $k_2$ there is a pair of continuously differentiable bijections $h_{m, k_1 \to k_2} : \mathbb{R}^{d(m)} \to \mathbb{R}^{d(m)}$ and $h_{m, k_2 \to k_1} := h_{m, k_1 \to k_2}^{-1}$, and a pair of proposal densities $q_{m,k_1 \to k_2}(u_{k_1 \to k_2})$ and $q_{m,k_2 \to k_1}(u_{k_2 \to k_1})$ where $u_{k_1 \to k_2} \in \mathbb{R}^{d(m) - n(k_1)}$ and $u_{k_2 \to k_1} \in \mathbb{R}^{d(m) - n(k_2)}$.
+A proposal is made from state $x = (k, \theta_k)$ by (i) sampling a move type $m \sim q_{x}(\cdot)$, and (ii) sampling continuous variable $u \sim q_{m,k \to k'}(\cdot)$ for $(k, k')$ associated with $m$, and (iii) computing $(\theta_{k}', u') := h_{m,k \to k'}(\theta_k, u)$, and proposing new state $x' = (k', \theta_k')$.
+The move is accepted with probability:
+\begin{equation}
+{
+\min\left\{ 1,
+\frac{p(k')}{p(k)} \frac{p_{k'}(\theta_{k'})}{p_k(\theta_k)} \frac{L_{D,k'}(\theta_k')}{L_{D,k}(\theta_k)}
+\frac{q_{x'}(m)}{q_{x}(m)}
+\frac{q_{m,k' \to k}(u)}{q_{m,k \to k'}(u)}
+\right\}
+}
+\end{equation}
+
+To represent reversible jump MCMC using involutive MCMC implemented with probabilistic programs, we write a probabilistic program $P$ that encodes the space of models $\mathcal{K}$, the prior distribution on models, $p(k)$, the per-model priors $p_k(\theta_k)$ and the per-model likelihoods $L_{D,k}(\theta_k)$.
+The set of all models $k$ corresponds to the set of all pairs $(A_P, \mathbf{d}_P)$ where $A_P$ represent possible trace structures (i.e.
+control-flow paths through $P$) and $\mathbf{d}_P$ are the set of assignments to discrete random choices made by $P$.
+The per-model continuous parameters $\theta$ correspond to continuous random choices $\mathbf{c}_P$.
+% TODO explain the likelihood
+An auxiliary probabilistic program $Q$ encodes both the probability distribution on moves types using discrete random choices and possibly stochastic control flow ($(A_Q, \mathbf{d}_Q)$), and the per-move-type probability densities on $u$ using continuous random choices $\mathbf{c}_Q$.
+The involution $f$ factors into an (i) involution $f_1$ on pairs $i = ((A_P, \mathbf{d}_P), (A_Q, \mathbf{d}_Q))$ that defines the association between move types $\mathcal{M}$ and the model pairs ($k_1, k_2$); and (ii) a family of bijections $f_{1,i}$ on the space of pairs $(\mathbf{c}_P, \mathbf{c}_Q)$ of continuous random choices for both programs for fixed values of the discrete random choices and fixed trace structure.
+
+Figure~\ref{fig:mixture} shows a split-merge reversible jump kernel for an infinite Gaussian mixture model~\citep{richardson1997bayesian} implemented using the probabilistic and differentiable programming languages described in this paper.
+Figure~\ref{fig:mixture}(b) shows the infinite Gaussian mixture model, specified as a probabilistic program $P$.
+The program takes the number of data points as input, then samples the number of clusters from a Poisson distribution, then samples cluster parameters and mixture proportions, and finally samples the data points from the resulting finite mixture.
+Figure~\ref{fig:mixture}(c) shows the auxiliary probabilistic program $Q$ for the split-merge kernel.
+This program takes a trace of the model program as input, and randomly decides whether to split a cluster and increase the number of clusters by one or merge two clusters and decrease the number of clusters by one.
+Then, the program randomly picks which cluster to split, or which clusters to merge.
+This kernel always merges the last cluster with a random other cluster; for ergodicity the move can be composed with a simple move (that has acceptance probability $1$) that swaps a random cluster with the last cluster.
+If a split is chosen, then the program also samples the three degrees of freedom necessary to generate the new parameters for the clusters in an invertible manner.
+Figure~\ref{fig:mixture}(d) shows a differentiable program specifying the involution for the split-merge kernel, and Figure~\ref{fig:mixture}(f) shows graphically how this involution acts on pairs of traces.
+The yellow section (1) defines an involution $f_1$ on the discrete random choices that specifies that (i) the \texttt{split} choice should be flipped (so that split moves are always mapped to merge moves and vice versa) and that (ii) the number of clusters should be increased by one for a split move and decreased by one for a merge move, and (iii) which merged cluster corresponds to which split clusters.
+The green section (2) specifies the continuous bijections that govern the transformation of continuous random choices during split moves and the purple section (3) specifies the inverses of these bijections, which govern the transformation of continuous choices during merge moves.
+
+\subsection{State-Dependent Mixture Proposal}
+
+State-dependent mixture proposal.
+Depending on the current state of the model, a different mixture distribution is used for the proposal.
+Implementing such a move requires computing the mixture probability under the forward move and under the reverse move, which may differ.
+
+$q_x(m)$ -- the mixture distribution
+
+$q_m(x')$ -- the proposal we are mixing over
+
+The first part of the auxiliary probabilistic program $Q$ determines which proposal to apply by sampling from a mixture distribution.
+Here, the program picks a random node in the tree, by doing a stochastic walk of the tree that terminates at the chosen node.
+\begin{center}
+\begin{tabular}{c}
+{\begin{lstlisting}[basicstyle=\small\ttfamily]
+prev_cov_function = trace[:cov_function]
+path ~ walk_tree(prev_cov_function, ..)
+\end{lstlisting}}
+\end{tabular}
+\end{center}
+The first part of the involution copies the random choices corresponding to the mixture from the input auxiliary trace to the output auxiliary trace.
+\begin{center}
+\begin{tabular}{c}
+{\begin{lstlisting}[basicstyle=\small\ttfamily]
+@copy(q_in, :path, q_out, :path)
+\end{lstlisting}}
+\end{tabular}
+\end{center}
+
+The code that walks the tree uses the following recursion, which results in a probability distribution that assigns exponentially lower probability to nodes that are deeper in the tree.
+\begin{center}
+\begin{tabular}{c}
+{\begin{lstlisting}[basicstyle=\small\ttfamily]
+if ({:done} ~ bernoulli(0.5))
+  return path
+elseif ({:recurse_left} ~ bernoulli(0.5))
+  path = (path..., :left_node)
+  return ({:left} ~ walk_tree(node.left, path))
+else
+  path = (path..., :right_node)
+  return ({:right} ~ walk_tree(node.right, path))
+end
+\end{lstlisting}}
+\end{tabular}
+\end{center}
+The resulting mixture distributions for two trees are:
+\begin{center}
+\includegraphics[width=0.8\linewidth]{figures/biased-tree-distribution.pdf} % TODO improve quality
+\end{center}
+Because the mixture is specified using a probabilistic program, it is straightforward to modify the program to define a different mixture distribution.
+The code below specifies a mixture distribution that is uniform over all nodes in the tree.
+\begin{center}
+\begin{tabular}{c}
+{\begin{lstlisting}[basicstyle=\small\ttfamily]
+n1 = size(node.left); n2 = size(node.right)
+if ({:done} ~ bernoulli(1 / (1 + n1 + n2)))
+  return path
+elseif ({:recurse_left} ~ bernoulli(n1 / (n1+n2))
+  path = (path..., :left_node)
+  return ({:left} ~ walk_tree(node.left, path))
+else
+  path = (path..., :right_node)
+  return ({:right} ~ walk_tree(node.right, path))
+end
+\end{lstlisting}}
+\end{tabular}
+\end{center}
+The resulting distributions are:
+\begin{center}
+\includegraphics[width=0.8\linewidth]{figures/unbiased-tree-distribution.pdf}
+\end{center}
+Note that in the first case, the ratio of mixture probabilities is either $1$, $0.5$, or $2$ depending on whether the previous and new subtrees are leaf or internal nodes.
+In the second case, the ratio of mixture probabilities is the ratio of sizes of the two trees.
+%Automating the computation of the mixture probabilities
+
+The rest of the kernel specifies what happens for each chosen node.
+The second part of the auxiliary probabilistic program $Q$ implements a common motif for Metropolis-Hastings algorithms, where the proposal samples new values for some random choices:
+\begin{center}
+\begin{tabular}{c}
+{\begin{lstlisting}[basicstyle=\small\ttfamily]
+new_subtree ~ cov_function_prior()
+\end{lstlisting}}
+\end{tabular}
+\end{center}
+and the involution swaps the old value with the newly proposed value:
+\begin{center}
+\begin{tabular}{c}
+{\begin{lstlisting}[basicstyle=\small\ttfamily]
+@copy(p_in, subtree_address, q_out, :new_subtree)
+@copy(q_in, :new_subtree, p_out, subtree_address)
+\end{lstlisting}}
+\end{tabular}
+\end{center}
+Here the values happen to be trees that themselves have a random number of random choices, but the same motif is also used in simpler applications of Metropolis-Hastings.
+
+Here, the proposal we are mixing over is easy to implement -- it involvs resampling a subtree from the prior, so the Jacobian is 1 and there is cancellation between the model and the proposal.
+
+But computing the mixture probabilities is complex, because the mixture is a mixture of either $n_1$ or $n_2$ elements, which depends on the size of the tree.
+
+Show the mixture probabilities in math notation?
+If both subtrees not leaf nodes, then the probabilities are the same.
+If both subtrees are leaf nodes, then the probabilities are different.
+If one subtree is a leaf node and the other is an internal node then the ratio is either $0.5$ or $2$ respectively.
+\begin{equation}
+q
+\end{equation}
+
+TODO: relate to `resimulation Metropolis-Hastings' in general-purpose probabilistic programs.
+
+Of course, it is straightforward to replace the prior with a custom data-driven proposal distribution on subtrees.
 

### 2020-07-13 paper.tex
add sparse jacobian fig

@@
 Let $y \sim q_x(\cdot)$ denote a random sample from the measure defined by $A \mapsto \int_A q_x(y) \mu_Q(dy)$ for $A \in \Sigma_Q$.
+
+% TODO what is a trace? and what is a p? do we need parents?
+%\begin{algorithm}[h]
+%\begin{algorithmic}
+%\Procedure{trimcmc-kernel}{$p$, $q$, $f$, $x$}
+    %\State $y \sim q_x(\cdot)$ \Comment{Sample auxiliary trace}
+    %\State $(x', y') \gets f(x, y)$ \Comment{Apply involution}
+    %\State $w_1 \gets \prod_{a \in A(x)} p_{A(x)}(x_a | x_{\mathrm{Pa}^{A(x)}_a})$
+    %\State $w_2 \gets \prod_{a \in A(y)} q_{x,A(y)}(y_a | x_{\mathrm{Pa}^{A(x)}_a})$
+    %\State $w_3 \gets \prod_{a \in A(x')} p_{A(x')}(x'_a | (x'_b)_{b \in \mathrm{Pa}^{A(x)}_a})$
+    %\State $w_4 \gets \prod_{a \in A(y')} q_{x',A(y')}(y'_b | (y'_b)_{b \in \mathrm{Pa}^{A(x)}_a})$
+    %\State $|J| \gets F.\textproc{jacdetabs}(x, y)$
+    %\State $\alpha \gets
+        %\displaystyle \frac{w_3 \cdot w_4}{w_1 \cdot w_2} \cdot w_5$
+    %\State $r \sim \mathrm{Uniform}(0, 1)$
+    %\State \algorithmicif \, $r \le \alpha$ \algorithmicthen \, \Return $x'$ \algorithmicelse \, \Return $x$ 
+%\EndProcedure
+
+\Procedure{$P$.score}{$x$}
+    %\State \Return $p(x)$ % TODO or product of densities? % TODO state notation is non-standard
+    % if we don't use this product then what is the point -- why don't we just use p(x)
+    \State \Return $\prod_{a \in A(x)} p_{\sigma(a)}(x[a])$ (?) TODO % TODO
+\EndProcedure
+
+\Procedure{$Q$.score}{$y$; $x$}
+    %\State \Return $q_x(y)$ % TODO or product of densities?
+    \State \Return $\prod_{a \in A(y)} q_{x,\sigma(a)}(y[a])$ (?) TODO % TODO
+\EndProcedure
+
+\Procedure{$F$.jacdetabs}{$x$,$y$}
+    \For{$a$ in $\textproc{continuous-addresses}(x)$}
+        \State $J1[a,:] \gets \nabla_{x[a]} \left( F(x, y)_{\mathrm{continuous}} \right)$ (?) TODO % TODO
+    \EndFor
+    \For{$a$ in $\textproc{continuous-addresses}(y)$}
+        \State $J2[a,:] \gets \nabla_{y[a]} \left( F(x, y)_{\mathrm{continuous}} \right)$ (?) TODO % TODO
+    \EndFor
+    \State $J \gets \left[\begin{array}{c}J1\\J2\end{array}\right]$
+    \State \Return $|\det(J)|$
+\EndProcedure
+\end{algorithmic}
+\caption{Trace Involutive MCMC}
+\label{alg:trimcmc}
+\end{algorithm}
 
@@
 For example, consider the Jacobian for the split-merge move of~\citet{richardson1997bayesian}, implemented in Figure~\ref{fig:mixture}:
-%\includegraphics[width=\linewidth]{figures/sparse-jacobian.pdf}
+\includegraphics[width=\linewidth]{figures/sparse-jacobian.pdf}
 In this case, the Jacobian determinant can be simplified to the determinant of a $6$-by-$6$ matrix that does not grow in size with the number of clusters and number of data points, because unaffected clusters and data points have their values \emph{copied} from the input model trace to the output model trace.

### 2020-07-13 paper.tex
finish commenting out
Typos

--- 2020-07-13 paper.tex
@@
 
-See Section~\ref{sec:radon-nikodym-special-case} for a derivation of the Radon-Nikodym derivative $d (\mu \circ f^{-1}) / d \mu$ for a special case, when the involution $f$ can be factored into (i) an involution on a countable partition of $Z$ and (ii) a family of continuously differentiable bijection on Euclidean spaces.
+See Section~\ref{sec:radon-nikodym-special-case} for a derivation of the Radon-Nikodym derivative $d (\mu \circ f^{-1}) / d \mu$ for a special case, when the involution $f$ can be factored into (i) an involution on a countable partition of $Z$ and (ii) a family of continuously differentiable bijections on Euclidean spaces.
 
@@
 
-The probabilistic programming language used in this paper is based on the dynamic modeling language in the Gen probabilistic programming system~\citep{gen-pldi}.
+The probabilistic programming language used in this paper is 
+% TODO: why "based on"?
+based on 
+the dynamic modeling language in the Gen probabilistic programming system~\citep{gen-pldi}.
 In this language, user tags each random choice with an explicit address in the code, as part of a \emph{random-choice expression} of the following form:
@@
 To handle programs that sample continuous probability distributions the notion of a probability mass function on traces needs to be generalized to a probability density function on traces with respect to a certain reference measure.
-The probabilistic program still defines a probability distibution (i.e. measure) on traces, but viavia  a density instead of a probability mass function.
+The probabilistic program still defines a probability distribution (i.e. measure) on traces, but via  a density instead of a probability mass function.
 
@@
 The random choice at address $\mathtt{:}\mathtt{degree}$ is discrete with $V_a = \integers_{\ge 0}$ and the other random choices are continuous.
-Address $\mathtt{:}\mathtt{:noise}$ has $V_a = \reals_{> 0}$ and $(\mathtt{"c"}, i)$ and $(\mathtt{"y"}, i)$ have $V_a = \reals$, and all of these use the Lebesgue measure. % TODO double-check this
+The address $\mathtt{:}\mathtt{noise}$ has $V_a = \reals_{> 0}$ and $(\mathtt{"c"}, i)$ and $(\mathtt{"y"}, i)$ have $V_a = \reals$, and all of these use the Lebesgue measure. % TODO double-check this
 The density with respect to the reference measure $\mu$ is:
@@
      @write(tr3, :k, k-1, :discrete)
-     new_weights = delete_weight(weights, i)
+     new_weights = delete_weight(weights, idx)
    end

### 2020-07-13 paper.tex

--- 2020-07-13 paper.tex
@@
 %\EndProcedure
-
-\Procedure{$P$.score}{$x$}
-    %\State \Return $p(x)$ % TODO or product of densities? % TODO state notation is non-standard
-    % if we don't use this product then what is the point -- why don't we just use p(x)
-    \State \Return $\prod_{a \in A(x)} p_{\sigma(a)}(x[a])$ (?) TODO % TODO
-\EndProcedure
-
-\Procedure{$Q$.score}{$y$; $x$}
-    %\State \Return $q_x(y)$ % TODO or product of densities?
-    \State \Return $\prod_{a \in A(y)} q_{x,\sigma(a)}(y[a])$ (?) TODO % TODO
-\EndProcedure
-
-\Procedure{$F$.jacdetabs}{$x$,$y$}
-    \For{$a$ in $\textproc{continuous-addresses}(x)$}
-        \State $J1[a,:] \gets \nabla_{x[a]} \left( F(x, y)_{\mathrm{continuous}} \right)$ (?) TODO % TODO
-    \EndFor
-    \For{$a$ in $\textproc{continuous-addresses}(y)$}
-        \State $J2[a,:] \gets \nabla_{y[a]} \left( F(x, y)_{\mathrm{continuous}} \right)$ (?) TODO % TODO
-    \EndFor
-    \State $J \gets \left[\begin{array}{c}J1\\J2\end{array}\right]$
-    \State \Return $|\det(J)|$
-\EndProcedure
-\end{algorithmic}
-\caption{Trace Involutive MCMC}
-\label{alg:trimcmc}
-\end{algorithm}
+%
+%\Procedure{$P$.score}{$x$}
+    %%\State \Return $p(x)$ % TODO or product of densities? % TODO state notation is non-standard
+    %% if we don't use this product then what is the point -- why don't we just use p(x)
+    %\State \Return $\prod_{a \in A(x)} p_{\sigma(a)}(x[a])$ (?) TODO % TODO
+%\EndProcedure
+%
+%\Procedure{$Q$.score}{$y$; $x$}
+    %%\State \Return $q_x(y)$ % TODO or product of densities?
+    %\State \Return $\prod_{a \in A(y)} q_{x,\sigma(a)}(y[a])$ (?) TODO % TODO
+%\EndProcedure
+%
+%\Procedure{$F$.jacdetabs}{$x$,$y$}
+    %\For{$a$ in $\textproc{continuous-addresses}(x)$}
+        %\State $J1[a,:] \gets \nabla_{x[a]} \left( F(x, y)_{\mathrm{continuous}} \right)$ (?) TODO % TODO
+    %\EndFor
+    %\For{$a$ in $\textproc{continuous-addresses}(y)$}
+        %\State $J2[a,:] \gets \nabla_{y[a]} \left( F(x, y)_{\mathrm{continuous}} \right)$ (?) TODO % TODO
+    %\EndFor
+    %\State $J \gets \left[\begin{array}{c}J1\\J2\end{array}\right]$
+    %\State \Return $|\det(J)|$
+%\EndProcedure
+%\end{algorithmic}
+%\caption{Trace Involutive MCMC}
+%\label{alg:trimcmc}
+%\end{algorithm}
 

### 2020-07-14 experiments/pytorch/involutions.py
add prototype pytorch implementation for jacobian calc

@@ -1,0 +1,64 @@
+import torch
+
+continuous = "continuous"
+discrete = "discrete"
+
+def polar_to_cartesian(r, theta):
+    x = torch.cos(r) * theta
+    y = torch.sin(r) * theta
+    return (x, y)
+
+def cartesian_to_polar(x, y):
+    theta = torch.atan2(y, x)
+    y = torch.sqrt(x * x + y * y)
+    return (theta, y)
+
+def write_cont(trace, addr, value):
+    trace[addr] = (value, continuous)
+
+def write_disc(trace, addr, value):
+    trace[addr] = (value, discrete)
+
+def f(input_trace):
+    output_trace = {}
+    if input_trace["polar"]:
+        (x, y) = polar_to_cartesian(input_trace["r"], input_trace["theta"])
+        write_cont(output_trace, "x", x)
+        write_cont(output_trace, "y", y)
+    else:
+        (r, theta) = cartesian_to_polar(input_trace["x"], input_trace["y"])
+        write_cont(output_trace, "r", r)
+        write_cont(output_trace, "theta", theta)
+    write_disc(output_trace, "polar", not input_trace["polar"])
+    return output_trace
+
+def involution_with_grad(f, input_trace):
+    stripped_input_trace = {}
+    for var in input_trace.keys():
+        if input_trace[var][1] == continuous:
+            stripped_input_trace[var] = torch.tensor(input_trace[var][0], requires_grad=True)
+        else:
+            stripped_input_trace[var] = input_trace[var][0]
+    output_trace = f(stripped_input_trace)
+    grads = []
+    for (output_var, (output_val, output_label)) in output_trace.items():
+        if output_label == continuous:
+            output_val.backward(retain_graph=True)
+            grad = []
+            for (input_var, input_val) in stripped_input_trace.items():
+                if input_trace[input_var][1] == continuous:
+                    grad.append(input_val.grad.clone())
+            for (input_var, input_val) in stripped_input_trace.items():
+                if input_trace[input_var][1] == continuous:
+                    input_val.grad.zero_()
+            grads.append(grad)
+    (_, logabsdet) = torch.tensor(grads).slogdet()
+    print(logabsdet)
+    
+input_trace = {
+        "polar" : (True, discrete),
+        "r": (1.2, continuous),
+        "theta" : (0.12, continuous)
+}
+
+print(involution_with_grad(f, input_trace))

### 2020-07-14 experiments/pytorch/involutions.py
checkpoint on pytorch implementation

@@ -1,17 +1,11 @@
 import torch
 
+##########################
+# generic implementation #
+##########################
+
 continuous = "continuous"
 discrete = "discrete"
-
-def polar_to_cartesian(r, theta):
-    x = torch.cos(r) * theta
-    y = torch.sin(r) * theta
-    return (x, y)
-
-def cartesian_to_polar(x, y):
-    theta = torch.atan2(y, x)
-    y = torch.sqrt(x * x + y * y)
-    return (theta, y)
 
 def write_cont(trace, addr, value):
     trace[addr] = (value, continuous)
@@ -19,20 +13,7 @@ def write_cont(trace, addr, value):
 def write_disc(trace, addr, value):
     trace[addr] = (value, discrete)
 
-def f(input_trace):
-    output_trace = {}
-    if input_trace["polar"]:
-        (x, y) = polar_to_cartesian(input_trace["r"], input_trace["theta"])
-        write_cont(output_trace, "x", x)
-        write_cont(output_trace, "y", y)
-    else:
-        (r, theta) = cartesian_to_polar(input_trace["x"], input_trace["y"])
-        write_cont(output_trace, "r", r)
-        write_cont(output_trace, "theta", theta)
-    write_disc(output_trace, "polar", not input_trace["polar"])
-    return output_trace
-
-def involution_with_grad(f, input_trace):
+def involution_with_jacobian_det(f, input_trace):
     stripped_input_trace = {}
     for var in input_trace.keys():
         if input_trace[var][1] == continuous:
@@ -53,12 +34,79 @@ def involution_with_grad(f, input_trace):
                     input_val.grad.zero_()
             grads.append(grad)
     (_, logabsdet) = torch.tensor(grads).slogdet()
-    print(logabsdet)
-    
-input_trace = {
+    return (output_trace, logabsdet)
+
+def strip_type_labels_for_logpdf(input_trace):
+    output_trace = {}
+    for (var, (val, label)) in input_trace.items():
+        output_trace[var] = val
+    return output_trace
+
+def involution_mcmc_step(p, q, f, input_trace):
+    prev_log_score = p(strip_type_labels_for_logpdf(input_trace))
+    (output_trace, logabsdet) = involution_with_jacobian_det(f, input_trace)
+    new_log_score = p(strip_type_labels_for_logpdf(output_trace))
[…]

### 2020-07-14 experiments/pytorch/involutions.py
checkpoint on pytorch implementation

@@ -4,79 +4,125 @@ import torch
 # generic implementation #
 ##########################
 
-continuous = "continuous"
-discrete = "discrete"
+CONTINUOUS = "CONTINUOUS"
+DISCRETE = "DISCRETE"
 
-def write_cont(trace, addr, value):
-    trace[addr] = (value, continuous)
+InvolutionRunnerState = collections.namedtuple("InvolutionRunnerState",
+    [   "input_model_trace", "input_aux_trace",
+        "output_model_trace", "output_aux_trace",
+        "input_grad_tensors"
+    ])
 
-def write_disc(trace, addr, value):
-    trace[addr] = (value, discrete)
+def read_model_cont(state, addr):
+    val = torch.tensor(state.input_model_trace.[addr], requires_grad=True)
+    state.input_grad_tensors.append(val.grad)
+    return val
 
-def involution_with_jacobian_det(f, input_trace):
-    stripped_input_trace = {}
-    for var in input_trace.keys():
-        if input_trace[var][1] == continuous:
-            stripped_input_trace[var] = torch.tensor(input_trace[var][0], requires_grad=True)
-        else:
-            stripped_input_trace[var] = input_trace[var][0]
-    output_trace = f(stripped_input_trace)
+def read_model_disc(state, addr):
+    return state.input_model_trace[addr]
+
+def write_model_cont(state, addr, value):
+    state.output_model_trace.trace[addr] = (value, CONTINUOUS)
+
+def write_model_disc(state, addr, value):
+    state.output_model_trace[addr] = (value, DISCRETE)
+
+def read_aux_cont(state, addr):
+    val = torch.tensor(state.input_aux_trace.[addr], requires_grad=True)
+    state.input_grad_tensors.append(val.grad)
+    return val
+
+def read_aux_disc(state, addr):
+    return state.input_aux_trace[addr]
+
+def write_aux_cont(state, addr, value):
+    state.output_aux_trace.trace[addr] = (value, CONTINUOUS)
+
+def write_aux_disc(state, addr, value):
+    state.output_aux_trace[addr] = (value, DISCRETE)
+
+
+from itertools import chain
+
+def involution_with_jacobian_det(f, input_model_trace, input_auxiliary_trace):
+    state = InvolutionRunnerState(input_model_trace, input_auxiliary_trace, {}, {}, [])
+    f(state)
     grads = []
-    for (output_var, (output_val, output_label)) in output_trace.items():
[…]

### 2020-07-14 experiments/pytorch/involutions.py
simple minimal PPL and involution

@@ -1,4 +1,6 @@
 import torch
+from collections import namedtuple
+from itertools import chain
 
 ##########################
 # generic implementation #
@@ -7,75 +9,75 @@ import torch
 CONTINUOUS = "CONTINUOUS"
 DISCRETE = "DISCRETE"
 
-InvolutionRunnerState = collections.namedtuple("InvolutionRunnerState",
+InvolutionRunnerState = namedtuple("InvolutionRunnerState",
     [   "input_model_trace", "input_aux_trace",
         "output_model_trace", "output_aux_trace",
-        "input_grad_tensors"
+        "input_cont_tensors", "output_cont_tensors"
     ])
 
 def read_model_cont(state, addr):
-    val = torch.tensor(state.input_model_trace.[addr], requires_grad=True)
-    state.input_grad_tensors.append(val.grad)
+    val = torch.tensor(state.input_model_trace[addr], requires_grad=True)
+    state.input_cont_tensors.append(val)
     return val
 
 def read_model_disc(state, addr):
     return state.input_model_trace[addr]
 
-def write_model_cont(state, addr, value):
-    state.output_model_trace.trace[addr] = (value, CONTINUOUS)
+def write_model_cont(state, addr, val):
+    state.output_model_trace[addr] =val 
+    state.output_cont_tensors.append(val)
 
-def write_model_disc(state, addr, value):
-    state.output_model_trace[addr] = (value, DISCRETE)
+def write_model_disc(state, addr, val):
+    state.output_model_trace[addr] = val
 
 def read_aux_cont(state, addr):
-    val = torch.tensor(state.input_aux_trace.[addr], requires_grad=True)
-    state.input_grad_tensors.append(val.grad)
+    val = torch.tensor(state.input_aux_trace[addr], requires_grad=True)
+    state.input_cont_tensors.append(val)
     return val
 
 def read_aux_disc(state, addr):
     return state.input_aux_trace[addr]
 
 def write_aux_cont(state, addr, value):
-    state.output_aux_trace.trace[addr] = (value, CONTINUOUS)
+    state.output_aux_trace[addr] = val
+    state.output_cont_tensors.append(val)
 
-def write_aux_disc(state, addr, value):
-    state.output_aux_trace[addr] = (value, DISCRETE)
+def write_aux_disc(state, addr, val):
+    state.output_aux_trace[addr] = val
 
 
-from itertools import chain
 
[…]

### 2020-07-14 experiments/pytorch/involutions.py
pytorch impl working

@@ -2,12 +2,34 @@ import torch
 from collections import namedtuple
 from itertools import chain
 
-##########################
-# generic implementation #
-##########################
 
-CONTINUOUS = "CONTINUOUS"
-DISCRETE = "DISCRETE"
+##############################################
+# minimal probabilistic programming language #
+##############################################
+
+def sample(p, *args):
+    trace = {}
+    def trace_choice(dist, addr):
+        val = dist.sample()
+        trace[addr] = val
+        return val
+    p(trace_choice, *args)
+    return trace
+
+def logpdf(trace, p, *args):
+    lpdf = torch.tensor(0.0)
+    def trace_choice(dist, addr):
+        nonlocal lpdf
+        val = trace[addr]
+        lpdf += dist.log_prob(val)
+        return val
+    p(trace_choice, *args)
+    return lpdf
+
+
+###############################################
+# minimal differentiable programming language #
+###############################################
 
 InvolutionRunnerState = namedtuple("InvolutionRunnerState",
     [   "input_model_trace", "input_aux_trace",
@@ -45,8 +67,6 @@ def write_aux_cont(state, addr, value):
 def write_aux_disc(state, addr, val):
     state.output_aux_trace[addr] = val
 
-
-
 def involution_with_jacobian_det(f, input_model_trace, input_auxiliary_trace):
     state = InvolutionRunnerState(input_model_trace, input_auxiliary_trace, {}, {}, [], [])
     f(state)
@@ -55,30 +75,16 @@ def involution_with_jacobian_det(f, input_model_trace, input_auxiliary_trace):
         output_cont_tensor.backward(retain_graph=True)
         grad = []
         for input_cont_tensor in state.input_cont_tensors:
-            grad.append(input_cont_tensor.clone())
+            grad.append(input_cont_tensor.grad.clone())
             input_cont_tensor.grad.zero_()
         grads.append(grad)
     (_, logabsdet) = torch.tensor(grads).slogdet()
     return (state.output_model_trace, state.output_aux_trace, logabsdet)
 
-def sample(p, *args):
-    trace = {}
-    def trace_choice(dist, addr):
-        val = dist.sample()
-        trace[addr] = val
-        return val
-    p(trace_choice, *args)
-    return trace
 
-def logpdf(trace, p, *args):
-    lpdf = torch.tensor(0.0)
-    def trace_choice(dist, addr):
-        nonlocal lpdf
-        val = trace[addr]
[…]

### 2020-07-14 paper.tex
Checkpoint

@@
 
-%This is the best paper~\citep{cusumano2019gen}.
-
-\subsection{Second Level Heading}
-
-\subsubsection{Third Level Heading}
-
-\paragraph{Fourth Level Heading}
-
 \section{BACKGROUND AND RELATED WORK}
-
-%\subsection{Traces}
 
@@
 
+\section{INVOLUTIVE MCMC ON GENERAL STATE SPACES}
+
+Involutive MCMC is a general framework for constructing MCMC kernels that are stationary
+for a target probability measure $p$. Informally, the algorithm works as follows: starting at some state $x$, we first sample
+an auxiliary variable $y \sim q_x$ from a state-dependent auxiliary distribution. We then apply an \textit{involution} $f$ to the pair $(x, y)$ to obtain $(x', y')$. 
+Finally, we computes an acceptance probability $\alpha$ and either accept $x'$ as the new state, or reject it and repeat the previous state $x$.
+Different choices of $q$ and $f$ recover many algorithms from the literature~\citep{imcmc}.
+
+In this section, we derive the involutive MCMC algorithm for models $p$ and auxiliary kernels
+$q$ defined over general state spaces.
+By emphasizing general state spaces, we intend to clarify
+a potential point of confusion regarding the involutive MCMC algorithm: as presented by
+\citet{imcmc},
+the acceptance probability $\alpha$ depends on the Jacobian of the involution $f$, but
+it is not immediately clear how to define this Jacobian when $f$ may operate on samples
+from arbitrary measurable spaces, rather than on vectors in $\mathbb{R}^n$. Our 
+reformulation of the algorithm below is general enough to handle arbitrary model and auxiliary
+distributions, and precise enough to enable automation via probabilistic and differentiable
+programming: the rest of this paper uses it to develop a technique for deriving efficient implementations
+of involutive MCMC algorithms automatically, 
+given only declarative specifications of $p$, $q$, and $f$.
+
+\subsection{Distributions on dictionaries}
+
+Instead of vectors of real numbers, we consider in this section distributions over 
+\textit{finite dictionaries} that map (possibly random) keys to (possibly random) values. 
+Dictionaries include vectors as a special case (a vector $\mathbf{x} \in \mathbb{R}^n$ can
+be represented as a dictionary mapping the keys $1, \dots, n$ to the values $x_1, \dots, x_n$),
+but are in general more flexible: different keys can hold values of different types, and we
+can also consider distributions in which the set of keys is itself random, useful for model
+selection problems and structure uncertainty more generally.
+
+\textbf{Measure space for finite dictionaries.} We fix a countably infinite set $\mathcal{A}$ of possible keys, and for each key $a \in \mathcal{A}$,
+we fix a measure space $(V_a, \Sigma_a, \mu_a)$ determining the possible values $V_a$ that may be
+associated with $a$.
+Given a set of keys $A$, we can consider the set $V_A = \times_{a \in A} V_a$ of \textit{assignments} of values to each key.
+($V_A$ can also be made a measure space, using the standard product $\sigma$-algebra $\Sigma_A = \otimes_{a \in A} \Sigma_a$ and the product measure $\mu_A = \times_{a \in A} \mu_a$.)  Then the set of all finite dictionaries is $\mathcal{D} := \bigcup_{A \subset \mathcal{A}, |A| < \infty} \{(A, \mathbf{x}) \mid \mathbf{x} \in V_A\}$.
+That is, a dictionary specifies
+a finite set of keys $A \subset \mathcal{A}$ at which it has values, and an assignment $\mathbf{x}$ of values $x_a$ for each.
+We equip $\mathcal{D}$ with the $\sigma$-algebra $\Sigma_{\mathcal{D}} := \{\bigcup_{A \subset \mathcal{A}, |A| < \infty} \{(A, \mathbf{x}) \mid \mathbf{x} \in B_A\} \mid B_A \in \Sigma_A \, \text{for each finite } A \subset \mathcal{A}\}$ to obtain a measurable space
+of dictionaries.
+A reference measure $\mu_\mathcal{D}$ on this space can be constructed using the product measures $\mu_A$: we set $\mu(B) := \sum_{A \in \mathcal{A}, |A| < \infty} \mu_A(\{\mathbf{x} \mid (A, \mathbf{x}) \in B\})$.
+
+\textbf{Notation for dictionaries.} Given a dictionary $m = (A, \mathbf{x})$, we write $A_m$ for $A$ and $m[a]$ for the value $x_a$ associated with a key $a \in A$.
+
+\textbf{Densities on finite dictionaries.} Probability distributions on dictionaries can be defined by (i) choosing 
+domains $V_a$ for each address $a$ of interest, and (ii) choosing a probability density $p : \mathcal{D} \rightarrow [0, \infty)$ such that $\int p(m) \mu_\mathcal{D}(\text{d}m) = 1$.
+We now give several examples
+to build intuition.
+
+\textit{Gaussian mixture model with unknown number of components.} Consider a generative model of univariate data points $y_1, \dots, y_n$ from a Gaussian mixture with an unknown number of components $k$, each with unknown mean $m_i$ and standard deviation $s_i$.
+If we place a Gaussian prior on $m_i$, a Gamma prior on $s_i$, and a Poisson prior on $k$, the resulting density is
+%
+$$p(d) = TODO$$
+%
+when $A_d$ is of the form $\{\texttt{k}, \texttt{y}_\texttt{1}, \dots, \texttt{y}_\texttt{n}\} \cup \{\texttt{m}_i \mid 1 \leq i \leq d[\texttt{k}]\} \cup \{\texttt{s}_i \mid 1 \leq i \leq d[\texttt{k}]\}$, and 0 otherwise.
+In this case, we have $V_\texttt{k} = \mathbb{N}$ with the counting measure for $\mu_\texttt{k}$, and for all other addresses $a \in \mathcal{A}$, $V_a = \mathbb{R}$ with the Lebesgue measure for $\mu_a$.
+
+\textit{Polynomial regression.}
+
+\textbf{Conditional distributions via disintegration.} Consider a probability density $p$ on the space $\mathcal{D}$ of dictionaries. 
+%For a set of keys $A \subset \mathcal{A}$, let $p(A) := \int 1[A_m = A] p(m) \mu_\mathcal{D}(\text{d}m)$, the probability under $p$ of a dictionary with exactly the key set $A$. 
+We say a key $a \in \mathcal{A}$ \textit{almost always appears} if $\int \mathbf{1}[a \in A_m] p(m) \mu_\mathcal{D}(\text{d}m) = 1$.
+Suppose $B$ is a set of keys that almost always appear for $p$, and that $b = (B, \mathbf{b})$ is a dictionary with keys $B$.
+Furthermore, let $(A_m, \mathbf{m}) \oplus (A_n, \mathbf{n}) := (A_m \cup A_n, \mathbf{m} \times \mathbf{n})$ be the \textit{merge} of two dictionaries $m$ and $n$ defined on disjoint key sets $A_m$ and $A_n$.
+Then we can define the conditional density $p(d \mid b) := \mathbf{1}[A_d \cap B = \emptyset] \frac{p(d \oplus b)}{\int_{\{m \mid A_m \cap B = \emptyset\}} p(m \oplus b) \mu_\mathcal{D}(\text{d}m)}$ when the denominator is finite.
+If the reference measure $\mu_a$ associated with each key $a \in B$ is discrete (e.g., the counting measure), then this definition corresponds to the ordinary notion of conditioning on an event (namely, the event that a sample from $p$ agrees with the dictionary $b$ on all addresses in $B$).
+When this is not the case, it corresponds to a more general measure-theoretic notion called disintegration~\citep{disintegration1997}.
+
+\subsection{Involutive MCMC}
+
 \section{INVOLUTIVE MCMC} \label{sec:involutive-mcmc}
-This section gives a measure-theoretic formulation of involutive MCMC on general state spaces.
-This generality is necessary for applications of involutive MCMC to models with structure uncertainty and for describing how involutive MCMC can be automated using probabilistic programming.
+Involutive MCMC (Algorithm~\ref{alg:involutive-mcmc}) defines a transition kernel that is stationary for a probabilistic model $p$, based on an auxiliary probability kernel $q$ and a deterministic involution $f$.
+Different choices of $q$ and $f$ recover instances of classic algorithms from the literature (e.g., Reversible-Jump MCMC~\citep{green}, TODO: others).
+This section briefly reviews involutive MCMC and gives a general measure-theoretic formulation of the algorithm on arbitrary state-spaces.
+In the following sections, we will describe how the implementations of involutive MCMC algorithms can be automated using probabilistic and differentiable programming.
+
+\begin{algorithm}[h]
+\begin{algorithmic}
+\Procedure{involutive-mcmc-move}{$p$, $q$, $f$, $x$}
+    \State $y \sim q_x(\cdot)$ \Comment{Sample auxiliary variables}
+    \State $(x', y') \gets f(x, y)$ \Comment{Apply involution}
+    \State $\alpha \gets
+        \displaystyle \frac{p(x') q_{x'}(y')}{p(x) q_{x}(y)} \left( \frac{d (\mu \circ f^{-1})}{d \mu} (x, y)\right)$
+    % TODO check it
+    \State $r \sim \mathrm{Uniform}(0, 1)$
+    \State \algorithmicif \, $r \le \alpha$ \algorithmicthen \, \Return $x'$ \algorithmicelse \, \Return $x$ 
+\EndProcedure
+\end{algorithmic}
+\caption{Involutive MCMC}
+\label{alg:involutive-mcmc}
+\end{algorithm}
 
 Let $(X, \Sigma_P, \mu_P)$ and $(Y, \Sigma_Q, \mu_Q)$ denote two measure spaces with $\sigma$-finite $\mu_P$ and $\mu_Q$. 
-Let $p : X \to [0, \infty)$ be a probability density with respect to $\mu_P$ and for each $x \in X$ such that $p(x) > 0$ let $q_x : Y \to [0, \infty)$ be probability density with respect to $\mu_Q$.
-We call $p$ the \emph{model density} and we call $q_x$ \emph{auxiliary densities}.
-Let $\pi(x, y) := p(x) q_x(y)$ and $Z := \{(x, y) \in X \times Y : \pi(x, y) > 0 \}$.
-Suppose that $Z \in \Sigma_P \otimes \Sigma_Q$ ($Z$ is a $\mu_P \times \mu_Q$-measurable set).
-Then $\Sigma := \{A \in \Sigma_P \otimes \Sigma_Q : A \subseteq Z\}$ is a $\sigma$-algebra on $Z$ and $\mu : \Sigma \to [0, \infty)$ with $\mu(A) := (\mu_P \times \mu_Q)(A)$ is a $\sigma$-finite measure on measurable space $(Z, \Sigma)$.
+Involutive MCMC implements a transition kernel that is invariant for a \textit{model distribution} given by $p : X \to [0, \infty)$, a probability density over $X$ with respect to $\mu_P$.
+Each iteration, the algorithm first samples auxiliary variables $y \in Y$ from an auxiliary distribution $q_x$ based on the model's current state $x$: for each $x \in X$ such that $p(x) > 0$, $q_x : Y \to [0, \infty)$ is a probability density with respect to $\mu_Q$.
+% We call $p$ the \emph{model density} and we call $q_x$ \emph{auxiliary densities}.
+
+The resulting pair $(x, y)$ of the current model state and the newly sampled auxiliary state will be an element of the joint space $Z := \{(x, y) \in X \times Y \mid p(x) q_x(y) > 0 \}$.
+We can equip $Z$ with the $\sigma$-algebra
+$\Sigma := \{A \cap Z \mid A \in \Sigma_P \otimes \Sigma_Q\}$ (assuming $Z$ is a $\mu_P \times \mu_Q$-measurable set), and a reference measure $\mu(A) := (\mu_P \times \mu_Q)(A)$. 
+%Let $\pi(x, y) := p(x) q_x(y)$ and $Z := \{(x, y) \in X \times Y : \pi(x, y) > 0 \}$.
+%Suppose that $Z \in \Sigma_P \otimes \Sigma_Q$ ($Z$ is a $\mu_P \times \mu_Q$-measurable set).
+%Then $\Sigma := \{A \in \Sigma_P \otimes \Sigma_Q : A \subseteq Z\}$ is a $\sigma$-algebra on $Z$ and $\mu : \Sigma \to [0, \infty)$ with $\mu(A) := (\mu_P \times \mu_Q)(A)$ is a $\sigma$-finite measure on measurable space $(Z, \Sigma)$.
 
 Let $f : Z \to Z$ denote an involution ($f^{-1} = f$) such that the pushforward of $\mu$ under $f$, denoted $\mu \circ f^{-1}$, is absolutely continuous with respect to $\mu$, with Radon-Nikodym derivative $d (\mu \circ f^{-1}) / d\mu : Z \to [0, \infty)$.
-Let $y \sim q_x(\cdot)$ denote a random sample from the measure defined by $A \mapsto \int_A q_x(y) \mu_Q(dy)$ for $A \in \Sigma_Q$.
+Involutive MCMC runs $f$ on $(x, y)$ to obtain $(x', y')$, then computes an acceptance probability $\alpha$.
+With probability $\alpha$, the new state $x'$ is returned; otherwise, the previous state $x$ is repeated.
+%Let $y \sim q_x(\cdot)$ denote a random sample from the measure defined by $A \mapsto \int_A q_x(y) \mu_Q(dy)$ for $A \in \Sigma_Q$.
 
@@
 
-\begin{algorithm}[h]
-\begin{algorithmic}
-\Procedure{involutive-mcmc-move}{$p$, $q$, $f$, $x$}
-    \State $y \sim q_x(\cdot)$ \Comment{Sample auxiliary variables}
-    \State $(x', y') \gets f(x, y)$ \Comment{Apply involution}
-    \State $\alpha \gets
-        \displaystyle \frac{p(x') q_{x'}(y')}{p(x) q_{x}(y)} \left( \frac{d (\mu \circ f^{-1})}{d \mu} (x, y)\right)$
-    % TODO check it
-    \State $r \sim \mathrm{Uniform}(0, 1)$
-    \State \algorithmicif \, $r \le \alpha$ \algorithmicthen \, \Return $x'$ \algorithmicelse \, \Return $x$ 
-\EndProcedure
-\end{algorithmic}
-\caption{Involutive MCMC}
-\label{alg:involutive-mcmc}
-\end{algorithm}
-
-See Section~\ref{sec:radon-nikodym-special-case} for a derivation of the Radon-Nikodym derivative $d (\mu \circ f^{-1}) / d \mu$ for a special case, when the involution $f$ can be factored into (i) an involution on a countable partition of $Z$ and (ii) a family of continuously differentiable bijections on Euclidean spaces.
+Then we have the following result:
 
@@
 
-\section{PROBABILISTIC PROGRAMMING WITH TRACES} \label{sec:prob-prog-traces}
-
-Probabilistic programs are programs that define probability distributions.
-Probabilistic programs can be used to define standard joint probability distributions over vectors of random-variables, as well as more complex probability distributions with uncertainty over the structure in which the set of random variables itself random.
-Involutive MCMC can be applied for target probability distributions on non-standard reference measures, including models with structure uncertainty.
-For example, reversible jump MCMC, which is a special case of involutive MCMC, can be applied to switch between different model structures.
-This section describes a framework for probabilistic programming that is expressive enough to represent probability distributions with structure uncertainty.
-
-In our framework, a probabilistic program defines a probability distribution on finite dictionaries that map the identifer for a random choice (called its \emph{address}) to its value.
-Following standard terminology~\citep{?}, we call these dictionaries \emph{traces} because they are often sampled by running a probabilistic program, and `tracing' the execution the program by recording the value of eacn random choice encountered.
-The next section makes the concept of a trace more concrete using code in a probabilistic programming language.
+See Section~\ref{sec:radon-nikodym-special-case} for a derivation of the Radon-Nikodym derivative $d (\mu \circ f^{-1}) / d \mu$ for a special case, when the involution $f$ can be factored into (i) an involution on a countable partition of $Z$ and (ii) a family of continuously differentiable bijections on Euclidean spaces.
+
+\section{AUTOMATING INVOLUTIVE MCMC} \label{sec:automating}
+
+Involutive MCMC is a general framework that can be used to develop diverse MCMC algorithms
+for models over arbitrary state spaces. We wish to \textit{automate} the implementation
+details for involutive MCMC algorithms, given only a specification of the model $p$, the
+auxiliary distribution $q$, and the involution $f$.
+To do so, we require a representation for the distributions $p$ and $q$ that is flexible
+enough to represent the full variety of models and auxiliary distributions of interest to
+practitioners. The representation must support density evaluation and sampling. It is also desirable that the representation be \textit{structured}: the more
+information available to us (e.g., about the decomposition of a distribution's state space into 
+individual univariate and multivariate random variables, or about conditional independence relationships in a model),
+the easier it will be for the implementation to exploit this structure automatically by using
+more efficient data structures and low-level manipulations.
+
+\textit{Probabilistic programs} are flexible and structured representations for arbitrary
+computable probability distributions. Unlike densities (but like Bayesian networks), probabilistic programs can be efficiently sampled and contain explicitly represented information
+about some conditional independence relationships in a model. 
+But unlike Bayesian networks, they do not assume a fixed number of random variables, state space dimension, or dependency structure.
+This section introduces one language for probabilistic programs, with which users can encode both their
+models $p$ and their auxiliary distributions $q_x$.
 
@@
 
+In our framework, a probabilistic program defines a probability distribution on finite maps (dictionaries) that map the names of random choices (their \emph{addresses}) to values.
+Following standard terminology~\citep{?}, we call these dictionaries \emph{traces} because they are often sampled by running a probabilistic program, and `tracing' the execution of the program by recording the value of each random choice encountered.
+We now make the concept of a trace more concrete using code in a probabilistic programming language.
+
 The probabilistic programming language used in this paper is 
-% TODO: why "based on"?
-based on 
 the dynamic modeling language in the Gen probabilistic programming system~\citep{gen-pldi}.
-In this language, user tags each random choice with an explicit address in the code, as part of a \emph{random-choice expression} of the following form:
+The language is embedded in the general-purpose Julia programming language~\citep{julia}.
+Julia functions can be turned into Gen probabilistic programs by tagging
+each random choice with an explicit address in the code, 
+as part of a \emph{random-choice expression} of the following form:
 \noindent
@@
 That is, the random choice expression evaluates to the sampled value.
-For example, the expression below evaluates to a an even non-negative integer:
+For example, the expression below samples the random variable \texttt{"a"} and multiplies it by 2,
+yielding an even non-negative integer:
 \noindent
@@
 \end{center}
-Probabilistic programs are typically constructed by combining multiple random choice expressions with regular Julia code (Gen is embedded in the Julia programming language~\citep{julia}).
+A probabilistic program will typically combine multiple random choice expressions. For example,
+the following code samples a random choice at address \texttt{"a"}, assigns it to the Julia variable
+\texttt{a}, and then samples a Bernoulli random variable at address \texttt{"b"} with a parameter
+that depends on \texttt{a}:
 \noindent
@@
 One possible trace samples the value 1 at $\texttt{"a"}$ and the Boolean value $\false$ at \texttt{"b"}, and we denote this trace by $\{\mathtt{"a"} \mapsto 1, \mathtt{"b"} \mapsto \false\}$.
-Note that the variable to which the first choice was assigned happens to be the same as the address.
-This is not required, and often the random choice has no variable name in the program, as in the second random choice with address \texttt{"b"}.
-But to reduce possibe unecessary typing, the following optional syntactic sugar will simultaneously assign the value to the program variable $\mathtt{a}$ and record it at trace address $\mathtt{a}$.\footnote{
+Note that the Julia variable to which the first choice is assigned (\texttt{a}) happens to be the same as the address (\texttt{"a"}).
+This is not required, and often the random choice has no corresponding Julia variable in the program, as in the second random choice with address \texttt{"b"}.
+But to reduce unnecessary typing, the following optional syntactic sugar will simultaneously assign the value to the program variable $\mathtt{a}$ and record it at trace address $\mathtt{"a"}$.\footnote{
 More precisely, the address is the Julia symbol (interned string) \texttt{a}, constructed with Julia syntax \texttt{:a}.}
@@
 \end{center}
-Probabilistic programs can make use of familiar control flow constructs, like loops and branching.
-Stochastic control flow is also possible:
+
+What gives probabilistic programs much of their expressive power is that they can make use of familiar control flow constructs, like loops and if statements, and this control flow can both depend on random choices and influence which random choices are made downstream.
+Consider the following program implementing a geometric
+distribution as the result of iterated Bernoulli draws:
 \noindent
@@
 \end{center}
-Note that for the probabilistic program to be valid, we require that every address is used at most once within every possible execution.
+Note that each Bernoulli flip uses a different address (the number \texttt{i} increases at each iteration).
+For the probabilistic program to be valid, we require that every address is used at most once within every possible execution.
 
@@
 The random choice at address $\mathtt{:}\mathtt{degree}$ is discrete with $V_a = \integers_{\ge 0}$ and the other random choices are continuous.
-The address $\mathtt{:}\mathtt{noise}$ has $V_a = \reals_{> 0}$ and $(\mathtt{"c"}, i)$ and $(\mathtt{"y"}, i)$ have $V_a = \reals$, and all of these use the Lebesgue measure. % TODO double-check this
-The density with respect to the reference measure $\mu$ is:
+The address $\mathtt{:}\mathtt{noise}$ has $V_a = \reals_{> 0}$ and $(\mathtt{"c"}, i)$ and $(\mathtt{"y"}, i)$ have $V_a = \reals$, and at all of these continuous addresses, we use the Lebesgue measure for $\mu_a$.
+% TODO double-check this
+The density with respect to the reference measure $\mu$ is then:
 \begin{align*}
@@
 
-To define the model and auxiliary densitites used in involutive MCMC in Section~\ref{sec:involutive-mcmc} using probabilistic programs, we write one probabilistic program ($P$) to define the model's probability distribution, and another probabilistic program ($Q$) to define the set of auxiliary distributions.
+To define the model and auxiliary densities used in involutive MCMC in Section~\ref{sec:involutive-mcmc} using probabilistic programs, we write one probabilistic program ($P$) to define the model's probability distribution, and another probabilistic program ($Q$) to define the set of auxiliary distributions.
 The model reference measure $(X, \Sigma_P, \mu_P)$ and density that were required in Section~\ref{sec:involutive-mcmc} are set to the probabilistic program's reference measure and density $p$ on traces as defined above (so that $X := T$).

### 2020-07-14 paper.tex

@@
 
-% TODO what is a trace? and what is a p? do we need parents?
-%\begin{algorithm}[h]
-%\begin{algorithmic}
-%\Procedure{trimcmc-kernel}{$p$, $q$, $f$, $x$}
-    %\State $y \sim q_x(\cdot)$ \Comment{Sample auxiliary trace}
-    %\State $(x', y') \gets f(x, y)$ \Comment{Apply involution}
-    %\State $w_1 \gets \prod_{a \in A(x)} p_{A(x)}(x_a | x_{\mathrm{Pa}^{A(x)}_a})$
-    %\State $w_2 \gets \prod_{a \in A(y)} q_{x,A(y)}(y_a | x_{\mathrm{Pa}^{A(x)}_a})$
-    %\State $w_3 \gets \prod_{a \in A(x')} p_{A(x')}(x'_a | (x'_b)_{b \in \mathrm{Pa}^{A(x)}_a})$
-    %\State $w_4 \gets \prod_{a \in A(y')} q_{x',A(y')}(y'_b | (y'_b)_{b \in \mathrm{Pa}^{A(x)}_a})$
-    %\State $|J| \gets F.\textproc{jacdetabs}(x, y)$
-    %\State $\alpha \gets
-        %\displaystyle \frac{w_3 \cdot w_4}{w_1 \cdot w_2} \cdot w_5$
-    %\State $r \sim \mathrm{Uniform}(0, 1)$
-    %\State \algorithmicif \, $r \le \alpha$ \algorithmicthen \, \Return $x'$ \algorithmicelse \, \Return $x$ 
-%\EndProcedure
-%
-%\Procedure{$P$.score}{$x$}
-    %%\State \Return $p(x)$ % TODO or product of densities? % TODO state notation is non-standard
-    %% if we don't use this product then what is the point -- why don't we just use p(x)
-    %\State \Return $\prod_{a \in A(x)} p_{\sigma(a)}(x[a])$ (?) TODO % TODO
-%\EndProcedure
-%
-%\Procedure{$Q$.score}{$y$; $x$}
-    %%\State \Return $q_x(y)$ % TODO or product of densities?
-    %\State \Return $\prod_{a \in A(y)} q_{x,\sigma(a)}(y[a])$ (?) TODO % TODO
-%\EndProcedure
-%
-%\Procedure{$F$.jacdetabs}{$x$,$y$}
-    %\For{$a$ in $\textproc{continuous-addresses}(x)$}
-        %\State $J1[a,:] \gets \nabla_{x[a]} \left( F(x, y)_{\mathrm{continuous}} \right)$ (?) TODO % TODO
-    %\EndFor
-    %\For{$a$ in $\textproc{continuous-addresses}(y)$}
-        %\State $J2[a,:] \gets \nabla_{y[a]} \left( F(x, y)_{\mathrm{continuous}} \right)$ (?) TODO % TODO
-    %\EndFor
-    %\State $J \gets \left[\begin{array}{c}J1\\J2\end{array}\right]$
-    %\State \Return $|\det(J)|$
-%\EndProcedure
-%\end{algorithmic}
-%\caption{Trace Involutive MCMC}
-%\label{alg:trimcmc}
-%\end{algorithm}
+\begin{algorithm}[h]
+\begin{algorithmic}
+\Procedure{involutive-mcmc-move}{$p$, $q$, $f$, $x$}
+    \State $y \sim q_x(\cdot)$ \Comment{Sample auxiliary variables}
+    \State $(x', y') \gets f(x, y)$ \Comment{Apply involution}
+    \State $\alpha \gets
+        \displaystyle \frac{p(x') q_{x'}(y')}{p(x) q_{x}(y)} \left( \frac{d (\mu \circ f^{-1})}{d \mu} (x, y)\right)$
+    % TODO check it
+    \State $r \sim \mathrm{Uniform}(0, 1)$
+    \State \algorithmicif \, $r \le \alpha$ \algorithmicthen \, \Return $x'$ \algorithmicelse \, \Return $x$ 
+\EndProcedure
+\end{algorithmic}
+\caption{Involutive MCMC}
+\label{alg:involutive-mcmc}
+\end{algorithm}
 

### 2020-07-14 paper.tex
added some content to involutive mcmc on dictionaries

@@
 
-\section{INVOLUTIVE MCMC ON GENERAL STATE SPACES}
+\section{INVOLUTIVE MCMC ON GENERAL STATE SPACES} % on traces ??
 
@@
 
-\subsection{Distributions on dictionaries}
+\subsection{Distributions on Dictionaries}
 
@@
 
-\textbf{Notation for dictionaries.} Given a dictionary $m = (A, \mathbf{x})$, we write $A_m$ for $A$ and $m[a]$ for the value $x_a$ associated with a key $a \in A$.
-
-\textbf{Densities on finite dictionaries.} Probability distributions on dictionaries can be defined by (i) choosing 
-domains $V_a$ for each address $a$ of interest, and (ii) choosing a probability density $p : \mathcal{D} \rightarrow [0, \infty)$ such that $\int p(m) \mu_\mathcal{D}(\text{d}m) = 1$.
+\textbf{Notation for dictionaries.} Given a dictionary $x = (A, \mathbf{x})$, we write $A_m$ for $A$ and $m[a]$ for the value $x_a$ associated with a key $a \in A$.
+
+\textbf{Probability distributions on finite dictionaries.} 
+When all keys are discrete (have countable $V_a$ and counting measure for $\mu_{a}$), a probability distribution on dictionaries is defined by a $p$ is a probability mass function $p : \mathcal{D} \to [0, 1]$ that assigns a probability $p(m)$ to each dictionary $m \in \mathcal{P}$ such that $\sum_{x \in \mathcal{D}} p(m) = 1$.
+More generally a probability distribution in dictionaries is defined by a probability density $p : \mathcal{D} \rightarrow [0, \infty)$ such that $\int p(m) \mu_\mathcal{D}(\text{d}m) = 1$.
 We now give several examples
 to build intuition.
+
+% TODO keys vs addresses
+
+% TODO how to notate dictionaries -- 'm' or 'd' or 'x' or 't'?
 
@@
 \textit{Polynomial regression.}
+
+TODO
 
@@
 
-\section{INVOLUTIVE MCMC} \label{sec:involutive-mcmc}
-Involutive MCMC (Algorithm~\ref{alg:involutive-mcmc}) defines a transition kernel that is stationary for a probabilistic model $p$, based on an auxiliary probability kernel $q$ and a deterministic involution $f$.
-Different choices of $q$ and $f$ recover instances of classic algorithms from the literature (e.g., Reversible-Jump MCMC~\citep{green}, TODO: others).
-This section briefly reviews involutive MCMC and gives a general measure-theoretic formulation of the algorithm on arbitrary state-spaces.
-In the following sections, we will describe how the implementations of involutive MCMC algorithms can be automated using probabilistic and differentiable programming.
-
+Let $(X, \Sigma_P, \mu_P)$ and $(Y, \Sigma_Q, \mu_Q)$ denote two measure spaces with $\sigma$-finite $\mu_P$ and $\mu_Q$. 
+Involutive MCMC implements a transition kernel that is invariant for a \textit{model distribution} given by $p : X \to [0, \infty)$, a probability density over $X$ with respect to $\mu_P$.
+Each iteration, the algorithm first samples auxiliary variables $y \in Y$ from an auxiliary distribution $q_x$ based on the model's current state $x$: for each $x \in X$ such that $p(x) > 0$, $q_x : Y \to [0, \infty)$ is a probability density with respect to $\mu_Q$.
+% We call $p$ the \emph{model density} and we call $q_x$ \emph{auxiliary densities}.
+
+The resulting pair $(x, y)$ of the current model state and the newly sampled auxiliary state will be an element of the joint space $Z := \{(x, y) \in X \times Y \mid p(x) q_x(y) > 0 \}$.
+We can equip $Z$ with the $\sigma$-algebra
+$\Sigma := \{A \cap Z \mid A \in \Sigma_P \otimes \Sigma_Q\}$ (assuming $Z$ is a $\mu_P \times \mu_Q$-measurable set), and a reference measure $\mu(A) := (\mu_P \times \mu_Q)(A)$. 
+%Let $\pi(x, y) := p(x) q_x(y)$ and $Z := \{(x, y) \in X \times Y : \pi(x, y) > 0 \}$.
+%Suppose that $Z \in \Sigma_P \otimes \Sigma_Q$ ($Z$ is a $\mu_P \times \mu_Q$-measurable set).
+%Then $\Sigma := \{A \in \Sigma_P \otimes \Sigma_Q : A \subseteq Z\}$ is a $\sigma$-algebra on $Z$ and $\mu : \Sigma \to [0, \infty)$ with $\mu(A) := (\mu_P \times \mu_Q)(A)$ is a $\sigma$-finite measure on measurable space $(Z, \Sigma)$.
+
+Let $f : Z \to Z$ denote an involution ($f^{-1} = f$) such that the pushforward of $\mu$ under $f$, denoted $\mu \circ f^{-1}$, is absolutely continuous with respect to $\mu$, with Radon-Nikodym derivative $d (\mu \circ f^{-1}) / d\mu : Z \to [0, \infty)$.
+Involutive MCMC runs $f$ on $(x, y)$ to obtain $(x', y')$, then computes an acceptance probability $\alpha$.
+With probability $\alpha$, the new state $x'$ is returned; otherwise, the previous state $x$ is repeated.
+%Let $y \sim q_x(\cdot)$ denote a random sample from the measure defined by $A \mapsto \int_A q_x(y) \mu_Q(dy)$ for $A \in \Sigma_Q$.
+
+% \left( \frac{d (\mu \circ f^{-1})}{d \mu} (x, y)\right)
+
+Suppose the set of possible keys $\mathcal{A}$ can be partitioned $\mathcal{A} = \mathcal{I} \cup \mathcal{K}$ where is a set of \emph{discrete keys} $\mathcal{I}$ that use countable $V_a$ and counting measure $\mu_a$ for $a \in \mathcal{I}$ and $\mathcal{K}$ is a set of \emph{continuous keys} that use $V_a = \mathbb{R}^{d_a}$ for some $d_a$ and Lebesgue-measure $\mu_a$ for $a \in \mathcal{K}$.
+%And suppose that there exists an involution $f : Z \to Z$ such that (i) 
+Suppose there is a countable partition of $Z$ into $\{Z_e : e \in E\}$ such that across each $Z_e$ the set of keys $A$ and the values $\mathbf{x}_{\mathrm{d}}$ of discrete keys are both constant (so that each set $Z_e$ is isomorphic to a Euclidean space of assignments to the continuous keys).
+Suppose there is an involution $g : E \to E$ between elements of the partition (a function with $g = g^{-1}$), and a family of continuously differentiable bijections $h_{e} : Z_e \to Z_{g(e)}$ for each $e$ with $h_{e} = h_{g(e)}^{-1}$.
+Let $e(z) \in E$ denote the element of the partition for an alement $z \in Z$.
+Then, $f : Z \to Z$ given by $f(z) := h_{e(z)}(z)$ is an involution:
+\begin{equation}
+f(f(z)) = h_{e(f(z))}(h_{e(z)}(z)) = h_{g(e(z))}(h_{e(z)}(z)) = z
+\end{equation}
+Let $|J h_e|(z)$ denote the absolute value of the Jacobian (determinant) of $h_e$, evaluated at $z$.
+Then, involutive MCMC is:
 \begin{algorithm}[h]
@@
     \State $\alpha \gets
-        \displaystyle \frac{p(x') q_{x'}(y')}{p(x) q_{x}(y)} \left( \frac{d (\mu \circ f^{-1})}{d \mu} (x, y)\right)$
-    % TODO check it
+        \displaystyle \frac{p(x') q_{x'}(y')}{p(x) q_{x}(y)} \cdot |J h_e|(z)$
     \State $r \sim \mathrm{Uniform}(0, 1)$
@@
 
-Let $(X, \Sigma_P, \mu_P)$ and $(Y, \Sigma_Q, \mu_Q)$ denote two measure spaces with $\sigma$-finite $\mu_P$ and $\mu_Q$. 
-Involutive MCMC implements a transition kernel that is invariant for a \textit{model distribution} given by $p : X \to [0, \infty)$, a probability density over $X$ with respect to $\mu_P$.
-Each iteration, the algorithm first samples auxiliary variables $y \in Y$ from an auxiliary distribution $q_x$ based on the model's current state $x$: for each $x \in X$ such that $p(x) > 0$, $q_x : Y \to [0, \infty)$ is a probability density with respect to $\mu_Q$.
-% We call $p$ the \emph{model density} and we call $q_x$ \emph{auxiliary densities}.
-
-The resulting pair $(x, y)$ of the current model state and the newly sampled auxiliary state will be an element of the joint space $Z := \{(x, y) \in X \times Y \mid p(x) q_x(y) > 0 \}$.
-We can equip $Z$ with the $\sigma$-algebra
-$\Sigma := \{A \cap Z \mid A \in \Sigma_P \otimes \Sigma_Q\}$ (assuming $Z$ is a $\mu_P \times \mu_Q$-measurable set), and a reference measure $\mu(A) := (\mu_P \times \mu_Q)(A)$. 
-%Let $\pi(x, y) := p(x) q_x(y)$ and $Z := \{(x, y) \in X \times Y : \pi(x, y) > 0 \}$.
-%Suppose that $Z \in \Sigma_P \otimes \Sigma_Q$ ($Z$ is a $\mu_P \times \mu_Q$-measurable set).
-%Then $\Sigma := \{A \in \Sigma_P \otimes \Sigma_Q : A \subseteq Z\}$ is a $\sigma$-algebra on $Z$ and $\mu : \Sigma \to [0, \infty)$ with $\mu(A) := (\mu_P \times \mu_Q)(A)$ is a $\sigma$-finite measure on measurable space $(Z, \Sigma)$.
-
-Let $f : Z \to Z$ denote an involution ($f^{-1} = f$) such that the pushforward of $\mu$ under $f$, denoted $\mu \circ f^{-1}$, is absolutely continuous with respect to $\mu$, with Radon-Nikodym derivative $d (\mu \circ f^{-1}) / d\mu : Z \to [0, \infty)$.
-Involutive MCMC runs $f$ on $(x, y)$ to obtain $(x', y')$, then computes an acceptance probability $\alpha$.
-With probability $\alpha$, the new state $x'$ is returned; otherwise, the previous state $x$ is repeated.
-%Let $y \sim q_x(\cdot)$ denote a random sample from the measure defined by $A \mapsto \int_A q_x(y) \mu_Q(dy)$ for $A \in \Sigma_Q$.
-
-\begin{algorithm}[h]
-\begin{algorithmic}
-\Procedure{involutive-mcmc-move}{$p$, $q$, $f$, $x$}
-    \State $y \sim q_x(\cdot)$ \Comment{Sample auxiliary variables}
-    \State $(x', y') \gets f(x, y)$ \Comment{Apply involution}
-    \State $\alpha \gets
-        \displaystyle \frac{p(x') q_{x'}(y')}{p(x) q_{x}(y)} \left( \frac{d (\mu \circ f^{-1})}{d \mu} (x, y)\right)$
-    % TODO check it
-    \State $r \sim \mathrm{Uniform}(0, 1)$
-    \State \algorithmicif \, $r \le \alpha$ \algorithmicthen \, \Return $x'$ \algorithmicelse \, \Return $x$ 
-\EndProcedure
-\end{algorithmic}
-\caption{Involutive MCMC}
-\label{alg:involutive-mcmc}
-\end{algorithm}
+\section{AUTOMATING INVOLUTIVE MCMC} \label{sec:involutive-mcmc}
+
+Involutive MCMC (Algorithm~\ref{alg:involutive-mcmc}) defines a transition kernel that is stationary for a probabilistic model $p$, based on an auxiliary probability kernel $q$ and a deterministic involution $f$.
+Different choices of $q$ and $f$ recover instances of classic algorithms from the literature (e.g., Reversible-Jump MCMC~\citep{green}, TODO: others).
+This section briefly reviews involutive MCMC and gives a general measure-theoretic formulation of the algorithm on arbitrary state-spaces.
+In the following sections, we will describe how the implementations of involutive MCMC algorithms can be automated using probabilistic and differentiable programming.
 

### 2020-07-14 paper.tex
Checkpoint on involutive mcmc with dictionaries

@@
 
-Let $(X, \Sigma_P, \mu_P)$ and $(Y, \Sigma_Q, \mu_Q)$ denote two measure spaces with $\sigma$-finite $\mu_P$ and $\mu_Q$. 
+Let $(X, \Sigma_P, \mu_P)$ and $(Y, \Sigma_Q, \mu_Q)$ denote two general measure spaces with $\sigma$-finite $\mu_P$ and $\mu_Q$. 
 Involutive MCMC implements a transition kernel that is invariant for a \textit{model distribution} given by $p : X \to [0, \infty)$, a probability density over $X$ with respect to $\mu_P$.
-Each iteration, the algorithm first samples auxiliary variables $y \in Y$ from an auxiliary distribution $q_x$ based on the model's current state $x$: for each $x \in X$ such that $p(x) > 0$, $q_x : Y \to [0, \infty)$ is a probability density with respect to $\mu_Q$.
+Each iteration, the algorithm first samples auxiliary variables $y \in Y$ from an \emph{auxiliary distribution} $q_x$ based on the model's current state $x$: for each $x \in X$ such that $p(x) > 0$, $q_x : Y \to [0, \infty)$ is a probability density with respect to $\mu_Q$.
 % We call $p$ the \emph{model density} and we call $q_x$ \emph{auxiliary densities}.
@@
 Let $f : Z \to Z$ denote an involution ($f^{-1} = f$) such that the pushforward of $\mu$ under $f$, denoted $\mu \circ f^{-1}$, is absolutely continuous with respect to $\mu$, with Radon-Nikodym derivative $d (\mu \circ f^{-1}) / d\mu : Z \to [0, \infty)$.
-Involutive MCMC runs $f$ on $(x, y)$ to obtain $(x', y')$, then computes an acceptance probability $\alpha$.
+Involutive MCMC runs $f$ on $(x, y)$ to obtain $(x', y')$, then computes an acceptance probability:
+\begin{equation}
+\alpha(x, y) = \frac{p(x') q_{x'}(y')}{p(x) q_{x}(y)} \cdot \left( \frac{d (\mu \circ f^{-1})}{d \mu} (x, y)\right)
+\end{equation}
 With probability $\alpha$, the new state $x'$ is returned; otherwise, the previous state $x$ is repeated.
@@
 
-% \left( \frac{d (\mu \circ f^{-1})}{d \mu} (x, y)\right)
-
-Suppose the set of possible keys $\mathcal{A}$ can be partitioned $\mathcal{A} = \mathcal{I} \cup \mathcal{K}$ where is a set of \emph{discrete keys} $\mathcal{I}$ that use countable $V_a$ and counting measure $\mu_a$ for $a \in \mathcal{I}$ and $\mathcal{K}$ is a set of \emph{continuous keys} that use $V_a = \mathbb{R}^{d_a}$ for some $d_a$ and Lebesgue-measure $\mu_a$ for $a \in \mathcal{K}$.
+\begin{theorem}[Involutive MCMC is stationary]
+Involutive MCMC defines a probability kernel $k$ on $X$ that is stationary with respect to the model probability distribution.
+That is, $\int_X k_x(B) p(x) d\mu_X(dx) = \int_B p(x) d\mu_X(dx)$ for all $B \in \Sigma_X$.
+\end{theorem}
+\begin{proof}
+The proof is presented in stages in the appendix (see Section~\ref{sec:involution-detailed-balance}, Section~\ref{sec:involution-is-stationary}, and Section~\ref{sec:involutive-mcmc-is-stationary}).
+\end{proof}
+
+\subsection{Involutive MCMC with Dictionaries}
+
+The general measure-theoretic formulation of involutive MCMC above does not specify how to compute a Radon-Nikodym derivative, or the densities.
+We now derive an algorithm for involutive MCMC when the model and auxiliary distributions are defined on spaces of dictionaries.
+
+% TODO should we just segment the addresses into discrete and continuous addresses as part of the dictionary formalism above -- that would allow us to move this up.
+the added flexibility of full per-address measure is not used anyways..
+
+Suppose that the model distribution $p$ and auxiliary distributions $q_x$ are all probability distributions on dictionaries.
+Then, $X$ and $Y$ are both sets of dictionaries, and the joint space $Z$ is a set of pairs $(x, y)$ of dictionaries.
+Suppose that both $X$ and $Y$ use the set of keys $\mathcal{A}$, and that $\mathcal{A}$ can be partitioned $\mathcal{A} = \mathcal{I} \cup \mathcal{K}$ where is a set of \emph{discrete keys} $\mathcal{I}$ that use countable $V_a$ and counting measure $\mu_a$ for $a \in \mathcal{I}$ and $\mathcal{K}$ is a set of \emph{continuous keys} that use $V_a = \mathbb{R}^{d_a}$ for some $d_a$ and Lebesgue-measure $\mu_a$ for $a \in \mathcal{K}$.
 %And suppose that there exists an involution $f : Z \to Z$ such that (i) 
-Suppose there is a countable partition of $Z$ into $\{Z_e : e \in E\}$ such that across each $Z_e$ the set of keys $A$ and the values $\mathbf{x}_{\mathrm{d}}$ of discrete keys are both constant (so that each set $Z_e$ is isomorphic to a Euclidean space of assignments to the continuous keys).
+Suppose there is a countable partition of $Z$ into $\{Z_e : e \in E\}$ such that across each $Z_e$ the sets of keys $A_x$ and $A_y$ and the values $\mathbf{x}_{\mathrm{d}}$ and $\mathbf{y}_{\mathrm{d}}$ of discrete keys are both constant (so that each set $Z_e$ is isomorphic to a Euclidean space of assignments to the continuous keys in the two dictionaries).
 Suppose there is an involution $g : E \to E$ between elements of the partition (a function with $g = g^{-1}$), and a family of continuously differentiable bijections $h_{e} : Z_e \to Z_{g(e)}$ for each $e$ with $h_{e} = h_{g(e)}^{-1}$.
@@
 Let $|J h_e|(z)$ denote the absolute value of the Jacobian (determinant) of $h_e$, evaluated at $z$.
-Then, involutive MCMC is:
+Algorithm~\ref{alg:involutive-mcmc} defines involutive MCMC in this more structured setting, in which the Radon-Nikodym is replaced with a Jacobian (see Section~\ref{sec:radon-nikodym-special-case} for a derivation).
 \begin{algorithm}[h]
 \begin{algorithmic}
-\Procedure{involutive-mcmc-move}{$p$, $q$, $f$, $x$}
-    \State $y \sim q_x(\cdot)$ \Comment{Sample auxiliary variables}
+\Procedure{involutive-mcmc-kernel}{$p$, $q$, $f$, $x$}
+    \State $y \sim q_x(\cdot)$ \Comment{Sample auxiliary dictionary}
     \State $(x', y') \gets f(x, y)$ \Comment{Apply involution}
@@
 \end{algorithmic}
-\caption{Involutive MCMC}
+\caption{Involutive MCMC for Dictionaries}
 \label{alg:involutive-mcmc}
@@
 
-\section{AUTOMATING INVOLUTIVE MCMC} \label{sec:involutive-mcmc}
-
-Involutive MCMC (Algorithm~\ref{alg:involutive-mcmc}) defines a transition kernel that is stationary for a probabilistic model $p$, based on an auxiliary probability kernel $q$ and a deterministic involution $f$.
+%\section{AUTOMATING INVOLUTIVE MCMC} \label{sec:involutive-mcmc}
+
+%Involutive MCMC (Algorithm~\ref{alg:involutive-mcmc}) defines a transition kernel that is stationary for a probabilistic model $p$, based on an auxiliary probability kernel $q$ and a deterministic involution $f$.
 Different choices of $q$ and $f$ recover instances of classic algorithms from the literature (e.g., Reversible-Jump MCMC~\citep{green}, TODO: others).
@@
 
-Then we have the following result:
-
-\begin{theorem}[Involutive MCMC is stationary]
-Algorithm~\ref{alg:involutive-mcmc} defines a probability kernel $k$ on $X$ that is stationary with respect to the model probability distribution.
-That is, $\int_X k_x(B) p(x) d\mu_X(dx) = \int_B p(x) d\mu_X(dx)$ for all $B \in \Sigma_X$.
-\end{theorem}
-\begin{proof}
-The proof is presented in stages in the appendix (see Section~\ref{sec:involution-detailed-balance}, Section~\ref{sec:involution-is-stationary}, and Section~\ref{sec:involutive-mcmc-is-stationary}).
-\end{proof}
-
-See Section~\ref{sec:radon-nikodym-special-case} for a derivation of the Radon-Nikodym derivative $d (\mu \circ f^{-1}) / d \mu$ for a special case, when the involution $f$ can be factored into (i) an involution on a countable partition of $Z$ and (ii) a family of continuously differentiable bijections on Euclidean spaces.
+%Then we have the following result:
 

### 2020-07-14 paper.tex
simplify measure formalism for dictionaries

@@
 
-\textbf{Measure space for finite dictionaries.} We fix a countably infinite set $\mathcal{A}$ of possible keys, and for each key $a \in \mathcal{A}$,
-we fix a measure space $(V_a, \Sigma_a, \mu_a)$ determining the possible values $V_a$ that may be
-associated with $a$.
-Given a set of keys $A$, we can consider the set $V_A = \times_{a \in A} V_a$ of \textit{assignments} of values to each key.
-($V_A$ can also be made a measure space, using the standard product $\sigma$-algebra $\Sigma_A = \otimes_{a \in A} \Sigma_a$ and the product measure $\mu_A = \times_{a \in A} \mu_a$.)  Then the set of all finite dictionaries is $\mathcal{D} := \bigcup_{A \subset \mathcal{A}, |A| < \infty} \{(A, \mathbf{x}) \mid \mathbf{x} \in V_A\}$.
-That is, a dictionary specifies
+\textbf{The space of finite dictionaries.}
+We fix a countably infinite set $\mathcal{A}$ of possible keys, such that each key $a$ is either \emph{discrete} ($a \in \mathcal{I}$) or \emph{continuous} ($a \in \mathcal{J}$), where $\mathcal{A} = \mathcal{I} \cup \mathcal{J}$.\footnote{
+It is straighforward to assign a general meaure space to each address, but this is not necessary for our purposes.}
+Let $V_a$ denote the set of possible values for key $a$, where $V_a$ is a countable set for each discrete key, and where $V_a = \mathbb{R}^{d_a}$ for each continuous key for some $d_a$.
+Given a set of keys $A$, let $V_A = \times_{a \in A} V_a$ denote the set of \textit{assignments} of values to each key.
+Then the set of all finite dictionaries is $\mathcal{D} := \bigcup_{A \subset \mathcal{A}, |A| < \infty} \{(A, \mathbf{x}) \mid \mathbf{x} \in V_A\}$. That is, a dictionary specifies 
 a finite set of keys $A \subset \mathcal{A}$ at which it has values, and an assignment $\mathbf{x}$ of values $x_a$ for each.
+
+\textbf{A measure space of finite dictionaries.}
+We associate a measure $\mu_a$ on $V_a$ for each key---the counting measure for each discrete key and the Lebesgue-measure on $\mathbf{R}^{d_a}$ for each continuous key.
+For each finite set of keys $A$, we make $V_A$ a measure space using the standard product $\sigma$-algebra $\Sigma_A = \otimes_{a \in A} \Sigma_a$ and the product measure $\mu_A = \times_{a \in A} \mu_a$.
 We equip $\mathcal{D}$ with the $\sigma$-algebra $\Sigma_{\mathcal{D}} := \{\bigcup_{A \subset \mathcal{A}, |A| < \infty} \{(A, \mathbf{x}) \mid \mathbf{x} \in B_A\} \mid B_A \in \Sigma_A \, \text{for each finite } A \subset \mathcal{A}\}$ to obtain a measurable space
@@
 \textbf{Notation for dictionaries.} Given a dictionary $x = (A, \mathbf{x})$, we write $A_m$ for $A$ and $m[a]$ for the value $x_a$ associated with a key $a \in A$.
+We also denote specific dictionaries using notation $\{k_1 \mapsto v_1, k_2 \mapsto v_2, \ldots\}$.
+For example, the dictionary $(A, \mathbf{x})$ with $A = \{1, \mathtt{"foo"}\}$ and $x_1 = 0.123$ and $x_{\mathtt{"foo"}} = 5$ is denoted 
+$\{1 \mapsto 0.123, \mathtt{"foo"} \mapsto 5\}$.
 
 \textbf{Probability distributions on finite dictionaries.} 
-When all keys are discrete (have countable $V_a$ and counting measure for $\mu_{a}$), a probability distribution on dictionaries is defined by a $p$ is a probability mass function $p : \mathcal{D} \to [0, 1]$ that assigns a probability $p(m)$ to each dictionary $m \in \mathcal{P}$ such that $\sum_{x \in \mathcal{D}} p(m) = 1$.
+When all keys are discrete, a probability distribution on dictionaries is defined by a $p$ is a probability mass function $p : \mathcal{D} \to [0, 1]$ that assigns a probability $p(m)$ to each dictionary $m \in \mathcal{P}$ such that $\sum_{x \in \mathcal{D}} p(m) = 1$.
 More generally a probability distribution in dictionaries is defined by a probability density $p : \mathcal{D} \rightarrow [0, \infty)$ such that $\int p(m) \mu_\mathcal{D}(\text{d}m) = 1$.
@@
 Suppose that the model distribution $p$ and auxiliary distributions $q_x$ are all probability distributions on dictionaries.
-Then, $X$ and $Y$ are both sets of dictionaries, and the joint space $Z$ is a set of pairs $(x, y)$ of dictionaries.
-Suppose that both $X$ and $Y$ use the set of keys $\mathcal{A}$, and that $\mathcal{A}$ can be partitioned $\mathcal{A} = \mathcal{I} \cup \mathcal{K}$ where is a set of \emph{discrete keys} $\mathcal{I}$ that use countable $V_a$ and counting measure $\mu_a$ for $a \in \mathcal{I}$ and $\mathcal{K}$ is a set of \emph{continuous keys} that use $V_a = \mathbb{R}^{d_a}$ for some $d_a$ and Lebesgue-measure $\mu_a$ for $a \in \mathcal{K}$.
+Then, $X$ and $Y$ are both sets of dictionaries, and the joint space $Z$ is a set of pairs $(x, y)$ of dictionaries with keys $\mathcal{A}$, so that $Z = X \times Y \subseteq \mathcal{D} \times \mathcal{D}$.
 %And suppose that there exists an involution $f : Z \to Z$ such that (i) 
-Suppose there is a countable partition of $Z$ into $\{Z_e : e \in E\}$ such that across each $Z_e$ the sets of keys $A_x$ and $A_y$ and the values $\mathbf{x}_{\mathrm{d}}$ and $\mathbf{y}_{\mathrm{d}}$ of discrete keys are both constant (so that each set $Z_e$ is isomorphic to a Euclidean space of assignments to the continuous keys in the two dictionaries).
+Suppose there is a countable partition of $Z$ into $\{Z_e : e \in E\}$ such that across each $Z_e$ the sets of keys $A_x$ and $A_y$ and the values $x_a$ and $y_a$ for all discrete keys $a$ are both constant (so that each set $Z_e$ is isomorphic to a Euclidean space of assignments to the continuous keys in the two dictionaries).
 Suppose there is an involution $g : E \to E$ between elements of the partition (a function with $g = g^{-1}$), and a family of continuously differentiable bijections $h_{e} : Z_e \to Z_{g(e)}$ for each $e$ with $h_{e} = h_{g(e)}^{-1}$.

### 2020-07-14 paper.tex
Rearranging

@@
 Algorithm~\ref{alg:involutive-mcmc} defines involutive MCMC in this more structured setting, in which the Radon-Nikodym is replaced with a Jacobian (see Section~\ref{sec:radon-nikodym-special-case} for a derivation).
+
 \begin{algorithm}[h]
@@
 
-%Then we have the following result:
-
 \section{AUTOMATING INVOLUTIVE MCMC} \label{sec:automating}
@@
 
+\subsection{Trace-Based Probabilistic Programming}
+
 \textit{Probabilistic programs} are flexible and structured representations for arbitrary
@@
 
-\subsection{A Probabilistic Programming Language}
+\subsubsection{A Probabilistic Programming Language}
 
 In our framework, a probabilistic program defines a probability distribution on finite maps (dictionaries) that map the names of random choices (their \emph{addresses}) to values.
+% TODO trace terminology
 Following standard terminology~\citep{?}, we call these dictionaries \emph{traces} because they are often sampled by running a probabilistic program, and `tracing' the execution of the program by recording the value of each random choice encountered.
@@
 
-\subsection{Handling Continuous Distributions}
-To handle programs that sample continuous probability distributions the notion of a probability mass function on traces needs to be generalized to a probability density function on traces with respect to a certain reference measure.
-The probabilistic program still defines a probability distribution (i.e. measure) on traces, but via  a density instead of a probability mass function.
-
-Let $\mathcal{A}$ denote a countable set of all possible addresses.
-For each address $a \in \mathcal{A}$ we assign a reference measure space $(V_a, \Sigma_a, \mu_a)$ where $\mu_a$ is $\sigma$-finite ($V$ stands for `values').
-Let $T := \cup_{A \subseteq \mathcal{A} : |A| < \infty} \{(A, \mathbf{x}) : \mathbf{x} \in \times_{a \in A} V_a\}\}$ denote the set of all finite traces (i.e.
+%\subsubsection{Handling Continuous Distributions}
+%To handle programs that sample continuous probability distributions the notion of a probability mass function on traces needs to be generalized to a probability density function on traces with respect to a certain reference measure.
+%The probabilistic program still defines a probability distribution (i.e. measure) on traces, but via  a density instead of a probability mass function.
+%
+%Let $\mathcal{A}$ denote a countable set of all possible addresses.
+%For each address $a \in \mathcal{A}$ we assign a reference measure space $(V_a, \Sigma_a, \mu_a)$ where $\mu_a$ is $\sigma$-finite ($V$ stands for `values').
+%Let $T := \cup_{A \subseteq \mathcal{A} : |A| < \infty} \{(A, \mathbf{x}) : \mathbf{x} \in \times_{a \in A} V_a\}\}$ denote the set of all finite traces (i.e.
 dictionaries) from addresses in $\mathcal{A}$ to their values.
-For each finite $A \subseteq \mathcal{A}$, let $\Sigma_A := \otimes_{a \in A} \Sigma_a$ (the product $\sigma$-algebra over $\Sigma_a$ for $a \in A$), and let $\mu_A := \times_{a \in A} \mu_a$ (the product measure).
-Let $\Sigma \subseteq \mathcal{P}(T)$ be the $\sigma$-algebra formed by sets of the form $B := \cup_{A \subseteq \mathcal{A}, |A| < \infty} \{(A, \mathbf{x}) : \mathbf{x} \in B_A\}$ for some $B_A \in \Sigma_A$ for each $A$.
-Let $\mu : \Sigma \to [0, \infty)$ be a measure on measurable space $(T, \Sigma)$ defined by $\mu(B) := \sum_{A \subseteq \mathcal{A}, |A| < \infty} \mu_A(B_A)$.
-
-A probabilistic program $P$ denotes a probability density $p : T \to [0, \infty)$ with respect to a reference measure $\mu$ on traces.
-Describing the denotational semantics of a particular probabilistic programming language (the function mapping programs $P$ to density functions $p$) is out of scope for this paper, and we refer to the reader to~\citet{popltraces} for an example of denotational semantics of a probabilistic programming language that uses explicit addresses for random-choices.
-But, informally, the density of a trace in our formalism (and existing formalisms) is the product of densities of each individual random choice, as an example below illustrates.
-
-The measure-theoretic formalism above is capable of expressing probabilistic programs that use continuous distributions by setting the reference measure space $(V_a, \Sigma_a, \mu_a)$ for some addresses $a$ to use the Lebesgue-measure.
+%For each finite $A \subseteq \mathcal{A}$, let $\Sigma_A := \otimes_{a \in A} \Sigma_a$ (the product $\sigma$-algebra over $\Sigma_a$ for $a \in A$), and let $\mu_A := \times_{a \in A} \mu_a$ (the product measure).
+%Let $\Sigma \subseteq \mathcal{P}(T)$ be the $\sigma$-algebra formed by sets of the form $B := \cup_{A \subseteq \mathcal{A}, |A| < \infty} \{(A, \mathbf{x}) : \mathbf{x} \in B_A\}$ for some $B_A \in \Sigma_A$ for each $A$.
+%Let $\mu : \Sigma \to [0, \infty)$ be a measure on measurable space $(T, \Sigma)$ defined by $\mu(B) := \sum_{A \subseteq \mathcal{A}, |A| < \infty} \mu_A(B_A)$.
+%
+
+\subsubsection{Automatically Computing Densities}
+% TODO explain the non-standard interpretation idea.
+A probabilistic program $P$ denotes a probability density $p : T \to [0, \infty)$ with respect to a reference measure $\mu$ on dictionaries.
+Defining the denotational semantics of a particular probabilistic programming language (the function mapping programs $P$ to density functions $p$) is out of scope for this paper, and we refer to the reader to~\citet{popltraces} for an example of denotational semantics of a probabilistic programming language similar to that of this paper.
+%But, informally, the density of a trace in our formalism (and existing formalisms) is the product of densities of each individual random choice, as an example below illustrates.
+But this section illustrates what form density functions for probabilistics programs take and how to they can be automatically evaluated by probabilistic programming systems.
 For example, consider the following probabilistic program, which encodes a generative model for data generated from a polynomial with unknown degree and unknown coefficients:
@@
 \end{align*}
-where $m(t, i) := \sum_{j=0}^{x[\mathtt{:}\mathtt{degree}]} \mathtt{xcoords}[i]^j \cdot x[(\mathtt{"c"}, j)]$.
-
-\subsection{Specifying the Densities for Involutive MCMC with Probabilistic Programs}
-
-To define the model and auxiliary densities used in involutive MCMC in Section~\ref{sec:involutive-mcmc} using probabilistic programs, we write one probabilistic program ($P$) to define the model's probability distribution, and another probabilistic program ($Q$) to define the set of auxiliary distributions.
-The model reference measure $(X, \Sigma_P, \mu_P)$ and density that were required in Section~\ref{sec:involutive-mcmc} are set to the probabilistic program's reference measure and density $p$ on traces as defined above (so that $X := T$).
-To define the auxiliary densities, we use a second probabilistic program $Q$ that is parametrized by (i.e. takes as input) traces of the model program $P$.
-The set of possible inputs to $Q$ is $\{x \in T : p(x) > 0\}$, and for each such $x$ the probabilistic program $Q$ defines a density $q_x : T \to [0, \infty)$ with respect to the reference measure on traces.\footnote{
-Note that we the same measure space $(T, \Sigma, \mu)$ on traces for both the model and auxiliary probabilistic program.
-If both programs sample some random choice at the same address $a$ and require different measure spaces $(V_a, \Sigma_a, \mu_a)$ for the address, this can be handled formally by making the addresses sampled by the two programs disjoint by prepending either the prefix `p.' or `q.' to addresses.
-This is only required for book-keeping in the formalism, and these prefixes do not need to be actually added to the addresses used in the programs.}
-
-The involutive MCMC framework of Section~\ref{sec:involutive-mcmc} requires that $Z := \{(x, y) \in X \times Y : \pi(x, y) > 0\}$ is $\mu_P \times \mu_Q$-measurable.
-We now give a sufficient condition for this to hold.
-Let $D \subseteq \mathcal{A}$ denote the subset of addresses that are discrete (i.e. where $V_a$ is a countable set and $\mu_a$ is the counting measure).
-For $\mathbf{x} \in \times_{a \in A} V_a$ let $\mathbf{x} = (\mathbf{d}, \mathbf{c})$ where $\mathbf{d} \in \times_{a \in A \cap D} V_a$ and $\mathbf{c} = \times_{a \in A \setminus D} V_a$, so that $\mathbf{d}$ is the discrete part of $\mathbf{x}$ and $\mathbf{c}$ is the non-discrete part.
- %TODO make this a proposition
-\begin{lemma}
-Suppose that $p$ and $q$ are such that $p(A, (\mathbf{d}, \mathbf{c})) > 0$ implies
-$p(A, (\mathbf{d}, \mathbf{c}')) > 0$ for all $\mathbf{c}' \in \prod_{a \in A \setminus D} V_a$,
-and that $q_x(A, (\mathbf{d}, \mathbf{c})) > 0$ where $x = (\tilde{A}, \tilde{\mathbf{d}}, \tilde{\mathbf{c}})$ implies that
-$q_{x'}(A, (\mathbf{d}, \mathbf{c}')) > 0$ for all $\mathbf{c}' \in \prod_{a \in A \setminus D} V_a$ and all $x' = (\tilde{A}, \tilde{\mathbf{d}}, \tilde{\mathbf{c}}')$ where $\tilde{\mathbf{c}}' \in \prod_{a \in \tilde{A} \setminus D} V_a$.
-Then, $Z := \{(x, y) \in X \times Y : \pi(x, y) > 0\}$ is $\mu \times \mu$-measurable where $\mu$ is the reference measure on traces.
-\end{lemma}
-\begin{proof}
-For $p$ and $q$ satisfying these conditions,
-$Z = \cup_{(A_1,A_2,\mathbf{d}_1,\mathbf{d}_2) \in E} \{((A_1, (\mathbf{d}_1, \mathbf{c}_1)), (A_2, (\mathbf{d}_2, \mathbf{c}_2))) : \mathbf{c}_1 \in \times_{a \in A_1 \setminus D} V_a, \mathbf{c}_2 \in \times_{a \in A_2 \setminus D} V_a\}$ for some countable set $E \subseteq \{(A_1, A_2, \mathbf{d}_1, \mathbf{d}_2) : A_1, A_2 \subseteq \mathcal{A}, |A_1| < \infty, |A_2| < \infty, \mathbf{d}_1 \in \times_{a \in A_1 \cap D} V_a, \mathbf{d}_2 \in \times_{a \in A_2 \cap D} V_a\}$ of address sets and discrete choice values for both programs.
-The measure of $Z$ is
-$(\mu \times \mu)(Z) = \sum_{(A_1,A_2,\mathbf{d}_1,\mathbf{d}_2) \in E} \prod_{a \in (A_1 \setminus D) \cup (A_2 \setminus D)} \mu_a(V_a)$.
-% TODO double check me.,
-\end{proof}
-
-Intuitively, this requirement means that for both the model probabilistic program $P$ and the auxiliary probabilistic program $Q$, the support of a random choice that is not discrete cannot depend on the value of another non-discrete random choice.
-Additionally, the support of non-discrete random choices in $Q$ cannot depend on the value of non-discrete random choices in the input $x$, which is a trace of $P$.
-This requirement defines a notion \emph{well-behavedness} for a probabilistic program ($P$) and an additional notion of well-behavedness for a pair of probabilistic programs that are sequenced one after the other ($P$ and $Q$).
-% TODO: prove that if Q satisfeies the criterion, then the composition P circ Q satisfies the single-program criterion..
-
-\subsubsection{Conditional Distributions on Traces}
+where $m(t, i) := \sum_{j=0}^{x[\mathtt{:}\mathtt{degree}]} \mathtt{xcoords}[i]^j \cdot x[(\mathtt{"c"}, j)]$, and where $p_{\mathrm{geometric}(r)}$, $p_{\mathrm{inversegamma}(\alpha, \beta)}$, and $p_{\mathrm{normal}(\mu, \sigma)}$ are density functions for geometric, inverse gamma, and normal distributions, respectively.
+
+\paragraph{Traces of Probabilistic Programs.}
+In general, the density $p$ of a probabilistic program with respect to the reference measure $\mu$ on dictionaries, evaluated at a dictionary $x = (A, \mathbf{x})$, can be computed as the product of individual density functions evaluated at each random choice $x[a]$.
+To compute the density of a dictionary $x$, many probabilistic programming systems including Gen employ traces.
+In the context of probabilistic programming, a \emph{trace} $t$ is a record of each of the random choices made during an execution of a probabilistic program, including (i) the addresses $A$ of the choices, (ii) their values $\mathbf{x}$, and (iii) the order in which the addresses was visited (as a list of addresses $t_1, \ldots, t_{|A|} \in A$), and (iv) the probability distributions $p^{(t)}_a$ from which each address $a \in A$ was sampled during the execution.
+That is, each trace $t$ contains a dictionary $x = (A, \mathbf{x})$ such that $p(x) > 0$, as well as the order information $t_1, \ldots, t_{|A|}$ and the distribution information $p^{(t)}_a$ for each $a$.
+If a trace $t$ is available for a dictionary $x$, then the probability density of the dictionary is the product of the densities of each random choice in the trace:
+\footnote{Probabilistic programming implementations use sums of log-densities instead of products for numerical reasons.}
+\begin{equation} \label{eq:density-product}
+p((A, \mathbf{x})) = \prod_{i=1}^{|A|} p^{(t)}_{a}(x[a_{t_i}] | x[a_{t_1}], \ldots, x[a_{t_{i-1}}])
+\end{equation}
+See~\citet{popltraces} for an example of a formal proof that the product of densities is the density of the dictionary $x$ with respect to the reference measure.
+Note that although in general, the probability density $p^{(t)}_a$ of address $a$ is a function of the value of the choice $x[a]$ as well the values of all previously sampled choices, in practice it is only a function of a subset of previous choices due to conditional independence structure in the model.
+
+Given a dictionary $x$ for which $p(x) > 0$, it is possible to obtain a trace $t$ containing that dictionary by executing the probabilistic program, but intercepting random choice expressions and modifying their behavior.
+For each intercepted random choice expression during the execution, instead of sampling the value of the random choice at address $a$, we (i) append the address $a$ to the list of addresses ($t_1, t_2, \ldots$), (ii) record the probability distribution $p^{(t)}_a$, (iii) look up the address $a$ in the dictionary (obtaining $v \gets x[a]$), and (iv) resume regular program execution, evaluating the random choice expression to the value $v$ obtained from the dictionary.
+We denote a trace $t$ obtained from a dictionary $x$ and a probabilistic program $P$:
+\begin{equation}
+t \gets \mathrm{trace}(P, x)
+\end{equation}
+
+% TODO add the algorithm box involving traces..
+
+\subsubsection{Conditioning on Observed Data}
+% TODO update this
 The formalism in this section so far defines the model distribution (with respect to which stationarity of the kernel is required) as the distribution on traces of an (unconditioned) probabilistic program $P$ that encodes a generative model.
@@
 
-\section{DIFFERENTIABLE PROGRAMMING WITH TRACES} \label{sec:differentiable-programming-traces}
+\subsubsection{Automatically Computing Density Ratios in Involutive MCMC}
+
+Provided the two probabilistic programs $P$ and $Q$ satisfy a simple technical requirement (described in Section~\ref{sec:technical-requirement}), they can be used to specify the model density $p$ and auxiliary densities $p_x$, respectively for use with involutive MCMC on dictionaries (Algorithm~\ref{alg:involutive-mcmc}).
+The acceptance probability of involutive MCMC includes the following ratio of densities:
+\begin{equation}
+\frac{p(x') q_{x'}(y')}{p(x) q_{x}(y)}
+\end{equation}
+To evaluate this density ratio, we can obtain four traces, one for each dictionary, and use those traces to compute the four probability densities, as shown in Algorithm~\ref{alg:trace-based-involutive-mcmc}.
+Section~\ref{sec:sparsity} describes a more efficent approach that exploits sparsity in the involution and resulting cancellation in the ratio for improved efficiency, but requires a more sophisticated probabilistic programming runtime system.
+
+\subsection{Differentiable Programming with Dictionaries} \label{sec:differentiable-programming-traces}
 
@@
 
-\subsection{A Differentiable Programming Language for Manipulating Traces}
+\subsubsection{A Differentiable Programming Language for Manipulating Dictionaries}
 Like the probabilistic programming languages described above, our differentiable language for involutions on traces is embedded based on Julia.
@@
 
-\subsection{Implicit Copies}
-
-If an address is present in the input model trace and is needed to form a complete output model trace but was not written to or copied to in by the function body, then it will be implicitly copied from the previous trace by the runtime implementation.
-This significantly reduces the amount of code that has to be written when only a small subset of the model trace is being modified during the application of the involution MCMC move (which is common in applications of reversible jump MCMC).
-As a simple example, if there are $N$ latent variables, and the move only updates 1 of them, then the involution needs to contain only $O(1)$ lines of code instead of $O(N)$ lines.
-
-TODO: show an example
-
-\subsection{Reasoning About Involutions}
+\subsubsection{Reasoning About Involutions}
 Recall the earlier code snippet that maps from polar to cartesian coordinates.
@@
 
-\subsection{The Radon-Nikodym Derivative}
-
-The differentiable programming language introduced in this section assumes that the reference measure of each random choice address is either the counting measure (corresponding to the $\mathtt{:}\mathtt{discrete}$ tag in \texttt{@read} and \texttt{@write} operations) or a Lebesgue measure ($\mathtt{:}\mathtt{continuous}$ tag).
-Consider a bijection $f : Z \to Z$ (recall that every involution is a bijection).
-If the space $Z$ can be partitioned into a countable number of sets $Z_1, Z_2, \ldots$, such that the each element of the partition is in one-to-one correspondence with another element ($f_1 : \{Z_1, Z_2, \ldots\} \to \{Z_1, Z_2, \ldots\}$ for some bijection $f_1$), and such that $f_{2,i} : Z_i \to Z_{f_1(i)}$ is a continuously differentiable bijection, and such that $f(z) = f_{2,i(z)}(z)$, then a Radon-Nikodym can be constructed from the Jacobian of the function $f_{2,i(z)}$ where $i(z) = Z_j$ for $z \in Z_j$.
-We denote the Jacobian (determinant) of a function $g$ by $| J g |$.
-See Section~\ref{sec:radon-nikodym-special-case} of the appendix for details.
-Applying this to the setting where $Z$ is a set of pairs of traces of probabilistic programs as formalized in the preivous section is straightforward when the partition of $Z$ is the countable disjoint union over all pairs $(A, \mathbf{d})$ where $A \subseteq \mathcal{A}$ are trace structures and $\mathbf{d}$ are assignments to discrete choices.
-
-% TODO do a PyTorch implementation??
-
-\section{AUTOMATIC ACCEPTANCE PROBABILITY CALCULATION} \label{sec:automating}
-
-The general form of the acceptance probability in Algorithm~\ref{alg:involutive-mcmc} is:
-\begin{equation}
-\min\left\{1, \frac{p(x') q_{x'}(y')}{p(x) q_{x}(y)} \left( \frac{d (\mu \circ f^{-1})}{d \mu} (x, y)\right) \right\}
-\end{equation}
-For the setting when $p$, $q$, and $f$ are defined using the probabilistic and differentiable programming languages of this section, the probability simplifies to:
-\begin{equation} \label{eq:accept-prob-jacobian}
-\min\left\{1, \frac{p(x') q_{x'}(y')}{p(x) q_{x}(y)} |J f_{2,i(x,y)} | \right\}
-\end{equation}
-Implementing the computation of this probability by hand is error-prone due to conceptual, arithmetic, and implementation errors; as well as time consuming and tedious.
-This section describes how the computation of this probability is automated by Gen using probabilistic and differentiable programming techniques.
-
-\subsection{Automatically Computing Probability Density Ratios}
-
-When the model density function $p$ and the auxiliary density functions $q_x$ in Algorithm~\ref{alg:involutive-mcmc} are specified by writing probabilistic programs in the language of Section~\ref{sec:prob-prog-traces}, Gen automatically computes the following ratio of densities, given the previous traces $x$ and $y$ and the proposed traces $x'$ and $y'$:
-\begin{equation} \label{eq:density-ratio}
-\frac{p(x') q_{x'}(y')}{p(x) q_x(y)}
-\end{equation}
-The density $p$ of a probabilistic program with respect to the reference measure $\mu$, evaluated at a trace $x = (A, \mathbf{x})$, is the product of individual densities for each random choice (and similarly for the auxiliary densities and the auxiliary probabilistic program):
-\begin{equation} \label{eq:density-product}
-p((A, \mathbf{x})) = \prod_{a \in A} p_{\sigma_a}(\mathbf{x}[a])
-\end{equation}
-where $\mathbf{x}[a]$ denotes the value of address $a$ in the trace, $p_{\sigma_a}$ is the probability density function for distribution from which random choice $a$ is sampled, given that the state of the program when sampling is $\sigma_a$.
-(See~\citet{popltraces} for a formal proof for a somewhat more restricted probabilistic programming language.)
-Therefore, the probabilistic programming language can compute the density at a trace $x$ by executing the program, but at each random choice $a$, reading the value of the choice $\mathbf{x}[a]$ instead of sampling it, computing the density at $\mathbf{x}[a]$ and accumulating it into the product above.\footnote{Probabilistic programming implementations use sums of log-densities instead of products for numerical reasons.}
-
-To evaluate this density ratio, we utilize Gen's API for probabilistic programs and traces, which automatically computes the density of a trace of a probabilistic program, whether that trace was sampled from the probabilistic program (as in the case of computing $q_x(y)$ for $y \sim q_{x}(\cdot)$) or when that trace was produced externally from the probabilistic program (as in the case of computing $q_{x'}(y')$, where $y'$ was produced from the involution $f$).
+%\subsubsection{The Radon-Nikodym Derivative}
+%
+%The differentiable programming language introduced in this section assumes that the reference measure of each random choice address is either the counting measure (corresponding to the $\mathtt{:}\mathtt{discrete}$ tag in \texttt{@read} and \texttt{@write} operations) or a Lebesgue measure ($\mathtt{:}\mathtt{continuous}$ tag).
+%Consider a bijection $f : Z \to Z$ (recall that every involution is a bijection).
+%If the space $Z$ can be partitioned into a countable number of sets $Z_1, Z_2, \ldots$, such that the each element of the partition is in one-to-one correspondence with another element ($f_1 : \{Z_1, Z_2, \ldots\} \to \{Z_1, Z_2, \ldots\}$ for some bijection $f_1$), and such that $f_{2,i} : Z_i \to Z_{f_1(i)}$ is a continuously differentiable bijection, and such that $f(z) = f_{2,i(z)}(z)$, then a Radon-Nikodym can be constructed from the Jacobian of the function $f_{2,i(z)}$ where $i(z) = Z_j$ for $z \in Z_j$.
+%We denote the Jacobian (determinant) of a function $g$ by $| J g |$.
+%See Section~\ref{sec:radon-nikodym-special-case} of the appendix for details.
+%Applying this to the setting where $Z$ is a set of pairs of traces of probabilistic programs as formalized in the preivous section is straightforward when the partition of $Z$ is the countable disjoint union over all pairs $(A, \mathbf{d})$ where $A \subseteq \mathcal{A}$ are trace structures and $\mathbf{d}$ are assignments to discrete choices.
+%
+
+\section{EXPLOITING SPARSITY FOR IMPROVED PERFORAMNCE} \label{sec:sparsity}
 
@@
 This allows for $O(1)$ computation of $p(x') / p(x)$ when the modeling language runtime system is able to detect and exploit the cancellation.
+
+\subsubsection{Implicit Copies}
+
+If an address is present in the input model trace and is needed to form a complete output model trace but was not written to or copied to in by the function body, then it will be implicitly copied from the previous trace by the runtime implementation.
+This significantly reduces the amount of code that has to be written when only a small subset of the model trace is being modified during the application of the involution MCMC move (which is common in applications of reversible jump MCMC).
+As a simple example, if there are $N$ latent variables, and the move only updates 1 of them, then the involution needs to contain only $O(1)$ lines of code instead of $O(N)$ lines.
+
+TODO: show an example
 
@@
 
+\subsection{A requirement on Densities for Involutive MCMC with Probabilistic Programs} \label{sec:technical-requirement}
+TODO: consider cutting this or moving to appendix -- it is a complicated technical requirement on programs
+
+% TODO cut this??
+To define the model and auxiliary densities used in involutive MCMC in Section~\ref{sec:involutive-mcmc} using probabilistic programs, we write one probabilistic program ($P$) to define the model's probability distribution, and another probabilistic program ($Q$) to define the set of auxiliary distributions.
+The model reference measure $(X, \Sigma_P, \mu_P)$ and density that were required in Section~\ref{sec:involutive-mcmc} are set to the probabilistic program's reference measure and density $p$ on traces as defined above (so that $X := T$).
+To define the auxiliary densities, we use a second probabilistic program $Q$ that is parametrized by (i.e. takes as input) traces of the model program $P$.
+The set of possible inputs to $Q$ is $\{x \in T : p(x) > 0\}$, and for each such $x$ the probabilistic program $Q$ defines a density $q_x : T \to [0, \infty)$ with respect to the reference measure on traces.\footnote{
+Note that we the same measure space $(T, \Sigma, \mu)$ on traces for both the model and auxiliary probabilistic program.
+If both programs sample some random choice at the same address $a$ and require different measure spaces $(V_a, \Sigma_a, \mu_a)$ for the address, this can be handled formally by making the addresses sampled by the two programs disjoint by prepending either the prefix `p.' or `q.' to addresses.
+This is only required for book-keeping in the formalism, and these prefixes do not need to be actually added to the addresses used in the programs.}
+
+The involutive MCMC framework of Section~\ref{sec:involutive-mcmc} requires that $Z := \{(x, y) \in X \times Y : \pi(x, y) > 0\}$ is $\mu_P \times \mu_Q$-measurable.
+We now give a sufficient condition for this to hold.
+Let $D \subseteq \mathcal{A}$ denote the subset of addresses that are discrete (i.e. where $V_a$ is a countable set and $\mu_a$ is the counting measure).
+For $\mathbf{x} \in \times_{a \in A} V_a$ let $\mathbf{x} = (\mathbf{d}, \mathbf{c})$ where $\mathbf{d} \in \times_{a \in A \cap D} V_a$ and $\mathbf{c} = \times_{a \in A \setminus D} V_a$, so that $\mathbf{d}$ is the discrete part of $\mathbf{x}$ and $\mathbf{c}$ is the non-discrete part.
+ %TODO make this a proposition
+\begin{lemma}
+Suppose that $p$ and $q$ are such that $p(A, (\mathbf{d}, \mathbf{c})) > 0$ implies
+$p(A, (\mathbf{d}, \mathbf{c}')) > 0$ for all $\mathbf{c}' \in \prod_{a \in A \setminus D} V_a$,
+and that $q_x(A, (\mathbf{d}, \mathbf{c})) > 0$ where $x = (\tilde{A}, \tilde{\mathbf{d}}, \tilde{\mathbf{c}})$ implies that
+$q_{x'}(A, (\mathbf{d}, \mathbf{c}')) > 0$ for all $\mathbf{c}' \in \prod_{a \in A \setminus D} V_a$ and all $x' = (\tilde{A}, \tilde{\mathbf{d}}, \tilde{\mathbf{c}}')$ where $\tilde{\mathbf{c}}' \in \prod_{a \in \tilde{A} \setminus D} V_a$.
+Then, $Z := \{(x, y) \in X \times Y : \pi(x, y) > 0\}$ is $\mu \times \mu$-measurable where $\mu$ is the reference measure on traces.
+\end{lemma}
+\begin{proof}
+For $p$ and $q$ satisfying these conditions,
+$Z = \cup_{(A_1,A_2,\mathbf{d}_1,\mathbf{d}_2) \in E} \{((A_1, (\mathbf{d}_1, \mathbf{c}_1)), (A_2, (\mathbf{d}_2, \mathbf{c}_2))) : \mathbf{c}_1 \in \times_{a \in A_1 \setminus D} V_a, \mathbf{c}_2 \in \times_{a \in A_2 \setminus D} V_a\}$ for some countable set $E \subseteq \{(A_1, A_2, \mathbf{d}_1, \mathbf{d}_2) : A_1, A_2 \subseteq \mathcal{A}, |A_1| < \infty, |A_2| < \infty, \mathbf{d}_1 \in \times_{a \in A_1 \cap D} V_a, \mathbf{d}_2 \in \times_{a \in A_2 \cap D} V_a\}$ of address sets and discrete choice values for both programs.
+The measure of $Z$ is
+$(\mu \times \mu)(Z) = \sum_{(A_1,A_2,\mathbf{d}_1,\mathbf{d}_2) \in E} \prod_{a \in (A_1 \setminus D) \cup (A_2 \setminus D)} \mu_a(V_a)$.
+% TODO double check me.,
+\end{proof}
+
+Intuitively, this requirement means that for both the model probabilistic program $P$ and the auxiliary probabilistic program $Q$, the support of a random choice that is not discrete cannot depend on the value of another non-discrete random choice.
+Additionally, the support of non-discrete random choices in $Q$ cannot depend on the value of non-discrete random choices in the input $x$, which is a trace of $P$.
+This requirement defines a notion \emph{well-behavedness} for a probabilistic program ($P$) and an additional notion of well-behavedness for a pair of probabilistic programs that are sequenced one after the other ($P$ and $Q$).
+% TODO: prove that if Q satisfeies the criterion, then the composition P circ Q satisfies the single-program criterion..
+
 \end{document}

### 2020-07-16 paper.tex
start implementation alg box
some tweaks to Section 3 and changes to Algortihm boxes

@@
 \AtBeginDocument{\mathcode`"="8000 }
+
+% algorithmic
+\algrenewcomment[1]{\(\triangleright\) #1}
+\algnewcommand{\LineComment}[1]{\State \(\triangleright\) #1}
+%\algblock[Foo]{execute}{end}
+
+\algblockx[exec]{execute}{END}%
+    [1][Unknown]{Execute #1}%
+    [1][Unknown]{\vspace{-4mm}}
 
@@
 Let $(X, \Sigma_P, \mu_P)$ and $(Y, \Sigma_Q, \mu_Q)$ denote two general measure spaces with $\sigma$-finite $\mu_P$ and $\mu_Q$. 
-Involutive MCMC implements a transition kernel that is invariant for a \textit{model distribution} given by $p : X \to [0, \infty)$, a probability density over $X$ with respect to $\mu_P$.
+Involutive MCMC (Algorithm~\ref{alg:involutive-mcmc}) implements a transition kernel that is invariant for a \textit{model distribution} given by $p : X \to [0, \infty)$, a probability density over $X$ with respect to $\mu_P$.
 Each iteration, the algorithm first samples auxiliary variables $y \in Y$ from an \emph{auxiliary distribution} $q_x$ based on the model's current state $x$: for each $x \in X$ such that $p(x) > 0$, $q_x : Y \to [0, \infty)$ is a probability density with respect to $\mu_Q$.
@@
 Let $f : Z \to Z$ denote an involution ($f^{-1} = f$) such that the pushforward of $\mu$ under $f$, denoted $\mu \circ f^{-1}$, is absolutely continuous with respect to $\mu$, with Radon-Nikodym derivative $d (\mu \circ f^{-1}) / d\mu : Z \to [0, \infty)$.
-Involutive MCMC runs $f$ on $(x, y)$ to obtain $(x', y')$, then computes an acceptance probability:
-\begin{equation}
-\alpha(x, y) = \frac{p(x') q_{x'}(y')}{p(x) q_{x}(y)} \cdot \left( \frac{d (\mu \circ f^{-1})}{d \mu} (x, y)\right)
-\end{equation}
+Involutive MCMC runs $f$ on $(x, y)$ to obtain $(x', y')$, then computes an acceptance probability $\alpha$.
 With probability $\alpha$, the new state $x'$ is returned; otherwise, the previous state $x$ is repeated.
@@
 
-The general measure-theoretic formulation of involutive MCMC above does not specify how to compute a Radon-Nikodym derivative, or the densities.
-We now derive an algorithm for involutive MCMC when the model and auxiliary distributions are defined on spaces of dictionaries.
-
-% TODO should we just segment the addresses into discrete and continuous addresses as part of the dictionary formalism above -- that would allow us to move this up.
-the added flexibility of full per-address measure is not used anyways..
+The general measure-theoretic formulation of involutive MCMC in Algorithm~\ref{alg:involutive-mcmc} does not specify how to compute the Radon-Nikodym derivative that is required for the acceptance probability.
+When the algorithm is specialized for state spaces consisting of dictionaries, the acceptance probability simplifies and can be computed from standard Jacobian and probability density computations.
+Because dictionaries are flexible objects, the algorithm remains quite general.
 
@@
 Let $|J h_e|(z)$ denote the absolute value of the Jacobian (determinant) of $h_e$, evaluated at $z$.
-Algorithm~\ref{alg:involutive-mcmc} defines involutive MCMC in this more structured setting, in which the Radon-Nikodym is replaced with a Jacobian (see Section~\ref{sec:radon-nikodym-special-case} for a derivation).
+Then, the acceptance probability simplifies to:
+\begin{equation}
+\alpha(x, y) = \frac{p(x') q_{x'}(y')}{p(x) q_{x}(y)} \cdot |J h_e|(z)
+\end{equation}
+Section~\ref{sec:radon-nikodym-special-case} of the appendix for details.
 
@@
 \begin{algorithmic}
-\Procedure{involutive-mcmc-kernel}{$p$, $q$, $f$, $x$}
-    \State $y \sim q_x(\cdot)$ \Comment{Sample auxiliary dictionary}
+\Procedure{involutive-mcmc}{$p$, $q$, $f$, $x$}
+    \State $y \sim q_x(\cdot)$ \Comment{Sample auxiliary state}
     \State $(x', y') \gets f(x, y)$ \Comment{Apply involution}
     \State $\alpha \gets
-        \displaystyle \frac{p(x') q_{x'}(y')}{p(x) q_{x}(y)} \cdot |J h_e|(z)$
+        \displaystyle \frac{p(x') q_{x'}(y')}{p(x) q_{x}(y)} \cdot \left( \frac{d (\mu \circ f^{-1})}{d \mu} (x, y)\right)$
     \State $r \sim \mathrm{Uniform}(0, 1)$
@@
 \end{algorithmic}
-\caption{Involutive MCMC for Dictionaries}
+\caption{Involutive MCMC}
 \label{alg:involutive-mcmc}
@@
 \section{AUTOMATING INVOLUTIVE MCMC} \label{sec:automating}
+
+% TODO add the algorithm box involving traces..
+
+\begin{algorithm*}[h]
+\begin{algorithmic}
+\Procedure{auto-involutive-mcmc}{$\mathcal{P}$, $\mathcal{Q}$, $\mathcal{F}$, $x$}
+    \State $(y, \log w_2) \gets \textproc{trace-and-score}(\mathcal{Q}_x)$ \Comment{Sample $y \sim q_x(\cdot)$ and compute log-density $\log q_x(y)$}
+    \State $(x', y', \log D) \gets \textproc{run-involution}(\mathcal{F}, (x, y))$ \Comment{Compute $(x', y') = f(x, y)$, log Radon-Nikodym deriv. $\log D$}
+    \State $\log w_1 \gets \textproc{score}(\mathcal{P}, x)$ \Comment{Compute log-density $\log p(x)$}
+    \State $\log w_3 \gets \textproc{score}(\mathcal{P}, x')$ \Comment{Compute log-density $\log p(x')$}
+    \State $\log w_4 \gets \textproc{score}(\mathcal{Q}_{x'}, y')$ \Comment{Compute log-density $\log q_{x'}(y')$}
+    \State $\alpha \gets \min\{1, \exp(\log w_3 - \log w_1 + \log w_4 - \log w_2 + \log J)\}$ \Comment{Compute acceptance probability}
+    \State \textbf{with probability} \,$\alpha$ \,\Return $x'$ \algorithmicelse \, \Return $x$ 
+\EndProcedure\\
+
+\begin{minipage}[t]{0.5\linewidth}
+\Procedure{trace-and-score}{$\mathcal{P}$}
+   \State $\log w \gets 0$
+   \State $x \gets \{\}$
+   \execute[$\mathcal{P}$], but with $\mathtt{"a \sim d"} \equiv$ (
+        \State \hspace{5mm} 1. set $v \sim d$
+        \State \hspace{5mm} 2. set $x[a] \gets v$
+        \State \hspace{5mm} 3. set $\log w \gets \log w + \textproc{logpdf}(d, v)$
+        \State \hspace{5mm} 4. evaluate to $v$)
+    \END
+   \State \Return $(x, \log w)$
+\EndProcedure
+\end{minipage}%
+\begin{minipage}[t]{0.5\linewidth}
+\Procedure{score}{$\mathcal{P}$, $x$}
+   \State $\log w \gets 0$
+   \execute[$\mathcal{P}$], but with $\mathtt{"a \sim d"} \equiv$ (
+        \State \hspace{5mm} 1. set $v \gets x[a]$
+        \State \hspace{5mm} 2. set $\log w \gets \log w + \textproc{logpdf}(d, v)$
+        \State \hspace{5mm} 3. evaluate to $v$)
+    \END
+   \State \Return $\log w$
+\EndProcedure
+\end{minipage}\\
+
+\Procedure{run-involution}{$\mathcal{F}$, $(x, y)$}
+% TODO would be much simpler to describe if we didn't have to separate out x and y..
+    \LineComment{Run the involution, and keep track of continuous addresses}
+    \State $x' \gets \{\}$ \Comment{Initialize empty output dictionary}
+    \execute[$\mathcal{F}$], but with
+        \State $\mathtt{"write\_cont}(a, v)\mathtt{"} \equiv$ (1. set $x'[a] \gets v$ and 2. set $W \gets W \cup \{a\}$)
+        \State $\mathtt{"read\_cont}(a)\mathtt{"} \equiv$ (1. set $R \gets R \cup \{a\}$ and 2. evaluate to $x[a]$)
+        \State $\mathtt{"copy}(a_1, a_2)\mathtt{"} \equiv$ (1. set $x'[a_2] \gets x[a_1]$ and 2. set $C \gets C \cup \{a_2\}$)
+    \END
+    \LineComment{Use automatic differentiation to compute Jacobian, skipping copied addresses}
+    \For{$i$ in $1, \ldots, |W \setminus C|$}
+        \State $J[i,:] \gets \nabla_{x_{R \setminus C}} \left( h_{e(x, y)}[(W \setminus C)_i] \right)$ \Comment{Gradient of $x'[(W \setminus C)_i]$ w.r.t. non-copied continuous inputs}
+    \EndFor
+    \State \Return $(x', \log |\det(J)|)$
+\EndProcedure
+
+\end{algorithmic}
+\caption{Automated Involutive MCMC}
+\label{alg:auto-involutive-mcmc}
+\end{algorithm*}
 
@@
 
-% TODO add the algorithm box involving traces..
-
 \subsubsection{Conditioning on Observed Data}

### 2020-07-16 paper.tex
4.1.1 rewrite

--- 2020-07-14 paper.tex
@@
 We fix a countably infinite set $\mathcal{A}$ of possible keys, such that each key $a$ is either \emph{discrete} ($a \in \mathcal{I}$) or \emph{continuous} ($a \in \mathcal{J}$), where $\mathcal{A} = \mathcal{I} \cup \mathcal{J}$.\footnote{
-It is straighforward to assign a general meaure space to each address, but this is not necessary for our purposes.}
+It is straightforward to assign a general measure space to each address, but this is not necessary for our purposes.}
 Let $V_a$ denote the set of possible values for key $a$, where $V_a$ is a countable set for each discrete key, and where $V_a = \mathbb{R}^{d_a}$ for each continuous key for some $d_a$.
@@
 \begin{algorithmic}
-\Procedure{involutive-mcmc-kernel}{$p$, $q$, $f$, $x$}
+\Procedure{involutive-mcmc}{$p$, $q$, $f$, $x$}
     \State $y \sim q_x(\cdot)$ \Comment{Sample auxiliary dictionary}
@@
 
+% TODO add the algorithm box involving traces..
+
+\begin{algorithm}[h]
+\begin{algorithmic}
+\Procedure{simulate-and-score}{$P(\beta)$}
+   \State $\log w \gets 0$
+   \State $x \gets \{\}$
+   \State Execute $P$ on input $\beta$, replacing $"\mathtt{a \sim d}"$ with:
+    \State $\;\;\;\;$ 1. $\; v \sim d$
+    \State $\;\;\;\;$ 2. $\; x[a] \gets v$
+    \State $\;\;\;\;$ 3. $\; \log w \gets \log w + \textproc{logpdf}(d, x[a])$
+    \State $\;\;\;\;$ 4. $\;$ \Return $v$
+   \State \Return $(x, \log w)$
+\EndProcedure
+
+\Procedure{score}{$P(\beta), x$}
+   \State $\log w \gets 0$
+   \State Execute $Q$ on input $x$, replacing $"\mathtt{a \sim d}"$ with:
+    \State $\;\;\;\;$ 1. $\; v \sim x[a]$
+    \State $\;\;\;\;$ 3. $\; \log w \gets \log w + \textproc{logpdf}(d, x[a])$
+    \State $\;\;\;\;$ 4. $\;$ \Return $v$
+   \State \Return $\log w$
+\EndProcedure
+
+\Procedure{run-involution}{$F$, $(x, y)$}
+    TODO
+\EndProcedure
+\end{algorithmic}
+\caption{Probabilistic and Differentiable Program Operations}
+\label{alg:prob-prog}
+\end{algorithm}
+
+\begin{algorithm}[h]
+\begin{algorithmic}
+\Procedure{auto-involutive-mcmc}{$P$, $Q$, $F$, $x$}
+    \State $\log w_1 \gets \textproc{score}(P(), x)$
+    \State $(y, \log w_2) \gets \textproc{simulate-and-score}(Q(x))$
+    \State $(x', y', \log |J|) \gets \textproc{run-involution}(F, (x, y))$
+    \State $\log w_2 \gets \textproc{score}(Q(x), y)$
+    \State $\log w_3 \gets \textproc{score}(P(), x')$
+    \State $\log w_4 \gets \textproc{score}(Q(x'), y')$
+    \State $\log \alpha \gets \log w_3 - \log w_1 + \log w_4 - \log w_2 + \log |J|$
+    \State $r \sim \mathrm{Uniform}(0, 1)$
+    \State \algorithmicif \, $r \le \exp(\alpha)$ \algorithmicthen \, \Return $x'$ \algorithmicelse \, \Return $x$ 
+\EndProcedure
+\end{algorithmic}
+\caption{Automated Involutive MCMC}
+\label{alg:involutive-mcmc}
+\end{algorithm}
+
 Involutive MCMC is a general framework that can be used to develop diverse MCMC algorithms
@@
 
-\textit{Probabilistic programs} are flexible and structured representations for arbitrary
-computable probability distributions. Unlike densities (but like Bayesian networks), probabilistic programs can be efficiently sampled and contain explicitly represented information
+\textit{Probabilistic programs} are flexible and structured representations for  probability distributions.
+Unlike densities (but like Bayesian networks), probabilistic programs can be efficiently sampled and contain explicitly represented information
 about some conditional independence relationships in a model. 
 But unlike Bayesian networks, they do not assume a fixed number of random variables, state space dimension, or dependency structure.
-This section introduces one language for probabilistic programs, with which users can encode both their
-models $p$ and their auxiliary distributions $q_x$.
-
-\subsubsection{A Probabilistic Programming Language}
-
-In our framework, a probabilistic program defines a probability distribution on finite maps (dictionaries) that map the names of random choices (their \emph{addresses}) to values.
-% TODO trace terminology
-Following standard terminology~\citep{?}, we call these dictionaries \emph{traces} because they are often sampled by running a probabilistic program, and `tracing' the execution of the program by recording the value of each random choice encountered.
-We now make the concept of a trace more concrete using code in a probabilistic programming language.
-
-The probabilistic programming language used in this paper is 
-the dynamic modeling language in the Gen probabilistic programming system~\citep{gen-pldi}.
-The language is embedded in the general-purpose Julia programming language~\citep{julia}.
-Julia functions can be turned into Gen probabilistic programs by tagging
-each random choice with an explicit address in the code, 
-as part of a \emph{random-choice expression} of the following form:
-\noindent
-\begin{center}
-\begin{tabular}{c}
-{\begin{lstlisting}[basicstyle=\small\ttfamily]
-{<address>} ~ <distribution>
-\end{lstlisting}}
-\end{tabular}
-\end{center}
-The left-hand side of the expression (green) specifies the address of the choice, and the right-hand side specifies the probability distribution to be sampled from.
-For example, the following expression samples a non-negative integer from a geometric distribution and the value gets recorded in the trace at the address \texttt{"a"}:
-\noindent
-\begin{center}
-\begin{tabular}{c}
-{\begin{lstlisting}[basicstyle=\small\ttfamily]
-{"a"} ~ geometric(0.5)
-\end{lstlisting}}
-\end{tabular}
-\end{center}
-In addition to being recorded in the trace, the sampled value is returned to the running program.
-That is, the random choice expression evaluates to the sampled value.
-For example, the expression below samples the random variable \texttt{"a"} and multiplies it by 2,
-yielding an even non-negative integer:
-\noindent
-\begin{center}
-\begin{tabular}{c}
-{\begin{lstlisting}[basicstyle=\small\ttfamily]
-({"a"} ~ geometric(0.5)) * 2
-\end{lstlisting}}
-\end{tabular}
-\end{center}
-A probabilistic program will typically combine multiple random choice expressions. For example,
-the following code samples a random choice at address \texttt{"a"}, assigns it to the Julia variable
-\texttt{a}, and then samples a Bernoulli random variable at address \texttt{"b"} with a parameter
-that depends on \texttt{a}:
-\noindent
-\begin{center}
-\begin{tabular}{c}
-{\begin{lstlisting}[basicstyle=\small\ttfamily]
-a = ({"a"} ~ geometric(0.5))
-{"b"} ~ bernoulli(a / (a + 1))
-\end{lstlisting}}
-\end{tabular}
-\end{center}
-This probabilistic program has an unbounded (countably infinite) number of possible traces it could produce.
-One possible trace samples the value 1 at $\texttt{"a"}$ and the Boolean value $\false$ at \texttt{"b"}, and we denote this trace by $\{\mathtt{"a"} \mapsto 1, \mathtt{"b"} \mapsto \false\}$.
-Note that the Julia variable to which the first choice is assigned (\texttt{a}) happens to be the same as the address (\texttt{"a"}).
-This is not required, and often the random choice has no corresponding Julia variable in the program, as in the second random choice with address \texttt{"b"}.
-But to reduce unnecessary typing, the following optional syntactic sugar will simultaneously assign the value to the program variable $\mathtt{a}$ and record it at trace address $\mathtt{"a"}$.\footnote{
-More precisely, the address is the Julia symbol (interned string) \texttt{a}, constructed with Julia syntax \texttt{:a}.}
-\noindent
-\begin{center}
-\begin{tabular}{c}
-{\begin{lstlisting}[basicstyle=\small\ttfamily]
-a ~ geometric(0.5)
-\end{lstlisting}}
-\end{tabular}
-\end{center}
-
-What gives probabilistic programs much of their expressive power is that they can make use of familiar control flow constructs, like loops and if statements, and this control flow can both depend on random choices and influence which random choices are made downstream.
-Consider the following program implementing a geometric
-distribution as the result of iterated Bernoulli draws:
-\noindent
-\begin{center}
-\begin{tabular}{c}
-{\begin{lstlisting}[basicstyle=\small\ttfamily]
-i = 1
-while ({i} ~ bernoulli(0.5))
-    i = i + 1
-end
-\end{lstlisting}}
-\end{tabular}
-\end{center}
-Note that each Bernoulli flip uses a different address (the number \texttt{i} increases at each iteration).
-For the probabilistic program to be valid, we require that every address is used at most once within every possible execution.
-
-For probabilistic programs that sample from discrete probability distributions like those used above, the meaning of the program is a probability mass function $p$ that assigns a probability to every possible trace.
-For example, for the program with the \texttt{while} loop above, the distribution is:
-\[
-\begin{array}{c|c}
-    x & p(x)\\
-    \hline
-    \{1 \mapsto \false\} & 0.5\\
-    \{1 \mapsto \true, 2 \mapsto \false\} & 0.5^2\\
-    \{1 \mapsto \true, 2 \mapsto \true, 3 \mapsto \false\} & 0.5^3\\
-    .. & ..
-\end{array}
-\]
-We assume that probabilistic programs halt with probability $1$, meaning that the sum of probabilities over all (finite) traces is one.
-Note that traces are always finite, but there may be an infinite number of possible traces, as in the previous example.
-
-In order to construct more complex distributions, we wrap probabilistic programs in functions, called
-`generative functions' in Gen.
-These functions can then be invoked by other functions, including recursively, which allows programs encode probability distributions on more complex and combinatorial objects.
-When invoking such a function, we give an \emph{address namespace} under which every random choice made in that function should be recorded.
-This helps prevent addresses of choices defined in different functions from colliding.
-The syntax for calling a generative function and assigning the address namespace is the same as that of random choice expressions.
-For example:
-\begin{center}
-\begin{tabular}{c}
-\begin{lstlisting}[basicstyle=\small\ttfamily]
-@gen function g()
-  if ({"go"} ~ bernoulli(0.2))
-    n1 = ({"L"} ~ g())
-    n2 = ({"R"} ~ g())
-    return n1 + n2 + 1
-  else
-    return 1
-  end
-end
-\end{lstlisting}
-\end{tabular}
-\end{center}
-Because this is a recursive function, a the trace is populated with a hierarchy of addresses.
-For example, the first random choice in the first recursive call has address \texttt{"L"."go"}.
-When this function is invoked it encodes the following probability distribution on traces (quotation marks omitted from addresses to improve readability):
-\[
-\begin{array}{c|c}
-    x & p(x)\\
-    \hline
-    \left\{\mathtt{go} \mapsto \false\right\} & 0.8\\
-& \\
-    \left\{\begin{array}{c}
-        \mathtt{go} \mapsto \true, \mathtt{L}.\mathtt{go} \mapsto \false, \mathtt{R}.\mathtt{go} \mapsto \false
-    \end{array}\right\} & 0.2 \cdot 0.8^2\\
-& \\
-    \left\{\begin{array}{c}
-        \mathtt{go} \mapsto \true,
-        \mathtt{L}.\mathtt{go} \mapsto \true,
-        \mathtt{R}.\mathtt{go} \mapsto \false,\\
-        \mathtt{L}.\mathtt{L}.\mathtt{go} \mapsto \false,
-        \mathtt{L}.\mathtt{R}.\mathtt{go} \mapsto \false
-    \end{array}\right\}& 0.2^2 \cdot 0.8^3\\
-     .. & ..
-\end{array}
-\]
-
-%\subsubsection{Handling Continuous Distributions}
-%To handle programs that sample continuous probability distributions the notion of a probability mass function on traces needs to be generalized to a probability density function on traces with respect to a certain reference measure.
-%The probabilistic program still defines a probability distribution (i.e. measure) on traces, but via  a density instead of a probability mass function.
-%
-%Let $\mathcal{A}$ denote a countable set of all possible addresses.
-%For each address $a \in \mathcal{A}$ we assign a reference measure space $(V_a, \Sigma_a, \mu_a)$ where $\mu_a$ is $\sigma$-finite ($V$ stands for `values').
-%Let $T := \cup_{A \subseteq \mathcal{A} : |A| < \infty} \{(A, \mathbf{x}) : \mathbf{x} \in \times_{a \in A} V_a\}\}$ denote the set of all finite traces (i.e.
-dictionaries) from addresses in $\mathcal{A}$ to their values.
-%For each finite $A \subseteq \mathcal{A}$, let $\Sigma_A := \otimes_{a \in A} \Sigma_a$ (the product $\sigma$-algebra over $\Sigma_a$ for $a \in A$), and let $\mu_A := \times_{a \in A} \mu_a$ (the product measure).
-%Let $\Sigma \subseteq \mathcal{P}(T)$ be the $\sigma$-algebra formed by sets of the form $B := \cup_{A \subseteq \mathcal{A}, |A| < \infty} \{(A, \mathbf{x}) : \mathbf{x} \in B_A\}$ for some $B_A \in \Sigma_A$ for each $A$.
-%Let $\mu : \Sigma \to [0, \infty)$ be a measure on measurable space $(T, \Sigma)$ defined by $\mu(B) := \sum_{A \subseteq \mathcal{A}, |A| < \infty} \mu_A(B_A)$.
-%
-
-\subsubsection{Automatically Computing Densities}
-% TODO explain the non-standard interpretation idea.
-A probabilistic program $P$ denotes a probability density $p : T \to [0, \infty)$ with respect to a reference measure $\mu$ on dictionaries.
-Defining the denotational semantics of a particular probabilistic programming language (the function mapping programs $P$ to density functions $p$) is out of scope for this paper, and we refer to the reader to~\citet{popltraces} for an example of denotational semantics of a probabilistic programming language similar to that of this paper.
-%But, informally, the density of a trace in our formalism (and existing formalisms) is the product of densities of each individual random choice, as an example below illustrates.
-But this section illustrates what form density functions for probabilistics programs take and how to they can be automatically evaluated by probabilistic programming systems.
-For example, consider the following probabilistic program, which encodes a generative model for data generated from a polynomial with unknown degree and unknown coefficients:
-\noindent
-\begin{center}
-\begin{tabular}{c}
-{\begin{lstlisting}[basicstyle=\small\ttfamily]
-@gen function polynomials(x_coords)
-    degree ~ |geometric(0.5)|
-    noise ~ |inv_gamma(1, 1)|
-    coeffs = [ ({("c", i)} ~ |normal(0, 2)|)
-               for i in 0:degree ]
-    for i=1:length(x_coords)
-        x = x_coords[i]
-        mu = coeffs' * x.^(0:length(coeffs)-1)
-        {("y", i)} ~ |normal(mu, noise + 0.1)|
-    end
-end
-\end{lstlisting}}
-\end{tabular}
-\end{center}
-The addresses used by this generative function are:
-\[
-\mathcal{A} := \begin{array}{c}
-\{\mathtt{:}\mathtt{noise}, \mathtt{:}\mathtt{degree}\} \bigcup
-\{(\mathtt{"c"}, i) : i \in \integers_{\ge 0}\} \bigcup\\
-\{(\mathtt{"y"}, i) : i \in \integers_{> 0}\}
-\end{array}
-\]
-(Note that $\mathcal{A}$ could contain additional addresses that the program never samples.)
-The random choice at address $\mathtt{:}\mathtt{degree}$ is discrete with $V_a = \integers_{\ge 0}$ and the other random choices are continuous.
-The address $\mathtt{:}\mathtt{noise}$ has $V_a = \reals_{> 0}$ and $(\mathtt{"c"}, i)$ and $(\mathtt{"y"}, i)$ have $V_a = \reals$, and at all of these continuous addresses, we use the Lebesgue measure for $\mu_a$.
-% TODO double-check this
-The density with respect to the reference measure $\mu$ is then:
+
+At the most basic level, a probabilistic program is a program that makes random choices. 
+Any such program induces a probability distribution over its possible \textit{execution traces}, records of each random choice it makes.
+If each random choice is associated with a unique \textit{address} from a countable set
+of addresses $\mathcal{A}$, 
+then these traces can be viewed as finite dictionaries, mapping the address of each random choice to its value. 
+The distribution induced by a probabilistic program over
+its execution traces can thus be understood as a measure on the space $(\mathcal{D}, \Sigma_\mathcal{D}, \mu_\mathcal{D})$ introduced in the previous section.
+Furthermore, densities of trace distributions with respect to $\mu_\mathcal{D}$ are typically easy to compute.
+
+In this section, we introduce the Gen language~\citep{gen-pldi} for probabilistic programs, and present a technique for automating the implementation of involutive MCMC algorithms when the model $p$ and auxiliary distribution $q_x$ are both represented as
+Gen programs.
+
+\subsubsection{A Language for Probabilistic Programs with Traces}
+
+The Gen probabilistic programming language augments the syntax of Julia~\citep{julia}
+with a single new construct, the \texttt{\{a\} $\sim$ d} expression, for making a \textit{named} random choice.
+An execution trace of a Gen probabilistic program can be sampled by running the program according to Julia's usual semantics, and upon encountering an expression of the form \texttt{\{a\} $\sim$ d}, (i) evaluating the \textit{address expression} \texttt{a} to obtain an address $a \in \mathcal{A}$; (ii) evaluating the \textit{distribution expression} $d$ to obtain a distribution $d$ over the measurable space $V_a$; (iii) sampling a value $x_a \sim d$ and adding the mapping $\{a \mapsto x_a\}$ to a dictionary called the program trace; and (iv) returning the sampled value $x_a$ to the program, to continue execution.
+When execution terminates, the program trace has accumulated a mapping for each random choice encountered during the program's execution.
+
+As a concrete example, consider Figure~\ref{fig:mixture}a, which defines a Gen program for a Gaussian mixture model with an unknown number of components.
+The program \texttt{p} accepts as input an integer \texttt{n}, a number of data points.
+Each $\texttt{n}$ defines
+a distinct distribution over dictionaries.
+
+The first line of the program samples a number of mixture components from a Poisson prior using the address \texttt{:k}.
+(This line could also be written \texttt{k = \{:k\} $\sim$ poisson\_plus\_one(1)}: the \textit{address} is the symbol \texttt{:k}, and the result of the choice is assigned to a Julia variable called \texttt{k}.
+Because this is a common pattern, Gen provides the syntactic sugar \texttt{x $\sim$ d} as shorthand for \texttt{x = \{:x\} $\sim$ d}.) The program then samples \texttt{k} means and \texttt{k} variances from Gaussian and inverse Gamma priors, respectively.
+Each of these 2\texttt{k} random choices has its own address, determined by the address expression preceding it.
+For example, the mean for the fourth mixture component (if \texttt{k} $\geq 4$) has address \texttt{(:mu, 4)}.
+The mixture weights are then sampled at the address \texttt{:weights} from a Dirichlet distribution, and $\texttt{n}$ data points are sampled at addresses \texttt{(:x, 1)}, \dots, \texttt{(:x, n)}.
+The random choice at address $\mathtt{:}\mathtt{k}$ is discrete with $V_{\texttt{:k}} = \integers_{\ge 1}$, and the other random choices are continuous.
+Overall, the program defines a distribution over traces with the following density over traces $(A, \mathbf{x})$ with respect to $\mu_\mathcal{D}$:
 \begin{align*}
-p(x) =
+p(n)((A, \mathbf{x})) =
 \begin{array}{l}
-p_{\mathrm{geometric}(0.5)}(x[\mathtt{:}\mathtt{degree}]) \cdot\\
-p_{\mathrm{inversegamma}(1, 1)}(x[\mathtt{:}\mathtt{noise}]) \cdot\\
-\prod_{0=1}^{x[\mathtt{:}\mathtt{degree}]} p_{\mathrm{normal}(0,1)}(x[(\mathtt{"c"}, i)])\cdot\\
-\prod_{i=1}^n p_{\mathrm{normal}(m(t, i), 0.1 + x[\mathtt{noise}])}(x[(\mathtt{"y"}, i)])
+p_{\mathrm{poisson}(1)}(x_{\mathtt{:}\mathtt{k}}-1) \cdot\\
+\prod_{i=1}^{x_{\mathtt{:}\mathtt{k}}} p_{\mathrm{normal}(0, 10)}(x_{(\mathtt{:}\mathtt{mu}, i)}) \cdot\\
+\prod_{i=1}^{x_{\mathtt{:}\mathtt{k}}} p_{\mathrm{inversegamma}(1, 10)}(x_{(\mathtt{:}\mathtt{var}, i)}) \cdot\\
+p_{\mathrm{dirichlet}(2.0,\dots,2.0)}(x_{\mathtt{:}\mathtt{weights}})\cdot\\
+\prod_{i=1}^n \sum_{j=1}^{x_{\mathtt{:}\mathtt{k}}} x_{\mathtt{:}\mathtt{weights}}[j] \cdot\\
+\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\; p_{\mathrm{normal}(x_{\mathtt{(:mu, j)}}, x_{(\mathtt{:}\mathtt{var}, j)})}(x_{(\mathtt{:}\mathtt{x}, i)})
 \end{array}
 \end{align*}
-where $m(t, i) := \sum_{j=0}^{x[\mathtt{:}\mathtt{degree}]} \mathtt{xcoords}[i]^j \cdot x[(\mathtt{"c"}, j)]$, and where $p_{\mathrm{geometric}(r)}$, $p_{\mathrm{inversegamma}(\alpha, \beta)}$, and $p_{\mathrm{normal}(\mu, \sigma)}$ are density functions for geometric, inverse gamma, and normal distributions, respectively.
-
-\paragraph{Traces of Probabilistic Programs.}
-In general, the density $p$ of a probabilistic program with respect to the reference measure $\mu$ on dictionaries, evaluated at a dictionary $x = (A, \mathbf{x})$, can be computed as the product of individual density functions evaluated at each random choice $x[a]$.
-To compute the density of a dictionary $x$, many probabilistic programming systems including Gen employ traces.
-In the context of probabilistic programming, a \emph{trace} $t$ is a record of each of the random choices made during an execution of a probabilistic program, including (i) the addresses $A$ of the choices, (ii) their values $\mathbf{x}$, and (iii) the order in which the addresses was visited (as a list of addresses $t_1, \ldots, t_{|A|} \in A$), and (iv) the probability distributions $p^{(t)}_a$ from which each address $a \in A$ was sampled during the execution.
-That is, each trace $t$ contains a dictionary $x = (A, \mathbf{x})$ such that $p(x) > 0$, as well as the order information $t_1, \ldots, t_{|A|}$ and the distribution information $p^{(t)}_a$ for each $a$.
-If a trace $t$ is available for a dictionary $x$, then the probability density of the dictionary is the product of the densities of each random choice in the trace:
-\footnote{Probabilistic programming implementations use sums of log-densities instead of products for numerical reasons.}
-\begin{equation} \label{eq:density-product}
-p((A, \mathbf{x})) = \prod_{i=1}^{|A|} p^{(t)}_{a}(x[a_{t_i}] | x[a_{t_1}], \ldots, x[a_{t_{i-1}}])
-\end{equation}
-See~\citet{popltraces} for an example of a formal proof that the product of densities is the density of the dictionary $x$ with respect to the reference measure.
-Note that although in general, the probability density $p^{(t)}_a$ of address $a$ is a function of the value of the choice $x[a]$ as well the values of all previously sampled choices, in practice it is only a function of a subset of previous choices due to conditional independence structure in the model.
-
-Given a dictionary $x$ for which $p(x) > 0$, it is possible to obtain a trace $t$ containing that dictionary by executing the probabilistic program, but intercepting random choice expressions and modifying their behavior.
-For each intercepted random choice expression during the execution, instead of sampling the value of the random choice at address $a$, we (i) append the address $a$ to the list of addresses ($t_1, t_2, \ldots$), (ii) record the probability distribution $p^{(t)}_a$, (iii) look up the address $a$ in the dictionary (obtaining $v \gets x[a]$), and (iv) resume regular program execution, evaluating the random choice expression to the value $v$ obtained from the dictionary.
-We denote a trace $t$ obtained from a dictionary $x$ and a probabilistic program $P$:
-\begin{equation}
-t \gets \mathrm{trace}(P, x)
-\end{equation}
-
-% TODO add the algorithm box involving traces..
+when $A$ contains exactly the addresses $\mathtt{:}\mathtt{k}, \mathtt{:}\mathtt{weights}, (\mathtt{:}\mathtt{x}, i)$ for $i = 1, \dots, n$, and $(\mathtt{:}\mathtt{mu}, j)$ and $(\mathtt{:}\mathtt{var}, j)$ for $j = 1, \dots, x_{\mathtt{:}\mathtt{k}}$; otherwise, the density is 0.
+
+% where $m(t, i) := \sum_{j=0}^{x[\mathtt{:}\mathtt{degree}]} \mathtt{xcoords}[i]^j \cdot x[(\mathtt{"c"}, j)]$, and where $p_{\mathrm{geometric}(r)}$, $p_{\mathrm{inversegamma}(\alpha, \beta)}$, and $p_{\mathrm{normal}(\mu, \sigma)}$ are density functions for geometric, inverse gamma, and normal distributions, respectively.
+
+% The addresses used by this generative function are:
+% \[
+% \mathcal{A} := \begin{array}{c}
+% \{\mathtt{:}\mathtt{noise}, \mathtt{:}\mathtt{degree}\} \bigcup
+% \{(\mathtt{"c"}, i) : i \in \integers_{\ge 0}\} \bigcup\\
+% \{(\mathtt{"y"}, i) : i \in \integers_{> 0}\}
+% \end{array}
+% \]
+% (Note that $\mathcal{A}$ could contain additional addresses that the program never samples.)
+% The random choice at address $\mathtt{:}\mathtt{degree}$ is discrete with $V_a = \integers_{\ge 0}$ and the other random choices are continuous.
+% The address $\mathtt{:}\mathtt{noise}$ has $V_a = \reals_{> 0}$ and $(\mathtt{"c"}, i)$ and $(\mathtt{"y"}, i)$ have $V_a = \reals$, and at all of these continuous addresses, we use the Lebesgue measure for $\mu_a$.
+% TODO double-check this
+% The density with respect to the reference measure $\mu$ is then:
+% \begin{align*}
+% p(x) =
+% \begin{array}{l}
+% p_{\mathrm{geometric}(0.5)}(x[\mathtt{:}\mathtt{degree}]) \cdot\\
+% p_{\mathrm{inversegamma}(1, 1)}(x[\mathtt{:}\mathtt{noise}]) \cdot\\
+% \prod_{0=1}^{x[\mathtt{:}\mathtt{degree}]} p_{\mathrm{normal}(0,1)}(x[(\mathtt{"c"}, i)])\cdot\\
+% \prod_{i=1}^n p_{\mathrm{normal}(m(t, i), 0.1 + x[\mathtt{noise}])}(x[(\mathtt{"y"}, i)])
+% \end{array}
+% \end{align*}
+% where $m(t, i) := \sum_{j=0}^{x[\mathtt{:}\mathtt{degree}]} \mathtt{xcoords}[i]^j \cdot x[(\mathtt{"c"}, j)]$, and where $p_{\mathrm{geometric}(r)}$, $p_{\mathrm{inversegamma}(\alpha, \beta)}$, and $p_{\mathrm{normal}(\mu, \sigma)}$ are density functions for geometric, inverse gamma, and normal distributions, respectively.
+
+% In our framework, a probabilistic program defines a probability distribution on finite maps (dictionaries) that map the names of random choices (their \emph{addresses}) to values.
+% % TODO trace terminology
+% Following standard terminology~\citep{?}, we call these dictionaries \emph{traces} because they are often sampled by running a probabilistic program, and `tracing' the execution of the program by recording the value of each random choice encountered.
+% We now make the concept of a trace more concrete using code in a probabilistic programming language.
+
+% The probabilistic programming language used in this paper is 
+% the dynamic modeling language in the Gen probabilistic programming system~\citep{gen-pldi}.
+% The language is embedded in the general-purpose Julia programming language~\citep{julia}.
+% Julia functions can be turned into Gen probabilistic programs by tagging
+% each random choice with an explicit address in the code, 
+% as part of a \emph{random-choice expression} of the following form:
+% \noindent
+% \begin{center}
+% \begin{tabular}{c}
+% {\begin{lstlisting}[basicstyle=\small\ttfamily]
+% {<address>} ~ <distribution>
+% \end{lstlisting}}
+% \end{tabular}
+% \end{center}
+% The left-hand side of the expression (green) specifies the address of the choice, and the right-hand side specifies the probability distribution to be sampled from.
+% For example, the following expression samples a non-negative integer from a geometric distribution and the value gets recorded in the trace at the address \texttt{"a"}:
+% \noindent
+% \begin{center}
+% \begin{tabular}{c}
+% {\begin{lstlisting}[basicstyle=\small\ttfamily]
+% {"a"} ~ geometric(0.5)
+% \end{lstlisting}}
+% \end{tabular}
+% \end{center}
+% In addition to being recorded in the trace, the sampled value is returned to the running program.
+% That is, the random choice expression evaluates to the sampled value.
+% For example, the expression below samples the random variable \texttt{"a"} and multiplies it by 2,
+% yielding an even non-negative integer:
+% \noindent
+% \begin{center}
+% \begin{tabular}{c}
+% {\begin{lstlisting}[basicstyle=\small\ttfamily]
+% ({"a"} ~ geometric(0.5)) * 2
+% \end{lstlisting}}
+% \end{tabular}
+% \end{center}
+% A probabilistic program will typically combine multiple random choice expressions. For example,
+% the following code samples a random choice at address \texttt{"a"}, assigns it to the Julia variable
+% \texttt{a}, and then samples a Bernoulli random variable at address \texttt{"b"} with a parameter
+% that depends on \texttt{a}:
+% \noindent
+% \begin{center}
+% \begin{tabular}{c}
+% {\begin{lstlisting}[basicstyle=\small\ttfamily]
+% a = ({"a"} ~ geometric(0.5))
+% {"b"} ~ bernoulli(a / (a + 1))
+% \end{lstlisting}}
+% \end{tabular}
+% \end{center}
+% This probabilistic program has an unbounded (countably infinite) number of possible traces it could produce.
+% One possible trace samples the value 1 at $\texttt{"a"}$ and the Boolean value $\false$ at \texttt{"b"}, and we denote this trace by $\{\mathtt{"a"} \mapsto 1, \mathtt{"b"} \mapsto \false\}$.
+% Note that the Julia variable to which the first choice is assigned (\texttt{a}) happens to be the same as the address (\texttt{"a"}).
+% This is not required, and often the random choice has no corresponding Julia variable in the program, as in the second random choice with address \texttt{"b"}.
+% But to reduce unnecessary typing, the following optional syntactic sugar will simultaneously assign the value to the program variable $\mathtt{a}$ and record it at trace address $\mathtt{"a"}$.\footnote{
+% More precisely, the address is the Julia symbol (interned string) \texttt{a}, constructed with Julia syntax \texttt{:a}.}
+% \noindent
+% \begin{center}
+% \begin{tabular}{c}
+% {\begin{lstlisting}[basicstyle=\small\ttfamily]
+% a ~ geometric(0.5)
+% \end{lstlisting}}
+% \end{tabular}
+% \end{center}
+
+% What gives probabilistic programs much of their expressive power is that they can make use of familiar control flow constructs, like loops and if statements, and this control flow can both depend on random choices and influence which random choices are made downstream.
+Consider the following program implementing a geometric
+% distribution as the result of iterated Bernoulli draws:
+% \noindent
+% \begin{center}
+% \begin{tabular}{c}
+% {\begin{lstlisting}[basicstyle=\small\ttfamily]
+% i = 1
+% while ({i} ~ bernoulli(0.5))
+%     i = i + 1
+% end
+% \end{lstlisting}}
+% \end{tabular}
+% \end{center}
+% Note that each Bernoulli flip uses a different address (the number \texttt{i} increases at each iteration).
+For the probabilistic program to be valid, we require that every address is used at most once within every possible execution.
+
+% For probabilistic programs that sample from discrete probability distributions like those used above, the meaning of the program is a probability mass function $p$ that assigns a probability to every possible trace.
+% For example, for the program with the \texttt{while} loop above, the distribution is:
+% \[
+% \begin{array}{c|c}
+%     x & p(x)\\
+%     \hline
+%     \{1 \mapsto \false\} & 0.5\\
+%     \{1 \mapsto \true, 2 \mapsto \false\} & 0.5^2\\
+%     \{1 \mapsto \true, 2 \mapsto \true, 3 \mapsto \false\} & 0.5^3\\
+%     .. & ..
+% \end{array}
+% \]
+% We assume that probabilistic programs halt with probability $1$, meaning that the sum of probabilities over all (finite) traces is one.
+% Note that traces are always finite, but there may be an infinite number of possible traces, as in the previous example.
+
+% In order to construct more complex distributions, we wrap probabilistic programs in functions, called
+% `generative functions' in Gen.
+% These functions can then be invoked by other functions, including recursively, which allows programs encode probability distributions on more complex and combinatorial objects.
+% When invoking such a function, we give an \emph{address namespace} under which every random choice made in that function should be recorded.
+% This helps prevent addresses of choices defined in different functions from colliding.
+% The syntax for calling a generative function and assigning the address namespace is the same as that of random choice expressions.
+% For example:
+% \begin{center}
+% \begin{tabular}{c}
+% \begin{lstlisting}[basicstyle=\small\ttfamily]
+% @gen function g()
+%   if ({"go"} ~ bernoulli(0.2))
+%     n1 = ({"L"} ~ g())
+%     n2 = ({"R"} ~ g())
+%     return n1 + n2 + 1
+%   else
+%     return 1
+%   end
+% end
+% \end{lstlisting}
+% \end{tabular}
+% \end{center}
+% Because this is a recursive function, a the trace is populated with a hierarchy of addresses.
+% For example, the first random choice in the first recursive call has address \texttt{"L"."go"}.
+% When this function is invoked it encodes the following probability distribution on traces (quotation marks omitted from addresses to improve readability):
+% \[
+% \begin{array}{c|c}
+%     x & p(x)\\
+%     \hline
+%     \left\{\mathtt{go} \mapsto \false\right\} & 0.8\\
+% & \\
+%     \left\{\begin{array}{c}
+%         \mathtt{go} \mapsto \true, \mathtt{L}.\mathtt{go} \mapsto \false, \mathtt{R}.\mathtt{go} \mapsto \false
+%     \end{array}\right\} & 0.2 \cdot 0.8^2\\
+% & \\
+%     \left\{\begin{array}{c}
+%         \mathtt{go} \mapsto \true,
+%         \mathtt{L}.\mathtt{go} \mapsto \true,
+%         \mathtt{R}.\mathtt{go} \mapsto \false,\\
+%         \mathtt{L}.\mathtt{L}.\mathtt{go} \mapsto \false,
+%         \mathtt{L}.\mathtt{R}.\mathtt{go} \mapsto \false
+%     \end{array}\right\}& 0.2^2 \cdot 0.8^3\\
+%      .. & ..
+% \end{array}
+% \]
+
+% %\subsubsection{Handling Continuous Distributions}
+% %To handle programs that sample continuous probability distributions the notion of a probability mass function on traces needs to be generalized to a probability density function on traces with respect to a certain reference measure.
+% %The probabilistic program still defines a probability distribution (i.e. measure) on traces, but via  a density instead of a probability mass function.
+% %
+% %Let $\mathcal{A}$ denote a countable set of all possible addresses.
+% %For each address $a \in \mathcal{A}$ we assign a reference measure space $(V_a, \Sigma_a, \mu_a)$ where $\mu_a$ is $\sigma$-finite ($V$ stands for `values').
+% %Let $T := \cup_{A \subseteq \mathcal{A} : |A| < \infty} \{(A, \mathbf{x}) : \mathbf{x} \in \times_{a \in A} V_a\}\}$ denote the set of all finite traces (i.e.
+dictionaries) from addresses in $\mathcal{A}$ to their values.
+% %For each finite $A \subseteq \mathcal{A}$, let $\Sigma_A := \otimes_{a \in A} \Sigma_a$ (the product $\sigma$-algebra over $\Sigma_a$ for $a \in A$), and let $\mu_A := \times_{a \in A} \mu_a$ (the product measure).
+% %Let $\Sigma \subseteq \mathcal{P}(T)$ be the $\sigma$-algebra formed by sets of the form $B := \cup_{A \subseteq \mathcal{A}, |A| < \infty} \{(A, \mathbf{x}) : \mathbf{x} \in B_A\}$ for some $B_A \in \Sigma_A$ for each $A$.
+% %Let $\mu : \Sigma \to [0, \infty)$ be a measure on measurable space $(T, \Sigma)$ defined by $\mu(B) := \sum_{A \subseteq \mathcal{A}, |A| < \infty} \mu_A(B_A)$.
+% %
+
+% \subsubsection{Automatically Computing Densities}
+% % TODO explain the non-standard interpretation idea.
+% A probabilistic program $P$ denotes a probability density $p : T \to [0, \infty)$ with respect to a reference measure $\mu$ on dictionaries.
+% Defining the denotational semantics of a particular probabilistic programming language (the function mapping programs $P$ to density functions $p$) is out of scope for this paper, and we refer to the reader to~\citet{popltraces} for an example of denotational semantics of a probabilistic programming language similar to that of this paper.
+% %But, informally, the density of a trace in our formalism (and existing formalisms) is the product of densities of each individual random choice, as an example below illustrates.
+% But this section illustrates what form density functions for probabilistics programs take and how to they can be automatically evaluated by probabilistic programming systems.
+% For example, consider the following probabilistic program, which encodes a generative model for data generated from a polynomial with unknown degree and unknown coefficients:
+% \noindent
+% \begin{center}
+% \begin{tabular}{c}
+% {\begin{lstlisting}[basicstyle=\small\ttfamily]
+% @gen function polynomials(x_coords)
+%     degree ~ |geometric(0.5)|
+%     noise ~ |inv_gamma(1, 1)|
+%     coeffs = [ ({("c", i)} ~ |normal(0, 2)|)
+%                for i in 0:degree ]
+%     for i=1:length(x_coords)
+%         x = x_coords[i]
+%         mu = coeffs' * x.^(0:length(coeffs)-1)
+%         {("y", i)} ~ |normal(mu, noise + 0.1)|
+%     end
+% end
+% \end{lstlisting}}
+% \end{tabular}
+% \end{center}
+% The addresses used by this generative function are:
+% \[
+% \mathcal{A} := \begin{array}{c}
+% \{\mathtt{:}\mathtt{noise}, \mathtt{:}\mathtt{degree}\} \bigcup
+% \{(\mathtt{"c"}, i) : i \in \integers_{\ge 0}\} \bigcup\\
+% \{(\mathtt{"y"}, i) : i \in \integers_{> 0}\}
+% \end{array}
+% \]
+% (Note that $\mathcal{A}$ could contain additional addresses that the program never samples.)
+% The random choice at address $\mathtt{:}\mathtt{degree}$ is discrete with $V_a = \integers_{\ge 0}$ and the other random choices are continuous.
+% The address $\mathtt{:}\mathtt{noise}$ has $V_a = \reals_{> 0}$ and $(\mathtt{"c"}, i)$ and $(\mathtt{"y"}, i)$ have $V_a = \reals$, and at all of these continuous addresses, we use the Lebesgue measure for $\mu_a$.
+% TODO double-check this
+% The density with respect to the reference measure $\mu$ is then:
+% \begin{align*}
+% p(x) =
+% \begin{array}{l}
+% p_{\mathrm{geometric}(0.5)}(x[\mathtt{:}\mathtt{degree}]) \cdot\\
+% p_{\mathrm{inversegamma}(1, 1)}(x[\mathtt{:}\mathtt{noise}]) \cdot\\
+% \prod_{0=1}^{x[\mathtt{:}\mathtt{degree}]} p_{\mathrm{normal}(0,1)}(x[(\mathtt{"c"}, i)])\cdot\\
+% \prod_{i=1}^n p_{\mathrm{normal}(m(t, i), 0.1 + x[\mathtt{noise}])}(x[(\mathtt{"y"}, i)])
+% \end{array}
+% \end{align*}
+% where $m(t, i) := \sum_{j=0}^{x[\mathtt{:}\mathtt{degree}]} \mathtt{xcoords}[i]^j \cdot x[(\mathtt{"c"}, j)]$, and where $p_{\mathrm{geometric}(r)}$, $p_{\mathrm{inversegamma}(\alpha, \beta)}$, and $p_{\mathrm{normal}(\mu, \sigma)}$ are density functions for geometric, inverse gamma, and normal distributions, respectively.
+
+\paragraph{Automating Density Computation.}
+% In general, the density $p$ of a probabilistic program with respect to the reference measure $\mu$ on dictionaries, evaluated at a dictionary $x = (A, \mathbf{x})$, can be computed as the product of individual density functions evaluated at each random choice $x[a]$.
+% To compute the density of a dictionary $x$, many probabilistic programming systems including Gen employ traces.
+% In the context of probabilistic programming, a \emph{trace} $t$ is a record of each of the random choices made during an execution of a probabilistic program, including (i) the addresses $A$ of the choices, (ii) their values $\mathbf{x}$, and (iii) the order in which the addresses was visited (as a list of addresses $t_1, \ldots, t_{|A|} \in A$), and (iv) the probability distributions $p^{(t)}_a$ from which each address $a \in A$ was sampled during the execution.
+% That is, each trace $t$ contains a dictionary $x = (A, \mathbf{x})$ such that $p(x) > 0$, as well as the order information $t_1, \ldots, t_{|A|}$ and the distribution information $p^{(t)}_a$ for each $a$.
+% If a trace $t$ is available for a dictionary $x$, then the probability density of the dictionary is the product of the densities of each random choice in the trace:
+% \footnote{Probabilistic programming implementations use sums of log-densities instead of products for numerical reasons.}
+% \begin{equation} \label{eq:density-product}
+% p((A, \mathbf{x})) = \prod_{i=1}^{|A|} p^{(t)}_{a}(x[a_{t_i}] | x[a_{t_1}], \ldots, x[a_{t_{i-1}}])
+% \end{equation}
+% See~\citet{popltraces} for an example of a formal proof that the product of densities is the density of the dictionary $x$ with respect to the reference measure.
+% Note that although in general, the probability density $p^{(t)}_a$ of address $a$ is a function of the value of the choice $x[a]$ as well the values of all previously sampled choices, in practice it is only a function of a subset of previous choices due to conditional independence structure in the model.
+
+% Given a dictionary $x$ for which $p(x) > 0$, it is possible to obtain a trace $t$ containing that dictionary by executing the probabilistic program, but intercepting random choice expressions and modifying their behavior.
+% For each intercepted random choice expression during the execution, instead of sampling the value of the random choice at address $a$, we (i) append the address $a$ to the list of addresses ($t_1, t_2, \ldots$), (ii) record the probability distribution $p^{(t)}_a$, (iii) look up the address $a$ in the dictionary (obtaining $v \gets x[a]$), and (iv) resume regular program execution, evaluating the random choice expression to the value $v$ obtained from the dictionary.
+% We denote a trace $t$ obtained from a dictionary $x$ and a probabilistic program $P$:
+% \begin{equation}
+% t \gets \mathrm{trace}(P, x)
+% \end{equation}
+The density of a dictionary $(A, \mathbf{x})$ under any probabilistic program's distribution on traces can be computed using Algorithm~\ref{alg:prob-prog} (\textsc{trace-and-score}).
+The idea is to run the probabilistic program, and whenever a random choice \texttt{\{a\} $\sim$ d} is encountered, to look up the value $x_a$ in the dictionary, compute its density under the primitive distribution $d$, multiply this density into a running total, and return control to the probabilistic program as if the sampling instruction had executed and returned $x_a$.
+At the end, the running total can be returned as the trace's density.
+(If at any point an address $a \not\in A$ is encountered, or if not all addresses in $A$ have been visited at the end of execution, the algorithm returns $0$ as the density.) This is a standard technique in probabilistic programming.
 

### 2020-07-16 paper.tex

@@
 \Procedure{involutive-mcmc}{$p$, $q$, $f$, $x$}
-    \State $y \sim q_x(\cdot)$ \Comment{Sample auxiliary dictionary}
+    \State $y \sim q_x(\cdot)$ \Comment{Sample auxiliary state}
     \State $(x', y') \gets f(x, y)$ \Comment{Apply involution}
     \State $\alpha \gets
-        \displaystyle \frac{p(x') q_{x'}(y')}{p(x) q_{x}(y)} \cdot |J h_e|(z)$
+        \displaystyle \frac{p(x') q_{x'}(y')}{p(x) q_{x}(y)} \cdot \left( \frac{d (\mu \circ f^{-1})}{d \mu} (x, y)\right)$
     \State $r \sim \mathrm{Uniform}(0, 1)$
@@
 
-\begin{algorithm}[h]
+\begin{algorithm*}[h]
 \begin{algorithmic}
-\Procedure{simulate-and-score}{$P(\beta)$}
+\Procedure{auto-involutive-mcmc}{$\mathcal{P}$, $\mathcal{Q}$, $\mathcal{F}$, $x$}
+    \State $(y, \log w_2) \gets \textproc{trace-and-score}(\mathcal{Q}_x)$ \Comment{Sample $y \sim q_x(\cdot)$ and compute log-density $\log q_x(y)$}
+    \State $(x', y', \log D) \gets \textproc{run-involution}(\mathcal{F}, (x, y))$ \Comment{Compute $(x', y') = f(x, y)$, log Radon-Nikodym deriv. $\log D$}
+    \State $\log w_1 \gets \textproc{score}(\mathcal{P}, x)$ \Comment{Compute log-density $\log p(x)$}
+    \State $\log w_3 \gets \textproc{score}(\mathcal{P}, x')$ \Comment{Compute log-density $\log p(x')$}
+    \State $\log w_4 \gets \textproc{score}(\mathcal{Q}_{x'}, y')$ \Comment{Compute log-density $\log q_{x'}(y')$}
+    \State $\alpha \gets \min\{1, \exp(\log w_3 - \log w_1 + \log w_4 - \log w_2 + \log J)\}$ \Comment{Compute acceptance probability}
+    \State \textbf{with probability} \,$\alpha$ \,\Return $x'$ \algorithmicelse \, \Return $x$ 
+\EndProcedure\\
+
+\begin{minipage}[t]{0.5\linewidth}
+\Procedure{trace-and-score}{$\mathcal{P}$}
    \State $\log w \gets 0$
    \State $x \gets \{\}$
-   \State Execute $P$ on input $\beta$, replacing $"\mathtt{a \sim d}"$ with:
-    \State $\;\;\;\;$ 1. $\; v \sim d$
-    \State $\;\;\;\;$ 2. $\; x[a] \gets v$
-    \State $\;\;\;\;$ 3. $\; \log w \gets \log w + \textproc{logpdf}(d, x[a])$
-    \State $\;\;\;\;$ 4. $\;$ \Return $v$
+   \execute[$\mathcal{P}$], but with $\mathtt{"a \sim d"} \equiv$ (
+        \State \hspace{5mm} 1. set $v \sim d$
+        \State \hspace{5mm} 2. set $x[a] \gets v$
+        \State \hspace{5mm} 3. set $\log w \gets \log w + \textproc{logpdf}(d, v)$
+        \State \hspace{5mm} 4. evaluate to $v$)
+    \END
    \State \Return $(x, \log w)$
 \EndProcedure
-
-\Procedure{score}{$P(\beta), x$}
+\end{minipage}%
+\begin{minipage}[t]{0.5\linewidth}
+\Procedure{score}{$\mathcal{P}$, $x$}
    \State $\log w \gets 0$
-   \State Execute $Q$ on input $x$, replacing $"\mathtt{a \sim d}"$ with:
-    \State $\;\;\;\;$ 1. $\; v \sim x[a]$
-    \State $\;\;\;\;$ 3. $\; \log w \gets \log w + \textproc{logpdf}(d, x[a])$
-    \State $\;\;\;\;$ 4. $\;$ \Return $v$
+   \execute[$\mathcal{P}$], but with $\mathtt{"a \sim d"} \equiv$ (
+        \State \hspace{5mm} 1. set $v \gets x[a]$
+        \State \hspace{5mm} 2. set $\log w \gets \log w + \textproc{logpdf}(d, v)$
+        \State \hspace{5mm} 3. evaluate to $v$)
+    \END
    \State \Return $\log w$
 \EndProcedure
-
-\Procedure{run-involution}{$F$, $(x, y)$}
-    TODO
+\end{minipage}\\
+
+\Procedure{run-involution}{$\mathcal{F}$, $(x, y)$}
+% TODO would be much simpler to describe if we didn't have to separate out x and y..
+    \LineComment{Run the involution, and keep track of continuous addresses}
+    \State $x' \gets \{\}$ \Comment{Initialize empty output dictionary}
+    \execute[$\mathcal{F}$], but with
+        \State $\mathtt{"write\_cont}(a, v)\mathtt{"} \equiv$ (1. set $x'[a] \gets v$ and 2. set $W \gets W \cup \{a\}$)
+        \State $\mathtt{"read\_cont}(a)\mathtt{"} \equiv$ (1. set $R \gets R \cup \{a\}$ and 2. evaluate to $x[a]$)
+        \State $\mathtt{"copy}(a_1, a_2)\mathtt{"} \equiv$ (1. set $x'[a_2] \gets x[a_1]$ and 2. set $C \gets C \cup \{a_2\}$)
+    \END
+    \LineComment{Use automatic differentiation to compute Jacobian, skipping copied addresses}
+    \For{$i$ in $1, \ldots, |W \setminus C|$}
+        \State $J[i,:] \gets \nabla_{x_{R \setminus C}} \left( h_{e(x, y)}[(W \setminus C)_i] \right)$ \Comment{Gradient of $x'[(W \setminus C)_i]$ w.r.t. non-copied continuous inputs}
+    \EndFor
+    \State \Return $(x', \log |\det(J)|)$
 \EndProcedure
-\end{algorithmic}
-\caption{Probabilistic and Differentiable Program Operations}
-\label{alg:prob-prog}
-\end{algorithm}
-
-\begin{algorithm}[h]
-\begin{algorithmic}
-\Procedure{auto-involutive-mcmc}{$P$, $Q$, $F$, $x$}
-    \State $\log w_1 \gets \textproc{score}(P(), x)$
-    \State $(y, \log w_2) \gets \textproc{simulate-and-score}(Q(x))$
-    \State $(x', y', \log |J|) \gets \textproc{run-involution}(F, (x, y))$
-    \State $\log w_2 \gets \textproc{score}(Q(x), y)$
-    \State $\log w_3 \gets \textproc{score}(P(), x')$
-    \State $\log w_4 \gets \textproc{score}(Q(x'), y')$
-    \State $\log \alpha \gets \log w_3 - \log w_1 + \log w_4 - \log w_2 + \log |J|$
-    \State $r \sim \mathrm{Uniform}(0, 1)$
-    \State \algorithmicif \, $r \le \exp(\alpha)$ \algorithmicthen \, \Return $x'$ \algorithmicelse \, \Return $x$ 
-\EndProcedure
+
 \end{algorithmic}
 \caption{Automated Involutive MCMC}
-\label{alg:involutive-mcmc}
-\end{algorithm}
+\label{alg:auto-involutive-mcmc}
+\end{algorithm*}
 

### 2020-07-16 paper.tex
Checkpoint

@@
 
-\textbf{Notation for dictionaries.} Given a dictionary $x = (A, \mathbf{x})$, we write $A_m$ for $A$ and $m[a]$ for the value $x_a$ associated with a key $a \in A$.
+\textbf{Notation for dictionaries.} Given a dictionary $d = (A, \mathbf{x})$, we write $A_d$ for $A$ and $d[a]$ for the value $x_a$ associated with a key $a \in A$.
 We also denote specific dictionaries using notation $\{k_1 \mapsto v_1, k_2 \mapsto v_2, \ldots\}$.
@@
 
-\textit{Gaussian mixture model with unknown number of components.} Consider a generative model of univariate data points $y_1, \dots, y_n$ from a Gaussian mixture with an unknown number of components $k$, each with unknown mean $m_i$ and standard deviation $s_i$.
-If we place a Gaussian prior on $m_i$, a Gamma prior on $s_i$, and a Poisson prior on $k$, the resulting density is
+\textit{Gaussian mixture model with unknown number of components.} Consider a generative model of univariate data points $y_1, \dots, y_n$ from a Gaussian mixture with an unknown number of components $k$, each with unknown mean $m_i$ and variance $s_i$.
+If we place a Gaussian prior on $m_i$, an inverse Gamma prior on $s_i$, and a Poisson prior on $k$, the resulting density is
 %
-$$p(d) = TODO$$
+\begin{align*}
+p(d) =
+\begin{array}{l}
+p_{\mathrm{poisson}(3)}(d[\mathtt{k}]) \cdot\\
+\prod_{i=1}^{d[\mathtt{k}]} p_{\mathrm{normal}(0, 1)}(d[\mathtt{m}_i]) \cdot\\
+\prod_{i=1}^{d[\mathtt{k}]} p_{\mathrm{inversegamma}(1, 10)}(d[\mathtt{s}_i]) \cdot\\
+\prod_{i=1}^n  \frac{1}{d[\mathtt{k}]} \sum_{j=1}^{d[\mathtt{k}]} \cdot p_{\mathrm{normal}(d[\mathtt{m}_j], d[\mathtt{s}_i])}(d[\mathtt{y}_i])
+\end{array}
+\end{align*}
 %
@@
 
-Provided the two probabilistic programs $P$ and $Q$ satisfy a simple technical requirement (described in Section~\ref{sec:technical-requirement}), they can be used to specify the model density $p$ and auxiliary densities $p_x$, respectively for use with involutive MCMC on dictionaries (Algorithm~\ref{alg:involutive-mcmc}).
+Provided the two probabilistic programs $P$ and $Q$ satisfy a simple technical requirement (described in Section~\ref{sec:technical-requirement}), they can be used to specify the model density $p$ and auxiliary densities $q_x$ for use with involutive MCMC on dictionaries (Algorithm~\ref{alg:involutive-mcmc}).
 The acceptance probability of involutive MCMC includes the following ratio of densities:
@@
 \end{equation}
-To evaluate this density ratio, we can obtain four traces, one for each dictionary, and use those traces to compute the four probability densities, as shown in Algorithm~\ref{alg:trace-based-involutive-mcmc}.
-Section~\ref{sec:sparsity} describes a more efficent approach that exploits sparsity in the involution and resulting cancellation in the ratio for improved efficiency, but requires a more sophisticated probabilistic programming runtime system.
+This ratio can be computed term-by-term using \textsc{trace-and-score} (Algorithm~\ref{alg:trace-based-involutive-mcmc}). 
+Section~\ref{sec:sparsity} describes a more efficient approach that exploits sparsity in the involution and resulting cancellation in the ratio for improved efficiency, but requires a more sophisticated probabilistic programming runtime system.
 
@@
 % TODO show an example of the type of bug it can detect (an edge case?)
+
+\textbf{Dynamic checks during inference.} These dynamic assertions can also be run during inference, at each application of the transition kernel.
+This can be useful to catch bugs that only occur in regions of the state space with low prior mass (but perhaps high posterior mass).
+When enabled, we can run dynamic checks after each application of the kernel, and when they fail, write to a debugging log and reject the proposed new state.
+As it turns out, the kernel induced by this procedure is still stationary for $p$:
+
+\begin{lemma}
+Let $p$ and $q_x$ be model and auxiliary densities as above, but suppose $f : \mathcal{D} \times \mathcal{D} \rightarrow \mathcal{D} \times \mathcal{D}$ may not be an involution on $Z$.
+Trace-based involutive MCMC with dynamic checks enabled, rejecting whenever such a check fails, still yields a kernel that is stationary for $p$.
+\end{lemma}
+\begin{proof}
+Let $R = \{x \in Z \mid f(f(x)) = x \wedge \pi(f(x)) > 0\}$, and let $h(x) := \mathbf{1}[x \in R]f(x) + \mathbf{1}[x \not\in R]x$.
+Then $h$ is an involution on $Z$, and trace-based involutive MCMC with $p$, $q_x$, and $h$ yields a stationary kernel.
+But this kernel is the same one induced by using $f$ with dynamic checks.
+For $x$ on which dynamic checks succeed, $h$ is equivalent to $f$.
+For $x$ on which dynamic checks fail, $h$ is equivalent to the identity; thus, accepting a move produced by $h$ is equivalent to rejecting.
+\end{proof}
 

### 2020-07-16 paper.tex
checkpoint on intro

@@
 \section{INTRODUCTION}
+Markov chain Monte Carlo (MCMC) algorithms are powerful tools for approximate sampling from probability distributions and are central to modern Bayesian statistics, probabilistic machine learning, statistical physics, and numerous applications of probabilistic inference from robotics to computational biology.
+But designing and deriving efficient MCMC algorithms is mathematically involved, and implementing MCMC kernels is tedious and notoriously error-prone.
+These challenges are especially pertinent when sampling from probability distributions on complex hybrid symbolic-numeric state spaces such as those arising in computational biology~\citep{huelsenbeck2004bayesian}, robotics and scene understanding~\citep{geiger2011generative}, and models of human cognition~\citep{tenenbaum2011grow}.
+
+Involutive MCMC is a mathematical construction for MCMC kernels that provides a simplifying and unifying perspective on a number of previously disparate classes of kernels,
+including reversible jump MCMC~\citep{green1995reversible}, the dominant mathematical framework for MCMC on complex hybrid state spaces.
+Involutive MCMC constructs an MCMC kernel from three components: (i) the unnormalized target density, (ii) a sampler and density for an auxiliary probability distribution (iii) an involution\footnote{a bijection that is its own inverse ($f$ where $f(f(z)) = z$)} on an extended state space.
+While this construction is mathematically clarifying, correctly implementing involutive MCMC kernels on complex state spaces remains challenging due to tedious density and Jacobian computations and the need for careful reasoning about the state space.
+
+This paper formulates involutive MCMC on general state spaces and shows how to automatically generate the implementation of an involutive MCMC kernel from high-level probabilistic programs and differentiable programs that define the relevant distributions and the involution, respectively.
+
+This paper introduces (1) a declarative interface for specifying involutive MCMC kernels and (2) an algorithm that automates the implementation of an involutive MCMC kernel from the declarative specification.
+
+The declarative interface takes as input two probabilistic programs that define the model distribution and auxiliary distribution respectively, and a differentiable program that defines the involution.
+
+The algorithm applies probabilistic programming techniques and automatic differentiation to the programs to fully automate the acceptance probability computation.
+
+The paper also shows how to improve the algorithm's asymptotic efficiency by exploiting the structure of the involution.
+
+The paper also gives an algorithm that dynamically detects mathematical and programming errors in the specification of an involutive MCMC algorithm.
+
+The paper also describes the implementation of this interface and these algorithms as part of the Gen probabilistic programming system.
+
+The paper also shows examples of involutive MCMC kernels specified using the interface that implement (1) a split-merge reversible jump move in an infinite mixture model, and a (2) state-dependent mixture of Metropolis-Hastings proposals on an infinite combinatorial space of covariance functions for a Gaussian process.
+
+The paper also gives a general measure-theoretic formulation of involutive MCMC, and a formulation of involutive MCMC on the more specialized but still very flexible state space of dictionaries that is the state space used by the Gen probabilistic programming system.
+
+%TODO: add
+    %representation of (Gen's) state space
+    %the language for specifying bijections on state space
 

### 2020-07-16 paper.tex
checkpoint on intro

--- 2020-07-16 paper.tex
@@
 % graphics
+\usepackage{hyperref}
 \usepackage{float}
@@
 \section{INTRODUCTION}
+Markov chain Monte Carlo (MCMC) algorithms are powerful tools for approximate sampling from probability distributions and are central to modern Bayesian statistics, probabilistic machine learning, statistical physics, and numerous applications of probabilistic inference.
+But designing and deriving efficient MCMC algorithms is mathematically involved, and implementing MCMC kernels is tedious and notoriously error-prone.
+These challenges are especially pertinent when sampling from probability distributions on complex state spaces that combine symbolic, numeric, and structural uncertainty, such as those arising in computational biology~\citep{huelsenbeck2004bayesian}, robotics and scene understanding~\citep{geiger2011generative}, and models of human cognition~\citep{tenenbaum2011grow}.
+
+Involutive MCMC is a mathematical construction for MCMC kernels that provides a simplifying and unifying perspective on a number of previously disparate classes of kernels,
+including reversible jump MCMC~\citep{green1995reversible}, the dominant mathematical framework for MCMC on complex hybrid state spaces.
+Involutive MCMC constructs an MCMC kernel from three components: (i) the unnormalized target density, (ii) a sampler and density for an auxiliary probability distribution (iii) an involution\footnote{a bijection that is its own inverse ($f$ where $f(f(z)) = z$)} on an extended state space.
+While this construction is mathematically clarifying, correctly implementing involutive MCMC kernels on complex state spaces remains challenging due to tedious density and Jacobian computations and the need for careful reasoning about the state space.
+
+This paper formulates involutive MCMC on general state spaces and shows how to automatically generate the implementation of an involutive MCMC kernel from three declarative programs that define the target probability distribution, auxiliary probability distribution, and the involution, respectively.
+The probability distributions are defined as probabilistic programs, and the involution is defined as a differentiable program that transforms the execution traces of the probabilistic programs.
+We use probabilistic programming techniques and automatic differentiation to automatically compute the acceptance probability.
+We also show how to automatically detect mathematical errors in the specification of an involutive MCMC kernel, and how to improve the efficiency by automatically exploiting the sparsity structure in the involution.
+We implemented the approach within the Gen probabilistic programming system\footnote{\url{https://www.gen.dev}}~\citep{cusumano2019pldi}.
+The paper shows examples of involutive MCMC kernels implemented in Gen for (1) a split-merge reversible jump move in an infinite mixture model, and a (2) state-dependent mixture of Metropolis-Hastings proposals on an infinite combinatorial space of covariance functions for a Gaussian process.
+We also provide a lightweight PyTorch implementation of the basic approach at:
+\begin{center}
+\vspace{-2mm}
+\url{https://github.com/probcomp/autoimcmc}.
+\end{center}
+
+The contributions of this paper include:
+\begin{enumerate}
+\item A 
+\end{enumerate}
+
+TODO: list contributiosn
 

### 2020-07-16 paper.tex
finish intro

@@
 
-This paper formulates involutive MCMC on general state spaces and shows how to automatically generate the implementation of an involutive MCMC kernel from three declarative programs that define the target probability distribution, auxiliary probability distribution, and the involution, respectively.
+This paper formulates involutive MCMC on general state spaces and shows how to automate the implementation of an involutive MCMC kernel from three declarative programs that define the target probability distribution, auxiliary probability distribution, and the involution, respectively.
 The probability distributions are defined as probabilistic programs, and the involution is defined as a differentiable program that transforms the execution traces of the probabilistic programs.
@@
 We also show how to automatically detect mathematical errors in the specification of an involutive MCMC kernel, and how to improve the efficiency by automatically exploiting the sparsity structure in the involution.
-We implemented the approach within the Gen probabilistic programming system\footnote{\url{https://www.gen.dev}}~\citep{cusumano2019pldi}.
+We implemented the approach within the Gen probabilistic programming system\footnote{\url{https://www.gen.dev}}~\citep{cusumano2019gen}.
 The paper shows examples of involutive MCMC kernels implemented in Gen for (1) a split-merge reversible jump move in an infinite mixture model, and a (2) state-dependent mixture of Metropolis-Hastings proposals on an infinite combinatorial space of covariance functions for a Gaussian process.
-We also provide a lightweight PyTorch implementation of the basic approach at:
-\begin{center}
-\vspace{-2mm}
-\url{https://github.com/probcomp/autoimcmc}.
-\end{center}
+We also provide a lightweight PyTorch implementation of the basic approach at
+{\small\url{https://github.com/probcomp/autoimcmc}}.
 
@@
 \begin{enumerate}
-\item A 
+\item A measure-theoretic formulation of involutive MCMC on general state spaces.
+\item A mathematical formulation of state spaces consisting of arbitrary key-value stores (i.e.
+dictionaries) that formalizes the space of execution traces of Gen probabilistic programs, and a formulation of involutive MCMC on these spaces.
+\item A differentiable programming language for defining bijections between spaces of traces of probabilistic programs.
+\item An algorithm that automates the implementation of an involutive MCMC kernel given two probabilistic programs and a differentiable program encoding the involution, using automatic differentiation and tracing of probabilistic programs.
+\item An extension to the algorithm that exploits sparsity in the involution to reduce the computational complexity of an involutive MCMC kernel from $O(N^2)$ to $O(1)$ in common cases.
+\item An algorithm that dynamically detects errors in the specification of an involutive MCMC kernel.
 \end{enumerate}
-
-TODO: list contributiosn
 

### 2020-07-16 paper.tex
tweak intro

@@
 
-Involutive MCMC is a mathematical construction for MCMC kernels that provides a simplifying and unifying perspective on a number of previously disparate classes of kernels,
-including reversible jump MCMC~\citep{green1995reversible}, the dominant mathematical framework for MCMC on complex hybrid state spaces.
+Involutive MCMC is a mathematical construction for MCMC kernels that gives a simplifying and unifying perspective on a number of previously disparate classes of kernels,
+including reversible jump MCMC~\citep{green1995reversible}, which is the dominant mathematical framework for MCMC on complex state spaces.
 Involutive MCMC constructs an MCMC kernel from three components: (i) the unnormalized target density, (ii) a sampler and density for an auxiliary probability distribution (iii) an involution\footnote{a bijection that is its own inverse ($f$ where $f(f(z)) = z$)} on an extended state space.

### 2020-07-16 paper.tex
tweak intro

@@
 We implemented the approach within the Gen probabilistic programming system\footnote{\url{https://www.gen.dev}}~\citep{cusumano2019gen}.
-The paper shows examples of involutive MCMC kernels implemented in Gen for (1) a split-merge reversible jump move in an infinite mixture model, and a (2) state-dependent mixture of Metropolis-Hastings proposals on an infinite combinatorial space of covariance functions for a Gaussian process.
+The paper shows examples of involutive MCMC kernels implemented in Gen for (i) a split-merge reversible jump move in an infinite mixture model, and a (ii) state-dependent mixture of Metropolis-Hastings proposals on an infinite combinatorial space of covariance functions for a Gaussian process.
 We also provide a lightweight PyTorch implementation of the basic approach at

### 2020-07-16 paper.tex
tweak intro

@@
 \item An algorithm that automates the implementation of an involutive MCMC kernel given two probabilistic programs and a differentiable program encoding the involution, using automatic differentiation and tracing of probabilistic programs.
-\item An extension to the algorithm that exploits sparsity in the involution to reduce the computational complexity of an involutive MCMC kernel from $O(N^2)$ to $O(1)$ in common cases.
+\item An extension to the algorithm that exploits sparsity in the involution to reduce the computational complexity of an involutive MCMC kernel from $O(N^2)$ to $O(1)$ in common cases where $N$ is the dimensionality of the latent space.
 \item An algorithm that dynamically detects errors in the specification of an involutive MCMC kernel.

### 2020-07-16 paper.tex
write a version of related work

@@
 
-\section{BACKGROUND AND RELATED WORK}
+\section{RELATED WORK}
+
+The involutive MCMC construction was previously implemented~\citep{gen-inv-mcmc} by the authors as part of the inference library of the Gen probabilistic programming system~\citep{cusumano2019gen}.
+The construction was motivated in part by a desire for a simple interface that automated the implementation of reversible jump MCMC samplers~\citep{green1995reversible}, state-dependent mixtures of proposals on complex state spaces, and data-driven neural proposals.
+Gen's involutive MCMC construction has since been used by a number of researchers to design and implement MCMC algorithms in diverse domains including computational biology~\citep{merrell2020inferring} and artificial intelligence~\citep{zhi2020online}.
+
+\citet{imcmc} independently identified the involutive MCMC construction as a unifying framework for MCMC algorithms, and showed how a number of classic and recent MCMC algorithms can be cast within this framework.
+\citet{imcmc} also identified design principles for developing new MCMC algorithms using the involutive MCMC construction, and showed that the framework aids in the derivation of novel efficient MCMC algorithms.
+
+The involutive MCMC construction encompasses many existing classes of MCMC kernels, some of which explicitly make use of bijective or involutive deterministic maps.
+In particular, the reversible jump framework~\citep{green1995reversible,hastie2012model} employs a family of continuously differentiable bijections between the Euclidean parameter spaces of different models.
+\citet{tierney1998note} described a family of deterministic proposals based on a deterministic involution that is equivalent to involutive MCMC but without the auxiliary proposal distribution.
+More recently,~\citet{spanbauer2020deep} defined a class of deep generative model based on differentiable involutions and trained these models to serve as efficient proposal distributions on continuous state spaces; the resulting algorithm is an instance of the construction presented in this paper.
+%Gen~\citet{cusumano2019gen}
+%Gen's involution MCMC construct~\citet{gen-inv-mcmc}
+%Reversible jump MCMC~\citet{green1995reversible}
+%^Using probabilistic programs as proposals~\citet{cusumano2018using}
+%Reversible jump probabilistic programming~\citet{roberts2019reversible}
 
@@
 %Let $y \sim q_x(\cdot)$ denote a random sample from the measure defined by $A \mapsto \int_A q_x(y) \mu_Q(dy)$ for $A \in \Sigma_Q$.
-
-\begin{theorem}[Involutive MCMC is stationary]
-Involutive MCMC defines a probability kernel $k$ on $X$ that is stationary with respect to the model probability distribution.
-That is, $\int_X k_x(B) p(x) d\mu_X(dx) = \int_B p(x) d\mu_X(dx)$ for all $B \in \Sigma_X$.
-\end{theorem}
-\begin{proof}
-The proof is presented in stages in the appendix (see Section~\ref{sec:involution-detailed-balance}, Section~\ref{sec:involution-is-stationary}, and Section~\ref{sec:involutive-mcmc-is-stationary}).
-\end{proof}
-
-\subsection{Involutive MCMC with Dictionaries}
-
-The general measure-theoretic formulation of involutive MCMC in Algorithm~\ref{alg:involutive-mcmc} does not specify how to compute the Radon-Nikodym derivative that is required for the acceptance probability.
-When the algorithm is specialized for state spaces consisting of dictionaries, the acceptance probability simplifies and can be computed from standard Jacobian and probability density computations.
-Because dictionaries are flexible objects, the algorithm remains quite general.
-
-Suppose that the model distribution $p$ and auxiliary distributions $q_x$ are all probability distributions on dictionaries.
-Then, $X$ and $Y$ are both sets of dictionaries, and the joint space $Z$ is a set of pairs $(x, y)$ of dictionaries with keys $\mathcal{A}$, so that $Z = X \times Y \subseteq \mathcal{D} \times \mathcal{D}$.
-%And suppose that there exists an involution $f : Z \to Z$ such that (i) 
-Suppose there is a countable partition of $Z$ into $\{Z_e : e \in E\}$ such that across each $Z_e$ the sets of keys $A_x$ and $A_y$ and the values $x_a$ and $y_a$ for all discrete keys $a$ are both constant (so that each set $Z_e$ is isomorphic to a Euclidean space of assignments to the continuous keys in the two dictionaries).
-Suppose there is an involution $g : E \to E$ between elements of the partition (a function with $g = g^{-1}$), and a family of continuously differentiable bijections $h_{e} : Z_e \to Z_{g(e)}$ for each $e$ with $h_{e} = h_{g(e)}^{-1}$.
-Let $e(z) \in E$ denote the element of the partition for an alement $z \in Z$.
-Then, $f : Z \to Z$ given by $f(z) := h_{e(z)}(z)$ is an involution:
-\begin{equation}
-f(f(z)) = h_{e(f(z))}(h_{e(z)}(z)) = h_{g(e(z))}(h_{e(z)}(z)) = z
-\end{equation}
-Let $|J h_e|(z)$ denote the absolute value of the Jacobian (determinant) of $h_e$, evaluated at $z$.
-Then, the acceptance probability simplifies to:
-\begin{equation}
-\alpha(x, y) = \frac{p(x') q_{x'}(y')}{p(x) q_{x}(y)} \cdot |J h_e|(z)
-\end{equation}
-Section~\ref{sec:radon-nikodym-special-case} of the appendix for details.
-
 \begin{algorithm}[h]
@@
 \end{algorithm}
+
+\begin{theorem}[Involutive MCMC is stationary]
+Involutive MCMC defines a probability kernel $k$ on $X$ that is stationary with respect to the model probability distribution.
+That is, $\int_X k_x(B) p(x) d\mu_X(dx) = \int_B p(x) d\mu_X(dx)$ for all $B \in \Sigma_X$.
+\end{theorem}
+\begin{proof}
+The proof is presented in stages in the appendix (see Section~\ref{sec:involution-detailed-balance}, Section~\ref{sec:involution-is-stationary}, and Section~\ref{sec:involutive-mcmc-is-stationary}).
+\end{proof}
+
+\subsection{Involutive MCMC with Dictionaries}
+
+The general measure-theoretic formulation of involutive MCMC in Algorithm~\ref{alg:involutive-mcmc} does not specify how to compute the Radon-Nikodym derivative that is required for the acceptance probability.
+When the algorithm is specialized for state spaces consisting of dictionaries, the acceptance probability simplifies and can be computed from standard Jacobian and probability density computations.
+Because dictionaries are flexible objects, the algorithm remains quite general.
+
+Suppose that the model distribution $p$ and auxiliary distributions $q_x$ are all probability distributions on dictionaries.
+Then, $X$ and $Y$ are both sets of dictionaries, and the joint space $Z$ is a set of pairs $(x, y)$ of dictionaries with keys $\mathcal{A}$, so that $Z = X \times Y \subseteq \mathcal{D} \times \mathcal{D}$.
+%And suppose that there exists an involution $f : Z \to Z$ such that (i) 
+Suppose there is a countable partition of $Z$ into $\{Z_e : e \in E\}$ such that across each $Z_e$ the sets of keys $A_x$ and $A_y$ and the values $x_a$ and $y_a$ for all discrete keys $a$ are both constant (so that each set $Z_e$ is isomorphic to a Euclidean space of assignments to the continuous keys in the two dictionaries).
+Suppose there is an involution $g : E \to E$ between elements of the partition (a function with $g = g^{-1}$), and a family of continuously differentiable bijections $h_{e} : Z_e \to Z_{g(e)}$ for each $e$ with $h_{e} = h_{g(e)}^{-1}$.
+Let $e(z) \in E$ denote the element of the partition for an alement $z \in Z$.
+Then, $f : Z \to Z$ given by $f(z) := h_{e(z)}(z)$ is an involution:
+\begin{equation}
+f(f(z)) = h_{e(f(z))}(h_{e(z)}(z)) = h_{g(e(z))}(h_{e(z)}(z)) = z
+\end{equation}
+Let $|J h_e|(z)$ denote the absolute value of the Jacobian (determinant) of $h_e$, evaluated at $z$.
+Then, the acceptance probability simplifies to:
+\begin{equation}
+\alpha(x, y) = \frac{p(x') q_{x'}(y')}{p(x) q_{x}(y)} \cdot |J h_e|(z)
+\end{equation}
+Section~\ref{sec:radon-nikodym-special-case} of the appendix for details.
 

### 2020-07-16 paper.tex
Rewrite abstract

@@
 \begin{abstract}
-Involutive MCMC is a unifying framework for MCMC that encompasses many standard
-and recent MCMC algorithms from the literature, from reversible jump MCMC to
-Hamiltonian Monte Carlo to kernels based on deep neural networks. The key idea
-in involutive MCMC is that a combination of (i) an auxiliary probability
-distribution and (ii) an involution on the extended state space that includes
-the model latent variables and the auxiliary variables results in MCMC kernels
-that satisfy detailed balance with respect to the target distribution. The
-involutive MCMC framework is appealing for its simplicity and generality, and
-promises to be a useful conceptual and mathematical tool for exploring the
-design space of MCMC kernels. However, like other Monte Carlo samplers,
-implementing involutive samplers is time consuming and error prone. This paper
-describes a technique for automatically generating the implementation of
-involutive MCMC samplers from two probabilistic programs that define the target
-distribution and the auxiliary probability distribution respectively, and a
-differentiable program that defines the involution. The technique also detects
-common conceptual and programming errors that arise when designing and
-specifying involutive MCMC algorithms. This paper describes the support for
-involutive MCMC in the Gen probabilistic system that uses this technique, and
-shows involutive samplers specified using Gen's high-level probabilistic and
-differentiable programming languages.
+Involutive MCMC is a unifying mathematical construction for MCMC kernels that generalizes many classic and state-of-the-art MCMC algorithms, from reversible jump MCMC to kernels based on deep neural networks.
+But like MCMC samplers more generally, implementing involutive MCMC kernels is often tedious and error-prone, especially when sampling on complex state spaces.
+This paper describes a technique for automating the implementation of involutive MCMC kernels given (i) a pair of probabilistic programs defining the target distribution and an auxiliary distribution respectively and (ii) a differentiable program that transforms the execution traces of these probabilistic programs.
+The technique, which is implemented as part of the Gen probabilistic programming system, also automatically detects user errors in the specification of involutive MCMC kernels and exploits sparsity in the kernels for improved efficiency.
+The paper shows example Gen code for a split-merge reversible jump move in an infinite Gaussian mixture model and a state-dependent mixture of proposals on a combinatorial space of covariance functions for a Gaussian process.
 \end{abstract}

### 2020-07-16 paper.tex
Add some more clarification to dictionaries section

@@
 Involutive MCMC is a general framework for constructing MCMC kernels that are stationary
-for a target probability measure $p$. Informally, the algorithm works as follows: starting at some state $x$, we first sample
+for a target probability distribution $p$. Informally, the algorithm works as follows: starting at some state $x$, we first sample
 an auxiliary variable $y \sim q_x$ from a state-dependent auxiliary distribution. We then apply an \textit{involution} $f$ to the pair $(x, y)$ to obtain $(x', y')$. 
@@
 
-In this section, we derive the involutive MCMC algorithm for models $p$ and auxiliary kernels
+In this section, we present involutive MCMC for models $p$ and auxiliary kernels
 $q$ defined over general state spaces.
@@
 
-\subsection{Distributions on Dictionaries}
-
-Instead of vectors of real numbers, we consider in this section distributions over 
-\textit{finite dictionaries} that map (possibly random) keys to (possibly random) values. 
-Dictionaries include vectors as a special case (a vector $\mathbf{x} \in \mathbb{R}^n$ can
-be represented as a dictionary mapping the keys $1, \dots, n$ to the values $x_1, \dots, x_n$),
-but are in general more flexible: different keys can hold values of different types, and we
-can also consider distributions in which the set of keys is itself random, useful for model
-selection problems and structure uncertainty more generally.
-
-\textbf{The space of finite dictionaries.}
-We fix a countably infinite set $\mathcal{A}$ of possible keys, such that each key $a$ is either \emph{discrete} ($a \in \mathcal{I}$) or \emph{continuous} ($a \in \mathcal{J}$), where $\mathcal{A} = \mathcal{I} \cup \mathcal{J}$.\footnote{
-It is straightforward to assign a general measure space to each address, but this is not necessary for our purposes.}
-Let $V_a$ denote the set of possible values for key $a$, where $V_a$ is a countable set for each discrete key, and where $V_a = \mathbb{R}^{d_a}$ for each continuous key for some $d_a$.
-Given a set of keys $A$, let $V_A = \times_{a \in A} V_a$ denote the set of \textit{assignments} of values to each key.
-Then the set of all finite dictionaries is $\mathcal{D} := \bigcup_{A \subset \mathcal{A}, |A| < \infty} \{(A, \mathbf{x}) \mid \mathbf{x} \in V_A\}$. That is, a dictionary specifies 
-a finite set of keys $A \subset \mathcal{A}$ at which it has values, and an assignment $\mathbf{x}$ of values $x_a$ for each.
-
-\textbf{A measure space of finite dictionaries.}
-We associate a measure $\mu_a$ on $V_a$ for each key---the counting measure for each discrete key and the Lebesgue-measure on $\mathbf{R}^{d_a}$ for each continuous key.
-For each finite set of keys $A$, we make $V_A$ a measure space using the standard product $\sigma$-algebra $\Sigma_A = \otimes_{a \in A} \Sigma_a$ and the product measure $\mu_A = \times_{a \in A} \mu_a$.
-We equip $\mathcal{D}$ with the $\sigma$-algebra $\Sigma_{\mathcal{D}} := \{\bigcup_{A \subset \mathcal{A}, |A| < \infty} \{(A, \mathbf{x}) \mid \mathbf{x} \in B_A\} \mid B_A \in \Sigma_A \, \text{for each finite } A \subset \mathcal{A}\}$ to obtain a measurable space
-of dictionaries.
-A reference measure $\mu_\mathcal{D}$ on this space can be constructed using the product measures $\mu_A$: we set $\mu(B) := \sum_{A \in \mathcal{A}, |A| < \infty} \mu_A(\{\mathbf{x} \mid (A, \mathbf{x}) \in B\})$.
-
-\textbf{Notation for dictionaries.} Given a dictionary $d = (A, \mathbf{x})$, we write $A_d$ for $A$ and $d[a]$ for the value $x_a$ associated with a key $a \in A$.
-We also denote specific dictionaries using notation $\{k_1 \mapsto v_1, k_2 \mapsto v_2, \ldots\}$.
-For example, the dictionary $(A, \mathbf{x})$ with $A = \{1, \mathtt{"foo"}\}$ and $x_1 = 0.123$ and $x_{\mathtt{"foo"}} = 5$ is denoted 
-$\{1 \mapsto 0.123, \mathtt{"foo"} \mapsto 5\}$.
-
-\textbf{Probability distributions on finite dictionaries.} 
-When all keys are discrete, a probability distribution on dictionaries is defined by a $p$ is a probability mass function $p : \mathcal{D} \to [0, 1]$ that assigns a probability $p(m)$ to each dictionary $m \in \mathcal{P}$ such that $\sum_{x \in \mathcal{D}} p(m) = 1$.
-More generally a probability distribution in dictionaries is defined by a probability density $p : \mathcal{D} \rightarrow [0, \infty)$ such that $\int p(m) \mu_\mathcal{D}(\text{d}m) = 1$.
-We now give several examples
-to build intuition.
-
-% TODO keys vs addresses
-
-% TODO how to notate dictionaries -- 'm' or 'd' or 'x' or 't'?
-
-\textit{Gaussian mixture model with unknown number of components.} Consider a generative model of univariate data points $y_1, \dots, y_n$ from a Gaussian mixture with an unknown number of components $k$, each with unknown mean $m_i$ and variance $s_i$.
-If we place a Gaussian prior on $m_i$, an inverse Gamma prior on $s_i$, and a Poisson prior on $k$, the resulting density is
-%
-\begin{align*}
-p(d) =
-\begin{array}{l}
-p_{\mathrm{poisson}(3)}(d[\mathtt{k}]) \cdot\\
-\prod_{i=1}^{d[\mathtt{k}]} p_{\mathrm{normal}(0, 1)}(d[\mathtt{m}_i]) \cdot\\
-\prod_{i=1}^{d[\mathtt{k}]} p_{\mathrm{inversegamma}(1, 10)}(d[\mathtt{s}_i]) \cdot\\
-\prod_{i=1}^n  \frac{1}{d[\mathtt{k}]} \sum_{j=1}^{d[\mathtt{k}]} \cdot p_{\mathrm{normal}(d[\mathtt{m}_j], d[\mathtt{s}_i])}(d[\mathtt{y}_i])
-\end{array}
-\end{align*}
-%
-when $A_d$ is of the form $\{\texttt{k}, \texttt{y}_\texttt{1}, \dots, \texttt{y}_\texttt{n}\} \cup \{\texttt{m}_i \mid 1 \leq i \leq d[\texttt{k}]\} \cup \{\texttt{s}_i \mid 1 \leq i \leq d[\texttt{k}]\}$, and 0 otherwise.
-In this case, we have $V_\texttt{k} = \mathbb{N}$ with the counting measure for $\mu_\texttt{k}$, and for all other addresses $a \in \mathcal{A}$, $V_a = \mathbb{R}$ with the Lebesgue measure for $\mu_a$.
-
-\textit{Polynomial regression.}
-
-TODO
-
-\textbf{Conditional distributions via disintegration.} Consider a probability density $p$ on the space $\mathcal{D}$ of dictionaries. 
-%For a set of keys $A \subset \mathcal{A}$, let $p(A) := \int 1[A_m = A] p(m) \mu_\mathcal{D}(\text{d}m)$, the probability under $p$ of a dictionary with exactly the key set $A$. 
-We say a key $a \in \mathcal{A}$ \textit{almost always appears} if $\int \mathbf{1}[a \in A_m] p(m) \mu_\mathcal{D}(\text{d}m) = 1$.
-Suppose $B$ is a set of keys that almost always appear for $p$, and that $b = (B, \mathbf{b})$ is a dictionary with keys $B$.
-Furthermore, let $(A_m, \mathbf{m}) \oplus (A_n, \mathbf{n}) := (A_m \cup A_n, \mathbf{m} \times \mathbf{n})$ be the \textit{merge} of two dictionaries $m$ and $n$ defined on disjoint key sets $A_m$ and $A_n$.
-Then we can define the conditional density $p(d \mid b) := \mathbf{1}[A_d \cap B = \emptyset] \frac{p(d \oplus b)}{\int_{\{m \mid A_m \cap B = \emptyset\}} p(m \oplus b) \mu_\mathcal{D}(\text{d}m)}$ when the denominator is finite.
-If the reference measure $\mu_a$ associated with each key $a \in B$ is discrete (e.g., the counting measure), then this definition corresponds to the ordinary notion of conditioning on an event (namely, the event that a sample from $p$ agrees with the dictionary $b$ on all addresses in $B$).
-When this is not the case, it corresponds to a more general measure-theoretic notion called disintegration~\citep{disintegration1997}.
-
-\subsection{Involutive MCMC}
+\subsection{General Involutive MCMC}
 
@@
 Each iteration, the algorithm first samples auxiliary variables $y \in Y$ from an \emph{auxiliary distribution} $q_x$ based on the model's current state $x$: for each $x \in X$ such that $p(x) > 0$, $q_x : Y \to [0, \infty)$ is a probability density with respect to $\mu_Q$.
-% We call $p$ the \emph{model density} and we call $q_x$ \emph{auxiliary densities}.
 
@@
 $\Sigma := \{A \cap Z \mid A \in \Sigma_P \otimes \Sigma_Q\}$ (assuming $Z$ is a $\mu_P \times \mu_Q$-measurable set), and a reference measure $\mu(A) := (\mu_P \times \mu_Q)(A)$.
-%Let $\pi(x, y) := p(x) q_x(y)$ and $Z := \{(x, y) \in X \times Y : \pi(x, y) > 0 \}$.
-%Suppose that $Z \in \Sigma_P \otimes \Sigma_Q$ ($Z$ is a $\mu_P \times \mu_Q$-measurable set).
-%Then $\Sigma := \{A \in \Sigma_P \otimes \Sigma_Q : A \subseteq Z\}$ is a $\sigma$-algebra on $Z$ and $\mu : \Sigma \to [0, \infty)$ with $\mu(A) := (\mu_P \times \mu_Q)(A)$ is a $\sigma$-finite measure on measurable space $(Z, \Sigma)$.
 
@@
 With probability $\alpha$, the new state $x'$ is returned; otherwise, the previous state $x$ is repeated.
-%Let $y \sim q_x(\cdot)$ denote a random sample from the measure defined by $A \mapsto \int_A q_x(y) \mu_Q(dy)$ for $A \in \Sigma_Q$.
+
 \begin{algorithm}[h]
@@
 
+\subsection{Probability Distributions on Dictionaries}
+
+While maximally general, the measure-theoretic formulation of involutive MCMC in Algorithm~\ref{alg:involutive-mcmc} is not amenable to an automated implementation, because it does not indicate how to compute the Radon-Nikodym derivative that is required for the acceptance probability, and it is unclear how to specify the probability measures involved.
+%When the algorithm is specialized for state spaces consisting of dictionaries, the acceptance probability simplifies and can be computed from standard Jacobian and probability density computations.
+%Because dictionaries are flexible objects, the algorithm remains quite general.
+
+While restricting the state space to vectors of real numbers would address these issues, we seek a 
+representation that remains flexible enough to represent complex hybrid state spaces with numeric, symbolic, and structure uncertainty.
+Therefore, we use state spaces consisting of 
+\textit{finite dictionaries} that map (possibly random) keys to (possibly random) values. 
+Dictionaries include vectors as a special case (a vector $\mathbf{x} \in \mathbb{R}^n$ can
+be represented as a dictionary mapping the keys $1, \dots, n$ to the values $x_1, \dots, x_n$),
+but are more flexible: different keys can hold values of different types (e.g. integers, strings), and we
+can also consider distributions in which the set of keys is itself random, which is useful for model
+selection problems and structure uncertainty more generally.
+
+This section describes probability distributions on dictionaries, and gives a constructive definition of the involutive MCMC acceptance probability in this setting in terms of a Jacobian.
+Section~\ref{sec:automating} will then show how probability distributions on dictionaries can be specified with probabilistic programs, and how probabilistic programming techniques can automatically compute probability densities spaces of dictionaries.
+
+\textbf{The space of finite dictionaries.}
+We fix a countably infinite set $\mathcal{A}$ of possible keys, such that each key $a$ is either called \emph{discrete} ($a \in \mathcal{I}$) or \emph{continuous} ($a \in \mathcal{J}$), where $\mathcal{A} = \mathcal{I} \cup \mathcal{J}$.\footnote{
+It is possible to assign a general measure space to each address, but this is not necessary for our purposes.}
+Let $V_a$ denote the set of possible values for key $a$, where $V_a$ is a countable set for each discrete key, and where $V_a = \mathbb{R}^{d_a}$ for each continuous key for some $d_a$.
+Given a set of keys $A$, let $V_A = \times_{a \in A} V_a$ denote the set of \textit{assignments} of values to each key.
+Then the set of all finite dictionaries is $\mathcal{D} := \bigcup_{A \subset \mathcal{A}, |A| < \infty} \{(A, \mathbf{x}) \mid \mathbf{x} \in V_A\}$. That is, a dictionary specifies 
+a finite set of keys $A \subset \mathcal{A}$ at which it has values, and an assignment $\mathbf{x}$ of values $x_a$ for each.
+
+\textbf{Relationship to representation of~\citet{green1995reversible}.}
+\citet{green1995reversible} uses a state space that is the countable union of `models', where for each model there is a Eucliean space of parameters.
+Dictionaries possesses substantially more structure:
+Parameters have keys, so it is possible to express that parameters are shared across different structures.
+Instead of a monolithic `model', dictionaries use a more elaborate discrete state that includes the set of keys and the assignment to the discrete keys.
+The additional structure of dictionaries enables the automation technique Section~\ref{sec:automating}.
+
+\textbf{A measure space of finite dictionaries.}
+We associate a measure $\mu_a$ on $V_a$ for each key---the counting measure for each discrete key and the Lebesgue-measure on $\mathbb{R}^{d_a}$ for each continuous key.
+For each finite set of keys $A$, we make $V_A$ a measure space using the standard product $\sigma$-algebra $\Sigma_A = \otimes_{a \in A} \Sigma_a$ and the product measure $\mu_A = \times_{a \in A} \mu_a$.
+We equip $\mathcal{D}$ with the $\sigma$-algebra $\Sigma_{\mathcal{D}} := \{\bigcup_{A \subset \mathcal{A}, |A| < \infty} \{(A, \mathbf{x}) \mid \mathbf{x} \in B_A\} \mid B_A \in \Sigma_A \, \text{for each finite } A \subset \mathcal{A}\}$ to obtain a measurable space
+of dictionaries.
+A reference measure $\mu_\mathcal{D}$ on this space can be constructed using the product measures $\mu_A$: we set $\mu(B) := \sum_{A \in \mathcal{A}, |A| < \infty} \mu_A(\{\mathbf{x} \mid (A, \mathbf{x}) \in B\})$.
+
+\textbf{Notation for dictionaries.} Given a dictionary $d = (A, \mathbf{x})$, we write $A_d$ for $A$ and $d[a]$ for the value $x_a$ associated with a key $a \in A$.
+We also denote specific dictionaries using notation $\{k_1 \mapsto v_1, k_2 \mapsto v_2, \ldots\}$.
+For example, the dictionary $(A, \mathbf{x})$ with $A = \{1, \mathtt{"foo"}\}$ and $x_1 = 0.123$ and $x_{\mathtt{"foo"}} = 5$ is denoted 
+$\{1 \mapsto 0.123, \mathtt{"foo"} \mapsto 5\}$.
+
+\textbf{Probability distributions on finite dictionaries.} 
+When all keys are discrete, a probability distribution on dictionaries is defined by a $p$ is a probability mass function $p : \mathcal{D} \to [0, 1]$ that assigns a probability $p(m)$ to each dictionary $m \in \mathcal{P}$ such that $\sum_{x \in \mathcal{D}} p(m) = 1$.
+More generally a probability distribution on dictionaries is defined by a probability density $p : \mathcal{D} \rightarrow [0, \infty)$ such that $\int p(m) \mu_\mathcal{D}(\text{d}m) = 1$. 
+The probability is distributed the finite sets of keys $A \subseteq \mathcal{A}$ where $|A| < \infty$:
+\begin{equation}
+1 = \int p(m) \mu_\mathcal{D}(\text{d}m) = \sum_{\substack{A \subseteq \mathcal{A}\\|A| < \infty}} \int p(m) \mu_{A}(dm)
+\end{equation}
+where the probability for each key set $A$ is:
+\begin{equation}\label{eq:integral}
+\int p(m) \mu_{A}(dm) = \sum_{\mathbf{x}_1} \int_{\mathbb{R}^{d}} p((A, (\mathbf{x}_1, \mathbf{x}_2))) d\mathbf{x}_2
+\end{equation}
+where $d: = \sum_{a \in A \cap \mathcal{J}} d_{a}$ and where $\mathbf{x}_1$ is an assignment to the discrete choices in $A$ and $\mathbf{x}_2$ is an assignment to the continuous choices.
+
+We now give an example to build intuition.
+Consider a generative model of univariate data points $y_1, \dots, y_n$ from a Gaussian mixture with an unknown number of components $k$, each with unknown mean $m_i$ and variance $s_i$.
+If we place a Gaussian prior on $m_i$, an inverse Gamma prior on $s_i$, and a Poisson prior on $k$, the resulting density on dictionaries $d$ is:
+%
+\begin{align*}
+p(d) =
+\begin{array}{l}
+p_{\mathrm{poisson}(3)}(d[\mathtt{k}]) \cdot\\
+\prod_{i=1}^{d[\mathtt{k}]} p_{\mathrm{normal}(0, 1)}(d[\mathtt{m}_i]) \cdot\\
+\prod_{i=1}^{d[\mathtt{k}]} p_{\mathrm{inversegamma}(1, 10)}(d[\mathtt{s}_i]) \cdot\\
+\prod_{i=1}^n  \frac{1}{d[\mathtt{k}]} \sum_{j=1}^{d[\mathtt{k}]} \cdot p_{\mathrm{normal}(d[\mathtt{m}_j], d[\mathtt{s}_i])}(d[\mathtt{y}_i])
+\end{array}
+\end{align*}
+%
+when $A_d = \{\texttt{k}, \texttt{y}_\texttt{1}, \dots, \texttt{y}_\texttt{n}\} \cup \{\texttt{m}_i \mid 1 \leq i \leq d[\texttt{k}]\} \cup \{\texttt{s}_i \mid 1 \leq i \leq d[\texttt{k}]\}$, and 0 otherwise.
+In this case, we have $V_\texttt{k} = \mathbb{N}$ with the counting measure for $\mu_\texttt{k}$, and for all other addresses $a \in \mathcal{A}$, $V_a = \mathbb{R}$ with the Lebesgue measure for $\mu_a$.
+For each $k \in \{0, 1, \ldots\}$, the probability for key set
+$\{\texttt{k}, \texttt{y}_\texttt{1}, \dots, \texttt{y}_\texttt{n}\} \cup \{\texttt{m}_i \mid 1 \leq i \leq k\} \cup \{\texttt{s}_i \mid 1 \leq i \leq k\}$
+is $p_{\mathrm{poisson}(s)}(k)$ (via Equation~(\ref{eq:integral})).
+
+\textbf{Conditional distributions via disintegration.} Consider a probability density $p$ on the space $\mathcal{D}$ of dictionaries. 
+%For a set of keys $A \subset \mathcal{A}$, let $p(A) := \int 1[A_m = A] p(m) \mu_\mathcal{D}(\text{d}m)$, the probability under $p$ of a dictionary with exactly the key set $A$. 
+We say a key $a \in \mathcal{A}$ \textit{almost always appears} if $\int \mathbf{1}[a \in A_m] p(m) \mu_\mathcal{D}(\text{d}m) = 1$.
+Suppose $B$ is a set of keys that almost always appear for $p$, and that $b = (B, \mathbf{b})$ is a dictionary with keys $B$.
+Furthermore, let $(A_m, \mathbf{m}) \oplus (A_n, \mathbf{n}) := (A_m \cup A_n, \mathbf{m} \times \mathbf{n})$ be the \textit{merge} of two dictionaries $m$ and $n$ defined on disjoint key sets $A_m$ and $A_n$.
+Then we can define the conditional density $p(d \mid b) := \mathbf{1}[A_d \cap B = \emptyset] \frac{p(d \oplus b)}{\int_{\{m \mid A_m \cap B = \emptyset\}} p(m \oplus b) \mu_\mathcal{D}(\text{d}m)}$ when the denominator is finite.
+If the reference measure $\mu_a$ associated with each key $a \in B$ is discrete (e.g., the counting measure), then this definition corresponds to the ordinary notion of conditioning on an event (namely, the event that a sample from $p$ agrees with the dictionary $b$ on all addresses in $B$).
+When this is not the case, it corresponds to a more general measure-theoretic notion called disintegration~\citep{chang1997conditioning}.
+
+Consider the infinite univariate mixture model, and the conditional density given observed data $(B, \mathbf{b}) := \{\mathtt{y}_1 \mapsto y_1, \ldots, \mathtt{y}_{\mathtt{n}} \mapsto y_{\mathtt{n}}\}$.
+The conditional density $p(d | b)$ is nonzero only if $A_d = \{\texttt{k}\} \cup \{\texttt{m}_i \mid 1 \leq i \leq k\} \cup \{\texttt{s}_i \mid 1 \leq i \leq k\}$ for some $k$ ($d$ does not contain y-values), and the denominator in the definition of $p(d | b)$ simplifies to the familiar sum of marginal likelihoods over all $k$, where each marginal likelihood is an integral over $\mathbb{R}^{2k}$.
+
 \subsection{Involutive MCMC with Dictionaries}
 
-The general measure-theoretic formulation of involutive MCMC in Algorithm~\ref{alg:involutive-mcmc} does not specify how to compute the Radon-Nikodym derivative that is required for the acceptance probability.
-When the algorithm is specialized for state spaces consisting of dictionaries, the acceptance probability simplifies and can be computed from standard Jacobian and probability density computations.
-Because dictionaries are flexible objects, the algorithm remains quite general.
-
-Suppose that the model distribution $p$ and auxiliary distributions $q_x$ are all probability distributions on dictionaries.
+%The general measure-theoretic formulation of involutive MCMC in Algorithm~\ref{alg:involutive-mcmc} does not specify how to compute the Radon-Nikodym derivative that is required for the acceptance probability.
+%When the algorithm is specialized for state spaces consisting of dictionaries, the acceptance probability simplifies and can be computed from standard Jacobian and probability density computations.
+%Because dictionaries are flexible objects, the algorithm remains quite general.
+
+Suppose that the model distribution and auxiliary distributions are probability distributions on dictionaries, with densities $p$ and $q_x$.
 Then, $X$ and $Y$ are both sets of dictionaries, and the joint space $Z$ is a set of pairs $(x, y)$ of dictionaries with keys $\mathcal{A}$, so that $Z = X \times Y \subseteq \mathcal{D} \times \mathcal{D}$.

### 2020-07-17 paper.tex
clean up automated involutive mccm algorihtm box

@@
 \textbf{Relationship to representation of~\citet{green1995reversible}.}
-\citet{green1995reversible} uses a state space that is the countable union of `models', where for each model there is a Eucliean space of parameters.
+\citet{green1995reversible} uses a state space that is the countable union of `models', where for each model there is a vector of real-valued parameters.
 Dictionaries possesses substantially more structure:
-Parameters have keys, so it is possible to express that parameters are shared across different structures.
-Instead of a monolithic `model', dictionaries use a more elaborate discrete state that includes the set of keys and the assignment to the discrete keys.
-The additional structure of dictionaries enables the automation technique Section~\ref{sec:automating}.
+Instead of monolithic `models', dictionaries use a more elaborate discrete state that includes the set of keys and the assignment to the discrete keys.
+Also, continuous keys play the role of real-valued parameters, so it is possible to express that a given real-valued parameter is shared between models.
+The additional structure of dictionaries enables the automation techniques in Section~\ref{sec:automating}.
 
@@
 Consider the infinite univariate mixture model, and the conditional density given observed data $(B, \mathbf{b}) := \{\mathtt{y}_1 \mapsto y_1, \ldots, \mathtt{y}_{\mathtt{n}} \mapsto y_{\mathtt{n}}\}$.
-The conditional density $p(d | b)$ is nonzero only if $A_d = \{\texttt{k}\} \cup \{\texttt{m}_i \mid 1 \leq i \leq k\} \cup \{\texttt{s}_i \mid 1 \leq i \leq k\}$ for some $k$ ($d$ does not contain y-values), and the denominator in the definition of $p(d | b)$ simplifies to the familiar sum of marginal likelihoods over all $k$, where each marginal likelihood is an integral over $\mathbb{R}^{2k}$.
+The conditional density $p(d | b)$ is nonzero only if $A_d = \{\texttt{k}\} \cup \{\texttt{m}_i \mid 1 \leq i \leq k\} \cup \{\texttt{s}_i \mid 1 \leq i \leq k\}$ for some $k$ ($d$ does not contain y-values), and the denominator in the definition of $p(d | b)$ simplifies to the familiar sum of marginal likelihoods over all $k$, where each marginal likelihood is a Riemann integral over $\mathbb{R}^{2k}$.
 
@@
 Suppose that the model distribution and auxiliary distributions are probability distributions on dictionaries, with densities $p$ and $q_x$.
-Then, $X$ and $Y$ are both sets of dictionaries, and the joint space $Z$ is a set of pairs $(x, y)$ of dictionaries with keys $\mathcal{A}$, so that $Z = X \times Y \subseteq \mathcal{D} \times \mathcal{D}$.
-%And suppose that there exists an involution $f : Z \to Z$ such that (i) 
-Suppose there is a countable partition of $Z$ into $\{Z_e : e \in E\}$ such that across each $Z_e$ the sets of keys $A_x$ and $A_y$ and the values $x_a$ and $y_a$ for all discrete keys $a$ are both constant (so that each set $Z_e$ is isomorphic to a Euclidean space of assignments to the continuous keys in the two dictionaries).
+Then, $X$ and $Y$ are both sets of dictionaries, and the joint space $Z$ is a set of pairs $(x, y)$ of dictionaries with keys $\mathcal{A}_P$ and $\mathcal{A}_Q$ respectively, so that $Z = X \times Y \subseteq \mathcal{D}_P \times \mathcal{D}_Q$ where $\mathcal{D}_P$ is the set of dictionaries on keys taken from $\mathcal{A}_P$ and similarly for $\mathcal{D}_Q$.
+% TODO: this is only needed for Algorithm 2, and it is relevant for the 
+To simplify the notation, and without loss of generality, we will assume that $\mathcal{A}_P$ and $\mathcal{A}_Q$ are disjoint\footnote{If $\mathcal{A}_P$ and $\mathcal{A}_Q$ are not disjoint then, they can be made so by adding a different prefix to each address.}, and we define $Z := \{ x \oplus y : p(x)q_x(y) > 0\} \subseteq \mathcal{D}$.
+where $\mathcal{D}$ is the set of dictionaries on keys from $\mathcal{A}_P \cup \mathcal{A}_Q$
+(recall $x \oplus y$ denotes the dictionary resulting from merging dictionaries $x$ and $y$ with disjoint keys).
+
+Suppose there is a countable partition of $Z$ into $\{Z_e : e \in E\}$ such that across each $Z_e$ the sets of keys $A_x$ and $A_y$ and the values $x_a$ and $y_a$ for all discrete keys $a$ are both constant.
+Then each set $Z_e$ is isomorphic to a Euclidean space of assignments to the continuous keys in the two dictionaries.
 Suppose there is an involution $g : E \to E$ between elements of the partition (a function with $g = g^{-1}$), and a family of continuously differentiable bijections $h_{e} : Z_e \to Z_{g(e)}$ for each $e$ with $h_{e} = h_{g(e)}^{-1}$.
@@
 Then, $f : Z \to Z$ given by $f(z) := h_{e(z)}(z)$ is an involution:
-\begin{equation}
+\begin{equation*}
 f(f(z)) = h_{e(f(z))}(h_{e(z)}(z)) = h_{g(e(z))}(h_{e(z)}(z)) = z
-\end{equation}
+\end{equation*}
 Let $|J h_e|(z)$ denote the absolute value of the Jacobian (determinant) of $h_e$, evaluated at $z$.
-Then, the acceptance probability simplifies to:
-\begin{equation}
-\alpha(x, y) = \frac{p(x') q_{x'}(y')}{p(x) q_{x}(y)} \cdot |J h_e|(z)
-\end{equation}
+Then, the acceptance probability in Algorithm~\ref{alg:involutive-mcmc} simplifies to:
+\begin{equation}
+\frac{p(x') q_{x'}(y')}{p(x) q_{x}(y)} \cdot |J h_e|(z)
+\end{equation}
+One example of a valid partition of $Z$ is given by equivalence classes of the following equivalence relation:
+\[
+z_1 \sim z_2 \iff (A_{z_1} = A_{z_2}) \land (z_1[a] = z_2[a] \; \forall a \in A_{z_1} \cap \mathcal{I})
+\]
+(dictionaries are equivalent if they contain the same keys and they agree on the value of all discrete keys).
 Section~\ref{sec:radon-nikodym-special-case} of the appendix for details.
@@
 \begin{algorithmic}
+\Require probabilistic programs $\mathcal{P}$ and $\mathcal{Q}$; differentiable program $\mathcal{F}$; initial state $x$
+\vspace{1mm}
 \Procedure{auto-involutive-mcmc}{$\mathcal{P}$, $\mathcal{Q}$, $\mathcal{F}$, $x$}
-    \State $(y, \log w_2) \gets \textproc{trace-and-score}(\mathcal{Q}_x)$ \Comment{Sample $y \sim q_x(\cdot)$ and compute log-density $\log q_x(y)$}
-    \State $(x', y', \log D) \gets \textproc{run-involution}(\mathcal{F}, (x, y))$ \Comment{Compute $(x', y') = f(x, y)$, log Radon-Nikodym deriv. $\log D$}
-    \State $\log w_1 \gets \textproc{score}(\mathcal{P}, x)$ \Comment{Compute log-density $\log p(x)$}
-    \State $\log w_3 \gets \textproc{score}(\mathcal{P}, x')$ \Comment{Compute log-density $\log p(x')$}
-    \State $\log w_4 \gets \textproc{score}(\mathcal{Q}_{x'}, y')$ \Comment{Compute log-density $\log q_{x'}(y')$}
-    \State $\alpha \gets \min\{1, \exp(\log w_3 - \log w_1 + \log w_4 - \log w_2 + \log J)\}$ \Comment{Compute acceptance probability}
+    \State $(y, \log q_x(y)) \gets \textproc{trace-and-score}(\mathcal{Q}_x)$ \Comment{Sample $y \sim q_x(\cdot)$; compute log-density via probabilistic program}
+    \State $(x' \oplus y', \log D) \gets \textproc{run-involution}(\mathcal{F}, x \oplus y)$ \Comment{{\small Compute $(x' \oplus y') = f(x \oplus y)$ and log(Radon-Nikodym derivative)}}
+    \State $\log p(x) \gets \textproc{score}(\mathcal{P}, x)$ \Comment{Compute log-density via probabilistic program}
+    \State $\log p(x') \gets \textproc{score}(\mathcal{P}, x')$ \Comment{Compute log-density via probabilistic program}
+    \State $\log q_{x'}(y') \gets \textproc{score}(\mathcal{Q}_{x'}, y')$ \Comment{Compute log-density via probabilistic program}
+    \State $\alpha \gets \min\{1, \exp(\log p(x') - \log p(x) + \log q_{x'}(y') - \log q_x(y) + \log D)\}$ \Comment{Compute acceptance probability}
     \State \textbf{with probability} \,$\alpha$ \,\Return $x'$ \algorithmicelse \, \Return $x$ 
 \EndProcedure\\
-
+\vspace{1mm}
 \begin{minipage}[t]{0.5\linewidth}
@@
 \end{minipage}\\
-
-\Procedure{run-involution}{$\mathcal{F}$, $(x, y)$}
+\vspace{1mm}
+\Procedure{run-involution}{$\mathcal{F}$, $z$}
 % TODO would be much simpler to describe if we didn't have to separate out x and y..
-    \LineComment{Run the involution, and keep track of continuous addresses}
-    \State $x' \gets \{\}$ \Comment{Initialize empty output dictionary}
+    \LineComment{Run the involution, and keep track of copied, read, and written continuous keys}
+    \State $z' \gets \{\}$ \Comment{Initialize empty output dictionary}
     \execute[$\mathcal{F}$], but with
-        \State $\mathtt{"write\_cont}(a, v)\mathtt{"} \equiv$ (1. set $x'[a] \gets v$ and 2. set $W \gets W \cup \{a\}$)
-        \State $\mathtt{"read\_cont}(a)\mathtt{"} \equiv$ (1. set $R \gets R \cup \{a\}$ and 2. evaluate to $x[a]$)
-        \State $\mathtt{"copy}(a_1, a_2)\mathtt{"} \equiv$ (1. set $x'[a_2] \gets x[a_1]$ and 2. set $C \gets C \cup \{a_2\}$)
+        \State $\mathtt{"write\_disc}(a, v)\mathtt{"} \equiv$ (set $z'[a] \gets v$)
+        \State $\mathtt{"read\_disc}(a)\mathtt{"} \equiv$ (evaluate to $z[a]$)
+        \State $\mathtt{"write\_cont}(a, v)\mathtt{"} \equiv$ (1. set $z'[a] \gets v$ and 2. set $W \gets W \cup \{a\}$)
+        \State $\mathtt{"read\_cont}(a)\mathtt{"} \equiv$ (1. set $R \gets R \cup \{a\}$ and 2. evaluate to $z[a]$)
+        \State $\mathtt{"copy}(a_1, a_2)\mathtt{"} \equiv$ (1. set $z'[a_2] \gets z[a_1]$ and 2. set $C \gets C \cup \{a_2\}$)
     \END
-    \LineComment{Use automatic differentiation to compute Jacobian, skipping copied addresses}
+    \LineComment{Use automatic differentiation of $\mathcal{F}$ to compute Jacobian, skipping copied addresses}
     \For{$i$ in $1, \ldots, |W \setminus C|$}
-        \State $J[i,:] \gets \nabla_{x_{R \setminus C}} \left( h_{e(x, y)}[(W \setminus C)_i] \right)$ \Comment{Gradient of $x'[(W \setminus C)_i]$ w.r.t. non-copied continuous inputs}
+        \State $a_i \gets (W \setminus C)_i$ \Comment{Pick $i$th key in $(W \setminus C)$; the order does not matter}
+        \State $J[i,:] \gets \nabla_{\mathbf{z}_{R \setminus C}} \left( h_{e(z)}[a_i] \right)$ \Comment{{\small Gradient of $z'[a_i] = h_{e(z)}[a_i]$ wrt non-copied continuous inputs ($z_a$ for $a \in R \setminus C$)}}
     \EndFor
-    \State \Return $(x', \log |\det(J)|)$
+    \State \Return $(z', \log(|\det(J)|))$
 \EndProcedure
-
 \end{algorithmic}
@@
 Therefore, the involution is a map from pairs of traces to pairs of traces.
-%To simplify the notation, we can represent a pair of traces $(x, y)$ as a single trace $z$, in which each address in $x$ and $x'$ is prefixed with some unique identifier (e.g.
-\texttt{model}) and each address in $y$ is prefixed with some unique identifier $y'$ (e.g.
-\texttt{auxiliary}).
 Then, $Z$ is a set of such trace pairs, and the involution $f$ is a function from $Z$ onto itself that is its own invverse.

### 2020-07-17 experiments/pytorch/minigen.py
change syntax of pytorch impl

@@ -37,35 +37,59 @@ InvolutionRunnerState = namedtuple("InvolutionRunnerState",
         "input_cont_tensors", "output_cont_tensors"
     ])
 
-def read_model_cont(state, addr):
-    val = torch.tensor(state.input_model_trace[addr], requires_grad=True)
-    state.input_cont_tensors.append(val)
-    return val
+# model and address identifiers
+MODEL = "model"
+AUX = "aux"
 
-def read_model_disc(state, addr):
-    return state.input_model_trace[addr]
+def check_which(which):
+    if which != MODEL and which != AUX:
+        raise Exception("address argument must be (MODEL, *) or (AUX, *)")
 
-def write_model_cont(state, addr, val):
-    state.output_model_trace[addr] =val 
-    state.output_cont_tensors.append(val)
+# user-provided type information for reads and writes
+CONTINUOUS = "continuous"
+DISCRETE = "discrete"
 
-def write_model_disc(state, addr, val):
-    state.output_model_trace[addr] = val
+def check_type_label(type_label):
+    if type_label != CONTINUOUS and type_label != DISCRETE:
+        raise Exception("type label argument must be CONTINUOUS or DISCRETE")
 
-def read_aux_cont(state, addr):
-    val = torch.tensor(state.input_aux_trace[addr], requires_grad=True)
-    state.input_cont_tensors.append(val)
-    return val
+def read(state, addr, type_label):
+    (which, addr) = addr
+    check_which(which)
+    check_type_label(type_label)
+    if type_label == CONTINUOUS:
+        if which == MODEL:
+            val = torch.tensor(state.input_model_trace[addr], requires_grad=True)
+        else:
+            val = torch.tensor(state.input_aux_trace[addr], requires_grad=True)
+        state.input_cont_tensors.append(val)
+        return val
+    else:
+        if which == MODEL:
+            return state.input_model_trace[addr]
+        else:
+            return state.input_aux_trace[addr]
 
-def read_aux_disc(state, addr):
-    return state.input_aux_trace[addr]
+def write(state, addr, val, type_label):
+    (which, addr) = addr
+    check_which(which)
+    check_type_label(type_label)
+    if type_label == CONTINUOUS:
+        if which == MODEL:
+            state.output_model_trace[addr] = val 
+        else:
[…]

### 2020-07-17 experiments/pytorch/minigen.py
add copy with smarter jacobian and involution check to pytorch impl

@@ -34,16 +34,22 @@ def logpdf(trace, p, *args):
 InvolutionRunnerState = namedtuple("InvolutionRunnerState",
     [   "input_model_trace", "input_aux_trace",
         "output_model_trace", "output_aux_trace",
-        "input_cont_tensors", "output_cont_tensors"
+        "input_cont_tensors", "output_cont_tensors",
+        "input_copied_addrs"
     ])
 
 # model and address identifiers
 MODEL = "model"
 AUX = "aux"
 
-def check_which(which):
+def get_which(addr):
+    exc = Exception("address argument must be (MODEL, *) or (AUX, *)")
+    if len(addr) != 2:
+        raise exc
+    (which, addr) = addr
     if which != MODEL and which != AUX:
-        raise Exception("address argument must be (MODEL, *) or (AUX, *)")
+        raise exc
+    return (which, addr)
 
 # user-provided type information for reads and writes
 CONTINUOUS = "continuous"
@@ -54,15 +60,14 @@ def check_type_label(type_label):
         raise Exception("type label argument must be CONTINUOUS or DISCRETE")
 
 def read(state, addr, type_label):
-    (which, addr) = addr
-    check_which(which)
+    (which, addr) = get_which(addr)
     check_type_label(type_label)
     if type_label == CONTINUOUS:
         if which == MODEL:
             val = torch.tensor(state.input_model_trace[addr], requires_grad=True)
         else:
             val = torch.tensor(state.input_aux_trace[addr], requires_grad=True)
-        state.input_cont_tensors.append(val)
+        state.input_cont_tensors[(which, addr)] = val
         return val
     else:
         if which == MODEL:
@@ -71,8 +76,7 @@ def read(state, addr, type_label):
             return state.input_aux_trace[addr]
 
 def write(state, addr, val, type_label):
-    (which, addr) = addr
-    check_which(which)
+    (which, addr) = get_which(addr)
     check_type_label(type_label)
     if type_label == CONTINUOUS:
         if which == MODEL:
@@ -86,23 +90,56 @@ def write(state, addr, val, type_label):
         else:
             state.output_aux_trace[addr] = val
 
-#def copy_model_to_model(state, addr1, addr2):
-#def copy_model_to_aux(state, addr1, addr2):
-#def copy_aux_to_model(state, addr1, addr2):
-#def copy_aux_to_aux(state, addr1, addr2):
-
[…]

### 2020-07-17 experiments/pytorch/minigen.py
clean up syntax of minipygen languages

@@ -2,28 +2,50 @@ import torch
 from collections import namedtuple
 from itertools import chain
 
+# Helper: inject variables into a function's scope:
+from functools import wraps
+def inject_variables(context, func):
+    @wraps(func)
+    def new_func(*args, **kwargs):
+        func_globals = func.__globals__
+        saved_values = func_globals.copy()
+        func_globals.update(context)
+        try:
+            result = func(*args, **kwargs)
+        finally:
+            for (var, val) in context.items():
+                if var in saved_values:
+                    func_globals.update({var: saved_values[var]})
+                else:
+                    del func_globals[var]
+        return result
+    return new_func
 
 ##############################################
 # minimal probabilistic programming language #
 ##############################################
 
-def sample(p, *args):
+def trace_and_score(p, *args):
     trace = {}
-    def trace_choice(dist, addr):
+    lpdf = torch.tensor(0.0)
+    def sample(dist, addr):
         val = dist.sample()
         trace[addr] = val
+        lpdf += dist.log_prob(val)
         return val
-    p(trace_choice, *args)
-    return trace
+    p = inject_variables({"sample" : sample}, p)
+    p(*args)
+    return (trace, lpdf)
 
-def logpdf(trace, p, *args):
+def score(trace, p, *args):
     lpdf = torch.tensor(0.0)
-    def trace_choice(dist, addr):
+    def sample(dist, addr):
         nonlocal lpdf
         val = trace[addr]
         lpdf += dist.log_prob(val)
         return val
-    p(trace_choice, *args)
+    p = inject_variables({"sample" : sample}, p)
+    p(*args)
     return lpdf
 
 
@@ -59,7 +81,7 @@ def check_type_label(type_label):
     if type_label != CONTINUOUS and type_label != DISCRETE:
         raise Exception("type label argument must be CONTINUOUS or DISCRETE")
 
-def read(state, addr, type_label):
+def _read(state, addr, type_label):
     (which, addr) = get_which(addr)
     check_type_label(type_label)
     if type_label == CONTINUOUS:
@@ -75,7 +97,7 @@ def read(state, addr, type_label):
[…]

### 2020-07-17 paper.tex
improve automated involutive mcmc alg box

@@
 \vspace{1mm}
-\begin{minipage}[t]{0.5\linewidth}
+\begin{minipage}[t]{0.49\linewidth}
 \Procedure{trace-and-score}{$\mathcal{P}$}
@@
    \State $x \gets \{\}$
-   \execute[$\mathcal{P}$], but with $\mathtt{"a \sim d"} \equiv$ (
-        \State \hspace{5mm} 1. set $v \sim d$
-        \State \hspace{5mm} 2. set $x[a] \gets v$
-        \State \hspace{5mm} 3. set $\log w \gets \log w + \textproc{logpdf}(d, v)$
-        \State \hspace{5mm} 4. evaluate to $v$)
+   \execute[$\mathcal{P}$], but with $\mathtt{"}a \sim \mathrm{distribution}\mathtt{"} \equiv$ (
+        \State 1. set $v \sim \mathrm{distribution}$
+        \State 2. set $x[a] \gets v$
+        \State 3. set $\log w \gets \log w + \textproc{logpdf}(\mathrm{distribution}, v)$
+        \State 4. evaluate to $v$)
     \END
@@
 \end{minipage}%
-\begin{minipage}[t]{0.5\linewidth}
+\hfill
+\begin{minipage}[t]{0.49\linewidth}
 \Procedure{score}{$\mathcal{P}$, $x$}
    \State $\log w \gets 0$
-   \execute[$\mathcal{P}$], but with $\mathtt{"a \sim d"} \equiv$ (
-        \State \hspace{5mm} 1. set $v \gets x[a]$
-        \State \hspace{5mm} 2. set $\log w \gets \log w + \textproc{logpdf}(d, v)$
-        \State \hspace{5mm} 3. evaluate to $v$)
+   \execute[$\mathcal{P}$], but with $\mathtt{"}a \sim \mathrm{distribution}\mathtt{"} \equiv$ (
+        \State 1. set $v \gets x[a]$
+        \State 2. set $\log w \gets \log w + \textproc{logpdf}(\mathrm{distribution}, v)$
+        \State 3. evaluate to $v$)
     \END
@@
     \execute[$\mathcal{F}$], but with
-        \State $\mathtt{"write\_disc}(a, v)\mathtt{"} \equiv$ (set $z'[a] \gets v$)
-        \State $\mathtt{"read\_disc}(a)\mathtt{"} \equiv$ (evaluate to $z[a]$)
-        \State $\mathtt{"write\_cont}(a, v)\mathtt{"} \equiv$ (1. set $z'[a] \gets v$ and 2. set $W \gets W \cup \{a\}$)
-        \State $\mathtt{"read\_cont}(a)\mathtt{"} \equiv$ (1. set $R \gets R \cup \{a\}$ and 2. evaluate to $z[a]$)
-        \State $\mathtt{"copy}(a_1, a_2)\mathtt{"} \equiv$ (1. set $z'[a_2] \gets z[a_1]$ and 2. set $C \gets C \cup \{a_2\}$)
+        \State $\mathtt{"write}(a, v)\mathtt{"} \equiv$
+            (\algorithmicif\hspace{1mm} $a \in \mathcal{I}$
+            \algorithmicthen\hspace{1mm} set $z'[a] \gets v$
+            \algorithmicelse\hspace{1mm} set $z'[a] \gets v$ and set $W \gets W \cup \{a\})$
+        \State $\mathtt{"read}(a)\mathtt{"} \equiv$
+            (\algorithmicif\hspace{1mm} $a \in \mathcal{I}$
+            \algorithmicthen\hspace{1mm} evaluate to $z[a]$
+            \algorithmicelse\hspace{1mm} set $R \gets R \cup \{a\}$ and evaluate to $z[a]$)
+        \State $\mathtt{"copy}(a_1, a_2)\mathtt{"} \equiv$ (set $z'[a_2] \gets z[a_1]$ and set $C \gets C \cup \{a_1\}$)
     \END
     \LineComment{Use automatic differentiation of $\mathcal{F}$ to compute Jacobian, skipping copied addresses}
-    \For{$i$ in $1, \ldots, |W \setminus C|$}
-        \State $a_i \gets (W \setminus C)_i$ \Comment{Pick $i$th key in $(W \setminus C)$; the order does not matter}
-        \State $J[i,:] \gets \nabla_{\mathbf{z}_{R \setminus C}} \left( h_{e(z)}[a_i] \right)$ \Comment{{\small Gradient of $z'[a_i] = h_{e(z)}[a_i]$ wrt non-copied continuous inputs ($z_a$ for $a \in R \setminus C$)}}
+    \For{$i$ in $1$ to $|W|$}
+        \State $a_i \gets W_i$ \Comment{Pick $i$th key in $W$; the order does not matter}
+        \State $J[i,:] \gets \nabla_{\mathbf{z}_{R \setminus C}} \left( h_{e(z)}[a_i] \right)$ \Comment{{\small Gradient of $z'[a_i] = h_{e(z)}[a_i]$ w.r.t.
+non-copied continuous inputs ($z_a$ for $a \in R \setminus C$)}}
     \EndFor

### 2020-07-17 paper.tex
add check for address set match
conditioning

@@
 \subsection{Probability Distributions on Dictionaries}
-
+\label{sec:dists-on-dicts}
 While maximally general, the measure-theoretic formulation of involutive MCMC in Algorithm~\ref{alg:involutive-mcmc} is not amenable to an automated implementation, because it does not indicate how to compute the Radon-Nikodym derivative that is required for the acceptance probability, and it is unclear how to specify the probability measures involved.
@@
 
-% where $m(t, i) := \sum_{j=0}^{x[\mathtt{:}\mathtt{degree}]} \mathtt{xcoords}[i]^j \cdot x[(\mathtt{"c"}, j)]$, and where $p_{\mathrm{geometric}(r)}$, $p_{\mathrm{inversegamma}(\alpha, \beta)}$, and $p_{\mathrm{normal}(\mu, \sigma)}$ are density functions for geometric, inverse gamma, and normal distributions, respectively.
-
-% The addresses used by this generative function are:
-% \[
-% \mathcal{A} := \begin{array}{c}
-% \{\mathtt{:}\mathtt{noise}, \mathtt{:}\mathtt{degree}\} \bigcup
-% \{(\mathtt{"c"}, i) : i \in \integers_{\ge 0}\} \bigcup\\
-% \{(\mathtt{"y"}, i) : i \in \integers_{> 0}\}
-% \end{array}
-% \]
-% (Note that $\mathcal{A}$ could contain additional addresses that the program never samples.)
-% The random choice at address $\mathtt{:}\mathtt{degree}$ is discrete with $V_a = \integers_{\ge 0}$ and the other random choices are continuous.
-% The address $\mathtt{:}\mathtt{noise}$ has $V_a = \reals_{> 0}$ and $(\mathtt{"c"}, i)$ and $(\mathtt{"y"}, i)$ have $V_a = \reals$, and at all of these continuous addresses, we use the Lebesgue measure for $\mu_a$.
-% TODO double-check this
-% The density with respect to the reference measure $\mu$ is then:
-% \begin{align*}
-% p(x) =
-% \begin{array}{l}
-% p_{\mathrm{geometric}(0.5)}(x[\mathtt{:}\mathtt{degree}]) \cdot\\
-% p_{\mathrm{inversegamma}(1, 1)}(x[\mathtt{:}\mathtt{noise}]) \cdot\\
-% \prod_{0=1}^{x[\mathtt{:}\mathtt{degree}]} p_{\mathrm{normal}(0,1)}(x[(\mathtt{"c"}, i)])\cdot\\
-% \prod_{i=1}^n p_{\mathrm{normal}(m(t, i), 0.1 + x[\mathtt{noise}])}(x[(\mathtt{"y"}, i)])
-% \end{array}
-% \end{align*}
-% where $m(t, i) := \sum_{j=0}^{x[\mathtt{:}\mathtt{degree}]} \mathtt{xcoords}[i]^j \cdot x[(\mathtt{"c"}, j)]$, and where $p_{\mathrm{geometric}(r)}$, $p_{\mathrm{inversegamma}(\alpha, \beta)}$, and $p_{\mathrm{normal}(\mu, \sigma)}$ are density functions for geometric, inverse gamma, and normal distributions, respectively.
-
-% In our framework, a probabilistic program defines a probability distribution on finite maps (dictionaries) that map the names of random choices (their \emph{addresses}) to values.
-% % TODO trace terminology
-% Following standard terminology~\citep{?}, we call these dictionaries \emph{traces} because they are often sampled by running a probabilistic program, and `tracing' the execution of the program by recording the value of each random choice encountered.
-% We now make the concept of a trace more concrete using code in a probabilistic programming language.
-
-% The probabilistic programming language used in this paper is 
-% the dynamic modeling language in the Gen probabilistic programming system~\citep{gen-pldi}.
-% The language is embedded in the general-purpose Julia programming language~\citep{julia}.
-% Julia functions can be turned into Gen probabilistic programs by tagging
-% each random choice with an explicit address in the code, 
-% as part of a \emph{random-choice expression} of the following form:
-% \noindent
-% \begin{center}
-% \begin{tabular}{c}
-% {\begin{lstlisting}[basicstyle=\small\ttfamily]
-% {<address>} ~ <distribution>
-% \end{lstlisting}}
-% \end{tabular}
-% \end{center}
-% The left-hand side of the expression (green) specifies the address of the choice, and the right-hand side specifies the probability distribution to be sampled from.
-% For example, the following expression samples a non-negative integer from a geometric distribution and the value gets recorded in the trace at the address \texttt{"a"}:
-% \noindent
-% \begin{center}
-% \begin{tabular}{c}
-% {\begin{lstlisting}[basicstyle=\small\ttfamily]
-% {"a"} ~ geometric(0.5)
-% \end{lstlisting}}
-% \end{tabular}
-% \end{center}
-% In addition to being recorded in the trace, the sampled value is returned to the running program.
-% That is, the random choice expression evaluates to the sampled value.
-% For example, the expression below samples the random variable \texttt{"a"} and multiplies it by 2,
-% yielding an even non-negative integer:
-% \noindent
-% \begin{center}
-% \begin{tabular}{c}
-% {\begin{lstlisting}[basicstyle=\small\ttfamily]
-% ({"a"} ~ geometric(0.5)) * 2
-% \end{lstlisting}}
-% \end{tabular}
-% \end{center}
-% A probabilistic program will typically combine multiple random choice expressions. For example,
-% the following code samples a random choice at address \texttt{"a"}, assigns it to the Julia variable
-% \texttt{a}, and then samples a Bernoulli random variable at address \texttt{"b"} with a parameter
-% that depends on \texttt{a}:
-% \noindent
-% \begin{center}
-% \begin{tabular}{c}
-% {\begin{lstlisting}[basicstyle=\small\ttfamily]
-% a = ({"a"} ~ geometric(0.5))
-% {"b"} ~ bernoulli(a / (a + 1))
-% \end{lstlisting}}
-% \end{tabular}
-% \end{center}
-% This probabilistic program has an unbounded (countably infinite) number of possible traces it could produce.
-% One possible trace samples the value 1 at $\texttt{"a"}$ and the Boolean value $\false$ at \texttt{"b"}, and we denote this trace by $\{\mathtt{"a"} \mapsto 1, \mathtt{"b"} \mapsto \false\}$.
-% Note that the Julia variable to which the first choice is assigned (\texttt{a}) happens to be the same as the address (\texttt{"a"}).
-% This is not required, and often the random choice has no corresponding Julia variable in the program, as in the second random choice with address \texttt{"b"}.
-% But to reduce unnecessary typing, the following optional syntactic sugar will simultaneously assign the value to the program variable $\mathtt{a}$ and record it at trace address $\mathtt{"a"}$.\footnote{
-% More precisely, the address is the Julia symbol (interned string) \texttt{a}, constructed with Julia syntax \texttt{:a}.}
-% \noindent
-% \begin{center}
-% \begin{tabular}{c}
-% {\begin{lstlisting}[basicstyle=\small\ttfamily]
-% a ~ geometric(0.5)
-% \end{lstlisting}}
-% \end{tabular}
-% \end{center}
-
-% What gives probabilistic programs much of their expressive power is that they can make use of familiar control flow constructs, like loops and if statements, and this control flow can both depend on random choices and influence which random choices are made downstream.
-Consider the following program implementing a geometric
-% distribution as the result of iterated Bernoulli draws:
-% \noindent
-% \begin{center}
-% \begin{tabular}{c}
-% {\begin{lstlisting}[basicstyle=\small\ttfamily]
-% i = 1
-% while ({i} ~ bernoulli(0.5))
-%     i = i + 1
-% end
-% \end{lstlisting}}
-% \end{tabular}
-% \end{center}
-% Note that each Bernoulli flip uses a different address (the number \texttt{i} increases at each iteration).
-For the probabilistic program to be valid, we require that every address is used at most once within every possible execution.
-
-% For probabilistic programs that sample from discrete probability distributions like those used above, the meaning of the program is a probability mass function $p$ that assigns a probability to every possible trace.
-% For example, for the program with the \texttt{while} loop above, the distribution is:
-% \[
-% \begin{array}{c|c}
-%     x & p(x)\\
-%     \hline
-%     \{1 \mapsto \false\} & 0.5\\
-%     \{1 \mapsto \true, 2 \mapsto \false\} & 0.5^2\\
-%     \{1 \mapsto \true, 2 \mapsto \true, 3 \mapsto \false\} & 0.5^3\\
-%     .. & ..
-% \end{array}
-% \]
-% We assume that probabilistic programs halt with probability $1$, meaning that the sum of probabilities over all (finite) traces is one.
-% Note that traces are always finite, but there may be an infinite number of possible traces, as in the previous example.
-
-% In order to construct more complex distributions, we wrap probabilistic programs in functions, called
-% `generative functions' in Gen.
-% These functions can then be invoked by other functions, including recursively, which allows programs encode probability distributions on more complex and combinatorial objects.
-% When invoking such a function, we give an \emph{address namespace} under which every random choice made in that function should be recorded.
-% This helps prevent addresses of choices defined in different functions from colliding.
-% The syntax for calling a generative function and assigning the address namespace is the same as that of random choice expressions.
-% For example:
-% \begin{center}
-% \begin{tabular}{c}
-% \begin{lstlisting}[basicstyle=\small\ttfamily]
-% @gen function g()
-%   if ({"go"} ~ bernoulli(0.2))
-%     n1 = ({"L"} ~ g())
-%     n2 = ({"R"} ~ g())
-%     return n1 + n2 + 1
-%   else
-%     return 1
-%   end
-% end
-% \end{lstlisting}
-% \end{tabular}
-% \end{center}
-% Because this is a recursive function, a the trace is populated with a hierarchy of addresses.
-% For example, the first random choice in the first recursive call has address \texttt{"L"."go"}.
-% When this function is invoked it encodes the following probability distribution on traces (quotation marks omitted from addresses to improve readability):
-% \[
-% \begin{array}{c|c}
-%     x & p(x)\\
-%     \hline
-%     \left\{\mathtt{go} \mapsto \false\right\} & 0.8\\
-% & \\
-%     \left\{\begin{array}{c}
-%         \mathtt{go} \mapsto \true, \mathtt{L}.\mathtt{go} \mapsto \false, \mathtt{R}.\mathtt{go} \mapsto \false
-%     \end{array}\right\} & 0.2 \cdot 0.8^2\\
-% & \\
-%     \left\{\begin{array}{c}
-%         \mathtt{go} \mapsto \true,
-%         \mathtt{L}.\mathtt{go} \mapsto \true,
-%         \mathtt{R}.\mathtt{go} \mapsto \false,\\
-%         \mathtt{L}.\mathtt{L}.\mathtt{go} \mapsto \false,
-%         \mathtt{L}.\mathtt{R}.\mathtt{go} \mapsto \false
-%     \end{array}\right\}& 0.2^2 \cdot 0.8^3\\
-%      .. & ..
-% \end{array}
-% \]
-
-% %\subsubsection{Handling Continuous Distributions}
-% %To handle programs that sample continuous probability distributions the notion of a probability mass function on traces needs to be generalized to a probability density function on traces with respect to a certain reference measure.
-% %The probabilistic program still defines a probability distribution (i.e. measure) on traces, but via  a density instead of a probability mass function.
-% %
-% %Let $\mathcal{A}$ denote a countable set of all possible addresses.
-% %For each address $a \in \mathcal{A}$ we assign a reference measure space $(V_a, \Sigma_a, \mu_a)$ where $\mu_a$ is $\sigma$-finite ($V$ stands for `values').
-% %Let $T := \cup_{A \subseteq \mathcal{A} : |A| < \infty} \{(A, \mathbf{x}) : \mathbf{x} \in \times_{a \in A} V_a\}\}$ denote the set of all finite traces (i.e.
-dictionaries) from addresses in $\mathcal{A}$ to their values.
-% %For each finite $A \subseteq \mathcal{A}$, let $\Sigma_A := \otimes_{a \in A} \Sigma_a$ (the product $\sigma$-algebra over $\Sigma_a$ for $a \in A$), and let $\mu_A := \times_{a \in A} \mu_a$ (the product measure).
-% %Let $\Sigma \subseteq \mathcal{P}(T)$ be the $\sigma$-algebra formed by sets of the form $B := \cup_{A \subseteq \mathcal{A}, |A| < \infty} \{(A, \mathbf{x}) : \mathbf{x} \in B_A\}$ for some $B_A \in \Sigma_A$ for each $A$.
-% %Let $\mu : \Sigma \to [0, \infty)$ be a measure on measurable space $(T, \Sigma)$ defined by $\mu(B) := \sum_{A \subseteq \mathcal{A}, |A| < \infty} \mu_A(B_A)$.
-% %
-
-% \subsubsection{Automatically Computing Densities}
-% % TODO explain the non-standard interpretation idea.
-% A probabilistic program $P$ denotes a probability density $p : T \to [0, \infty)$ with respect to a reference measure $\mu$ on dictionaries.
-% Defining the denotational semantics of a particular probabilistic programming language (the function mapping programs $P$ to density functions $p$) is out of scope for this paper, and we refer to the reader to~\citet{popltraces} for an example of denotational semantics of a probabilistic programming language similar to that of this paper.
-% %But, informally, the density of a trace in our formalism (and existing formalisms) is the product of densities of each individual random choice, as an example below illustrates.
-% But this section illustrates what form density functions for probabilistics programs take and how to they can be automatically evaluated by probabilistic programming systems.
-% For example, consider the following probabilistic program, which encodes a generative model for data generated from a polynomial with unknown degree and unknown coefficients:
-% \noindent
-% \begin{center}
-% \begin{tabular}{c}
-% {\begin{lstlisting}[basicstyle=\small\ttfamily]
-% @gen function polynomials(x_coords)
-%     degree ~ |geometric(0.5)|
-%     noise ~ |inv_gamma(1, 1)|
-%     coeffs = [ ({("c", i)} ~ |normal(0, 2)|)
-%                for i in 0:degree ]
-%     for i=1:length(x_coords)
-%         x = x_coords[i]
-%         mu = coeffs' * x.^(0:length(coeffs)-1)
-%         {("y", i)} ~ |normal(mu, noise + 0.1)|
-%     end
-% end
-% \end{lstlisting}}
-% \end{tabular}
-% \end{center}
-% The addresses used by this generative function are:
-% \[
-% \mathcal{A} := \begin{array}{c}
-% \{\mathtt{:}\mathtt{noise}, \mathtt{:}\mathtt{degree}\} \bigcup
-% \{(\mathtt{"c"}, i) : i \in \integers_{\ge 0}\} \bigcup\\
-% \{(\mathtt{"y"}, i) : i \in \integers_{> 0}\}
-% \end{array}
-% \]
-% (Note that $\mathcal{A}$ could contain additional addresses that the program never samples.)
-% The random choice at address $\mathtt{:}\mathtt{degree}$ is discrete with $V_a = \integers_{\ge 0}$ and the other random choices are continuous.
-% The address $\mathtt{:}\mathtt{noise}$ has $V_a = \reals_{> 0}$ and $(\mathtt{"c"}, i)$ and $(\mathtt{"y"}, i)$ have $V_a = \reals$, and at all of these continuous addresses, we use the Lebesgue measure for $\mu_a$.
-% TODO double-check this
-% The density with respect to the reference measure $\mu$ is then:
-% \begin{align*}
-% p(x) =
-% \begin{array}{l}
-% p_{\mathrm{geometric}(0.5)}(x[\mathtt{:}\mathtt{degree}]) \cdot\\
-% p_{\mathrm{inversegamma}(1, 1)}(x[\mathtt{:}\mathtt{noise}]) \cdot\\
-% \prod_{0=1}^{x[\mathtt{:}\mathtt{degree}]} p_{\mathrm{normal}(0,1)}(x[(\mathtt{"c"}, i)])\cdot\\
-% \prod_{i=1}^n p_{\mathrm{normal}(m(t, i), 0.1 + x[\mathtt{noise}])}(x[(\mathtt{"y"}, i)])
-% \end{array}
-% \end{align*}
-% where $m(t, i) := \sum_{j=0}^{x[\mathtt{:}\mathtt{degree}]} \mathtt{xcoords}[i]^j \cdot x[(\mathtt{"c"}, j)]$, and where $p_{\mathrm{geometric}(r)}$, $p_{\mathrm{inversegamma}(\alpha, \beta)}$, and $p_{\mathrm{normal}(\mu, \sigma)}$ are density functions for geometric, inverse gamma, and normal distributions, respectively.
-
-\paragraph{Automating Density Computation.}
-% In general, the density $p$ of a probabilistic program with respect to the reference measure $\mu$ on dictionaries, evaluated at a dictionary $x = (A, \mathbf{x})$, can be computed as the product of individual density functions evaluated at each random choice $x[a]$.
-% To compute the density of a dictionary $x$, many probabilistic programming systems including Gen employ traces.
-% In the context of probabilistic programming, a \emph{trace} $t$ is a record of each of the random choices made during an execution of a probabilistic program, including (i) the addresses $A$ of the choices, (ii) their values $\mathbf{x}$, and (iii) the order in which the addresses was visited (as a list of addresses $t_1, \ldots, t_{|A|} \in A$), and (iv) the probability distributions $p^{(t)}_a$ from which each address $a \in A$ was sampled during the execution.
-% That is, each trace $t$ contains a dictionary $x = (A, \mathbf{x})$ such that $p(x) > 0$, as well as the order information $t_1, \ldots, t_{|A|}$ and the distribution information $p^{(t)}_a$ for each $a$.
-% If a trace $t$ is available for a dictionary $x$, then the probability density of the dictionary is the product of the densities of each random choice in the trace:
-% \footnote{Probabilistic programming implementations use sums of log-densities instead of products for numerical reasons.}
-% \begin{equation} \label{eq:density-product}
-% p((A, \mathbf{x})) = \prod_{i=1}^{|A|} p^{(t)}_{a}(x[a_{t_i}] | x[a_{t_1}], \ldots, x[a_{t_{i-1}}])
-% \end{equation}
-% See~\citet{popltraces} for an example of a formal proof that the product of densities is the density of the dictionary $x$ with respect to the reference measure.
-% Note that although in general, the probability density $p^{(t)}_a$ of address $a$ is a function of the value of the choice $x[a]$ as well the values of all previously sampled choices, in practice it is only a function of a subset of previous choices due to conditional independence structure in the model.
-
-% Given a dictionary $x$ for which $p(x) > 0$, it is possible to obtain a trace $t$ containing that dictionary by executing the probabilistic program, but intercepting random choice expressions and modifying their behavior.
-% For each intercepted random choice expression during the execution, instead of sampling the value of the random choice at address $a$, we (i) append the address $a$ to the list of addresses ($t_1, t_2, \ldots$), (ii) record the probability distribution $p^{(t)}_a$, (iii) look up the address $a$ in the dictionary (obtaining $v \gets x[a]$), and (iv) resume regular program execution, evaluating the random choice expression to the value $v$ obtained from the dictionary.
-% We denote a trace $t$ obtained from a dictionary $x$ and a probabilistic program $P$:
-% \begin{equation}
-% t \gets \mathrm{trace}(P, x)
-% \end{equation}
-The density of a dictionary $(A, \mathbf{x})$ under any probabilistic program's distribution on traces can be computed using Algorithm~\ref{alg:prob-prog} (\textsc{trace-and-score}).
+\paragraph{Automatic Sampling and Density Computation for Probabilistic Programs.}
+Sampling traces from and computing densities of dictionaries under, the distribution on traces
+induced by a probabilistic program is straight-forward to do, using a standard technique
+in probabilistic programming.
+
+We illustrate this technique in the \textsc{trace-and-score} and \textsc{score} subroutines in Algorithm~\ref{alg:auto-involutive-mcmc}.
+The \textsc{trace-and-score} subroutine samples a trace by running a probabilistic program,
+but recording the value of every encountered choice into a dictionary, which it returns once execution has terminated.
+It also returns the log density of the trace, calculated by accumulating densities of the individual choices it encounters.
+(Note that this procedure is only valid if the program
+halts with probability 1. Otherwise, it could loop infinitely, and even if it terminates,
+the density will be incorrect.)
+
+The density of an arbitrary dictionary $(A, \mathbf{x})$ under any probabilistic program's distribution on traces can be computed using \textsc{score}.
 The idea is to run the probabilistic program, and whenever a random choice \texttt{\{a\} $\sim$ d} is encountered, to look up the value $x_a$ in the dictionary, compute its density under the primitive distribution $d$, multiply this density into a running total, and return control to the probabilistic program as if the sampling instruction had executed and returned $x_a$.
 At the end, the running total can be returned as the trace's density.
-(If at any point an address $a \not\in A$ is encountered, or if not all addresses in $A$ have been visited at the end of execution, the algorithm returns $0$ as the density.) This is a standard technique in probabilistic programming.
-
-\subsubsection{Conditioning on Observed Data}
-% TODO update this
-The formalism in this section so far defines the model distribution (with respect to which stationarity of the kernel is required) as the distribution on traces of an (unconditioned) probabilistic program $P$ that encodes a generative model.
-However, typically in applications of MCMC the target distribution is obtained by conditioning on evidence.
-Our formalism does not distinguish between latent and observed random choices, and it is straightforward to show that the involution MCMC kernel is stationary with respect to a conditional distribution on the latent part of a partially observed trace as long as the involution does not modify the existence or values of observed random choices (when such conditional distribution is well-defined).
-Gen provides a mechanism for obtaining initial trace(s) with observed random choices fixed deterministically to the observed values.
-% TODO could add more content from my thesis here to make this more formal.
-% (define marginal density, then define conditional distribution)
-% could also do this in an updated version of the tech report..
+If at any point an address $a \not\in A$ is encountered, or if not all addresses in $A$ have been visited at the end of execution, the algorithm returns $0$ as the density.
 
@@
 
-Provided the two probabilistic programs $P$ and $Q$ satisfy a simple technical requirement (described in Section~\ref{sec:technical-requirement}), they can be used to specify the model density $p$ and auxiliary densities $q_x$ for use with involutive MCMC on dictionaries (Algorithm~\ref{alg:involutive-mcmc}).
+Given two probabilistic programs $P$ and $Q$ satisfying a simple technical requirement (described in Section~\ref{sec:technical-requirement}), we can interpret them as specifying the model density $p$ and auxiliary densities $q_x$ for use with involutive MCMC on dictionaries (Algorithm~\ref{alg:involutive-mcmc}).
+(In this case, the program $Q$
+accepts a trace $x$ of $P$ as input, and the distribution for $q_x$ is given by $Q(x)$.)
 The acceptance probability of involutive MCMC includes the following ratio of densities:
@@
 \end{equation}
-This ratio can be computed term-by-term using \textsc{trace-and-score} (Algorithm~\ref{alg:trace-based-involutive-mcmc}). 
+
+This ratio can be computed term-wise using \textsc{score} on the probabilistic programs $P$ and $Q$.
+
+However, typically, we cannot use a probabilistic program $P$ to represent the target distribution directly: probabilistic programs implement simulators for a distribution, but target densities are typically not tractable to simulate (hence the need for MCMC).
+Instead, we may wish to sample from a target $p$ that arises from conditioning a probabilistic program $P$'s distribution over traces on observations of the values at some addresses.
+Let $\tilde{p}$ denote $P$'s distribution over traces, and let $b = (B, \textbf{b})$ be a dictionary of observations, as described in Section~\ref{sec:dists-on-dicts}.
+Then, as shown in that section, the target distribution $p(x) = \tilde{p}(x \mid b) = \frac{\tilde{p}(x \oplus b)}{Z(b)}$, where $Z(b) = \int_{\{m \mid m \cap B = \emptyset\}} p(m \oplus b) \mu_\mathcal{D}(\text{d}m)$ is the partition function.
+Then $\tilde{p}(x \oplus b) = Z(b)p(x)$.
+If we use $\tilde{p}(x \oplus b)$ in place of $p(x)$ to compute the ratio above, we will wind up with the same output, because $Z(b)$ will cancel in the numerator and denominator.
+
+Thus, this ratio can be computed term-by-term using \textsc{score} (Algorithm~\ref{alg:auto-involutive-mcmc}): to compute $q_x(y')$, we run the algorithm directly on the program $Q(x)$, and to compute $p(x)$ and $p(x')$, we actually merge the dictionary $x$ with the observations $b$ and compute $Z(b)p(x) = \tilde{p}(x \oplus b)$ instead, by running the algorithm on $P()$ with trace $x \oplus b$.
+The density $q_x(y)$ can be computed while $y$ is being sampled, using \textsc{trace-and-score}.
+
 Section~\ref{sec:sparsity} describes a more efficient approach that exploits sparsity in the involution and resulting cancellation in the ratio for improved efficiency, but requires a more sophisticated probabilistic programming runtime system.

### 2020-07-17 paper.tex
Missing comma

@@
     \State $\alpha \gets \min\{1, \exp(\log p(x') - \log p(x) + \log q_{x'}(y') - \log q_x(y) + \log D)\}$ \Comment{Compute acceptance probability}
-    \State \textbf{with probability} \,$\alpha$ \,\Return $x'$ \algorithmicelse \, \Return $x$ 
+    \State \textbf{with probability} \,$\alpha$ \,\Return $x'$ \algorithmicelse \, \Return $x$  \algorithmicend
 \EndProcedure\\
@@
 \Procedure{trace-and-score}{$\mathcal{P}$}
-   \State $\log w \gets 0$
+   \State $s \gets 0$
    \State $x \gets \{\}$
@@
         \State 2. set $x[a] \gets v$
-        \State 3. set $\log w \gets \log w + \textproc{logpdf}(\mathrm{distribution}, v)$
+        \State 3. set $s \gets s + \textproc{logpdf}(\mathrm{distribution}, v)$
         \State 4. evaluate to $v$)
     \END
-   \State \Return $(x, \log w)$
+   \State \Return $(x, s)$
 \EndProcedure
@@
 \Procedure{score}{$\mathcal{P}$, $x$}
-   \State $\log w \gets 0$
+    \State $s \gets 0$
+    \State $A \gets\{\}$
     \execute[$\mathcal{P}$], but with $\mathtt{"}a \sim \mathrm{distribution}\mathtt{"} \equiv$ (
         \State 1. set $v \gets x[a]$
-        \State 2. set $\log w \gets \log w + \textproc{logpdf}(\mathrm{distribution}, v)$
-        \State 3. evaluate to $v$)
+        \State 2. set $A \gets A \cup \{a\}$
+        \State 3. set $s \gets s + \textproc{logpdf}(\mathrm{distribution}, v)$
+        \State 4. evaluate to $v$)
     \END
-   \State \Return $\log w$
+   \State \algorithmicif\hspace{1mm} $A \ne A_x$ \algorithmicthen\hspace{1mm} $s \gets -\infty$ \algorithmicend
+   \State \Return $s$
 \EndProcedure
@@
 \paragraph{Automatic Sampling and Density Computation for Probabilistic Programs.}
-Sampling traces from and computing densities of dictionaries under, the distribution on traces
+Sampling traces from, and computing densities of dictionaries under, the distribution on traces
 induced by a probabilistic program is straight-forward to do, using a standard technique

### 2020-07-17 paper.tex
Notation (mathcal P, mathcal Q, marginal likelihood instead of partition function)

@@
 
-Given two probabilistic programs $P$ and $Q$ satisfying a simple technical requirement (described in Section~\ref{sec:technical-requirement}), we can interpret them as specifying the model density $p$ and auxiliary densities $q_x$ for use with involutive MCMC on dictionaries (Algorithm~\ref{alg:involutive-mcmc}).
-(In this case, the program $Q$
-accepts a trace $x$ of $P$ as input, and the distribution for $q_x$ is given by $Q(x)$.)
+Given two probabilistic programs $\mathcal{P}$ and $\mathcal{Q}$ satisfying a simple technical requirement (described in Section~\ref{sec:technical-requirement}), we can interpret them as specifying the model density $p$ and auxiliary densities $q_x$ for use with involutive MCMC on dictionaries (Algorithm~\ref{alg:involutive-mcmc}).
+(In this case, the program $\mathcal{Q}$
+accepts a trace $x$ of $\mathcal{P}$ as input, and the distribution for $q_x$ is given by $\mathcal{Q}_x$.)
 The acceptance probability of involutive MCMC includes the following ratio of densities:
@@
 
-This ratio can be computed term-wise using \textsc{score} on the probabilistic programs $P$ and $Q$.
-
-However, typically, we cannot use a probabilistic program $P$ to represent the target distribution directly: probabilistic programs implement simulators for a distribution, but target densities are typically not tractable to simulate (hence the need for MCMC).
-Instead, we may wish to sample from a target $p$ that arises from conditioning a probabilistic program $P$'s distribution over traces on observations of the values at some addresses.
-Let $\tilde{p}$ denote $P$'s distribution over traces, and let $b = (B, \textbf{b})$ be a dictionary of observations, as described in Section~\ref{sec:dists-on-dicts}.
-Then, as shown in that section, the target distribution $p(x) = \tilde{p}(x \mid b) = \frac{\tilde{p}(x \oplus b)}{Z(b)}$, where $Z(b) = \int_{\{m \mid m \cap B = \emptyset\}} p(m \oplus b) \mu_\mathcal{D}(\text{d}m)$ is the partition function.
-Then $\tilde{p}(x \oplus b) = Z(b)p(x)$.
-If we use $\tilde{p}(x \oplus b)$ in place of $p(x)$ to compute the ratio above, we will wind up with the same output, because $Z(b)$ will cancel in the numerator and denominator.
-
-Thus, this ratio can be computed term-by-term using \textsc{score} (Algorithm~\ref{alg:auto-involutive-mcmc}): to compute $q_x(y')$, we run the algorithm directly on the program $Q(x)$, and to compute $p(x)$ and $p(x')$, we actually merge the dictionary $x$ with the observations $b$ and compute $Z(b)p(x) = \tilde{p}(x \oplus b)$ instead, by running the algorithm on $P()$ with trace $x \oplus b$.
+This ratio can be computed term-wise using \textsc{score} on the probabilistic programs $\mathcal{P}$ and $\mathcal{Q}$.
+
+However, typically, we cannot use a probabilistic program $\mathcal{P}$ to represent the target distribution directly: probabilistic programs implement simulators for a distribution, but target densities are typically not tractable to simulate (hence the need for MCMC).
+Instead, we may wish to sample from a target $p$ that arises from conditioning a probabilistic program $\mathcal{P}$'s distribution over traces on observations of the values at some addresses.
+Let $\tilde{p}$ denote $\mathcal{P}$'s distribution over traces, and let $b = (B, \textbf{b})$ be a dictionary of observations, as described in Section~\ref{sec:dists-on-dicts}.
+Then, as shown in that section, the target distribution $p(x) = \tilde{p}(x \mid b) = \frac{\tilde{p}(x \oplus b)}{\mathcal{L}(b)}$, where $\mathcal{L}(b) = \int_{\{m \mid m \cap B = \emptyset\}} p(m \oplus b) \mu_\mathcal{D}(\text{d}m)$ is the marginal likelihood of $b$, and does not depend on $x$.
+Then $\tilde{p}(x \oplus b) = \mathcal{L}(b)p(x)$.
+If we use $\tilde{p}(x \oplus b)$ in place of $p(x)$ to compute the ratio above, we will wind up with the same output, because $\mathcal{L}(b)$ will cancel in the numerator and denominator.
+
+Thus, this ratio can be computed term-by-term using \textsc{score} (Algorithm~\ref{alg:auto-involutive-mcmc}): to compute $q_x(y')$, we run the algorithm directly on the program $\mathcal{Q}_x$, and to compute $p(x)$ and $p(x')$, we actually merge the dictionary $x$ with the observations $b$ and compute $\mathcal{L}(b)p(x) = \tilde{p}(x \oplus b)$ instead, by running the algorithm on $\mathcal{P}$ with trace $x \oplus b$.
 The density $q_x(y)$ can be computed while $y$ is being sampled, using \textsc{trace-and-score}.
@@
 \end{center}
-In order to apply this function in the context of involution MCMC, we require probabilistic programs $P$ and $Q$ that define space of input and output traces for the involution.
-A minimal example does not require any auxiliary random choices, so $Q$ can be an empty probabilistic program.
-A minimal toy example of a probabilistic program $P$ that is compatible with the involution \texttt{h} defined above is:
+In order to apply this function in the context of involution MCMC, we require probabilistic programs $\mathcal{P}$ and $\mathcal{Q}$ that define space of input and output traces for the involution.
+A minimal example does not require any auxiliary random choices, so $\mathcal{Q}$ can be an empty probabilistic program.
+A minimal toy example of a probabilistic program $\mathcal{P}$ that is compatible with the involution \texttt{h} defined above is:
 \noindent
@@
 
-To represent reversible jump MCMC using involutive MCMC implemented with probabilistic programs, we write a probabilistic program $P$ that encodes the space of models $\mathcal{K}$, the prior distribution on models, $p(k)$, the per-model priors $p_k(\theta_k)$ and the per-model likelihoods $L_{D,k}(\theta_k)$.
-The set of all models $k$ corresponds to the set of all pairs $(A_P, \mathbf{d}_P)$ where $A_P$ represent possible trace structures (i.e.
-control-flow paths through $P$) and $\mathbf{d}_P$ are the set of assignments to discrete random choices made by $P$.
-The per-model continuous parameters $\theta$ correspond to continuous random choices $\mathbf{c}_P$.
+To represent reversible jump MCMC using involutive MCMC implemented with probabilistic programs, we write a probabilistic program $\mathcal{P}$ that encodes the space of models $\mathcal{K}$, the prior distribution on models, $p(k)$, the per-model priors $p_k(\theta_k)$ and the per-model likelihoods $L_{D,k}(\theta_k)$.
+The set of all models $k$ corresponds to the set of all pairs $(A_P, \mathbf{d}_P)$ where $A_\mathcal{P}$ represent possible trace structures (i.e.
+control-flow paths through $\mathcal{P}$) and $\mathbf{d}_P$ are the set of assignments to discrete random choices made by $\mathcal{P}$.
+The per-model continuous parameters $\theta$ correspond to continuous random choices $\mathbf{c}_\mathcal{P}$.
 % TODO explain the likelihood
-An auxiliary probabilistic program $Q$ encodes both the probability distribution on moves types using discrete random choices and possibly stochastic control flow ($(A_Q, \mathbf{d}_Q)$), and the per-move-type probability densities on $u$ using continuous random choices $\mathbf{c}_Q$.
-The involution $f$ factors into an (i) involution $f_1$ on pairs $i = ((A_P, \mathbf{d}_P), (A_Q, \mathbf{d}_Q))$ that defines the association between move types $\mathcal{M}$ and the model pairs ($k_1, k_2$); and (ii) a family of bijections $f_{1,i}$ on the space of pairs $(\mathbf{c}_P, \mathbf{c}_Q)$ of continuous random choices for both programs for fixed values of the discrete random choices and fixed trace structure.
+An auxiliary probabilistic program $\mathcal{Q}$ encodes both the probability distribution on moves types using discrete random choices and possibly stochastic control flow ($(A_\mathcal{Q}, \mathbf{d}_\mathcal{Q})$), and the per-move-type probability densities on $u$ using continuous random choices $\mathbf{c}_\mathcal{Q}$.
+The involution $f$ factors into an (i) involution $f_1$ on pairs $i = ((A_\mathcal{P}, \mathbf{d}_\mathcal{P}), (A_\mathcal{Q}, \mathbf{d}_\mathcal{Q}))$ that defines the association between move types $\mathcal{M}$ and the model pairs ($k_1, k_2$); and (ii) a family of bijections $f_{1,i}$ on the space of pairs $(\mathbf{c}_\mathcal{P}, \mathbf{c}_\mathcal{Q})$ of continuous random choices for both programs for fixed values of the discrete random choices and fixed trace structure.
 
 Figure~\ref{fig:mixture} shows a split-merge reversible jump kernel for an infinite Gaussian mixture model~\citep{richardson1997bayesian} implemented using the probabilistic and differentiable programming languages described in this paper.
-Figure~\ref{fig:mixture}(b) shows the infinite Gaussian mixture model, specified as a probabilistic program $P$.
+Figure~\ref{fig:mixture}(b) shows the infinite Gaussian mixture model, specified as a probabilistic program $\mathcal{P}$.
 The program takes the number of data points as input, then samples the number of clusters from a Poisson distribution, then samples cluster parameters and mixture proportions, and finally samples the data points from the resulting finite mixture.
-Figure~\ref{fig:mixture}(c) shows the auxiliary probabilistic program $Q$ for the split-merge kernel.
+Figure~\ref{fig:mixture}(c) shows the auxiliary probabilistic program $\mathcal{Q}$ for the split-merge kernel.
 This program takes a trace of the model program as input, and randomly decides whether to split a cluster and increase the number of clusters by one or merge two clusters and decrease the number of clusters by one.

### 2020-07-17 paper.tex
add some very rough content to section 4.2

@@
 %Reversible jump probabilistic programming~\citet{roberts2019reversible}
-
-\begin{figure*}[ht]
-    \centering
-    \includegraphics[width=\textwidth]{figures/mixture.pdf}
-    \caption{
-Example of reversible jump MCMC~\citep{green1995reversible} implemented using involutive MCMC in Gen.
-The example implements a `split-merge move' in a infinite Gaussian mixture model~\citep{richardson1997bayesian} using three Gen programs:
-(1) a probabilistic program $\mathtt{p}$ encoding the generative model (shown in b),
-(2) a probabilistic program $\mathtt{q}$ encoding an auxiliary probability distribution (shown in c),
-and (3) a differentiable program $\mathtt{h}$ that encodes an involution on the space of pairs of traces of $\mathtt{p}$ and $\mathtt{q}$ (shown in d).
-Gen's involutive MCMC operator (shown in e) automatically computes the acceptance probability.
-}
-    \label{fig:mixture}
-\end{figure*}
-
-\begin{figure*}[ht]
-    \centering
-    \includegraphics[width=\textwidth]{figures/structure-learning.pdf}
-    \caption{
-A mixture kernel implemented using involutive MCMC in Gen, applied to infer the covariance function of a Gaussian process.
-The prior on covariance functions is based on a probabilistic context-free grammar.
-Each component kernel in the mixture replaces a subtree of the covariance function parse tree with a new subtree.
-The mixture kernel chooses a random subtree to replace via a random walk on the parse tree.
-The mixture kernel is composed from three Gen programs:
-(1) a probabilistic program $\mathtt{p}$ encoding the generative model (shown in a),
-(2) a probabilistic program $\mathtt{q}$ encoding an auxiliary probability distribution (shown in b), and
-(3) a differentiable program $\mathtt{h}$ that encodes an involution (shown in d).
-}
-    \label{fig:structure-learning}
-\end{figure*}
 
@@
 
+\begin{figure*}[h!]
+    \centering
+    \includegraphics[width=0.93\textwidth]{figures/mixture.pdf}
+    \caption{
+Example of reversible jump MCMC~\citep{green1995reversible} implemented using involutive MCMC in Gen.
+The example implements a `split-merge move' in a infinite Gaussian mixture model~\citep{richardson1997bayesian} using three Gen programs:
+(1) a probabilistic program $\mathtt{p}$ encoding the generative model (shown in b),
+(2) a probabilistic program $\mathtt{q}$ encoding an auxiliary probability distribution (shown in c),
+and (3) a differentiable program $\mathtt{h}$ that encodes an involution on the space of pairs of traces of $\mathtt{p}$ and $\mathtt{q}$ (shown in d).
+Gen's involutive MCMC operator (shown in e) automatically computes the acceptance probability.
+}
+    \label{fig:mixture}
+\end{figure*}
+
 \subsection{Involutive MCMC with Dictionaries}
@@
 
-%\section{AUTOMATING INVOLUTIVE MCMC} \label{sec:involutive-mcmc}
-
-%Involutive MCMC (Algorithm~\ref{alg:involutive-mcmc}) defines a transition kernel that is stationary for a probabilistic model $p$, based on an auxiliary probability kernel $q$ and a deterministic involution $f$.
-Different choices of $q$ and $f$ recover instances of classic algorithms from the literature (e.g., Reversible-Jump MCMC~\citep{green}, TODO: others).
-This section briefly reviews involutive MCMC and gives a general measure-theoretic formulation of the algorithm on arbitrary state-spaces.
-In the following sections, we will describe how the implementations of involutive MCMC algorithms can be automated using probabilistic and differentiable programming.
-
-\section{AUTOMATING INVOLUTIVE MCMC} \label{sec:automating}
-
-% TODO add the algorithm box involving traces..
+\section{AUTOMATING INVOLUTIVE MCMC WITH TRACES} \label{sec:automating}
 
@@
 
-\subsection{Differentiable Programming with Dictionaries} \label{sec:differentiable-programming-traces}
-
-The previous section described how the probability densities ($p$ and $q$) required for involutive MCMC can be specified using probabilistic programs.
-This section introduces a differentiable programming language that can be used to specify the involution $f$ that is used in involutive MCMC.
-The interpreter for the language can compute the Radon-Nikodym derivative required for the acceptance probability using automatic differentiation, which will be discussed in Section~\ref{sec:automating}.
-The language has been implemented as part of the Gen system~\citep{Gen}.
-This section describes the language and informally introduces its semantics.
-
-In Algorithm~\ref{alg:involutive-mcmc} the involution $f : Z \to Z$ maps a pair $(x, y) \in Z$ to a new pair $(x', y') \in Z$.
-When the densities $p$ and $q$ are defined with probabilistic programs, each element of these pairs is a trace.
-Recall that a trace is mathematically, a finite dictionary that maps addresses to values.
-Therefore, the involution is a map from pairs of traces to pairs of traces.
-Then, $Z$ is a set of such trace pairs, and the involution $f$ is a function from $Z$ onto itself that is its own invverse.
-
-\subsubsection{A Differentiable Programming Language for Manipulating Dictionaries}
-Like the probabilistic programming languages described above, our differentiable language for involutions on traces is embedded based on Julia.
+\subsection{Differentiable Programming with Traces} \label{sec:differentiable-programming-traces}
+
+The previous section described how the probability densities ($p$ and $q$) required for involutive MCMC on spaces of dictionaries can be specified and automatically computed using probabilistic programs.
+This section introduces a lightweight differentiable programming language for specifying involutions ($f : Z \to Z$) on dictionaries.
+The interpreter for the language computes the Radon-Nikodym derivative required for the acceptance probability using automatic differentiation.
+The language is implemented as part of the Gen system~\citep{cusumano2019gen}, which is embedded in the Julia language.
+We also provide a simplified PyTorch-based implementation of the language.
+
+Recall that for involutive MCMC on a state space of dictionaries, we define $Z := \{x \oplus y : p(x) q_x(y) > 0\} \subseteq \mathcal{D}$, where $x \oplus y$ is the dictionary resulting from merging dictionaries $x$ and $y$ with disjoint keys.
+Intuitively, $z = x \oplus y$ the trace of a composite probabilistic program constructed by running $\mathcal{P}$ first, and then running $\mathcal{Q}$, passing the trace of $\mathcal{P}$ as its input.
+
+\subsubsection{A Differentiable Programming Language for Manipulating Traces}
+We denote a program in the differentiable language by $\mathcal{F}$ and the involution it defines as $f$.
+Like the probabilistic programming language described in the previous section, our differentiable language for involutions on traces is embedded based on Julia.
 The differentiable language has specialized expressions for reading values from addresses in a trace, writing values to the addresses in a trace, and copying values from an address in one trace to an address in another trace.
@@
 Hierarchical addresses of the form $\mathtt{a.b.c}$ are constructed in the language using the syntax \texttt{a=>b=>c}.}
-The third argument is either $\mathtt{:}\mathtt{discrete}$ or $\mathtt{:}\mathtt{continuous}$, and informs the language runtime whether the random choice at that address is drawn from a discrete or continuous distribution (this information is used to support automatic differentiation, as discussed in Section~\ref{sec:automating}).
+The third argument is either $\mathtt{:}\mathtt{discrete}$ or $\mathtt{:}\mathtt{continuous}$, and informs the language runtime whether the random choice at that address is drawn from a discrete or continuous distribution (this information will be used to support automatic differentiation).
 Similarly, to write a value to a trace, we use a \texttt{@write} keyword:
@@
 \end{center}
-As we will see in Section~\ref{sec:automating}, it is preferable to copy when possible instead of reading and then writing, as the additional information can improve the efficiency of automatic differentiation.
+As we will see in Section~\ref{sec:sparsity}, it is preferable to copy when possible instead of reading and then writing, as the additional information can improve the efficiency of automatic differentiation.
 
@@
  % TODO change it from trace_bijection to tracefn in the figures and implementation...
-Each input and output trace is given a name (e.g. \texttt{tr1}, \texttt{tr2}, \texttt{tr3}, and \texttt{tr4}) and these names must be unique.
+Each input and output trace is given a name (e.g. \texttt{tr1}, \texttt{tr2}, \texttt{tr3}, and \texttt{tr4}).
 For involution MCMC, the function must have two input traces and two output traces.
-The input trace is the input model trace ($x$), the second input trace is the input auxiliary trace ($y$), the first output trace is the output model trace ($x'$) and the second output trace is the output auxiliary trace ($y'$).
+The first input trace is the input model trace ($x$), the second input trace is the input auxiliary trace ($y$), the first output trace is the output model trace ($x'$) and the second output trace is the output auxiliary trace ($y'$).
 
@@
 
-%\subsubsection{The Radon-Nikodym Derivative}
-%
-%The differentiable programming language introduced in this section assumes that the reference measure of each random choice address is either the counting measure (corresponding to the $\mathtt{:}\mathtt{discrete}$ tag in \texttt{@read} and \texttt{@write} operations) or a Lebesgue measure ($\mathtt{:}\mathtt{continuous}$ tag).
-%Consider a bijection $f : Z \to Z$ (recall that every involution is a bijection).
-%If the space $Z$ can be partitioned into a countable number of sets $Z_1, Z_2, \ldots$, such that the each element of the partition is in one-to-one correspondence with another element ($f_1 : \{Z_1, Z_2, \ldots\} \to \{Z_1, Z_2, \ldots\}$ for some bijection $f_1$), and such that $f_{2,i} : Z_i \to Z_{f_1(i)}$ is a continuously differentiable bijection, and such that $f(z) = f_{2,i(z)}(z)$, then a Radon-Nikodym can be constructed from the Jacobian of the function $f_{2,i(z)}$ where $i(z) = Z_j$ for $z \in Z_j$.
-%We denote the Jacobian (determinant) of a function $g$ by $| J g |$.
-%See Section~\ref{sec:radon-nikodym-special-case} of the appendix for details.
-%Applying this to the setting where $Z$ is a set of pairs of traces of probabilistic programs as formalized in the preivous section is straightforward when the partition of $Z$ is the countable disjoint union over all pairs $(A, \mathbf{d})$ where $A \subseteq \mathcal{A}$ are trace structures and $\mathbf{d}$ are assignments to discrete choices.
-%
+\subsubsection{Example: Split-Merge Reversible Jump}
+
+Figure~\ref{fig:mixture}(d) shows an involution expressed in the differentiable programming language introduced in this section, and Figure~\ref{fig:mixture}(f) shows a schematic of the information flow through the involution as it is applied twice in a row.
+The involution implements a split or a merge proposal in the infinite mixture model of Figure~\ref{fig:mixture}(b), depending on the trace of the auxiliary program in Figure~\ref{fig:mixture}(c).
+If there are at least two clusters, and the value of address $\mathtt{split}$ is false, then the involution proposes to merge two clusters by running code block 2 shown in green.
+Otherwise, the involution proposes to split a cluster by running code block 1 shown in yellow.
+Note that that code block 1 (yellow) is itself an involution on the discrete addresses in the traces---each application of code block 1 to the traces causes them to flip back and forth between a split and a merge proposal.
+Also note that code blocks 2 and 3 are continuously differentiable bijections on the continuous addresses in the traces, and they are inverses of one another.
+
+\subsubsection{Computing the Jacobian with Automatic Differentiation}
+Recall that the involution $f$ is composed of (i) an involution $g$ on elements $e$ of a partition of the state space, and (ii) a pair of continuous differentiable bijections $h_{e}$ and $h_{g(e)} = h_{e}^{-1}$ between each pair of corresponding elements of the partition.
+Each function $h_e$ is a function from the values at continuous addresses in the input trace ($z[a]$ for $a \in A_z \cap \mathcal{J}$) to the values at continuous addresses in the output trace ($z'[a]$ for $a \in A_{z'} \cap \mathcal{J}$).
+We compute the Jacobian using automatic differentiation (AD).
+The Gen implementation uses forward-mode AD whereas our PyTorch implementation uses reverse-mode AD.
+The procedure \textproc{run-involution} in Algorithm~\ref{alg:auto-involutive-mcmc} shows the implementation of the interpreter for the language that uses reverse-mode AD.
+The interpreter executes $\mathcal{F}$ using regular Julia or Python semantics, but intercepts calls to \texttt{write}, \texttt{read}, and \texttt{copy} statements, and in addition to performing the desired operation, records the set of continuous addresses that are read, written, and copied.
+After $\mathcal{F}$ is finished executing, AD is used to compute the Jacobian $|J h_e|(z)$; in the case of reverse-mode AD by iterating over output continuous addresses and backpropagating from each one to all of the input continous addresses, computing the Jacobian row-by-row.
+The address metadata that was recorded during the first pass through the function is used to make AD more efficient, as discussed in the next section.
 
@@
 Of course, it is straightforward to replace the prior with a custom data-driven proposal distribution on subtrees.
+
+\begin{figure*}[ht]
+    \centering
+    \includegraphics[width=0.8\textwidth]{figures/structure-learning.pdf}
+    \caption{
+A mixture kernel implemented using involutive MCMC in Gen, applied to infer the covariance function of a Gaussian process.
+The prior on covariance functions is based on a probabilistic context-free grammar.
+Each component kernel in the mixture replaces a subtree of the covariance function parse tree with a new subtree.
+The mixture kernel chooses a random subtree to replace via a random walk on the parse tree.
+The mixture kernel is composed from three Gen programs:
+(1) a probabilistic program $\mathtt{p}$ encoding the generative model (shown in a),
+(2) a probabilistic program $\mathtt{q}$ encoding an auxiliary probability distribution (shown in b), and
+(3) a differentiable program $\mathtt{h}$ that encodes an involution (shown in d).
+}
+    \label{fig:structure-learning}
+\end{figure*}
 

### 2020-07-17 paper.tex
Notation (replace A with K)

--- 2020-07-17 paper.tex
@@
 \textbf{The space of finite dictionaries.}
-We fix a countably infinite set $\mathcal{A}$ of possible keys, such that each key $a$ is either called \emph{discrete} ($a \in \mathcal{I}$) or \emph{continuous} ($a \in \mathcal{J}$), where $\mathcal{A} = \mathcal{I} \cup \mathcal{J}$.\footnote{
-It is possible to assign a general measure space to each address, but this is not necessary for our purposes.}
-Let $V_a$ denote the set of possible values for key $a$, where $V_a$ is a countable set for each discrete key, and where $V_a = \mathbb{R}^{d_a}$ for each continuous key for some $d_a$.
-Given a set of keys $A$, let $V_A = \times_{a \in A} V_a$ denote the set of \textit{assignments} of values to each key.
-Then the set of all finite dictionaries is $\mathcal{D} := \bigcup_{A \subset \mathcal{A}, |A| < \infty} \{(A, \mathbf{x}) \mid \mathbf{x} \in V_A\}$. That is, a dictionary specifies 
-a finite set of keys $A \subset \mathcal{A}$ at which it has values, and an assignment $\mathbf{x}$ of values $x_a$ for each.
+We fix a countably infinite set $\mathcal{K}$ of possible keys, such that each key $k$ is either called \emph{discrete} ($k \in \mathcal{I}$) or \emph{continuous} ($k \in \mathcal{J}$), where $\mathcal{K} = \mathcal{I} \cup \mathcal{J}$.\footnote{
+It is possible to assign a general measure space to each key, but this is not necessary for our purposes.}
+Let $V_k$ denote the set of possible values for key $k$, where $V_k$ is a countable set for each discrete key, and where $V_k = \mathbb{R}^{d_k}$ for each continuous key for some $d_k$.
+Given a set of keys $K$, let $V_K = \times_{k \in A} V_k$ denote the set of \textit{assignments} of values to each key.
+Then the set of all finite dictionaries is $\mathcal{D} := \bigcup_{K \subset \mathcal{K}, |K| < \infty} \{(K, \mathbf{x}) \mid \mathbf{x} \in V_K\}$. That is, a dictionary specifies 
+a finite set of keys $K \subset \mathcal{K}$ at which it has values, and an assignment $\mathbf{x}$ of values $x_k$ for each.
 
@@
 \textbf{A measure space of finite dictionaries.}
-We associate a measure $\mu_a$ on $V_a$ for each key---the counting measure for each discrete key and the Lebesgue-measure on $\mathbb{R}^{d_a}$ for each continuous key.
-For each finite set of keys $A$, we make $V_A$ a measure space using the standard product $\sigma$-algebra $\Sigma_A = \otimes_{a \in A} \Sigma_a$ and the product measure $\mu_A = \times_{a \in A} \mu_a$.
-We equip $\mathcal{D}$ with the $\sigma$-algebra $\Sigma_{\mathcal{D}} := \{\bigcup_{A \subset \mathcal{A}, |A| < \infty} \{(A, \mathbf{x}) \mid \mathbf{x} \in B_A\} \mid B_A \in \Sigma_A \, \text{for each finite } A \subset \mathcal{A}\}$ to obtain a measurable space
+We associate a measure $\mu_k$ on $V_k$ for each key---the counting measure for each discrete key and the Lebesgue-measure on $\mathbb{R}^{d_k}$ for each continuous key.
+For each finite set of keys $K$, we make $V_K$ a measure space using the standard product $\sigma$-algebra $\Sigma_K = \otimes_{k \in A} \Sigma_k$ and the product measure $\mu_K = \times_{k \in A} \mu_k$.
+We equip $\mathcal{D}$ with the $\sigma$-algebra $\Sigma_{\mathcal{D}} := \{\bigcup_{K \subset \mathcal{K}, |K| < \infty} \{(K, \mathbf{x}) \mid \mathbf{x} \in B_K\} \mid B_K \in \Sigma_K \, \text{for each finite } K \subset \mathcal{K}\}$ to obtain a measurable space
 of dictionaries.
-A reference measure $\mu_\mathcal{D}$ on this space can be constructed using the product measures $\mu_A$: we set $\mu(B) := \sum_{A \in \mathcal{A}, |A| < \infty} \mu_A(\{\mathbf{x} \mid (A, \mathbf{x}) \in B\})$.
-
-\textbf{Notation for dictionaries.} Given a dictionary $d = (A, \mathbf{x})$, we write $A_d$ for $A$ and $d[a]$ for the value $x_a$ associated with a key $a \in A$.
+A reference measure $\mu_\mathcal{D}$ on this space can be constructed using the product measures $\mu_K$: we set $\mu(B) := \sum_{A \in \mathcal{K}, |K| < \infty} \mu_K(\{\mathbf{x} \mid (K, \mathbf{x}) \in B\})$.
+
+\textbf{Notation for dictionaries.} Given a dictionary $d = (K, \mathbf{x})$, we write $K_d$ for $K$ and $d[k]$ for the value $x_k$ associated with a key $k \in K$.
 We also denote specific dictionaries using notation $\{k_1 \mapsto v_1, k_2 \mapsto v_2, \ldots\}$.
-For example, the dictionary $(A, \mathbf{x})$ with $A = \{1, \mathtt{"foo"}\}$ and $x_1 = 0.123$ and $x_{\mathtt{"foo"}} = 5$ is denoted 
+For example, the dictionary $(K, \mathbf{x})$ with $K = \{1, \mathtt{"foo"}\}$ and $x_1 = 0.123$ and $x_{\mathtt{"foo"}} = 5$ is denoted 
 $\{1 \mapsto 0.123, \mathtt{"foo"} \mapsto 5\}$.
@@
 More generally a probability distribution on dictionaries is defined by a probability density $p : \mathcal{D} \rightarrow [0, \infty)$ such that $\int p(m) \mu_\mathcal{D}(\text{d}m) = 1$. 
-The probability is distributed the finite sets of keys $A \subseteq \mathcal{A}$ where $|A| < \infty$:
-\begin{equation}
-1 = \int p(m) \mu_\mathcal{D}(\text{d}m) = \sum_{\substack{A \subseteq \mathcal{A}\\|A| < \infty}} \int p(m) \mu_{A}(dm)
-\end{equation}
-where the probability for each key set $A$ is:
+The probability is distributed the finite sets of keys $K \subseteq \mathcal{K}$ where $|K| < \infty$:
+\begin{equation}
+1 = \int p(m) \mu_\mathcal{D}(\text{d}m) = \sum_{\substack{K \subseteq \mathcal{K}\\|K| < \infty}} \int p(m) \mu_{K}(dm)
+\end{equation}
+where the probability for each key set $K$ is:
 \begin{equation}\label{eq:integral}
-\int p(m) \mu_{A}(dm) = \sum_{\mathbf{x}_1} \int_{\mathbb{R}^{d}} p((A, (\mathbf{x}_1, \mathbf{x}_2))) d\mathbf{x}_2
-\end{equation}
-where $d: = \sum_{a \in A \cap \mathcal{J}} d_{a}$ and where $\mathbf{x}_1$ is an assignment to the discrete choices in $A$ and $\mathbf{x}_2$ is an assignment to the continuous choices.
+\int p(m) \mu_{K}(dm) = \sum_{\mathbf{x}_1} \int_{\mathbb{R}^{d}} p((K, (\mathbf{x}_1, \mathbf{x}_2))) d\mathbf{x}_2
+\end{equation}
+where $d: = \sum_{k \in A \cap \mathcal{J}} d_{k}$ and where $\mathbf{x}_1$ is an assignment to the discrete choices in $K$ and $\mathbf{x}_2$ is an assignment to the continuous choices.
 
@@
 %
-when $A_d = \{\texttt{k}, \texttt{y}_\texttt{1}, \dots, \texttt{y}_\texttt{n}\} \cup \{\texttt{m}_i \mid 1 \leq i \leq d[\texttt{k}]\} \cup \{\texttt{s}_i \mid 1 \leq i \leq d[\texttt{k}]\}$, and 0 otherwise.
-In this case, we have $V_\texttt{k} = \mathbb{N}$ with the counting measure for $\mu_\texttt{k}$, and for all other addresses $a \in \mathcal{A}$, $V_a = \mathbb{R}$ with the Lebesgue measure for $\mu_a$.
+when $K_d = \{\texttt{k}, \texttt{y}_\texttt{1}, \dots, \texttt{y}_\texttt{n}\} \cup \{\texttt{m}_i \mid 1 \leq i \leq d[\texttt{k}]\} \cup \{\texttt{s}_i \mid 1 \leq i \leq d[\texttt{k}]\}$, and 0 otherwise.
+In this case, we have $V_\texttt{k} = \mathbb{N}$ with the counting measure for $\mu_\texttt{k}$, and for all other keys $k \in \mathcal{K}$, $V_k = \mathbb{R}$ with the Lebesgue measure for $\mu_k$.
 For each $k \in \{0, 1, \ldots\}$, the probability for key set
@@
 \textbf{Conditional distributions via disintegration.} Consider a probability density $p$ on the space $\mathcal{D}$ of dictionaries. 
-%For a set of keys $A \subset \mathcal{A}$, let $p(A) := \int 1[A_m = A] p(m) \mu_\mathcal{D}(\text{d}m)$, the probability under $p$ of a dictionary with exactly the key set $A$. 
-We say a key $a \in \mathcal{A}$ \textit{almost always appears} if $\int \mathbf{1}[a \in A_m] p(m) \mu_\mathcal{D}(\text{d}m) = 1$.
+%For a set of keys $K \subset \mathcal{K}$, let $p(A) := \int 1[K_m = A] p(m) \mu_\mathcal{D}(\text{d}m)$, the probability under $p$ of a dictionary with exactly the key set $K$. 
+We say a key $k \in \mathcal{K}$ \textit{almost always appears} if $\int \mathbf{1}[k \in K_m] p(m) \mu_\mathcal{D}(\text{d}m) = 1$.
 Suppose $B$ is a set of keys that almost always appear for $p$, and that $b = (B, \mathbf{b})$ is a dictionary with keys $B$.
-Furthermore, let $(A_m, \mathbf{m}) \oplus (A_n, \mathbf{n}) := (A_m \cup A_n, \mathbf{m} \times \mathbf{n})$ be the \textit{merge} of two dictionaries $m$ and $n$ defined on disjoint key sets $A_m$ and $A_n$.
-Then we can define the conditional density $p(d \mid b) := \mathbf{1}[A_d \cap B = \emptyset] \frac{p(d \oplus b)}{\int_{\{m \mid A_m \cap B = \emptyset\}} p(m \oplus b) \mu_\mathcal{D}(\text{d}m)}$ when the denominator is finite.
-If the reference measure $\mu_a$ associated with each key $a \in B$ is discrete (e.g., the counting measure), then this definition corresponds to the ordinary notion of conditioning on an event (namely, the event that a sample from $p$ agrees with the dictionary $b$ on all addresses in $B$).
+Furthermore, let $(K_m, \mathbf{m}) \oplus (K_n, \mathbf{n}) := (K_m \cup K_n, \mathbf{m} \times \mathbf{n})$ be the \textit{merge} of two dictionaries $m$ and $n$ defined on disjoint key sets $K_m$ and $K_n$.
+Then we can define the conditional density $p(d \mid b) := \mathbf{1}[K_d \cap B = \emptyset] \frac{p(d \oplus b)}{\int_{\{m \mid K_m \cap B = \emptyset\}} p(m \oplus b) \mu_\mathcal{D}(\text{d}m)}$ when the denominator is finite.
+If the reference measure $\mu_k$ associated with each key $k \in B$ is discrete (e.g., the counting measure), then this definition corresponds to the ordinary notion of conditioning on an event (namely, the event that a sample from $p$ agrees with the dictionary $b$ on all keys in $B$).
 When this is not the case, it corresponds to a more general measure-theoretic notion called disintegration~\citep{chang1997conditioning}.
@@
 Consider the infinite univariate mixture model, and the conditional density given observed data $(B, \mathbf{b}) := \{\mathtt{y}_1 \mapsto y_1, \ldots, \mathtt{y}_{\mathtt{n}} \mapsto y_{\mathtt{n}}\}$.
-The conditional density $p(d | b)$ is nonzero only if $A_d = \{\texttt{k}\} \cup \{\texttt{m}_i \mid 1 \leq i \leq k\} \cup \{\texttt{s}_i \mid 1 \leq i \leq k\}$ for some $k$ ($d$ does not contain y-values), and the denominator in the definition of $p(d | b)$ simplifies to the familiar sum of marginal likelihoods over all $k$, where each marginal likelihood is a Riemann integral over $\mathbb{R}^{2k}$.
+The conditional density $p(d | b)$ is nonzero only if $K_d = \{\texttt{k}\} \cup \{\texttt{m}_i \mid 1 \leq i \leq k\} \cup \{\texttt{s}_i \mid 1 \leq i \leq k\}$ for some $k$ ($d$ does not contain y-values), and the denominator in the definition of $p(d | b)$ simplifies to the familiar sum of marginal likelihoods over all $k$, where each marginal likelihood is a Riemann integral over $\mathbb{R}^{2k}$.
 
@@
 Suppose that the model distribution and auxiliary distributions are probability distributions on dictionaries, with densities $p$ and $q_x$.
-Then, $X$ and $Y$ are both sets of dictionaries, and the joint space $Z$ is a set of pairs $(x, y)$ of dictionaries with keys $\mathcal{A}_P$ and $\mathcal{A}_Q$ respectively, so that $Z = X \times Y \subseteq \mathcal{D}_P \times \mathcal{D}_Q$ where $\mathcal{D}_P$ is the set of dictionaries on keys taken from $\mathcal{A}_P$ and similarly for $\mathcal{D}_Q$.
+Then, $X$ and $Y$ are both sets of dictionaries, and the joint space $Z$ is a set of pairs $(x, y)$ of dictionaries with keys $\mathcal{K}_P$ and $\mathcal{K}_Q$ respectively, so that $Z = X \times Y \subseteq \mathcal{D}_P \times \mathcal{D}_Q$ where $\mathcal{D}_P$ is the set of dictionaries on keys taken from $\mathcal{K}_P$ and similarly for $\mathcal{D}_Q$.
 % TODO: this is only needed for Algorithm 2, and it is relevant for the 
-To simplify the notation, and without loss of generality, we will assume that $\mathcal{A}_P$ and $\mathcal{A}_Q$ are disjoint\footnote{If $\mathcal{A}_P$ and $\mathcal{A}_Q$ are not disjoint then, they can be made so by adding a different prefix to each address.}, and we define $Z := \{ x \oplus y : p(x)q_x(y) > 0\} \subseteq \mathcal{D}$.
-where $\mathcal{D}$ is the set of dictionaries on keys from $\mathcal{A}_P \cup \mathcal{A}_Q$
+To simplify the notation, and without loss of generality, we will assume that $\mathcal{K}_P$ and $\mathcal{K}_Q$ are disjoint\footnote{If $\mathcal{K}_P$ and $\mathcal{K}_Q$ are not disjoint then, they can be made so by adding a different prefix to the keys of each set.}, and we define $Z := \{ x \oplus y : p(x)q_x(y) > 0\} \subseteq \mathcal{D}$.
+where $\mathcal{D}$ is the set of dictionaries on keys from $\mathcal{K}_P \cup \mathcal{K}_Q$
 (recall $x \oplus y$ denotes the dictionary resulting from merging dictionaries $x$ and $y$ with disjoint keys).
 
-Suppose there is a countable partition of $Z$ into $\{Z_e : e \in E\}$ such that across each $Z_e$ the sets of keys $A_x$ and $A_y$ and the values $x_a$ and $y_a$ for all discrete keys $a$ are both constant.
+Suppose there is a countable partition of $Z$ into $\{Z_e : e \in E\}$ such that across each $Z_e$ the sets of keys $K_x$ and $K_y$ and the values $x_k$ and $y_k$ for all discrete keys $k$ are both constant.
 Then each set $Z_e$ is isomorphic to a Euclidean space of assignments to the continuous keys in the two dictionaries.
@@
 \[
-z_1 \sim z_2 \iff (A_{z_1} = A_{z_2}) \land (z_1[a] = z_2[a] \; \forall a \in A_{z_1} \cap \mathcal{I})
+z_1 \sim z_2 \iff (K_{z_1} = K_{z_2}) \land (z_1[k] = z_2[k] \; \forall k \in K_{z_1} \cap \mathcal{I})
 \]
@@
    \State $x \gets \{\}$
-   \execute[$\mathcal{P}$], but with $\mathtt{"}a \sim \mathrm{distribution}\mathtt{"} \equiv$ (
+   \execute[$\mathcal{P}$], but with $\mathtt{"}k \sim \mathrm{distribution}\mathtt{"} \equiv$ (
         \State 1. set $v \sim \mathrm{distribution}$
-        \State 2. set $x[a] \gets v$
+        \State 2. set $x[k] \gets v$
         \State 3. set $s \gets s + \textproc{logpdf}(\mathrm{distribution}, v)$
@@
     \State $s \gets 0$
-    \State $A \gets\{\}$
-    \execute[$\mathcal{P}$], but with $\mathtt{"}a \sim \mathrm{distribution}\mathtt{"} \equiv$ (
-        \State 1. set $v \gets x[a]$
-        \State 2. set $A \gets A \cup \{a\}$
+    \State $K \gets\{\}$
+    \execute[$\mathcal{P}$], but with $\mathtt{"}k \sim \mathrm{distribution}\mathtt{"} \equiv$ (
+        \State 1. set $v \gets x[k]$
+        \State 2. set $K \gets K \cup \{k\}$
         \State 3. set $s \gets s + \textproc{logpdf}(\mathrm{distribution}, v)$
@@
     \END
-   \State \algorithmicif\hspace{1mm} $A \ne A_x$ \algorithmicthen\hspace{1mm} $s \gets -\infty$ \algorithmicend
+   \State \algorithmicif\hspace{1mm} $K \ne K_x$ \algorithmicthen\hspace{1mm} $s \gets -\infty$ \algorithmicend
    \State \Return $s$
@@
     \execute[$\mathcal{F}$], but with
-        \State $\mathtt{"write}(a, v)\mathtt{"} \equiv$
-            (\algorithmicif\hspace{1mm} $a \in \mathcal{I}$
-            \algorithmicthen\hspace{1mm} set $z'[a] \gets v$
-            \algorithmicelse\hspace{1mm} set $z'[a] \gets v$ and set $W \gets W \cup \{a\})$
-        \State $\mathtt{"read}(a)\mathtt{"} \equiv$
-            (\algorithmicif\hspace{1mm} $a \in \mathcal{I}$
-            \algorithmicthen\hspace{1mm} evaluate to $z[a]$
-            \algorithmicelse\hspace{1mm} set $R \gets R \cup \{a\}$ and evaluate to $z[a]$)
-        \State $\mathtt{"copy}(a_1, a_2)\mathtt{"} \equiv$ (set $z'[a_2] \gets z[a_1]$ and set $C \gets C \cup \{a_1\}$)
+        \State $\mathtt{"write}(k, v)\mathtt{"} \equiv$
+            (\algorithmicif\hspace{1mm} $k \in \mathcal{I}$
+            \algorithmicthen\hspace{1mm} set $z'[k] \gets v$
+            \algorithmicelse\hspace{1mm} set $z'[k] \gets v$ and set $W \gets W \cup \{k\})$
+        \State $\mathtt{"read}(k)\mathtt{"} \equiv$
+            (\algorithmicif\hspace{1mm} $k \in \mathcal{I}$
+            \algorithmicthen\hspace{1mm} evaluate to $z[k]$
+            \algorithmicelse\hspace{1mm} set $R \gets R \cup \{k\}$ and evaluate to $z[k]$)
+        \State $\mathtt{"copy}(k_1, k_2)\mathtt{"} \equiv$ (set $z'[k_2] \gets z[k_1]$ and set $C \gets C \cup \{k_1\}$)
     \END
@@
     \For{$i$ in $1$ to $|W|$}
-        \State $a_i \gets W_i$ \Comment{Pick $i$th key in $W$; the order does not matter}
-        \State $J[i,:] \gets \nabla_{\mathbf{z}_{R \setminus C}} \left( h_{e(z)}[a_i] \right)$ \Comment{{\small Gradient of $z'[a_i] = h_{e(z)}[a_i]$ w.r.t.
-non-copied continuous inputs ($z_a$ for $a \in R \setminus C$)}}
+        \State $k_i \gets W_i$ \Comment{Pick $i$th key in $W$; the order does not matter}
+        \State $J[i,:] \gets \nabla_{\mathbf{z}_{R \setminus C}} \left( h_{e(z)}[a_i] \right)$ \Comment{{\small Gradient of $z'[k_i] = h_{e(z)}[k_i]$ w.r.t.
+non-copied continuous inputs ($z_k$ for $k \in R \setminus C$)}}
     \EndFor
@@
 Any such program induces a probability distribution over its possible \textit{execution traces}, records of each random choice it makes.
-If each random choice is associated with a unique \textit{address} from a countable set
-of addresses $\mathcal{A}$, 
+If each random choice is associated with a unique \textit{address} from the set of
+dictionary keys $\mathcal{K}$, 
 then these traces can be viewed as finite dictionaries, mapping the address of each random choice to its value. 
@@
 with a single new construct, the \texttt{\{a\} $\sim$ d} expression, for making a \textit{named} random choice.
-An execution trace of a Gen probabilistic program can be sampled by running the program according to Julia's usual semantics, and upon encountering an expression of the form \texttt{\{a\} $\sim$ d}, (i) evaluating the \textit{address expression} \texttt{a} to obtain an address $a \in \mathcal{A}$; (ii) evaluating the \textit{distribution expression} $d$ to obtain a distribution $d$ over the measurable space $V_a$; (iii) sampling a value $x_a \sim d$ and adding the mapping $\{a \mapsto x_a\}$ to a dictionary called the program trace; and (iv) returning the sampled value $x_a$ to the program, to continue execution.
+An execution trace of a Gen probabilistic program can be sampled by running the program according to Julia's usual semantics, and upon encountering an expression of the form \texttt{\{a\} $\sim$ d}, (i) evaluating the \textit{address expression} \texttt{a} to obtain an address $k \in \mathcal{K}$; (ii) evaluating the \textit{distribution expression} $d$ to obtain a distribution $d$ over the measurable space $V_k$; (iii) sampling a value $x_k \sim d$ and adding the mapping $\{k \mapsto x_k\}$ to a dictionary called the program trace; and (iv) returning the sampled value $x_k$ to the program, to continue execution.
 When execution terminates, the program trace has accumulated a mapping for each random choice encountered during the program's execution.
@@
 The random choice at address $\mathtt{:}\mathtt{k}$ is discrete with $V_{\texttt{:k}} = \integers_{\ge 1}$, and the other random choices are continuous.
-Overall, the program defines a distribution over traces with the following density over traces $(A, \mathbf{x})$ with respect to $\mu_\mathcal{D}$:
+Overall, the program defines a distribution over traces with the following density over traces $(K, \mathbf{x})$ with respect to $\mu_\mathcal{D}$:
 \begin{align*}
-p(n)((A, \mathbf{x})) =
+p(n)((K, \mathbf{x})) =
 \begin{array}{l}
@@
 \end{align*}
-when $A$ contains exactly the addresses $\mathtt{:}\mathtt{k}, \mathtt{:}\mathtt{weights}, (\mathtt{:}\mathtt{x}, i)$ for $i = 1, \dots, n$, and $(\mathtt{:}\mathtt{mu}, j)$ and $(\mathtt{:}\mathtt{var}, j)$ for $j = 1, \dots, x_{\mathtt{:}\mathtt{k}}$; otherwise, the density is 0.
+when $K$ contains exactly the addresses $\mathtt{:}\mathtt{k}, \mathtt{:}\mathtt{weights}, (\mathtt{:}\mathtt{x}, i)$ for $i = 1, \dots, n$, and $(\mathtt{:}\mathtt{mu}, j)$ and $(\mathtt{:}\mathtt{var}, j)$ for $j = 1, \dots, x_{\mathtt{:}\mathtt{k}}$; otherwise, the density is 0.
 
@@
 
-The density of an arbitrary dictionary $(A, \mathbf{x})$ under any probabilistic program's distribution on traces can be computed using \textsc{score}.
-The idea is to run the probabilistic program, and whenever a random choice \texttt{\{a\} $\sim$ d} is encountered, to look up the value $x_a$ in the dictionary, compute its density under the primitive distribution $d$, multiply this density into a running total, and return control to the probabilistic program as if the sampling instruction had executed and returned $x_a$.
+The density of an arbitrary dictionary $(K, \mathbf{x})$ under any probabilistic program's distribution on traces can be computed using \textsc{score}.
+The idea is to run the probabilistic program, and whenever a random choice \texttt{\{a\} $\sim$ d} is encountered, to look up the value $x_k$ in the dictionary, compute its density under the primitive distribution $d$, multiply this density into a running total, and return control to the probabilistic program as if the sampling instruction had executed and returned $x_k$.
 At the end, the running total can be returned as the trace's density.
-If at any point an address $a \not\in A$ is encountered, or if not all addresses in $A$ have been visited at the end of execution, the algorithm returns $0$ as the density.
+If at any point an address $k \not\in K$ is encountered, or if not all addresses in $K$ have been visited at the end of execution, the algorithm returns $0$ as the density.
 
@@
 %See Section~\ref{sec:radon-nikodym-special-case} of the appendix for details.
-%Applying this to the setting where $Z$ is a set of pairs of traces of probabilistic programs as formalized in the preivous section is straightforward when the partition of $Z$ is the countable disjoint union over all pairs $(A, \mathbf{d})$ where $A \subseteq \mathcal{A}$ are trace structures and $\mathbf{d}$ are assignments to discrete choices.
+%Applying this to the setting where $Z$ is a set of pairs of traces of probabilistic programs as formalized in the preivous section is straightforward when the partition of $Z$ is the countable disjoint union over all pairs $(K, \mathbf{d})$ where $K \subseteq \mathcal{K}$ are trace structures and $\mathbf{d}$ are assignments to discrete choices.
 %
@@
 
-The reversible jump MCMC framework involves a set of `models' $k \in \mathcal{K}$, and a prior distribution on models $p(k)$.
-For each model, there is a latent continuous parameter vector $\theta_k \in \mathbb{R}^{n(k)}$ where $n(k)$ is the dimension of model $k$, and a likelihood function $L_{D,k}(\theta_k)$ for each $k$ given data $D$.
-The latent state $x$ is a pair $(k, \theta_k)$ of model and continuous parameter.
+The reversible jump MCMC framework involves a set of `models' $h \in \mathcal{H}$, and a prior distribution on models $p(h)$.
+For each model, there is a latent continuous parameter vector $\theta_h \in \mathbb{R}^{n(h)}$ where $n(h)$ is the dimension of model $h$, and a likelihood function $L_{D,h}(\theta_h)$ for each $h$ given data $D$.
+The latent state $x$ is a pair $(h, \theta_h)$ of model and continuous parameter.
 There is a set of \emph{move types} $\mathcal{M}$.
-Each move type $m \in \mathcal{M}$ is associated with an unordered pair of models $(k_1, k_2)$ and a dimensionality $d(m)$ such that $d(m) \ge n(k_1)$ and $d(m) \ge n(k_2)$ (zero, one, or more than one move types maybe assocaite with a given pair of models).
-For each latent state $x = (k, \theta_k)$, there is a probability distribution $q_{x}(m)$ on move types such that $q_{x}(m) > 0$ implies that $k$ is one of the models for move type $m$.
-For each move type $m \in \mathcal{K}$ between $k_1$ and $k_2$ there is a pair of continuously differentiable bijections $h_{m, k_1 \to k_2} : \mathbb{R}^{d(m)} \to \mathbb{R}^{d(m)}$ and $h_{m, k_2 \to k_1} := h_{m, k_1 \to k_2}^{-1}$, and a pair of proposal densities $q_{m,k_1 \to k_2}(u_{k_1 \to k_2})$ and $q_{m,k_2 \to k_1}(u_{k_2 \to k_1})$ where $u_{k_1 \to k_2} \in \mathbb{R}^{d(m) - n(k_1)}$ and $u_{k_2 \to k_1} \in \mathbb{R}^{d(m) - n(k_2)}$.
-A proposal is made from state $x = (k, \theta_k)$ by (i) sampling a move type $m \sim q_{x}(\cdot)$, and (ii) sampling continuous variable $u \sim q_{m,k \to k'}(\cdot)$ for $(k, k')$ associated with $m$, and (iii) computing $(\theta_{k}', u') := h_{m,k \to k'}(\theta_k, u)$, and proposing new state $x' = (k', \theta_k')$.
+Each move type $m \in \mathcal{M}$ is associated with an unordered pair of models $(h_1, h_2)$ and a dimensionality $d(m)$ such that $d(m) \ge n(h_1)$ and $d(m) \ge n(h_2)$ (zero, one, or more than one move types may be associated with a given pair of models).
+For each latent state $x = (h, \theta_h)$, there is a probability distribution $q_{x}(m)$ on move types such that $q_{x}(m) > 0$ implies that $h$ is one of the models for move type $m$.
+For each move type $m \in \mathcal{M}$ between $h_1$ and $h_2$ there is a pair of continuously differentiable bijections $g_{m, h_1 \to h_2} : \mathbb{R}^{d(m)} \to \mathbb{R}^{d(m)}$ and $g_{m, h_2 \to h_1} := g_{m, h_1 \to h_2}^{-1}$, and a pair of proposal densities $q_{m,h_1 \to h_2}(u_{h_1 \to h_2})$ and $q_{m,h_2 \to h_1}(u_{h_2 \to h_1})$ where $u_{h_1 \to h_2} \in \mathbb{R}^{d(m) - n(h_1)}$ and $u_{h_2 \to h_1} \in \mathbb{R}^{d(m) - n(h_2)}$.
+A proposal is made from state $x = (h, \theta_h)$ by (i) sampling a move type $m \sim q_{x}(\cdot)$, and (ii) sampling continuous variable $u \sim q_{m,h \to h'}(\cdot)$ for $(h, h')$ associated with $m$, and (iii) computing $(\theta_{h}', u') := g_{m,h \to h'}(\theta_h, u)$, and proposing new state $x' = (h', \theta_h')$.
 The move is accepted with probability:
@@
 \min\left\{ 1,
-\frac{p(k')}{p(k)} \frac{p_{k'}(\theta_{k'})}{p_k(\theta_k)} \frac{L_{D,k'}(\theta_k')}{L_{D,k}(\theta_k)}
+\frac{p(h')}{p(h)} \frac{p_{h'}(\theta_{h'})}{p_h(\theta_h)} \frac{L_{D,h'}(\theta_h')}{L_{D,h}(\theta_h)}
 \frac{q_{x'}(m)}{q_{x}(m)}
-\frac{q_{m,k' \to k}(u)}{q_{m,k \to k'}(u)}
+\frac{q_{m,h' \to h}(u)}{q_{m,h \to h'}(u)}
 \right\}
@@
 
-To represent reversible jump MCMC using involutive MCMC implemented with probabilistic programs, we write a probabilistic program $\mathcal{P}$ that encodes the space of models $\mathcal{K}$, the prior distribution on models, $p(k)$, the per-model priors $p_k(\theta_k)$ and the per-model likelihoods $L_{D,k}(\theta_k)$.
-The set of all models $k$ corresponds to the set of all pairs $(A_P, \mathbf{d}_P)$ where $A_\mathcal{P}$ represent possible trace structures (i.e.
-control-flow paths through $\mathcal{P}$) and $\mathbf{d}_P$ are the set of assignments to discrete random choices made by $\mathcal{P}$.
+To represent reversible jump MCMC using involutive MCMC implemented with probabilistic programs, we write a probabilistic program $\mathcal{P}$ that encodes the space of models $\mathcal{H}$, the prior distribution on models, $p(h)$, the per-model priors $p_h(\theta_h)$ and the per-model likelihoods $L_{D,h}(\theta_h)$.
+The set of all models $h$ corresponds to the set of all pairs $(K_\mathcal{P}, \mathbf{d}_\mathcal{P})$ where $K_\mathcal{P}$ represent possible trace structures (i.e.
+control-flow paths through $\mathcal{P}$) and $\mathbf{d}_\mathcal{P}$ are the set of assignments to discrete random choices made by $\mathcal{P}$.
 The per-model continuous parameters $\theta$ correspond to continuous random choices $\mathbf{c}_\mathcal{P}$.
 % TODO explain the likelihood
-An auxiliary probabilistic program $\mathcal{Q}$ encodes both the probability distribution on moves types using discrete random choices and possibly stochastic control flow ($(A_\mathcal{Q}, \mathbf{d}_\mathcal{Q})$), and the per-move-type probability densities on $u$ using continuous random choices $\mathbf{c}_\mathcal{Q}$.
-The involution $f$ factors into an (i) involution $f_1$ on pairs $i = ((A_\mathcal{P}, \mathbf{d}_\mathcal{P}), (A_\mathcal{Q}, \mathbf{d}_\mathcal{Q}))$ that defines the association between move types $\mathcal{M}$ and the model pairs ($k_1, k_2$); and (ii) a family of bijections $f_{1,i}$ on the space of pairs $(\mathbf{c}_\mathcal{P}, \mathbf{c}_\mathcal{Q})$ of continuous random choices for both programs for fixed values of the discrete random choices and fixed trace structure.
+An auxiliary probabilistic program $\mathcal{Q}$ encodes both the probability distribution on moves types using discrete random choices and possibly stochastic control flow ($(K_\mathcal{Q}, \mathbf{d}_\mathcal{Q})$), and the per-move-type probability densities on $u$ using continuous random choices $\mathbf{c}_\mathcal{Q}$.
+The involution $f$ factors into an (i) involution $f_1$ on pairs $i = ((K_\mathcal{P}, \mathbf{d}_\mathcal{P}), (H_\mathcal{Q}, \mathbf{d}_\mathcal{Q}))$ that defines the association between move types $\mathcal{M}$ and the model pairs ($h_1, h_2$); and (ii) a family of bijections $f_{1,i}$ on the space of pairs $(\mathbf{c}_\mathcal{P}, \mathbf{c}_\mathcal{Q})$ of continuous random choices for both programs for fixed values of the discrete random choices and fixed trace structure.
 
@@
 {\begin{lstlisting}[basicstyle=\small\ttfamily]
-@copy(p_in, subtree_address, q_out, :new_subtree)
-@copy(q_in, :new_subtree, p_out, subtree_address)
+@copy(p_in, subtree_kddress, q_out, :new_subtree)
+@copy(q_in, :new_subtree, p_out, subtree_kddress)
 \end{lstlisting}}
@@
 %\end{equation}
-%where $\ell_x(A)$ is the probability of transitioning into set $A$ from $x$, and is given by:
+%where $\ell_x(A)$ is the probability of transitioning into set $K$ from $x$, and is given by:
 %\begin{equation}
@@
 Let $\Sigma_n$ and $\mu_n$ denote the Lebesgue $\sigma$-algebra and Lebesgue measure on $\mathbb{R}^n$, respectively.
-Let $\Sigma \subset \mathcal{P}(Z)$ be the $\sigma$-algebra of sets of the form $\cup_{i \in I} \{(i, \mathbf{x}) : \mathbf{x} \in A_i\}$ for some $A_i \in \Sigma_{n_i}$ for each $i \in I$.
+Let $\Sigma \subset \mathcal{P}(Z)$ be the $\sigma$-algebra of sets of the form $\cup_{i \in I} \{(i, \mathbf{x}) : \mathbf{x} \in K_i\}$ for some $K_i \in \Sigma_{n_i}$ for each $i \in I$.
 %Let $\Sigma$ be the $\sigma$-algebra on $Z$ generated by the sets $G := \{ \{(i, \mathbf{x}) : \mathbf{x} \in A\} : i \in I, A \in \Sigma_{n_i} \}$.
@@
 \begin{equation}
-\mu(\cup_{i \in I} \{(i, \mathbf{x}) : \mathbf{x} \in A_i\}) := \sum_{i \in I} \mu_{n_i}(A_i)
+\mu(\cup_{i \in I} \{(i, \mathbf{x}) : \mathbf{x} \in K_i\}) := \sum_{i \in I} \mu_{n_i}(K_i)
 \end{equation}
@@
 \end{equation}
-Consider $(\mu \circ f^{-1})(A)$ for $A \in \Sigma$:
+Consider $(\mu \circ f^{-1})(A)$ for $K \in \Sigma$:
 \begin{align}
-(\mu \circ f^{-1})(\cup_{i \in I} \{ (i, \mathbf{x}) : \mathbf{x} \in A_i \})
-&= \mu(f^{-1}(\cup_{i \in I} \{ (i, \mathbf{x}) : \mathbf{x} \in A_i \}))\\
-&= \mu(\cup_{i \in I} f^{-1}(\{ (i, \mathbf{x}) : \mathbf{x} \in A_i \}))\\
-&= \mu(\cup_{i \in I} \{ (f_1(i), \mathbf{x}) : \mathbf{x} \in f_{2,i}(A_i) \})\\
-&= \sum_{i \in I} \mu_{n_i}(f_{2,i}(A_i))
+(\mu \circ f^{-1})(\cup_{i \in I} \{ (i, \mathbf{x}) : \mathbf{x} \in K_i \})
+&= \mu(f^{-1}(\cup_{i \in I} \{ (i, \mathbf{x}) : \mathbf{x} \in K_i \}))\\
+&= \mu(\cup_{i \in I} f^{-1}(\{ (i, \mathbf{x}) : \mathbf{x} \in K_i \}))\\
+&= \mu(\cup_{i \in I} \{ (f_1(i), \mathbf{x}) : \mathbf{x} \in f_{2,i}(K_i) \})\\
+&= \sum_{i \in I} \mu_{n_i}(f_{2,i}(K_i))
 \end{align}
-It suffices to show that for all $A \in \Sigma$:
-\begin{equation}
-\int_A \left| (Jf_{2,i})(\mathbf{x})\right| \mu(dz) = \sum_{i \in I} \mu_{n_i}(f_{2,i}(A_i))
+It suffices to show that for all $K \in \Sigma$:
+\begin{equation}
+\int_K \left| (Jf_{2,i})(\mathbf{x})\right| \mu(dz) = \sum_{i \in I} \mu_{n_i}(f_{2,i}(K_i))
 \end{equation}
@@
 \begin{align}
-\int_A \left| (Jf_{2,i})(\mathbf{x})\right| \mu(dz)
-= \sum_{i \in I} \int_{A_i} \left| (Jf_{2,i})(\mathbf{x})\right| \mu_{n_i}(d \mathbf{x}) % TODO how to justify this? it seems obvious
-= \sum_{i \in I} \mu_{n_i}(f_{2,i}(A_i)) % Borgachev Theorem 4.7.1 
+\int_K \left| (Jf_{2,i})(\mathbf{x})\right| \mu(dz)
+= \sum_{i \in I} \int_{K_i} \left| (Jf_{2,i})(\mathbf{x})\right| \mu_{n_i}(d \mathbf{x}) % TODO how to justify this? it seems obvious
+= \sum_{i \in I} \mu_{n_i}(f_{2,i}(K_i)) % Borgachev Theorem 4.7.1 
 \end{align}
@@
 Given $\sigma$-finite measures $(X_P, \Sigma_P, \mu_P)$ and $(X_Q, \Sigma_Q, \mu_Q)$,
-if $X \in \Sigma_P \otimes \Sigma_Q$ then $\Sigma = \{A \in \Sigma_P \otimes \Sigma_Q : A \subseteq X\}$ is a $\sigma$-algebra on $X$, and $\mu : \Sigma \to [0, \infty)$ given by $\mu(A) := (\mu_P \times \mu_Q)(A)$ is $\sigma$-finite measure on measurable space $(X, \Sigma)$.
+if $X \in \Sigma_P \otimes \Sigma_Q$ then $\Sigma = \{A \in \Sigma_P \otimes \Sigma_Q : K \subseteq X\}$ is a $\sigma$-algebra on $X$, and $\mu : \Sigma \to [0, \infty)$ given by $\mu(A) := (\mu_P \times \mu_Q)(A)$ is $\sigma$-finite measure on measurable space $(X, \Sigma)$.
 \end{lemma}
@@
 Let $h(z)$ be a density for $\pi$ with respect to $\nu$.
-Let $A := \{z \in Z : h(z) > 0 \mbox{ and } h(f(z)) > 0\}$.
+Let $K := \{z \in Z : h(z) > 0 \mbox{ and } h(f(z)) > 0\}$.
 $k$ satisfies detailed balance with respect to $\pi$ if and only if:
 \begin{enumerate}
-\item $\alpha(z, f(z)) = 0$ for $\pi$-almost all $z \not \in A$
+\item $\alpha(z, f(z)) = 0$ for $\pi$-almost all $z \not \in K$
 \item $\alpha(z, f(z)) \frac{h(z)}{h(f(z))} = \alpha(f(z), z)$
@@
 $(\mu \circ f^{-1})(A) = 0$ implies $\mu(f^{-1}(A)) = 0$ implies $\nu(f^{-1}(A)) = 0$ implies $(\nu \circ f^{-1})(A) = 0$.
-To show that $z \mapsto (d \nu / d \mu)(f(z))$ is the Radon-Nikodym derivative $d (\nu \circ f^{-1}) / d (\mu \circ f^{-1})$, it suffices to show that for all $A \in \Sigma$:
-\begin{equation}
-\int_A \left( \frac{d \nu}{d \mu} (f(z)) \right) ( \mu \circ f^{-1} )(dz) = (\nu \circ f^{-1})(A) := \nu(f^{-1}(A))
-\end{equation}
-Applying Theorem 3.6.1 in Bogachev with $Y := A$, $y := z$, $x := z'$, $X := f^{-1}(A)$, and $g(y) := (d\nu / d\mu)(f(y))$:
+To show that $z \mapsto (d \nu / d \mu)(f(z))$ is the Radon-Nikodym derivative $d (\nu \circ f^{-1}) / d (\mu \circ f^{-1})$, it suffices to show that for all $K \in \Sigma$:
+\begin{equation}
+\int_K \left( \frac{d \nu}{d \mu} (f(z)) \right) ( \mu \circ f^{-1} )(dz) = (\nu \circ f^{-1})(A) := \nu(f^{-1}(A))
+\end{equation}
+Applying Theorem 3.6.1 in Bogachev with $Y := K$, $y := z$, $x := z'$, $X := f^{-1}(A)$, and $g(y) := (d\nu / d\mu)(f(y))$:
 \begin{align}
-\int_A \left( \frac{d \nu}{d \mu} (f(z)) \right) ( \mu \circ f^{-1} )(dz)
+\int_K \left( \frac{d \nu}{d \mu} (f(z)) \right) ( \mu \circ f^{-1} )(dz)
 &= \int_Y g(y) ( \mu \circ f^{-1} )(dy)\\
@@
 \begin{equation}
-\int_B \pi(z) k'_z(A) \mu(dz) = \int_A \pi(z) k'_z(B) \mu(dz) \;\; \mbox{ for all } A, B \in \Sigma
+\int_B \pi(z) k'_z(A) \mu(dz) = \int_K \pi(z) k'_z(B) \mu(dz) \;\; \mbox{ for all } K, B \in \Sigma
 \end{equation}
@@
 \begin{align}
-\int_Z \pi(z) k'_z(A) \mu(dz) &= \int_A \pi(z) k'_z(Z) \mu(dz) = \int_A \pi(z) \mu(dz) \;\; \mbox{ for all } A \in \Sigma\\
+\int_Z \pi(z) k'_z(A) \mu(dz) &= \int_K \pi(z) k'_z(Z) \mu(dz) = \int_K \pi(z) \mu(dz) \;\; \mbox{ for all } A \in \Sigma\\
 \end{align}
@@
 \begin{equation}
-\int_Z k'_z(A) \pi(z) \mu(dz) = \int_A \pi(z) \mu(dz) \mbox{ for all } A \in \Sigma
+\int_Z k'_z(A) \pi(z) \mu(dz) = \int_K \pi(z) \mu(dz) \mbox{ for all } A \in \Sigma
 \end{equation}
@@
 \begin{equation}
-\int_X k_x(A) p(x) \mu_P(dx) = \int_A p(x) \mu_P(dx) \mbox{ for all } A \in \Sigma_P
+\int_X k_x(A) p(x) \mu_P(dx) = \int_K p(x) \mu_P(dx) \mbox{ for all } A \in \Sigma_P
 \end{equation}
@@
     &= \int_{\{(x, u) \in Z : x \in A\}} q_x(u) p(x) \mu(dz)\\
-    &= \int_A \left( \int_U q_x(u) \mu_Q(du) \right) p(x) \mu_P(dx)\\
-    &= \int_A p(x) \mu_P(dx)
+    &= \int_K \left( \int_U q_x(u) \mu_Q(du) \right) p(x) \mu_P(dx)\\
+    &= \int_K p(x) \mu_P(dx)
 \end{align*}
@@
 %
-%the set of addresses $\mathcal{A}$ is partitioned into two spaces $\mathcal{A}_1$ and $\mathcal{A}_2$\\
+%the set of addresses $\mathcal{K}$ is partitioned into two spaces $\mathcal{K}_1$ and $\mathcal{K}_2$\\
 %
@@
 %
-%then $Z := \cup_{S \in \mathcal{S}} \{(S, \mathbf{x}) : \mathbf{x} \in \times_{a \in S} X_a \}$\\
+%then $Z := \cup_{S \in \mathcal{S}} \{(S, \mathbf{x}) : \mathbf{x} \in \times_{k \in S} X_k \}$\\
 %
@@
 %\[
-%\cup_{S \in \mathcal{S}} \{((\mathcal{A}_1 \cap S) \cup (\mathcal{A}_2 \cap S), \mathbf{x_1}, \mathbf{x_2}) : \mathbf{x_1} \in \times_{a \in S_1} X_a, \mathbf{x_2} \in \times_{a \in S_2} X_a\}
+%\cup_{S \in \mathcal{S}} \{((\mathcal{K}_1 \cap S) \cup (\mathcal{K}_2 \cap S), \mathbf{x_1}, \mathbf{x_2}) : \mathbf{x_1} \in \times_{k \in S_1} X_k, \mathbf{x_2} \in \times_{k \in S_2} X_k\}
 %\]
@@
 %\[
-%\cup_{S_1 \in \mathcal{S}} \{(S_1, \mathbf{x_1}) : \mathbf{x_1} \in \times_{a \in S_1}\}
+%\cup_{S_1 \in \mathcal{S}} \{(S_1, \mathbf{x_1}) : \mathbf{x_1} \in \times_{k \in S_1}\}
 %\]
@@
 %\pi(
-%\cup_{S \in \mathcal{S} : A_1 \cap S = S_1} \{(S_1 \cup (\mathcal{A}_2 \cap S), \mathbf{x_1}, \mathbf{x_2}) : \mathbf{x_1} \in B_{S_1}, \mathbf{x_2} \in \times_{a \in S_2} X_a\}
+%\cup_{S \in \mathcal{S} : K_1 \cap S = S_1} \{(S_1 \cup (\mathcal{K}_2 \cap S), \mathbf{x_1}, \mathbf{x_2}) : \mathbf{x_1} \in B_{S_1}, \mathbf{x_2} \in \times_{k \in S_2} X_k\}
 %) &= p(B_T)\\
-%\sum_{S \in \mathcal{S} : A_1 \cap S = S_1} \int_{B_{S_1} \times \otimes_{a \in S_2} X_a} \pi(S_1, S_2, \mathbf{x}_1, \mathbf{x}_2) \mu_{S_1 \cup S_2}(d \mathbf{x}_1 \mathbf{x}_2) &= p(B_T)\\
-%\sum_{S \in \mathcal{S} : A_1 \cap S = S_1} \int_{B_{S_1} \times \otimes_{a \in S_2} X_a} p(S_1, \mathbf{x}_1) q_{(S_1,\mathbf{x}_1)}(S_2, \mathbf{x}_2) \mu_{S_1 \cup S_2}(d \mathbf{x}_1 \mathbf{x}_2) &= p(B_T)\\
-%\sum_{S \in \mathcal{S} : A_1 \cap S = S_1} \int_{B_{S_1}} p(S_1, \mathbf{x}_1) \left( \int_{\otimes_{a \in S_2 X_a}} q_{(S_1,\mathbf{x}_1)}(S_2, \mathbf{x}_2) \mu_{S_2}(d \mathbf{x}_2) \right) \mu_{S_1}(d \mathbf{x}_1) &= p(B_T) [\mbox{Fubini / Tonnelli?}]\\
-%\sum_{S_2 \subseteq A_2 : S_1 \cup S_2 \in \mathcal{S}} \int_{B_{S_1}} p(S_1, \mathbf{x}_1) \left( \int_{\otimes_{a \in S_2 X_a}} q_{(S_1,\mathbf{x}_1)}(S_2, \mathbf{x}_2) \mu_{S_2}(d \mathbf{x}_2) \right) \mu_{S_1}(d \mathbf{x}_1) &= p(B_T)\\
-%\int_{B_{S_1}} p(S_1, \mathbf{x}_1) \left( \sum_{S_2 \subseteq A_2 : S_1 \cup S_2 \in \mathcal{S}} \int_{\otimes_{a \in S_2 X_a}} q_{(S_1,\mathbf{x}_1)}(S_2, \mathbf{x}_2) \mu_{S_2}(d \mathbf{x}_2) \right) \mu_{S_1}(d \mathbf{x}_1) &= p(B_T) [\mbox{Fubini / Tonnelli?}]\\
+%\sum_{S \in \mathcal{S} : K_1 \cap S = S_1} \int_{B_{S_1} \times \otimes_{k \in S_2} X_k} \pi(S_1, S_2, \mathbf{x}_1, \mathbf{x}_2) \mu_{S_1 \cup S_2}(d \mathbf{x}_1 \mathbf{x}_2) &= p(B_T)\\
+%\sum_{S \in \mathcal{S} : K_1 \cap S = S_1} \int_{B_{S_1} \times \otimes_{k \in S_2} X_k} p(S_1, \mathbf{x}_1) q_{(S_1,\mathbf{x}_1)}(S_2, \mathbf{x}_2) \mu_{S_1 \cup S_2}(d \mathbf{x}_1 \mathbf{x}_2) &= p(B_T)\\
+%\sum_{S \in \mathcal{S} : K_1 \cap S = S_1} \int_{B_{S_1}} p(S_1, \mathbf{x}_1) \left( \int_{\otimes_{k \in S_2 X_k}} q_{(S_1,\mathbf{x}_1)}(S_2, \mathbf{x}_2) \mu_{S_2}(d \mathbf{x}_2) \right) \mu_{S_1}(d \mathbf{x}_1) &= p(B_T) [\mbox{Fubini / Tonnelli?}]\\
+%\sum_{S_2 \subseteq K_2 : S_1 \cup S_2 \in \mathcal{S}} \int_{B_{S_1}} p(S_1, \mathbf{x}_1) \left( \int_{\otimes_{k \in S_2 X_k}} q_{(S_1,\mathbf{x}_1)}(S_2, \mathbf{x}_2) \mu_{S_2}(d \mathbf{x}_2) \right) \mu_{S_1}(d \mathbf{x}_1) &= p(B_T)\\
+%\int_{B_{S_1}} p(S_1, \mathbf{x}_1) \left( \sum_{S_2 \subseteq K_2 : S_1 \cup S_2 \in \mathcal{S}} \int_{\otimes_{k \in S_2 X_k}} q_{(S_1,\mathbf{x}_1)}(S_2, \mathbf{x}_2) \mu_{S_2}(d \mathbf{x}_2) \right) \mu_{S_1}(d \mathbf{x}_1) &= p(B_T) [\mbox{Fubini / Tonnelli?}]\\
 %\end{align}
@@
 %\begin{align}
-%\sum_{S_2 \subseteq A_2 : S_1 \cup S_2 \in \mathcal{S}} \int_{\otimes_{a \in S_2 X_a}} q_{(S_1,\mathbf{x}_1)}(S_2, \mathbf{x}_2) \mu_{S_2}(d \mathbf{x}_2) = 1 \mbox{ for all } (S_1, \mathbf{x}_1) \in T
+%\sum_{S_2 \subseteq K_2 : S_1 \cup S_2 \in \mathcal{S}} \int_{\otimes_{k \in S_2 X_k}} q_{(S_1,\mathbf{x}_1)}(S_2, \mathbf{x}_2) \mu_{S_2}(d \mathbf{x}_2) = 1 \mbox{ for all } (S_1, \mathbf{x}_1) \in T
 %\end{align}
@@
 %\[
-%\mathcal{S}_2 = \{S_2 \subseteq A_2 : S_1 \cup S_2 \in \mathcal{S}\}
+%\mathcal{S}_2 = \{S_2 \subseteq K_2 : S_1 \cup S_2 \in \mathcal{S}\}
 %\]
@@
 %\[
-%B_1 = \cup_{S_1 \in \mathcal{S}_1} \{(S_1, \mathbf{x}_1) : \mathbf{x}_1 \in B_{S_1}\}\;\;\; \mbox{where} \;\; B_{S_1} \in \otimes_{a \in S_1} \Sigma_{a} \;\;\; [\mbox{Measurable set } B_1]
+%B_1 = \cup_{S_1 \in \mathcal{S}_1} \{(S_1, \mathbf{x}_1) : \mathbf{x}_1 \in B_{S_1}\}\;\;\; \mbox{where} \;\; B_{S_1} \in \otimes_{k \in S_1} \Sigma_{k} \;\;\; [\mbox{Measurable set } B_1]
 %\]
@@
 Note that we the same measure space $(T, \Sigma, \mu)$ on traces for both the model and auxiliary probabilistic program.
-If both programs sample some random choice at the same address $a$ and require different measure spaces $(V_a, \Sigma_a, \mu_a)$ for the address, this can be handled formally by making the addresses sampled by the two programs disjoint by prepending either the prefix `p.' or `q.' to addresses.
+If both programs sample some random choice at the same address $k$ and require different measure spaces $(V_k, \Sigma_k, \mu_k)$ for the address, this can be handled formally by making the addresses sampled by the two programs disjoint by prepending either the prefix `p.' or `q.' to addresses.
 This is only required for book-keeping in the formalism, and these prefixes do not need to be actually added to the addresses used in the programs.}
@@
 We now give a sufficient condition for this to hold.
-Let $D \subseteq \mathcal{A}$ denote the subset of addresses that are discrete (i.e. where $V_a$ is a countable set and $\mu_a$ is the counting measure).
-For $\mathbf{x} \in \times_{a \in A} V_a$ let $\mathbf{x} = (\mathbf{d}, \mathbf{c})$ where $\mathbf{d} \in \times_{a \in A \cap D} V_a$ and $\mathbf{c} = \times_{a \in A \setminus D} V_a$, so that $\mathbf{d}$ is the discrete part of $\mathbf{x}$ and $\mathbf{c}$ is the non-discrete part.
+Let $D \subseteq \mathcal{K}$ denote the subset of addresses that are discrete (i.e. where $V_k$ is a countable set and $\mu_k$ is the counting measure).
+For $\mathbf{x} \in \times_{k \in A} V_k$ let $\mathbf{x} = (\mathbf{d}, \mathbf{c})$ where $\mathbf{d} \in \times_{k \in A \cap D} V_k$ and $\mathbf{c} = \times_{k \in A \setminus D} V_k$, so that $\mathbf{d}$ is the discrete part of $\mathbf{x}$ and $\mathbf{c}$ is the non-discrete part.
  %TODO make this a proposition
 \begin{lemma}
-Suppose that $p$ and $q$ are such that $p(A, (\mathbf{d}, \mathbf{c})) > 0$ implies
-$p(A, (\mathbf{d}, \mathbf{c}')) > 0$ for all $\mathbf{c}' \in \prod_{a \in A \setminus D} V_a$,
-and that $q_x(A, (\mathbf{d}, \mathbf{c})) > 0$ where $x = (\tilde{A}, \tilde{\mathbf{d}}, \tilde{\mathbf{c}})$ implies that
-$q_{x'}(A, (\mathbf{d}, \mathbf{c}')) > 0$ for all $\mathbf{c}' \in \prod_{a \in A \setminus D} V_a$ and all $x' = (\tilde{A}, \tilde{\mathbf{d}}, \tilde{\mathbf{c}}')$ where $\tilde{\mathbf{c}}' \in \prod_{a \in \tilde{A} \setminus D} V_a$.
+Suppose that $p$ and $q$ are such that $p(K, (\mathbf{d}, \mathbf{c})) > 0$ implies
+$p(K, (\mathbf{d}, \mathbf{c}')) > 0$ for all $\mathbf{c}' \in \prod_{k \in A \setminus D} V_k$,
+and that $q_x(K, (\mathbf{d}, \mathbf{c})) > 0$ where $x = (\tilde{A}, \tilde{\mathbf{d}}, \tilde{\mathbf{c}})$ implies that
+$q_{x'}(K, (\mathbf{d}, \mathbf{c}')) > 0$ for all $\mathbf{c}' \in \prod_{k \in A \setminus D} V_k$ and all $x' = (\tilde{A}, \tilde{\mathbf{d}}, \tilde{\mathbf{c}}')$ where $\tilde{\mathbf{c}}' \in \prod_{k \in \tilde{A} \setminus D} V_k$.
 Then, $Z := \{(x, y) \in X \times Y : \pi(x, y) > 0\}$ is $\mu \times \mu$-measurable where $\mu$ is the reference measure on traces.
@@
 For $p$ and $q$ satisfying these conditions,
-$Z = \cup_{(A_1,A_2,\mathbf{d}_1,\mathbf{d}_2) \in E} \{((A_1, (\mathbf{d}_1, \mathbf{c}_1)), (A_2, (\mathbf{d}_2, \mathbf{c}_2))) : \mathbf{c}_1 \in \times_{a \in A_1 \setminus D} V_a, \mathbf{c}_2 \in \times_{a \in A_2 \setminus D} V_a\}$ for some countable set $E \subseteq \{(A_1, A_2, \mathbf{d}_1, \mathbf{d}_2) : A_1, A_2 \subseteq \mathcal{A}, |A_1| < \infty, |A_2| < \infty, \mathbf{d}_1 \in \times_{a \in A_1 \cap D} V_a, \mathbf{d}_2 \in \times_{a \in A_2 \cap D} V_a\}$ of address sets and discrete choice values for both programs.
+$Z = \cup_{(K_1,K_2,\mathbf{d}_1,\mathbf{d}_2) \in E} \{((K_1, (\mathbf{d}_1, \mathbf{c}_1)), (K_2, (\mathbf{d}_2, \mathbf{c}_2))) : \mathbf{c}_1 \in \times_{k \in K_1 \setminus D} V_k, \mathbf{c}_2 \in \times_{k \in K_2 \setminus D} V_k\}$ for some countable set $E \subseteq \{(K_1, K_2, \mathbf{d}_1, \mathbf{d}_2) : K_1, K_2 \subseteq \mathcal{K}, |K_1| < \infty, |K_2| < \infty, \mathbf{d}_1 \in \times_{k \in K_1 \cap D} V_k, \mathbf{d}_2 \in \times_{k \in K_2 \cap D} V_k\}$ of address sets and discrete choice values for both programs.
 The measure of $Z$ is
-$(\mu \times \mu)(Z) = \sum_{(A_1,A_2,\mathbf{d}_1,\mathbf{d}_2) \in E} \prod_{a \in (A_1 \setminus D) \cup (A_2 \setminus D)} \mu_a(V_a)$.
+$(\mu \times \mu)(Z) = \sum_{(K_1,K_2,\mathbf{d}_1,\mathbf{d}_2) \in E} \prod_{k \in (K_1 \setminus D) \cup (K_2 \setminus D)} \mu_k(V_k)$.
 % TODO double check me.,

### 2020-07-17 paper.tex
tweaks to sec 3 and 4

--- 2020-07-17 paper.tex
@@
 
-In this section, we introduce the Gen language~\citep{gen-pldi} for probabilistic programs, and present a technique for automating the implementation of involutive MCMC algorithms when the model $p$ and auxiliary distribution $q_x$ are both represented as
-Gen programs.
-
-\subsubsection{A Language for Probabilistic Programs with Traces}
-
-The Gen probabilistic programming language augments the syntax of Julia~\citep{julia}
-with a single new construct, the \texttt{\{a\} $\sim$ d} expression, for making a \textit{named} random choice.
+In this section, we introduce a probabilistic programming language from the Gen probabilistic programming system~\citep{cusumano2019gen}, and present a technique for automating the implementation of involutive MCMC algorithms when the model $p$ and auxiliary distribution $q_x$ are both represented as
+probabilistic programs in this language.
+But the technique is not limited to the Gen system---we also provide the implementation of a minimal probabilistic programming language in PyTorch that supports the automation technique.
+
+\subsubsection{A Probabilistic Programming Language}
+
+Our probabilistic programming language augments the syntax of Julia~\citep{bezanson2017julia}
+with a new construct, the \texttt{\{a\} $\sim$ d} expression, for making a \textit{named} random choice.
 An execution trace of a Gen probabilistic program can be sampled by running the program according to Julia's usual semantics, and upon encountering an expression of the form \texttt{\{a\} $\sim$ d}, (i) evaluating the \textit{address expression} \texttt{a} to obtain an address $a \in \mathcal{A}$; (ii) evaluating the \textit{distribution expression} $d$ to obtain a distribution $d$ over the measurable space $V_a$; (iii) sampling a value $x_a \sim d$ and adding the mapping $\{a \mapsto x_a\}$ to a dictionary called the program trace; and (iv) returning the sampled value $x_a$ to the program, to continue execution.
@@
 
-As a concrete example, consider Figure~\ref{fig:mixture}a, which defines a Gen program for a Gaussian mixture model with an unknown number of components.
-The program \texttt{p} accepts as input an integer \texttt{n}, a number of data points.
-Each $\texttt{n}$ defines
+As a concrete example, consider the probabilistic program in Figure~\ref{fig:mixture}a, which defines a Gaussian mixture model with an unknown number of components:
+\noindent
+\begin{center}
+\begin{tabular}{c}
+{\begin{lstlisting}[basicstyle=\small\ttfamily]
+@gen function p(n::Int)
+  k ~ poisson_plus_one(1)
+  means = [
+    ({(:mu, j)} ~ normal(0, 10)) for j in 1:k]
+  vars = [
+    ({(:var, j)} ~ inv_gamma(1, 10)) for j in 1:k]
+  weights ~ dirichlet([2.0 for j in 1:k])
+  for i in 1:n
+    {(:x, i)} ~ mixture_of_normals(
+        weights, means, vars)
+  end
+end
+\end{lstlisting}}
+\end{tabular}
+\end{center}
+The program \texttt{p} accepts as input an integer \texttt{n}, a number of data points. Each $\texttt{n}$ defines
 a distinct distribution over dictionaries.

### 2020-07-17 paper.tex
tweaks to section 4

@@
 
-As a concrete example, consider the probabilistic program in Figure~\ref{fig:mixture}a, which defines a Gaussian mixture model with an unknown number of components:
+As a concrete example, consider the probabilistic program in Figure~\ref{fig:mixture}b, which defines a Gaussian mixture model with an unknown number of components:
 \noindent
@@
 
-Given two probabilistic programs $\mathcal{P}$ and $\mathcal{Q}$ satisfying a simple technical requirement (described in Section~\ref{sec:technical-requirement}), we can interpret them as specifying the model density $p$ and auxiliary densities $q_x$ for use with involutive MCMC on dictionaries (Algorithm~\ref{alg:involutive-mcmc}).
+If the two probabilistic programs $\mathcal{P}$ and $\mathcal{Q}$ satisfy a simple technical requirement (described in Section~\ref{sec:technical-requirement}), we can interpret them as specifying the model density $p$ and auxiliary densities $q_x$ for use with involutive MCMC on dictionaries (Algorithm~\ref{alg:involutive-mcmc}).
 (In this case, the program $\mathcal{Q}$
 accepts a trace $x$ of $\mathcal{P}$ as input, and the distribution for $q_x$ is given by $\mathcal{Q}_x$.)
-The acceptance probability of involutive MCMC includes the following ratio of densities:
-\begin{equation}
-\frac{p(x') q_{x'}(y')}{p(x) q_{x}(y)}
-\end{equation}
-
-This ratio can be computed term-wise using \textsc{score} on the probabilistic programs $\mathcal{P}$ and $\mathcal{Q}$.
 
@@
 Then $\tilde{p}(x \oplus b) = \mathcal{L}(b)p(x)$.
-If we use $\tilde{p}(x \oplus b)$ in place of $p(x)$ to compute the ratio above, we will wind up with the same output, because $\mathcal{L}(b)$ will cancel in the numerator and denominator.
-
-Thus, this ratio can be computed term-by-term using \textsc{score} (Algorithm~\ref{alg:auto-involutive-mcmc}): to compute $q_x(y')$, we run the algorithm directly on the program $\mathcal{Q}_x$, and to compute $p(x)$ and $p(x')$, we actually merge the dictionary $x$ with the observations $b$ and compute $\mathcal{L}(b)p(x) = \tilde{p}(x \oplus b)$ instead, by running the algorithm on $\mathcal{P}$ with trace $x \oplus b$.
+If we use $\tilde{p}(x \oplus b)$ in place of $p(x)$ to compute the ratio of densities in Algorithm~\ref{alg:involutive-mcmc}, we will wind up with the same output, because $\mathcal{L}(b)$ will cancel in the numerator and denominator:
+\begin{equation}
+\frac{p(x') q_{x'}(y')}{p(x) q_{x}(y)}
+= \frac{\tilde{p}(x' \oplus b) q_{x'}(y')}{\tilde{p}(x \oplus b) q_x(y)}
+= \frac{\tilde{p}(x' | b) q_{x'}(y')}{\tilde{p}(x | b) q_x(y)}
+\end{equation}
+
+Thus, this ratio can be computed term-by-term using \textsc{score} on the probabilistic programs $\mathcal{P}$ and $\mathcal{Q}$: (Algorithm~\ref{alg:auto-involutive-mcmc}): to compute $q_x(y')$, we run the algorithm directly on the program $\mathcal{Q}_x$, and to compute $p(x)$ and $p(x')$, we actually merge the dictionary $x$ with the observations $b$ and compute $\mathcal{L}(b)p(x) = \tilde{p}(x \oplus b)$ instead, by running the algorithm on $\mathcal{P}$ with trace $x \oplus b$.
 The density $q_x(y)$ can be computed while $y$ is being sampled, using \textsc{trace-and-score}.
@@
 Section~\ref{sec:sparsity} describes a more efficient approach that exploits sparsity in the involution and resulting cancellation in the ratio for improved efficiency, but requires a more sophisticated probabilistic programming runtime system.
+
+Figure~\ref{fig:mixture}b and Figure~\ref{fig:mixture}c show examples of probabilistic programs $\mathcal{P}$ and $\mathcal{Q}$ respectively, for a split-merge reversible jump move.
 

### 2020-07-17 paper.tex
clean up 4.2

@@
 
+% TODO consider adding this back, to explain to reader why they are reading this:
 %The general measure-theoretic formulation of involutive MCMC in Algorithm~\ref{alg:involutive-mcmc} does not specify how to compute the Radon-Nikodym derivative that is required for the acceptance probability.
@@
 Then, the acceptance probability in Algorithm~\ref{alg:involutive-mcmc} simplifies to:
-\begin{equation}
+\begin{equation} \label{eq:acceptance-ratio-dictionaries}
 \frac{p(x') q_{x'}(y')}{p(x) q_{x}(y)} \cdot |J h_e|(z)
@@
     \execute[$\mathcal{F}$], but with
-        \State $\mathtt{"write}(a, v)\mathtt{"} \equiv$
+        \State $\mathtt{"@write}(a, v)\mathtt{"} \equiv$
             (\algorithmicif\hspace{1mm} $a \in \mathcal{I}$
@@
             \algorithmicelse\hspace{1mm} set $z'[a] \gets v$ and set $W \gets W \cup \{a\})$
-        \State $\mathtt{"read}(a)\mathtt{"} \equiv$
+        \State $\mathtt{"@read}(a)\mathtt{"} \equiv$
             (\algorithmicif\hspace{1mm} $a \in \mathcal{I}$
@@
             \algorithmicelse\hspace{1mm} set $R \gets R \cup \{a\}$ and evaluate to $z[a]$)
-        \State $\mathtt{"copy}(a_1, a_2)\mathtt{"} \equiv$ (set $z'[a_2] \gets z[a_1]$ and set $C \gets C \cup \{a_1\}$)
+        \State $\mathtt{"@copy}(a_1, a_2)\mathtt{"} \equiv$ (set $z'[a_2] \gets z[a_1]$ and set $C \gets C \cup \{a_1\}$)
     \END
@@
 
-\subsection{Trace-Based Probabilistic Programming}
+\subsection{Trace-Based Probabilistic Programming} \label{sec:probabilistic-programming}
 
@@
 
-The previous section described how the probability densities ($p$ and $q$) required for involutive MCMC on spaces of dictionaries can be specified and automatically computed using probabilistic programs.
-This section introduces a lightweight differentiable programming language for specifying involutions ($f : Z \to Z$) on dictionaries.
-The interpreter for the language computes the Radon-Nikodym derivative required for the acceptance probability using automatic differentiation.
-The language is implemented as part of the Gen system~\citep{cusumano2019gen}, which is embedded in the Julia language.
-We also provide a simplified PyTorch-based implementation of the language.
-
+Section~\ref{sec:probabilistic-programming} showed that if the densities $p$ and $q$ are specified using probabilistic programs $\mathcal{P}$ and $\mathcal{Q}$, then density ratio in the acceptance probability for involutive MCMC on dictionaries Equation~(\ref{eq:acceptance-ratio-dictionaries}) can be automated using probabilistic programming techniques.
+This section shows that if the involution $f$ is specified in using a differentiable program $\mathcal{F}$ that transforms the execution traces of probabilistic programs, then the Jacobian factor in Equation~(\ref{eq:acceptance-ratio-dictionaries}) can be automated using automatic differentiation.
+The procedure \textproc{auto-involutive-mcmc} in Algorithm~\ref{alg:auto-involutive-mcmc} combines these two ideas and automates involutive MCMC given the programs $\mathcal{P}$, $\mathcal{Q}$, and $\mathcal{F}$.
+
+%The previous section described how the probability densities ($p$ and $q$) required for involutive MCMC on spaces of dictionaries can be specified and automatically computed using probabilistic programs.
+%This section introduces a lightweight differentiable programming language for specifying involutions ($f : Z \to Z$) on dictionaries.
+%The interpreter for the language computes the Radon-Nikodym derivative required for the acceptance probability using automatic differentiation.
+%The language is implemented as part of the Gen system~\citep{cusumano2019gen}, which is embedded in the Julia language.
+%We also provide a simplified PyTorch-based implementation of the language.
+
+\subsubsection{A Differentiable Programming Language for Manipulating Traces}
 Recall that for involutive MCMC on a state space of dictionaries, we define $Z := \{x \oplus y : p(x) q_x(y) > 0\} \subseteq \mathcal{D}$, where $x \oplus y$ is the dictionary resulting from merging dictionaries $x$ and $y$ with disjoint keys.
-Intuitively, $z = x \oplus y$ the trace of a composite probabilistic program constructed by running $\mathcal{P}$ first, and then running $\mathcal{Q}$, passing the trace of $\mathcal{P}$ as its input.
-
-\subsubsection{A Differentiable Programming Language for Manipulating Traces}
-We denote a program in the differentiable language by $\mathcal{F}$ and the involution it defines as $f$.
-Like the probabilistic programming language described in the previous section, our differentiable language for involutions on traces is embedded based on Julia.
-The differentiable language has specialized expressions for reading values from addresses in a trace, writing values to the addresses in a trace, and copying values from an address in one trace to an address in another trace.
-For example, to read a value from a trace, we use the \texttt{@read} keyword:
+The involution is a function $f : Z \to Z$.
+%Intuitively, $z = x \oplus y$ the trace of a composite probabilistic program constructed by running $\mathcal{P}$ first, and then running $\mathcal{Q}$, passing the trace of $\mathcal{P}$ as its input.
+
+We now introduce a simple differentiable programming language for specifying involutions $f$.
+The language needs to have syntax for reading the value from an address in $x \oplus y \in Z$ and writing to an address in $x' \oplus y' \in Z$.
+To read a value $(x \oplus y)[a]$ at address $a$, we use the \texttt{@read} keyword:
 \noindent
@@
 {\begin{lstlisting}[basicstyle=\small\ttfamily]
-    value = @read(<trace>, <address>, <type>)
-\end{lstlisting}}
-\end{tabular}
-\end{center}
-The first argument defines what trace we are reading from.
-The second argument is the address in that trace, which may be hierarchical.\footnote{
-Hierarchical addresses of the form $\mathtt{a.b.c}$ are constructed in the language using the syntax \texttt{a=>b=>c}.}
-The third argument is either $\mathtt{:}\mathtt{discrete}$ or $\mathtt{:}\mathtt{continuous}$, and informs the language runtime whether the random choice at that address is drawn from a discrete or continuous distribution (this information will be used to support automatic differentiation).
-Similarly, to write a value to a trace, we use a \texttt{@write} keyword:
+    value = @read(<address>, <type>)
+\end{lstlisting}}
+\end{tabular}
+\end{center}
+The first argument is the address $a$ and the second argument is either $\mathtt{discrete}$ or $\mathtt{continuous}$, and informs the interpreter whether the random choice at that address is drawn from a discrete or continuous distribution (this information will be used to support automatic differentiation).
+Recall that $x$ is the trace of the model probabilistic program, and $y$ is the trace of the auxiliary probabilistic program.
+Each address $a$ therefore is a pair, where the first element is either $\mathtt{model}$ or $\mathtt{aux}$, and the second element is the address within either $x$ or $y$.
+For example, to read the value of a continuous address $\mathtt{:}\mathtt{r}$ from the model trace ($x$):
 \noindent
@@
 {\begin{lstlisting}[basicstyle=\small\ttfamily]
-    @write(<trace>, <address>, <value>, <type>)
-\end{lstlisting}}
-\end{tabular}
-\end{center}
-The mathematical behavior of the function is defined simply by writing Julia code that implements it.
-For example, the following code reads the values of parameters in polar coordinates and writes out their values in cartesian coordinates:
+    val = @read((model, :r), continuous)
+\end{lstlisting}}
+\end{tabular}
+\end{center}
+The syntax for writing to an address in $x' \oplus y' \in Z$ is similar.
+For example to write a value $\mathtt{val}$ to $x'[\mathtt{:}\mathtt{r}]$:
 \noindent
@@
 {\begin{lstlisting}[basicstyle=\small\ttfamily]
-r = @read(trace1, :r, :continuous)
-theta = @read(trace1, :theta, :continuous)
+    @write((model, :r), val, continuous)
+\end{lstlisting}}
+\end{tabular}
+\end{center}
+Note that the input traces $x, y$ are distinct from the output traces $x', y'$;
+input traces can only be read from, and output traces can only be written to.
+For example, it is not possible to write an output trace and then read the written value from it later.
+Often, we want to simply copy the value from some address in the input traces to some address in the output trace.
+While this is possible via a $\mathtt{@read}$ followed by a $\mathtt{@write}$, the language provides a special syntax for copies:
+\noindent
+\begin{center}
+\begin{tabular}{c}
+{\begin{lstlisting}[basicstyle=\small\ttfamily]
+    @copy(<from-address>, <to-address>)
+\end{lstlisting}}
+\end{tabular}
+\end{center}
+Like above, the two addresses here are both pairs of the trace and the address within that trace.
+For example, to copy the value from address $\mathtt{:}\mathtt{u}$ in $x$ to address $\mathtt{:}\mathtt{v}$ in $x'$, we use:
+\noindent
+\begin{center}
+\begin{tabular}{c}
+{\begin{lstlisting}[basicstyle=\small\ttfamily]
+    @copy((model, :u), (model, :v))
+\end{lstlisting}}
+\end{tabular}
+\end{center}
+Of course, it is also possible to copy from $x$ to $y'$, from $y$ to $x'$ and from $y$ to $y'$.
+As we will see in Section~\ref{sec:sparsity}, it is preferable to copy when possible instead of reading and then writing, as this can improve the efficiency of automatic differentiation.
+
+%We denote a program in the differentiable language by $\mathcal{F}$ and the involution it defines as $f$.
+%Like the probabilistic programming language described in the previous section, our differentiable language for involutions on traces is embedded based on Julia.
+%The differentiable language has specialized expressions for reading values from addresses in a trace, writing values to the addresses in a trace, and copying values from an address in one trace to an address in another trace.
+%The first argument defines what trace we are reading from.
+%The second argument is the address in that trace, which may be hierarchical.\footnote{
+%Hierarchical addresses of the form $\mathtt{a.b.c}$ are constructed in the language using the syntax \texttt{a=>b=>c}.}
+%The third argument is either $\mathtt{:}\mathtt{discrete}$ or $\mathtt{:}\mathtt{continuous}$, and informs the language runtime whether the random choice at that address is drawn from a discrete or continuous distribution (this information will be used to support automatic differentiation).
+%Similarly, to write a value to a trace, we use a \texttt{@write} keyword:
+%\noindent
+%\begin{center}
+%\begin{tabular}{c}
+%{\begin{lstlisting}[basicstyle=\small\ttfamily]
+    %@write(<trace>, <address>, <value>, <type>)
+%\end{lstlisting}}
+%\end{tabular}
+%\end{center}
+%The mathematical behavior of the function is defined simply by writing Julia code that implements it.
+\paragraph{Constructing an Involution}
+The following code reads the values of parameters from the model in polar coordinates and writes out their values in cartesian coordinates:
+\noindent
+\begin{center}
+\begin{tabular}{c}
+{\begin{lstlisting}[basicstyle=\small\ttfamily]
+r = @read((model, :r), :continuous)
+theta = @read((model, :theta), :continuous)
 x = r * sin(theta)
 y = r * cos(theta)
-@write(trace2, :x, x, :continuous)
-@write(trace2, :y, y, :continuous)
-\end{lstlisting}}
-\end{tabular}
-\end{center}
-It is also possible to copy values from an address in one trace to an address in another trace:
-\noindent
-\begin{center}
-\begin{tabular}{c}
-{\begin{lstlisting}[basicstyle=\small\ttfamily]
-@copy(<from-trace>, <from-addr>,
-      <to-trace>, <to-addr>)
-\end{lstlisting}}
-\end{tabular}
-\end{center}
-As we will see in Section~\ref{sec:sparsity}, it is preferable to copy when possible instead of reading and then writing, as the additional information can improve the efficiency of automatic differentiation.
-
-The names of the input traces and output traces are defined as part of the function signature, which as the following form, where \texttt{h} is the name of the function being defined.
-\noindent
-\begin{center}
-\begin{tabular}{c}
-{\begin{lstlisting}[basicstyle=\small\ttfamily]
-@tracefn h from (tr1, tr2) to (tr3, tr4) begin
-    ..
-end
-\end{lstlisting}}
-\end{tabular}
-\end{center}
- % TODO change it from trace_bijection to tracefn in the figures and implementation...
-Each input and output trace is given a name (e.g. \texttt{tr1}, \texttt{tr2}, \texttt{tr3}, and \texttt{tr4}).
-For involution MCMC, the function must have two input traces and two output traces.
-The first input trace is the input model trace ($x$), the second input trace is the input auxiliary trace ($y$), the first output trace is the output model trace ($x'$) and the second output trace is the output auxiliary trace ($y'$).
-
-\subsubsection{Reasoning About Involutions}
-Recall the earlier code snippet that maps from polar to cartesian coordinates.
+@write((model, :x), x, :continuous)
+@write((model, :y), y, :continuous)
+\end{lstlisting}}
+\end{tabular}
+\end{center}
+
 This code snippet defines a bijection from one coordinate system to another.
@@
 How can we make the function an involution?
-
-First, the function will need the ability to go from cartesian coordinates to polar coordinates, so the following code for the inverse of the forward transformation must appear in the function somewhere:
-\noindent
-\begin{center}
-\begin{tabular}{c}
-{\begin{lstlisting}[basicstyle=\small\ttfamily]
-r = (x^2 + y^2)
-theta = atan(y, x)
-\end{lstlisting}}
-\end{tabular}
-\end{center}
-To be an involution, the input and output traces must contain some information indicating whether they are using polar or cartesian coordinates, so the involution code knows which of the two transformations to read from, and which addresses to read from and write to.
+First, the function will need the ability to go from cartesian coordinates to polar coordinates, so the code for the inverse of the forward transformation must appear in the function somewhere.
+Also, to be an involution, the input and output traces must contain some information indicating whether they are using polar or cartesian coordinates, so the involution code knows which of the two transformations to read from, and which addresses to read from and write to.
 Below is a minimal example of the involution, that uses a Boolean address polar.
@@
 {\begin{lstlisting}[basicstyle=\small\ttfamily]
-@tracefn h from (tr1, tr2) to (tr3, tr4)
-   polar = @read(tr1, :polar, :discrete)
+@tracefn f
+   polar = @read((model, :polar), discrete)
    if polar
-      r = @read(tr1, :r, :continuous)
-      theta = @read(tr1, :theta, :continuous)
+      r = @read((model, :r), continuous)
+      theta = @read((model, :theta), continuous)
       x = r * sin(theta)
       y = r * cos(theta)
-      @write(tr3, :x, x, :continuous)
-      @write(tr3, :y, y, :continuous)
+      @write((model, :x), x, continuous)
+      @write((model, :y), y, continuous)
    else
-      x = @read(tr1, :x, :continuous)
-      y = @read(tr1, :y, :continuous)
+      x = @read((model, :x), continuous)
+      y = @read((model, :y), continuous)
       r = (x^2 + y^2)
       theta = atan(y, x)
-      @write(tr3, :r, r, :continuous)
-      @write(tr3, :theta, theta, :continuous)
-   end
-   @write(tr3, :polar, !polar, :discrete)
-end
-\end{lstlisting}}
-\end{tabular}
-\end{center}
-In order to apply this function in the context of involution MCMC, we require probabilistic programs $\mathcal{P}$ and $\mathcal{Q}$ that define space of input and output traces for the involution.
-A minimal example does not require any auxiliary random choices, so $\mathcal{Q}$ can be an empty probabilistic program.
-A minimal toy example of a probabilistic program $\mathcal{P}$ that is compatible with the involution \texttt{h} defined above is:
+      @write((model, :r), r, continuous)
+      @write((model, :theta), theta, continuous)
+   end
+   @write((model, :polar), !polar, discrete)
+end
+\end{lstlisting}}
+\end{tabular}
+\end{center}
+In order to apply this function in the context of involutive MCMC, we require probabilistic programs $\mathcal{P}$ and $\mathcal{Q}$ that define space of input and output traces for the involution.
+The involution only reads and writes to model addresses, so $\mathcal{Q}$ can be an empty program:
 \noindent
@@
 end
-\end{lstlisting}}
-\end{tabular}
-\end{center}
-
-\subsubsection{Example: Split-Merge Reversible Jump}
-
-Figure~\ref{fig:mixture}(d) shows an involution expressed in the differentiable programming language introduced in this section, and Figure~\ref{fig:mixture}(f) shows a schematic of the information flow through the involution as it is applied twice in a row.
-The involution implements a split or a merge proposal in the infinite mixture model of Figure~\ref{fig:mixture}(b), depending on the trace of the auxiliary program in Figure~\ref{fig:mixture}(c).
-If there are at least two clusters, and the value of address $\mathtt{split}$ is false, then the involution proposes to merge two clusters by running code block 2 shown in green.
-Otherwise, the involution proposes to split a cluster by running code block 1 shown in yellow.
-Note that that code block 1 (yellow) is itself an involution on the discrete addresses in the traces---each application of code block 1 to the traces causes them to flip back and forth between a split and a merge proposal.
-Also note that code blocks 2 and 3 are continuously differentiable bijections on the continuous addresses in the traces, and they are inverses of one another.
+
+@gen function q(x)
+end
+\end{lstlisting}}
+\end{tabular}
+\end{center}
 
 \subsubsection{Computing the Jacobian with Automatic Differentiation}
-Recall that the involution $f$ is composed of (i) an involution $g$ on elements $e$ of a partition of the state space, and (ii) a pair of continuous differentiable bijections $h_{e}$ and $h_{g(e)} = h_{e}^{-1}$ between each pair of corresponding elements of the partition.
+
+Recall that the involution $f$ must decompose into (i) an involution $g$ on elements $e \in E$ of a partition of the state space, and (ii) a pair of continuous differentiable bijections $h_{e}$ and $h_{g(e)} = h_{e}^{-1}$ between each pair of corresponding elements of the partition.
 Each function $h_e$ is a function from the values at continuous addresses in the input trace ($z[a]$ for $a \in A_z \cap \mathcal{J}$) to the values at continuous addresses in the output trace ($z'[a]$ for $a \in A_{z'} \cap \mathcal{J}$).
-We compute the Jacobian using automatic differentiation (AD).
+We compute the Jacobian using automatic differentiation (AD);
+different approaches are possible.
 The Gen implementation uses forward-mode AD whereas our PyTorch implementation uses reverse-mode AD.
 The procedure \textproc{run-involution} in Algorithm~\ref{alg:auto-involutive-mcmc} shows the implementation of the interpreter for the language that uses reverse-mode AD.
-The interpreter executes $\mathcal{F}$ using regular Julia or Python semantics, but intercepts calls to \texttt{write}, \texttt{read}, and \texttt{copy} statements, and in addition to performing the desired operation, records the set of continuous addresses that are read, written, and copied.
-After $\mathcal{F}$ is finished executing, AD is used to compute the Jacobian $|J h_e|(z)$; in the case of reverse-mode AD by iterating over output continuous addresses and backpropagating from each one to all of the input continous addresses, computing the Jacobian row-by-row.
-The address metadata that was recorded during the first pass through the function is used to make AD more efficient, as discussed in the next section.
+The interpreter executes $\mathcal{F}$ using regular Julia or Python semantics, but intercepts calls to \texttt{@write}, \texttt{@read}, and \texttt{@copy} statements, and in addition to performing the desired operation, records the set of continuous addresses that are read, written, and copied.
+After $\mathcal{F}$ is finished executing, AD uses the recorded addresses to compute the Jacobian $|J h_e|(z)$; in the case of reverse-mode AD by iterating over output continuous addresses and backpropagating from each one to all of the input continous addresses, computing the Jacobian row-by-row.
+
+%\subsubsection{Example: Involution for a Split-Merge Reversible Jump Kernel}
+%
+%Figure~\ref{fig:mixture}(d) shows an involution expressed in the differentiable programming language introduced in this section, and Figure~\ref{fig:mixture}(f) shows a schematic of the information flow through the involution as it is applied twice in a row.
+%The involution implements a split or a merge proposal in the infinite mixture model of Figure~\ref{fig:mixture}(b), depending on the trace of the auxiliary program in Figure~\ref{fig:mixture}(c).
+%If there are at least two clusters, and the value of address $\mathtt{split}$ is false, then the involution proposes to merge two clusters by running code block 2 shown in green.
+%Otherwise, the involution proposes to split a cluster by running code block 1 shown in yellow.
+%Note that that code block 1 (yellow) is itself an involution on the discrete addresses in the traces---each application of code block 1 to the traces causes them to flip back and forth between a split and a merge proposal.
+%Also note that code blocks 2 and 3 are continuously differentiable bijections on the continuous addresses in the traces, and they are inverses of one another.
 

### 2020-07-17 paper.tex
remove split merge example from sec 4.2
tweak sec 4.2
remove ref to fig 1 from sec 4.1

@@
 
-As a concrete example, consider the probabilistic program in Figure~\ref{fig:mixture}b, which defines a Gaussian mixture model with an unknown number of components:
+For example, consider the probabilistic program below, which defines a Gaussian mixture model with an unknown number of components:
 \noindent
@@
 a distinct distribution over dictionaries.
-
 The first line of the program samples a number of mixture components from a Poisson prior using the address \texttt{:k}.
@@
 After $\mathcal{F}$ is finished executing, AD uses the recorded addresses to compute the Jacobian $|J h_e|(z)$; in the case of reverse-mode AD by iterating over output continuous addresses and backpropagating from each one to all of the input continous addresses, computing the Jacobian row-by-row.
-
-%\subsubsection{Example: Involution for a Split-Merge Reversible Jump Kernel}
-%
-%Figure~\ref{fig:mixture}(d) shows an involution expressed in the differentiable programming language introduced in this section, and Figure~\ref{fig:mixture}(f) shows a schematic of the information flow through the involution as it is applied twice in a row.
-%The involution implements a split or a merge proposal in the infinite mixture model of Figure~\ref{fig:mixture}(b), depending on the trace of the auxiliary program in Figure~\ref{fig:mixture}(c).
-%If there are at least two clusters, and the value of address $\mathtt{split}$ is false, then the involution proposes to merge two clusters by running code block 2 shown in green.
-%Otherwise, the involution proposes to split a cluster by running code block 1 shown in yellow.
-%Note that that code block 1 (yellow) is itself an involution on the discrete addresses in the traces---each application of code block 1 to the traces causes them to flip back and forth between a split and a merge proposal.
-%Also note that code blocks 2 and 3 are continuously differentiable bijections on the continuous addresses in the traces, and they are inverses of one another.
+Note that the $\mathtt{discrete}$ and $\mathtt{continuous}$ labels are omitted from the syntax in \textproc{run-involution} to simplify notation ($a \in \mathcal{I}$ indicates a discrete choice and otherwise a choice is continuous).
 

### 2020-07-17 paper.tex

--- 2020-07-17 paper.tex
@@
 
-In this section, we introduce a probabilistic programming language from the Gen probabilistic programming system~\citep{cusumano2019gen}, and present a technique for automating the implementation of involutive MCMC algorithms when the model $p$ and auxiliary distribution $q_x$ are both represented as
-probabilistic programs in this language.
-But the technique is not limited to the Gen system---we also provide the implementation of a minimal probabilistic programming language in PyTorch that supports the automation technique.
-
-\subsubsection{A Probabilistic Programming Language}
-
-Our probabilistic programming language augments the syntax of Julia~\citep{bezanson2017julia}
-with a new construct, the \texttt{\{a\} $\sim$ d} expression, for making a \textit{named} random choice.
-An execution trace of a Gen probabilistic program can be sampled by running the program according to Julia's usual semantics, and upon encountering an expression of the form \texttt{\{a\} $\sim$ d}, (i) evaluating the \textit{address expression} \texttt{a} to obtain an address $a \in \mathcal{A}$; (ii) evaluating the \textit{distribution expression} $d$ to obtain a distribution $d$ over the measurable space $V_a$; (iii) sampling a value $x_a \sim d$ and adding the mapping $\{a \mapsto x_a\}$ to a dictionary called the program trace; and (iv) returning the sampled value $x_a$ to the program, to continue execution.
+In this section, we introduce the Gen language~\citep{gen-pldi} for probabilistic programs, and present a technique for automating the implementation of involutive MCMC algorithms when the model $p$ and auxiliary distribution $q_x$ are both represented as
+Gen programs.
+
+\subsubsection{A Language for Probabilistic Programs with Traces}
+
+The Gen probabilistic programming language augments the syntax of Julia~\citep{julia}
+with a single new construct, the \texttt{\{a\} $\sim$ d} expression, for making a \textit{named} random choice.
+An execution trace of a Gen probabilistic program can be sampled by running the program according to Julia's usual semantics, and upon encountering an expression of the form \texttt{\{a\} $\sim$ d}, (i) evaluating the \textit{address expression} \texttt{a} to obtain an address $k \in \mathcal{K}$; (ii) evaluating the \textit{distribution expression} $d$ to obtain a distribution $d$ over the measurable space $V_k$; (iii) sampling a value $x_k \sim d$ and adding the mapping $\{k \mapsto x_k\}$ to a dictionary called the program trace; and (iv) returning the sampled value $x_k$ to the program, to continue execution.
 When execution terminates, the program trace has accumulated a mapping for each random choice encountered during the program's execution.
 
-As a concrete example, consider the probabilistic program in Figure~\ref{fig:mixture}b, which defines a Gaussian mixture model with an unknown number of components:
-\noindent
-\begin{center}
-\begin{tabular}{c}
-{\begin{lstlisting}[basicstyle=\small\ttfamily]
-@gen function p(n::Int)
-  k ~ poisson_plus_one(1)
-  means = [
-    ({(:mu, j)} ~ normal(0, 10)) for j in 1:k]
-  vars = [
-    ({(:var, j)} ~ inv_gamma(1, 10)) for j in 1:k]
-  weights ~ dirichlet([2.0 for j in 1:k])
-  for i in 1:n
-    {(:x, i)} ~ mixture_of_normals(
-        weights, means, vars)
-  end
-end
-\end{lstlisting}}
-\end{tabular}
-\end{center}
-The program \texttt{p} accepts as input an integer \texttt{n}, a number of data points. Each $\texttt{n}$ defines
+As a concrete example, consider Figure~\ref{fig:mixture}a, which defines a Gen program for a Gaussian mixture model with an unknown number of components.
+The program \texttt{p} accepts as input an integer \texttt{n}, a number of data points.
+Each $\texttt{n}$ defines
 a distinct distribution over dictionaries.
@@
 
-If the two probabilistic programs $\mathcal{P}$ and $\mathcal{Q}$ satisfy a simple technical requirement (described in Section~\ref{sec:technical-requirement}), we can interpret them as specifying the model density $p$ and auxiliary densities $q_x$ for use with involutive MCMC on dictionaries (Algorithm~\ref{alg:involutive-mcmc}).
+Given two probabilistic programs $\mathcal{P}$ and $\mathcal{Q}$ satisfying a simple technical requirement (described in Section~\ref{sec:technical-requirement}), we can interpret them as specifying the model density $p$ and auxiliary densities $q_x$ for use with involutive MCMC on dictionaries (Algorithm~\ref{alg:involutive-mcmc}).
 (In this case, the program $\mathcal{Q}$
 accepts a trace $x$ of $\mathcal{P}$ as input, and the distribution for $q_x$ is given by $\mathcal{Q}_x$.)
+The acceptance probability of involutive MCMC includes the following ratio of densities:
+\begin{equation}
+\frac{p(x') q_{x'}(y')}{p(x) q_{x}(y)}
+\end{equation}
+
+This ratio can be computed term-wise using \textsc{score} on the probabilistic programs $\mathcal{P}$ and $\mathcal{Q}$.
 
@@
 Then $\tilde{p}(x \oplus b) = \mathcal{L}(b)p(x)$.
-If we use $\tilde{p}(x \oplus b)$ in place of $p(x)$ to compute the ratio of densities in Algorithm~\ref{alg:involutive-mcmc}, we will wind up with the same output, because $\mathcal{L}(b)$ will cancel in the numerator and denominator:
-\begin{equation}
-\frac{p(x') q_{x'}(y')}{p(x) q_{x}(y)}
-= \frac{\tilde{p}(x' \oplus b) q_{x'}(y')}{\tilde{p}(x \oplus b) q_x(y)}
-= \frac{\tilde{p}(x' | b) q_{x'}(y')}{\tilde{p}(x | b) q_x(y)}
-\end{equation}
-
-Thus, this ratio can be computed term-by-term using \textsc{score} on the probabilistic programs $\mathcal{P}$ and $\mathcal{Q}$: (Algorithm~\ref{alg:auto-involutive-mcmc}): to compute $q_x(y')$, we run the algorithm directly on the program $\mathcal{Q}_x$, and to compute $p(x)$ and $p(x')$, we actually merge the dictionary $x$ with the observations $b$ and compute $\mathcal{L}(b)p(x) = \tilde{p}(x \oplus b)$ instead, by running the algorithm on $\mathcal{P}$ with trace $x \oplus b$.
+If we use $\tilde{p}(x \oplus b)$ in place of $p(x)$ to compute the ratio above, we will wind up with the same output, because $\mathcal{L}(b)$ will cancel in the numerator and denominator.
+
+Thus, this ratio can be computed term-by-term using \textsc{score} (Algorithm~\ref{alg:auto-involutive-mcmc}): to compute $q_x(y')$, we run the algorithm directly on the program $\mathcal{Q}_x$, and to compute $p(x)$ and $p(x')$, we actually merge the dictionary $x$ with the observations $b$ and compute $\mathcal{L}(b)p(x) = \tilde{p}(x \oplus b)$ instead, by running the algorithm on $\mathcal{P}$ with trace $x \oplus b$.
 The density $q_x(y)$ can be computed while $y$ is being sampled, using \textsc{trace-and-score}.
@@
 Section~\ref{sec:sparsity} describes a more efficient approach that exploits sparsity in the involution and resulting cancellation in the ratio for improved efficiency, but requires a more sophisticated probabilistic programming runtime system.
-
-Figure~\ref{fig:mixture}b and Figure~\ref{fig:mixture}c show examples of probabilistic programs $\mathcal{P}$ and $\mathcal{Q}$ respectively, for a split-merge reversible jump move.
 
@@
 
-\section{EXPLOITING SPARSITY FOR IMPROVED PERFORAMNCE} \label{sec:sparsity}
+\section{EXPLOITING SPARSITY FOR IMPROVED PERFORMANCE} \label{sec:sparsity}
 

### 2020-07-17 paper.tex

@@
     \execute[$\mathcal{F}$], but with
-        \State $\mathtt{"write}(k, v)\mathtt{"} \equiv$
+        \State $\mathtt{"@write}(k, v)\mathtt{"} \equiv$
             (\algorithmicif\hspace{1mm} $k \in \mathcal{I}$
@@
             \algorithmicelse\hspace{1mm} set $z'[k] \gets v$ and set $W \gets W \cup \{k\})$
-        \State $\mathtt{"read}(k)\mathtt{"} \equiv$
+        \State $\mathtt{"@read}(k)\mathtt{"} \equiv$
             (\algorithmicif\hspace{1mm} $k \in \mathcal{I}$
@@
 
-In this section, we introduce the Gen language~\citep{gen-pldi} for probabilistic programs, and present a technique for automating the implementation of involutive MCMC algorithms when the model $p$ and auxiliary distribution $q_x$ are both represented as
-Gen programs.
-
-\subsubsection{A Language for Probabilistic Programs with Traces}
-
-The Gen probabilistic programming language augments the syntax of Julia~\citep{julia}
+In this section, we introduce a probabilistic programming language from the Gen probabilistic programming system~\citep{cusumano2019gen}, and present a technique for automating the implementation of involutive MCMC algorithms when the model $p$ and auxiliary distribution $q_x$ are both represented as
+probabilistic programs in this language.
+But the technique is not limited to the Gen system---we also provide the implementation of a minimal probabilistic programming language in PyTorch that supports the automation technique.
+
+\subsubsection{A Probabilistic Programming Language}
+
+Our probabilistic programming language augments the syntax of Julia~\citep{bezanson2017julia}
 with a single new construct, the \texttt{\{a\} $\sim$ d} expression, for making a \textit{named} random choice.
@@
 
-As a concrete example, consider Figure~\ref{fig:mixture}a, which defines a Gen program for a Gaussian mixture model with an unknown number of components.
-The program \texttt{p} accepts as input an integer \texttt{n}, a number of data points.
-Each $\texttt{n}$ defines
+For example, consider the probabilistic program below, which defines a Gaussian mixture model with an unknown number of components:
+\noindent
+\begin{center}
+\begin{tabular}{c}
+{\begin{lstlisting}[basicstyle=\small\ttfamily]
+@gen function p(n::Int)
+  k ~ poisson_plus_one(1)
+  means = [
+    ({(:mu, j)} ~ normal(0, 10)) for j in 1:k]
+  vars = [
+    ({(:var, j)} ~ inv_gamma(1, 10)) for j in 1:k]
+  weights ~ dirichlet([2.0 for j in 1:k])
+  for i in 1:n
+    {(:x, i)} ~ mixture_of_normals(
+        weights, means, vars)
+  end
+end
+\end{lstlisting}}
+\end{tabular}
+\end{center}
+The program \texttt{p} accepts as input an integer \texttt{n}, a number of data points. Each $\texttt{n}$ defines
 a distinct distribution over dictionaries.
-
 The first line of the program samples a number of mixture components from a Poisson prior using the address \texttt{:k}.
@@
 
+Figure~\ref{fig:mixture}b and Figure~\ref{fig:mixture}c show examples of probabilistic programs $\mathcal{P}$ and $\mathcal{Q}$ respectively, for a split-merge reversible jump move.
+
 \subsection{Differentiable Programming with Traces} \label{sec:differentiable-programming-traces}
 
-The previous section described how the probability densities ($p$ and $q$) required for involutive MCMC on spaces of dictionaries can be specified and automatically computed using probabilistic programs.
-This section introduces a lightweight differentiable programming language for specifying involutions ($f : Z \to Z$) on dictionaries.
-The interpreter for the language computes the Radon-Nikodym derivative required for the acceptance probability using automatic differentiation.
-The language is implemented as part of the Gen system~\citep{cusumano2019gen}, which is embedded in the Julia language.
-We also provide a simplified PyTorch-based implementation of the language.
-
+Section~\ref{sec:probabilistic-programming} showed that if the densities $p$ and $q$ are specified using probabilistic programs $\mathcal{P}$ and $\mathcal{Q}$, then density ratio in the acceptance probability for involutive MCMC on dictionaries Equation~(\ref{eq:acceptance-ratio-dictionaries}) can be automated using probabilistic programming techniques.
+This section shows that if the involution $f$ is specified in using a differentiable program $\mathcal{F}$ that transforms the execution traces of probabilistic programs, then the Jacobian factor in Equation~(\ref{eq:acceptance-ratio-dictionaries}) can be automated using automatic differentiation.
+The procedure \textproc{auto-involutive-mcmc} in Algorithm~\ref{alg:auto-involutive-mcmc} combines these two ideas and automates involutive MCMC given the programs $\mathcal{P}$, $\mathcal{Q}$, and $\mathcal{F}$.
+
+%The previous section described how the probability densities ($p$ and $q$) required for involutive MCMC on spaces of dictionaries can be specified and automatically computed using probabilistic programs.
+%This section introduces a lightweight differentiable programming language for specifying involutions ($f : Z \to Z$) on dictionaries.
+%The interpreter for the language computes the Radon-Nikodym derivative required for the acceptance probability using automatic differentiation.
+%The language is implemented as part of the Gen system~\citep{cusumano2019gen}, which is embedded in the Julia language.
+%We also provide a simplified PyTorch-based implementation of the language.
+
+\subsubsection{A Differentiable Programming Language for Manipulating Traces}
 Recall that for involutive MCMC on a state space of dictionaries, we define $Z := \{x \oplus y : p(x) q_x(y) > 0\} \subseteq \mathcal{D}$, where $x \oplus y$ is the dictionary resulting from merging dictionaries $x$ and $y$ with disjoint keys.
-Intuitively, $z = x \oplus y$ the trace of a composite probabilistic program constructed by running $\mathcal{P}$ first, and then running $\mathcal{Q}$, passing the trace of $\mathcal{P}$ as its input.
-
-\subsubsection{A Differentiable Programming Language for Manipulating Traces}
-We denote a program in the differentiable language by $\mathcal{F}$ and the involution it defines as $f$.
-Like the probabilistic programming language described in the previous section, our differentiable language for involutions on traces is embedded based on Julia.
-The differentiable language has specialized expressions for reading values from addresses in a trace, writing values to the addresses in a trace, and copying values from an address in one trace to an address in another trace.
-For example, to read a value from a trace, we use the \texttt{@read} keyword:
+The involution is a function $f : Z \to Z$.
+%Intuitively, $z = x \oplus y$ the trace of a composite probabilistic program constructed by running $\mathcal{P}$ first, and then running $\mathcal{Q}$, passing the trace of $\mathcal{P}$ as its input.
+
+We now introduce a simple differentiable programming language for specifying involutions $f$.
+The language needs to have syntax for reading the value from an address in $x \oplus y \in Z$ and writing to an address in $x' \oplus y' \in Z$.
+To read a value $(x \oplus y)[a]$ at address $a$, we use the \texttt{@read} keyword:
 \noindent
@@
 end
-\end{lstlisting}}
-\end{tabular}
-\end{center}
-
-\subsubsection{Example: Split-Merge Reversible Jump}
-
-Figure~\ref{fig:mixture}(d) shows an involution expressed in the differentiable programming language introduced in this section, and Figure~\ref{fig:mixture}(f) shows a schematic of the information flow through the involution as it is applied twice in a row.
-The involution implements a split or a merge proposal in the infinite mixture model of Figure~\ref{fig:mixture}(b), depending on the trace of the auxiliary program in Figure~\ref{fig:mixture}(c).
-If there are at least two clusters, and the value of address $\mathtt{split}$ is false, then the involution proposes to merge two clusters by running code block 2 shown in green.
-Otherwise, the involution proposes to split a cluster by running code block 1 shown in yellow.
-Note that that code block 1 (yellow) is itself an involution on the discrete addresses in the traces---each application of code block 1 to the traces causes them to flip back and forth between a split and a merge proposal.
-Also note that code blocks 2 and 3 are continuously differentiable bijections on the continuous addresses in the traces, and they are inverses of one another.
+
+@gen function q(x)
+end
+\end{lstlisting}}
+\end{tabular}
+\end{center}
 
 \subsubsection{Computing the Jacobian with Automatic Differentiation}
-Recall that the involution $f$ is composed of (i) an involution $g$ on elements $e$ of a partition of the state space, and (ii) a pair of continuous differentiable bijections $h_{e}$ and $h_{g(e)} = h_{e}^{-1}$ between each pair of corresponding elements of the partition.
+
+Recall that the involution $f$ must decompose into (i) an involution $g$ on elements $e \in E$ of a partition of the state space, and (ii) a pair of continuous differentiable bijections $h_{e}$ and $h_{g(e)} = h_{e}^{-1}$ between each pair of corresponding elements of the partition.
 Each function $h_e$ is a function from the values at continuous addresses in the input trace ($z[a]$ for $a \in A_z \cap \mathcal{J}$) to the values at continuous addresses in the output trace ($z'[a]$ for $a \in A_{z'} \cap \mathcal{J}$).
-We compute the Jacobian using automatic differentiation (AD).
+We compute the Jacobian using automatic differentiation (AD);
+different approaches are possible.
 The Gen implementation uses forward-mode AD whereas our PyTorch implementation uses reverse-mode AD.
 The procedure \textproc{run-involution} in Algorithm~\ref{alg:auto-involutive-mcmc} shows the implementation of the interpreter for the language that uses reverse-mode AD.
-The interpreter executes $\mathcal{F}$ using regular Julia or Python semantics, but intercepts calls to \texttt{write}, \texttt{read}, and \texttt{copy} statements, and in addition to performing the desired operation, records the set of continuous addresses that are read, written, and copied.
-After $\mathcal{F}$ is finished executing, AD is used to compute the Jacobian $|J h_e|(z)$; in the case of reverse-mode AD by iterating over output continuous addresses and backpropagating from each one to all of the input continous addresses, computing the Jacobian row-by-row.
-The address metadata that was recorded during the first pass through the function is used to make AD more efficient, as discussed in the next section.
+The interpreter executes $\mathcal{F}$ using regular Julia or Python semantics, but intercepts calls to \texttt{@write}, \texttt{@read}, and \texttt{@copy} statements, and in addition to performing the desired operation, records the set of continuous addresses that are read, written, and copied.
+After $\mathcal{F}$ is finished executing, AD uses the recorded addresses to compute the Jacobian $|J h_e|(z)$; in the case of reverse-mode AD by iterating over output continuous addresses and backpropagating from each one to all of the input continous addresses, computing the Jacobian row-by-row.
+Note that the $\mathtt{discrete}$ and $\mathtt{continuous}$ labels are omitted from the syntax in \textproc{run-involution} to simplify notation ($a \in \mathcal{I}$ indicates a discrete choice and otherwise a choice is continuous).
 

### 2020-07-17 paper.tex
finished 5.1

@@
 
-\subsubsection{Incremental Computation of Model Density Ratio}
+Suppose $N$ and $N'$ are the total number of random choices in the input traces $x \oplus y$ and output traces $x' \oplus y'$ respectively, and $M$ is the number of continuous random choices (which is the same in the input traces and output traces).
+Then, the number of operations requried in Algorithm~\ref{alg:auto-involutive-mcmc} grows as $O(N) + O(M^2)$ (where each call to \textproc{logpdf} and the computation of each entry in the Jacobian both count as one operation).
+The linear term represents the complexity of sampling $y'$ and computing the four log-densities used to compute the acceptance probability.
+The quadratic term represents the complexity of computing the Jacobian matrix and its determinant.
+It is possible to reduce the computational complexity of an automated involutive MCMC kernel by exploiting special structure in the involution $f$, in some cases to $O(1)$.
+This section describes how an automated involutive MCMC implementation can reduce the computational complexity of the Jacobian computation and the density ratio computation.
+These techniques are used in the Gen implementation of automated involutive MCMC, and one of the techniques is used in our minimal PyTorch implementation.
+
+\subsection{Sparsity-Aware Automatic Jacobian Computation}
+
+The naive implementation of Algorithm~\ref{alg:auto-involutive-mcmc} computes by the Jacobian by first computing the $M$-by-$M$ Jacobian matrix $J h_{e(z)}$ via automatic differentiation, and them computing the absolute value of its determinant.
+However, we observe that in many applications of involutive MCMC, the values at continuous choices in the input traces are directly copied into the output traces (either at the same, or different, keys).
+These copy operations result in columns in the Jacobian matrix that have a single $1$ entry with remaining entries are $0$.
+For example, for the function $(u, v, x, y) \mapsto (u, 2 u - v, y, x) = (u', v', x', y')$, the Jacobian is (with columns corresponding to $u'$, $v'$, $x'$, and $y'$ and rows corresponding to $u$, $v$, $x$, and $y$):
+\begin{equation}
+\left[
+\begin{array}{cccc}
+1 & 2 & 0 & 0\\
+0 & -1 & 0 & 0\\
+0 & 0 & 0 & 1\\
+0 & 0 & 1 & 0
+\end{array}
+\right]
+\begin{array}{c}
+u\\ v\\ x\\ y
+\end{array}
+\end{equation}
+Using the cofactor expansion of the determinant, we observe that for any (`copy') row or column in an $M$-by-$M$ Jacobian matrix with a single $1$ and all other entries $0$, that absolute value of the determinant is equivalent to that of the $(M-1)$-by-$(M-1)$ sub-matrix with the corresponding row and column omitted (even if that would remove other nonzero entries from the matrix).
+By applying this rule recursively, we can instead compute the determinant of a much smaller matrix; for the example with $M=4$, the absolute value of the determinant simplifies to the absolute value of a single entry ($|-1|$).
+Indeed, if those output variables that are the result of copying an input variable, then we can avoid computing those entries completely, and the computational complexity reduces from $M^2$ to $K^2$ where $K$ is the number of output variables that were copied from some input variable.
+The $\mathtt{@copy}$ statement in our differentiable programming language makes the set of output variables that were copied immediately available to the interpreter.
+Many involutive MCMC kernels update only a constant number of variables that does not depend on the size $N$ of the trace or the total number of continous keys $M$.
+For example, consider the Jacobian for a split-merge reversible jump move~\citet{richardson1997bayesian}, which is implemented in Figure~\ref{fig:mixture}:
+\includegraphics[width=\linewidth]{figures/sparse-jacobian.pdf}
+In this case, the size $N$ of the trace grows linearly in the number of clusters, and the number of data points, but the Jacobian determinant can be calculated from only the $6$-by-$6$ submatrix of the Jacobian that involves the parameters of the cluster being split (or the clusters being merged).
+While a reasonable hand-coded implementation of this algorithm would likely exploit this sparsity as well, our differentiable programming languages for traces automates it.
+
+% TODO show an performance table of moves per second with and without the sparsity reduction (?)
+\subsection{Incremental Computation of Model Density Ratio and Implicit Copies}
 While it is possible to compute the ratio of densities in Equation~(\ref{eq:density-ratio}) by separately computing the four densities of the four traces, this scales linearly in the total number of random choices in the traces, because every random choice in each trace needs to be visited to compute the products in Equation~(\ref{eq:density-product}).
@@
 
-\subsubsection{Implicit Copies}
-
 If an address is present in the input model trace and is needed to form a complete output model trace but was not written to or copied to in by the function body, then it will be implicitly copied from the previous trace by the runtime implementation.
@@
 
-\subsection{Sparsity-Aware Automatic Jacobian Computation}
-
-It is possible to automatically compute the Jacobian determinant in Equation~(\ref{eq:accept-prob-jacobian}) using automatic differentiation of the involution $f$.
-Note that while the involution can manipulate both discrete and continuous random choices, the Jacobian is the matrix of partial derivatives associated with the continuous random choices only.
-The differentiable programming language implementation distinguishes between continuous and discrete choices with the help of the user-provided type tags.
-The sums of the number of continuous dimensions across all addresses in the input and the output traces respectively must match, because otherwise $f$ cannot factor into continuously differentiable bijections $f_{2,i}$, which is necessary for the involutive MCMC acceptance ratio to simplify to Equation~(\ref{eq:accept-prob-jacobian}) (see Section~\ref{sec:radon-nikodym-special-case}).
-
-The Jacobian is computed automatically by applying forward-mode automatic differentiation separately for each continuous random choice in the input traces, which scales as $O(N^2)$ where $N$ is the number of such choices.
-Common determinant implementations based on the LU-decomposition scale on the order $O(N^2)$ for an $N$-by-$N$ matrix.
-Therefore, the basic algorithm computes $|J f_{2,i(x,y)} |$ in $O(N^2)$ where $N$ is the number of continuous addresses in $x$ or $y$.
-
-To derive a more efficient algorithm, we use the cofactor expansion of the determinant, and observe that any row or column in the Jacobian matrix that contains a single $1$ and all other ($N-1$) entries $0$ implies that the absolute value of the determinant is equivalent to the absolute value of the $N-1$-by-$N-1$ sub-matrix with the corresponding row and column omitted (even if that would remove other nonzero entries from the matrix).
-Applying this rule recursively allows a significant reduction in the size of the determinant computation when the Jacobian matrix possesses sparsity.
-Sparse Jacobians are common in applications of involutive MCMC, and reversible jump MCMC in particular.
-For example, consider the Jacobian for the split-merge move of~\citet{richardson1997bayesian}, implemented in Figure~\ref{fig:mixture}:
-\includegraphics[width=\linewidth]{figures/sparse-jacobian.pdf}
-In this case, the Jacobian determinant can be simplified to the determinant of a $6$-by-$6$ matrix that does not grow in size with the number of clusters and number of data points, because unaffected clusters and data points have their values \emph{copied} from the input model trace to the output model trace.
-While a reasonable hand-coded implementation of this algorithm would likely exploit this sparsity as well, it would require careful and error-prone reasoning and implementation, which our approach automates.
-
-% TODO show an performance table of moves per second with and without the sparsity reduction (?)
+\subsection{Incremental Computation of Model Density Ratio and Implicit Copies}
+While it is possible to compute the ratio of densities in Equation~(\ref{eq:density-ratio}) by separately computing the four densities of the four traces, this scales linearly in the total number of random choices in the traces, because every random choice in each trace needs to be visited to compute the products in Equation~(\ref{eq:density-product}).
+Often, the number of random choices in the auxiliary traces $y$ and $y'$ are constant size $O(1)$ whereas the number of random choices in the model traces $x$ and $x'$ is $O(N)$ where $N$ is the number of data points, or latent components
+(see Figure~\ref{fig:mixture} for an example, where at most two clusters are affected by each application of the MCMC kernel).
+In these cases, the ratio of model trace densities $p(x') / p(x)$ dominates the asymptotic running time of computing the density ratio of Equation~\ref{eq:density-ratio}.
+However, due to conditional independencies in the model, the densities $p(x')$ and $p(x)$ often share common factors that cancel, which affords an opportunity for $O(1)$ computation of the ratio.
+Gen's trace API includes an \emph{incremental update} operation~\citep{gen-paper}, which takes a previous trace $x$, and a partial trace $u$ containing the values of random choices that should be changed (or added if there is stochastic structure in the model), and returns a new trace $x'$.
+This allows for $O(1)$ computation of $p(x') / p(x)$ when the modeling language runtime system is able to detect and exploit the cancellation.
+
+If an address is present in the input model trace and is needed to form a complete output model trace but was not written to or copied to in by the function body, then it will be implicitly copied from the previous trace by the runtime implementation.
+This significantly reduces the amount of code that has to be written when only a small subset of the model trace is being modified during the application of the involution MCMC move (which is common in applications of reversible jump MCMC).
+As a simple example, if there are $N$ latent variables, and the move only updates 1 of them, then the involution needs to contain only $O(1)$ lines of code instead of $O(N)$ lines.
+
+TODO: show an example
 

### 2020-07-17 paper.tex
fixes to 5.1

@@
 Using the cofactor expansion of the determinant, we observe that for any (`copy') row or column in an $M$-by-$M$ Jacobian matrix with a single $1$ and all other entries $0$, that absolute value of the determinant is equivalent to that of the $(M-1)$-by-$(M-1)$ sub-matrix with the corresponding row and column omitted (even if that would remove other nonzero entries from the matrix).
-By applying this rule recursively, we can instead compute the determinant of a much smaller matrix; for the example with $M=4$, the absolute value of the determinant simplifies to the absolute value of a single entry ($|-1|$).
-Indeed, if those output variables that are the result of copying an input variable, then we can avoid computing those entries completely, and the computational complexity reduces from $M^2$ to $K^2$ where $K$ is the number of output variables that were copied from some input variable.
-The $\mathtt{@copy}$ statement in our differentiable programming language makes the set of output variables that were copied immediately available to the interpreter.
-Many involutive MCMC kernels update only a constant number of variables that does not depend on the size $N$ of the trace or the total number of continous keys $M$.
+By applying this rule recursively, we can instead compute the determinant of a much smaller matrix; for the example above with $M=4$, the absolute value of the determinant simplifies to the absolute value of a single entry ($|-1|$).
+Indeed, if some input variable is copied to some output variable, then we can entirely avoid computing its row (and corresponding column) of the Jacobian, and the computational complexity reduces from $M^2$ to $(M-K)^2$ where $K$ is the number of input variables that were copied from some input variable.
+The $\mathtt{@copy}$ statement in our differentiable programming language makes the set of input variables that were copied immediately available to the interpreter, which makes automating this optimization possible, as shown in \textproc{run-involution} in Algorithm~\ref{alg:auto-involutive-mcmc}.
+
+Many involutive MCMC kernels update only a constant number of variables that does not depend on the size $N$ of the trace or the total number of continous keys $M$, and the other variables are copied over unchanged.
 For example, consider the Jacobian for a split-merge reversible jump move~\citet{richardson1997bayesian}, which is implemented in Figure~\ref{fig:mixture}:
 \includegraphics[width=\linewidth]{figures/sparse-jacobian.pdf}
-In this case, the size $N$ of the trace grows linearly in the number of clusters, and the number of data points, but the Jacobian determinant can be calculated from only the $6$-by-$6$ submatrix of the Jacobian that involves the parameters of the cluster being split (or the clusters being merged).
-While a reasonable hand-coded implementation of this algorithm would likely exploit this sparsity as well, our differentiable programming languages for traces automates it.
+Black squares indicate nonzero entries.
+In this case, the size $N$ of the trace grows linearly both in the number of clusters and the number of data points, but the Jacobian determinant can be calculated from only the $6$-by-$6$ submatrix of the Jacobian that involves the parameters of the one cluster being split (or the two clusters being merged).
+While a reasonable hand-coded implementation of this algorithm would likely perform this optimization as well, Algorithm~\ref{alg:auto-involutive-mcmc} automates it.
 
@@
 \subsection{Incremental Computation of Model Density Ratio and Implicit Copies}
+
 While it is possible to compute the ratio of densities in Equation~(\ref{eq:density-ratio}) by separately computing the four densities of the four traces, this scales linearly in the total number of random choices in the traces, because every random choice in each trace needs to be visited to compute the products in Equation~(\ref{eq:density-product}).
@@
 
-\subsection{Incremental Computation of Model Density Ratio and Implicit Copies}
 While it is possible to compute the ratio of densities in Equation~(\ref{eq:density-ratio}) by separately computing the four densities of the four traces, this scales linearly in the total number of random choices in the traces, because every random choice in each trace needs to be visited to compute the products in Equation~(\ref{eq:density-product}).

### 2020-07-18 paper.tex
section 5.2

@@
             \algorithmicelse\hspace{1mm} set $R \gets R \cup \{k\}$ and evaluate to $z[k]$)
-        \State $\mathtt{"copy}(k_1, k_2)\mathtt{"} \equiv$ (set $z'[k_2] \gets z[k_1]$ and set $C \gets C \cup \{k_1\}$)
+        \State $\mathtt{"@copy}(k_1, k_2)\mathtt{"} \equiv$ (set $z'[k_2] \gets z[k_1]$ and set $C \gets C \cup \{k_1\}$)
     \END
@@
 Suppose $N$ and $N'$ are the total number of random choices in the input traces $x \oplus y$ and output traces $x' \oplus y'$ respectively, and $M$ is the number of continuous random choices (which is the same in the input traces and output traces).
-Then, the number of operations requried in Algorithm~\ref{alg:auto-involutive-mcmc} grows as $O(N) + O(M^2)$ (where each call to \textproc{logpdf} and the computation of each entry in the Jacobian both count as one operation).
+Then, the number of operations requried in Algorithm~\ref{alg:auto-involutive-mcmc} grows as $O(N + N') + O(M^2)$ (where each call to \textproc{logpdf} and the computation of each entry in the Jacobian both count as one operation).
 The linear term represents the complexity of sampling $y'$ and computing the four log-densities used to compute the acceptance probability.
@@
 
-Many involutive MCMC kernels update only a constant number of variables that does not depend on the size $N$ of the trace or the total number of continous keys $M$, and the other variables are copied over unchanged.
+Many involutive MCMC kernels update only a constant number of variables that does not depend on the sizes $N$ and $N'$ of the input and output traces or the total number of continous keys $M$; the other variables are copied over unchanged.
 For example, consider the Jacobian for a split-merge reversible jump move~\citet{richardson1997bayesian}, which is implemented in Figure~\ref{fig:mixture}:
@@
 In this case, the size $N$ of the trace grows linearly both in the number of clusters and the number of data points, but the Jacobian determinant can be calculated from only the $6$-by-$6$ submatrix of the Jacobian that involves the parameters of the one cluster being split (or the two clusters being merged).
-While a reasonable hand-coded implementation of this algorithm would likely perform this optimization as well, Algorithm~\ref{alg:auto-involutive-mcmc} automates it.
+Therefore, the computation complexity of the acceptance ratio computation reduces from $O(N + N') + O(M^2)$ to $O(N + N')$.
+While a reasonable hand-coded implementation of this algorithm would likely perform this sort of optimization as well, Algorithm~\ref{alg:auto-involutive-mcmc} automates it.
 
 % TODO show an performance table of moves per second with and without the sparsity reduction (?)
-\subsection{Incremental Computation of Model Density Ratio and Implicit Copies}
-
-While it is possible to compute the ratio of densities in Equation~(\ref{eq:density-ratio}) by separately computing the four densities of the four traces, this scales linearly in the total number of random choices in the traces, because every random choice in each trace needs to be visited to compute the products in Equation~(\ref{eq:density-product}).
-Often, the number of random choices in the auxiliary traces $y$ and $y'$ are constant size $O(1)$ whereas the number of random choices in the model traces $x$ and $x'$ is $O(N)$ where $N$ is the number of data points, or latent components
-(see Figure~\ref{fig:mixture} for an example, where at most two clusters are affected by each application of the MCMC kernel).
-In these cases, the ratio of model trace densities $p(x') / p(x)$ dominates the asymptotic running time of computing the density ratio of Equation~\ref{eq:density-ratio}.
-However, due to conditional independencies in the model, the densities $p(x')$ and $p(x)$ often share common factors that cancel, which affords an opportunity for $O(1)$ computation of the ratio.
-Gen's trace API includes an \emph{incremental update} operation~\citep{gen-paper}, which takes a previous trace $x$, and a partial trace $u$ containing the values of random choices that should be changed (or added if there is stochastic structure in the model), and returns a new trace $x'$.
-This allows for $O(1)$ computation of $p(x') / p(x)$ when the modeling language runtime system is able to detect and exploit the cancellation.
-
-If an address is present in the input model trace and is needed to form a complete output model trace but was not written to or copied to in by the function body, then it will be implicitly copied from the previous trace by the runtime implementation.
-This significantly reduces the amount of code that has to be written when only a small subset of the model trace is being modified during the application of the involution MCMC move (which is common in applications of reversible jump MCMC).
-As a simple example, if there are $N$ latent variables, and the move only updates 1 of them, then the involution needs to contain only $O(1)$ lines of code instead of $O(N)$ lines.
-
-TODO: show an example
-
-While it is possible to compute the ratio of densities in Equation~(\ref{eq:density-ratio}) by separately computing the four densities of the four traces, this scales linearly in the total number of random choices in the traces, because every random choice in each trace needs to be visited to compute the products in Equation~(\ref{eq:density-product}).
-Often, the number of random choices in the auxiliary traces $y$ and $y'$ are constant size $O(1)$ whereas the number of random choices in the model traces $x$ and $x'$ is $O(N)$ where $N$ is the number of data points, or latent components
-(see Figure~\ref{fig:mixture} for an example, where at most two clusters are affected by each application of the MCMC kernel).
-In these cases, the ratio of model trace densities $p(x') / p(x)$ dominates the asymptotic running time of computing the density ratio of Equation~\ref{eq:density-ratio}.
-However, due to conditional independencies in the model, the densities $p(x')$ and $p(x)$ often share common factors that cancel, which affords an opportunity for $O(1)$ computation of the ratio.
-Gen's trace API includes an \emph{incremental update} operation~\citep{gen-paper}, which takes a previous trace $x$, and a partial trace $u$ containing the values of random choices that should be changed (or added if there is stochastic structure in the model), and returns a new trace $x'$.
-This allows for $O(1)$ computation of $p(x') / p(x)$ when the modeling language runtime system is able to detect and exploit the cancellation.
-
-If an address is present in the input model trace and is needed to form a complete output model trace but was not written to or copied to in by the function body, then it will be implicitly copied from the previous trace by the runtime implementation.
-This significantly reduces the amount of code that has to be written when only a small subset of the model trace is being modified during the application of the involution MCMC move (which is common in applications of reversible jump MCMC).
-As a simple example, if there are $N$ latent variables, and the move only updates 1 of them, then the involution needs to contain only $O(1)$ lines of code instead of $O(N)$ lines.
-
-TODO: show an example
+\subsection{Incremental Computation of Output Traces and Density Ratios}
+
+It is possible to further reduce the complexity of applying an automated involutive MCMC kernel from $O(N + N')$ to $O(1)$ for many involutions, using additional optimizations that reduce the complexity of computing the density ratio and the complexity of computing the output of the involution to $O(1)$.
+
+Consider input traces $z$ and output traces $z'$, with sizes $N$ and $N'$ respectively, and an involution $f$ such that $z' = f(z)$.
+The approach in Algorithm~\ref{alg:auto-involutive-mcmc} involves running a program $\mathcal{F}$ that explicitly writes the value for each element of $f$, which requires $N'$ write operations.
+We now show how to specify the involution $f$ such that computing $z'$ from $z$ can take less than $N'$ operations for involutions that possess sparse structure.
+Recall that $K_{z}$ and $K_{z'}$ are the sets of keys in $z$ and $z'$, respectively, and that $z$ and $z'$ may both have values for certain keys, and $z$ may contains keys not present in $z'$, and vice-versa.
+Often, an involution does not modify the values of many keys in the trace---this is the case in the split-merge reversible jump move described earlier.
+Suppose that $A \subseteq K_{z} \cap K_{z'}$ is the set of keys $k$ in both $z$ and $z'$ for which $z'[k] = z[k]$.
+Let $\Delta := K_{z'} \setminus A$ be the set of keys in $z'$ that are either (i) not in $z$, or (ii) in $z$ but have their value changed.
+Then, the previous trace $z$ and a dictionary $\delta$ with $K_{\delta} = \Delta$ and $\delta[k] = z'[k]$ and the density function $p$ suffice to uniquely define the trace $z'$, provided the density satisfies the following condition:
+$p(x) > 0$ and $p(x') > 0$ and $x \ne x'$ implies there exists $k \in K_x \cap K_{x'}$ with $x[k] \ne x'[k]$.
+Let $x|_A$ denote the \emph{restriction} of a dictionary $x$ to only keys in $A$.
+Then, \textproc{naive-trace-update} (Algorithm~\ref{alg:naive-update}) computes the new trace $x'$ and difference in log-densities $\log p(x') - \log p(x)$ of a probabilistic program from a previous trace $x$ and a dictionary $\delta$, provided there exists $x' = \delta \oplus x|_A$ for some $A$ where $p(x') > 0$.
+\begin{algorithm}[h]
+\begin{algorithmic}
+\Require probabilistic program $\mathcal{P}$; previous trace $x$;\\
+dictionary $\delta$ such that there exists $x' = (\delta \oplus x|_{A})$ for some $A$ where $p(x') > 0$.
+\vspace{1mm}
+\Procedure{naive-trace-update}{$\mathcal{P}$, $x$, $\delta$}
+   \State $x' \gets \{\}$
+   \State $s' \gets 0$
+   \execute[$\mathcal{P}$], but with $\mathtt{"}k \sim \mathrm{distribution}\mathtt{"} \equiv$ (
+        \State 1. if $k \in K_{\delta}$ then $v \gets \delta[k]$ else $v \gets x[k]$
+        \State 2. set $x'[k] \gets v$
+        \State 3. set $s' \gets s' + \textproc{logpdf}(\mathrm{distribution}, v)$
+        \State 4. evaluate to $v$)
+    \END
+   \State $s \gets \textproc{score}(\mathcal{P}, x)$
+   \State \Return $(x', s' - s)$
+\EndProcedure
+\end{algorithmic}
+\caption{Naive Trace Update}
+\label{alg:naive-update}
+\end{algorithm}
+\textproc{naive-trace-update} is a naive implementation of a general \emph{trace-update} operation that is part of Gen's API.
+While the naive implementation requires $O(|K_x| + |K_{x'}|)$ operations, it is possible for more sophisticated implementations of \emph{trace-update} to run in $O(|K_{\delta}|)$ operations for certain probabilistic programs $\mathcal{P}$ by exploiting conditional independence structure program that leads to cancellations in the density ratio $p(x') / p(x)$.
+The details of these more efficient implementations are outside the scope of this paper, but are implemented as part of the Gen system.
+
+Given an implementation of \emph{trace-update} (that has the same input and output signature of \textproc{naive-trace-update}) we can optimize Algorithm~\ref{alg:auto-involutive-mcmc} by
+modifying the requirement for the involution program $\mathcal{F}$:
+Instead of explicitly specifying all of $z'$, it only needs to specify a subset of the trace $\delta$.
+Therefore, the involution program can run in $O(|K_{\delta}|)$ operations.
+We also remove the two calls to \textproc{score} on $x$ and $x'$, and replace them with a single call to \emph{trace-update} on $\mathcal{P}$ that takes in $x$, and $\delta$ and returns $x'$ and $\log p(x') - \log p(x)$.
+When $|K_{\delta}|$ and $|K_{y}|$ and $|K_{y}$ are all constant in the size $N$ of the model trace $x$, the resulting algorithm scales as $O(1)$ instead of $O(N + N')$.
 
@@
 \end{figure*}
+
+\section{DISCUSSION}
+
+- using probabilistic programming systems to develop new MCMC algorithms.
+
+- trace translators
+
+- work on compilers for inference programs that generate more specialized, high
+  performance implementations, from user specification of an inference
+algorithm.
 

### 2020-07-18 paper.tex
finish 5.2

@@
 
-It is possible to further reduce the complexity of applying an automated involutive MCMC kernel from $O(N + N')$ to $O(1)$ for many involutions, using additional optimizations that reduce the complexity of computing the density ratio and the complexity of computing the output of the involution to $O(1)$.
-
-Consider input traces $z$ and output traces $z'$, with sizes $N$ and $N'$ respectively, and an involution $f$ such that $z' = f(z)$.
-The approach in Algorithm~\ref{alg:auto-involutive-mcmc} involves running a program $\mathcal{F}$ that explicitly writes the value for each element of $f$, which requires $N'$ write operations.
-We now show how to specify the involution $f$ such that computing $z'$ from $z$ can take less than $N'$ operations for involutions that possess sparse structure.
-Recall that $K_{z}$ and $K_{z'}$ are the sets of keys in $z$ and $z'$, respectively, and that $z$ and $z'$ may both have values for certain keys, and $z$ may contains keys not present in $z'$, and vice-versa.
+It is possible to further reduce the the number of operations in automated involutive MCMC from $O(N + N')$ to $O(N_2 + N_2')$ where $N = N_1 + N_2$ and $N' = N_1' + N_2'$ and $N_1$ and $N_2$ are the sizes of the model trace and auxiliary trace respectively (and similarly for $N_1'$ and $N_2'$).
+In cases where $N_2$ and $N_2'$ have fixed size even as $N_1$ and $N_1'$ grow, this approach can result in $O(1)$ operations per kernel application.
+The efficiency gains are due to a combination of exploiting sparsity in the involution and conditional independencies in the model.
+
+Consider input traces $(x, y)$ and output traces $(x', y')$ where $x$ and $x'$ have sizes $N_1$ and $N_2$ respectively.
+The approach in Algorithm~\ref{alg:auto-involutive-mcmc} involves running a program $\mathcal{F}$ that explicitly writes the value for each element of $(x', y')$, which requires greater than $N_1'$ `write' or `copy' operations.
+We now show how to specify the involution $f$ such that computing $z'$ from $z$ can take fewer operations for involutions that possess sparse structure.
+Recall that $K_{x}$ and $K_{x'}$ are the sets of keys in $x$ and $x'$, respectively, and that $x$ and $x'$ may both have values for certain keys, and $x$ may contains keys not present in $x'$, and vice-versa.
 Often, an involution does not modify the values of many keys in the trace---this is the case in the split-merge reversible jump move described earlier.
-Suppose that $A \subseteq K_{z} \cap K_{z'}$ is the set of keys $k$ in both $z$ and $z'$ for which $z'[k] = z[k]$.
-Let $\Delta := K_{z'} \setminus A$ be the set of keys in $z'$ that are either (i) not in $z$, or (ii) in $z$ but have their value changed.
-Then, the previous trace $z$ and a dictionary $\delta$ with $K_{\delta} = \Delta$ and $\delta[k] = z'[k]$ and the density function $p$ suffice to uniquely define the trace $z'$, provided the density satisfies the following condition:
+Suppose that $A \subseteq K_{x} \cap K_{x'}$ is the set of keys $k$ in both $x$ and $x'$ for which $x'[k] = x[k]$.
+Let $\Delta := K_{x'} \setminus A$ be the set of keys in $x'$ that are either (i) not in $x$, or (ii) in $x$ but have their value changed.
+Then, the previous trace $x$ and a dictionary $\delta$ with $K_{\delta} = \Delta$ and $\delta[k] = x'[k]$ and the density function $p$ suffice to uniquely define the trace $x'$, provided the density satisfies the following condition:
 $p(x) > 0$ and $p(x') > 0$ and $x \ne x'$ implies there exists $k \in K_x \cap K_{x'}$ with $x[k] \ne x'[k]$.
@@
 Instead of explicitly specifying all of $z'$, it only needs to specify a subset of the trace $\delta$.
-Therefore, the involution program can run in $O(|K_{\delta}|)$ operations.
+Therefore, the involution program can run in $O(|K_{\delta}| + |K_{y}| + |K_{y'}|)$ operations.
 We also remove the two calls to \textproc{score} on $x$ and $x'$, and replace them with a single call to \emph{trace-update} on $\mathcal{P}$ that takes in $x$, and $\delta$ and returns $x'$ and $\log p(x') - \log p(x)$.
-When $|K_{\delta}|$ and $|K_{y}|$ and $|K_{y}$ are all constant in the size $N$ of the model trace $x$, the resulting algorithm scales as $O(1)$ instead of $O(N + N')$.
+When $|K_{\delta}|$ and $|K_{y}|$ and $|K_{y}|$ are all constant in the size $N_1$ of the model trace $x$, the resulting algorithm scales as $O(1)$ instead of $O(N + N')$.
 

### 2020-07-18 paper.tex
Pass through section 7

@@
 
-\section{A DYNAMIC CHECK FOR DETECTING BUGS}
+\section{DYNAMICALLY DETECTING BUGS IN INVOLUTIONS}
 
@@
 \begin{algorithmic}
-\Procedure{imcmc-dynamic-check}{$p$, $q$, $f$}
+\Procedure{involutive-mcmc-check}{$p$, $q$, $f$}
     \State $x \sim p(\cdot)$ \Comment{Sample state from prior}
@@
 
-To test the user's algorithm, Gen first samples a model trace $x$ from the model prior $p$, and auxiliary randomness $y$ from the kernel $q_x$. It then runs three tests:
+To test the user's algorithm, Gen first samples a model trace $x$ from the model prior $p$, and auxiliary trace $y$ from the $q_x$. It then runs three tests:
 
@@
 The move is accepted with probability:
-\begin{equation}
-{
+\begin{equation*}
+{\scriptstyle
 \min\left\{ 1,
@@
 \frac{q_{m,h' \to h}(u)}{q_{m,h \to h'}(u)}
+|J g_{m, h\to h'}|(\theta_h, u)
 \right\}
 }
-\end{equation}
+\end{equation*}
 
@@
 
-State-dependent mixture proposal.
-Depending on the current state of the model, a different mixture distribution is used for the proposal.
-Implementing such a move requires computing the mixture probability under the forward move and under the reverse move, which may differ.
-
-$q_x(m)$ -- the mixture distribution
-
-$q_m(x')$ -- the proposal we are mixing over
-
-The first part of the auxiliary probabilistic program $Q$ determines which proposal to apply by sampling from a mixture distribution.
-Here, the program picks a random node in the tree, by doing a stochastic walk of the tree that terminates at the chosen node.
+Figure~\ref{fig:structure-learning} shows automated involutive MCMC being applied to fully Bayesian inference over the covariance function of a Gaussian process, where the prior on covariance functions (Figure~\ref{fig:structure-learning}a) is based on a probabilistic context-free grammar.
+The inference algorithm is based on an involutive MCMC kernel that uses a state-dependent mixture of proposals.
+A variant of this inference algorithm was previously studied in~\citep{schaechtle2016time,saad2019bayesian} based on a model of~\citet{grosse2012exploiting}.
+
+At each iteration of the MCMC algorithm, the auxiliary probabilistic program $\mathcal{Q}$ (Figure~\ref{fig:structure-learning}d) first picks a random node in the parse tree of the covariance function, by doing a stochastic walk of the existing parse tree that terminates at the chosen node.
 \begin{center}
@@
 \end{center}
-The first part of the involution copies the random choices corresponding to the mixture from the input auxiliary trace to the output auxiliary trace.
+The first part of the involution (Figure~\ref{fig:structure-learning}e) copies the random choices made during this walk from the input auxiliary trace to the output auxiliary trace.
 \begin{center}
@@
 \end{center}
-
 The code that walks the tree uses the following recursion, which results in a probability distribution that assigns exponentially lower probability to nodes that are deeper in the tree.
@@
 \end{center}
-The resulting mixture distributions for two trees are:
+The resulting distributions on selected nodes for two example trees are shown below:
 \begin{center}
@@
 \end{center}
-Because the mixture is specified using a probabilistic program, it is straightforward to modify the program to define a different mixture distribution.
+The other parts of the auxiliary probabilistic program and the involution program specifies a proposal distribution for the subtree of the parse tree that is rooted at the chosen node.
+In particular, the rest of the auxiliary probabilistic program $\mathcal{Q}$ proposes a new subtree by sampling from the same process used to recursively define the prior distribution:
+\begin{center}
+\begin{tabular}{c}
+{\begin{lstlisting}[basicstyle=\small\ttfamily]
+new_subtree ~ cov_function_prior()
+\end{lstlisting}}
+\end{tabular}
+\end{center}
+The involution program swaps the old subtree with the newly proposed subtree:
+\begin{center}
+\begin{tabular}{c}
+{\begin{lstlisting}[basicstyle=\small\ttfamily]
+@copy(p_in, subtree_kddress, q_out, :new_subtree)
+@copy(q_in, :new_subtree, p_out, subtree_kddress)
+\end{lstlisting}}
+\end{tabular}
+\end{center}
+Note that the subtrees here happen to be trees that themselves have a random number of random choices, but the same general construction for a mixture of proposals below can also be used for simpler Metropolis-Hastings kernels:
+\begin{enumerate}
+\item The auxiliary probabilistic program samples from a distribution over different sets of random choices that will be proposed to
+(in this case, each set is a different subtree of the parse tree).
+\item The auxiliary probabilistic program them samples new values for those random choices (in this case, a new subtree).
+\item The involution program swaps the previous values of those random choices with their new values, by swapping data between the model trace and the auxiliary trace.
+\item The involution program copies the choice of what subset of random choices to propose to from the input auxiliary trace to the output auxiliary trace.
+\end{enumerate}
+
+Because the mixture distribution is specified using a probabilistic program, it is straightforward to modify the program to define a different mixture distribution.
 The code below specifies a mixture distribution that is uniform over all nodes in the tree.
@@
 \end{center}
-The resulting distributions are:
+The resulting distributions, for two input trees, are:
 \begin{center}
@@
 \end{center}
-Note that in the first case, the ratio of mixture probabilities is either $1$, $0.5$, or $2$ depending on whether the previous and new subtrees are leaf or internal nodes.
-In the second case, the ratio of mixture probabilities is the ratio of sizes of the two trees.
-%Automating the computation of the mixture probabilities
-
-The rest of the kernel specifies what happens for each chosen node.
-The second part of the auxiliary probabilistic program $Q$ implements a common motif for Metropolis-Hastings algorithms, where the proposal samples new values for some random choices:
-\begin{center}
-\begin{tabular}{c}
-{\begin{lstlisting}[basicstyle=\small\ttfamily]
-new_subtree ~ cov_function_prior()
-\end{lstlisting}}
-\end{tabular}
-\end{center}
-and the involution swaps the old value with the newly proposed value:
-\begin{center}
-\begin{tabular}{c}
-{\begin{lstlisting}[basicstyle=\small\ttfamily]
-@copy(p_in, subtree_kddress, q_out, :new_subtree)
-@copy(q_in, :new_subtree, p_out, subtree_kddress)
-\end{lstlisting}}
-\end{tabular}
-\end{center}
-Here the values happen to be trees that themselves have a random number of random choices, but the same motif is also used in simpler applications of Metropolis-Hastings.
-
-Here, the proposal we are mixing over is easy to implement -- it involvs resampling a subtree from the prior, so the Jacobian is 1 and there is cancellation between the model and the proposal.
-
-But computing the mixture probabilities is complex, because the mixture is a mixture of either $n_1$ or $n_2$ elements, which depends on the size of the tree.
-
-Show the mixture probabilities in math notation?
-If both subtrees not leaf nodes, then the probabilities are the same.
-If both subtrees are leaf nodes, then the probabilities are different.
-If one subtree is a leaf node and the other is an internal node then the ratio is either $0.5$ or $2$ respectively.
-\begin{equation}
-q
-\end{equation}
-
-TODO: relate to `resimulation Metropolis-Hastings' in general-purpose probabilistic programs.
-
-Of course, it is straightforward to replace the prior with a custom data-driven proposal distribution on subtrees.
+Note that the mixture probabilities do not in general cancel, and must be accounted for in the acceptance probability calculation.
+For the original mixture distribution, the ratio of mixture probabilities is either $1$, $0.5$, or $2$ depending on whether the previous and new subtrees are leaf or internal nodes.
+For this alternative mixture distribution, the ratio of mixture probabilities is the ratio of sizes of the two trees.
+In both cases, our automated involutive MCMC algorithm automatically computes the acceptance probability.
 

### 2020-07-18 paper.tex
add initial discussion section

@@
 {\begin{lstlisting}[basicstyle=\small\ttfamily]
-@copy(q_in, :path, q_out, :path)
+@copy((aux, :path), (aux, :path))
 \end{lstlisting}}
@@
 {\begin{lstlisting}[basicstyle=\small\ttfamily]
-@copy(p_in, subtree_kddress, q_out, :new_subtree)
-@copy(q_in, :new_subtree, p_out, subtree_kddress)
+@copy((model, subtree_address), (model, :new_subtree))
+@copy((aux, :new_subtree), (aux, subtree_address))
 \end{lstlisting}}
@@
 \end{center}
-Note that the mixture probabilities do not in general cancel, and must be accounted for in the acceptance probability calculation.
+Note that the probability of choosing a given subtree to propose to is itself changed when the subtree changes.
+Therefore, the mixture probabilities must be accounted for in the acceptance probability calculation.
 For the original mixture distribution, the ratio of mixture probabilities is either $1$, $0.5$, or $2$ depending on whether the previous and new subtrees are leaf or internal nodes.
@@
 
-- using probabilistic programming systems to develop new MCMC algorithms.
-
-- trace translators
-
-- work on compilers for inference programs that generate more specialized, high
-  performance implementations, from user specification of an inference
-algorithm.
+Involutive MCMC is a unifying construction for MCMC algorithms that encompasses both classic approaches to constructing kernels like reversible jump MCMC~\citep{green1995reversible}, but also recently introduced classes of MCMC kernels based on neural networks~\citep{spanbauer2020deep}.
+Therefore, the approach to automating involutive MCMC kernels presented in this paper makes a number of classic MCMC techniques easier to use and broadens their accessibility, and may potentially aid in development of novel MCMC techniques as well.
+The implementation of our approach in the Gen probabilistic programming system has already been used by researchers in computational biology~\citep{merrell2020inferring} and artificial intelligence~\citep{zhi2020online} to prototype and develop new reversible-jump MCMC algorithms.
+
+The technique for automating involutive MCMC presented in this paper can be generalized to the setting of sequential Monte Carlo samplers~\citep{del2006sequential}.
+Instead of one model probabilistic program, one auxiliary probabilistic program and one involution program, there are two model probabilistic programs, two auxiliary probabilistic programs, and a pair of bijective differentiable programs that transform traces of one model into traces of the other.
+This more general construct, which builds on earlier work on sequential Monte Carlo and probabilistic programs~\citep{cusumano2018incremental}, has already been implemented as part of the Gen probabilistic programming system.
+
+Improving the performance of automated involutive MCMC and of flexible probabilistic programming systems like Gen more generally is an important area for future work.
+The approach described in this paper is largely dynamic and is not performance-competitive with optimized hand-coded implementations in performance-oriented languages like C.
+While our approach is already valuable for use cases where the best performance is not necessary or the expertise or time needed for an optimized hand-coded implementation is not available,
+more research into compilers and automatic code generation of custom inference algorithms from high-level user specifications
+would broaden the applicability of systems like Gen.
 

### 2020-07-18 paper.tex
add to related work

@@
 More recently,~\citet{spanbauer2020deep} defined a class of deep generative model based on differentiable involutions and trained these models to serve as efficient proposal distributions on continuous state spaces; the resulting algorithm is an instance of the construction presented in this paper.
-%Gen~\citet{cusumano2019gen}
-%Gen's involution MCMC construct~\citet{gen-inv-mcmc}
-%Reversible jump MCMC~\citet{green1995reversible}
-%^Using probabilistic programs as proposals~\citet{cusumano2018using}
-%Reversible jump probabilistic programming~\citet{roberts2019reversible}
+
+In recent decades, a number of probabilistic programming systems have provided automated implementations of probabilistic inference algorithms~\citep{gilks1994language,milch2007,pfeffer200714,goodman2012church,gehr2016psi,carpenter2017stan}.
+The majority of these systems have focused on the automation of one or more built-in inference algorithms, with customization limited tuning parameters of the algorithm.
+The programmable inference paradigm~\citep{mansinghka2018probabilistic} to probabilistic programming instead emphasizes user control over the inference algorithm and automation of the lower-level implementation.
+In particular, the Gen probabilistic programming system~\citep{cusumano2019gen} offers a high degree of user programmability by exposing an API that supports high-level user implementations of inference algorithms, instead of a set of a built-in inference engine.
+One other probabilistic programming system besides Gen supports custom reversible jump samplers:
+\citet{roberts2019} presents a system embedded in Haskell that can automatically generate the implementation of certain reversible jump MCMC kernels from a high-level specification, but does not handle many kernels that can be handled by our technique, including the example in Figure~\ref{fig:structure-learning}.
 

### 2020-07-18 paper.tex
remove cycle ga

@@
 
-As a final, rather different example in this class, suppose approximate inverses $F_\theta, G_\theta : \mathbb{R}^d \rightarrow \mathbb{R}^d$ are learned neural networks trained via a Cycle GAN objective~\citep{cyclegan} to translate between the latent spaces of two models in a Bayesian model selection problem.
+%As a final, rather different example in this class, suppose approximate inverses $F_\theta, G_\theta : \mathbb{R}^d \rightarrow \mathbb{R}^d$ are learned neural networks trained via a Cycle GAN objective~\citep{cyclegan} to translate between the latent spaces of two models in a Bayesian model selection problem.
 A user might implement the following involution:
-
-\begin{center}
-\begin{tabular}{c}
-{\begin{lstlisting}[basicstyle=\small\ttfamily]
-@tracefn h from (tr1, tr2) to (tr3, tr4)
-   which_model = @read(tr1, :model, :discrete)
-   if which_model == 1
-      z = @read(tr1, :z, :continuous)
-      z_in_model_2 = F(z)
-      @write(tr3, :z, z_in_model_2, :continuous)
-      @write(tr3, :model, 2, :discrete)
-   else
-      z = @read(tr1, :z, :continuous)
-      z_in_model_1 = G(z)
-      @write(tr3, :z, z_in_model_1, :continuous)
-      @write(tr3, :model, 1, :discrete)
-   end
-end
-\end{lstlisting}}
-\end{tabular}
-\end{center}
-
-However, even though $F_\theta$ and $G_\theta$ are trained with
-an objective that encourages $F_\theta(G_\theta(z)) \approx z$, 
-this relationship is not exact equality, so the involutive MCMC
-framework cannot be directly applied, and the involution check
-will fail. This error could be fixed by sampling a proposed value
-\textit{near} $F_\theta(z)$ or $G_\theta(z)$, rather than directly applying $F_\theta$
-and $G_\theta$ to $z$.
+%
+%\begin{center}
+%\begin{tabular}{c}
+%{\begin{lstlisting}[basicstyle=\small\ttfamily]
+%@tracefn h from (tr1, tr2) to (tr3, tr4)
+   %which_model = @read(tr1, :model, :discrete)
+   %if which_model == 1
+      %z = @read(tr1, :z, :continuous)
+      %z_in_model_2 = F(z)
+      %@write(tr3, :z, z_in_model_2, :continuous)
+      %@write(tr3, :model, 2, :discrete)
+   %else
+      %z = @read(tr1, :z, :continuous)
+      %z_in_model_1 = G(z)
+      %@write(tr3, :z, z_in_model_1, :continuous)
+      %@write(tr3, :model, 1, :discrete)
+   %end
+%end
+%\end{lstlisting}}
+%\end{tabular}
+%\end{center}
+
+%However, even though $F_\theta$ and $G_\theta$ are trained with
+%an objective that encourages $F_\theta(G_\theta(z)) \approx z$, 
+%this relationship is not exact equality, so the involutive MCMC
+%framework cannot be directly applied, and the involution check
+%will fail. This error could be fixed by sampling a proposed value
+%\textit{near} $F_\theta(z)$ or $G_\theta(z)$, rather than directly applying $F_\theta$
+%and $G_\theta$ to $z$.
 

### 2020-07-18 experiments/pytorch/minigen.py
use new syntax in minigen

@@ -64,6 +64,20 @@ InvolutionRunnerState = namedtuple("InvolutionRunnerState",
 MODEL = "model"
 AUX = "aux"
 
+# for syntactic sugar
+class ModelSugar:
+    def __getitem__(self, key):
+        return (MODEL, key)
+
+class AuxSugar:
+    def __getitem__(self, key):
+        return (AUX, key)
+
+model_in = ModelSugar()
+model_out = ModelSugar()
+aux_in = AuxSugar()
+aux_out = AuxSugar()
+
 def get_which(addr):
     exc = Exception("address argument must be (MODEL, *) or (AUX, *)")
     if len(addr) != 2:
@@ -238,24 +252,24 @@ def cartesian_to_polar(x, y):
     return (theta, y)
 
 def f():
-    polar = read((MODEL, "polar"), DISCRETE)
+    polar = read(model_in["polar"], DISCRETE)
     if polar:
-        r = read((MODEL, "r"), CONTINUOUS)
-        theta = read((MODEL, "theta"), CONTINUOUS)
+        r = read(model_in["r"], CONTINUOUS)
+        theta = read(model_in["theta"], CONTINUOUS)
         (x, y) = polar_to_cartesian(r, theta)
-        write((MODEL, "x"), x, CONTINUOUS)
-        write((MODEL, "y"), y, CONTINUOUS)
+        write(model_out["x"], x, CONTINUOUS)
+        write(model_out["y"], y, CONTINUOUS)
     else:
-        x = read((MODEL, "x"), CONTINUOUS)
-        y = read((MODEL, "y"), CONTINUOUS)
+        x = read(model_in["x"], CONTINUOUS)
+        y = read(model_in["y"], CONTINUOUS)
         (r, theta) = cartesian_to_polar(x, y)
-        write((MODEL, "r"), r, CONTINUOUS)
-        write((MODEL, "theta"), theta, CONTINUOUS)
-    write((MODEL, "polar"), not polar, DISCRETE)
-    copy((MODEL, "u"), (MODEL, "u"))
-    u = read((MODEL, "u"), CONTINUOUS)
-    v = read((MODEL, "v"), CONTINUOUS)
-    write((MODEL, "v"), u - v, CONTINUOUS)
+        write(model_out["r"], r, CONTINUOUS)
+        write(model_out["theta"], theta, CONTINUOUS)
+    write(model_out["polar"], not polar, DISCRETE)
[…]

### 2020-07-18 experiments/pytorch/minigen.py
update syntax

@@ -88,17 +88,17 @@ def get_which(addr):
     return (which, addr)
 
 # user-provided type information for reads and writes
-CONTINUOUS = "continuous"
-DISCRETE = "discrete"
+continuous = "continuous"
+discrete = "discrete"
 
 def check_type_label(type_label):
-    if type_label != CONTINUOUS and type_label != DISCRETE:
-        raise Exception("type label argument must be CONTINUOUS or DISCRETE")
+    if type_label != continuous and type_label != discrete:
+        raise Exception("type label argument must be continuous or discrete")
 
 def _read(state, addr, type_label):
     (which, addr) = get_which(addr)
     check_type_label(type_label)
-    if type_label == CONTINUOUS:
+    if type_label == continuous:
         if which == MODEL:
             val = torch.tensor(state.input_model_trace[addr], requires_grad=True)
         else:
@@ -114,7 +114,7 @@ def _read(state, addr, type_label):
 def _write(state, addr, val, type_label):
     (which, addr) = get_which(addr)
     check_type_label(type_label)
-    if type_label == CONTINUOUS:
+    if type_label == continuous:
         if which == MODEL:
             state.output_model_trace[addr] = val 
         else:
@@ -252,24 +252,24 @@ def cartesian_to_polar(x, y):
     return (theta, y)
 
 def f():
-    polar = read(model_in["polar"], DISCRETE)
+    polar = read(model_in["polar"], discrete)
     if polar:
-        r = read(model_in["r"], CONTINUOUS)
-        theta = read(model_in["theta"], CONTINUOUS)
+        r = read(model_in["r"], continuous)
+        theta = read(model_in["theta"], continuous)
         (x, y) = polar_to_cartesian(r, theta)
-        write(model_out["x"], x, CONTINUOUS)
-        write(model_out["y"], y, CONTINUOUS)
+        write(model_out["x"], x, continuous)
+        write(model_out["y"], y, continuous)
     else:
-        x = read(model_in["x"], CONTINUOUS)
-        y = read(model_in["y"], CONTINUOUS)
+        x = read(model_in["x"], continuous)
+        y = read(model_in["y"], continuous)
         (r, theta) = cartesian_to_polar(x, y)
-        write(model_out["r"], r, CONTINUOUS)
-        write(model_out["theta"], theta, CONTINUOUS)
-    write(model_out["polar"], not polar, DISCRETE)
[…]

### 2020-07-18 paper.tex
minor proofreading corrections

@@
 Involutive MCMC is a unifying mathematical construction for MCMC kernels that generalizes many classic and state-of-the-art MCMC algorithms, from reversible jump MCMC to kernels based on deep neural networks.
-But like MCMC samplers more generally, implementing involutive MCMC kernels is often tedious and error-prone, especially when sampling on complex state spaces.
+But as with MCMC samplers more generally, implementing involutive MCMC kernels is often tedious and error-prone, especially when sampling on complex state spaces.
 This paper describes a technique for automating the implementation of involutive MCMC kernels given (i) a pair of probabilistic programs defining the target distribution and an auxiliary distribution respectively and (ii) a differentiable program that transforms the execution traces of these probabilistic programs.
@@
 \section{INTRODUCTION}
-Markov chain Monte Carlo (MCMC) algorithms are powerful tools for approximate sampling from probability distributions and are central to modern Bayesian statistics, probabilistic machine learning, statistical physics, and numerous applications of probabilistic inference.
+Markov chain Monte Carlo (MCMC) algorithms are powerful tools for approximate sampling from probability distributions and are central to modern Bayesian statistics, probabilistic machine learning, statistical physics, and numerous application areas where probabilistic modeling and inference are used.
 But designing and deriving efficient MCMC algorithms is mathematically involved, and implementing MCMC kernels is tedious and notoriously error-prone.
-These challenges are especially pertinent when sampling from probability distributions on complex state spaces that combine symbolic, numeric, and structural uncertainty, such as those arising in computational biology~\citep{huelsenbeck2004bayesian}, robotics and scene understanding~\citep{geiger2011generative}, and models of human cognition~\citep{tenenbaum2011grow}.
+These challenges are especially pronounced when sampling from probability distributions on complex state spaces that combine symbolic, numeric, and structural uncertainty, such as those arising in computational biology~\citep{huelsenbeck2004bayesian}, robotics and scene understanding~\citep{geiger2011generative}, and models of human cognition~\citep{tenenbaum2011grow}.
 
@@
 including reversible jump MCMC~\citep{green1995reversible}, which is the dominant mathematical framework for MCMC on complex state spaces.
-Involutive MCMC constructs an MCMC kernel from three components: (i) the unnormalized target density, (ii) a sampler and density for an auxiliary probability distribution (iii) an involution\footnote{a bijection that is its own inverse ($f$ where $f(f(z)) = z$)} on an extended state space.
+Involutive MCMC constructs an MCMC kernel from three components: (i) the unnormalized target density, (ii) a sampler and density for an auxiliary probability distribution, and (iii) an involution\footnote{a bijection that is its own inverse ($f$ where $f(f(z)) = z$)} on an extended state space.
 While this construction is mathematically clarifying, correctly implementing involutive MCMC kernels on complex state spaces remains challenging due to tedious density and Jacobian computations and the need for careful reasoning about the state space.
@@
 We implemented the approach within the Gen probabilistic programming system\footnote{\url{https://www.gen.dev}}~\citep{cusumano2019gen}.
-The paper shows examples of involutive MCMC kernels implemented in Gen for (i) a split-merge reversible jump move in an infinite mixture model, and a (ii) state-dependent mixture of Metropolis-Hastings proposals on an infinite combinatorial space of covariance functions for a Gaussian process.
+The paper shows examples of involutive MCMC kernels implemented in Gen for (i) a split-merge reversible jump move in an infinite mixture model, and (ii) a state-dependent mixture of Metropolis-Hastings proposals on an infinite combinatorial space of covariance functions for a Gaussian process.
 We also provide a lightweight PyTorch implementation of the basic approach at
@@
 \item An algorithm that automates the implementation of an involutive MCMC kernel given two probabilistic programs and a differentiable program encoding the involution, using automatic differentiation and tracing of probabilistic programs.
-\item An extension to the algorithm that exploits sparsity in the involution to reduce the computational complexity of an involutive MCMC kernel from $O(N^2)$ to $O(1)$ in common cases where $N$ is the dimensionality of the latent space.
+\item An extension to the algorithm that exploits sparsity in the involution to reduce the computational complexity of an involutive MCMC kernel from $O(N^2)$ to $O(1)$ in common cases, where $N$ is the dimensionality of the latent space.
 \item An algorithm that dynamically detects errors in the specification of an involutive MCMC kernel.
@@
 The construction was motivated in part by a desire for a simple interface that automated the implementation of reversible jump MCMC samplers~\citep{green1995reversible}, state-dependent mixtures of proposals on complex state spaces, and data-driven neural proposals.
-Gen's involutive MCMC construction has since been used by a number of researchers to design and implement MCMC algorithms in diverse domains including computational biology~\citep{merrell2020inferring} and artificial intelligence~\citep{zhi2020online}.
-
-\citet{imcmc} independently identified the involutive MCMC construction as a unifying framework for MCMC algorithms, and showed how a number of classic and recent MCMC algorithms can be cast within this framework.
+Gen's involutive MCMC construction has since been used by a number of researchers to design and implement MCMC algorithms in diverse domains, including computational biology~\citep{merrell2020inferring} and artificial intelligence~\citep{zhi2020online}.
+
+\citet{imcmc} independently identified the involutive MCMC construction as a unifying framework for MCMC algorithms, and showed how more than a dozen classic and recent MCMC algorithms can be cast within this framework.
 \citet{imcmc} also identified design principles for developing new MCMC algorithms using the involutive MCMC construction, and showed that the framework aids in the derivation of novel efficient MCMC algorithms.
@@
 In recent decades, a number of probabilistic programming systems have provided automated implementations of probabilistic inference algorithms~\citep{gilks1994language,milch2007,pfeffer200714,goodman2012church,gehr2016psi,carpenter2017stan}.
-The majority of these systems have focused on the automation of one or more built-in inference algorithms, with customization limited tuning parameters of the algorithm.
-The programmable inference paradigm~\citep{mansinghka2018probabilistic} to probabilistic programming instead emphasizes user control over the inference algorithm and automation of the lower-level implementation.
+The majority of these systems have focused on the automation of one or more built-in inference algorithms, with customization limited to tuning a handful of parameters of the algorithm.
+The programmable inference paradigm~\citep{mansinghka2018probabilistic} in probabilistic programming instead emphasizes user control over the inference algorithm and automation of the lower-level implementation.
 In particular, the Gen probabilistic programming system~\citep{cusumano2019gen} offers a high degree of user programmability by exposing an API that supports high-level user implementations of inference algorithms, instead of a set of a built-in inference engine.
 One other probabilistic programming system besides Gen supports custom reversible jump samplers:
-\citet{roberts2019} presents a system embedded in Haskell that can automatically generate the implementation of certain reversible jump MCMC kernels from a high-level specification, but does not handle many kernels that can be handled by our technique, including the example in Figure~\ref{fig:structure-learning}.
+\citet{roberts2019} present a system embedded in Haskell that can automatically generate the implementation of certain reversible jump MCMC kernels from a high-level specification, and \citet{narayanan2020symbolic} also give a technique that can be used to automatically compute the Metropolis-Hastings acceptance probabilities in some settings; however, these approaches do not handle many kernels that can be handled by our technique, including the example in Figure~\ref{fig:structure-learning}.
 
@@
 an auxiliary variable $y \sim q_x$ from a state-dependent auxiliary distribution. We then apply an \textit{involution} $f$ to the pair $(x, y)$ to obtain $(x', y')$. 
-Finally, we computes an acceptance probability $\alpha$ and either accept $x'$ as the new state, or reject it and repeat the previous state $x$.
+Finally, we compute an acceptance probability $\alpha$ and either accept $x'$ as the new state, or reject it and repeat the previous state $x$.
 Different choices of $q$ and $f$ recover many algorithms from the literature~\citep{imcmc}.
@@
 This section describes probability distributions on dictionaries, and gives a constructive definition of the involutive MCMC acceptance probability in this setting in terms of a Jacobian.
-Section~\ref{sec:automating} will then show how probability distributions on dictionaries can be specified with probabilistic programs, and how probabilistic programming techniques can automatically compute probability densities spaces of dictionaries.
+Section~\ref{sec:automating} will then show how probability distributions on dictionaries can be specified with probabilistic programs, and how probabilistic programming techniques can automatically compute probability densities on spaces of dictionaries.
 
@@
 \citet{green1995reversible} uses a state space that is the countable union of `models', where for each model there is a vector of real-valued parameters.
-Dictionaries possesses substantially more structure:
+Dictionaries possess substantially more structure:
 Instead of monolithic `models', dictionaries use a more elaborate discrete state that includes the set of keys and the assignment to the discrete keys. 
-Also, continuous keys play the role of real-valued parameters, so it is possible to express that a given real-valued parameter is shared between models.
+Also, because continuous keys play the role of real-valued parameters, it is possible to express that a given real-valued parameter is shared between models.
 The additional structure of dictionaries enables the automation techniques in Section~\ref{sec:automating}.
@@
 \textbf{A measure space of finite dictionaries.}
-We associate a measure $\mu_k$ on $V_k$ for each key---the counting measure for each discrete key and the Lebesgue-measure on $\mathbb{R}^{d_k}$ for each continuous key.
+We associate a measure $\mu_k$ on $V_k$ with each key $k$---the counting measure for each discrete key and the Lebesgue-measure on $\mathbb{R}^{d_k}$ for each continuous key.
 For each finite set of keys $K$, we make $V_K$ a measure space using the standard product $\sigma$-algebra $\Sigma_K = \otimes_{k \in A} \Sigma_k$ and the product measure $\mu_K = \times_{k \in A} \mu_k$.
@@
 \textbf{Probability distributions on finite dictionaries.} 
-When all keys are discrete, a probability distribution on dictionaries is defined by a $p$ is a probability mass function $p : \mathcal{D} \to [0, 1]$ that assigns a probability $p(m)$ to each dictionary $m \in \mathcal{P}$ such that $\sum_{x \in \mathcal{D}} p(m) = 1$.
+When $V_k$ is discrete for all keys $k$, a probability distribution on dictionaries is defined by a probability mass function $p : \mathcal{D} \to [0, 1]$ that assigns a probability $p(m)$ to each dictionary $m \in \mathcal{P}$ such that $\sum_{m \in \mathcal{D}} p(m) = 1$.
 More generally a probability distribution on dictionaries is defined by a probability density $p : \mathcal{D} \rightarrow [0, \infty)$ such that $\int p(m) \mu_\mathcal{D}(\text{d}m) = 1$. 
-The probability is distributed the finite sets of keys $K \subseteq \mathcal{K}$ where $|K| < \infty$:
+The probability is distributed among the finite sets of keys $K \subseteq \mathcal{K}$ where $|K| < \infty$:
 \begin{equation}
@@
 In this case, we have $V_\texttt{k} = \mathbb{N}$ with the counting measure for $\mu_\texttt{k}$, and for all other keys $k \in \mathcal{K}$, $V_k = \mathbb{R}$ with the Lebesgue measure for $\mu_k$.
-For each $k \in \{0, 1, \ldots\}$, the probability for key set
-$\{\texttt{k}, \texttt{y}_\texttt{1}, \dots, \texttt{y}_\texttt{n}\} \cup \{\texttt{m}_i \mid 1 \leq i \leq k\} \cup \{\texttt{s}_i \mid 1 \leq i \leq k\}$
-is $p_{\mathrm{poisson}(s)}(k)$ (via Equation~(\ref{eq:integral})).
+For each $j \in \{0, 1, \ldots\}$, the probability for key set
+$\{\texttt{k}, \texttt{y}_\texttt{1}, \dots, \texttt{y}_\texttt{n}\} \cup \{\texttt{m}_i \mid 1 \leq i \leq j\} \cup \{\texttt{s}_i \mid 1 \leq i \leq j\}$
+is $p_{\mathrm{poisson}(s)}(j)$ (via Equation~(\ref{eq:integral})).
 
@@
 % TODO: this is only needed for Algorithm 2, and it is relevant for the 
-To simplify the notation, and without loss of generality, we will assume that $\mathcal{K}_P$ and $\mathcal{K}_Q$ are disjoint\footnote{If $\mathcal{K}_P$ and $\mathcal{K}_Q$ are not disjoint then, they can be made so by adding a different prefix to the keys of each set.}, and we define $Z := \{ x \oplus y : p(x)q_x(y) > 0\} \subseteq \mathcal{D}$.
+To simplify the notation, and without loss of generality, we will assume that $\mathcal{K}_P$ and $\mathcal{K}_Q$ are disjoint\footnote{If $\mathcal{K}_P$ and $\mathcal{K}_Q$ are not disjoint then, they can be made so by adding a different prefix to the keys of each set.}, and we define $Z := \{ x \oplus y : p(x)q_x(y) > 0\} \subseteq \mathcal{D}$,
 where $\mathcal{D}$ is the set of dictionaries on keys from $\mathcal{K}_P \cup \mathcal{K}_Q$
@@
 
-Suppose there is a countable partition of $Z$ into $\{Z_e : e \in E\}$ such that across each $Z_e$ the sets of keys $K_x$ and $K_y$ and the values $x_k$ and $y_k$ for all discrete keys $k$ are both constant.
+Suppose there is a countable partition of $Z$ into $\{Z_e : e \in E\}$ such that if $(K_1, \mathbf{x}_1)$ and $(K_2, \mathbf{x}_2)$ are two dictionaries in the same component $Z_e$, then $K_1 = K_2$ and they agree on all \textit{discrete} values: $x_{1k} = x_{2k}$ for all $k \in K_1 \cap \mathcal{I}$.
 Then each set $Z_e$ is isomorphic to a Euclidean space of assignments to the continuous keys in the two dictionaries.
-Suppose there is an involution $g : E \to E$ between elements of the partition (a function with $g = g^{-1}$), and a family of continuously differentiable bijections $h_{e} : Z_e \to Z_{g(e)}$ for each $e$ with $h_{e} = h_{g(e)}^{-1}$.
-Let $e(z) \in E$ denote the element of the partition for an alement $z \in Z$.
-Then, $f : Z \to Z$ given by $f(z) := h_{e(z)}(z)$ is an involution:
+Suppose there is an involution $g : E \to E$ between elements of the partition (a function with $g = g^{-1}$), and a family of continuously differentiable bijections $h_{e} : Z_e \to Z_{g(e)}$ indexed by $e \in E$, with $h_{e} = h_{g(e)}^{-1}$.
+Let $e(z) \in E$ denote which element of the partition a dictionary $z \in Z$ belongs to.
+Then $f : Z \to Z$ given by $f(z) := h_{e(z)}(z)$ is an involution:
 \begin{equation*}
-f(f(z)) = h_{e(f(z))}(h_{e(z)}(z)) = h_{g(e(z))}(h_{e(z)}(z)) = z
+f(f(z)) = h_{e(f(z))}(h_{e(z)}(z)) = h_{g(e(z))}(h_{e(z)}(z)) = z.
 \end{equation*}
-Let $|J h_e|(z)$ denote the absolute value of the Jacobian (determinant) of $h_e$, evaluated at $z$.
+Let $|J h_e|(z)$ denote the absolute value of the determinant of the Jacobian of $h_e$, evaluated at $z$.
 Then, the acceptance probability in Algorithm~\ref{alg:involutive-mcmc} simplifies to:
@@
 (dictionaries are equivalent if they contain the same keys and they agree on the value of all discrete keys).
-Section~\ref{sec:radon-nikodym-special-case} of the appendix for details.
+See Section~\ref{sec:radon-nikodym-special-case} of the appendix for details.
 
@@
 
-Thus, this ratio can be computed term-by-term using \textsc{score} on the probabilistic programs $\mathcal{P}$ and $\mathcal{Q}$: (Algorithm~\ref{alg:auto-involutive-mcmc}): to compute $q_x(y')$, we run the algorithm directly on the program $\mathcal{Q}_x$, and to compute $p(x)$ and $p(x')$, we actually merge the dictionary $x$ with the observations $b$ and compute $\mathcal{L}(b)p(x) = \tilde{p}(x \oplus b)$ instead, by running the algorithm on $\mathcal{P}$ with trace $x \oplus b$.
+Thus, this ratio can be computed term-by-term using \textsc{score} on the probabilistic programs $\mathcal{P}$ and $\mathcal{Q}$ (Algorithm~\ref{alg:auto-involutive-mcmc}): to compute $q_x(y')$, we run the algorithm directly on the program $\mathcal{Q}_x$, and to compute $p(x)$ and $p(x')$, we actually merge the dictionary $x$ with the observations $b$ and compute $\mathcal{L}(b)p(x) = \tilde{p}(x \oplus b)$ instead, by running the algorithm on $\mathcal{P}$ with trace $x \oplus b$.
 The density $q_x(y)$ can be computed while $y$ is being sampled, using \textsc{trace-and-score}.
 
-Section~\ref{sec:sparsity} describes a more efficient approach that exploits sparsity in the involution and resulting cancellation in the ratio for improved efficiency, but requires a more sophisticated probabilistic programming runtime system.
+Section~\ref{sec:sparsity} describes a more efficient approach that exploits sparsity in the involution (and the resulting cancellations in the acceptance ratio) for improved efficiency, but requires a more sophisticated probabilistic programming runtime system.
 
@@
 Section~\ref{sec:probabilistic-programming} showed that if the densities $p$ and $q$ are specified using probabilistic programs $\mathcal{P}$ and $\mathcal{Q}$, then density ratio in the acceptance probability for involutive MCMC on dictionaries Equation~(\ref{eq:acceptance-ratio-dictionaries}) can be automated using probabilistic programming techniques.
-This section shows that if the involution $f$ is specified in using a differentiable program $\mathcal{F}$ that transforms the execution traces of probabilistic programs, then the Jacobian factor in Equation~(\ref{eq:acceptance-ratio-dictionaries}) can be automated using automatic differentiation.
+This section shows that if the involution $f$ is specified in using a differentiable program $\mathcal{F}$ that transforms the execution traces of probabilistic programs, then the Jacobian factor in Equation~(\ref{eq:acceptance-ratio-dictionaries}) can also be automated, using automatic differentiation.
 The procedure \textproc{auto-involutive-mcmc} in Algorithm~\ref{alg:auto-involutive-mcmc} combines these two ideas and automates involutive MCMC given the programs $\mathcal{P}$, $\mathcal{Q}$, and $\mathcal{F}$.
@@
 Recall that the involution $f$ must decompose into (i) an involution $g$ on elements $e \in E$ of a partition of the state space, and (ii) a pair of continuous differentiable bijections $h_{e}$ and $h_{g(e)} = h_{e}^{-1}$ between each pair of corresponding elements of the partition.
-Each function $h_e$ is a function from the values at continuous addresses in the input trace ($z[a]$ for $a \in A_z \cap \mathcal{J}$) to the values at continuous addresses in the output trace ($z'[a]$ for $a \in A_{z'} \cap \mathcal{J}$).
-We compute the Jacobian using automatic differentiation (AD);
-different approaches are possible.
-The Gen implementation uses forward-mode AD whereas our PyTorch implementation uses reverse-mode AD.
+Each function $h_e$ is a function from the values at continuous addresses in the input trace ($z[k]$ for $k \in K_z \cap \mathcal{J}$) to the values at continuous addresses in the output trace ($z'[k]$ for $k \in K_{z'} \cap \mathcal{J}$).
+We compute the Jacobian using automatic differentiation. (The Gen implementation uses forward-mode AD whereas our PyTorch implementation uses reverse-mode AD.)
 The procedure \textproc{run-involution} in Algorithm~\ref{alg:auto-involutive-mcmc} shows the implementation of the interpreter for the language that uses reverse-mode AD.
 The interpreter executes $\mathcal{F}$ using regular Julia or Python semantics, but intercepts calls to \texttt{@write}, \texttt{@read}, and \texttt{@copy} statements, and in addition to performing the desired operation, records the set of continuous addresses that are read, written, and copied.
-After $\mathcal{F}$ is finished executing, AD uses the recorded addresses to compute the Jacobian $|J h_e|(z)$; in the case of reverse-mode AD by iterating over output continuous addresses and backpropagating from each one to all of the input continous addresses, computing the Jacobian row-by-row.
+After $\mathcal{F}$ is finished executing, AD uses the recorded addresses to compute the Jacobian $|J h_e|(z)$; in the case of reverse-mode AD, this is accomplished by iterating over output continuous addresses and backpropagating from each one to all of the input continuous addresses, computing the Jacobian row-by-row.
 Note that the $\mathtt{discrete}$ and $\mathtt{continuous}$ labels are omitted from the syntax in \textproc{run-involution} to simplify notation ($a \in \mathcal{I}$ indicates a discrete choice and otherwise a choice is continuous).
@@
 The quadratic term represents the complexity of computing the Jacobian matrix and its determinant.
+
 It is possible to reduce the computational complexity of an automated involutive MCMC kernel by exploiting special structure in the involution $f$, in some cases to $O(1)$.
@@
 
-The naive implementation of Algorithm~\ref{alg:auto-involutive-mcmc} computes by the Jacobian by first computing the $M$-by-$M$ Jacobian matrix $J h_{e(z)}$ via automatic differentiation, and them computing the absolute value of its determinant.
+The naive implementation of Algorithm~\ref{alg:auto-involutive-mcmc} computes by the Jacobian by first computing the $M$-by-$M$ Jacobian matrix $J h_{e(z)}$ via automatic differentiation, and then computing the absolute value of its determinant.
 However, we observe that in many applications of involutive MCMC, the values at continuous choices in the input traces are directly copied into the output traces (either at the same, or different, keys).
-These copy operations result in columns in the Jacobian matrix that have a single $1$ entry with remaining entries are $0$.
+These copy operations result in columns in the Jacobian matrix that have a single $1$ entry with remaining entries equal to $0$.
 For example, for the function $(u, v, x, y) \mapsto (u, 2 u - v, y, x) = (u', v', x', y')$, the Jacobian is (with columns corresponding to $u'$, $v'$, $x'$, and $y'$ and rows corresponding to $u$, $v$, $x$, and $y$):
@@
 \end{equation}
-Using the cofactor expansion of the determinant, we observe that for any (`copy') row or column in an $M$-by-$M$ Jacobian matrix with a single $1$ and all other entries $0$, that absolute value of the determinant is equivalent to that of the $(M-1)$-by-$(M-1)$ sub-matrix with the corresponding row and column omitted (even if that would remove other nonzero entries from the matrix).
+Using the cofactor expansion of the determinant, we observe that for any (`copy') row or column in an $M$-by-$M$ Jacobian matrix with a single $1$ and all other entries $0$, the absolute value of the determinant is equivalent to that of the $(M-1)$-by-$(M-1)$ sub-matrix with the corresponding row \textit{and} column omitted (even if that would remove other nonzero entries from the matrix).
 By applying this rule recursively, we can instead compute the determinant of a much smaller matrix; for the example above with $M=4$, the absolute value of the determinant simplifies to the absolute value of a single entry ($|-1|$).
@@
 (in this case, each set is a different subtree of the parse tree).
-\item The auxiliary probabilistic program them samples new values for those random choices (in this case, a new subtree).
+\item The auxiliary probabilistic program then samples new values for those random choices (in this case, a new subtree).
 \item The involution program swaps the previous values of those random choices with their new values, by swapping data between the model trace and the auxiliary trace.

### 2020-07-18 experiments/gen/mixture.jl
Update mixture example to use new syntax

@@ -183,18 +183,18 @@ end
    end
 end
 
-@bijection function split_merge_inv(_, _, _)
-    k = @read_discrete_from_model(:k)
-    split = (k == 1) ? true : @read_discrete_from_proposal(:split)
+@transform split_merge_inv (model_in, aux_in) to (model_out, aux_out) begin
+    k = @read(model_in[:k], :discrete)
+    split = (k == 1) ? true : @read(aux_in[:split], :discrete)
     if split
 
-        cluster_to_split = @read_discrete_from_proposal(:cluster_to_split)
-        u1 = @read_continuous_from_proposal(:u1)
-        u2 = @read_continuous_from_proposal(:u2)
-        u3 = @read_continuous_from_proposal(:u3)
-        weights = @read_continuous_from_model(:weights)
-        mu = @read_continuous_from_model((:mu, cluster_to_split))
-        var = @read_continuous_from_model((:var, cluster_to_split))
+        cluster_to_split = @read(aux_in[:cluster_to_split], :discrete)
+        u1 = @read(aux_in[:u1], :continuous)
+        u2 = @read(aux_in[:u2], :continuous)
+        u3 = @read(aux_in[:u3], :continuous)
+        weights = @read(model_in[:weights], :continuous)
+        mu = @read(model_in[(:mu, cluster_to_split)], :continuous)
+        var = @read(model_in[(:var, cluster_to_split)], :continuous)
 
         new_weights = split_weights(weights, cluster_to_split, u1, k)
         w1 = new_weights[cluster_to_split]
@@ -202,23 +202,23 @@ end
         (mu1, mu2) = split_means(mu, var, u2, w1, w2)
         (var1, var2) = split_vars(w1, w2, var, u2, u3)
 
-        @write_discrete_to_model(:k, k+1)
-        @copy_proposal_to_proposal(:cluster_to_split, :cluster_to_merge)
-        @write_discrete_to_proposal(:split, false)
-        @write_continuous_to_model(:weights, new_weights)
-        @write_continuous_to_model((:mu, cluster_to_split), mu1)
-        @write_continuous_to_model((:mu, k+1), mu2)
-        @write_continuous_to_model((:var, cluster_to_split), var1)
-        @write_continuous_to_model((:var, k+1), var2)
+        @write(model_out[:k], k+1, :discrete)
+        @copy(aux_in[:cluster_to_split], aux_out[:cluster_to_merge])
[…]

### 2020-07-18 paper.tex
update syntax in code figures and text

@@
 (2) a probabilistic program $\mathtt{q}$ encoding an auxiliary probability distribution (shown in c),
-and (3) a differentiable program $\mathtt{h}$ that encodes an involution on the space of pairs of traces of $\mathtt{p}$ and $\mathtt{q}$ (shown in d).
+and (3) a differentiable program $\mathtt{f}$ that encodes an involution on the space of pairs of traces of $\mathtt{p}$ and $\mathtt{q}$ (shown in d).
 Gen's involutive MCMC operator (shown in e) automatically computes the acceptance probability.
@@
 Recall that $x$ is the trace of the model probabilistic program, and $y$ is the trace of the auxiliary probabilistic program.
-Each address $a$ therefore is a pair, where the first element is either $\mathtt{model}$ or $\mathtt{aux}$, and the second element is the address within either $x$ or $y$.
-For example, to read the value of a continuous address $\mathtt{:}\mathtt{r}$ from the model trace ($x$):
+Each address $a$ therefore needs to specify which of these traces to read from, and the address within that trace.
+The traces are given names in the function signature:
 \noindent
@@
 \begin{tabular}{c}
+{\begin{lstlisting}[basicstyle=\footnotesize\ttfamily]
+@transform f (model_in, aux_in) to (model_out, aux_out)
+begin
+..
+end
+\end{lstlisting}}
+\end{tabular}
+\end{center}
+Here, the traces $x$, $y$, $x'$ and $y'$ are given names $\mathtt{model\_in}$, $\mathtt{aux\_in}$, $\mathtt{model\_out}$, and $\mathtt{aux\_out}$, respectively.
+The syntax for address $\mathtt{a}$ within trace $\mathtt{trace}$ is $\mathtt{trace[a]}$.
+For example, to read the value of a continuous address $\mathtt{:}\mathtt{r}$ from the input model trace ($x$):
+\noindent
+\begin{center}
+\begin{tabular}{c}
 {\begin{lstlisting}[basicstyle=\small\ttfamily]
-    val = @read((model, :r), continuous)
+    val = @read(model_in[:r], continuous)
 \end{lstlisting}}
@@
 {\begin{lstlisting}[basicstyle=\small\ttfamily]
-    @write((model, :r), val, continuous)
+    @write(model_in[:r], val, continuous)
 \end{lstlisting}}
@@
 {\begin{lstlisting}[basicstyle=\small\ttfamily]
-    @copy((model, :u), (model, :v))
+    @copy(model_in[:u], model_in[:v])
 \end{lstlisting}}
@@
 {\begin{lstlisting}[basicstyle=\small\ttfamily]
-r = @read((model, :r), :continuous)
-theta = @read((model, :theta), :continuous)
+r = @read(model_in[:r], :continuous)
+theta = @read(model_in[:theta], :continuous)
 x = r * sin(theta)
 y = r * cos(theta)
-@write((model, :x), x, :continuous)
-@write((model, :y), y, :continuous)
+@write((model_out[:x], x, :continuous)
+@write((model_out[:y], y, :continuous)
 \end{lstlisting}}
@@
 @tracefn f
-   polar = @read((model, :polar), discrete)
+   polar = @read(model_in[:polar], discrete)
    if polar
-      r = @read((model, :r), continuous)
-      theta = @read((model, :theta), continuous)
+      r = @read(model_in[:r]), continuous)
+      theta = @read(model_in[:theta], continuous)
       x = r * sin(theta)
       y = r * cos(theta)
-      @write((model, :x), x, continuous)
-      @write((model, :y), y, continuous)
+      @write(model_out[:x], x, continuous)
+      @write(model_out[:y], y, continuous)
    else
-      x = @read((model, :x), continuous)
-      y = @read((model, :y), continuous)
+      x = @read(model_in[:x], continuous)
+      y = @read(model_in[:y], continuous)
       r = (x^2 + y^2)
       theta = atan(y, x)
-      @write((model, :r), r, continuous)
-      @write((model, :theta), theta, continuous)
+      @write(model_out[:r], r, continuous)
+      @write(model_out[:theta], theta, continuous)
    end
-   @write((model, :polar), !polar, discrete)
+   @write(model_out[:polar], !polar, discrete)
 end
@@
 implementation of a \textit{birth-death} move for the mixture model in Figure~\ref{fig:mixture}, which either adds a new mixture component or selects an existing one
-at random to delete:
+at random to delete
+(we name the traces $\mathtt{tr1}$, $\mathtt{tr2}$, $\mathtt{tr3}$, ad $\mathtt{tr4}$ for reasons of space):
 
@@
 
-@tracefn h from (tr1, tr2) to (tr3, tr4)
+@transform h (tr1, tr2) to (tr3, tr4) begin
    is_birth = @read(tr2, :is_birth, :discrete)
-   @write(tr4, :is_birth, !is_birth, :discrete)
-   k = @read(tr1, :k, :discrete)
-   weights = @read(tr1, :weights, :continuous)
+   @write(tr4[:is_birth], !is_birth, :discrete)
+   k = @read(tr1[:k], :discrete)
+   weights = @read(tr1[:weights], :continuous)
    if is_birth
-     @write(tr3, :k, k+1, :discrete)
-     new_mu = @read(tr2, :new_mu, :continuous)
-     new_var = @read(tr2, :new_var, :continuous)
-     @write(tr3, (:mu, k+1), new_mu, :continuous)
-     @write(tr3, (:var, k+1), new_var, :continuous)
+     @write(tr3[:k], k+1, :discrete)
+     new_mu = @read(tr2[:new_mu], :continuous)
+     new_var = @read(tr2[:new_var], :continuous)
+     @write(tr3[(:mu, k+1)], new_mu, :continuous)
+     @write(tr3[(:var, k+1)], new_var, :continuous)
      new_weights = add_weight(weights)
-     @write(tr4, :deletion_idx, k+1, :discrete)
+     @write(tr4[:deletion_idx], k+1, :discrete)
    else
-     idx = @read(tr2, :deletion_idx, :discrete)
-     @copy(tr1, (:mu, idx), tr4, :new_mu)
-     @copy(tr1, (:var, idx), tr4, :new_var)
+     idx = @read(tr2[:deletion_idx], :discrete)
+     @copy(tr1[(:mu, idx)], tr4[:new_mu])
+     @copy(tr1[(:var, idx)], tr4[:new_var])
      for i in (idx+1):k
-       @copy(tr1, (:mu, i), tr3, (:mu, i-1))
-       @copy(tr1, (:var, i), tr3, (:var, i-1))
+       @copy(tr1[(:mu, i)], tr3[(:mu, i-1)])
+       @copy(tr1[(:var, i)], tr3[(:var, i-1)])
      end
-     @write(tr3, :k, k-1, :discrete)
+     @write(tr3[:k], k-1, :discrete)
      new_weights = delete_weight(weights, idx)
    end
-   @write(tr3, :weights, new_weights, :continuous)
+   @write(tr3[:weights], new_weights, :continuous)
 end
@@
 {\begin{lstlisting}[basicstyle=\small\ttfamily]
-@copy((aux, :path), (aux, :path))
+@copy(aux_in[:path], aux_out[:path])
 \end{lstlisting}}
@@
 {\begin{lstlisting}[basicstyle=\small\ttfamily]
-@copy((model, subtree_address), (model, :new_subtree))
-@copy((aux, :new_subtree), (aux, subtree_address))
+@copy(model_in[subtree_address], aux_out[:new_subtree])
+@copy(aux_in[:new_subtree], model_out[subtree_address])
 \end{lstlisting}}
@@
     \centering
-    \includegraphics[width=0.8\textwidth]{figures/structure-learning.pdf}
+    \includegraphics[width=0.95\textwidth]{figures/structure-learning.pdf}
     \caption{
@@
 (2) a probabilistic program $\mathtt{q}$ encoding an auxiliary probability distribution (shown in b), and
-(3) a differentiable program $\mathtt{h}$ that encodes an involution (shown in d).
+(3) a differentiable program $\mathtt{f}$ that encodes an involution (shown in d).
 }

### 2020-07-19 paper.tex
fix cubic

@@
 \item An algorithm that automates the implementation of an involutive MCMC kernel given two probabilistic programs and a differentiable program encoding the involution, using automatic differentiation and tracing of probabilistic programs.
-\item An extension to the algorithm that exploits sparsity in the involution to reduce the computational complexity of an involutive MCMC kernel from $O(N^2)$ to $O(1)$ in common cases, where $N$ is the dimensionality of the latent space.
+\item An extension to the algorithm that exploits sparsity in the involution to reduce the computational complexity of an involutive MCMC kernel from $O(N^3)$ to $O(1)$ in common cases, where $N$ is the dimensionality of the latent space.
 \item An algorithm that dynamically detects errors in the specification of an involutive MCMC kernel.
@@
 Suppose $N$ and $N'$ are the total number of random choices in the input traces $x \oplus y$ and output traces $x' \oplus y'$ respectively, and $M$ is the number of continuous random choices (which is the same in the input traces and output traces).
-Then, the number of operations requried in Algorithm~\ref{alg:auto-involutive-mcmc} grows as $O(N + N') + O(M^2)$ (where each call to \textproc{logpdf} and the computation of each entry in the Jacobian both count as one operation).
+Then, the number of operations used in Algorithm~\ref{alg:auto-involutive-mcmc} grows as $O(N + N') + O(M^3)$ (where each call to \textproc{logpdf} and the computation of each entry in the Jacobian both count as one operation).
 The linear term represents the complexity of sampling $y'$ and computing the four log-densities used to compute the acceptance probability.
-The quadratic term represents the complexity of computing the Jacobian matrix and its determinant.
+The cubic term represents the complexity of computing the Jacobian matrix and its determinant.
 
@@
 By applying this rule recursively, we can instead compute the determinant of a much smaller matrix; for the example above with $M=4$, the absolute value of the determinant simplifies to the absolute value of a single entry ($|-1|$).
-Indeed, if some input variable is copied to some output variable, then we can entirely avoid computing its row (and corresponding column) of the Jacobian, and the computational complexity reduces from $M^2$ to $(M-K)^2$ where $K$ is the number of input variables that were copied from some input variable.
+Indeed, if some input variable is copied to some output variable, then we can entirely avoid computing its row (and corresponding column) of the Jacobian, and the computational complexity reduces from $M^3$ to $(M-K)^3$ where $K$ is the number of input variables that were copied from some input variable.
 The $\mathtt{@copy}$ statement in our differentiable programming language makes the set of input variables that were copied immediately available to the interpreter, which makes automating this optimization possible, as shown in \textproc{run-involution} in Algorithm~\ref{alg:auto-involutive-mcmc}.
@@
 In this case, the size $N$ of the trace grows linearly both in the number of clusters and the number of data points, but the Jacobian determinant can be calculated from only the $6$-by-$6$ submatrix of the Jacobian that involves the parameters of the one cluster being split (or the two clusters being merged).
-Therefore, the computation complexity of the acceptance ratio computation reduces from $O(N + N') + O(M^2)$ to $O(N + N')$.
+Therefore, the computation complexity of the acceptance ratio computation reduces from $O(N + N') + O(M^3)$ to $O(N + N')$.
 While a reasonable hand-coded implementation of this algorithm would likely perform this sort of optimization as well, Algorithm~\ref{alg:auto-involutive-mcmc} automates it.

### 2020-07-19 paper.tex
checkpoint on section 5 pass

@@
 The involutive MCMC construction encompasses many existing classes of MCMC kernels, some of which explicitly make use of bijective or involutive deterministic maps.
-In particular, the reversible jump framework~\citep{green1995reversible,hastie2012model} employs a family of continuously differentiable bijections between the Euclidean parameter spaces of different models.
-\citet{tierney1998note} described a family of deterministic proposals based on a deterministic involution that is equivalent to involutive MCMC but without the auxiliary proposal distribution.
+In particular, the reversible jump framework~\citep{green1995reversible,hastie2012model} employs a family of continuously differentiable bijections between the parameter spaces of different models.
+\citet{tierney1998note} described a family of deterministic proposals based on a deterministic involution that is equivalent to involutive MCMC but without the auxiliary probability distribution.
 More recently,~\citet{spanbauer2020deep} defined a class of deep generative model based on differentiable involutions and trained these models to serve as efficient proposal distributions on continuous state spaces; the resulting algorithm is an instance of the construction presented in this paper.
 
-In recent decades, a number of probabilistic programming systems have provided automated implementations of probabilistic inference algorithms~\citep{gilks1994language,milch2007,pfeffer200714,goodman2012church,gehr2016psi,carpenter2017stan}.
-The majority of these systems have focused on the automation of one or more built-in inference algorithms, with customization limited to tuning a handful of parameters of the algorithm.
-The programmable inference paradigm~\citep{mansinghka2018probabilistic} in probabilistic programming instead emphasizes user control over the inference algorithm and automation of the lower-level implementation.
-In particular, the Gen probabilistic programming system~\citep{cusumano2019gen} offers a high degree of user programmability by exposing an API that supports high-level user implementations of inference algorithms, instead of a set of a built-in inference engine.
+In recent decades, a number of probabilistic programming systems have automated the implementations of probabilistic inference algorithms~\citep{gilks1994language,milch2007,pfeffer200714,goodman2012church,gehr2016psi,carpenter2017stan}.
+Most of these systems support generic built-in inference algorithms, with user-customization limited to tweaking algorithm parameters.
+Some systems allow for user-defined variational families or proposal distributions~\citep{ritchie2016deep,bingham2019pyro}.
+Programmable inference~\citep{mansinghka2018probabilistic} proposes that inference algorithms be specified by users in high-level programming languages.
+The Gen probabilistic programming system~\citep{cusumano2019gen} exposes an API that supports high-level user implementations of an open-ended set of inference algorithms, and abstracts away low-level implementation details of inference algorithms.
+
 One other probabilistic programming system besides Gen supports custom reversible jump samplers:
-\citet{roberts2019} present a system embedded in Haskell that can automatically generate the implementation of certain reversible jump MCMC kernels from a high-level specification, and \citet{narayanan2020symbolic} also give a technique that can be used to automatically compute the Metropolis-Hastings acceptance probabilities in some settings; however, these approaches do not handle many kernels that can be handled by our technique, including the example in Figure~\ref{fig:structure-learning}.
+\citet{roberts2019} present a system embedded in Haskell that automatically generates the implementation of some reversible jump MCMC kernels from a high-level specification.
+\citet{narayanan2020symbolic} give a technique that automatically computes Metropolis-Hastings acceptance probabilities in some settings; however, these approaches do not handle many kernels that can be handled by our technique, including the example in Figure~\ref{fig:structure-learning}.
 
@@
 \textbf{Relationship to representation of~\citet{green1995reversible}.}
-\citet{green1995reversible} uses a state space that is the countable union of `models', where for each model there is a vector of real-valued parameters.
+\citet{green1995reversible} uses a state space that is the countable union of `models', where each model uses a vector of real-valued parameters.
 Dictionaries possess substantially more structure:
@@
 Suppose $N$ and $N'$ are the total number of random choices in the input traces $x \oplus y$ and output traces $x' \oplus y'$ respectively, and $M$ is the number of continuous random choices (which is the same in the input traces and output traces).
-Then, the number of operations used in Algorithm~\ref{alg:auto-involutive-mcmc} grows as $O(N + N') + O(M^3)$ (where each call to \textproc{logpdf} and the computation of each entry in the Jacobian both count as one operation).
-The linear term represents the complexity of sampling $y'$ and computing the four log-densities used to compute the acceptance probability.
-The cubic term represents the complexity of computing the Jacobian matrix and its determinant.
-
-It is possible to reduce the computational complexity of an automated involutive MCMC kernel by exploiting special structure in the involution $f$, in some cases to $O(1)$.
-This section describes how an automated involutive MCMC implementation can reduce the computational complexity of the Jacobian computation and the density ratio computation.
-These techniques are used in the Gen implementation of automated involutive MCMC, and one of the techniques is used in our minimal PyTorch implementation.
+Then, the number of operations used in Algorithm~\ref{alg:auto-involutive-mcmc} grows as $O(N + N') + O(M^3)$.
+The linear term is due to sampling $y'$ and computing the four log-densities required for the acceptance probability.
+The cubic term is due to computing the Jacobian determinant, which is also required for the acceptance probability (Equation~\ref{eq:acceptance-ratio-dictionaries}).
+
+It is possible to reduce the number of operations performed in an automated involutive MCMC kernel by exploiting special structure in the involution $f$.
+In some cases, this structure can lead to $O(1)$ operations per kernel application (i.e. constant in $N$, $N'$, and $M$).
+This section describes techniques for exploiting involution structure within an automated involutive MCMC implementation.
+These techniques are used in the Gen implementation, and one of the techniques is used in our minimal PyTorch implementation.
 
@@
 The naive implementation of Algorithm~\ref{alg:auto-involutive-mcmc} computes by the Jacobian by first computing the $M$-by-$M$ Jacobian matrix $J h_{e(z)}$ via automatic differentiation, and then computing the absolute value of its determinant.
-However, we observe that in many applications of involutive MCMC, the values at continuous choices in the input traces are directly copied into the output traces (either at the same, or different, keys).
+However, we observe that in many applications of involutive MCMC, the values at continuous choices in the input traces are directly copied into the output traces (either at the same key or a different key).
 These copy operations result in columns in the Jacobian matrix that have a single $1$ entry with remaining entries equal to $0$.
@@
 \end{equation}
-Using the cofactor expansion of the determinant, we observe that for any (`copy') row or column in an $M$-by-$M$ Jacobian matrix with a single $1$ and all other entries $0$, the absolute value of the determinant is equivalent to that of the $(M-1)$-by-$(M-1)$ sub-matrix with the corresponding row \textit{and} column omitted (even if that would remove other nonzero entries from the matrix).
+Using the cofactor expansion of the determinant, we observe that for any (`copy') column in an $M$-by-$M$ Jacobian matrix (a column with a single $1$ and all other entries $0$), the absolute value of the determinant is equivalent to that of the $(M-1)$-by-$(M-1)$ sub-matrix with the corresponding row \textit{and} column omitted (even if that would remove other nonzero entries from the matrix).
 By applying this rule recursively, we can instead compute the determinant of a much smaller matrix; for the example above with $M=4$, the absolute value of the determinant simplifies to the absolute value of a single entry ($|-1|$).
-Indeed, if some input variable is copied to some output variable, then we can entirely avoid computing its row (and corresponding column) of the Jacobian, and the computational complexity reduces from $M^3$ to $(M-K)^3$ where $K$ is the number of input variables that were copied from some input variable.
-The $\mathtt{@copy}$ statement in our differentiable programming language makes the set of input variables that were copied immediately available to the interpreter, which makes automating this optimization possible, as shown in \textproc{run-involution} in Algorithm~\ref{alg:auto-involutive-mcmc}.
-
-Many involutive MCMC kernels update only a constant number of variables that does not depend on the sizes $N$ and $N'$ of the input and output traces or the total number of continous keys $M$; the other variables are copied over unchanged.
+Indeed, if some input key is copied to some output key, then we can entirely avoid computing its row (and corresponding column) of the Jacobian.
+Tsome herefore, the number of operations (which is dominated by the determinant) reduces from $M^3$ to $(M-|C|)^3$ where $|C|$ is the number of input keys that were copied to some output input key.
+The $\mathtt{@copy}$ statement in our differentiable programming language makes the set of input keys that were copied explicitly available to the interpreter, which makes automating this optimization straightforward, as shown in \textproc{run-involution} in Algorithm~\ref{alg:auto-involutive-mcmc}.
+
+Many involutive MCMC kernels update only a constant number of keys that does not depend on the sizes $N$ and $N'$ of the input and output traces or the total number of continous keys $M$; the other keys are copied over unchanged.
 For example, consider the Jacobian for a split-merge reversible jump move~\citet{richardson1997bayesian}, which is implemented in Figure~\ref{fig:mixture}:
@@
 In this case, the size $N$ of the trace grows linearly both in the number of clusters and the number of data points, but the Jacobian determinant can be calculated from only the $6$-by-$6$ submatrix of the Jacobian that involves the parameters of the one cluster being split (or the two clusters being merged).
-Therefore, the computation complexity of the acceptance ratio computation reduces from $O(N + N') + O(M^3)$ to $O(N + N')$.
+Therefore, the acceptance ratio computation reduces from $O(N + N') + O(M^3)$ to $O(N + N')$.
 While a reasonable hand-coded implementation of this algorithm would likely perform this sort of optimization as well, Algorithm~\ref{alg:auto-involutive-mcmc} automates it.
@@
 
-It is possible to further reduce the the number of operations in automated involutive MCMC from $O(N + N')$ to $O(N_2 + N_2')$ where $N = N_1 + N_2$ and $N' = N_1' + N_2'$ and $N_1$ and $N_2$ are the sizes of the model trace and auxiliary trace respectively (and similarly for $N_1'$ and $N_2'$).
-In cases where $N_2$ and $N_2'$ have fixed size even as $N_1$ and $N_1'$ grow, this approach can result in $O(1)$ operations per kernel application.
-The efficiency gains are due to a combination of exploiting sparsity in the involution and conditional independencies in the model.
-
-Consider input traces $(x, y)$ and output traces $(x', y')$ where $x$ and $x'$ have sizes $N_1$ and $N_2$ respectively.
-The approach in Algorithm~\ref{alg:auto-involutive-mcmc} involves running a program $\mathcal{F}$ that explicitly writes the value for each element of $(x', y')$, which requires greater than $N_1'$ `write' or `copy' operations.
-We now show how to specify the involution $f$ such that computing $z'$ from $z$ can take fewer operations for involutions that possess sparse structure.
-Recall that $K_{x}$ and $K_{x'}$ are the sets of keys in $x$ and $x'$, respectively, and that $x$ and $x'$ may both have values for certain keys, and $x$ may contains keys not present in $x'$, and vice-versa.
+Consider input traces $(x, y)$ and output traces $(x', y')$ with $f(x, y) = (x', y')$ for involution $f$ encoded by program $\mathcal{F}$.
+Let $N_1 := |K_{x}|$, $N_2 := |K_{y}|$, $N_1' := |K_{x'}|$ and $N_2' := |K_{y'}|$, and
+$N := N_1 + N_2$ and $N' := N_1' + N_2'$.
+Suppose that $N_1 + N_1' \gg N_2 + N_2'$, which occurs when the size of the model traces grows in the size of a data set, but the size of the auxiliary trace is constant.
+Algorithm~\ref{alg:auto-involutive-mcmc} runs $\mathcal{F}$, which explicitly writes the value for each element of $(x', y')$, requiring $N' = N_1' + N_2' \approx N_1'$ `write' or `copy' operations.
+The algorithm also uses approximately $N_1 + N_1'$ operations to evaluate the log-densities, because it accumulates the log-density of each random choice in $x$, and $x'$.
+
+We now show how to modify the program $\mathcal{F}$ and its interpreter, so that both computing $z' = (x', y')$ from $z = (x, y)$, and computing the log density ratio $\log(p(x') / p(x))$, can take $O(1)$ operations (that is, constant in $N_1 + N_1'$) whn the involution $f$ has sparse structure and the model density $p$ has conditional independencies.
+
+Note that $x$ and $x'$ may both have values for certain keys, and $x$ may contains keys not present in $x'$, and vice-versa.
 Often, an involution does not modify the values of many keys in the trace---this is the case in the split-merge reversible jump move described earlier.
@@
 Let $\Delta := K_{x'} \setminus A$ be the set of keys in $x'$ that are either (i) not in $x$, or (ii) in $x$ but have their value changed.
-Then, the previous trace $x$ and a dictionary $\delta$ with $K_{\delta} = \Delta$ and $\delta[k] = x'[k]$ and the density function $p$ suffice to uniquely define the trace $x'$, provided the density satisfies the following condition:
+Then, the previous trace $x$ and a dictionary $\delta$ with $K_{\delta} = \Delta$ and $\delta[k] = x'[k]$ and the density function $p$ suffice to uniquely define the trace $x'$, provided the density satisfies the following condition, which is satisfied by densities represented as probabilistic programs:
 $p(x) > 0$ and $p(x') > 0$ and $x \ne x'$ implies there exists $k \in K_x \cap K_{x'}$ with $x[k] \ne x'[k]$.
 Let $x|_A$ denote the \emph{restriction} of a dictionary $x$ to only keys in $A$.
-Then, \textproc{naive-trace-update} (Algorithm~\ref{alg:naive-update}) computes the new trace $x'$ and difference in log-densities $\log p(x') - \log p(x)$ of a probabilistic program from a previous trace $x$ and a dictionary $\delta$, provided there exists $x' = \delta \oplus x|_A$ for some $A$ where $p(x') > 0$.
+Then, \textproc{naive-trace-update} (Algorithm~\ref{alg:naive-update}) computes the new trace $x'$ and log densities ratio $\log (p(x') / p(x))$ of a probabilistic program from a previous trace $x$ and a dictionary $\delta$, provided there exists $x' = \delta \oplus x|_A$ for some $A$ where $p(x') > 0$.
 \begin{algorithm}[h]
@@
 \end{algorithm}
+
 \textproc{naive-trace-update} is a naive implementation of a general \emph{trace-update} operation that is part of Gen's API.
@@
 
-To represent reversible jump MCMC using involutive MCMC implemented with probabilistic programs, we write a probabilistic program $\mathcal{P}$ that encodes the space of models $\mathcal{H}$, the prior distribution on models, $p(h)$, the per-model priors $p_h(\theta_h)$ and the per-model likelihoods $L_{D,h}(\theta_h)$.
+To represent reversible jump MCMC involutive MCMC implemented with probabilistic programs, we write a probabilistic program $\mathcal{P}$ that encodes the space of models $\mathcal{H}$, the prior distribution on models, $p(h)$, the per-model priors $p_h(\theta_h)$ and the per-model likelihoods $L_{D,h}(\theta_h)$.
 The set of all models $h$ corresponds to the set of all pairs $(K_\mathcal{P}, \mathbf{d}_\mathcal{P})$ where $K_\mathcal{P}$ represent possible trace structures (i.e.
@@
 
-Involutive MCMC is a unifying construction for MCMC algorithms that encompasses both classic approaches to constructing kernels like reversible jump MCMC~\citep{green1995reversible}, but also recently introduced classes of MCMC kernels based on neural networks~\citep{spanbauer2020deep}.
+Involutive MCMC is a unifying construction~\citep{imcmc} for MCMC algorithms that encompasses both classic approaches to constructing kernels like reversible jump MCMC~\citep{green1995reversible}, but also recently introduced classes of MCMC kernels based on neural networks~\citep{spanbauer2020deep}.
 Therefore, the approach to automating involutive MCMC kernels presented in this paper makes a number of classic MCMC techniques easier to use and broadens their accessibility, and may potentially aid in development of novel MCMC techniques as well.
@@
 
-\subsubsection*{Acknowledgements}
-All acknowledgments go at the end of the paper, including thanks to reviewers who gave useful comments, to colleagues who contributed to the ideas, and to funding agencies and corporate sponsors that provided financial support.
+%\subsubsection*{Acknowledgements}
 

### 2020-07-19 paper.tex
pass over section 5

@@
 Indeed, if some input key is copied to some output key, then we can entirely avoid computing its row (and corresponding column) of the Jacobian.
-Tsome herefore, the number of operations (which is dominated by the determinant) reduces from $M^3$ to $(M-|C|)^3$ where $|C|$ is the number of input keys that were copied to some output input key.
+Therefore, the number of operations (which is dominated by the determinant) reduces from $M^3$ to $(M-|C|)^3$ where $|C|$ is the number of input keys that were copied to some output input key.
 The $\mathtt{@copy}$ statement in our differentiable programming language makes the set of input keys that were copied explicitly available to the interpreter, which makes automating this optimization straightforward, as shown in \textproc{run-involution} in Algorithm~\ref{alg:auto-involutive-mcmc}.
@@
 
-We now show how to modify the program $\mathcal{F}$ and its interpreter, so that both computing $z' = (x', y')$ from $z = (x, y)$, and computing the log density ratio $\log(p(x') / p(x))$, can take $O(1)$ operations (that is, constant in $N_1 + N_1'$) whn the involution $f$ has sparse structure and the model density $p$ has conditional independencies.
+We now show how to modify the program $\mathcal{F}$ and its interpreter, so that both computing $z' = (x', y')$ from $z = (x, y)$, and computing the log density ratio $\log(p(x') / p(x))$, can take $O(1)$ operations (that is, constant in $N_1 + N_1'$) when the involution $f$ has sparse structure and the model density $p$ has conditional independencies.
 
@@
 \textproc{naive-trace-update} is a naive implementation of a general \emph{trace-update} operation that is part of Gen's API.
-While the naive implementation requires $O(|K_x| + |K_{x'}|)$ operations, it is possible for more sophisticated implementations of \emph{trace-update} to run in $O(|K_{\delta}|)$ operations for certain probabilistic programs $\mathcal{P}$ by exploiting conditional independence structure program that leads to cancellations in the density ratio $p(x') / p(x)$.
+While the naive implementation requires $O(N_1 + N_1')$ operations, it is possible for more sophisticated implementations of \emph{trace-update} to run in approximately $O(|K_{\delta}|)$ operations for certain probabilistic programs $\mathcal{P}$ by exploiting conditional independence structure program that leads to cancellations in the density ratio $p(x') / p(x)$.
 The details of these more efficient implementations are outside the scope of this paper, but are implemented as part of the Gen system.
+(As a simple example, consider a density $p$ that factors into a product of independent densities, one for each random choice; and an involution that only modifies one of these choices).
 
@@
 modifying the requirement for the involution program $\mathcal{F}$:
-Instead of explicitly specifying all of $z'$, it only needs to specify a subset of the trace $\delta$.
-Therefore, the involution program can run in $O(|K_{\delta}| + |K_{y}| + |K_{y'}|)$ operations.
-We also remove the two calls to \textproc{score} on $x$ and $x'$, and replace them with a single call to \emph{trace-update} on $\mathcal{P}$ that takes in $x$, and $\delta$ and returns $x'$ and $\log p(x') - \log p(x)$.
-When $|K_{\delta}|$ and $|K_{y}|$ and $|K_{y}|$ are all constant in the size $N_1$ of the model trace $x$, the resulting algorithm scales as $O(1)$ instead of $O(N + N')$.
+Instead of explicitly specifying a value for all keys in $x'$, it can specify $x'$ by only explicitly specifying values in $\delta$.
+Therefore, the involution program can run in $O(|K_{\delta}| + N_2 + N_2')$ operations.
+We replace the separate calls to \textproc{score} for the log-densities $\log p(x)$ and $\log p(x')$ and replace them with a single call to \emph{trace-update} on $\mathcal{P}$ that takes in $x$, and $\delta$ and returns $x'$ and $\log p(x') - \log p(x)$.
+When $|K_{\delta}|$ and $N_2$ and $N_2'$ are all constant in the size $N_1$ of the model trace $x$, the resulting algorithm uses a number of operations that is constant in $O(N_1 + N_1')$.
 

### 2020-07-19 paper.tex
be more explicit about observations

@@
 \begin{algorithmic}
-\Require probabilistic programs $\mathcal{P}$ and $\mathcal{Q}$; differentiable program $\mathcal{F}$; initial state $x$
+\Require probabilistic programs $\mathcal{P}$ and $\mathcal{Q}$; differentiable program $\mathcal{F}$; initial state $x$; observations $b$
 \vspace{1mm}
@@
     \State $(x' \oplus y', \log D) \gets \textproc{run-involution}(\mathcal{F}, x \oplus y)$ \Comment{{\small Compute $(x' \oplus y') = f(x \oplus y)$ and log(Radon-Nikodym derivative)}}
-    \State $\log p(x) \gets \textproc{score}(\mathcal{P}, x)$ \Comment{Compute log-density via probabilistic program}
-    \State $\log p(x') \gets \textproc{score}(\mathcal{P}, x')$ \Comment{Compute log-density via probabilistic program}
+    \State $\log \tilde{p}(x \oplus b) \gets \textproc{score}(\mathcal{P}, x \oplus b)$ \Comment{Compute log-density via probabilistic program}
+    \State $\log \tilde{p}(x' \oplus b) \gets \textproc{score}(\mathcal{P}, x' \oplus b)$ \Comment{Compute log-density via probabilistic program}
     \State $\log q_{x'}(y') \gets \textproc{score}(\mathcal{Q}_{x'}, y')$ \Comment{Compute log-density via probabilistic program}
-    \State $\alpha \gets \min\{1, \exp(\log p(x') - \log p(x) + \log q_{x'}(y') - \log q_x(y) + \log D)\}$ \Comment{Compute acceptance probability}
+    \State $\alpha \gets \min\{1, \exp(\log \tilde{p}(x' \oplus b) - \log \tilde{p}(x \oplus b) + \log q_{x'}(y') - \log q_x(y) + \log D)\}$ \Comment{Acceptance probability}
     \State \textbf{with probability} \,$\alpha$ \,\Return $x'$ \algorithmicelse \, \Return $x$  \algorithmicend
@@
         \State $k_i \gets W_i$ \Comment{Pick $i$th key in $W$; the order does not matter}
-        \State $J[i,:] \gets \nabla_{\mathbf{z}_{R \setminus C}} \left( h_{e(z)}[a_i] \right)$ \Comment{{\small Gradient of $z'[k_i] = h_{e(z)}[k_i]$ w.r.t.
+        \State $J[:,i] \gets \nabla_{\mathbf{z}_{R \setminus C}} \left( h_{e(z)}[a_i] \right)$ \Comment{{\small Gradient of $z'[k_i] = h_{e(z)}[k_i]$ w.r.t.
 non-copied continuous inputs ($z_k$ for $k \in R \setminus C$)}}
@@
 
+Note that in general there are observations $b$ that induce the target density $p(x)$ via $p(x) \propto \tilde{p}(x \oplus b)$ as described in Section~\ref{sec:dists-on-dicts}, and the involution $f$ in general depends on $b$.
+The involution program $\mathcal{F}$ can read the values of the $b$, which are stored alongside the unobserved random choices in the input model trace.
+
 \subsubsection{Computing the Jacobian with Automatic Differentiation}
@@
 The interpreter executes $\mathcal{F}$ using regular Julia or Python semantics, but intercepts calls to \texttt{@write}, \texttt{@read}, and \texttt{@copy} statements, and in addition to performing the desired operation, records the set of continuous addresses that are read, written, and copied.
-After $\mathcal{F}$ is finished executing, AD uses the recorded addresses to compute the Jacobian $|J h_e|(z)$; in the case of reverse-mode AD, this is accomplished by iterating over output continuous addresses and backpropagating from each one to all of the input continuous addresses, computing the Jacobian row-by-row.
-Note that the $\mathtt{discrete}$ and $\mathtt{continuous}$ labels are omitted from the syntax in \textproc{run-involution} to simplify notation ($a \in \mathcal{I}$ indicates a discrete choice and otherwise a choice is continuous).
+After $\mathcal{F}$ is finished executing, AD uses the recorded addresses to compute the Jacobian $|J h_e|(z)$; in the case of reverse-mode AD, this is accomplished by iterating over output continuous addresses and backpropagating from each one to all of the input continuous addresses, computing the Jacobian column-by-column.
+Keys with an $n$-dimensional Lebesgue measure as their reference measure, corresponding to vector-valued random choices, are unpacked into $n$ separate columns in the Jacobian; this detail is elided in Algorithm~\ref{alg:auto-involutive-mcmc}.
+The $\mathtt{discrete}$ and $\mathtt{continuous}$ labels are also omitted from the syntax in \textproc{run-involution} to simplify notation ($a \in \mathcal{I}$ indicates a discrete choice and otherwise a choice is continuous).
 
@@
 
-We now show how to modify the program $\mathcal{F}$ and its interpreter, so that both computing $z' = (x', y')$ from $z = (x, y)$, and computing the log density ratio $\log(p(x') / p(x))$, can take $O(1)$ operations (that is, constant in $N_1 + N_1'$) when the involution $f$ has sparse structure and the model density $p$ has conditional independencies.
+We now show how to modify the program $\mathcal{F}$ and its interpreter, so that both computing $z' = (x', y')$ from $z = (x, y)$, and computing the log density ratio $\log(\tilde{p}(x' \oplus b) / \tilde{p}(x \oplus b))$, can take $O(1)$ operations (that is, constant in $N_1 + N_1'$) when the involution $f$ has sparse structure and the model density $p$ has conditional independencies.
 
@@
 Let $\Delta := K_{x'} \setminus A$ be the set of keys in $x'$ that are either (i) not in $x$, or (ii) in $x$ but have their value changed.
-Then, the previous trace $x$ and a dictionary $\delta$ with $K_{\delta} = \Delta$ and $\delta[k] = x'[k]$ and the density function $p$ suffice to uniquely define the trace $x'$, provided the density satisfies the following condition, which is satisfied by densities represented as probabilistic programs:
+Then, the previous trace $x$ and a dictionary $\delta$ with $K_{\delta} = \Delta$ and $\delta[k] = x'[k]$ and the density function $p$ suffice to uniquely define the trace $x'$, provided the density satisfies the following condition, which is satisfied by densities as probabilistic programs:
 $p(x) > 0$ and $p(x') > 0$ and $x \ne x'$ implies there exists $k \in K_x \cap K_{x'}$ with $x[k] \ne x'[k]$.
 Let $x|_A$ denote the \emph{restriction} of a dictionary $x$ to only keys in $A$.
-Then, \textproc{naive-trace-update} (Algorithm~\ref{alg:naive-update}) computes the new trace $x'$ and log densities ratio $\log (p(x') / p(x))$ of a probabilistic program from a previous trace $x$ and a dictionary $\delta$, provided there exists $x' = \delta \oplus x|_A$ for some $A$ where $p(x') > 0$.
+Then, \textproc{naive-trace-update} (Algorithm~\ref{alg:naive-update}) computes the new trace $x'$ and log-densities ratio $\log (p(x') / p(x))$ of a probabilistic program from a previous trace $x$ and a dictionary $\delta$ (and observations $b$), provided there exists $x' = \delta \oplus x|_A$ for some $A$ where $p(x') > 0$.
 \begin{algorithm}[h]
 \begin{algorithmic}
-\Require probabilistic program $\mathcal{P}$; previous trace $x$;\\
+\Require probabilistic program $\mathcal{P}$; previous trace $x$; observations $b$;\\
 dictionary $\delta$ such that there exists $x' = (\delta \oplus x|_{A})$ for some $A$ where $p(x') > 0$.
 \vspace{1mm}
-\Procedure{naive-trace-update}{$\mathcal{P}$, $x$, $\delta$}
+\Procedure{naive-trace-update}{$\mathcal{P}$, $x$, $b$, $\delta$}
+   \State {\bf assert} $K_{x} \cap K_{b} = \varnothing$
    \State $x' \gets \{\}$
@@
    \execute[$\mathcal{P}$], but with $\mathtt{"}k \sim \mathrm{distribution}\mathtt{"} \equiv$ (
-        \State 1. if $k \in K_{\delta}$ then $v \gets \delta[k]$ else $v \gets x[k]$
-        \State 2. set $x'[k] \gets v$
+        \State 1. if $k \in K_{\delta}$ then $v \gets \delta[k]$ else $v \gets (x \oplus b)[k]$
+        \State 2. if $k \not \in K_{b}$ then set $x'[k] \gets v$
         \State 3. set $s' \gets s' + \textproc{logpdf}(\mathrm{distribution}, v)$
@@
     \END
-   \State $s \gets \textproc{score}(\mathcal{P}, x)$
+   \State $s \gets \textproc{score}(\mathcal{P}, x \oplus b)$
    \State \Return $(x', s' - s)$
@@
 \textproc{naive-trace-update} is a naive implementation of a general \emph{trace-update} operation that is part of Gen's API.
-While the naive implementation requires $O(N_1 + N_1')$ operations, it is possible for more sophisticated implementations of \emph{trace-update} to run in approximately $O(|K_{\delta}|)$ operations for certain probabilistic programs $\mathcal{P}$ by exploiting conditional independence structure program that leads to cancellations in the density ratio $p(x') / p(x)$.
+While the naive implementation requires $O(N_1 + N_1')$ operations, it is possible for more sophisticated implementations of \emph{trace-update} to run in approximately $O(|K_{\delta}|)$ operations for certain probabilistic programs $\mathcal{P}$ by exploiting conditional independence structure program that leads to cancellations in the density ratio $\tilde{p}(x' \oplus) / \tilde{p}(x \oplus b)$.
 The details of these more efficient implementations are outside the scope of this paper, but are implemented as part of the Gen system.
@@
 \begin{algorithmic}
-\Procedure{involutive-mcmc-check}{$p$, $q$, $f$}
-    \State $x \sim p(\cdot)$ \Comment{Sample state from prior}
-    \State $y \sim q_x(\cdot)$ \Comment{Sample auxiliary variables}
-    \State $(x', y') \gets f(x, y)$ \Comment{Apply involution}
-    \State {\bf assert} $p(x')q_{x'}(y') > 0$ \Comment{Support check}
-    \State {\bf assert} $f(x', y') = (x, y)$ \Comment{Involution check}
-    \State {\bf assert} $|\text{Rd}^{\textit{cont}}_{f,x,y}| = |\text{Wr}^{\textit{cont}}_{f,x,y}|$ \Comment{Dimension check}
+\Procedure{involutive-mcmc-check}{$\mathcal{P}$, $\mathcal{Q}$, $\mathcal{F}$}
+    \LineComment{Generate a randomized test case}
+    \State $(x \oplus b) \sim \textproc{trace-and-score}(\mathcal{P})$ % TODO not conditioned on data?
+    \State $y \sim \textproc{trace-and-score}(\mathcal{Q}_x)$
+    %\State $x \sim p(\cdot)$ \Comment{Sample state from prior}
+    %\State $y \sim q_x(\cdot)$ \Comment{Sample auxiliary variables}
+
+    \LineComment{Run involution and perform dimension check}
+    \State $x' \oplus y' \gets \textproc{run-involution}(\mathcal{F}, x \oplus y)$
+
+    \LineComment{Support check}
+    \State $\log p(x') \gets \textproc{score}(\mathcal{P}, x')$
+    \State $\log q_{x'}(y') \gets \textproc{score}(\mathcal{Q}_{x'}, y')$
+    \State {\bf assert} $(\log p(x') > -\infty) \land (\log q_{x'}(y') > -\infty)$
+
+    \LineComment{Involution check}
+    \State $\tilde{x} \oplus \tilde{y} \gets \textproc{run-involution}(\mathcal{F}, x' \oplus y')$
+    \State ${\bf assert} (\tilde{x} = x) \land (\tilde{y} = y)$
+    %\State $(x', y') \gets f(x, y)$ \Comment{Apply involution}
+    %\State {\bf assert} $f(x', y') = (x, y)$ \Comment{Involution check}
+    %\State {\bf assert} $p(x')q_{x'}(y') > 0$ \Comment{Support check}
+    %\State {\bf assert} $f(x', y') = (x, y)$ \Comment{Involution check}
+    %\State {\bf assert} $|R| = |W|$ \Comment{Dimension check}
 \EndProcedure
@@
 
-To test the user's algorithm, Gen first samples a model trace $x$ from the model prior $p$, and auxiliary trace $y$ from the $q_x$. It then runs three tests:
+To test the user's algorithm, Gen first samples a model trace $x$ and simulated observations $b$ from the model prior $p$, and auxiliary trace $y$ from the $q_x$. It then runs three tests:
 
@@
 
+\item \textbf{Dimension check.} The \textit{dimension check} runs the involution $f$ on $(x, y)$ and records the sets of addresses $R$ and $W$ at which $f$ performs continuous reads and writes, respectively.
+It then checks that the sizes of these sets match.
+This check is run when computing the determinant of the Jacobian in \textproc{run-involution}.
+
 \item \textbf{Involution check.} The \textit{involution check} runs $f \circ f$ on $(x, y)$ and checks that the result is equal to $(x, y)$.
 If $f$ is an involution on $Z$, then this check will always succeed; if there is a set $\tilde{Z}$ of positive $\pi$-measure on which $f(f(z)) \neq z$ (i.e., $f$ is not an involution), then this check has positive probability of failing.
-
-\item \textbf{Dimension check.} The \textit{dimension check} runs the involution $f$ on $(x, y)$ and records the sets of addresses $\text{Rd}_{f,x,y}^{\textit{cont}}$ and $\text{Wr}_{f,x,y}^{\textit{cont}}$ at which $f$ performs continuous reads and writes, respectively.
-It then checks that the sizes of these sets match.
-(Addresses with an $n$-dimensional Lebesgue measure as their reference measure, corresponding to vector-valued random choices, are unpacked into $n$ separate addresses for the purposes of this check.)
 \end{enumerate}

### 2020-07-19 paper.tex
a quick pass through the proofs section; cleaning up dead code

@@
 A proposal is made from state $x = (h, \theta_h)$ by (i) sampling a move type $m \sim q_{x}(\cdot)$, and (ii) sampling continuous variable $u \sim q_{m,h \to h'}(\cdot)$ for $(h, h')$ associated with $m$, and (iii) computing $(\theta_{h}', u') := g_{m,h \to h'}(\theta_h, u)$, and proposing new state $x' = (h', \theta_h')$.
-The move is accepted with probability:
-\begin{equation*}
-{\scriptstyle
-\min\left\{ 1,
-\frac{p(h')}{p(h)} \frac{p_{h'}(\theta_{h'})}{p_h(\theta_h)} \frac{L_{D,h'}(\theta_h')}{L_{D,h}(\theta_h)}
-\frac{q_{x'}(m)}{q_{x}(m)}
-\frac{q_{m,h' \to h}(u)}{q_{m,h \to h'}(u)}
-|J g_{m, h\to h'}|(\theta_h, u)
-\right\}
-}
-\end{equation*}
-
-To represent reversible jump MCMC involutive MCMC implemented with probabilistic programs, we write a probabilistic program $\mathcal{P}$ that encodes the space of models $\mathcal{H}$, the prior distribution on models, $p(h)$, the per-model priors $p_h(\theta_h)$ and the per-model likelihoods $L_{D,h}(\theta_h)$.
-The set of all models $h$ corresponds to the set of all pairs $(K_\mathcal{P}, \mathbf{d}_\mathcal{P})$ where $K_\mathcal{P}$ represent possible trace structures (i.e.
+%The move is accepted with probability:
+%\begin{equation*}
+%{
+%\min\left\{ 1,
+%\frac{p(h')}{p(h)} \frac{p_{h'}(\theta_{h'})}{p_h(\theta_h)} \frac{L_{D,h'}(\theta_h')}{L_{D,h}(\theta_h)}
+%\frac{q_{x'}(m)}{q_{x}(m)}
+%\frac{q_{m,h' \to h}(u)}{q_{m,h \to h'}(u)}
+%|J g_{m, h\to h'}|(\theta_h, u)
+%\right\}
+%}
+%\end{equation*}
+
+To encode reversible jump MCMC in our framework, we write a probabilistic program $\mathcal{P}$ that encodes the space of models $\mathcal{H}$, the prior distribution on models, $p(h)$, the per-model priors $p_h(\theta_h)$ and the per-model likelihoods $L_{D,h}(\theta_h)$.
+The set of all models $h$ is encoded in the set of all pairs $(K_\mathcal{P}, \mathbf{d}_\mathcal{P})$ where $K_\mathcal{P}$ represent possible trace structures (i.e.
 control-flow paths through $\mathcal{P}$) and $\mathbf{d}_\mathcal{P}$ are the set of assignments to discrete random choices made by $\mathcal{P}$.
-The per-model continuous parameters $\theta$ correspond to continuous random choices $\mathbf{c}_\mathcal{P}$.
+The per-model continuous parameters $\theta$ are encoded via continuous random choices $\mathbf{c}_\mathcal{P}$.
 % TODO explain the likelihood
-An auxiliary probabilistic program $\mathcal{Q}$ encodes both the probability distribution on moves types using discrete random choices and possibly stochastic control flow ($(K_\mathcal{Q}, \mathbf{d}_\mathcal{Q})$), and the per-move-type probability densities on $u$ using continuous random choices $\mathbf{c}_\mathcal{Q}$.
+The auxiliary probabilistic program $\mathcal{Q}$ encodes both the probability distribution on moves types using discrete random choices and possibly stochastic control flow ($(K_\mathcal{Q}, \mathbf{d}_\mathcal{Q})$), and the per-move-type probability densities on $u$ using continuous random choices $\mathbf{c}_\mathcal{Q}$.
 The involution $f$ factors into an (i) involution $f_1$ on pairs $i = ((K_\mathcal{P}, \mathbf{d}_\mathcal{P}), (H_\mathcal{Q}, \mathbf{d}_\mathcal{Q}))$ that defines the association between move types $\mathcal{M}$ and the model pairs ($h_1, h_2$); and (ii) a family of bijections $f_{1,i}$ on the space of pairs $(\mathbf{c}_\mathcal{P}, \mathbf{c}_\mathcal{Q})$ of continuous random choices for both programs for fixed values of the discrete random choices and fixed trace structure.
@@
 Figure~\ref{fig:mixture} shows a split-merge reversible jump kernel for an infinite Gaussian mixture model~\citep{richardson1997bayesian} implemented using the probabilistic and differentiable programming languages described in this paper.
-Figure~\ref{fig:mixture}(b) shows the infinite Gaussian mixture model, specified as a probabilistic program $\mathcal{P}$.
+Figure~\ref{fig:mixture}b shows the infinite Gaussian mixture model, specified as a probabilistic program $\mathcal{P}$.
 The program takes the number of data points as input, then samples the number of clusters from a Poisson distribution, then samples cluster parameters and mixture proportions, and finally samples the data points from the resulting finite mixture.
-Figure~\ref{fig:mixture}(c) shows the auxiliary probabilistic program $\mathcal{Q}$ for the split-merge kernel.
+Figure~\ref{fig:mixture}c shows the auxiliary probabilistic program $\mathcal{Q}$ for the split-merge kernel.
 This program takes a trace of the model program as input, and randomly decides whether to split a cluster and increase the number of clusters by one or merge two clusters and decrease the number of clusters by one.
@@
 If a split is chosen, then the program also samples the three degrees of freedom necessary to generate the new parameters for the clusters in an invertible manner.
-Figure~\ref{fig:mixture}(d) shows a differentiable program specifying the involution for the split-merge kernel, and Figure~\ref{fig:mixture}(f) shows graphically how this involution acts on pairs of traces.
+Figure~\ref{fig:mixture}d shows a differentiable program specifying the involution for the split-merge kernel, and Figure~\ref{fig:mixture}f shows graphically how this involution acts on pairs of traces.
 The yellow section (1) defines an involution $f_1$ on the discrete random choices that specifies that (i) the \texttt{split} choice should be flipped (so that split moves are always mapped to merge moves and vice versa) and that (ii) the number of clusters should be increased by one for a split move and decreased by one for a merge move, and (iii) which merged cluster corresponds to which split clusters.
@@
 = \sum_{i \in I} \int_{K_i} \left| (Jf_{2,i})(\mathbf{x})\right| \mu_{n_i}(d \mathbf{x}) % TODO how to justify this? it seems obvious
-= \sum_{i \in I} \mu_{n_i}(f_{2,i}(K_i)) % Borgachev Theorem 4.7.1 
+= \sum_{i \in I} \mu_{n_i}(f_{2,i}(K_i)) % Bogachev Theorem 4.7.1 
 \end{align}
-where the final step uses Borgachev Theorem 3.7.1 with $g := 1$, and $F := f_{2,i}$.
+where the final step uses Bogachev Theorem 3.7.1 with $g := 1$, and $F := f_{2,i}$.
 
 \subsection{Proof of detailed balance for involution} \label{sec:involution-detailed-balance}
-
-\begin{lemma}
-Given $\sigma$-finite measures $(X_P, \Sigma_P, \mu_P)$ and $(X_Q, \Sigma_Q, \mu_Q)$,
-if $X \in \Sigma_P \otimes \Sigma_Q$ then $\Sigma = \{A \in \Sigma_P \otimes \Sigma_Q : K \subseteq X\}$ is a $\sigma$-algebra on $X$, and $\mu : \Sigma \to [0, \infty)$ given by $\mu(A) := (\mu_P \times \mu_Q)(A)$ is $\sigma$-finite measure on measurable space $(X, \Sigma)$.
-\end{lemma}
-\begin{proof}
-TODO: write out the detailed steps, proving each property of a $\sigma$-algebra%$ TODO
-\end{proof}
-
-\begin{lemma}
-$\pi : X \to [0, \infty)$ is a probability density with respect to $\mu$ and the measure induced by $\pi$ and the measure $\mu$ are mutually absolutely continuous 
-\end{lemma}
-\begin{proof}
-% TODO need to show that pi is measurable by the product measure?
-\begin{equation}
-\int_X \pi(x) \mu(dx) = \int_{X_P \times X_Q} \pi(x) (\mu_P \times \mu_Q)(dx) = \int_{X_P} p(t) \left( \int_{X_Q} q_t(u) \mu_Q(du) \right) \mu_P(dt) = 1
-\end{equation}
-
-%(note that $\mu$ is $\sigma$-finite since $\mu_P \times \mu_Q$ is $\sigma$-finite. % TODO: prove this 
-%Suppose that $\pi$ is a probability density with respect to $\mu$. % TODO is this guaranteed?
-\end{proof}
+The involutive MCMC kernel is composed of two parts:
+An extension of the state space, and an involution on the extended state space.
+First, we show detailed balance for the deterministic involution move applied to the extended state space.
+
+%\begin{lemma}
+%Given $\sigma$-finite measures $(X_P, \Sigma_P, \mu_P)$ and $(X_Q, \Sigma_Q, \mu_Q)$,
+%if $X \in \Sigma_P \otimes \Sigma_Q$ then $\Sigma = \{A \in \Sigma_P \otimes \Sigma_Q : K \subseteq X\}$ is a $\sigma$-algebra on $X$, and $\mu : \Sigma \to [0, \infty)$ given by $\mu(A) := (\mu_P \times \mu_Q)(A)$ is $\sigma$-finite measure on measurable space $(X, \Sigma)$.
+%\end{lemma}
+%\begin{proof}
+%TODO: write out the detailed steps, proving each property of a $\sigma$-algebra%$ TODO
+%\end{proof}
+%
+%\begin{lemma}
+%$\pi : X \to [0, \infty)$ is a probability density with respect to $\mu$ and the measure induced by $\pi$ and the measure $\mu$ are mutually absolutely continuous 
+%\end{lemma}
+%\begin{proof}
+%% TODO need to show that pi is measurable by the product measure?
+%\begin{equation}
+%\int_X \pi(x) \mu(dx) = \int_{X_P \times X_Q} \pi(x) (\mu_P \times \mu_Q)(dx) = \int_{X_P} p(t) \left( \int_{X_Q} q_t(u) \mu_Q(du) \right) \mu_P(dt) = 1
+%\end{equation}
+%
+%%(note that $\mu$ is $\sigma$-finite since $\mu_P \times \mu_Q$ is $\sigma$-finite. % TODO: prove this 
+%%Suppose that $\pi$ is a probability density with respect to $\mu$. % TODO is this guaranteed?
+%\end{proof}
 
 \citet{tierney1998note} gives a class of MCMC kernels based on involutions that satisfy detailed balance.
-We reproduce the result in our notation:
+We now reproduce the result in our notation:
 \begin{lemma}[Detailed balance for involution move~\citep{tierney1998note}] \label{lemma:tierney-involution}
@@
     \int_X k_x(A) p(x) \mu_P(dx) &= \int_X \left( \int_U k'_{x,u}(A \times U) q_x(u) \mu_Q(du) \right) p(x) \mu_P(dx)\\
-    &= \int_{X \times U} k'_{x,u}(A \times U) q_x(u) p(x) (\mu_P \times \mu_Q)(dz) \;\; [\mbox{ Tonelli's theorem }]\\ % TODO check
+    &= \int_{X \times U} k'_{x,u}(A \times U) q_x(u) p(x) (\mu_P \times \mu_Q)(dz)\\% \;\; [\mbox{ Tonelli's theorem }]\\ % TODO check
     &= \int_{Z} k'_{x,u}(\{(x', u') \in Z : x' \in A\}) \pi(z) \mu(dz)\\ % TODO justify
@@
 
-\subsection{A requirement on Densities for Involutive MCMC with Probabilistic Programs} \label{sec:technical-requirement}
-TODO: consider cutting this or moving to appendix -- it is a complicated technical requirement on programs
+\subsection{A Sufficient Condition for Involutive MCMC with Dictionaries} \label{sec:technical-requirement}
+%TODO: consider cutting this or moving to appendix -- it is a complicated technical requirement on programs
 
 % TODO cut this??
-To define the model and auxiliary densities used in involutive MCMC in Section~\ref{sec:involutive-mcmc} using probabilistic programs, we write one probabilistic program ($P$) to define the model's probability distribution, and another probabilistic program ($Q$) to define the set of auxiliary distributions.
-The model reference measure $(X, \Sigma_P, \mu_P)$ and density that were required in Section~\ref{sec:involutive-mcmc} are set to the probabilistic program's reference measure and density $p$ on traces as defined above (so that $X := T$).
-To define the auxiliary densities, we use a second probabilistic program $Q$ that is parametrized by (i.e. takes as input) traces of the model program $P$.
-The set of possible inputs to $Q$ is $\{x \in T : p(x) > 0\}$, and for each such $x$ the probabilistic program $Q$ defines a density $q_x : T \to [0, \infty)$ with respect to the reference measure on traces.\footnote{
-Note that we the same measure space $(T, \Sigma, \mu)$ on traces for both the model and auxiliary probabilistic program.
-If both programs sample some random choice at the same address $k$ and require different measure spaces $(V_k, \Sigma_k, \mu_k)$ for the address, this can be handled formally by making the addresses sampled by the two programs disjoint by prepending either the prefix `p.' or `q.' to addresses.
-This is only required for book-keeping in the formalism, and these prefixes do not need to be actually added to the addresses used in the programs.}
-
-The involutive MCMC framework of Section~\ref{sec:involutive-mcmc} requires that $Z := \{(x, y) \in X \times Y : \pi(x, y) > 0\}$ is $\mu_P \times \mu_Q$-measurable.
-We now give a sufficient condition for this to hold.
+%To define the model and auxiliary densities used in involutive MCMC in Section~\ref{sec:involutive-mcmc} using probabilistic programs, we write one probabilistic program ($P$) to define the model's probability distribution, and another probabilistic program ($Q$) to define the set of auxiliary distributions.
+%The model reference measure $(X, \Sigma_P, \mu_P)$ and density that were required in Section~\ref{sec:involutive-mcmc} are set to the probabilistic program's reference measure and density $p$ on traces as defined above (so that $X := T$).
+%To define the auxiliary densities, we use a second probabilistic program $Q$ that is parametrized by (i.e. takes as input) traces of the model program $P$.
+%The set of possible inputs to $Q$ is $\{x \in T : p(x) > 0\}$, and for each such $x$ the probabilistic program $Q$ defines a density $q_x : T \to [0, \infty)$ with respect to the reference measure on traces.\footnote{
+%Note that we the same measure space $(T, \Sigma, \mu)$ on traces for both the model and auxiliary probabilistic program.
+%If both programs sample some random choice at the same address $k$ and require different measure spaces $(V_k, \Sigma_k, \mu_k)$ for the address, this can be handled formally by making the addresses sampled by the two programs disjoint by prepending either the prefix `p.' or `q.' to addresses.
+%This is only required for book-keeping in the formalism, and these prefixes do not need to be actually added to the addresses used in the programs.}
+
+Our formulation of involutive MCMC requires the following technical condition to hold:
+$Z := \{(x, y) \in X \times Y : \pi(x, y) > 0\}$ is $\mu_P \times \mu_Q$-measurable.
+We now give a sufficient condition for this to hold, when $X$ and $Y$ are spaces of dictionaries.
+%$\pi(x, y) := p(x) q_x(y)$ is defined via probabilistic programs $\mathcal{P}$ and $\mathcal{Q}$.
 Let $D \subseteq \mathcal{K}$ denote the subset of addresses that are discrete (i.e. where $V_k$ is a countable set and $\mu_k$ is the counting measure).
@@
 \end{proof}
-
-Intuitively, this requirement means that for both the model probabilistic program $P$ and the auxiliary probabilistic program $Q$, the support of a random choice that is not discrete cannot depend on the value of another non-discrete random choice.
-Additionally, the support of non-discrete random choices in $Q$ cannot depend on the value of non-discrete random choices in the input $x$, which is a trace of $P$.
-This requirement defines a notion \emph{well-behavedness} for a probabilistic program ($P$) and an additional notion of well-behavedness for a pair of probabilistic programs that are sequenced one after the other ($P$ and $Q$).
+When $p$ and $q$ are defined via probabilistic programs $\mathcal{P}$ and $\mathcal{Q}$ respectively, 
+this requirement means that for both the model probabilistic program $\mathcal{P}$ and the auxiliary probabilistic program $\mathcal{Q}$, the support of a random choice that is not discrete cannot depend on the value of another non-discrete random choice.
+Additionally, the support of non-discrete random choices in $\mathcal{Q}$ cannot depend on the value of non-discrete random choices in the input $x$, which is a trace of $\mathcal{P}$.
+This requirement defines a notion \emph{well-behavedness} for a probabilistic program ($\mathcal{P}$) and an additional notion of well-behavedness for a pair of probabilistic programs that are sequenced one after the other ($\mathcal{P}$ and $\mathcal{Q}$).
+%It is possible to satisfy this requirement via a `disciplined' approach to probabilistic modeling, where the support of a random choice at a given address is fixed.
 % TODO: prove that if Q satisfeies the criterion, then the composition P circ Q satisfies the single-program criterion..

### 2020-07-19 paper.tex
tweaks to sections 3 and 4

@@
 \item An algorithm that automates the implementation of an involutive MCMC kernel given two probabilistic programs and a differentiable program encoding the involution, using automatic differentiation and tracing of probabilistic programs.
-\item An extension to the algorithm that exploits sparsity in the involution to reduce the computational complexity of an involutive MCMC kernel from $O(N^3)$ to $O(1)$ in common cases, where $N$ is the dimensionality of the latent space.
+\item An extension to the algorithm that exploits sparsity in the involution to reduce the number of operations needed to apply an involutive MCMC kernel from $O(N^3)$ to $O(1)$ in some cases, where $N$ is the dimensionality of the latent space.
 \item An algorithm that dynamically detects errors in the specification of an involutive MCMC kernel.
@@
 Some systems allow for user-defined variational families or proposal distributions~\citep{ritchie2016deep,bingham2019pyro}.
-Programmable inference~\citep{mansinghka2018probabilistic} proposes that inference algorithms be specified by users in high-level programming languages.
+Programmable inference~\citep{mansinghka2018probabilistic} proposes that inference algorithms be programmed by users using new high-level inference abstractions.
 The Gen probabilistic programming system~\citep{cusumano2019gen} exposes an API that supports high-level user implementations of an open-ended set of inference algorithms, and abstracts away low-level implementation details of inference algorithms.
@@
 Let $V_k$ denote the set of possible values for key $k$, where $V_k$ is a countable set for each discrete key, and where $V_k = \mathbb{R}^{d_k}$ for each continuous key for some $d_k$.
-Given a set of keys $K$, let $V_K = \times_{k \in A} V_k$ denote the set of \textit{assignments} of values to each key.
+Given a set of keys $K$, let $V_K = \times_{k \in K} V_k$ denote the set of \textit{assignments} of values to each key.
 Then the set of all finite dictionaries is $\mathcal{D} := \bigcup_{K \subset \mathcal{K}, |K| < \infty} \{(K, \mathbf{x}) \mid \mathbf{x} \in V_K\}$. That is, a dictionary specifies 
@@
 \textbf{Relationship to representation of~\citet{green1995reversible}.}
-\citet{green1995reversible} uses a state space that is the countable union of `models', where each model uses a vector of real-valued parameters.
-Dictionaries possess substantially more structure:
+\citet{green1995reversible} uses a state space that is the countable union of `models', where each model is typically a vector of real-valued parameters.
+Dictionaries have substantially more structure:
 Instead of monolithic `models', dictionaries use a more elaborate discrete state that includes the set of keys and the assignment to the discrete keys. 
@@
 We associate a measure $\mu_k$ on $V_k$ with each key $k$---the counting measure for each discrete key and the Lebesgue-measure on $\mathbb{R}^{d_k}$ for each continuous key.
-For each finite set of keys $K$, we make $V_K$ a measure space using the standard product $\sigma$-algebra $\Sigma_K = \otimes_{k \in A} \Sigma_k$ and the product measure $\mu_K = \times_{k \in A} \mu_k$.
+For each finite set of keys $K$, we make $V_K$ a measure space using the standard product $\sigma$-algebra $\Sigma_K = \otimes_{k \in K} \Sigma_k$ and the product measure $\mu_K = \times_{k \in K} \mu_k$.
 We equip $\mathcal{D}$ with the $\sigma$-algebra $\Sigma_{\mathcal{D}} := \{\bigcup_{K \subset \mathcal{K}, |K| < \infty} \{(K, \mathbf{x}) \mid \mathbf{x} \in B_K\} \mid B_K \in \Sigma_K \, \text{for each finite } K \subset \mathcal{K}\}$ to obtain a measurable space
 of dictionaries.
-A reference measure $\mu_\mathcal{D}$ on this space can be constructed using the product measures $\mu_K$: we set $\mu(B) := \sum_{A \in \mathcal{K}, |K| < \infty} \mu_K(\{\mathbf{x} \mid (K, \mathbf{x}) \in B\})$.
-
-\textbf{Notation for dictionaries.} Given a dictionary $d = (K, \mathbf{x})$, we write $K_d$ for $K$ and $d[k]$ for the value $x_k$ associated with a key $k \in K$.
+A reference measure $\mu_\mathcal{D}$ on this space can be constructed using the product measures $\mu_K$: we set $\mu(B) := \sum_{K \in \mathcal{K}, |K| < \infty} \mu_K(\{\mathbf{x} \mid (K, \mathbf{x}) \in B\})$.
+
+\textbf{Notation for dictionaries.} Given a dictionary $m = (K, \mathbf{x})$, we write $K_m$ for $K$ and $m[k]$ for the value $x_k$ associated with a key $k \in K$.
 We also denote specific dictionaries using notation $\{k_1 \mapsto v_1, k_2 \mapsto v_2, \ldots\}$.
@@
 More generally a probability distribution on dictionaries is defined by a probability density $p : \mathcal{D} \rightarrow [0, \infty)$ such that $\int p(m) \mu_\mathcal{D}(\text{d}m) = 1$. 
-The probability is distributed among the finite sets of keys $K \subseteq \mathcal{K}$ where $|K| < \infty$:
-\begin{equation}
-1 = \int p(m) \mu_\mathcal{D}(\text{d}m) = \sum_{\substack{K \subseteq \mathcal{K}\\|K| < \infty}} \int p(m) \mu_{K}(dm)
-\end{equation}
-where the probability for each key set $K$ is:
-\begin{equation}\label{eq:integral}
-\int p(m) \mu_{K}(dm) = \sum_{\mathbf{x}_1} \int_{\mathbb{R}^{d}} p((K, (\mathbf{x}_1, \mathbf{x}_2))) d\mathbf{x}_2
-\end{equation}
-where $d: = \sum_{k \in A \cap \mathcal{J}} d_{k}$ and where $\mathbf{x}_1$ is an assignment to the discrete choices in $K$ and $\mathbf{x}_2$ is an assignment to the continuous choices.
+The probability mass is distributed among the finite sets of keys $K \subseteq \mathcal{K}$:
+\begin{equation} \label{eq:integral}
+1 = \sum_{\substack{K \subseteq \mathcal{K}\\|K| < \infty}}
+\left( \sum_{\mathbf{x}_1} \left( \int_{\mathbb{R}^{d_K}} p((K, (\mathbf{x}_1, \mathbf{x}_2))) d\mathbf{x}_2 \right) \right)
+\end{equation}
+where $d_K : = \sum_{k \in K \cap \mathcal{J}} d_{k}$ is the total continuous dimension for keys $K$, and where $\mathbf{x}_1$ is an assignment to the discrete choices in $K$ and $\mathbf{x}_2$ is an assignment to the continuous choices.
 
@@
 In this case, we have $V_\texttt{k} = \mathbb{N}$ with the counting measure for $\mu_\texttt{k}$, and for all other keys $k \in \mathcal{K}$, $V_k = \mathbb{R}$ with the Lebesgue measure for $\mu_k$.
-For each $j \in \{0, 1, \ldots\}$, the probability for key set
+For each $j \in \{0, 1, \ldots\}$, the probability mass assigned to key set
 $\{\texttt{k}, \texttt{y}_\texttt{1}, \dots, \texttt{y}_\texttt{n}\} \cup \{\texttt{m}_i \mid 1 \leq i \leq j\} \cup \{\texttt{s}_i \mid 1 \leq i \leq j\}$
@@
 Suppose $B$ is a set of keys that almost always appear for $p$, and that $b = (B, \mathbf{b})$ is a dictionary with keys $B$.
-Furthermore, let $(K_m, \mathbf{m}) \oplus (K_n, \mathbf{n}) := (K_m \cup K_n, \mathbf{m} \times \mathbf{n})$ be the \textit{merge} of two dictionaries $m$ and $n$ defined on disjoint key sets $K_m$ and $K_n$.
+Furthermore, let $(K_m, \mathbf{m}) \oplus (K_n, \mathbf{n}) := (K_m \cup K_n, (\mathbf{m}, \mathbf{n}))$ be the \textit{merge} of two dictionaries $m$ and $n$ defined on disjoint key sets $K_m$ and $K_n$.
 Then we can define the conditional density $p(d \mid b) := \mathbf{1}[K_d \cap B = \emptyset] \frac{p(d \oplus b)}{\int_{\{m \mid K_m \cap B = \emptyset\}} p(m \oplus b) \mu_\mathcal{D}(\text{d}m)}$ when the denominator is finite.
-If the reference measure $\mu_k$ associated with each key $k \in B$ is discrete (e.g., the counting measure), then this definition corresponds to the ordinary notion of conditioning on an event (namely, the event that a sample from $p$ agrees with the dictionary $b$ on all keys in $B$).
+If each each key $k \in B$ is discrete (e.g.
+$\mu_k$ is the counting measure), then this definition corresponds to the ordinary notion of conditioning on an event (namely, the event that a sample from $p$ agrees with the dictionary $b$ on all keys in $B$).
 When this is not the case, it corresponds to a more general measure-theoretic notion called disintegration~\citep{chang1997conditioning}.
@@
 Then each set $Z_e$ is isomorphic to a Euclidean space of assignments to the continuous keys in the two dictionaries.
-Suppose there is an involution $g : E \to E$ between elements of the partition (a function with $g = g^{-1}$), and a family of continuously differentiable bijections $h_{e} : Z_e \to Z_{g(e)}$ indexed by $e \in E$, with $h_{e} = h_{g(e)}^{-1}$.
+Suppose there is an involution $g : E \to E$ between elements of the partition, and a family of continuously differentiable bijections $h_{e} : Z_e \to Z_{g(e)}$ indexed by $e \in E$, with $h_{e} = h_{g(e)}^{-1}$.
 Let $e(z) \in E$ denote which element of the partition a dictionary $z \in Z$ belongs to.
@@
 Our probabilistic programming language augments the syntax of Julia~\citep{bezanson2017julia}
-with a single new construct, the \texttt{\{a\} $\sim$ d} expression, for making a \textit{named} random choice.
-An execution trace of a Gen probabilistic program can be sampled by running the program according to Julia's usual semantics, and upon encountering an expression of the form \texttt{\{a\} $\sim$ d}, (i) evaluating the \textit{address expression} \texttt{a} to obtain an address $k \in \mathcal{K}$; (ii) evaluating the \textit{distribution expression} $d$ to obtain a distribution $d$ over the measurable space $V_k$; (iii) sampling a value $x_k \sim d$ and adding the mapping $\{k \mapsto x_k\}$ to a dictionary called the program trace; and (iv) returning the sampled value $x_k$ to the program, to continue execution.
+with a single new construct, the `$\{\mathtt{address}\} \sim \mathtt{distribution}$' expression, for making a \textit{named} random choice.
+An execution trace of a Gen probabilistic program can be sampled by running the program according to Julia's usual semantics, and upon encountering an expression of the form `$\{\mathtt{address}\} \sim \mathtt{distribution}$', (i) evaluating the \textit{address expression} (`$\mathtt{address}$') to obtain an address $k \in \mathcal{K}$; (ii) evaluating the \textit{distribution expression} (`$\mathtt{distribution}$') to obtain a probability distribution over the measurable space $V_k$; (iii) sampling a value $x_k$ from this distribution and adding the mapping $\{k \mapsto x_k\}$ to a dictionary called the program trace; and (iv) returning the sampled value $x_k$ to the program, to continue execution.
 When execution terminates, the program trace has accumulated a mapping for each random choice encountered during the program's execution.
@@
 The density of an arbitrary dictionary $(K, \mathbf{x})$ under any probabilistic program's distribution on traces can be computed using \textsc{score}.
-The idea is to run the probabilistic program, and whenever a random choice \texttt{\{a\} $\sim$ d} is encountered, to look up the value $x_k$ in the dictionary, compute its density under the primitive distribution $d$, multiply this density into a running total, and return control to the probabilistic program as if the sampling instruction had executed and returned $x_k$.
+The idea is to run the probabilistic program, and whenever a random choice `\texttt{\{address\} $\sim$ distribution}' is encountered, to look up the value $x_k$ in the dictionary, compute its density under the primitive distribution $d$, multiply this density into a running total, and return control to the probabilistic program as if the sampling instruction had executed and returned $x_k$.
 At the end, the running total can be returned as the trace's density.
@@
 
-If the two probabilistic programs $\mathcal{P}$ and $\mathcal{Q}$ satisfy a simple technical requirement (described in Section~\ref{sec:technical-requirement}), we can interpret them as specifying the model density $p$ and auxiliary densities $q_x$ for use with involutive MCMC on dictionaries (Algorithm~\ref{alg:involutive-mcmc}).
-(In this case, the program $\mathcal{Q}$
-accepts a trace $x$ of $\mathcal{P}$ as input, and the distribution for $q_x$ is given by $\mathcal{Q}_x$.)
-
-However, typically, we cannot use a probabilistic program $\mathcal{P}$ to represent the target distribution directly: probabilistic programs implement simulators for a distribution, but target densities are typically not tractable to simulate (hence the need for MCMC).
+We can use two probabilistic programs $\mathcal{P}$ and $\mathcal{Q}$ to specify the model density $p$ and auxiliary densities $q_x$ that appear in involutive MCMC
+(Algorithm~\ref{alg:involutive-mcmc}).\footnote{The programs must satisfy a mild technical requirement for our formalism to go through; see Section~\ref{sec:technical-requirement}}
+In this case, the program $\mathcal{Q}$ 
+accepts a trace $x$ of $\mathcal{P}$ as input, we denote a probabilistic program $\mathcal{Q}$
+applied to input $x$ by $\mathcal{Q}_x$.
+
+Typically, we cannot use a probabilistic program $\mathcal{P}$ to represent the target distribution directly: probabilistic programs implement simulators for a distribution, but target densities are typically not tractable to simulate (hence the need for MCMC).
 Instead, we may wish to sample from a target $p$ that arises from conditioning a probabilistic program $\mathcal{P}$'s distribution over traces on observations of the values at some addresses.

### 2020-07-19 paper.tex
pass through poaper

@@
 \usepackage[usenames,dvipsnames,svgnames]{xcolor}
+
+% for scaling down code listings
+\newsavebox{\mybox}
 
@@
 
-Section~\ref{sec:probabilistic-programming} showed that if the densities $p$ and $q$ are specified using probabilistic programs $\mathcal{P}$ and $\mathcal{Q}$, then density ratio in the acceptance probability for involutive MCMC on dictionaries Equation~(\ref{eq:acceptance-ratio-dictionaries}) can be automated using probabilistic programming techniques.
+Section~\ref{sec:probabilistic-programming} showed that if the densities $p$ and $q$ are specified using probabilistic programs $\mathcal{P}$ and $\mathcal{Q}$, then the density ratio in the acceptance probability for involutive MCMC on dictionaries (Equation~(\ref{eq:acceptance-ratio-dictionaries})) can be automated using probabilistic programming techniques.
 This section shows that if the involution $f$ is specified in using a differentiable program $\mathcal{F}$ that transforms the execution traces of probabilistic programs, then the Jacobian factor in Equation~(\ref{eq:acceptance-ratio-dictionaries}) can also be automated, using automatic differentiation.
@@
 \end{center}
-The first argument is the address $a$ and the second argument is either $\mathtt{discrete}$ or $\mathtt{continuous}$, and informs the interpreter whether the random choice at that address is drawn from a discrete or continuous distribution (this information will be used to support automatic differentiation).
+The first argument is the address $a$ and the second argument is either $\mathtt{:}\mathtt{discrete}$ or $\mathtt{:}\mathtt{continuous}$, and informs the interpreter whether the random choice at that address is drawn from a discrete or continuous distribution (this information will be used to support automatic differentiation).
 Recall that $x$ is the trace of the model probabilistic program, and $y$ is the trace of the auxiliary probabilistic program.
@@
 {\begin{lstlisting}[basicstyle=\small\ttfamily]
-    val = @read(model_in[:r], continuous)
+    val = @read(model_in[:r], :continuous)
 \end{lstlisting}}
@@
 {\begin{lstlisting}[basicstyle=\small\ttfamily]
-    @write(model_in[:r], val, continuous)
+    @write(model_in[:r], val, :continuous)
 \end{lstlisting}}
@@
 For example, it is not possible to write an output trace and then read the written value from it later.
+
 Often, we want to simply copy the value from some address in the input traces to some address in the output trace.
-While this is possible via a $\mathtt{@read}$ followed by a $\mathtt{@write}$, the language provides a special syntax for copies:
+While this is possible via a $\mathtt{@read}$ followed by a $\mathtt{@write}$, the language provides a special syntax:
 \noindent
@@
 \end{center}
-Like above, the two addresses here are both pairs of the trace and the address within that trace.
 For example, to copy the value from address $\mathtt{:}\mathtt{u}$ in $x$ to address $\mathtt{:}\mathtt{v}$ in $x'$, we use:
@@
 Of course, it is also possible to copy from $x$ to $y'$, from $y$ to $x'$ and from $y$ to $y'$.
-As we will see in Section~\ref{sec:sparsity}, it is preferable to copy when possible instead of reading and then writing, as this can improve the efficiency of automatic differentiation.
+As we will see in Section~\ref{sec:sparsity}, it is preferable to use $\mathtt{@copy}$ when possible instead of reading and then writing, as this can improve the efficiency of automatic differentiation.
 
@@
 {\begin{lstlisting}[basicstyle=\small\ttfamily]
-@tracefn f
-   polar = @read(model_in[:polar], discrete)
+polar = @read(model_in[:polar], :discrete)
 if polar
-      r = @read(model_in[:r]), continuous)
-      theta = @read(model_in[:theta], continuous)
+   r = @read(model_in[:r]), :continuous)
+   theta = @read(model_in[:theta], :continuous)
    x = r * sin(theta)
    y = r * cos(theta)
-      @write(model_out[:x], x, continuous)
-      @write(model_out[:y], y, continuous)
+   @write(model_out[:x], x, :continuous)
+   @write(model_out[:y], y, :continuous)
 else
-      x = @read(model_in[:x], continuous)
-      y = @read(model_in[:y], continuous)
+   x = @read(model_in[:x], :continuous)
+   y = @read(model_in[:y], :continuous)
    r = (x^2 + y^2)
    theta = atan(y, x)
-      @write(model_out[:r], r, continuous)
-      @write(model_out[:theta], theta, continuous)
+   @write(model_out[:r], r, :continuous)
+   @write(model_out[:theta], theta, :continuous)
 end
-   @write(model_out[:polar], !polar, discrete)
-end
+@write(model_out[:polar], !polar, :discrete)
 \end{lstlisting}}
@@
 
-Suppose $N$ and $N'$ are the total number of random choices in the input traces $x \oplus y$ and output traces $x' \oplus y'$ respectively, and $M$ is the number of continuous random choices (which is the same in the input traces and output traces).
+Suppose $N$ and $N'$ are the total number of random choices in the input traces $x, y$ and output traces $x', y'$ respectively, and $M$ is the number of continuous random choices (which must be the same in the input traces and output traces).
 Then, the number of operations used in Algorithm~\ref{alg:auto-involutive-mcmc} grows as $O(N + N') + O(M^3)$.
@@
 \end{equation}
-Using the cofactor expansion of the determinant, we observe that for any (`copy') column in an $M$-by-$M$ Jacobian matrix (a column with a single $1$ and all other entries $0$), the absolute value of the determinant is equivalent to that of the $(M-1)$-by-$(M-1)$ sub-matrix with the corresponding row \textit{and} column omitted (even if that would remove other nonzero entries from the matrix).
+Using the cofactor expansion of the determinant, we observe that for any `copy' column in an $M$-by-$M$ Jacobian matrix (a column with a single $1$ and all other entries $0$), the absolute value of the determinant is equivalent to that of the $(M-1)$-by-$(M-1)$ sub-matrix with the corresponding column \textit{and} row omitted (even if that would remove other nonzero entries from the matrix).
 By applying this rule recursively, we can instead compute the determinant of a much smaller matrix; for the example above with $M=4$, the absolute value of the determinant simplifies to the absolute value of a single entry ($|-1|$).
@@
 
-Many involutive MCMC kernels update only a constant number of keys that does not depend on the sizes $N$ and $N'$ of the input and output traces or the total number of continous keys $M$; the other keys are copied over unchanged.
+Many involutive MCMC kernels modify only a constant number of keys that does not depend on the sizes $N$ and $N'$ of the input and output traces or the total number of continous keys $M$; the other keys are copied over unchanged.
 For example, consider the Jacobian for a split-merge reversible jump move~\citet{richardson1997bayesian}, which is implemented in Figure~\ref{fig:mixture}:
@@
 
-\textproc{naive-trace-update} is a naive implementation of a general \emph{trace-update} operation that is part of Gen's API.
-While the naive implementation requires $O(N_1 + N_1')$ operations, it is possible for more sophisticated implementations of \emph{trace-update} to run in approximately $O(|K_{\delta}|)$ operations for certain probabilistic programs $\mathcal{P}$ by exploiting conditional independence structure program that leads to cancellations in the density ratio $\tilde{p}(x' \oplus) / \tilde{p}(x \oplus b)$.
+\textproc{naive-trace-update} is a naive implementation of a general \emph{trace-update} operation that is part of Gen's API~\citep{cusumano2019gen}.
+While the naive implementation requires $O(N_1 + N_1')$ operations, it is possible for more sophisticated implementations of \emph{trace-update} to run in approximately $O(|K_{\delta}|)$ operations for certain probabilistic programs $\mathcal{P}$ by exploiting conditional independence structure program that leads to cancellations in the density ratio $\tilde{p}(x' \oplus b) / \tilde{p}(x \oplus b)$.
 The details of these more efficient implementations are outside the scope of this paper, but are implemented as part of the Gen system.
+% TODO be more precise:
 (As a simple example, consider a density $p$ that factors into a product of independent densities, one for each random choice; and an involution that only modifies one of these choices).
@@
 Therefore, the involution program can run in $O(|K_{\delta}| + N_2 + N_2')$ operations.
-We replace the separate calls to \textproc{score} for the log-densities $\log p(x)$ and $\log p(x')$ and replace them with a single call to \emph{trace-update} on $\mathcal{P}$ that takes in $x$, and $\delta$ and returns $x'$ and $\log p(x') - \log p(x)$.
-When $|K_{\delta}|$ and $N_2$ and $N_2'$ are all constant in the size $N_1$ of the model trace $x$, the resulting algorithm uses a number of operations that is constant in $O(N_1 + N_1')$.
+We replace the separate calls to \textproc{score} for the log-densities $\log \tilde{p}(x \oplus b)$ and $\log \tilde{p}(x' \oplus b)$ and replace them with a single call to \emph{trace-update} on $\mathcal{P}$ that takes in $x$ and $\delta$ and returns $x'$ and $\log (\tilde{p}(x' \oplus b) / \tilde{p}(x \oplus b))$.
+When $|K_{\delta}|$ and $N_2$ and $N_2'$ are all constant in the size $N_1$ of the model trace $x$, the resulting algorithm uses a number of operations that does not grow with $N_1 + N_1'$.
+% TODO need to differentiate between number of data and latent dimension here!
+% it is only constant in the latent dimension, not in the data.. (except in unusual cases)
 
@@
 
-Due to the automation described in the previous section, users of Gen can implement involutive MCMC algorithms by writing probabilistic programs for $p$ and $q$ and a differentiable program for the involution $f$, avoiding the need to derive and implement the accept/reject formula by hand.
+The automated technique described in the previous sections allows users to implement involutive MCMC algorithms by writing probabilistic programs for $p$ and $q$ and a differentiable program for the involution $f$, avoiding the need to derive and implement the accept/reject formula by hand.
 This eliminates certain classes of errors that could otherwise be difficult to root out in hand-coded implementations of MCMC algorithms.
@@
 
-To test the user's algorithm, Gen first samples a model trace $x$ and simulated observations $b$ from the model prior $p$, and auxiliary trace $y$ from the $q_x$. It then runs three tests:
+The check first randomly samples a model trace $x$ and simulated observations $b$ from the model prior $p$, and auxiliary trace $y$ from the $q_x$.
+It then runs three tests on this simulated test case:
 
 \begin{enumerate}
-\item \textbf{Support check.} The \textit{support check} runs the involution $f$ on $(x, y)$ and checks that the resulting pair of traces, $(x', y')$, are within the set $Z$ of positive-density elements for $\pi$.
-
-\item \textbf{Dimension check.} The \textit{dimension check} runs the involution $f$ on $(x, y)$ and records the sets of addresses $R$ and $W$ at which $f$ performs continuous reads and writes, respectively.
-It then checks that the sizes of these sets match.
-This check is run when computing the determinant of the Jacobian in \textproc{run-involution}.
-
-\item \textbf{Involution check.} The \textit{involution check} runs $f \circ f$ on $(x, y)$ and checks that the result is equal to $(x, y)$.
+\item \textbf{Support check.} The \textit{support check} runs the involution $\mathcal{F}$ on $(x, y)$ and checks that the resulting pair of traces, $(x', y')$, are within the set $Z$ of positive-density elements for $\pi$.
+
+\item \textbf{Dimension check.} The \textit{dimension check} runs the involution $\mathcal{F}$ on $(x, y)$ and records the sets of addresses $R$ and $W$ at which $\mathcal{F}$ performs continuous reads and writes, respectively, and checks that the sizes of these sets match.
+(This check is part of computing the determinant of the Jacobian in \textproc{run-involution}).
+
+\item \textbf{Involution check.} The \textit{involution check} runs $\mathcal{F}$ twice to compute $f(f(x, y))$ and checks that the result is equal to $(x, y)$.
 If $f$ is an involution on $Z$, then this check will always succeed; if there is a set $\tilde{Z}$ of positive $\pi$-measure on which $f(f(z)) \neq z$ (i.e., $f$ is not an involution), then this check has positive probability of failing.
@@
 %$x$ is the model state, $y$ is the momentum, and the involution $f$ first runs a leapfrog integrator to obtain a new state $x'$ and final momentum $-y'$, then negates the final momentum and returns $(x', y')$.
-When applied to HMC, Gen's dynamic bug check involves:
+When applied to HMC, the dynamic bug check involves:
 %Gen's dynamic checks will uncover mathematical errors in the implementation of the leapfrog integrator, and will also detect a failure to perform the final negation of the momentum.
@@
 
-As another, simpler example of this type of error, consider the \texttt{split\_params} and \texttt{merge\_params} functions invoked in Figure~\ref{fig:mixture}, as part of a reversible-jump MCMC kernel for learning a mixture of Gaussians with an unknown number of components.
+As another, simpler example of this type of error, consider the \texttt{split\_params} and \texttt{merge\_params} functions invoked in Figure~\ref{fig:mixture}, as part of a reversible-jump MCMC kernel for inferring a mixture of Gaussians with an unknown number of components.
 Considering only the $\mu$ parameter, suppose the \textit{split} move samples an auxiliary variable $u_1$ and computes $\mu_1 = \mu + u_1, \mu_2 = \mu - u_1$ as the means of the two new clusters.
@@
 a new component to the \textit{end} (index $k+1$), so the move is not
-reversible. Gen's involution check will discover that the deletion of a
+reversible. The involution check will discover that the deletion of a
 component with index $i < k$ is not reversed by a corresponding birth move,
@@
 Such errors can be difficult to detect statically, because the addresses at which a probabilistic program makes random choices, and the distributions of those choices, may change from sample to sample.
-(Previous work has explored static analyses based on types~\citep{popltraces} and abstract interpretation~\citep{verified_svi}, but these each work on limited subsets of the programs that Gen's full modeling language permits, and it is often precisely these more complex programs that require the flexibility of the involutive MCMC framework in the first place.) Gen's dynamic support and dimension checks can help to detect bugs like these.
+(Previous work has explored static analyses based on types~\citep{popltraces} and abstract interpretation~\citep{verified_svi}, but these each work on limited subsets of the programs that Gen's full modeling language permits, and it is often precisely these more complex programs that require the flexibility of the involutive MCMC framework in the first place).
+Our dynamic support and dimension checks can help to detect bugs like these.
 For example, if the involution $f$ writes to a misspelled address, the support check will determine that the resulting trace's density is 0.
@@
 \begin{tabular}{c}
-{\begin{lstlisting}[basicstyle=\small\ttfamily]
+\begin{lrbox}{\mybox}%
+\begin{lstlisting}[basicstyle=\small\ttfamily]
 @copy(model_in[subtree_address], aux_out[:new_subtree])
 @copy(aux_in[:new_subtree], model_out[subtree_address])
-\end{lstlisting}}
-\end{tabular}
-\end{center}
-Note that the subtrees here happen to be trees that themselves have a random number of random choices, but the same general construction for a mixture of proposals below can also be used for simpler Metropolis-Hastings kernels:
+\end{lstlisting}
+\end{lrbox}%
+\scalebox{0.9}{\usebox{\mybox}}
+\end{tabular}
+\end{center}
+This is an instance of a more general pattern for implementing state-dependent mixture proposals:
 \begin{enumerate}
-\item The auxiliary probabilistic program samples from a distribution over different sets of random choices that will be proposed to
+\item The auxiliary probabilistic program $\mathcal{Q}$ samples from a distribution over different sets of random choices that will be proposed to
 (in this case, each set is a different subtree of the parse tree).
-\item The auxiliary probabilistic program then samples new values for those random choices (in this case, a new subtree).
-\item The involution program swaps the previous values of those random choices with their new values, by swapping data between the model trace and the auxiliary trace.
-\item The involution program copies the choice of what subset of random choices to propose to from the input auxiliary trace to the output auxiliary trace.
+\item The auxiliary probabilistic program $\mathcal{Q}$ then samples new values for those random choices (in this case, a new subtree).
+\item The involution program $\mathcal{F}$ swaps the previous values of those random choices with their new values, by swapping data between the model trace and the auxiliary trace.
+\item The involution program $\mathcal{F}$ copies the random choices that determined what subset of random choices to propose to from the input auxiliary trace to the output auxiliary trace.
 \end{enumerate}
 
-Because the mixture distribution is specified using a probabilistic program, it is straightforward to modify the program to define a different mixture distribution.
+Note that because the mixture distribution is specified using a probabilistic program, it is straightforward to modify the program $\mathcal{Q}$ to define a different mixture distribution.
 The code below specifies a mixture distribution that is uniform over all nodes in the tree.
@@
 Note that the probability of choosing a given subtree to propose to is itself changed when the subtree changes.
-Therefore, the mixture probabilities must be accounted for in the acceptance probability calculation.
+Therefore, the mixture probabilities do not in general cancel in the the acceptance probability calculation, and must be accounted for.
 For the original mixture distribution, the ratio of mixture probabilities is either $1$, $0.5$, or $2$ depending on whether the previous and new subtrees are leaf or internal nodes.
-For this alternative mixture distribution, the ratio of mixture probabilities is the ratio of sizes of the two trees.
+For this alternative mixture distribution, the ratio of mixture probabilities is the ratio of sizes of the two trees (e.g. $9/5$ or $5/9$ for the trees above).
 In both cases, our automated involutive MCMC algorithm automatically computes the acceptance probability.

### 2020-07-19 paper.tex
making main algorithm box less cluttered

@@
 
-\section{AUTOMATING INVOLUTIVE MCMC WITH TRACES} \label{sec:automating}
-
-\begin{algorithm*}[h]
+\begin{algorithm*}[h!]
 \begin{algorithmic}
@@
 \Procedure{auto-involutive-mcmc}{$\mathcal{P}$, $\mathcal{Q}$, $\mathcal{F}$, $x$}
-    \State $(y, \log q_x(y)) \gets \textproc{trace-and-score}(\mathcal{Q}_x)$ \Comment{Sample $y \sim q_x(\cdot)$; compute log-density via probabilistic program}
-    \State $(x' \oplus y', \log D) \gets \textproc{run-involution}(\mathcal{F}, x \oplus y)$ \Comment{{\small Compute $(x' \oplus y') = f(x \oplus y)$ and log(Radon-Nikodym derivative)}}
-    \State $\log \tilde{p}(x \oplus b) \gets \textproc{score}(\mathcal{P}, x \oplus b)$ \Comment{Compute log-density via probabilistic program}
-    \State $\log \tilde{p}(x' \oplus b) \gets \textproc{score}(\mathcal{P}, x' \oplus b)$ \Comment{Compute log-density via probabilistic program}
-    \State $\log q_{x'}(y') \gets \textproc{score}(\mathcal{Q}_{x'}, y')$ \Comment{Compute log-density via probabilistic program}
-    \State $\alpha \gets \min\{1, \exp(\log \tilde{p}(x' \oplus b) - \log \tilde{p}(x \oplus b) + \log q_{x'}(y') - \log q_x(y) + \log D)\}$ \Comment{Acceptance probability}
+    \State\vspace{-2mm}
+    \LineComment{Sample $y \sim q_x(\cdot)$ and compute log-density via probabilistic program $\mathcal{Q}$}
+    \State $(y, \log q_x(y)) \gets \textproc{trace-and-score}(\mathcal{Q}_x)$
+    \State\vspace{-2mm}
+    \LineComment{{\small Compute $(x' \oplus y') = f(x \oplus y)$ and log(Radon-Nikodym derivative) via differentiable program $\mathcal{F}$}}
+    \State $(x' \oplus y', \log D) \gets \textproc{run-involution}(\mathcal{F}, x \oplus y)$
+    \State\vspace{-2mm}
+    \LineComment{Compute log-density via probabilistic program $\mathcal{P}$}
+    \State $\log \tilde{p}(x \oplus b) \gets \textproc{score}(\mathcal{P}, x \oplus b)$
+    \State\vspace{-2mm}
+    \LineComment{Compute log-density via probabilistic program $\mathcal{P}$}
+    \State $\log \tilde{p}(x' \oplus b) \gets \textproc{score}(\mathcal{P}, x' \oplus b)$ 
+    \State\vspace{-2mm}
+    \LineComment{Compute log-density via probabilistic program $\mathcal{Q}$}
+    \State $\log q_{x'}(y') \gets \textproc{score}(\mathcal{Q}_{x'}, y')$ 
+    \State\vspace{-2mm}
+    \LineComment{Compute acceptance probability and accept or reject}
+    \State $\alpha \gets \min\{1, \exp(\log \tilde{p}(x' \oplus b) - \log \tilde{p}(x \oplus b) + \log q_{x'}(y') - \log q_x(y) + \log D)\}$ 
     \State \textbf{with probability} \,$\alpha$ \,\Return $x'$ \algorithmicelse \, \Return $x$  \algorithmicend
@@
 
+\section{AUTOMATING INVOLUTIVE MCMC WITH TRACES} \label{sec:automating}
+
 Involutive MCMC is a general framework that can be used to develop diverse MCMC algorithms
@@
 \begin{tabular}{c}
-{\begin{lstlisting}[basicstyle=\footnotesize\ttfamily]
+\begin{lrbox}{\mybox}%
+\begin{lstlisting}[basicstyle=\small\ttfamily]
 @transform f (model_in, aux_in) to (model_out, aux_out)
@@
 end
-\end{lstlisting}}
+\end{lstlisting}
+\end{lrbox}%
+\scalebox{0.94}{\usebox{\mybox}}
 \end{tabular}

### 2020-07-19 paper.tex
add acks

@@
 
-%\subsubsection*{Acknowledgements}
+\subsubsection*{Acknowledgements}
+This research was supported in part by the US Department of Defense through the the National Defense Science \& Engineering Graduate Fellowship (NDSEG) Program, the DARPA SD2 program (contract FA8750-17-C-0239), the DARPA Machine Common Sense (MCS) program, the DARPA Synergistic Discovery and Design (SD2) program, support from the Intel Corporation, and a philanthropic gift from the Aphorism Foundation.
+The authors would also like to thank Feras Saad and Cameron Freer for helpful discussions.
 

### 2020-07-19 build_for_arxiv.sh
add build for arxiv script

@@ -1,0 +1,19 @@
+#!/bin/sh
+#set -e
+set -u
+
+
+# make sure bbl file exists
+pdflatex paper
+bibtex paper
+pdflatex paper
+pdflatex paper
+
+# put into tar
+if [ -d "for_arxiv" ] 
+then
+    rm -r for_arxiv
+fi
+mkdir for_arxiv
+cp -r paper.tex paper.bbl aistats2020.sty fancyhdr.sty genlisting.tex figures for_arxiv
+tar -cvzf for_arxiv.tar.gz for_arxiv

### 2020-07-20 paper.tex
add authors
use split merge example for involution lang

@@
 
-\usepackage{aistats2020}
+\usepackage[accepted]{aistats2020}
 % If your paper is accepted, change the options for the package
@@
 \vspace{1mm}
-\Procedure{auto-involutive-mcmc}{$\mathcal{P}$, $\mathcal{Q}$, $\mathcal{F}$, $x$}
+\Procedure{auto-involutive-mcmc}{$\mathcal{P}$, $\mathcal{Q}$, $\mathcal{F}$, $x$, $b$}
     \State\vspace{-2mm}
@@
     \State\vspace{-2mm}
-    \LineComment{{\small Compute $(x' \oplus y') = f(x \oplus y)$ and log(Radon-Nikodym derivative) via differentiable program $\mathcal{F}$}}
-    \State $(x' \oplus y', \log D) \gets \textproc{run-involution}(\mathcal{F}, x \oplus y)$
+    \LineComment{{\small Compute $(x' \oplus y') = f(x \oplus y)$ and log(Radon-Nikodym derivative) via differentiable program $\mathcal{F}_b$}}
+    \State $(x' \oplus y', \log D) \gets \textproc{run-involution}(\mathcal{F}_b, x \oplus y)$
     \State\vspace{-2mm}
@@
 \vspace{1mm}
-\Procedure{run-involution}{$\mathcal{F}$, $z$}
+\Procedure{run-involution}{$\mathcal{F}_b$, $z$}
 % TODO would be much simpler to describe if we didn't have to separate out x and y..
@@
 \paragraph{Constructing an Involution}
-The following code reads the values of parameters from the model in polar coordinates and writes out their values in cartesian coordinates:
+Consider the following generative model, which posits that a vector of univariate data is either generated from a single normal distribution or a mixture of two normal distributions.
+The model is expressed as a probabilistic program:
 \noindent
@@
 \begin{tabular}{c}
-{\begin{lstlisting}[basicstyle=\small\ttfamily]
-r = @read(model_in[:r], :continuous)
-theta = @read(model_in[:theta], :continuous)
-x = r * sin(theta)
-y = r * cos(theta)
-@write((model_out[:x], x, :continuous)
-@write((model_out[:y], y, :continuous)
-\end{lstlisting}}
-\end{tabular}
-\end{center}
-
-This code snippet defines a bijection from one coordinate system to another.
-However, it does not define an involution, because applying the function the first time produces a trace containing cartesian coordinates, and applying the function a second time to the resulting trace would fail, because the function reads in polar coordinates from its input trace.
-How can we make the function an involution?
-First, the function will need the ability to go from cartesian coordinates to polar coordinates, so the code for the inverse of the forward transformation must appear in the function somewhere.
-Also, to be an involution, the input and output traces must contain some information indicating whether they are using polar or cartesian coordinates, so the involution code knows which of the two transformations to read from, and which addresses to read from and write to.
-Below is a minimal example of the involution, that uses a Boolean address polar.
-Note that the involution negates the value of this address, to indicate that the coordinate system is switching with each application.
-\noindent
-\begin{center}
-\begin{tabular}{c}
-{\begin{lstlisting}[basicstyle=\small\ttfamily]
-polar = @read(model_in[:polar], :discrete)
-if polar
-   r = @read(model_in[:r]), :continuous)
-   theta = @read(model_in[:theta], :continuous)
-   x = r * sin(theta)
-   y = r * cos(theta)
-   @write(model_out[:x], x, :continuous)
-   @write(model_out[:y], y, :continuous)
-else
-   x = @read(model_in[:x], :continuous)
-   y = @read(model_in[:y], :continuous)
-   r = (x^2 + y^2)
-   theta = atan(y, x)
-   @write(model_out[:r], r, :continuous)
-   @write(model_out[:theta], theta, :continuous)
-end
-@write(model_out[:polar], !polar, :discrete)
-\end{lstlisting}}
-\end{tabular}
-\end{center}
-In order to apply this function in the context of involutive MCMC, we require probabilistic programs $\mathcal{P}$ and $\mathcal{Q}$ that define space of input and output traces for the involution.
-The involution only reads and writes to model addresses, so $\mathcal{Q}$ can be an empty program:
-\noindent
-\begin{center}
-\begin{tabular}{c}
-{\begin{lstlisting}[basicstyle=\small\ttfamily]
+\begin{lrbox}{\mybox}%
+\begin{lstlisting}[basicstyle=\small\ttfamily]
 @gen function p()
-    polar ~ bernoulli(0.5)
-    if polar
-        r ~ gamma(1, 1)
-        theta ~ uniform(-pi, pi)
-    else
-        x ~ normal(0, 1)
-        y ~ normal(0, 1)
+  k ~ uniform_discrete(1, 2)
+  means = [{(:mu, j)} ~ normal(0, 10) for j in 1:k]
+  weights = ones(k)/k
+  vars = ones(k)
+  for i in 1:100
+    {(:x, i)} ~ mixture_of_normals(weights, means, vars)
   end
 end
-
-@gen function q(x)
+\end{lstlisting}
+\end{lrbox}%
+\scalebox{0.90}{\usebox{\mybox}}
+\end{tabular}
+\end{center}
+This is a simplified version of the unfinite Gaussian mixture model in Figure~\ref{fig:mixture}b.
+Each key of the form $(\mathtt{:}\mathtt{x}, i)$ will be observed (in the dictionary $b$), and the latent part of the trace (the dictionary $x$) contains the other keys $\mathtt{:}\mathtt{k}$ and $(\mathtt{:}\mathtt{mu}, 1)$ and (only if $\mathtt{k} = 2$) $(\mathtt{:}\mathtt{mu}, 2)$.
+
+We will now walk through how to express a simple `split-merge' move~\citep{richardson1997bayesian} using our differentiable programming language.
+Suppose we want to construct an involutive MCMC kernel that changes $\mathtt{k}$ from $2$ to $1$ or vice-versa.
+The discrete part of the involution is straightforward:
+\noindent
+\begin{center}
+\begin{tabular}{c}
+\begin{lrbox}{\mybox}%
+\begin{lstlisting}[basicstyle=\small\ttfamily]
+k = @read(model_in["k"], discrete)
+if k == 1
+    @write(model_out["k"], 2, discrete)
+else
+    @write(model_out["k"], 1, discrete)
 end
-\end{lstlisting}}
-\end{tabular}
-\end{center}
-
-Note that in general there are observations $b$ that induce the target density $p(x)$ via $p(x) \propto \tilde{p}(x \oplus b)$ as described in Section~\ref{sec:dists-on-dicts}, and the involution $f$ in general depends on $b$.
-The involution program $\mathcal{F}$ can read the values of the $b$, which are stored alongside the unobserved random choices in the input model trace.
+\end{lstlisting}
+\end{lrbox}%
+\scalebox{1.0}{\usebox{\mybox}}
+\end{tabular}
+\end{center}
+Writing the involution code for the continuous choices is more complex.
+There is no one-to-one correspondence between the space of dictionaries with keys $\{\mathtt{:}\mathtt{k}, (\mathtt{:}\mathtt{mu}, 1)\}$ and the space of dictionaries $x$ with keys $\{\mathtt{:}\mathtt{k}, (\mathtt{:}\mathtt{mu}, 1), (\mathtt{:}\mathtt{mu}, 2)\}$.
+Therefore, we need to extend the state space using the auxiliary distribution, defined with the following probabilistic program:
+\noindent
+\begin{center}
+\begin{tabular}{c}
+\begin{lrbox}{\mybox}%
+\begin{lstlisting}[basicstyle=\small\ttfamily]
+@gen function q(model_trace)
+    if model_trace["k"] == 1
+        # we are doing a split, sample extra DoF
+        u ~ beta(2, 2)
+    end
+end
+\end{lstlisting}
+\end{lrbox}%
+\scalebox{1.00}{\usebox{\mybox}}
+\end{tabular}
+\end{center}
+Our involution is on the space of combined dictionaries $x \oplus y$ where $y$ are traces of $\mathtt{q}$.
+Note that for each $(x \oplus y)$ where $x[\mathtt{k}] = 1$, $y$ has a key $\mathtt{u}$, and for each $(x \oplus y)$ where $x[\mathtt{k}] = 2$, $y$ is empty.
+We extend the involution using a pair of bijections between $(\mu_1, \mu_2)$ and $(\mu_1, u)$:
+\[
+\mu_1, \mu_2 \mapsto ((\mu_1 + \mu_2) / 2, \mu_2 - (\mu_1 + \mu_2)/2)
+\]
+\[
+\mu_1, u \mapsto (\mu_1 - u, \mu_1 + u)
+\]
+The value of $x[\mathtt{k}]$ determines which of these functions is executed.
+The full involution program $\mathcal{F}$ is then:
+\noindent
+\begin{center}
+\begin{tabular}{c}
+\begin{lrbox}{\mybox}%
+\begin{lstlisting}[basicstyle=\small\ttfamily]
+@transform f (model_in,aux_in) to (model_out,aux_out)
+begin
+    k = @read(model_in["k"], discrete)
+    if k == 1
+        @write(model_out["k"], 2, discrete)
+        mu = @read(model_in[("mu", 1)], continuous)
+        u = @read(aux_in["u"], continuous)
+        mu1 = mu - u; mu2 = mu + u
+        @write(model_out[("mu", 1)], mu1, continuous)
+        @write(model_out[("mu", 2)], mu2, continuous)
+    else
+        @write(model_out["k"], 1, discrete)
+        mu1 = @read(model_in[("mu", 1)], continuous)
+        mu2 = @read(model_in[("mu", 2)], continuous)
+        mu = (mu2 + mu1) / 2
+        u = mu2 - mu
+        @write(model_out[("mu", 1)], mu, continuous)
+        @write(aux_out["u"], u, continuous)
+    end
+end
+\end{lstlisting}
+\end{lrbox}%
+\scalebox{0.90}{\usebox{\mybox}}
+\end{tabular}
+\end{center}
+The two continuous bijections (implemented in blocks of code in the two branches) are inverses of one another.
+This is a common pattern in involution programs---an involution on the discrete parts of the traces (in this case, just $x[\mathtt{k}]$) determines via control flow which continuous code blocks get executed, such that the end-to-end program defines an involution.
+
+Note that the observations $b$ (in this case, the x-coordinates) are included in traces of $\mathtt{p}$, which take the form $x \oplus b$ where $x$ is the latent part and $b$ is the observed part.
+The involution program $\mathcal{F}$ is allowed to read from the the observations in the model trace using the same syntax as used to read from the latent part of the trace $x$.
+To denote this dependency we use notation $\mathcal{F}_b$.
+(The example given here does not utilize this feature).
 

### 2020-07-20 paper.tex
updates to sparsity section

@@
 
-Suppose $N$ and $N'$ are the total number of random choices in the input traces $x, y$ and output traces $x', y'$ respectively, and $M$ is the number of continuous random choices (which must be the same in the input traces and output traces).
+Suppose $N$ and $N'$ are the total number of random choices in the input traces $x, y$ and output traces $x', y'$ respectively
+(note that the observations $b$ are excluded---$x$ and $x'$ are the \emph{latent} part of the model's traces only).
+Let $M$ be the number of continuous random choices among these (which must be the same in the input traces and output traces).
 Then, the number of operations used in Algorithm~\ref{alg:auto-involutive-mcmc} grows as $O(N + N') + O(M^3)$.
@@
 \includegraphics[width=\linewidth]{figures/sparse-jacobian.pdf}
+Here, the move is splitting cluster 2 into two clusters (cluster 2 and cluster 4).
 Black squares indicate nonzero entries.
-In this case, the size $N$ of the trace grows linearly both in the number of clusters and the number of data points, but the Jacobian determinant can be calculated from only the $6$-by-$6$ submatrix of the Jacobian that involves the parameters of the one cluster being split (or the two clusters being merged).
+In this case, the size $N$ of the latent part of the trace grows linearly in the number of clusters, but the Jacobian determinant can be calculated from only the $6$-by-$6$ submatrix of the Jacobian that involves the parameters of the one cluster being split (or the two clusters being merged).
 Therefore, the acceptance ratio computation reduces from $O(N + N') + O(M^3)$ to $O(N + N')$.
@@
 
-Consider input traces $(x, y)$ and output traces $(x', y')$ with $f(x, y) = (x', y')$ for involution $f$ encoded by program $\mathcal{F}$.
+Consider input traces $(x, y)$ and output traces $(x', y')$ with $f(x, y) = (x', y')$ for involution $f$ encoded by program $\mathcal{F}$ (again, $x$ and $x'$ include only the \emph{latent} part of the model trace and not the observed part $b$).
 Let $N_1 := |K_{x}|$, $N_2 := |K_{y}|$, $N_1' := |K_{x'}|$ and $N_2' := |K_{y'}|$, and
 $N := N_1 + N_2$ and $N' := N_1' + N_2'$.
-Suppose that $N_1 + N_1' \gg N_2 + N_2'$, which occurs when the size of the model traces grows in the size of a data set, but the size of the auxiliary trace is constant.
+Suppose that $N_1 + N_1' \gg N_2 + N_2'$, which often occurs when the the number of latent variables is large, but the involutive MCMC move only updates a portion of the latent variables.
 Algorithm~\ref{alg:auto-involutive-mcmc} runs $\mathcal{F}$, which explicitly writes the value for each element of $(x', y')$, requiring $N' = N_1' + N_2' \approx N_1'$ `write' or `copy' operations.
@@
 We replace the separate calls to \textproc{score} for the log-densities $\log \tilde{p}(x \oplus b)$ and $\log \tilde{p}(x' \oplus b)$ and replace them with a single call to \emph{trace-update} on $\mathcal{P}$ that takes in $x$ and $\delta$ and returns $x'$ and $\log (\tilde{p}(x' \oplus b) / \tilde{p}(x \oplus b))$.
-When $|K_{\delta}|$ and $N_2$ and $N_2'$ are all constant in the size $N_1$ of the model trace $x$, the resulting algorithm uses a number of operations that does not grow with $N_1 + N_1'$.
+When $|K_{\delta}|$ and $N_2$ and $N_2'$ are all constant in the size $N_1$ of the latent part of model trace ($x$), the resulting algorithm uses a number of operations that does not grow with $N_1 + N_1'$.
 % TODO need to differentiate between number of data and latent dimension here!

### 2020-07-20 paper.tex
mvoe figures up
tweaks

@@
 
+\begin{figure*}[ht!]
+    \centering
+    \includegraphics[width=0.93\textwidth]{figures/mixture.pdf}
+    \caption{
+Example of reversible jump MCMC~\citep{green1995reversible} implemented using involutive MCMC in Gen.
+The example implements a `split-merge move' in a infinite Gaussian mixture model~\citep{richardson1997bayesian} using three Gen programs:
+(1) a probabilistic program $\mathtt{p}$ encoding the generative model (shown in b),
+(2) a probabilistic program $\mathtt{q}$ encoding an auxiliary probability distribution (shown in c),
+and (3) a differentiable program $\mathtt{f}$ that encodes an involution on the space of pairs of traces of $\mathtt{p}$ and $\mathtt{q}$ (shown in d).
+Gen's involutive MCMC operator (shown in e) automatically computes the acceptance probability.
+}
+    \label{fig:mixture}
+\end{figure*}
+
+\begin{figure*}[ht!]
+    \centering
+    \includegraphics[width=0.95\textwidth]{figures/structure-learning.pdf}
+    \caption{
+A mixture kernel implemented using involutive MCMC in Gen, applied to infer the covariance function of a Gaussian process.
+The prior on covariance functions is based on a probabilistic context-free grammar.
+Each component kernel in the mixture replaces a subtree of the covariance function parse tree with a new subtree.
+The mixture kernel chooses a random subtree to replace via a random walk on the parse tree.
+The mixture kernel is composed from three Gen programs:
+(1) a probabilistic program $\mathtt{p}$ encoding the generative model (shown in a),
+(2) a probabilistic program $\mathtt{q}$ encoding an auxiliary probability distribution (shown in b), and
+(3) a differentiable program $\mathtt{f}$ that encodes an involution (shown in d).
+}
+    \label{fig:structure-learning}
+\end{figure*}
+
+\clearpage
 \section{INVOLUTIVE MCMC ON GENERAL STATE SPACES} % on traces ??
@@
 The conditional density $p(d | b)$ is nonzero only if $K_d = \{\texttt{k}\} \cup \{\texttt{m}_i \mid 1 \leq i \leq k\} \cup \{\texttt{s}_i \mid 1 \leq i \leq k\}$ for some $k$ ($d$ does not contain y-values), and the denominator in the definition of $p(d | b)$ simplifies to the familiar sum of marginal likelihoods over all $k$, where each marginal likelihood is a Riemann integral over $\mathbb{R}^{2k}$.
-
-\begin{figure*}[h!]
-    \centering
-    \includegraphics[width=0.93\textwidth]{figures/mixture.pdf}
-    \caption{
-Example of reversible jump MCMC~\citep{green1995reversible} implemented using involutive MCMC in Gen.
-The example implements a `split-merge move' in a infinite Gaussian mixture model~\citep{richardson1997bayesian} using three Gen programs:
-(1) a probabilistic program $\mathtt{p}$ encoding the generative model (shown in b),
-(2) a probabilistic program $\mathtt{q}$ encoding an auxiliary probability distribution (shown in c),
-and (3) a differentiable program $\mathtt{f}$ that encodes an involution on the space of pairs of traces of $\mathtt{p}$ and $\mathtt{q}$ (shown in d).
-Gen's involutive MCMC operator (shown in e) automatically computes the acceptance probability.
-}
-    \label{fig:mixture}
-\end{figure*}
 
@@
 The details of these more efficient implementations are outside the scope of this paper, but are implemented as part of the Gen system.
-% TODO be more precise:
-(As a simple example, consider a density $p$ that factors into a product of independent densities, one for each random choice; and an involution that only modifies one of these choices).
 
@@
 Instead of explicitly specifying a value for all keys in $x'$, it can specify $x'$ by only explicitly specifying values in $\delta$.
+(The other values in $x'$ are \emph{implicit copies} via \emph{trace-update}---they could be explicitly copied within $\mathcal{F}$, but this would result in uncessary code and possibly unecessary computation.)
 Therefore, the involution program can run in $O(|K_{\delta}| + N_2 + N_2')$ operations.
@@
 When $|K_{\delta}|$ and $N_2$ and $N_2'$ are all constant in the size $N_1$ of the latent part of model trace ($x$), the resulting algorithm uses a number of operations that does not grow with $N_1 + N_1'$.
-% TODO need to differentiate between number of data and latent dimension here!
-% it is only constant in the latent dimension, not in the data.. (except in unusual cases)
 
@@
 
-\subsection{State-Dependent Mixture Proposal}
+\subsection{State-Dependent Mixture Proposals}
 
@@
 In both cases, our automated involutive MCMC algorithm automatically computes the acceptance probability.
-
-\begin{figure*}[ht]
-    \centering
-    \includegraphics[width=0.95\textwidth]{figures/structure-learning.pdf}
-    \caption{
-A mixture kernel implemented using involutive MCMC in Gen, applied to infer the covariance function of a Gaussian process.
-The prior on covariance functions is based on a probabilistic context-free grammar.
-Each component kernel in the mixture replaces a subtree of the covariance function parse tree with a new subtree.
-The mixture kernel chooses a random subtree to replace via a random walk on the parse tree.
-The mixture kernel is composed from three Gen programs:
-(1) a probabilistic program $\mathtt{p}$ encoding the generative model (shown in a),
-(2) a probabilistic program $\mathtt{q}$ encoding an auxiliary probability distribution (shown in b), and
-(3) a differentiable program $\mathtt{f}$ that encodes an involution (shown in d).
-}
-    \label{fig:structure-learning}
-\end{figure*}
 

### 2020-07-20 paper.tex
tweaks

@@
 dictionaries) that formalizes the space of execution traces of Gen probabilistic programs, and a formulation of involutive MCMC on these spaces.
-\item A differentiable programming language for defining bijections between spaces of traces of probabilistic programs.
+\item A differentiable programming language for defining transformations between spaces of traces of probabilistic programs.
 \item An algorithm that automates the implementation of an involutive MCMC kernel given two probabilistic programs and a differentiable program encoding the involution, using automatic differentiation and tracing of probabilistic programs.
@@
 \citet{tierney1998note} described a family of deterministic proposals based on a deterministic involution that is equivalent to involutive MCMC but without the auxiliary probability distribution.
-More recently,~\citet{spanbauer2020deep} defined a class of deep generative model based on differentiable involutions and trained these models to serve as efficient proposal distributions on continuous state spaces; the resulting algorithm is an instance of the construction presented in this paper.
+More recently,~\citet{spanbauer2020deep} defined a class of deep generative models based on differentiable involutions and trained these models to serve as efficient proposal distributions on continuous state spaces; the resulting algorithm is an instance of the construction presented in this paper.
 
@@
     \LineComment{{\small Compute $(x' \oplus y') = f(x \oplus y)$ and log(Radon-Nikodym derivative) via differentiable program $\mathcal{F}_b$}}
-    \State $(x' \oplus y', \log D) \gets \textproc{run-involution}(\mathcal{F}_b, x \oplus y)$
+    \State $(x' \oplus y', \log D) \gets \textproc{run-involution}(\mathcal{F}, x \oplus y, b)$
     \State\vspace{-2mm}
@@
 \vspace{1mm}
-\Procedure{run-involution}{$\mathcal{F}_b$, $z$}
-% TODO would be much simpler to describe if we didn't have to separate out x and y..
+\Procedure{run-involution}{$\mathcal{F}$, $z$, $b$}
     \LineComment{Run the involution, and keep track of copied, read, and written continuous keys}
@@
         \State $\mathtt{"@copy}(k_1, k_2)\mathtt{"} \equiv$ (set $z'[k_2] \gets z[k_1]$ and set $C \gets C \cup \{k_1\}$)
+        \State (and reading from observations $b$ for any $k \in K_b$)
     \END
@@
 with a single new construct, the `$\{\mathtt{address}\} \sim \mathtt{distribution}$' expression, for making a \textit{named} random choice.
-An execution trace of a Gen probabilistic program can be sampled by running the program according to Julia's usual semantics, and upon encountering an expression of the form `$\{\mathtt{address}\} \sim \mathtt{distribution}$', (i) evaluating the \textit{address expression} (`$\mathtt{address}$') to obtain an address $k \in \mathcal{K}$; (ii) evaluating the \textit{distribution expression} (`$\mathtt{distribution}$') to obtain a probability distribution over the measurable space $V_k$; (iii) sampling a value $x_k$ from this distribution and adding the mapping $\{k \mapsto x_k\}$ to a dictionary called the program trace; and (iv) returning the sampled value $x_k$ to the program, to continue execution.
-When execution terminates, the program trace has accumulated a mapping for each random choice encountered during the program's execution.
+An execution trace of a Gen probabilistic program is a dictionary that can be sampled by running the program according to Julia's usual semantics, and upon encountering an expression of the form `$\{\mathtt{address}\} \sim \mathtt{distribution}$', (i) evaluating the \textit{address expression} (`$\mathtt{address}$') to obtain an address $k \in \mathcal{K}$; (ii) evaluating the \textit{distribution expression} (`$\mathtt{distribution}$') to obtain a probability distribution over the measurable space $V_k$; (iii) sampling a value $x_k$ from this distribution and adding the mapping $\{k \mapsto x_k\}$ to the execution trace; and (iv) returning the sampled value $x_k$ to the program, to continue execution.
+When execution terminates, the execution trace has accumulated a mapping for each random choice encountered during the program's execution.
 
@@
 Typically, we cannot use a probabilistic program $\mathcal{P}$ to represent the target distribution directly: probabilistic programs implement simulators for a distribution, but target densities are typically not tractable to simulate (hence the need for MCMC).
-Instead, we may wish to sample from a target $p$ that arises from conditioning a probabilistic program $\mathcal{P}$'s distribution over traces on observations of the values at some addresses.
+Instead, we may wish to sample from a target $p$ that arises from \emph{conditioning} a probabilistic program $\mathcal{P}$'s distribution over traces on observations of the values at some addresses.
 Let $\tilde{p}$ denote $\mathcal{P}$'s distribution over traces, and let $b = (B, \textbf{b})$ be a dictionary of observations, as described in Section~\ref{sec:dists-on-dicts}.
@@
 
-Section~\ref{sec:sparsity} describes a more efficient approach that exploits sparsity in the involution (and the resulting cancellations in the acceptance ratio) for improved efficiency, but requires a more sophisticated probabilistic programming runtime system.
+Section~\ref{sec:sparsity} describes a more efficient approach that exploits sparsity in the involution and cancellations in the acceptance ratio for improved efficiency, but requires a more sophisticated probabilistic programming runtime system.
 
@@
 Section~\ref{sec:probabilistic-programming} showed that if the densities $p$ and $q$ are specified using probabilistic programs $\mathcal{P}$ and $\mathcal{Q}$, then the density ratio in the acceptance probability for involutive MCMC on dictionaries (Equation~(\ref{eq:acceptance-ratio-dictionaries})) can be automated using probabilistic programming techniques.
-This section shows that if the involution $f$ is specified in using a differentiable program $\mathcal{F}$ that transforms the execution traces of probabilistic programs, then the Jacobian factor in Equation~(\ref{eq:acceptance-ratio-dictionaries}) can also be automated, using automatic differentiation.
+This section shows that if the involution $f$ is specified using a differentiable program $\mathcal{F}$ that transforms the execution traces of probabilistic programs, then the Jacobian factor in Equation~(\ref{eq:acceptance-ratio-dictionaries}) can also be automated, using automatic differentiation.
 The procedure \textproc{auto-involutive-mcmc} in Algorithm~\ref{alg:auto-involutive-mcmc} combines these two ideas and automates involutive MCMC given the programs $\mathcal{P}$, $\mathcal{Q}$, and $\mathcal{F}$.
@@
 The syntax for address $\mathtt{a}$ within trace $\mathtt{trace}$ is $\mathtt{trace[a]}$.
-For example, to read the value of a continuous address $\mathtt{:}\mathtt{r}$ from the input model trace ($x$):
+For example, to read the value of a continuous address $\mathtt{:}\mathtt{a}$ from the input model trace ($x$):
 \noindent
@@
 {\begin{lstlisting}[basicstyle=\small\ttfamily]
-    val = @read(model_in[:r], :continuous)
+val = @read(model_in[:a], :continuous)
 \end{lstlisting}}
@@
 The syntax for writing to an address in $x' \oplus y' \in Z$ is similar.
-For example to write a value $\mathtt{val}$ to $x'[\mathtt{:}\mathtt{r}]$:
+For example to write a value $\mathtt{val}$ to $x'[\mathtt{:}\mathtt{a}]$:
 \noindent
@@
 {\begin{lstlisting}[basicstyle=\small\ttfamily]
-    @write(model_in[:r], val, :continuous)
+@write(model_out[:a], val, :continuous)
 \end{lstlisting}}
@@
 {\begin{lstlisting}[basicstyle=\small\ttfamily]
-    @copy(<from-address>, <to-address>)
+@copy(<source-address>, <destination-address>)
 \end{lstlisting}}
@@
 {\begin{lstlisting}[basicstyle=\small\ttfamily]
-    @copy(model_in[:u], model_in[:v])
+@copy(model_in[:u], model_out[:v])
 \end{lstlisting}}
@@
 Of course, it is also possible to copy from $x$ to $y'$, from $y$ to $x'$ and from $y$ to $y'$.
-As we will see in Section~\ref{sec:sparsity}, it is preferable to use $\mathtt{@copy}$ when possible instead of reading and then writing, as this can improve the efficiency of automatic differentiation.
+As we will see in Section~\ref{sec:sparsity}, it is preferable to use $\mathtt{@copy}$ when possible instead of reading and then writing, as this can make the acceptance probability calculation more efficient.
 
@@
 \end{center}
-This is a simplified version of the unfinite Gaussian mixture model in Figure~\ref{fig:mixture}b.
-Each key of the form $(\mathtt{:}\mathtt{x}, i)$ will be observed (in the dictionary $b$), and the latent part of the trace (the dictionary $x$) contains the other keys $\mathtt{:}\mathtt{k}$ and $(\mathtt{:}\mathtt{mu}, 1)$ and (only if $\mathtt{k} = 2$) $(\mathtt{:}\mathtt{mu}, 2)$.
+This is a simplified version of the infinite Gaussian mixture model in Figure~\ref{fig:mixture}b.
+Each key of the form $(\mathtt{:}\mathtt{x}, i)$ will be observed (in the dictionary $b$), so the latent part of the trace (the dictionary $x$) contains the other keys $\mathtt{:}\mathtt{k}$ and $(\mathtt{:}\mathtt{mu}, 1)$ and (when $\mathtt{k} = 2$) $(\mathtt{:}\mathtt{mu}, 2)$.
 
 We will now walk through how to express a simple `split-merge' move~\citep{richardson1997bayesian} using our differentiable programming language.
-Suppose we want to construct an involutive MCMC kernel that changes $\mathtt{k}$ from $2$ to $1$ or vice-versa.
+Suppose we want to construct an involutive MCMC kernel that changes $\mathtt{k}$ from $2$ to $1$ or vice versa.
 The discrete part of the involution is straightforward:
@@
 Note that for each $(x \oplus y)$ where $x[\mathtt{k}] = 1$, $y$ has a key $\mathtt{u}$, and for each $(x \oplus y)$ where $x[\mathtt{k}] = 2$, $y$ is empty.
-We extend the involution using a pair of bijections between $(\mu_1, \mu_2)$ and $(\mu_1, u)$:
-\[
-\mu_1, \mu_2 \mapsto ((\mu_1 + \mu_2) / 2, \mu_2 - (\mu_1 + \mu_2)/2)
-\]
-\[
-\mu_1, u \mapsto (\mu_1 - u, \mu_1 + u)
-\]
+We extend the involution using a pair of bijections between $(\mu_1, \mu_2)$ and $(\mu_1, u)$ that show how two cluster means should be transformed into one cluster mean (merge), and vice versa (split):
+\begin{equation} \label{eq:split-merge}
+\begin{array}{c}
+\mu_1, \mu_2 \mapsto ((\mu_1 + \mu_2) / 2, \mu_2 - (\mu_1 + \mu_2)/2) \;\; [\mathrm{Merge}]\\
+\mu_1, u \mapsto (\mu_1 - u, \mu_1 + u) \;\; [\mathrm{Split}]
+\end{array}
+\end{equation}
 The value of $x[\mathtt{k}]$ determines which of these functions is executed.
@@
         @write(model_out["k"], 2, discrete)
+
+        # split bijection
         mu = @read(model_in[("mu", 1)], continuous)
@@
         @write(model_out["k"], 1, discrete)
+
+        # merge bijection
         mu1 = @read(model_in[("mu", 1)], continuous)
@@
 The involution program $\mathcal{F}$ is allowed to read from the the observations in the model trace using the same syntax as used to read from the latent part of the trace $x$.
-To denote this dependency we use notation $\mathcal{F}_b$.
 (The example given here does not utilize this feature).
@@
 Each function $h_e$ is a function from the values at continuous addresses in the input trace ($z[k]$ for $k \in K_z \cap \mathcal{J}$) to the values at continuous addresses in the output trace ($z'[k]$ for $k \in K_{z'} \cap \mathcal{J}$).
+In the example above, the partition $E$ is given by the equivalence classes of the relation $(x_1, y_1) \sim (x_2, y_2) \iff x_1[\mathtt{k}] = x_2[\mathtt{k}]$ (there are two equivalence classes), $g$ maps the $k = 1$ class to the $k = 2$ class and vice versa, and
+the continuous bijections $h_e$ and $h_{g(e)}$ are the functions in Equation~(\ref{eq:split-merge}).
+
 We compute the Jacobian using automatic differentiation. (The Gen implementation uses forward-mode AD whereas our PyTorch implementation uses reverse-mode AD.)
@@
 For example, for the function $(u, v, x, y) \mapsto (u, 2 u - v, y, x) = (u', v', x', y')$, the Jacobian is (with columns corresponding to $u'$, $v'$, $x'$, and $y'$ and rows corresponding to $u$, $v$, $x$, and $y$):
-\begin{equation}
+\begin{equation*}
 \left[
@@
 \end{array}
-\end{equation}
+\end{equation*}
 Using the cofactor expansion of the determinant, we observe that for any `copy' column in an $M$-by-$M$ Jacobian matrix (a column with a single $1$ and all other entries $0$), the absolute value of the determinant is equivalent to that of the $(M-1)$-by-$(M-1)$ sub-matrix with the corresponding column \textit{and} row omitted (even if that would remove other nonzero entries from the matrix).
@@
 Indeed, if some input key is copied to some output key, then we can entirely avoid computing its row (and corresponding column) of the Jacobian.
-Therefore, the number of operations (which is dominated by the determinant) reduces from $M^3$ to $(M-|C|)^3$ where $|C|$ is the number of input keys that were copied to some output input key.
+Therefore, the number of operations (which is dominated by the determinant) reduces from $M^3$ to $(M-|C|)^3$ where $|C|$ is the number of input keys that were copied to some output key.
 The $\mathtt{@copy}$ statement in our differentiable programming language makes the set of input keys that were copied explicitly available to the interpreter, which makes automating this optimization straightforward, as shown in \textproc{run-involution} in Algorithm~\ref{alg:auto-involutive-mcmc}.
@@
 
-Note that $x$ and $x'$ may both have values for certain keys, and $x$ may contains keys not present in $x'$, and vice-versa.
+Note that $x$ and $x'$ may both have values for certain keys, and $x$ may contains keys not present in $x'$, and vice versa.
 Often, an involution does not modify the values of many keys in the trace---this is the case in the split-merge reversible jump move described earlier.
@@
 Let $\Delta := K_{x'} \setminus A$ be the set of keys in $x'$ that are either (i) not in $x$, or (ii) in $x$ but have their value changed.
-Then, the previous trace $x$ and a dictionary $\delta$ with $K_{\delta} = \Delta$ and $\delta[k] = x'[k]$ and the density function $p$ suffice to uniquely define the trace $x'$, provided the density satisfies the following condition, which is satisfied by densities as probabilistic programs:
+Then, the previous trace $x$ and a dictionary $\delta$ with $K_{\delta} = \Delta$ and $\delta[k] = x'[k]$ and the density function $p$ suffice to uniquely define the trace $x'$, provided the density satisfies the following condition, which is satisfied by densities defined by probabilistic programs\footnote{Intuitively, two traces of a probabilistic program cannot have different sets of addresses unless they share an address and disagree on its value.}:
 $p(x) > 0$ and $p(x') > 0$ and $x \ne x'$ implies there exists $k \in K_x \cap K_{x'}$ with $x[k] \ne x'[k]$.
@@
 modifying the requirement for the involution program $\mathcal{F}$:
-Instead of explicitly specifying a value for all keys in $x'$, it can specify $x'$ by only explicitly specifying values in $\delta$.
-(The other values in $x'$ are \emph{implicit copies} via \emph{trace-update}---they could be explicitly copied within $\mathcal{F}$, but this would result in uncessary code and possibly unecessary computation.)
+Instead of explicitly specifying a value for all keys in $x'$, it need only explicitly specify values in $\delta$.
+(We call the other values in $x'$ \emph{implicit copies}---they could be explicitly copied within $\mathcal{F}$, but this would result in uncessary code and possibly unecessary computation.)
 Therefore, the involution program can run in $O(|K_{\delta}| + N_2 + N_2')$ operations.
-We replace the separate calls to \textproc{score} for the log-densities $\log \tilde{p}(x \oplus b)$ and $\log \tilde{p}(x' \oplus b)$ and replace them with a single call to \emph{trace-update} on $\mathcal{P}$ that takes in $x$ and $\delta$ and returns $x'$ and $\log (\tilde{p}(x' \oplus b) / \tilde{p}(x \oplus b))$.
+We remove the separate calls to \textproc{score} for the log-densities $\log \tilde{p}(x \oplus b)$ and $\log \tilde{p}(x' \oplus b)$ and replace them with a single call to \emph{trace-update} on $\mathcal{P}$ that takes in $x$ and $\delta$ and returns $x'$ and $\log (\tilde{p}(x' \oplus b) / \tilde{p}(x \oplus b))$.
 When $|K_{\delta}|$ and $N_2$ and $N_2'$ are all constant in the size $N_1$ of the latent part of model trace ($x$), the resulting algorithm uses a number of operations that does not grow with $N_1 + N_1'$.
@@
 \Procedure{involutive-mcmc-check}{$\mathcal{P}$, $\mathcal{Q}$, $\mathcal{F}$}
+    \State \vspace{-2mm}
     \LineComment{Generate a randomized test case}
-    \State $(x \oplus b) \sim \textproc{trace-and-score}(\mathcal{P})$ % TODO not conditioned on data?
-    \State $y \sim \textproc{trace-and-score}(\mathcal{Q}_x)$
+    \State $((x \oplus b), \_) \sim \textproc{trace-and-score}(\mathcal{P})$ % TODO not conditioned on data?
+    \State $(y, \_) \sim \textproc{trace-and-score}(\mathcal{Q}_x)$
     %\State $x \sim p(\cdot)$ \Comment{Sample state from prior}
     %\State $y \sim q_x(\cdot)$ \Comment{Sample auxiliary variables}
-
+    \State \vspace{-2mm}
     \LineComment{Run involution and perform dimension check}
-    \State $x' \oplus y' \gets \textproc{run-involution}(\mathcal{F}, x \oplus y)$
-
+    \State $x' \oplus y' \gets \textproc{run-involution}(\mathcal{F}, x \oplus y, b)$
+    \State \vspace{-2mm}
     \LineComment{Support check}
@@
     \State {\bf assert} $(\log p(x') > -\infty) \land (\log q_{x'}(y') > -\infty)$
-
+    \State \vspace{-2mm}
     \LineComment{Involution check}
-    \State $\tilde{x} \oplus \tilde{y} \gets \textproc{run-involution}(\mathcal{F}, x' \oplus y')$
+    \State $\tilde{x} \oplus \tilde{y} \gets \textproc{run-involution}(\mathcal{F}, x' \oplus y', b)$
     \State ${\bf assert} (\tilde{x} = x) \land (\tilde{y} = y)$
@@
 
+\paragraph{Review of reversible jump MCMC.}
 The reversible jump MCMC framework involves a set of `models' $h \in \mathcal{H}$, and a prior distribution on models $p(h)$.
@@
 
+\paragraph{Encoding reversible jump in involutive MCMC.}
 To encode reversible jump MCMC in our framework, we write a probabilistic program $\mathcal{P}$ that encodes the space of models $\mathcal{H}$, the prior distribution on models, $p(h)$, the per-model priors $p_h(\theta_h)$ and the per-model likelihoods $L_{D,h}(\theta_h)$.
@@
 
+\paragraph{Example: Split-merge reversible jump.}
 Figure~\ref{fig:mixture} shows a split-merge reversible jump kernel for an infinite Gaussian mixture model~\citep{richardson1997bayesian} implemented using the probabilistic and differentiable programming languages described in this paper.
-Figure~\ref{fig:mixture}b shows the infinite Gaussian mixture model, specified as a probabilistic program $\mathcal{P}$.
+Figure~\ref{fig:mixture}b shows the infinite Gaussian mixture model, specified as a probabilistic program $\mathtt{p}$.
 The program takes the number of data points as input, then samples the number of clusters from a Poisson distribution, then samples cluster parameters and mixture proportions, and finally samples the data points from the resulting finite mixture.
-Figure~\ref{fig:mixture}c shows the auxiliary probabilistic program $\mathcal{Q}$ for the split-merge kernel.
+Figure~\ref{fig:mixture}c shows the auxiliary probabilistic program $\mathtt{q}$ for the split-merge kernel.
 This program takes a trace of the model program as input, and randomly decides whether to split a cluster and increase the number of clusters by one or merge two clusters and decrease the number of clusters by one.
@@
 
+\paragraph{Example: Bayesian structure learning for Gaussian processes}
 Figure~\ref{fig:structure-learning} shows automated involutive MCMC being applied to fully Bayesian inference over the covariance function of a Gaussian process, where the prior on covariance functions (Figure~\ref{fig:structure-learning}a) is based on a probabilistic context-free grammar.
@@
 
-At each iteration of the MCMC algorithm, the auxiliary probabilistic program $\mathcal{Q}$ (Figure~\ref{fig:structure-learning}d) first picks a random node in the parse tree of the covariance function, by doing a stochastic walk of the existing parse tree that terminates at the chosen node.
+\paragraph{Hierarchical address spaces.}
+This example uses probabilistic programs that invoke other probabilistic programs, sometimes recursively.
+For example, the model probabilistic program $\mathtt{p}$ invokes the probabilistic program $\mathtt{cov\_function\_prior}$, which is itself recursive.
+Similarly, the auxiliary probabilistic program $\mathtt{q}$ invokes $\mathtt{walk\_tree}$ (which is recursive) as well as $\mathtt{cov\_function\_prior}$.
+Consider the syntax used to recursively invoke $\mathtt{walk\_tree}$ within $\mathtt{walk\_tree}$:
+\begin{center}
+\begin{tabular}{c}
+{\begin{lstlisting}[basicstyle=\small\ttfamily]
+({:left} ~ walk_tree(node.left, path))
+\end{lstlisting}}
+\end{tabular}
+\end{center}
+This expression resembles a random choice expression.
+However, instead of associating the return value of the function $\mathtt{walk\_tree}$ with the address $\mathtt{:}\mathtt{left}$, the address $\mathtt{:}\mathtt{left}$ is associated with the entire trace of random choices made within $\mathtt{walk\_tree}$.
+That is, $\mathtt{:}\mathtt{left}$ is the \emph{namespace} for the addresses of all random choice made within the invocation.
+Further invocations by the callee themselves result in nested namespaces.
+This process results in a \emph{hierarchical address space} for random choices, as shown in Figure~\ref{fig:structure-learning}b.
+This does not modify the mathematical formalism---each random choice made during the execution of a probabilistic program still has a unique address, but the address has multiple components that localize it within the hierarchy.
+For example, the choice $\mathtt{done}$ a recursive call to $\mathtt{walk\_tree}$ might have address:
+\[
+k = (\mathtt{:}\mathtt{left} \mbox{ \texttt{=>} } \mathtt{:}\mathtt{right} \mbox{ \texttt{=>} } \mathtt{:}\mathtt{done})
+\]
+(In Gen, `\texttt{=>}' constructs hierarchical addresses.)
+
+\paragraph{A complex state-dependent distribution.}
+At each iteration of the MCMC algorithm, the auxiliary probabilistic program $\mathtt{q}$ (Figure~\ref{fig:structure-learning}d) first picks a random node in the parse tree of the covariance function, by doing a stochastic walk of the existing parse tree that terminates at the chosen node.
 \begin{center}
@@
 path ~ walk_tree(prev_cov_function, ..)
-\end{lstlisting}}
-\end{tabular}
-\end{center}
-The first part of the involution (Figure~\ref{fig:structure-learning}e) copies the random choices made during this walk from the input auxiliary trace to the output auxiliary trace.
-\begin{center}
-\begin{tabular}{c}
-{\begin{lstlisting}[basicstyle=\small\ttfamily]
-@copy(aux_in[:path], aux_out[:path])
 \end{lstlisting}}
@@
 \end{center}
-The other parts of the auxiliary probabilistic program and the involution program specifies a proposal distribution for the subtree of the parse tree that is rooted at the chosen node.
-In particular, the rest of the auxiliary probabilistic program $\mathcal{Q}$ proposes a new subtree by sampling from the same process used to recursively define the prior distribution:
+The first part of the involution (Figure~\ref{fig:structure-learning}e) copies the random choices made during this walk from the input auxiliary trace to the output auxiliary trace.
 \begin{center}
@@
 {\begin{lstlisting}[basicstyle=\small\ttfamily]
-new_subtree ~ cov_function_prior()
+@copy(aux_in[:path], aux_out[:path])
 \end{lstlisting}}
@@
 \end{center}
-The involution program swaps the old subtree with the newly proposed subtree:
-\begin{center}
-\begin{tabular}{c}
-\begin{lrbox}{\mybox}%
-\begin{lstlisting}[basicstyle=\small\ttfamily]
-@copy(model_in[subtree_address], aux_out[:new_subtree])
-@copy(aux_in[:new_subtree], model_out[subtree_address])
-\end{lstlisting}
-\end{lrbox}%
-\scalebox{0.9}{\usebox{\mybox}}
-\end{tabular}
-\end{center}
-This is an instance of a more general pattern for implementing state-dependent mixture proposals:
-\begin{enumerate}
-\item The auxiliary probabilistic program $\mathcal{Q}$ samples from a distribution over different sets of random choices that will be proposed to
-(in this case, each set is a different subtree of the parse tree).
-\item The auxiliary probabilistic program $\mathcal{Q}$ then samples new values for those random choices (in this case, a new subtree).
-\item The involution program $\mathcal{F}$ swaps the previous values of those random choices with their new values, by swapping data between the model trace and the auxiliary trace.
-\item The involution program $\mathcal{F}$ copies the random choices that determined what subset of random choices to propose to from the input auxiliary trace to the output auxiliary trace.
-\end{enumerate}
-
-Note that because the mixture distribution is specified using a probabilistic program, it is straightforward to modify the program $\mathcal{Q}$ to define a different mixture distribution.
+Note that here, $\mathtt{@copy}$ is being used to copy the entire \emph{set} of random choices from the namespace $\mathtt{:}\mathtt{path}$ in $\mathtt{aux\_in}$ to the namespace $\mathtt{:}\mathtt{path}$ in $\mathtt{aux\_out}$.
+
+Because the mixture distribution is specified using a probabilistic program, it is straightforward to modify the program $\mathtt{p}$ to define a different mixture distribution.
 The code below specifies a mixture distribution that is uniform over all nodes in the tree.
@@
 In both cases, our automated involutive MCMC algorithm automatically computes the acceptance probability.
+
+\paragraph{A general pattern for state-dependent mixtures of proposals in Metropolis-Hastings}
+The other parts of the auxiliary probabilistic program and the involution program specifies a proposal distribution for the subtree of the parse tree that is rooted at the chosen node.
+In particular, the rest of the auxiliary probabilistic program $\mathtt{q}$ proposes a new subtree by sampling from the same process used to recursively define the prior distribution:
+\begin{center}
+\begin{tabular}{c}
+{\begin{lstlisting}[basicstyle=\small\ttfamily]
+new_subtree ~ cov_function_prior()
+\end{lstlisting}}
+\end{tabular}
+\end{center}
+The involution program swaps the old subtree with the newly proposed subtree:
+\begin{center}
+\begin{tabular}{c}
+\begin{lrbox}{\mybox}%
+\begin{lstlisting}[basicstyle=\small\ttfamily]
+@copy(model_in[subtree_address], aux_out[:new_subtree])
+@copy(aux_in[:new_subtree], model_out[subtree_address])
+\end{lstlisting}
+\end{lrbox}%
+\scalebox{0.9}{\usebox{\mybox}}
+\end{tabular}
+\end{center}
+
+This is an instance of a more general pattern for implementing state-dependent mixture proposals:
+\begin{enumerate}
+\item The auxiliary probabilistic program $\mathcal{Q}$ samples from a distribution over different sets of random choices that will be proposed to
+(in this case, each set is a different subtree of the parse tree).
+\item The auxiliary probabilistic program $\mathcal{Q}$ then samples new values for those random choices (in this case, a new subtree).
+\item The involution program $\mathcal{F}$ swaps the previous values of those random choices with their new values, by swapping data between the model trace and the auxiliary trace.
+\item The involution program $\mathcal{F}$ copies the random choices that determined what subset of random choices to propose to from the input auxiliary trace to the output auxiliary trace.
+\end{enumerate}
 

### 2020-07-20 paper.tex
tweaks

@@
 end
-
+\end{lstlisting}}
+\end{tabular}
+\end{center}
+
+\begin{center}
+\begin{tabular}{c}
+{\begin{lstlisting}[basicstyle=\small\ttfamily]
 @transform h (tr1, tr2) to (tr3, tr4) begin
@@
 \]
-(In Gen, `\texttt{=>}' constructs hierarchical addresses.)
+(`\texttt{=>}' is the Gen syntax for constructing hierarchical addresses).
 
@@
 \end{center}
-The resulting distributions on selected nodes for two example trees are shown below:
+The resulting distributions on selected nodes for two possible input trees are shown below:
 \begin{center}
@@
 \end{center}
-The resulting distributions, for two input trees, are:
+The resulting distributions, for two possible input trees, are:
 \begin{center}

### 2020-07-21 paper.tex
Fix appendix labeling
Improve defn of inverse

@@
 including reversible jump MCMC~\citep{green1995reversible}, which is the dominant mathematical framework for MCMC on complex state spaces.
-Involutive MCMC constructs an MCMC kernel from three components: (i) the unnormalized target density, (ii) a sampler and density for an auxiliary probability distribution, and (iii) an involution\footnote{a bijection that is its own inverse ($f$ where $f(f(z)) = z$)} on an extended state space.
+Involutive MCMC constructs an MCMC kernel from three components: (i) the unnormalized target density, (ii) a sampler and density for an auxiliary probability distribution, and (iii) an involution\footnote{An \emph{involution} is a bijection that is its own inverse (that is, $f$ where $f(f(z)) = z$).} on an extended state space.
 While this construction is mathematically clarifying, correctly implementing involutive MCMC kernels on complex state spaces remains challenging due to tedious density and Jacobian computations and the need for careful reasoning about the state space.
@@
 \onecolumn
-\section*{APPENDIX}
+\appendix
+\section{APPENDIX}
 
